### 2019-04-09 paper/draft.tex
jr notes

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{amsmath}

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}

\begin{document}

\title{Identifying representations from neuroimaging data with covarying features}

\author{King, Jean-Remi\inst{1} \and Lopez-Paz, David \inst{1}}

\institute{Facebook Artificial Intelligence Research, Paris, France
\email{\{jeanremi,d.lopez\}@fb.com}}

\maketitle              % typeset the header of the contribution

\begin{abstract}
The abstract should briefly summarize the contents of the paper in
15--250 words.

\keywords{Neuroimaging \and Decoding \and Encoding \and Multivariate Pattern Analyses.}
\end{abstract}

\section{Introduction}
Sixty years ago, Hubel and Wiesel implanted an anesthetized cat with a microelectrode, and discovered that distinct neurons reliably spiked when a visual stimulus was presented with a specific orientation, while other neurons only spiked when the stimulus was oriented at a different angle. Since, a systematic mapping has been engaged in cognitive neuroscience to identify the contents encoded in brain activity. This quest has revealed that some neurons specifically responds to individual locations (place cell), whereas other respond to specific emotions () , or to specific individuals (). The mapping between neuronal activity and hypothetical features has been considerably accelerating with the development of functional Magnetic Resonance Imaginge (fMRI), eletro- and magneto-encephalography (E/MEG) as as well as with the accumulation of sensors on each microelectrode (neuropixel).

The abundance of data produced by such recording devices now necessitate the design of statistical models optimized to identify neural representations. Two distinct analytical frameworks have been developed to address this issue: decoding and encoding. Both are based on the premise that a feature is represented in neuronal activity if and only if such a feature can be linearly and instantly read from that neuronal activity. This premise, discussed elsewhere (ref), capitalizes on the idea that neurons can be approximated as detector of particular combination of excitatory and inhibitory inputs, and can thus "read" information under a linear constrain.

Encoding consists in predicting brain activity from a model. For example, an encoding model can be used to predict the brain responses to hand movements (ref), the identity of faces in a visual stimulus (ref), or to the nature of a syntactic operation (ref).
Decoding consists in the reverse prediction: i.e. it is optimized to predict a feature from a brain activity pattern.

In practice, a majority of fMRI studies are based on encoding analyses, whereas a majority of M/EEG studies are based on decoding analyses. This difference probably reflects the distinct limitations of these two types of neuroimaging.

Indeed, fMRI produces brain images of relatively high spatial resolution: typically, each voxel specifically record neuronal activity over 3 mm cube, by measuring their blood-oxygen-level dependent (BOLD) response. The corresponding haemodynamic function that characterizes the BOLD is non-trivially linked to neuronal activity (ref), but is generally approximated as a temporal convolution of the neuronal activity. Consequently, in fMRI, the primary difficulty consists in disentangling the factors that have contributed to the modulation of the BOLD response. Encoding analyses can thus be used to quantify the contribution of multiple (potentially overlapping) features.

By contrast, M/EEG record the magnetic and electric fields elicited by neuronal activity thanks via sensors distant from the brain. Unlike fMRI, the relationship between M/EEG signals and brain activity is both largely instantaneous and linear. However, M/EEG activity is a linear mixture of the brain activity: a unique brain area likely projects its activity onto multiple sensors, and a unique sensor likely records the activity of many brain areas. Furthermore, M/EEG sensors also pick up non-neuronal signals. In particular, the magnetic field of the earth, the presence of surrounding electronic devices, the heart, the electrically charged retina and their associated movements, as well as breathing, swallowing and jaw clenching can all elicit electro-magnetic signals orders of magnitude larger than those of our brain. Consequently, in M/EEG, the primary difficulty consists in detecting the specific activity that is associated with a model feature. Decoding analyses are thus used to discriminate, in this linear mixture of noise and signals.

The respective advantages of encoding and decoding come at a costs. By focusing on univariate brain responses, encoding analyses cannot efficiently capture multivariate patterns of brain activity. For example, if the heart modulates the baseline activity of two voxels, then it can be difficult to show with a univariate analysis that each voxel also correlates with the feature of a model (Fig. X). To compensate for such issue, it is thus common to (1) incorporate physiological features (e.g. heart and breathing activity) in the encoding model and to (2) do repeated measurements in order to correct the variance explained by the model by the variance explained by an identical condition. This type of correction is thus limited to (1) observable artifacts as well as to (2) experimental effects that are independent from memory.

By contrast, decoding analyses are fit to multivariate brain activity patterns, and can thus be more robust to the structured modulations induced by artefacts. However, by focusing on a single feature of a model (e.g. the presence of a face, the pitch of a tone etc), decoding cannot easily disentangle two covarying features (e.g. the contrast and the luminance of an image). To compensate for such an issue, it is thus common restrict the experimental protocols with synthetic stimuli whose features are orthogonal by design (e.g. images whose luminance and contrast are independently varied).

In the present study, we introduce a novel method to get the best of the two approaches. We show on synthetic data that our approach systematically outperforms encoding and decoding methods, both in its specificity and sensitivity. We provide a proof that formalizes the conditions within which our method will be optimal. Finally we provide an extension to non-linear but monotonic activation functions in order to potentially in order to capture the saturation effects the BOLD and population responses.

# RSA: in discussion
- only applies to categorical effects
- analyzed at the stimulus level, not the feature level

\section{First Section}

\begin{equation}
Y\leftarrow (XS+N)F
\end{equation}

where Y is MEG activity, X are input features, S is encoding matrix, N is source noise and F is forward operator

\subsection{LS from Y to X}

\begin{equation}
\begin{aligned}
\min_G \left \| X-YG \right \| &= \min_G \left \| X - (XSF+NF)G\right\| \\
&{}= \min_G \left \| X - XSFG\right\|  + \left \| NFG \right \| \\
\end{aligned}
\end{equation}

Given that :
\begin{equation}
\mathop{\mathbb{E}}[N]=0
\end{equation}

and
\begin{equation}
\left \| X-XSFG \right \| \leq \left \| X \right \| \cdot  \left \| I-SFG \right \| \\
\end{equation}

Then
\begin{equation}
\min_G \left \| X-YG \right \| = \min_G \left \| I -SFG\right\| \\
\Rightarrow \hat  G= (SF)^{\dagger}
\end{equation}

\subsection{LS from X to Y\^{G}}

\begin{equation}
\begin{aligned}
\min_H \left \| Y \hat G - XH \right \| &= \min_H \left \| (XSF+NF) \hat G - XH \right \| \\
&= \min_H \left \| XSF \hat G - XH \right \| \\
&= \min_H \left \| SF \hat G - H \right \| \\
&= \min_H \left \| SF (SF)^{\dagger} - H \right \| \\
\end{aligned}
\end{equation}

Under the assumptions that S is an orthogonal projection
\begin{equation}
    S = S^2, S=S^T, S^{\dagger} = S
\end{equation}

And under the extra assumptions that :

\begin{equation}
(SF)^{\dagger} = F^{\dagger} S^{\dagger}
\end{equation}

Then,

\begin{equation}
\begin{aligned}
\min_H \left \| SF (SF)^{\dagger} - H \right \| &= \min_H \left \| SFF^{\dagger}S^{\dagger} - H \right \| \\
&= \min_H \left \| SS^{\dagger} - H \right \| \\
&= \min_H \left \| SS - H \right \| \\
&= \min_H \left \| S - H \right \|\\
\Rightarrow \hat H = S
\end{aligned}
\end{equation}

\section{Second Section}

\begin{equation}
Y\leftarrow \sigma((XS+N)F)
\end{equation}

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
% \bibliographystyle{splncs04}
% \bibliography{mybibliography}
%
\begin{thebibliography}{8}
\bibitem{ref_article1}
Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)

\bibitem{ref_lncs1}
Author, F., Author, S.: Title of a proceedings paper. In: Editor,
F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
Springer, Heidelberg (2016). \doi{10.10007/1234567890}

\bibitem{ref_book1}
Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
Location (1999)

\bibitem{ref_proc1}
Author, A.-B.: Contribution title. In: 9th International Proceedings
on Proceedings, pp. 1--2. Publisher, Location (2010)

\bibitem{ref_url1}
LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
Oct 2017
\end{thebibliography}
\end{document}

### 2019-04-10 paper/old_intro.tex
NIPS format

--- 2019-04-09 paper/draft.tex
@@
-% This is samplepaper.tex, a sample chapter demonstrating the
-% LLNCS macro package for Springer Computer Science proceedings;
-% Version 2.20 of 2017/10/04
-%
-\documentclass[runningheads]{llncs}
-%
-\usepackage{graphicx}
-\usepackage{amsmath}
-
-% Used for displaying a sample figure. If possible, figure files should
-% be included in EPS format.
-%
-% If you use the hyperref package, please uncomment the following line
-% to display URLs in blue roman font according to Springer's eBook style:
-% \renewcommand\UrlFont{\color{blue}\rmfamily}
-
-\begin{document}
-
-\title{Identifying representations from neuroimaging data with covarying features}
-
-\author{King, Jean-Remi\inst{1} \and Lopez-Paz, David \inst{1}}
-
-\institute{Facebook Artificial Intelligence Research, Paris, France
-\email{\{jeanremi,d.lopez\}@fb.com}}
-
-\maketitle              % typeset the header of the contribution
-
-\begin{abstract}
-The abstract should briefly summarize the contents of the paper in
-15--250 words.
-
-\keywords{Neuroimaging \and Decoding \and Encoding \and Multivariate Pattern Analyses.}
-\end{abstract}
-
-\section{Introduction}
 Sixty years ago, Hubel and Wiesel implanted an anesthetized cat with a microelectrode, and discovered that distinct neurons reliably spiked when a visual stimulus was presented with a specific orientation, while other neurons only spiked when the stimulus was oriented at a different angle.
@@
 
-# RSA: in discussion
+RSA: in discussion
 - only applies to categorical effects
 - analyzed at the stimulus level, not the feature level
-
-\section{First Section}
-
-\begin{equation}
-Y\leftarrow (XS+N)F
-\end{equation}
-
-where Y is MEG activity, X are input features, S is encoding matrix, N is source noise and F is forward operator
-
-\subsection{LS from Y to X}
-
-\begin{equation}
-\begin{aligned}
-\min_G \left \| X-YG \right \| &= \min_G \left \| X - (XSF+NF)G\right\| \\
-&{}= \min_G \left \| X - XSFG\right\|  + \left \| NFG \right \| \\
-\end{aligned}
-\end{equation}
-
-Given that :
-\begin{equation}
-\mathop{\mathbb{E}}[N]=0
-\end{equation}
-
-and
-\begin{equation}
-\left \| X-XSFG \right \| \leq \left \| X \right \| \cdot  \left \| I-SFG \right \| \\
-\end{equation}
-
-Then
-\begin{equation}
-\min_G \left \| X-YG \right \| = \min_G \left \| I -SFG\right\| \\
-\Rightarrow \hat  G= (SF)^{\dagger}
-\end{equation}
-
-\subsection{LS from X to Y\^{G}}
-
-\begin{equation}
-\begin{aligned}
-\min_H \left \| Y \hat G - XH \right \| &= \min_H \left \| (XSF+NF) \hat G - XH \right \| \\
-&= \min_H \left \| XSF \hat G - XH \right \| \\
-&= \min_H \left \| SF \hat G - H \right \| \\
-&= \min_H \left \| SF (SF)^{\dagger} - H \right \| \\
-\end{aligned}
-\end{equation}
-
-Under the assumptions that S is an orthogonal projection
-\begin{equation}
-    S = S^2, S=S^T, S^{\dagger} = S
-\end{equation}
-
-And under the extra assumptions that :
-
-\begin{equation}
-(SF)^{\dagger} = F^{\dagger} S^{\dagger}
-\end{equation}
-
-Then,
-
-\begin{equation}
-\begin{aligned}
-\min_H \left \| SF (SF)^{\dagger} - H \right \| &= \min_H \left \| SFF^{\dagger}S^{\dagger} - H \right \| \\
-&= \min_H \left \| SS^{\dagger} - H \right \| \\
-&= \min_H \left \| SS - H \right \| \\
-&= \min_H \left \| S - H \right \|\\
-\Rightarrow \hat H = S
-\end{aligned}
-\end{equation}
-
-\section{Second Section}
-
-\begin{equation}
-Y\leftarrow \sigma((XS+N)F)
-\end{equation}
-
-%
-% ---- Bibliography ----
-%
-% BibTeX users should specify bibliography style 'splncs04'.
-% References will then be sorted and formatted in the correct style.
-%
-% \bibliographystyle{splncs04}
-% \bibliography{mybibliography}
-%
-\begin{thebibliography}{8}
-\bibitem{ref_article1}
-Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
-
-\bibitem{ref_lncs1}
-Author, F., Author, S.: Title of a proceedings paper. In: Editor,
-F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
-Springer, Heidelberg (2016). \doi{10.10007/1234567890}
-
-\bibitem{ref_book1}
-Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
-Location (1999)
-
-\bibitem{ref_proc1}
-Author, A.-B.: Contribution title. In: 9th International Proceedings
-on Proceedings, pp. 1--2. Publisher, Location (2010)
-
-\bibitem{ref_url1}
-LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
-Oct 2017
-\end{thebibliography}
-\end{document}

### 2019-04-10 paper/paper.tex
NIPS format

--- 2019-04-09 paper/draft.tex
@@
-% This is samplepaper.tex, a sample chapter demonstrating the
-% LLNCS macro package for Springer Computer Science proceedings;
-% Version 2.20 of 2017/10/04
-%
-\documentclass[runningheads]{llncs}
-%
-\usepackage{graphicx}
+\documentclass{article}
+
+\usepackage[final]{neurips_2019}
+
+\usepackage[utf8]{inputenc} % allow utf-8 input
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{hyperref}       % hyperlinks
+\usepackage{url}            % simple URL typesetting
+\usepackage{booktabs}       % professional-quality tables
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
 \usepackage{amsmath}
 
-% Used for displaying a sample figure. If possible, figure files should
-% be included in EPS format.
-%
-% If you use the hyperref package, please uncomment the following line
-% to display URLs in blue roman font according to Springer's eBook style:
-% \renewcommand\UrlFont{\color{blue}\rmfamily}
+\title{Learning causal features with double regression}
+
+\author{%
+  Jean-Remi King\\
+  Facebook AI\\
+  \texttt{email} \\
+  \And
+  Fran\c{c}ois Charton\\
+  Facebook AI\\
+  \texttt{fcharton@fb.com}\\
+  \And
+  David Lopez-Paz\\
+  Facebook AI\\
+  \texttt{fcharton@fb.com}
+}
 
@@
 
-\title{Identifying representations from neuroimaging data with covarying features}
-
-\author{King, Jean-Remi\inst{1} \and Lopez-Paz, David \inst{1}}
-
-\institute{Facebook Artificial Intelligence Research, Paris, France
-\email{\{jeanremi,d.lopez\}@fb.com}}
-
-\maketitle              % typeset the header of the contribution
+\maketitle
 
 \begin{abstract}
-The abstract should briefly summarize the contents of the paper in
-15--250 words.
-
-\keywords{Neuroimaging \and Decoding \and Encoding \and Multivariate Pattern Analyses.}
+    To be written.
 \end{abstract}
@@
 \section{Introduction}
-Sixty years ago, Hubel and Wiesel implanted an anesthetized cat with a microelectrode, and discovered that distinct neurons reliably spiked when a visual stimulus was presented with a specific orientation, while other neurons only spiked when the stimulus was oriented at a different angle.
-Since, a systematic mapping has been engaged in cognitive neuroscience to identify the contents encoded in brain activity.
-This quest has revealed that some neurons specifically responds to individual locations (place cell), whereas other respond to specific emotions () , or to specific individuals ().
-The mapping between neuronal activity and hypothetical features has been considerably accelerating with the development of functional Magnetic Resonance Imaginge (fMRI), eletro- and magneto-encephalography (E/MEG) as as well as with the accumulation of sensors on each microelectrode (neuropixel).
 
-The abundance of data produced by such recording devices now necessitate the design of statistical models optimized to identify neural representations.
-Two distinct analytical frameworks have been developed to address this issue: decoding and encoding.
-Both are based on the premise that a feature is represented in neuronal activity if and only if such a feature can be linearly and instantly read from that neuronal activity.
-This premise, discussed elsewhere (ref), capitalizes on the idea that neurons can be approximated as detector of particular combination of excitatory and inhibitory inputs, and can thus "read" information under a linear constrain.
+A survey in deep learning \citep{deep_learning_nature}.
 
-Encoding consists in predicting brain activity from a model.
-For example, an encoding model can be used to predict the brain responses to hand movements (ref), the identity of faces in a visual stimulus (ref), or to the nature of a syntactic operation (ref).
-Decoding consists in the reverse prediction: i.e. it is optimized to predict a feature from a brain activity pattern.
+\subsection{Contributions}
 
-In practice, a majority of fMRI studies are based on encoding analyses, whereas a majority of M/EEG studies are based on decoding analyses.
-This difference probably reflects the distinct limitations of these two types of neuroimaging.
-
-Indeed, fMRI produces brain images of relatively high spatial resolution: typically, each voxel specifically record neuronal activity over 3 mm cube, by measuring their blood-oxygen-level dependent (BOLD) response.
-The corresponding haemodynamic function that characterizes the BOLD is non-trivially linked to neuronal activity (ref), but is generally approximated as a temporal convolution of the neuronal activity.
-Consequently, in fMRI, the primary difficulty consists in disentangling the factors that have contributed to the modulation of the BOLD response.
-Encoding analyses can thus be used to quantify the contribution of multiple (potentially overlapping) features.
-
-By contrast, M/EEG record the magnetic and electric fields elicited by neuronal activity thanks via sensors distant from the brain.
-Unlike fMRI, the relationship between M/EEG signals and brain activity is both largely instantaneous and linear.
-However, M/EEG activity is a linear mixture of the brain activity: a unique brain area likely projects its activity onto multiple sensors, and a unique sensor likely records the activity of many brain areas.
-Furthermore, M/EEG sensors also pick up non-neuronal signals.
-In particular, the magnetic field of the earth, the presence of surrounding electronic devices, the heart, the electrically charged retina and their associated movements, as well as breathing, swallowing and jaw clenching can all elicit electro-magnetic signals orders of magnitude larger than those of our brain.
-Consequently, in M/EEG, the primary difficulty consists in detecting the specific activity that is associated with a model feature.
-Decoding analyses are thus used to discriminate, in this linear mixture of noise and signals.
-
-The respective advantages of encoding and decoding come at a costs.
-By focusing on univariate brain responses, encoding analyses cannot efficiently capture multivariate patterns of brain activity.
-For example, if the heart modulates the baseline activity of two voxels, then it can be difficult to show with a univariate analysis that each voxel also correlates with the feature of a model (Fig.
-X).
-To compensate for such issue, it is thus common to (1) incorporate physiological features (e.g.
-heart and breathing activity) in the encoding model and to (2) do repeated measurements in order to correct the variance explained by the model by the variance explained by an identical condition.
-This type of correction is thus limited to (1) observable artifacts as well as to (2) experimental effects that are independent from memory.
-
-By contrast, decoding analyses are fit to multivariate brain activity patterns, and can thus be more robust to the structured modulations induced by artefacts.
-However, by focusing on a single feature of a model (e.g.
-the presence of a face, the pitch of a tone etc), decoding cannot easily disentangle two covarying features (e.g.
-the contrast and the luminance of an image).
-To compensate for such an issue, it is thus common restrict the experimental protocols with synthetic stimuli whose features are orthogonal by design (e.g.
-images whose luminance and contrast are independently varied).
-
-In the present study, we introduce a novel method to get the best of the two approaches.
-We show on synthetic data that our approach systematically outperforms encoding and decoding methods, both in its specificity and sensitivity.
-We provide a proof that formalizes the conditions within which our method will be optimal.
-Finally we provide an extension to non-linear but monotonic activation functions in order to potentially in order to capture the saturation effects the BOLD and population responses.
-
-# RSA: in discussion
-- only applies to categorical effects
-- analyzed at the stimulus level, not the feature level
-
-\section{First Section}
-
+\section{Problem formulation}
 \begin{equation}
-Y\leftarrow (XS+N)F
+    Y = F(SX + N)
+    \label{eq:model}
 \end{equation}
 
-where Y is MEG activity, X are input features, S is encoding matrix, N is source noise and F is forward operator
-
-\subsection{LS from Y to X}
+\subsection{Consistency proof}
 
@@
 
-\section{Second Section}
+\section{Experiments}
 
-\begin{equation}
-Y\leftarrow \sigma((XS+N)F)
-\end{equation}
+\section{Conclusion}
 
-%
-% ---- Bibliography ----
-%
-% BibTeX users should specify bibliography style 'splncs04'.
-% References will then be sorted and formatted in the correct style.
-%
-% \bibliographystyle{splncs04}
-% \bibliography{mybibliography}
-%
-\begin{thebibliography}{8}
-\bibitem{ref_article1}
-Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
+\clearpage
+\newpage
 
-\bibitem{ref_lncs1}
-Author, F., Author, S.: Title of a proceedings paper. In: Editor,
-F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
-Springer, Heidelberg (2016). \doi{10.10007/1234567890}
+\bibliographystyle{abbrvnat}
+\bibliography{paper}
 
-\bibitem{ref_book1}
-Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
-Location (1999)
-
-\bibitem{ref_proc1}
-Author, A.-B.: Contribution title. In: 9th International Proceedings
-on Proceedings, pp. 1--2. Publisher, Location (2010)
-
-\bibitem{ref_url1}
-LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
-Oct 2017
-\end{thebibliography}
 \end{document}

### 2019-04-11 paper/paper.tex
Done.
Added blah.
Added maths (1), pseudo inverse derivation

@@
   Facebook AI\\
-  \texttt{fcharton@fb.com}
+  \texttt{dlp@fb.com}
 }
@@
 \section{Problem formulation}
+Our problem may be formulated as
 \begin{equation}
-    Y = F(SX + N)
+    Y = F(EX + N)
     \label{eq:model}
 \end{equation}
+where F is an unknown matrix, presumably full or high rank, corresponding to the measurement apparatus, E is a low rank transformation carried on the features (X), and N a noise term, centred but not necessarily gaussian or diagonal.
 
-\subsection{Consistency proof}
+The first regression (X from Y), amounts to finding
+\begin{equation}
+\begin{aligned}
+\min_G \left \| X-GY \right \|^2 &= \min_G \left \| X - GF(EX+N)\right\|^2 \\
+&{}= \min_G \left \| (I-GFE)X\right\| ^2 + \left \| GFN\right \| ^2\\
+&{}\leq \min_G \left \| I-GFE\right\| ^2 \left \| X\right\| ^2 + k\left \| G\right \| ^2\\ 
+\end{aligned}
+\end{equation}
+since N is centred, and with k a positive constant which depends on the variance of FN.
+
+This amounts to finding the minimal norm generalized inverse of FE, or its Moore-Penrose inverse
+\begin{equation}
+\hat  G= (FE)^{\dagger}
+\end{equation}
+
+The second regression is from X to the predicted value of X in the above, or $\hat G  Y$
 
@@
 \begin{aligned}
-\min_G \left \| X-YG \right \| &= \min_G \left \| X - (XSF+NF)G\right\| \\
-&{}= \min_G \left \| X - XSFG\right\|  + \left \| NFG \right \| \\
+\min_H \left \| (FE)^{\dagger}Y - HX \right \|^2 &= \min_H \left \| (FE)^{\dagger}F(EX+N) - HX \right \|^2 \\
+&= \min_H \left \| ((FE)^{\dagger}(FE)-H)X \right \| ^2 + \left \| (FE)^{\dagger}FN \right \| ^2\\
+&= \min_H \left \| ((FE)^{\dagger}(FE)-H)X \right \| ^2\\
 \end{aligned}
@@
 
-Given that :
-\begin{equation}
-\mathop{\mathbb{E}}[N]=0
-\end{equation}
-
-and
-\begin{equation}
-\left \| X-XSFG \right \| \leq \left \| X \right \| \cdot  \left \| I-SFG \right \| \\
-\end{equation}
-
-Then
-\begin{equation}
-\min_G \left \| X-YG \right \| = \min_G \left \| I -SFG\right\| \\
-\Rightarrow \hat  G= (SF)^{\dagger}
-\end{equation}
-
-\subsection{LS from X to Y\^{G}}
-
+Therefore, under the hypothese that N is centred, our regression method estimates 
 \begin{equation}
 \begin{aligned}
-\min_H \left \| Y \hat G - XH \right \| &= \min_H \left \| (XSF+NF) \hat G - XH \right \| \\
-&= \min_H \left \| XSF \hat G - XH \right \| \\
-&= \min_H \left \| SF \hat G - H \right \| \\
-&= \min_H \left \| SF (SF)^{\dagger} - H \right \| \\
+\hat H =(FE)^{\dagger}(FE)
 \end{aligned}
@@
 
-Under the assumptions that S is an orthogonal projection
-\begin{equation}
-    S = S^2, S=S^T, S^{\dagger} = S
-\end{equation}
+For any matrix A, $A^\dagger A$ is the orthogonal projector over Img(A'), and $I-A^\dagger A$ the orthogonal projector over Ker(A).
 
-And under the extra assumptions that :
-
-\begin{equation}
-(SF)^{\dagger} = F^{\dagger} S^{\dagger}
-\end{equation}
-
-Then,
-
+If E selects some features of X (ie Ex=0 on any unselected feature, and E full ranked over the set of selected features) and if F is full rank over the subspace spanned by the features selected by E, then
 \begin{equation}
 \begin{aligned}
-\min_H \left \| SF (SF)^{\dagger} - H \right \| &= \min_H \left \| SFF^{\dagger}S^{\dagger} - H \right \| \\
-&= \min_H \left \| SS^{\dagger} - H \right \| \\
-&= \min_H \left \| SS - H \right \| \\
-&= \min_H \left \| S - H \right \|\\
-\Rightarrow \hat H = S
+Ker(FE) = Ker(E)  
 \end{aligned}
 \end{equation}
+$I - \hat H$ is the projector over Ker(E), the diagonal matrix that spans the unselected features, and $\hat H$ is the diagonal matrix corresponding to the selected features.
+
+Under such hypotheses, $\hat H$ retrieves the features of X that have a causal influence on Y, and should do it in a fairly resilient way, since it is diagonal, and all its eigenvalues (0 or 1) are well separated.
+
+\section{Estimator, and relation with Canonical Component Analysis}
 

### 2019-04-12 code/models/cca_jr.py
cca jr

@@ -1,0 +1,144 @@
+"""
+
+https://en.wikipedia.org/wiki/Canonical_correlation
+
+See github.com/Neuromorphs/telluride-decoding-toolbox/blob/master/cca.m
+for similar approach without scaling and bagging
+"""
+
+import numpy as np
+from scipy.linalg import eigh, pinv
+from sklearn.base import BaseEstimator
+from sklearn.preprocessing import StandardScaler
+from sklearn.model_selection import ShuffleSplit
+from sklearn.cross_decomposition import CCA as skCCA
+
+
+class CCA(BaseEstimator):
+
+    def __init__(self, n_components=-1, alpha=0., normalize=True,
+                 bagging=False):
+        self.n_components = n_components
+        self.alpha = alpha
+        self.bagging = bagging
+        self.normalize = normalize
+
+    def fit(self, X, Y):
+        assert len(X) == len(Y)
+
+        nx = X.shape[1]
+        ny = Y.shape[1]
+
+        if self.normalize:
+            self.xscale_ = StandardScaler()
+            X = self.xscale_.fit_transform(X)
+            self.yscale_ = StandardScaler()
+            Y = self.yscale_.fit_transform(Y)
+
+        if self.bagging in (False, 0, None):
+            Gset = range(len(X))
+            Hset = range(len(X))
+            ensemble = [(Gset, Hset)]
+        else:
+            bagging = ShuffleSplit(self.bagging, test_size=.5)
+            ensemble = [split for split in bagging.split(X, Y)]
+
+        nx, ny = X.shape[1], Y.shape[1]
+
+        # Bagging ensemble
+        Pxxs = list()
+        Cxys = list()
+        Pyys = list()
+        Cyxs = list()
+        for set1, set2 in ensemble:
+            # FIXME: suboptimal way to compute cov(X,Y)
+            XY1 = np.c_[X[set1], Y[set1]]
+            XY2 = np.c_[X[set2], Y[set2]]
+            Cxy = np.cov(XY1.T)[:nx, nx:]
+            Cyx = np.cov(XY2.T)[:nx, nx:].T
+
+            Cxx = np.cov(X[set1].T)
+            Cyy = np.cov(Y[set2].T)
+
+            # invert matrices with tikohnov regularization
+            Pxxs.append(pinv(Cxx + self.alpha*np.eye(nx)))
[…]

### 2019-04-12 code/models/cca_gallant.py
cca gallantlab

@@ -1,0 +1,446 @@
+"""Python module for regularized kernel canonical correlation analysis
+
+Slightly Adapted from https://github.com/gallantlab/pyrcca
+
+Licence
+Copyright (c) 2015, The Regents of the University of California (Regents).
+All rights reserved.
+
+Permission to use, copy, modify, and distribute this software and its
+documentation for educational, research, and not-for-profit purposes, without
+fee and without a signed licensing agreement, is hereby granted, provided that
+the above copyright notice, this paragraph and the following two paragraphs
+appear in all copies, modifications, and distributions. Contact The Office of
+Technology Licensing, UC Berkeley, 2150 Shattuck Avenue, Suite 510, Berkeley,
+CA 94720-1620, (510) 643-7201, for commercial licensing opportunities.
+
+Created by Natalia Bilenko, University of California, Berkeley.
+
+IN NO EVENT SHALL REGENTS BE LIABLE TO ANY PARTY FOR DIRECT, INDIRECT, SPECIAL,
+INCIDENTAL, OR CONSEQUENTIAL DAMAGES, INCLUDING LOST PROFITS, ARISING OUT OF
+THE USE OF THIS SOFTWARE AND ITS DOCUMENTATION, EVEN IF REGENTS HAS BEEN
+ADVISED OF THE POSSIBILITY OF SUCH DAMAGE. REGENTS SPECIFICALLY DISCLAIMS ANY
+WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF
+MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE. THE SOFTWARE AND
+ACCOMPANYING DOCUMENTATION, IF ANY, PROVIDED HEREUNDER IS PROVIDED "AS IS".
+REGENTS HAS NO OBLIGATION TO PROVIDE MAINTENANCE, SUPPORT, UPDATES,
+ENHANCEMENTS, OR MODIFICATIONS.
+"""
+
+# TODO: fix kernel prediction: currently only project to Z space
+
+import joblib
+import numpy as np
+from scipy.linalg import eigh
+from sklearn.preprocessing import StandardScaler
+
+
+__copyright__ = 'Copyright 2016, UC Berkeley, Gallant lab.'
+
+
+class _CCABase(object):
+    def __init__(self, numCV=None, reg=None, regs=None, numCC=None,
+                 numCCs=None, kernelcca=True, ktype=None, verbose=False,
[…]

### 2019-04-12 code/models/pls_kernel.py
pls kernel

@@ -1,0 +1,414 @@
+"""A module which implements kernel PLS.
+
+from github.com/jhumphry/regressions/blob/master/examples/kpls_example.py
+
+ISC License
+
+Regressions
+===========
+
+Provides various forms of regression which are not all covered by other
+Python statistical packages. The aim is to achieve clarity of
+implementation with speed a secondary goal. Python 3.5 and Numpy 1.10 or
+greater are required as the new '@' matrix multiplication operator is
+used.
+
+All of the regressions require the X and Y data to be provided in the form
+of matrices, with one row per data sample and the same number of data
+samples in each. Currently missing data or NaN are not supported.
+
+"""
+
+# Copyright (c) 2015, James Humphry - see LICENSE file for details
+
+import random
+import abc
+import numpy as np
+import math
+import scipy.linalg as linalg
+
+
+class ParameterError(Exception):
+    """Parameters passed to a regression routine are unacceptable
+
+    This is a generic exception used to indicate that the parameters
+    passed are mis-matched, nonsensical or otherwise problematic.
+    """
+    pass
+
+
+class ConvergenceError(Exception):
+    """Iterative algorithm failed to converge.
+
+    Many of the routines used for regressions are iterative and in some
+    cases may not converge. This is mainly likely to happen if the data
+    has pathological features, or if too many components of a data set
+    have been extracted by an iterative process and the residue is
+    becoming dominated by rounding or other errors.
+    """
+    pass
+
+
+DEFAULT_MAX_ITERATIONS = 250
+"""Default maximum number of iterations that iterative routines will
+attempt before raising a ConvergenceError."""
+
+DEFAULT_EPSILON = 1.0E-6
+"""A default epsilon value used in various places, such as to decide when
+iterations have converged sufficiently."""
+
+
+class RegressionBase(metaclass=abc.ABCMeta):
+
+    """Abstract base class for regressions
+
+    All the various types of regression objects will have at least the
+    attributes present here.
+
+    Args:
+        X (ndarray N x n): X calibration data, one row per data sample
+        Y (ndarray N x m): Y calibration data, one row per data sample
+        standardize_X (boolean, optional): Standardize the X data
+        standardize_Y (boolean, optional): Standardize the Y data
+
+    Attributes:
+        data_samples (int): number of calibration data samples (=N)
+        max_rank (int): maximum rank of calibration X-data (limits the
+            number of components that can be found)
[…]

### 2019-04-12 code/models/jrr.py
jrr

@@ -1,0 +1,103 @@
+import numpy as np
+from sklearn.linear_model import RidgeCV
+from sklearn.model_selection import ShuffleSplit
+
+
+class JRR(object):
+    """Finds E in:
+        Y = F(EX+N)
+
+    Where Y is the recordings (samples * sensors),
+          F is an unknown forward operator (neurons * sensors),
+          X is the factors (samples * sources),
+          N is a source noise (samples * sources)
+
+    Step 1: Bagging
+    for set1, set2 in partitions:
+        Step 1.1: G = ls(Y_1, X_1)
+        Step 1.2: H = ls(X_2, Y_2 * G)
+    Step 2 (optional): E = diag(H)
+    Step 3: F = ls(XE, Y)
+    """
+
+    def __init__(self, G=None, H=None, F=None, bagging=10, zero_off_diag=True):
+        self.G = RidgeCV() if G is None else G
+        self.H = RidgeCV() if H is None else H
+        self.F = RidgeCV() if F is None else F
+        self.bagging = bagging
+        self.zero_off_diag = zero_off_diag
+
+    def fit(self, X, Y):
+        if self.bagging is False:
+            Gset = range(len(X))
+            Hset = range(len(X))
+            ensemble = [(Gset, Hset)]
+        else:
+            bagging = ShuffleSplit(self.bagging, test_size=.5)
+            ensemble = [split for split in bagging.split(X, Y)]
+
+        H = list()
+        for train, test in ensemble:
+            X_hat = self.G.fit(Y[train], X[train]).predict(Y)
+            H += [self.H.fit(X[test], X_hat[test]).coef_, ]
+
+        # Estimate E
+        self.E_ = np.mean(H, 0).T  # TODO: check transpose
+        if self.zero_off_diag:
+            self.E_ = np.diag(np.diag(self.E_))
+        # Filter X by E hat and estimate F
+        self.F.fit(X @ self.E_, Y)
+        return self
+
+    def predict(self, X):
+        return self.F.predict(X @ self.E_)
+
+    def score(self, X, y, sample_weight=None):
+        from sklearn.metrics import r2_score
+        return r2_score(y, self.predict(X), sample_weight=sample_weight,
+                        multioutput='variance_weighted')
+
+
+if __name__ == '__main__':
[…]

### 2019-04-12 code/models/regression_nonspherical.py
non spherical regression

@@ -1,0 +1,79 @@
+# Author: Jean-Remi King <jeanremi.king@gmail.com>
+#
+# Licence: BSD 3-clause
+
+"""Implement regularized regression with non-spherical priors as described in:
+   Nunez-Elizalde*, Huth* & Gallant (2016) Improving predictive models using
+   non-spherical Gaussian priors, Society for Neuroscience (Poster).
+"""
+import numpy as np
+from numpy.random import rand, randn
+from scipy.linalg import inv, sqrtm
+from sklearn.linear_model import Ridge
+
+
+def r2(a, b):
+    return np.corrcoef(a.squeeze(), b.squeeze())[0, 1]
+
+
+def ridge(X, y, alpha=1.):
+    """Classic ridge (i.e. spherical priors)"""
+    Id = np.eye(X.shape[1])
+    return inv(X.T.dot(X) + alpha * Id) @ X.T.dot(y)
+
+
+def tikhonov_ns_prior(X, y, S, alpha=1.):
+    """Tikhonov with non spherical gaussian priors"""
+    S_sqrt = sqrtm(S)
+    A = X.dot(S_sqrt)
+    Id = np.eye(A.shape[1])
+    beta_A = inv(A.T.dot(A) + alpha * Id) @ A.T.dot(y)
+    return S_sqrt.dot(beta_A)
+
+
+class NonSphericalRidge(Ridge):
+    def __init__(self, S, alpha=1):
+        # TODO add intercept
+        self.S = S
+
+    def fit(self, X, y):
+        self.coef_ = tikhonov_ns_prior(X, y, self.S, self.alpha)
+        return self
+
+
+if __name__ == '__main__':
+    r_ridge, r_ns_prior = list(), list()  # initialize scoring metrics
+
+    for voxel in range(100):
+        print(voxel)
+        # Simulate data
+        n_samples, n_features = 200, 500
+
+        # Independent features
+        X = randn(n_samples, n_features)
+
+        # Make non-orthogonal features
+        S = rand(n_features, n_features)
+        S = S @ S.T  # ensure positive definite covariance
+
+        X = X @ S.T
+
+        # True coefficients
+        beta = randn(n_features, 1)
+
+        # Observables
+        noise = randn(n_samples, 1)
+        y = np.dot(X, beta) + noise
+
+        # Estimate beta with or without spherical priors
+        beta_hat = ridge(X, y, alpha=1.)
[…]

### 2019-04-12 code/models/regression_nonspherical.py
non spherical regression

@@ -5,6 +5,8 @@
 """Implement regularized regression with non-spherical priors as described in:
    Nunez-Elizalde*, Huth* & Gallant (2016) Improving predictive models using
    non-spherical Gaussian priors, Society for Neuroscience (Poster).
+
+   Unclear in which condition it will be useful?
 """
 import numpy as np
 from numpy.random import rand, randn
@@ -45,7 +47,6 @@ if __name__ == '__main__':
     r_ridge, r_ns_prior = list(), list()  # initialize scoring metrics
 
     for voxel in range(100):
-        print(voxel)
         # Simulate data
         n_samples, n_features = 200, 500
 

### 2019-04-12 code/models/encoding.py
encoding models: ols(PCA(Y), X)

@@ -1,0 +1,76 @@
+import numpy as np
+from sklearn.linear_model import RidgeCV, MultiTaskLassoCV, LinearRegression
+from sklearn.base import BaseEstimator
+from sklearn.decomposition import PCA
+from sklearn.preprocessing import scale
+from sklearn.metrics import roc_auc_score
+
+
+class Encoding(BaseEstimator):
+    def __init__(self, n_components=5, ols=None):
+        self.pca = PCA(n_components)
+        self.ols = RidgeCV() if ols is None else ols
+
+    def fit(self, X, Y):
+        Yt = self.pca.fit_transform(Y)
+        self.ols.fit(X, Yt)
+        self.coef_ = self.ols.coef_.T  # to be in similar format to PLS and CCA
+        return self
+
+    def predict(self, X):
+        Ypred = self.ols.predict(X)
+        return self.pca.inverse_transform(Ypred)
+
+    def score(self, X, y, sample_weight=None):
+        from sklearn.metrics import r2_score
+        return r2_score(y, self.predict(X), sample_weight=sample_weight,
+                        multioutput='variance_weighted')
+
+
+if __name__ == '__main__':
+    # Simulate data
+    """Y = F(EX+N)"""
+
+    np.random.seed(0)
+
+    # Problem dimensionality
+    n = 1000
+    nE = nX = 10
+    nY = 10
+    snr = .25  # signal to noise ratio
+    selected = .5  # number of X feature selected by E
+
+    selected = min(int(np.floor(selected*nX)) + 1, nX-1)
+    E = np.identity(nX)
+    E[selected:] = 0
+
+    # X covariance
+    Cx = np.random.randn(nX, nX)
+    Cx = Cx.dot(Cx.T) / nX  # sym pos-semidefin
+    X = np.random.multivariate_normal(np.zeros(nX), Cx, n)
+
+    # Noise (homosedastic in source space)
+    N = np.random.randn(n, nE)
+
+    # Forward operator (linear mixture)
+    F = np.random.randn(nY, nE)
+
+    Y = ((X @ E.T) * snr + N) @ F.T
+
+    X = scale(X)
+    Y = scale(Y)
+
+    # Fit method
+    alphas = np.logspace(-2, 2, 5)
+    penalties = dict(none=LinearRegression(),
+                     l2=RidgeCV(alphas=alphas),
+                     l1=MultiTaskLassoCV(alphas=alphas))
[…]

### 2019-04-12 code/models/jrr.py

@@ -70,7 +70,7 @@ if __name__ == '__main__':
     n = 1000
     nE = nX = 10
     nY = 10
-    snr = .5  # signal to noise ratio
+    snr = .25  # signal to noise ratio
     selected = .5  # number of X feature selected by E
 
     selected = min(int(np.floor(selected*nX)) + 1, nX-1)

### 2019-04-12 code/models/base.py
grid search cv wrapper for n_components

@@ -1,0 +1,45 @@
+import numpy as np
+from sklearn.base import BaseEstimator
+from sklearn.model_selection import GridSearchCV
+
+
+class EstimatorCV(BaseEstimator):
+    """Transform an operator into a grid search for n_componentsself.
+    This is useful because sklearn CCA and PLS restricts the n_components
+    to the X dimensions whereas our Encoding model resticts the n_components to
+    the Y.
+    """
+
+    def __init__(self, estimator, dim='X', n_components=5):
+        """
+        n_components is the number of models tested between 1 and size of 'dim'
+        e.g. if dim == 'X' and X.shape[1] == 100 and n_components = 3
+             then the grid will search through self.components_= [1, 10, 100]
+        """
+        self.n_components = n_components
+        self.dim = dim
+        self.base_estimator = estimator
+
+    def fit(self, X, y):
+        if self.dim == 'X':
+            n_max = X.shape[1]
+        else:
+            n_max = y.shape[1]
+
+        components = np.logspace(0, np.log10(n_max), self.n_components)
+        components = np.unique(components.astype(int))
+        self.estimator_ = GridSearchCV(self.base_estimator,
+                                       dict(n_components=components))
+        self.estimator_.fit(X, y)
+        return self
+
+    def predict(self, X):
+        return self.estimator_.predict(X)
+
+    def score(self, X, y):
+        return self.estimator_.score(X, y)
+
+
+if __name__ == '__main__':
+    from sklearn.cross_decomposition import CCA
+    cca_cv = EstimatorCV(CCA(), dim='X')

### 2019-04-12 code/models/rrr.py
reduced rank regression

@@ -1,0 +1,102 @@
+"""
+Reduced rank regression class.
+aka Optimal linear 'bottlenecking' or 'multitask learning'.
+
+Aapted from
+https://raw.githubusercontent.com/riscy/machine_learning_linear_models/master/reduced_rank_regressor.py
+
+(c) Chris Rayner (2015), dchrisrayner AT gmail DOT com
+
+For kernel implementation see
+https://github.com/rockNroll87q/RRRR/blob/master/reduced_rank_regressor.py
+"""
+
+
+import numpy as np
+from scipy import randn
+from scipy.sparse import eye
+from scipy.linalg import pinv, svd
+
+
+def ideal_data(num, dimX, dimY, rrank, noise=1):
+    """Low rank data"""
+    X = randn(num, dimX)
+    W = randn(dimX, rrank)  @ randn(rrank, dimY)
+    Y = X @ W + randn(num, dimY) * noise
+    return X, Y
+
+
+class ReducedRankRegressor(object):
+    """
+    Reduced Rank Regressor (linear 'bottlenecking' or 'multitask learning')
+    - X is an n-by-d matrix of features.
+    - Y is an n-by-D matrix of targets.
+    - rrank is a rank constraint.
+    - reg is a regularization parameter (optional).
+    """
+    def __init__(self, rank, reg=0):
+        self.reg = reg
+        self.rank = rank
+
+    def fit(self, X, Y):
+        CXX = X.T @ X + self.reg * eye(X.shape[1])
+        CXY = X.T @ Y
+        _U, _S, V = svd(CXY.T @ pinv(CXX) @ CXY)
+        self.W_ = V[:self.rank, :].T
+        self.A_ = (pinv(CXX) @ CXY @ self.W_).T
+        return self
+
+    def predict(self, X):
+        """Predict Y from X."""
+        return X @ self.A_.T @ self.W_.T
+
+    def score(self, X, y, sample_weight=None):
+        from sklearn.metrics import r2_score
+        return r2_score(y, self.predict(X), sample_weight=sample_weight,
+                        multioutput='variance_weighted')
+
+
+if __name__ == '__main__':
+    from sklearn.preprocessing import scale
+    from sklearn.metrics import roc_auc_score
+    # Simulate data
+    """Y = F(EX+N)"""
+
+    np.random.seed(0)
+
+    # Problem dimensionality
[…]

### 2019-04-12 code/models/partial_correlations.py
add david s partial correlation

@@ -1,0 +1,92 @@
+"""David's first approach when I exposed the problem.
+Reasonable to add in the comparison?
+"""
+import numpy as np
+from sklearn.linear_model import RidgeCV
+from sklearn.model_selection import ShuffleSplit
+
+
+def correlation(x, y):
+    a = (x - x.mean(0)) / x.std(0)
+    b = (y - y.mean(0)) / y.std(0)
+    return a.T @ b / x.shape[0]
+
+
+def partial_correlation_bagging(solver, x, y, z, ensemble=None):
+    if ensemble is None:
+        ensemble = [(range(len(x)), range(len(x))), ]
+    r = []
+    for set1, set2 in ensemble:
+        p_x = solver.fit(z[set1], x[set1]).predict(z[set2])
+        p_y = solver.fit(z[set1], y[set1]).predict(z[set2])
+        r.append(correlation(x[set2] - p_x, y[set2] - p_y))
+    return np.mean(r, 0)
+
+
+def partial_correlation_loop(solver, x, y, ensemble=None):
+    e_hat = np.zeros(y.shape[1])
+    for i in range(y.shape[1]):
+        y_i = y[:, i].reshape(-1, 1)
+        y_not_i = np.delete(y, i, axis=1)
+        r = partial_correlation_bagging(solver, x, y_i, y_not_i, ensemble)
+        e_hat[i] = np.sum(r**2)
+    return e_hat
+
+
+class PartialCorrelation(object):
+
+    def __init__(self, solver=None, bagging=False):
+        self.solver = RidgeCV() if solver is None else solver
+        self.bagging = bagging
+
+    def fit(self, X, Y):
+        ensemble = None
+        if self.bagging:
+            cv = ShuffleSplit(test_size=.5)
+            ensemble = [(train, test) for train, test in cv.split(X, Y)]
+        self.E_ = partial_correlation_loop(self.solver, X, Y, ensemble)
+        return self
+
+
+if __name__ == '__main__':
+    from sklearn.preprocessing import scale
+    from sklearn.metrics import roc_auc_score
+    # Simulate data
+    """Y = F(EX+N)"""
+
+    np.random.seed(0)
+
+    # Problem dimensionality
+    n = 1000
+    nE = nX = 10
+    nY = 10
+    snr = 25  # signal to noise ratio
+    selected = .5  # number of X feature selected by E
+
[…]

### 2019-04-12 paper/paper.tex

@@
 
-\section{Problem formulation}
+\section{Math corner}
+
+\subsection{Problem formulation, asymptotic estimates}
+
 Our problem may be formulated as
@@
 
-Therefore, under the hypothese that N is centred, our regression method estimates 
+Therefore, under the hypothesis that N is centred, our regression method estimates 
 \begin{equation}
@@
 
-For any matrix A, $A^\dagger A$ is the orthogonal projector over Img(A'), and $I-A^\dagger A$ the orthogonal projector over Ker(A).
+For any matrix A, $A^\dagger A$ is the orthogonal projector over $Img(A')$, and $I-A^\dagger A$ the orthogonal projector over $Ker(A)$.
 
-If E selects some features of X (ie Ex=0 on any unselected feature, and E full ranked over the set of selected features) and if F is full rank over the subspace spanned by the features selected by E, then
+If E selects some features of X (ie $Ex=0$ on any unselected feature, and E full ranked over the set of selected features) and if F is full rank over the subspace spanned by the features selected by E, then
 \begin{equation}
@@
 \end{equation}
-$I - \hat H$ is the projector over Ker(E), the diagonal matrix that spans the unselected features, and $\hat H$ is the diagonal matrix corresponding to the selected features.
+$I - \hat H$ is the projector over $Ker(E)$, the diagonal matrix that spans the unselected features, and $\hat H$ is the diagonal matrix corresponding to the selected features.
 
@@
 
-\section{Estimator, and relation with Canonical Component Analysis}
+\subsection{The double regression estimator, and canonical component analysis}
+
+For any real matrices of X and Y, the least square regression of Y from X is calculated by the formula (X' denoting the transpose of X)
+\begin{equation}
+\begin{aligned}
+(X'X)^{-1} X'Y
+\end{aligned}
+\end{equation}
+and the prediction of Y thus obtained (from X) can be written
+\begin{equation}
+\begin{aligned}
+X(X'X)^{-1} X'Y
+\end{aligned}
+\end{equation}
+
+In double regression, we first calculate the regression of X from Y, and calculate $\hat X$ the corresponding prediction as $Y(Y'Y)^{-1} Y'X$. The regression of $\hat X$ from X then becomes
+\begin{equation}
+\begin{aligned}
+(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X
+\end{aligned}
+\end{equation}
+Under ideal conditions (good conditioning of $X'X$ and $Y'Y$), this should recover a diagonal binary matrix, the 1 on the diagonal corresponding the the features of X used by Y.
+In a noisy environment, double regression tries to extract the useful features of X (ie those having an influence on Y) as non zero elements of the diagonal of (9).
+
+(here be statistical test for the diagonal, and rationale for it)
+
+Formula (9) shows that double regression can be understood as a special case of canonical component analysis (CCA).
+In CCA, given two matrixes (n,p and n,q) X and Y describing different features measured on the same data sample, one tries to characterise the closeness between X and Y (ie between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.
+This is done by finding vectors $a$ and $b$ such that $Xa$ and $Yb$ display maximal correlation, or by maximising $a'X'Yb$ over unit-normed $a$ and $b$.
+To find a, one calculates the eigenvectors of (9).
+The largest one is the direction with maximal correlation, the second one the maximal correlation direction orthogonal to the first, and so on.
+For each canonical dimension, the corresponding eigenvalue is the square of the cosine between the corresponding direction along X and the subspace spanned by Y.
+
+If (9) is diagonal, for instance in the model presented above, the features of X are eigenvectors.
+If all eigenvalues are either 0 or 1, since they correspond to the squared cosine of the angle between the corresponding feature and the subspace spanned by Y, all the features are either part of the Y subspace (eigenvalues of one) or orthogonal to it (for zero eigenvalues).
+
+Thus, double regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
+The closer we are to this hypothesis, the closer the results of CCA and double regression will be.
+
+\subsection{Stability analysis}
+
+\subsection{Related approaches}
+(PLS, 2 way, 3 fold...)
 

### 2019-04-12 code/models/cca_gallant.py
default to no n_components truncating + sklearn compatibility

@@ -33,6 +33,7 @@ import joblib
 import numpy as np
 from scipy.linalg import eigh
 from sklearn.preprocessing import StandardScaler
+from sklearn.base import BaseEstimator
 
 
 __copyright__ = 'Copyright 2016, UC Berkeley, Gallant lab.'
@@ -376,9 +377,9 @@ def _make_kernel(d, normalize=True, ktype='linear', gausigma=1.0, degree=2):
     return kernel
 
 
-class KernelCCA(object):
+class KernelCCA(BaseEstimator):
     """Object to make pyrcca have sklearn API"""
-    def __init__(self, reg=0, n_components=2, kernel='linear'):
+    def __init__(self, reg=0, n_components=-1, kernel='linear'):
         if kernel == 'linear':
             self.cca = CCA(reg=reg, kernelcca=False,
                            numCC=n_components)
@@ -389,18 +390,22 @@ class KernelCCA(object):
             self.cca = CCA(reg=reg, ktype=kernel_, kernelcca=True,
                            numCC=n_components)
         self.kernel = kernel
+        self.n_components = n_components
 
     def fit(self, X, Y):
         self.x_scaler_ = StandardScaler()
         X = self.x_scaler_.fit_transform(X)
         self.y_scaler_ = StandardScaler()
         Y = self.y_scaler_.fit_transform(Y)
+        if self.n_components == -1:
+            self.cca.numCC = X.shape[1]
         self.cca.train([X, Y])
         wx, wy = self.cca.ws
         self.x_weights_ = wx
         self.y_weights_ = wy
         iwy = np.linalg.pinv(wy, rcond=self.cca.cutoff)
         self.coef_ = iwy.T @ wx.T
+        self.coef_ = self.coef_.T  # to match sklearn CCA
         # TODO: find var mapping with sklearn
         return self
 
@@ -408,7 +413,7 @@ class KernelCCA(object):
         if self.kernel != 'linear':
             print('kernel prediction are probably not valid')
         X = self.x_scaler_.transform(X)
-        Y_hat = (self.coef_ @ X.T).T
+        Y_hat = (self.coef_.T @ X.T).T
         Y_hat = self.y_scaler_.inverse_transform(Y_hat)
         return Y_hat
 

### 2019-04-12 code/models/jrr.py
bagging param

@@ -28,7 +28,7 @@ class JRR(object):
         self.zero_off_diag = zero_off_diag
 
     def fit(self, X, Y):
-        if self.bagging is False:
+        if self.bagging in (0, False, None):
             Gset = range(len(X))
             Hset = range(len(X))
             ensemble = [(Gset, Hset)]

### 2019-04-12 code/models/rrr.py
sklearn-like variables

@@ -16,6 +16,7 @@ import numpy as np
 from scipy import randn
 from scipy.sparse import eye
 from scipy.linalg import pinv, svd
+from sklearn.base import BaseEstimator
 
 
 def ideal_data(num, dimX, dimY, rrank, noise=1):
@@ -26,7 +27,7 @@ def ideal_data(num, dimX, dimY, rrank, noise=1):
     return X, Y
 
 
-class ReducedRankRegressor(object):
+class ReducedRankRegressor(BaseEstimator):
     """
     Reduced Rank Regressor (linear 'bottlenecking' or 'multitask learning')
     - X is an n-by-d matrix of features.
@@ -34,21 +35,21 @@ class ReducedRankRegressor(object):
     - rrank is a rank constraint.
     - reg is a regularization parameter (optional).
     """
-    def __init__(self, rank, reg=0):
+    def __init__(self, n_components=-1, reg=0):
         self.reg = reg
-        self.rank = rank
+        self.n_components = n_components
 
     def fit(self, X, Y):
         CXX = X.T @ X + self.reg * eye(X.shape[1])
         CXY = X.T @ Y
         _U, _S, V = svd(CXY.T @ pinv(CXX) @ CXY)
-        self.W_ = V[:self.rank, :].T
-        self.A_ = (pinv(CXX) @ CXY @ self.W_).T
+        self.W_ = V[:self.n_components, :].T
+        self.coef_ = pinv(CXX) @ CXY @ self.W_
         return self
 
     def predict(self, X):
         """Predict Y from X."""
-        return X @ self.A_.T @ self.W_.T
+        return X @ self.coef_ @ self.W_.T
 
     def score(self, X, y, sample_weight=None):
         from sklearn.metrics import r2_score
@@ -94,8 +95,8 @@ if __name__ == '__main__':
     # Fit method
     rrr = ReducedRankRegressor(3)
     train, test = range(0, n, 2), range(1, n, 2)
-    coef = rrr.fit(X[train], Y[train]).A_
-    E_hat = np.mean(coef**2, 0)
+    coef = rrr.fit(X[train], Y[train]).coef_
+    E_hat = np.mean(coef**2, 1)
     score = rrr.score(X[test], Y[test])
 
     print('E_auc', roc_auc_score(np.diag(E), E_hat))

### 2019-04-12 code/base.py
clean up

@@ -1,0 +1,52 @@
+import numpy as np
+from sklearn.preprocessing import scale
+
+
+def make_data(n: int = 1000,  # number of samples
+              nX: int = 100,  # dimensionality of X
+              nE: int = 10,  # dimensionality of E
+              nY: int = 10,  # dimensionality of Y
+              random_seed: int = None,
+              snr: float = 1e-1,  # signal to noise ratio
+              selected: float = .5,  # proportion of selected feature in E
+              E: np.ndarray = None,  # selecting matrix
+              F: np.ndarray = None,  # mixing matrix
+              Cx: np.ndarray = None,  # feature covariance
+              heteroscedastic: bool = False,  # noise type
+              ) -> (np.ndarray, np.ndarray, np.ndarray):
+
+    """Y = F(EX+N)"""
+    np.random.seed(random_seed)
+
+    if E is None:
+        E = np.zeros((nE, nX))
+        selected = min(int(np.floor(selected*nX)) + 1, nX-1)
+        E[:, :selected] = np.random.randn(nE, selected)
+    else:
+        nE, nX = E.shape
+
+    if Cx is None:
+        Cx = np.random.randn(nX, nX)
+        Cx = Cx.dot(Cx.T) / nX  # sym pos-semidefin
+    else:
+        nX = len(Cx)
+    X = np.random.multivariate_normal(np.zeros(nX), Cx, n)
+
+    N = np.random.randn(n, nE)
+    if heteroscedastic:
+        Cn = np.random.randn(nE, nE)
+        Cn = Cn.dot(Cn.T) / nE  # sym pos-semidefin
+        N = np.random.multivariate_normal(np.zeros(nE), Cn, n)
+
+    if F is None:
+        F = np.random.randn(nY, nE)
+    else:
+        nY = len(F)
+
+    S = (X @ E.T) * snr
+    Y = (S+N) @ F.T
+
+    X = scale(X)
+    Y = scale(Y)
+
+    return X, Y, E

### 2019-04-12 code/models/__init__.py

@@ -1,0 +1,6 @@
+from .base import EstimatorCV
+from .jrr import JRR
+from .encoding import Encoding
+from .pls_kernel import KernelPLS
+from .cca_gallant import KernelCCA
+from .rrr import ReducedRankRegressor

### 2019-04-12 code/models/base.py

@@ -31,6 +31,7 @@ class EstimatorCV(BaseEstimator):
         self.estimator_ = GridSearchCV(self.base_estimator,
                                        dict(n_components=components))
         self.estimator_.fit(X, y)
+        self.coef_ = self.estimator_.best_estimator_.coef_
         return self
 
     def predict(self, X):

### 2019-04-26 paper/paper.tex
roc auc test notebook, and diagonal non zero element test in paper.tex

@@
 
-\subsection{Problem formulation, asymptotic estimates}
+\subsection{Model formulation, average behaviour}
 
@@
 \end{equation}
-where F is an unknown matrix, presumably full or high rank, corresponding to the measurement apparatus, E is a low rank transformation carried on the features (X), and N a noise term, centred but not necessarily gaussian or diagonal.
+where F is an unknown matrix, presumably full or high rank, corresponding to the measurement apparatus, E is a low rank transformation carried on the ftkvdkkdtgbcifnndifftgetecfdbjujjeatures (X), and N a noise term, centred but not necessarily gaussian or diagonal.
 
@@
 
-(here be statistical test for the diagonal, and rationale for it)
-
 Formula (9) shows that double regression can be understood as a special case of canonical component analysis (CCA).
@@
 The closer we are to this hypothesis, the closer the results of CCA and double regression will be.
+
+\subsection{Double regression as feature extraction}
+
+Testing for non zero elements on the diagonal of (9) can be viewed as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).
+This can be done on maximising the ratio of inter-group variance over total variance (ie minimising intra-group variance).
+In a one dimensional two clusters setting, this amounts to sorting the data, and separating the p smallest from the n-p largest values so that inter group inertia is maximal.
+
+If s and r are the average values of the two clusters, and p and n-p their size, we are maximising 
+\begin{equation}
+\begin{aligned}
+{p (n-p)\over n^2} (s-r)^2 
+\end{aligned}
+\end{equation}
+ and selecting the p features corresponding to the largest values.
 

### 2019-04-27 paper/paper.tex
added Maxime, intro, double regression description, plan for experiments in paper.tex

@@
   \texttt{dlp@fb.com}
+  \And
+  Maxime Oquab\\
+  Facebook AI\\
+  \texttt{qas@fb.com}
 }
@@
     To be written.
+This is an example of citation
+
+A survey in deep learning \citep{deep_learning_nature}. 
 \end{abstract}
 
-\section{Introduction}
+\section{Inverse problems}
+Many causal problems are inverse problems: trying to tell causes from their effects.
+We are given experimental data (effects), possible explanations (causes), and are tasked to find the causes that account for the observations, knowing that other non causal (contingent, random) factors are at work.
 
-A survey in deep learning \citep{deep_learning_nature}.
+This can be modelled in the following way.
+We measure a phenomenon Y, via an apparatus F.
+We are given a set X of possible causal features, some of which are active (account for the observed effects).
+We note EX the subset of active features (E being a binary diagonal matrix, ie all zeroes except a few ones on the diagonal).
+Besides EX, the measured phenomenon is influenced by other factors, which we represent as noise N.
+Assuming that the noise is centred (equal to zero on average) and that dependences can be represented as linear transformations, we can summarize our model by the following equation: $Y = F(EX + N)$.
+Our goal, given observations of X and Y, is to find E.
 
-\subsection{Contributions}
+ **********Description of the neuroscience side of the problem***********
 
-\section{Math corner}
+\section{Double regression}
+A natural approach to inverse problems is anticausal regression, where one tries to predict the different features of X from the observations Y.
+Because of noise, loss of information in the measuring apparatus (F not being full rank, or well conditioned), possible correlations between features of X, and the fact that only a few of these features account for Y, such an approach will not retrieve X, but some approximation of it, $\hat X$.
+
+On the other hand, since we are given a statistical model for calculating Y from X, we could, by assuming some properties of N, use the opposite approach, and predict the different features of Y from X.
+Again this discounts the correlations between features of Y, the fact that not all features of X are active and the effect of noise.
+
+Double regression tries to combine both approaches by conducting first an anticausal regression, and then trying to predict its result $\hat X$ from X.
+In other words, the anticausal step (from Y to X) provides us with intermediary features, which the causak step tries to predict from X.
+Since matrices X and Y are likely to be multicollinear, stable estimators are needed, and we use ridge regression.
+Besides, since conducting two regressions from the same data sample would cause some of the shared noise to be learnt, we split our sample, and use half of it for each of the regressions, therefore "decorrelating" the two steps.
+
+Under our hypotheses (centred noise, E a binary diagonal), and even under heavy noise, experiments suggest that H provides a good approximation to E.
+In particular, H is close to diagonal, and binary clustering of its diagonal elements using a Sonquist and Morgan statistics, recovers the original E.
+
+In a nutshell, double regression is performed as follows:
+
+	1- Randomly split the data sample
+
+	2- From the first half sample, find G such that GY is the regularized least square approximation of X
+
+	3- From the second half of the sample, calculate $\hat X=GY$, and find H such that HX is the regularized least square approximation of $\hat X$
+
+	4- Using the Sonquist and Morgan criterion, select the largest diagonal elements of H, set them to one and the rest of H to zero
+
+	5- Use the resulting matrix as a estimator of E, the active features in the causal process
+
+	6- Alternatively, use recovered E to predict Y from X
+
+\section{Mathematics}
 
@@
 \end{equation}
-where F is an unknown matrix, presumably full or high rank, corresponding to the measurement apparatus, E is a low rank transformation carried on the ftkvdkkdtgbcifnndifftgetecfdbjujjeatures (X), and N a noise term, centred but not necessarily gaussian or diagonal.
+where F is an unknown matrix, presumably full or high rank, corresponding to the measurement apparatus, E is a low rank transformation carried on the features (X), and N a noise term, centred but not necessarily gaussian or diagonal.
 
@@
 
+\section{Experiments - simulated data}
+Experiments on simulated data serve three goals: compare double regression with other known techniques (CCA, OLS, PLS...) both as regression and feature extraction methods, provide evidence for our claim on stability, invariance to environment, and behavior as a causal detector, and understand the strengths and limitations of double regression, and directions for future improvement.
+
+\subsection{Experimenting with double regression}
+In these experiments, we focus on double regression as a feature detector.
+Given p features X, exhibiting various levels of correlations and observed on a sample of size n, matrix E extracts a (small) set of d features.
+A noise term N (p, n) of various covariance and signal to noise ratio (compared to EX) is then added, and the result is transformed by a random (q,p) matrix F, to yield an (q,n) Y.
+
+In data generation, we experiment on the size of feature spaces (p,q), the proportion of selected features (d/p), signal to noise ratios, and the conditioning of X (feature multicollinearity), N (noise covariance) and F.
+In the algorithm, we test various ways of regularizing the two regressions (namely pseudo inversion of X and Y), of splitting data samples between the regressions, of extracting the selected features, and of using the information recovered to improve prediction of Y.
+
+In all experiments, the figure of merit is the quality of the reconstruction of E, measured as the Hamming distance (total number of misclassifications) between the selection performed by E and the features extracted
+
+\subsection{Comparison with other methods - feature extraction}
+
+\subsection{Comparison with other methods - predicting Y}
+
+\subsection{Comparison with other methods - stable regression}
+In this section, we perform the same tests as in the previous one, but the test set is taken from a different distribution of X than the training set.
+In practice, we test the same trained predictors of Y over test samples generated fro; a distribution of X with a different covariance.
+
+The rationale behind such tests is that whereas double regression, being a feature extraction method, might not be as efficient as dedicated porediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.
+
+********** here be results and discussion ************
+
 \section{Experiments}

### 2019-05-07 code/models/jrr2.py
added jrr2, with sonquist morgan demo

@@ -1,0 +1,103 @@
+import numpy as np
+from sklearn.linear_model import RidgeCV
+from sklearn.model_selection import ShuffleSplit
+
+
+class JRR(object):
+    """Finds E in:
+        Y = F(EX+N)
+
+    Where Y is the recordings (samples * sensors),
+          F is an unknown forward operator (neurons * sensors),
+          X is the factors (samples * sources),
+          N is a source noise (samples * sources)
+
+    Step 1: Bagging
+    for set1, set2 in partitions:
+        Step 1.1: G = ls(Y_1, X_1)
+        Step 1.2: H = ls(X_2, Y_2 * G)
+    Step 2 (optional): E = diag(H)
+    Step 3: F = ls(XE, Y)
+    """
+
+    def __init__(self, G=None, H=None, F=None, bagging=10, zero_off_diag=True):
+        self.G = RidgeCV() if G is None else G
+        self.H = RidgeCV() if H is None else H
+        self.F = RidgeCV() if F is None else F
+        self.bagging = bagging
+        self.zero_off_diag = zero_off_diag
+
+    def fit(self, X, Y):
+        if self.bagging in (0, False, None):
+            Gset = range(len(X))
+            Hset = range(len(X))
+            ensemble = [(Gset, Hset)]
+        else:
+            bagging = ShuffleSplit(self.bagging, test_size=.5)
+            ensemble = [split for split in bagging.split(X, Y)]
+
+        H = list()
+        for train, test in ensemble:
+            X_hat = self.G.fit(Y[train], X[train]).predict(Y)
+            H += [self.H.fit(X[test], X_hat[test]).coef_, ]
+
+        # Estimate E
+        self.E_ = np.mean(H, 0).T  # TODO: check transpose
+        if self.zero_off_diag:
+            self.E_ = np.diag(np.diag(self.E_))
+        # Filter X by E hat and estimate F
+        self.F.fit(X @ self.E_, Y)
+        return self
+
+    def predict(self, X):
+        return self.F.predict(X @ self.E_)
+
+    def score(self, X, y, sample_weight=None):
+        from sklearn.metrics import r2_score
+        return r2_score(y, self.predict(X), sample_weight=sample_weight,
+                        multioutput='variance_weighted')
+
+
+if __name__ == '__main__':
[…]

### 2019-05-07 code/models/jrr2.py
added jrr2, with sonquist morgan demo

@@ -3,6 +3,24 @@ from sklearn.linear_model import RidgeCV
 from sklearn.model_selection import ShuffleSplit
 
 
+def sonquist_morgan(x):
+    z=np.sort(x)
+    n=z.size
+    m1=0
+    m2=np.sum(z)
+    mx=0
+    best=-1
+    for i in range(n-1): 
+        m1+=z[i]
+        m2-=z[i]
+        ind=(i+1)*(n-i-1)*(m1/(i+1)-m2/(n-i-1))**2
+        if ind>mx :
+            mx=ind
+            best=z[i]
+    K=mx/(np.var(x)*n) #significance indicator, not used for now
+    return 1*(x>(best+1.0e-6))
+
+
 class JRR(object):
     """Finds E in:
         Y = F(EX+N)
@@ -44,7 +62,8 @@ class JRR(object):
         # Estimate E
         self.E_ = np.mean(H, 0).T  # TODO: check transpose
         if self.zero_off_diag:
-            self.E_ = np.diag(np.diag(self.E_))
+            #self.E_ = np.diag(np.diag(self.E_))
+            self.E_ = np.diag(sonquist_morgan(np.diag(self.E_)))
         # Filter X by E hat and estimate F
         self.F.fit(X @ self.E_, Y)
         return self
@@ -99,5 +118,10 @@ if __name__ == '__main__':
     E_hat = jrr.fit(X[train], Y[train]).E_
     score = jrr.score(X[test], Y[test])
 
+    #compare extraction to E
+    extraction_score=np.sum(abs(np.diag(E)-np.diag(E_hat))
+
+    #WARNING: this might change now that E_at has been replaced by the sonquist version...
+    # it should make us better, as we don't care anymore about the order of diagonal coeffs, just their class
     print('E_auc', roc_auc_score(np.diag(E), np.diag(E_hat)))
     print('Y_score', score)

### 2019-05-07 paper/paper.tex
udated the paper, to be reviewed by David

--- 2019-04-12 paper/paper.tex
@@
 \section{Introduction}
+Causal discovery, telling causes from their effects, is an inverse problem.
+Given experimental data (effects) and possible explanations (causes), we are tasked to find the causes that account for the observations, knowing that other non causal factors (contingent, random) are at work.
 
-A survey in deep learning \citep{deep_learning_nature}.
+This can be modelled in the following way.
+Through an apparatus F, we measure a phenomenon Y, which result from a set of possible causes X and from unknown other factors that will be modelled as a noise N.
+Of all possible causes in X, only a few are active, and account for the observed Y.
+Let EX be the subset of such active features, E being a binary diagonal matrix (all zeroes except a few 1 on the diagonal).
+Assuming that noise is centred (by adding if necessary an intercept on top of the causal features, to capture the bias in N), and that all dependences can be represented as linear transformation, we can write our model as $Y = F(MEX + N)$.
+We are given experimental measures of X and Y, and our goal is to recover E, the causes of Y, with M, N and F unknown.
 
-\subsection{Contributions}
+Such problems can be attacked in two different ways.
+In the direct approach, we may use the above equation as a statistical model, try to express Y as a function of X, and test whether the coefficients associated with specific features of X are zero.
+Alternatively, we may try an inverse approach, and express various features of X as a function of Y, and estimate their variation as Y changes.
+In the presence of noise, and a large number of features of X, Y and N, both methods are limited, because they fail to take advantage of the correlations between features of the variable they are modelling (or guard against their multicollinearity).
+In the direct approach, features of Y are modelled separately (which also complicates testing for nullity of coefficients).
+In the indirect method, correlations between features of X, and possibly with various random factors, are unaccounted for.
+In effect, we do not retrieve matrix X (or an approximation of it) but some estimator $\hat X$ of it.
 
-\section{Math corner}
+In this paper, we introduce double regression, a method that combines both approaches.
+We first predict X from Y, as in the inverse approach, and retrieve an estimator $\hat X$, that we then try to predict from X.
+We prove that the estimator of $\hat X$, from X provides a good approximation of E, is robust to noise and environment change, and can be used either to extract causal features or to better predict Y from X.
+We expose the links between double regression and canonical component analysis, show that it compares favourably with other methods, test its limiting cases and the influence of various aspects of the problem (dimensionality, signal to noise ratios) and illustrate its use on an example from neurosciences.
 
-\subsection{Problem formulation, asymptotic estimates}
+\section{Double regression}
+We are given two set of measurements, over a sample of N individuals.
+Y are multidimensional effects, and X are multidimensional causal features.
+We believe that the values of Y result from a small number of features of X, denoted as EX, and some large multidimensional noise N, and try to recover E, the list of causal features.
 
-Our problem may be formulated as
-\begin{equation}
-    Y = F(EX + N)
-    \label{eq:model}
-\end{equation}
-where F is an unknown matrix, presumably full or high rank, corresponding to the measurement apparatus, E is a low rank transformation carried on the features (X), and N a noise term, centred but not necessarily gaussian or diagonal.
+Double regression first performs a least squares estimation of X from Y, that is finds $\hat G$ such that $\hat X=\hat G Y$ is the least square predictor of X.
+This retrieve some best approximation of the causes of Y in X.
+Since the covariance matrix of Y might exhibit multicollinearity (the effects measured in Y being correlated), we used a regularised method for estimation the least square solution (Ridge regression with the regularisation coefficient determined by cross validation).
 
-The first regression (X from Y), amounts to finding
+The second step of double regression performs a least square estimation of $\hat X$ fromX, that finds $\hat H$ such that $\hat \hat X=\hat H X$ is the least square predictor of $\hat X$.
+Again, to avoid the effects of multicollinearity in the covariance matrix of X, cross validated ridge regression is used once again.
+Also, to prevent spurious correlations between the two regression steps (should the be performed on the same sample), our original sample is randomly split into two halves, and each step is performed over one of them, effectively decorrelating the two regressions.
+Optionally, these two regressions can be repeated on many split samples, and their results averaged.
+Over small samples, such bagging compensates for the reduction in sample size due to splitting.
+
+Under our hypotheses (centred noise, E a binary diagonal), and even under heavy noise, $\hat H$ is close to diagonal, and provides, up to a scaling factor that depends on the dimensions of X, Y and N and the signal to noise ratio of the problem considered, a good approximation to E.
+To recover a binary diagonal matrix (an estimate of E), we extract the diagonal of $\hat H$ and select its largest elements using a binary clustering algorithm (AID, Sonquist and Morgan).
+The diagonal matrix $\hat E$, having the largest elements of H as its unit diagonal coefficients (the rest being zero) is our estimate of E.
+This allow us to retrieve the causal features of E.
+Since E and $\hat E$ are binary, the Hamming distance between both matrices (or their diagonal) quantifies the goodness of fit.
+
+To use double regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform (on a test sample) a cross validated ridge regression of each feature of Y from $\hat E X$.
+
+In a nutshell, double regression is performed as follows:
+\begin{enumerate}
+\item Randomly split the data sample.
+\item From the first half sample, find $\hat G$ such that $\hat X = \hat G Y$ is the regularised least square approximation of X.
+\item From the second half sample, calculate $\hat X = \hat G Y$, and find $\hat H$ such that $\hat H X$ is the regularized least square approximation of $\hat X$.
+\item (optional) Repeat steps 1 to 3 and average the results.
+\item Using the Sonquist and Morgan criterion, select the largest diagonal elements of $\hat H$, set them to one and the rest of $\hat  H$ to zero.
+\item Use the resulting matrix $\hat E$ as an estimator of E, the active features in the causal process
+\item Alternatively, use $\hat E$ to predict Y from X, by performing a regularised least square regression of Y from $\hat E X$. 
+\end{enumerate}
+
+\section{Related work}
+not sure whether this should be here of after the math part...
+
+We need to discuss other methods of feature extraction (if any) and list the regression methods we will compare to
+
+we should also discuss the way we can us regression methods for feature extraction (and the opposite)
+
+\section{The mathematics of double regression}
+
+\subsection{Model formulation and asymptotic behaviour}
+
+As seen in the introduction, our problem is modelled as $Y = F(MEX + N)$, with X an (N, dx) matrix of possible causes, E a (dx,dx) binary diagonal matrix, that selects active features from X.
+M an unknown (possibly full rank) (dz,dx) transformation of EX, N a (dz,N) homoscedastic noise matrix, F and unknown (possibly full rank) (dy,dz) transformation corresponding to the measuring apparatus, and Y a (dx,dy) matrix of measured effects.
+
+Since N is centred, the first least square regression (X from Y), amounts to finding
 \begin{equation}
@@
 \end{equation}
-since N is centred, and with k a positive constant which depends on the variance of FN.
+with k a positive constant which depends on the variance of FN.
 
-This amounts to finding the minimal norm generalized inverse of FE, or its Moore-Penrose inverse
-\begin{equation}
-\hat  G= (FE)^{\dagger}
-\end{equation}
+This amounts to finding the minimal norm generalized inverse of FE, or its Moore-Penrose inverse $\hat  G= (FE)^{\dagger}$.
 
@@
 \begin{aligned}
-\min_H \left \| (FE)^{\dagger}Y - HX \right \|^2 &= \min_H \left \| (FE)^{\dagger}F(EX+N) - HX \right \|^2 \\
-&= \min_H \left \| ((FE)^{\dagger}(FE)-H)X \right \| ^2 + \left \| (FE)^{\dagger}FN \right \| ^2\\
-&= \min_H \left \| ((FE)^{\dagger}(FE)-H)X \right \| ^2\\
-\end{aligned}
-\end{equation}
-
-Therefore, under the hypothesis that N is centred, our regression method estimates 
-\begin{equation}
-\begin{aligned}
-\hat H =(FE)^{\dagger}(FE)
+arg \min_H \left \| (FE)^{\dagger}Y - HX \right \|^2 &= arg \min_H \left \| (FE)^{\dagger}F(EX+N) - HX \right \|^2 \\
+&= arg \min_H \left \| ((FE)^{\dagger}(FE)-H)X \right \| ^2 + \left \| (FE)^{\dagger}FN \right \| ^2\\
+&= arg \min_H \left \| ((FE)^{\dagger}(FE)-H)X \right \| ^2\\
+&=(FE)^{\dagger}(FE)
 \end{aligned}
@@
 
-If E selects some features of X (ie $Ex=0$ on any unselected feature, and E full ranked over the set of selected features) and if F is full rank over the subspace spanned by the features selected by E, then
-\begin{equation}
-\begin{aligned}
-Ker(FE) = Ker(E)  
-\end{aligned}
-\end{equation}
-$I - \hat H$ is the projector over $Ker(E)$, the diagonal matrix that spans the unselected features, and $\hat H$ is the diagonal matrix corresponding to the selected features.
+In our case, $Ker(FE)\supseteq Ker(E)$.
+If F has full rank over the subspace spanned by the features selected by E (this will happen for almost all random F), then $Ker(FE) = Ker(E)$,.
+This means $I - \hat H$ is the projector over $Ker(E)$, the diagonal matrix that spans the unselected features, and $\hat H$ is the diagonal matrix corresponding to the selected features, that is E.
 
-Under such hypotheses, $\hat H$ retrieves the features of X that have a causal influence on Y, and should do it in a fairly resilient way, since it is diagonal, and all its eigenvalues (0 or 1) are well separated.
+Under such hypotheses, $\hat H$ retrieves E, the features of X that have a causal influence on Y.
+Since it is diagonal with eigenvalues well separated (0 and 1), it should do so in a fairly resilient way.
 
-\subsection{The double regression estimator, and canonical component analysis}
+\subsection{Double regression estimator, link with canonical component analysis}
 
-For any real matrices of X and Y, the least square regression of Y from X is calculated by the formula (X' denoting the transpose of X)
-\begin{equation}
-\begin{aligned}
-(X'X)^{-1} X'Y
-\end{aligned}
-\end{equation}
-and the prediction of Y thus obtained (from X) can be written
-\begin{equation}
-\begin{aligned}
-X(X'X)^{-1} X'Y
-\end{aligned}
-\end{equation}
+For any real matrices of X and Y, the least square regression of Y from X is calculated by the formula (X' denoting the transpose of X) $(X'X)^{-1} X'Y$ and the prediction of Y thus obtained (from X) can be written $X(X'X)^{-1} X'Y$.
+(To account for regularisation, a diagonal term should be added to the covariance matrix, which becomes $X'X+kI$, we do not mention it here, to simplify notation).
 
@@
 \end{equation}
-Under ideal conditions (good conditioning of $X'X$ and $Y'Y$), this should recover a diagonal binary matrix, the 1 on the diagonal corresponding the the features of X used by Y.
-In a noisy environment, double regression tries to extract the useful features of X (ie those having an influence on Y) as non zero elements of the diagonal of (9).
+Under ideal conditions (good conditioning of $X'X$ and $Y'Y$, no noise), this should recover a diagonal binary matrix, the 1 on the diagonal corresponding the the features of X used by Y.
+In a noisy environment, double regression tries to extract the useful features of X (ie those having an influence on Y) as non zero elements of the diagonal of (3).
 
-(here be statistical test for the diagonal, and rationale for it)
-
-Formula (9) shows that double regression can be understood as a special case of canonical component analysis (CCA).
+Formula (3) shows that double regression can be understood as a special case of canonical component analysis (CCA).
 In CCA, given two matrixes (n,p and n,q) X and Y describing different features measured on the same data sample, one tries to characterise the closeness between X and Y (ie between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.
 This is done by finding vectors $a$ and $b$ such that $Xa$ and $Yb$ display maximal correlation, or by maximising $a'X'Yb$ over unit-normed $a$ and $b$.
-To find a, one calculates the eigenvectors of (9).
+To find a, one calculates the eigenvectors of (3).
 The largest one is the direction with maximal correlation, the second one the maximal correlation direction orthogonal to the first, and so on.
@@
 
-If (9) is diagonal, for instance in the model presented above, the features of X are eigenvectors.
+If (3) is diagonal, for instance in the model presented above, the features of X are eigenvectors.
 If all eigenvalues are either 0 or 1, since they correspond to the squared cosine of the angle between the corresponding feature and the subspace spanned by Y, all the features are either part of the Y subspace (eigenvalues of one) or orthogonal to it (for zero eigenvalues).
-
 Thus, double regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
@@
 
+\subsection{Double regression as feature extraction}
+
+Recovering non zero elements on the diagonal of (3) could be viewed as a statistical test problem.
+Since $\hat H$ is the coefficient matrix of a least square regression, a one-sided t value test, as used in regression analyses, could be used.
+However, this supposes that the coefficients of $\hat H$ are centered and follow a chi square distribution, something our method does not guarantee.
+Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).
+This can be done by maximising the ratio of inter-group variance over total variance (ie minimising intra-group variance).
+In a one dimensional two clusters setting, this amounts to sorting the data, and separating the p smallest from the n-p largest values so that inter group inertia is maximal.
+
+If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we are maximising the Sonquist and Morgan criterion $$K = {p (n-p) \over n} {(s-r)^2 \over V}$$ and selecting the p features corresponding to the largest values.
+If diagonal elements of (3) are normally distributed, K follows a chi-square distribution with one degree of freedom.
+At 95\% confidence level, a value of K superior to 3.84 makes our selection significant (for 99\% confidence, the corresponding value of K is 6.63) \citep{Kass_75}
+
 \subsection{Stability analysis}
 
-\subsection{Related approaches}
-(PLS, 2 way, 3 fold...)
+\section{Experiments - simulated data}
+Experiments on simulated data serve three goals: compare double regression with other known techniques (CCA, OLS, PLS...) both as regression and feature extraction methods, provide evidence for our claims on stability, invariance to environment, and behavior as a causal detector, and understand the strengths and limitations of double regression, and directions for future improvement.
+
+The model used for simulation is $Y=F(snr MEX+N)$, with X a random (dx,N) matrix, E a square binary diagonal of dimension dx with the nc last elements equal to 1 (this means that of the dx features of X, only the nc last have some influence on Y), M, N and F (dz,dx), (dz,N) and (dy,dz) random matrices.
+snr is a real number measuring the amount of noise.
+Our simulation therefore depends upon 6 free parameters :
+
+\begin{itemize}
+\item dx : number of features of X
+\item nc: number of active features (in EX)
+\item dz: number of noise features
+\item snr: signal to noise ratio
+\item dy: number of features in Y
+\item N: size of sample
+\end{itemize}
+
+Random matrices X, F, N and M are populated with centred gaussian random coefficients with unit variance, divided by $ \surd n$ where n is the smallest dimension of the matrix.
+Since a random square matrix with unit gaussian coefficients of dimension n has eigenvalues between $ \surd n $ and $\surd n $, this allows us to compare problems with different dimensions (values of dx, dy, dz and N).
+
+We evaluate the performance of double regression and other algorithms in two different roles : as a feature extractor, where we try to recover the non-zero elements of E (the active features of X), and as a regression method, where we try to predict over a test sample the values of Y from those of X.
+Feature extraction produces a list of active features, ie a binary vector of size dx.
+Its goodness of fit is judged by comparing this vector with the diagonal of E, and counting the number of disagreements (Hamming distance which should be minimised).
+For regression, we use a test set of N additional measurements, and calculate the R-squared ratio of variances (which should be maximised).
+
+\subsection{Experimenting with double regression}
+In these experiments, we focus on double regression as a feature detector.
+Given p features X, exhibiting various levels of correlations and observed on a sample of size n, matrix E extracts a (small) set of d features.
+A noise term N (p, n) of various covariance and signal to noise ratio (compared to EX) is then added, and the result is transformed by a random (q,p) matrix F, to yield an (q,n) Y.
+
+In data generation, we experiment on the size of feature spaces (p,q), the proportion of selected features (d/p), signal to noise ratios, and the conditioning of X (feature multicollinearity), N (noise covariance) and F.
+In the algorithm, we test various ways of regularizing the two regressions (namely pseudo inversion of X and Y), of splitting data samples between the regressions, of extracting the selected features, and of using the information recovered to improve prediction of Y.
+
+In all experiments, the figure of merit is the quality of the reconstruction of E, measured as the Hamming distance (total number of misclassifications) between the selection performed by E and the features extracted
+
+\subsection{Comparison with other methods - feature extraction}
+
+\subsection{Comparison with other methods - predicting Y}
+
+\subsection{Comparison with other methods - stable regression}
+In this section, we perform the same tests as in the previous one, but the test set is taken from a different distribution of X than the training set.
+In practice, we test the same trained predictors of Y over test samples generated fro; a distribution of X with a different covariance.
+\
+The rationale behind such tests is that whereas double regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.
+
+********** here be results and discussion ************
 

### 2019-05-08 paper/paper.tex
experiments

@@
 
-\section{Related work}
-not sure whether this should be here of after the math part...
-
-We need to discuss other methods of feature extraction (if any) and list the regression methods we will compare to
-
-we should also discuss the way we can us regression methods for feature extraction (and the opposite)
+\section{Related works}
+FC: this section either here or after the maths, unsure... 
+This should list other methods of feature extraction, if any (I don't know...)
+And discuss similar methods of regression, that can be used for feature extraction
+And maybe have something about other causal discovery methods
 
@@
 \begin{aligned}
-\min_G \left \| X-GY \right \|^2 &= \min_G \left \| X - GF(EX+N)\right\|^2 \\
-&{}= \min_G \left \| (I-GFE)X\right\| ^2 + \left \| GFN\right \| ^2\\
-&{}\leq \min_G \left \| I-GFE\right\| ^2 \left \| X\right\| ^2 + k\left \| G\right \| ^2\\ 
+\min_G \left \| X-GY \right \|^2 &= \min_G \left \| X - GF(MEX+N)\right\|^2 \\
+&{}= \min_G \left \| (I-GFME)X\right\| ^2 + \left \| GFN\right \| ^2\\
+&{}\leq \min_G \left \| I-GFME\right\| ^2 \left \| X\right\| ^2 + k\left \| G\right \| ^2\\ 
 \end{aligned}
@@
 
-This amounts to finding the minimal norm generalized inverse of FE, or its Moore-Penrose inverse $\hat  G= (FE)^{\dagger}$.
+This amounts to finding the minimal norm generalized inverse of FME, or its Moore-Penrose inverse $\hat  G= (FME)^{\dagger}$.
 
@@
 \begin{aligned}
-arg \min_H \left \| (FE)^{\dagger}Y - HX \right \|^2 &= arg \min_H \left \| (FE)^{\dagger}F(EX+N) - HX \right \|^2 \\
-&= arg \min_H \left \| ((FE)^{\dagger}(FE)-H)X \right \| ^2 + \left \| (FE)^{\dagger}FN \right \| ^2\\
-&= arg \min_H \left \| ((FE)^{\dagger}(FE)-H)X \right \| ^2\\
-&=(FE)^{\dagger}(FE)
+arg \min_H \left \| (FME)^{\dagger}Y - HX \right \|^2 &= arg \min_H \left \| (FME)^{\dagger}F(MEX+N) - HX \right \|^2 \\
+&= arg \min_H \left \| ((FME)^{\dagger}(FME)-H)X \right \| ^2 + \left \| (FME)^{\dagger}FN \right \| ^2\\
+&= arg \min_H \left \| ((FME)^{\dagger}(FME)-H)X \right \| ^2\\
+&=(FME)^{\dagger}(FME)
 \end{aligned}
@@
 
-In our case, $Ker(FE)\supseteq Ker(E)$.
-If F has full rank over the subspace spanned by the features selected by E (this will happen for almost all random F), then $Ker(FE) = Ker(E)$,.
+In our case, $Ker(FME)\supseteq Ker(E)$.
+If FM has full rank over the subspace spanned by the features selected by E (this is assumed for M and will happen for almost all random F), then $Ker(FME) = Ker(E)$,.
 This means $I - \hat H$ is the projector over $Ker(E)$, the diagonal matrix that spans the unselected features, and $\hat H$ is the diagonal matrix corresponding to the selected features, that is E.
@@
 
-\subsection{Double regression as feature extraction}
+\subsection{Feature extraction}
 
@@
 
+FC need to add something about values of $\hat H$, why active diagonal coefficients are not close to one, why inactive coeffs are not zero centered...  
+Discussion about bias induced by regularisation goes here
+
 \subsection{Stability analysis}
+FC: to be written, at a loss right now, but two weeks left...
 
@@
 Random matrices X, F, N and M are populated with centred gaussian random coefficients with unit variance, divided by $ \surd n$ where n is the smallest dimension of the matrix.
-Since a random square matrix with unit gaussian coefficients of dimension n has eigenvalues between $ \surd n $ and $\surd n $, this allows us to compare problems with different dimensions (values of dx, dy, dz and N).
+Since a random square matrix with unit gaussian coefficients of dimension n has eigenvalues between $ \surd n $ and $\surd n $, this allows meaningful comparison of results when dimensions N, dx, dy and dz vary.
 
-We evaluate the performance of double regression and other algorithms in two different roles : as a feature extractor, where we try to recover the non-zero elements of E (the active features of X), and as a regression method, where we try to predict over a test sample the values of Y from those of X.
-Feature extraction produces a list of active features, ie a binary vector of size dx.
-Its goodness of fit is judged by comparing this vector with the diagonal of E, and counting the number of disagreements (Hamming distance which should be minimised).
+We evaluate the performance of double regression in two different roles : as a feature extractor, recovering the non-zero elements of E (the active features of X), and as a regression method, predicting Y from X over a test sample.
+For feature extraction, the quality of the fit is measured as the number of disagreements between our estimator and the diagonal of E (ie, their Hamming distance, which should be minimised).
 For regression, we use a test set of N additional measurements, and calculate the R-squared ratio of variances (which should be maximised).
@@
 \subsection{Experimenting with double regression}
-In these experiments, we focus on double regression as a feature detector.
-Given p features X, exhibiting various levels of correlations and observed on a sample of size n, matrix E extracts a (small) set of d features.
-A noise term N (p, n) of various covariance and signal to noise ratio (compared to EX) is then added, and the result is transformed by a random (q,p) matrix F, to yield an (q,n) Y.
+In these experiments, we try to understand how various parameters of double regression (dimension, active features, signal to noise ratio) influence the quality of feature extraction and regression.
+For regression, we are also investigating different ways of taking extracted features into account, and the impact of varying environments (having training and test samples from different distributions).
 
-In data generation, we experiment on the size of feature spaces (p,q), the proportion of selected features (d/p), signal to noise ratios, and the conditioning of X (feature multicollinearity), N (noise covariance) and F.
-In the algorithm, we test various ways of regularizing the two regressions (namely pseudo inversion of X and Y), of splitting data samples between the regressions, of extracting the selected features, and of using the information recovered to improve prediction of Y.
+Working from a fixed sample of 1000, and 15 active features, and varying values of dx, dy, dz (from 50 to 200) and snr (from 0.2 to 1), we observe that, all other factors being equal, the quality of feature extraction and regression decrease when dx increases (X has more inactive features), dz increase (more noise dimensions) and snr decreases (less signal, more noise).
+The dimensionality of Y plays a much smaller role, except for very small values of dy (less than 10).
 
-In all experiments, the figure of merit is the quality of the reconstruction of E, measured as the Hamming distance (total number of misclassifications) between the selection performed by E and the features extracted
+For feature extraction, double regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
+The following graphs show the boundary between values of dx and dz that allow for features extractions (dx lower than boundary, or dz larger), for different values of dy (level curves on the graph), and different values of signal to noise ratio and number of active variables.
 
-\subsection{Comparison with other methods - feature extraction}
+FC: add graphs here, see whether it is better to have nc fixed, or a proportion of dx, note : part of this can be moved to appendix if lack of space
 
-\subsection{Comparison with other methods - predicting Y}
+For values of ... in ..., an empirical formula can be provided. Double regression allows for good feature extraction if :
 
-\subsection{Comparison with other methods - stable regression}
+For regression, we want to understand how feature extraction helps predict Y from X. To this effect, we compare four approaches to regression:
+\begin{itemize}
+\item standard ridge regression of Y from X, where no features are extracted this is our baseline
+\item weighted ridge regression, of Y from $\hat H X$, where features of X are weighted according to the diagonal of $\hat H$
+\item filtered ridge regression, of Y from $\hat E X$, where active features are first extracted (as above), and Y is predicted from the extracted features only  
+\item perfect filtered (oracle) ridge regression of Y from EX, where E is supposed known 
+\end{itemize}
+Over a large set of parameters, we observe that filtered regression is on average better than weighted, which is on average better than standard regression.
+This demonstrates that double regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.
+
+FC graphs go here
+
+The quality of regression is independent from dy (except for very small values), increase with snr, decreases with dx, and increases with dz if it is larger than dx 
+FC graphs go here
+
+\subsection{Comparison with other methods}
+Maxime territory
+
+\subsection{Double regression in multiple environments}
 In this section, we perform the same tests as in the previous one, but the test set is taken from a different distribution of X than the training set.
-In practice, we test the same trained predictors of Y over test samples generated fro; a distribution of X with a different covariance.
-\
+In practice, we test the same trained predictors of Y over test samples generated from a distribution of X with a different covariance.
+
 The rationale behind such tests is that whereas double regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.
 
-********** here be results and discussion ************
+FC: ideally, we'd like to assess the multiple env performance of jrr, and compare it with others.
+Are multiple environments limited to regression analyses?
+How should environments be generated (not just variance of X, I expect)?
 
-\section{Experiments}
+\section{Experiments - real case}
+Jean Remi and Maxime territory
 
@@
 
+\section{Appendices}
+
+This should hold part of the tests, explanations on simulation, detailed results and stuff on meg data
+
 \end{document}

### 2019-05-12 paper/paper.tex
better structure from david comments, improved intro and description of jrr, began experiments

@@
 
-This can be modelled in the following way.
-Through an apparatus F, we measure a phenomenon Y, which result from a set of possible causes X and from unknown other factors that will be modelled as a noise N.
-Of all possible causes in X, only a few are active, and account for the observed Y.
-Let EX be the subset of such active features, E being a binary diagonal matrix (all zeroes except a few 1 on the diagonal).
-Assuming that noise is centred (by adding if necessary an intercept on top of the causal features, to capture the bias in N), and that all dependences can be represented as linear transformation, we can write our model as $Y = F(MEX + N)$.
-We are given experimental measures of X and Y, and our goal is to recover E, the causes of Y, with M, N and F unknown.
+More precisely, we measure, via an apparatus F, a multidimensional signal Y, that is a function of possible causes X, and other, unknown, factors that can be understood as noise N: $Y=f(X,N)$.
+Of all possible cause (features of X), only a few account for Y, let them be represented as EX, with E a binary diagonal matrix: all zeroes except a few ones on the diagonal, which correspond to active causal features of X.
+Let us assume that noise N is centred (adding, if need be, an intercept to the active causal features, to capture the bias in N), and additive (not a very strong assumption in the absence of further constraints on N).
+Then, we can write our model of causal discovery as $Y=f(m(EX)+N)$, with f and m unknown functions.
+Up to some rearrangement of the features of X, we can also suppose m linear.
 
-Such problems can be attacked in two different ways.
-In the direct approach, we may use the above equation as a statistical model, try to express Y as a function of X, and test whether the coefficients associated with specific features of X are zero.
-Alternatively, we may try an inverse approach, and express various features of X as a function of Y, and estimate their variation as Y changes.
-In the presence of noise, and a large number of features of X, Y and N, both methods are limited, because they fail to take advantage of the correlations between features of the variable they are modelling (or guard against their multicollinearity).
-In the direct approach, features of Y are modelled separately (which also complicates testing for nullity of coefficients).
-In the indirect method, correlations between features of X, and possibly with various random factors, are unaccounted for.
-In effect, we do not retrieve matrix X (or an approximation of it) but some estimator $\hat X$ of it.
+We end up with the following model $Y=f(MEX+N)$, where N is centred noise E a binary diagonal matrix, M a full rank matrix and f some unknown transformation.
+Given experimental measures of X and Y (as matrices), and our goal is to recover E, the active causal features of the process under study.
 
-In this paper, we introduce double regression, a method that combines both approaches.
-We first predict X from Y, as in the inverse approach, and retrieve an estimator $\hat X$, that we then try to predict from X.
-We prove that the estimator of $\hat X$, from X provides a good approximation of E, is robust to noise and environment change, and can be used either to extract causal features or to better predict Y from X.
-We expose the links between double regression and canonical component analysis, show that it compares favourably with other methods, test its limiting cases and the influence of various aspects of the problem (dimensionality, signal to noise ratios) and illustrate its use on an example from neurosciences.
+*** description of the neuroscience problem
+
+Such problems have been attacked along two different lines.
+In the direct approach, we use the above equation as a statistical model, try to express each feature of Y as a function of X, test whether the coefficients associated with specific features of X are different from zero, and recover E as the features of X active over some dimension of Y.
+In the inverse approach, we try to express each feature of X as a function of Y, estimate its changes as Y varies, and recover E as the set of "non random" features of X.
+In the presence of noise and a large number of features of X, Y and N, both approaches are limited because they fail to take advantage of the correlations between features of the variable they are modelling (or guard against their multicollinearity).
+In the direct approach, features of Y are modelled separately.
+In the inverse approaches, correlations between features of X are not accounted for.
+
+In this paper, we introduce double regression, which combines direct and inverse approaches.
+We first predict X from Y, as in the inverse approach, and retrieve a regularised least square estimate $\hat X$.
+Then, we calculate a regularised least square estimator of $\hat X$ from X.
+If the active features of X are represented as EX, with E a binary diagonal, this provides a good approximation of E, robust to noise and environment change.
+The main purpose of double regression is feature extraction, but it can be turned into a regression method by filtering X on the selected features.
+
+In the next sections, we describe double regression, prove that it asymptotically extracts E, both in the linear and non linear case, and that it can be understood as a special case of canonical component analysis (cca).
+By experimenting with simulated data, we demonstrate its robustness and use cases, and show that it compares favourably with state of the art techniques for regression and feature extraction.
+Finally, we illustrate its use on an example from neuroscience.
 
 \section{Double regression}
-We are given two set of measurements, over a sample of N individuals.
-Y are multidimensional effects, and X are multidimensional causal features.
-We believe that the values of Y result from a small number of features of X, denoted as EX, and some large multidimensional noise N, and try to recover E, the list of causal features.
+\subsection{The method}
+We are given two set of measurements over a sample of N individuals.
+Y are ny multidimensional effects, and X are nx multidimensional possible causes.
+We assume that only some features of X influence Y, so that $Y=f(EX,N)$, E being a binary diagonal, N some unknown noise, and f an unknown function.
+We try to recover E, the active causal features.
 
-Double regression first performs a least squares estimation of X from Y, that is finds $\hat G$ such that $\hat X=\hat G Y$ is the least square predictor of X.
-This retrieve some best approximation of the causes of Y in X.
-Since the covariance matrix of Y might exhibit multicollinearity (the effects measured in Y being correlated), we used a regularised method for estimation the least square solution (Ridge regression with the regularisation coefficient determined by cross validation).
+Double regression consists in two steps.
+First, we perform a least squares estimation of X from Y, that is we find $\hat G$ such that $\hat X=\hat G Y$ is the least square predictor of X.
+This recovers an approximation of the causes of Y in X.
+Then, we perform a least squares estimation of $\hat X$ from X, and determine $\hat H$ such that $\hat {\hat X}=\hat H X$ is the least square predictor of $\hat X$.
+In matrix form, we have $\hat G=(Y'Y)^{-1} Y'X$, and
+\begin{equation} \hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X\end{equation}
+Under good conditions (close to linear dependence between X and Y, and covariance matrices of X and Y well conditioned), this recovers a close approximation to E.
 
-The second step of double regression performs a least square estimation of $\hat X$ fromX, that finds $\hat H$ such that $\hat \hat X=\hat H X$ is the least square predictor of $\hat X$.
-Again, to avoid the effects of multicollinearity in the covariance matrix of X, cross validated ridge regression is used once again.
-Also, to prevent spurious correlations between the two regression steps (should the be performed on the same sample), our original sample is randomly split into two halves, and each step is performed over one of them, effectively decorrelating the two regressions.
-Optionally, these two regressions can be repeated on many split samples, and their results averaged.
-Over small samples, such bagging compensates for the reduction in sample size due to splitting.
+In practice, since the covariance matrices of Y and X might exhibit multicollinearity, we use regularisation in both matrix inversions.
+This is done by using ridge regression, and determining at each step the regularisation parameter by cross validation.
+In the above formula, this amounts to replacing $(X'X)^{-1}$ and $(Y'Y)^{-1}$ by $(X'X+\lambda)^{-1}$ and $(Y'Y+\mu)^{-1}$.
+Also, to prevent spurious correlations between the two regressions (should they be performed on the same sample), our original sample is randomly split into two halves, and each step is performed over one of them, effectively decorrelating the two steps.
+This is repeated over several (10) splits, and the resulting $\hat H$ are averaged.
+This bagging compensates for the reduction in sample size caused by splitting.
 
-Under our hypotheses (centred noise, E a binary diagonal), and even under heavy noise, $\hat H$ is close to diagonal, and provides, up to a scaling factor that depends on the dimensions of X, Y and N and the signal to noise ratio of the problem considered, a good approximation to E.
-To recover a binary diagonal matrix (an estimate of E), we extract the diagonal of $\hat H$ and select its largest elements using a binary clustering algorithm (AID, Sonquist and Morgan).
-The diagonal matrix $\hat E$, having the largest elements of H as its unit diagonal coefficients (the rest being zero) is our estimate of E.
-This allow us to retrieve the causal features of E.
-Since E and $\hat E$ are binary, the Hamming distance between both matrices (or their diagonal) quantifies the goodness of fit.
+Under our hypotheses (centred noise, E a binary diagonal) and even under heavy noise, $\hat H$ is close to diagonal.
+Yet, because of regularisation which adds positive bias to small eigenvalues, and noise which attenuates large eigenvalues, $\hat H$ is in fact a scaled and noisy approximation to E.
+To recover a binary diagonal, as our estimate of E, we extract the diagonal of $\hat H$ and select its large elements.
+The diagonal matrix $\hat E$, having large elements of H as its unit diagonal coefficients (the rest being zero) is our estimate of E, the active features of X.
 
-To use double regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform (on a test sample) a cross validated ridge regression of each feature of Y from $\hat E X$.
+To select large elements of $\hat H$, the traditional approach in regression analysis consists in testing whether those coefficient differ from zero (or are larger from zero since we are only interested in large positive coefficients).
+This is usually done via Student t-test of variants of it.
+This will not work here, as double regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
+Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).
 
-In a nutshell, double regression is performed as follows:
+This can be done by maximising the ratio of inter-group variance over total variance (ie minimising intra-group variance).
+In a one dimensional two clusters setting, this amounts to sorting the data, and separating the p smallest from the n-p largest values so that inter group inertia is maximal.
+If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we are maximising the Sonquist and Morgan criterion $$K = {p (n-p) \over n} {(s-r)^2 \over V}$$ and selecting the p features corresponding to the largest values.
+If diagonal elements of $\hat H$ are normally distributed, K follows a chi-square distribution with one degree of freedom.
+At 95\% confidence level, a value of K superior to 3.84 makes our selection significant (for 99\% confidence, the corresponding value of K is 6.63) \citep{Kass_75}
+
+This extract from $\hat H$ a binary diagonal $\hat E$.
+To use double regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
+
+Summarizing, double regression is performed as follows:
 \begin{enumerate}
 \item Randomly split the data sample.
-\item From the first half sample, find $\hat G$ such that $\hat X = \hat G Y$ is the regularised least square approximation of X.
-\item From the second half sample, calculate $\hat X = \hat G Y$, and find $\hat H$ such that $\hat H X$ is the regularized least square approximation of $\hat X$.
-\item (optional) Repeat steps 1 to 3 and average the results.
-\item Using the Sonquist and Morgan criterion, select the largest diagonal elements of $\hat H$, set them to one and the rest of $\hat  H$ to zero.
+\item From the first half sample, using crossalidated ridge regression, find $\hat G$, the regularised least square regression of X from Y: $\hat G=(Y'Y)^{-1} Y'X$.
+\item From the second half sample, calculate $\hat X = \hat G Y$, and find $\hat H$, the regularized least square regression of $\hat X$ from X: $\hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X$.
+\item Repeat steps 1 to 3 and average the results as $\hat H$.
+\item Using the Sonquist and Morgan criterion, select the large diagonal elements of $\hat H$, set them to one and the rest of $\hat  H$ to zero.
 \item Use the resulting matrix $\hat E$ as an estimator of E, the active features in the causal process
@@
 
-\section{Related works}
-FC: this section either here or after the maths, unsure... 
-This should list other methods of feature extraction, if any (I don't know...)
-And discuss similar methods of regression, that can be used for feature extraction
-And maybe have something about other causal discovery methods
-
-\section{The mathematics of double regression}
-
-\subsection{Model formulation and asymptotic behaviour}
-
-As seen in the introduction, our problem is modelled as $Y = F(MEX + N)$, with X an (N, dx) matrix of possible causes, E a (dx,dx) binary diagonal matrix, that selects active features from X.
-M an unknown (possibly full rank) (dz,dx) transformation of EX, N a (dz,N) homoscedastic noise matrix, F and unknown (possibly full rank) (dy,dz) transformation corresponding to the measuring apparatus, and Y a (dx,dy) matrix of measured effects.
+\subsection{Asymptotic behaviour - linear case}
+As discussed in the introduction, our problem can be formulated (up to a redifinition of the features of X) as $Y=f(MEX+N)$, with F an unknown function.
+If F is linearm we can rewrite it as $Y = F(MEX + N)$, with X an (N, dx) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
+M an unknown full rank (dz,dx) transformation of EX, N a (dz,N) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dz) transformation corresponding to the measuring apparatus, and Y a (dx,dy) matrix of measured effects.
 
@@
 \end{equation}
-with k a positive constant which depends on the variance of FN.
-
-This amounts to finding the minimal norm generalized inverse of FME, or its Moore-Penrose inverse $\hat  G= (FME)^{\dagger}$.
+with k a positive constant which depends on the variance of FN. This amounts to finding the minimal norm generalized inverse of FME, or its Moore-Penrose inverse $\hat  G= (FME)^{\dagger}$.
 
 The second regression is from X to the predicted value of X in the above, or $\hat G Y$
-
 \begin{equation}
@@
 \end{equation}
+Thus, in the linear case, double regression retrieves $(FME)^{\dagger}(FME)$.
 
 For any matrix A, $A^\dagger A$ is the orthogonal projector over $Img(A')$, and $I-A^\dagger A$ the orthogonal projector over $Ker(A)$.
-
 In our case, $Ker(FME)\supseteq Ker(E)$.
-If FM has full rank over the subspace spanned by the features selected by E (this is assumed for M and will happen for almost all random F), then $Ker(FME) = Ker(E)$,.
+If FM has full rank over the subspace spanned by the features selected by E (this is assumed for M and will happen for almost any random F), then $Ker(FME) = Ker(E)$,.
 This means $I - \hat H$ is the projector over $Ker(E)$, the diagonal matrix that spans the unselected features, and $\hat H$ is the diagonal matrix corresponding to the selected features, that is E.
-
 Under such hypotheses, $\hat H$ retrieves E, the features of X that have a causal influence on Y.
@@
 
-\subsection{Double regression estimator, link with canonical component analysis}
+\subsection{Asymptotic behaviour - non linear case}
 
-For any real matrices of X and Y, the least square regression of Y from X is calculated by the formula (X' denoting the transpose of X) $(X'X)^{-1} X'Y$ and the prediction of Y thus obtained (from X) can be written $X(X'X)^{-1} X'Y$.
-(To account for regularisation, a diagonal term should be added to the covariance matrix, which becomes $X'X+kI$, we do not mention it here, to simplify notation).
+\subsection{Stability analysis}
 
-In double regression, we first calculate the regression of X from Y, and calculate $\hat X$ the corresponding prediction as $Y(Y'Y)^{-1} Y'X$. The regression of $\hat X$ from X then becomes
-\begin{equation}
-\begin{aligned}
-(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X
-\end{aligned}
-\end{equation}
-Under ideal conditions (good conditioning of $X'X$ and $Y'Y$, no noise), this should recover a diagonal binary matrix, the 1 on the diagonal corresponding the the features of X used by Y.
-In a noisy environment, double regression tries to extract the useful features of X (ie those having an influence on Y) as non zero elements of the diagonal of (3).
+\section{Related works}
+\subsection{Double regression as a special case of canonical component analysis}
+Formula (1) shows that double regression can be seen as a special case of canonical component analysis (CCA).
+In CCA, we are given two matrixes X and Y (or size N,p and N,q) that describe different features measured on the same data sample, and characterise their proximity (that is, between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.
 
-Formula (3) shows that double regression can be understood as a special case of canonical component analysis (CCA).
-In CCA, given two matrixes (n,p and n,q) X and Y describing different features measured on the same data sample, one tries to characterise the closeness between X and Y (ie between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.
 This is done by finding vectors $a$ and $b$ such that $Xa$ and $Yb$ display maximal correlation, or by maximising $a'X'Yb$ over unit-normed $a$ and $b$.
-To find a, one calculates the eigenvectors of (3).
+To find a, one calculates the eigenvectors of (1).
 The largest one is the direction with maximal correlation, the second one the maximal correlation direction orthogonal to the first, and so on.
@@
 
-If (3) is diagonal, for instance in the model presented above, the features of X are eigenvectors.
+If (1) is diagonal, for instance in the model presented above, the features of X are eigenvectors.
 If all eigenvalues are either 0 or 1, since they correspond to the squared cosine of the angle between the corresponding feature and the subspace spanned by Y, all the features are either part of the Y subspace (eigenvalues of one) or orthogonal to it (for zero eigenvalues).
@@
 
-\subsection{Feature extraction}
+There are differences between CCA and double regression.
+First, since our objective is not to characterise the proximity between X and Y, but just to extract causal features, we can dispense with the last step of CCA (diagonalisation of (1)).
+Second, since we operate in a noisy environment, we use regularisation, introducing bias in the correlation calculations.
+Finally, whereas CCA wants to leverage all correlations between X and Y, we filter some of them, through splitting and bagging.
+In other words, whereas we are working from the same formulae as CCA, we use them in different ways, for different purposes
 
-Recovering non zero elements on the diagonal of (3) could be viewed as a statistical test problem.
-Since $\hat H$ is the coefficient matrix of a least square regression, a one-sided t value test, as used in regression analyses, could be used.
-However, this supposes that the coefficients of $\hat H$ are centered and follow a chi square distribution, something our method does not guarantee.
-Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).
-This can be done by maximising the ratio of inter-group variance over total variance (ie minimising intra-group variance).
-In a one dimensional two clusters setting, this amounts to sorting the data, and separating the p smallest from the n-p largest values so that inter group inertia is maximal.
-
-If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we are maximising the Sonquist and Morgan criterion $$K = {p (n-p) \over n} {(s-r)^2 \over V}$$ and selecting the p features corresponding to the largest values.
-If diagonal elements of (3) are normally distributed, K follows a chi-square distribution with one degree of freedom.
-At 95\% confidence level, a value of K superior to 3.84 makes our selection significant (for 99\% confidence, the corresponding value of K is 6.63) \citep{Kass_75}
-
-FC need to add something about values of $\hat H$, why active diagonal coefficients are not close to one, why inactive coeffs are not zero centered...  
-Discussion about bias induced by regularisation goes here
-
-\subsection{Stability analysis}
-FC: to be written, at a loss right now, but two weeks left...
+\subsection{Related regression methods}
+FC: discuss similar methods of regression, that can be used for feature extraction
+And maybe have something about other causal discovery methods
 
@@
 
-The model used for simulation is $Y=F(snr MEX+N)$, with X a random (dx,N) matrix, E a square binary diagonal of dimension dx with the nc last elements equal to 1 (this means that of the dx features of X, only the nc last have some influence on Y), M, N and F (dz,dx), (dz,N) and (dy,dz) random matrices.
-snr is a real number measuring the amount of noise.
+The model used for simulation is $Y=f(snr MEX+N)$, with X a random (dx,N) matrix, E a square binary diagonal of dimension dx with the nc last elements equal to 1 (this means that of the dx features of X, only the nc last have some influence on Y), M and N (dz,dx) and (dz,N) random matrices, and f a function from $\mathbb{R}^{dz}$ to $\mathbb{R}^{dy}$, that can be linear or non linear.
+In the linear case, the model becomes $Y=F(MEX+N)$, with F a (dy,dz) matrix.
+In the non linear case, F becomes $FSH$ where F and H are matrices of dimensions (dz,dz) and (dy,dz), and S a vector of dz sigmoid transformations.
+Finally, $snr$ is a real number measuring the amount of noise.
+
 Our simulation therefore depends upon 6 free parameters : 
-
 \begin{itemize}
@@
 
-Random matrices X, F, N and M are populated with centred gaussian random coefficients with unit variance, divided by $ \surd n$ where n is the smallest dimension of the matrix.
-Since a random square matrix with unit gaussian coefficients of dimension n has eigenvalues between $ \surd n $ and $\surd n $, this allows meaningful comparison of results when dimensions N, dx, dy and dz vary.
+Random matrices X, M, N, H and F are populated with centred gaussian random coefficients with unit variance, divided by $ \surd n$ where n is the smallest dimension of the matrix.
+Since a random square matrix with unit gaussian coefficients of dimension n has eigenvalues between $ \surd n $ and $\surd n $, such  scaling allows for meaningful comparisons when dimensions dx, dy and dz vary.
 
-We evaluate the performance of double regression in two different roles : as a feature extractor, recovering the non-zero elements of E (the active features of X), and as a regression method, predicting Y from X over a test sample.
-For feature extraction, the quality of the fit is measured as the number of disagreements between our estimator and the diagonal of E (ie, their Hamming distance, which should be minimised).
+FC: insert comment about scaling sigmoids, and maybe something about uniform coeffs
+
+We evaluate the performance of double regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
+For feature extraction, the quality of fit is the number of disagreements between our estimator and the diagonal of E (ie, their Hamming distance, which should be minimised).
 For regression, we use a test set of N additional measurements, and calculate the R-squared ratio of variances (which should be maximised).
+Performance indicators are calculated by averaging the results of several experiments.
 
-\subsection{Experimenting with double regression}
-In these experiments, we try to understand how various parameters of double regression (dimension, active features, signal to noise ratio) influence the quality of feature extraction and regression.
-For regression, we are also investigating different ways of taking extracted features into account, and the impact of varying environments (having training and test samples from different distributions).
+\subsection{Feature extraction - linear case}
+All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by double regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
+The decrease is usually very steep until a Hamming distance under 2 is achieved.
+Figure 1 presents a number of such curves for varying values of dx=dy=dz, nc=30 and N=1000.
+
+Figure 1
+
+The distance between $\hat E$ and E incorporates two different errors: false positives, where features not in E are extracted, and false negatives where features in E are not extracted.
+For low values of the snr, and typical values of nc (much smaller than dx), false positives are high, but they fall faster as snr increases.
+False negatives begin lower, but decrease more slowly (figure 2).
+In other words, for each values of the parameters, one may determine a value of the signal to noise ratio, such that snr higher than this threshold allow for good feature extraction.
+At this level, all the errors are false negatives.
+
+Figure 2
 
@@
 
+\subsection{Feature extraction - nonlinear case}
+
+\subsection{Regression }
 For regression, we want to understand how feature extraction helps predict Y from X. To this effect, we compare four approaches to regression:
@@
 
-\subsection{Double regression in multiple environments}
+\subsection{Multiple environments}
 In this section, we perform the same tests as in the previous one, but the test set is taken from a different distribution of X than the training set.

### 2019-05-15 code/synthetic_v_1_0.py
Adding fixed synthetic generator.

@@ -1,0 +1,58 @@
+import numpy as np
+from scipy.linalg import sqrtm
+
+
+def rolling_covariance(rho, dim):
+    cov = np.zeros((dim, dim))
+    for i in range(cov.shape[0]):
+        for j in range(cov.shape[1]):
+            cov[i, j] = np.power(rho, np.abs(i - j) / dim)
+    return cov
+
+
+def synthetic_data_v_1_0(n_samples=1000,   # number of samples
+                         dim_x=50,         # number of features
+                         rho_x=0.5,        # correlation of features
+                         rho_n=0.5,        # correlation of noise
+                         dim_y=30,         # number of sensors
+                         snr=1.0,          # signal-to-noise ratio
+                         nc=5,             # number of selected features
+                         nonlinear=False):  # number of selected features
+    # linear transformation
+    F = np.random.randn(dim_x, dim_y) / np.sqrt(dim_x)
+
+    # masking transformation
+    E = np.array([0] * (dim_x - nc) + [1] * (nc))
+    np.random.shuffle(E)
+    E = np.diag(E)
+
+    # features
+    cov_X = rolling_covariance(rho_x, dim_x)
+    X = np.random.randn(n_samples, dim_x) @ sqrtm(cov_X)
+    np.random.shuffle(X.T)
+
+    # noise
+    cov_N = rolling_covariance(rho_n, dim_x)
+    N = np.random.randn(n_samples, dim_x) @ sqrtm(cov_N)
+    np.random.shuffle(N.T)
+
+    # observed sensor data
+    Y = (snr * X @ E + N) @ F
+
+    if nonlinear:
+        Y = 1. / (1. + math.exp(-Y))
+
+    # return inputs, outputs, and solution
+    return X, Y, E
+
+
+sweep_v_1_0 = {
+    "n_samples": [100, 1000, 10000],
+    "rho_x": [0, 0.01, 0.2, 0.95],
+    "rho_n": [0, 0.01, 0.2, 0.95],
+    "dim_x": [10, 50, 100, 500],
+    "dim_y": [10, 50, 100, 500],
+    "snr": [0.1, .25, 1],
+    "nc": [1, 10, 20],
+    "nonlinear": [0, 1],
+}

### 2019-05-15 code/pedogagical_example_bias_lambdax.py
add example to show jrr bias

@@ -1,0 +1,67 @@
+from sklearn.linear_model import LinearRegression, RidgeCV
+from sklearn.model_selection import ShuffleSplit
+from scipy.stats import ttest_1samp
+import numpy as np
+import matplotlib.pyplot as plt
+
+
+def make_data(seed=0):
+    np.random.seed(seed)
+
+    # make data
+    n = 1000  # number of samples
+    nX = 60  # dimensionality of X
+    nY = 60
+
+    Cx = np.random.randn(nX, nX)
+    Cx = Cx.dot(Cx.T) / nX  # sym pos-semidefin
+
+    X = np.random.multivariate_normal(np.zeros(nX), Cx, n)
+    N = np.random.randn(n, nX)
+    F = np.random.randn(nY, nX)
+    E = np.eye(nX)
+    E[:10] = 0
+    E[30:] = 0
+
+    Y = (X @ E + N) @ F.T
+    Y += np.random.randn(*Y.shape)
+    return X, Y, E
+
+
+if __name__ == '__main__':
+    X, Y, E = make_data()
+    n, nX = X.shape
+
+    alphas = np.logspace(-6, 6, 20)
+    ridge = RidgeCV(alphas=alphas, fit_intercept=False)
+    ols = LinearRegression(fit_intercept=False)
+
+    # cv
+    set1, set2 = range(n//2), range(n//2, n)
+    cv = ShuffleSplit(100, test_size=.5, random_state=0)
+
+    X, Y, E = make_data(seed=2)
+    
+    # vanilla JRR Bias
+    E_hats = list()
+    for set1, set2 in cv.split(X, Y):
+        # decode
+        G = ridge.fit(Y[set1], X[set1])
+        X_hat = G.predict(Y)
+        # Encode
+        E_hat = ridge.fit(X[set2], X_hat[set2]).coef_
+        E_hats.append(E_hat)
+    E_hats = np.mean(E_hats, 0)
+
+    # plot
+    vmax = .3
+    fig, axes = plt.subplots(1, 2)
+
+    axes[0].fill_between(range(len(E)), np.diag(E_hat))
+    not_causal = np.where(np.diag(E) == 0)[0]
+    _, p_value = ttest_1samp(np.diag(E_hat)[not_causal], 0)
+    axes[0].set_title('non causal > 0: p=%.4f' % p_value)
+    if vmax is None:
[…]

### 2019-05-15 paper/paper.tex
Figure

@@
 
-\usepackage[final]{neurips_2019}
+\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx}
+\usepackage[hidelinks]{hyperref}
+\usepackage[]{neurips_2019}
 
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{amsmath}
+\usetikzlibrary{calc}
+
+\tikzset{
+    ncbar angle/.initial=90,
+    ncbar/.style={
+        to path=(\tikztostart)
+        -- ($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)
+        -- ($(\tikztotarget)!($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztostart)$)
+        -- (\tikztotarget)
+    },
+    ncbar/.default=0.5cm,
+}
+
+\tikzset{round left paren/.style={ncbar=0.5cm,out=110,in=-110}}
+\tikzset{round right paren/.style={ncbar=0.5cm,out=70,in=-70}}
+
+\newcommand{\dlp}[1]{{\color{red} (DLP: #1)}}
+\newcommand{\mo}[1]{{\color{green} (MO: #1)}}
+\newcommand{\jrk}[1]{{\color{blue} (JRK: #1)}}
+\newcommand{\fc}[1]{{\color{orange} (FC: #1)}}
 
@@
 \section{Introduction}
+
 Causal discovery, telling causes from their effects, is an inverse problem.
+%
 Given experimental data (effects) and possible explanations (causes), we are tasked to find the causes that account for the observations, knowing that other non causal factors (contingent, random) are at work.
@@
 More precisely, we measure, via an apparatus F, a multidimensional signal Y, that is a function of possible causes X, and other, unknown, factors that can be understood as noise N: $Y=f(X,N)$.
+%
 Of all possible cause (features of X), only a few account for Y, let them be represented as EX, with E a binary diagonal matrix: all zeroes except a few ones on the diagonal, which correspond to active causal features of X.
+%
 Let us assume that noise N is centred (adding, if need be, an intercept to the active causal features, to capture the bias in N), and additive (not a very strong assumption in the absence of further constraints on N).
+%
 Then, we can write our model of causal discovery as $Y=f(m(EX)+N)$, with f and m unknown functions.
+%
 Up to some rearrangement of the features of X, we can also suppose m linear.
@@
 We end up with the following model $Y=f(MEX+N)$, where N is centred noise E a binary diagonal matrix, M a full rank matrix and f some unknown transformation.
+%
 Given experimental measures of X and Y (as matrices), and our goal is to recover E, the active causal features of the process under study.
@@
 Such problems have been attacked along two different lines.
+%
 In the direct approach, we use the above equation as a statistical model, try to express each feature of Y as a function of X, test whether the coefficients associated with specific features of X are different from zero, and recover E as the features of X active over some dimension of Y.
+%
 In the inverse approach, we try to express each feature of X as a function of Y, estimate its changes as Y varies, and recover E as the set of "non random" features of X.
+%
 In the presence of noise and a large number of features of X, Y and N, both approaches are limited because they fail to take advantage of the correlations between features of the variable they are modelling (or guard against their multicollinearity).
+%
 In the direct approach, features of Y are modelled separately.
+%
 In the inverse approaches, correlations between features of X are not accounted for.
@@
 In this paper, we introduce double regression, which combines direct and inverse approaches.
+%
 We first predict X from Y, as in the inverse approach, and retrieve a regularised least square estimate $\hat X$.
+%
 Then, we calculate a regularised least square estimator of $\hat X$ from X.
+%
 If the active features of X are represented as EX, with E a binary diagonal, this provides a good approximation of E, robust to noise and environment change.
+%
 The main purpose of double regression is feature extraction, but it can be turned into a regression method by filtering X on the selected features.
@@
 In the next sections, we describe double regression, prove that it asymptotically extracts E, both in the linear and non linear case, and that it can be understood as a special case of canonical component analysis (cca).
+%
 By experimenting with simulated data, we demonstrate its robustness and use cases, and show that it compares favourably with state of the art techniques for regression and feature extraction.
+%
 Finally, we illustrate its use on an example from neuroscience.
@@
 \section{Double regression}
+
+\begin{figure}[t!]
+    \centering
+    \begin{tikzpicture}
+    \newcommand\posY{0}
+    \newcommand\posX{3}
+    \newcommand\posE{5}
+    \newcommand\posN{7}
+    \newcommand\posF{9.5}
+
+    \node[thick, draw=black, minimum height=3cm, minimum width=2cm, fill=yellow!10] (Y) at (\posY, 0){};
+    \node[] (eq) at (1.5, 0){$=$};
+    \node[] (times) at (4, 0){$\times$};
+    \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=yellow!10] (X) at (\posX, 0){};
+    \node[thick, draw=black, minimum height=1cm, minimum width=1cm, fill=red!10] (E) at (\posE, 0){};
+
+    \draw[fill=white] (\posE - 0.5 + 0.0, 0.5 - 0.0) rectangle (\posE - 0.5 + 0.0 + 0.1, 0.5 - 0.0 - 0.1);
+    \draw[fill=black] (\posE - 0.5 + 0.1, 0.5 - 0.1) rectangle (\posE - 0.5 + 0.1 + 0.1, 0.5 - 0.1 - 0.1);
+    \draw[fill=white] (\posE - 0.5 + 0.2, 0.5 - 0.2) rectangle (\posE - 0.5 + 0.2 + 0.1, 0.5 - 0.2 - 0.1);
+    \draw[fill=white] (\posE - 0.5 + 0.3, 0.5 - 0.3) rectangle (\posE - 0.5 + 0.3 + 0.1, 0.5 - 0.3 - 0.1);
+    \draw[fill=white] (\posE - 0.5 + 0.4, 0.5 - 0.4) rectangle (\posE - 0.5 + 0.4 + 0.1, 0.5 - 0.4 - 0.1);
+    \draw[fill=white] (\posE - 0.5 + 0.5, 0.5 - 0.5) rectangle (\posE - 0.5 + 0.5 + 0.1, 0.5 - 0.5 - 0.1);
+    \draw[fill=white] (\posE - 0.5 + 0.6, 0.5 - 0.6) rectangle (\posE - 0.5 + 0.6 + 0.1, 0.5 - 0.6 - 0.1);
+    \draw[fill=black] (\posE - 0.5 + 0.7, 0.5 - 0.7) rectangle (\posE - 0.5 + 0.7 + 0.1, 0.5 - 0.7 - 0.1);
+    \draw[fill=white] (\posE - 0.5 + 0.8, 0.5 - 0.8) rectangle (\posE - 0.5 + 0.8 + 0.1, 0.5 - 0.8 - 0.1);
+    \draw[fill=black] (\posE - 0.5 + 0.9, 0.5 - 0.9) rectangle (\posE - 0.5 + 0.9 + 0.1, 0.5 - 0.9 - 0.1);
+
+    \node[] (plus) at (6, 0){$+$};
+    \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=blue!10] (N) at (\posN, 0){};
+    \node[thick, draw=black, minimum height=2cm, minimum width=1cm, fill=red!10] (F) at (\posF, 0){};
+    \draw [thick] (2.25, -1.5) to [round left paren ] (2.25, 1.5);
+    \draw [thick] (7.75, -1.5) to [round right paren ] (7.75, 1.5);
+
+    \node[] (times2) at (8.5, 0){$\times$};
+
+    \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{n \times q}$}};
+    \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{n \times d}$}};
+    \node[] (annE) at (\posE, -1.8){\scalebox{0.85}{$E \in \mathbb{D}^{d \times d}$}};
+    \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{n \times d}$}};
+    \node[] (annF) at (\posF, -1.8){\scalebox{0.85}{$F \in \mathbb{R}^{d \times q}$}};
+
+    \node[] (labY) at (\posY, 2){observed effects};
+    \node[] (labX) at (\posX, 2.3){world};
+    \node[] (labX) at (\posX, 2){features};
+    \node[] (labE) at (\posE, 2.3){cause};
+    \node[] (labE) at (\posE, 2){selection};
+    \node[] (labN) at (\posN, 2){noise};
+    \node[] (labF) at (\posF, 2.3){cause-effect};
+    \node[] (labF) at (\posF, 2){mapping};
+
+    \node[] (reg1) at (5,-3.25) {$\hat{E} = \overbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\underbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}_{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}^{\text{2) regression from } X \text{ to } \hat{X}} $};
+    \end{tikzpicture}
+    \caption{Summary of our method blah blah blah}
+    \label{fig:}
+\end{figure}
+
 \subsection{The method}
+
 We are given two set of measurements over a sample of N individuals.
+%
 Y are ny multidimensional effects, and X are nx multidimensional possible causes.
+%
 We assume that only some features of X influence Y, so that $Y=f(EX,N)$, E being a binary diagonal, N some unknown noise, and f an unknown function.
+%
 We try to recover E, the active causal features.
@@
 Double regression consists in two steps.
+%
 First, we perform a least squares estimation of X from Y, that is we find $\hat G$ such that $\hat X=\hat G Y$ is the least square predictor of X.
+%
 This recovers an approximation of the causes of Y in X.
+%
 Then, we perform a least squares estimation of $\hat X$ from X, and determine $\hat H$ such that $\hat {\hat X}=\hat H X$ is the least square predictor of $\hat X$.
+%
 In matrix form, we have $\hat G=(Y'Y)^{-1} Y'X$, and
@@
 In practice, since the covariance matrices of Y and X might exhibit multicollinearity, we use regularisation in both matrix inversions.
+%
 This is done by using ridge regression, and determining at each step the regularisation parameter by cross validation.
+%
 In the above formula, this amounts to replacing $(X'X)^{-1}$ and $(Y'Y)^{-1}$ by $(X'X+\lambda)^{-1}$ and $(Y'Y+\mu)^{-1}$.
+%
 Also, to prevent spurious correlations between the two regressions (should they be performed on the same sample), our original sample is randomly split into two halves, and each step is performed over one of them, effectively decorrelating the two steps.
+%
 This is repeated over several (10) splits, and the resulting $\hat H$ are averaged.
+%
 This bagging compensates for the reduction in sample size caused by splitting.
@@
 Under our hypotheses (centred noise, E a binary diagonal) and even under heavy noise, $\hat H$ is close to diagonal.
+%
 Yet, because of regularisation which adds positive bias to small eigenvalues, and noise which attenuates large eigenvalues, $\hat H$ is in fact a scaled and noisy approximation to E.
+%
 To recover a binary diagonal, as our estimate of E, we extract the diagonal of $\hat H$ and select its large elements.
+%
 The diagonal matrix $\hat E$, having large elements of H as its unit diagonal coefficients (the rest being zero) is our estimate of E, the active features of X.
@@
 To select large elements of $\hat H$, the traditional approach in regression analysis consists in testing whether those coefficient differ from zero (or are larger from zero since we are only interested in large positive coefficients).
+%
 This is usually done via Student t-test of variants of it.
+%
 This will not work here, as double regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
+%
 Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).
@@
 This can be done by maximising the ratio of inter-group variance over total variance (ie minimising intra-group variance).
+%
 In a one dimensional two clusters setting, this amounts to sorting the data, and separating the p smallest from the n-p largest values so that inter group inertia is maximal.
+%
 If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we are maximising the Sonquist and Morgan criterion $$K = {p (n-p) \over n} {(s-r)^2 \over V}$$ and selecting the p features corresponding to the largest values.
+%
 If diagonal elements of $\hat H$ are normally distributed, K follows a chi-square distribution with one degree of freedom.
+%
 At 95\% confidence level, a value of K superior to 3.84 makes our selection significant (for 99\% confidence, the corresponding value of K is 6.63) \citep{Kass_75}
@@
 This extract from $\hat H$ a binary diagonal $\hat E$.
+%
 To use double regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
@@
 \item Alternatively, use $\hat E$ to predict Y from X, by performing a regularised least square regression of Y from $\hat E X$.
+%
+
 \end{enumerate}
@@
 As discussed in the introduction, our problem can be formulated (up to a redifinition of the features of X) as $Y=f(MEX+N)$, with F an unknown function.
+%
 If F is linearm we can rewrite it as $Y = F(MEX + N)$, with X an (N, dx) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
+%
 M an unknown full rank (dz,dx) transformation of EX, N a (dz,N) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dz) transformation corresponding to the measuring apparatus, and Y a (dx,dy) matrix of measured effects.
@@
 \end{equation}
-with k a positive constant which depends on the variance of FN. This amounts to finding the minimal norm generalized inverse of FME, or its Moore-Penrose inverse $\hat  G= (FME)^{\dagger}$.
+with k a positive constant which depends on the variance of FN.
+%
+This amounts to finding the minimal norm generalized inverse of FME, or its Moore-Penrose inverse $\hat  G= (FME)^{\dagger}$.
 
@@
 For any matrix A, $A^\dagger A$ is the orthogonal projector over $Img(A')$, and $I-A^\dagger A$ the orthogonal projector over $Ker(A)$.
+%
 In our case, $Ker(FME)\supseteq Ker(E)$.
+%
 If FM has full rank over the subspace spanned by the features selected by E (this is assumed for M and will happen for almost any random F), then $Ker(FME) = Ker(E)$,.
+%
 This means $I - \hat H$ is the projector over $Ker(E)$, the diagonal matrix that spans the unselected features, and $\hat H$ is the diagonal matrix corresponding to the selected features, that is E.
+%
 Under such hypotheses, $\hat H$ retrieves E, the features of X that have a causal influence on Y.
+%
 Since it is diagonal with eigenvalues well separated (0 and 1), it should do so in a fairly resilient way.
@@
 Formula (1) shows that double regression can be seen as a special case of canonical component analysis (CCA).
+%
 In CCA, we are given two matrixes X and Y (or size N,p and N,q) that describe different features measured on the same data sample, and characterise their proximity (that is, between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.
@@
 This is done by finding vectors $a$ and $b$ such that $Xa$ and $Yb$ display maximal correlation, or by maximising $a'X'Yb$ over unit-normed $a$ and $b$.
+%
 To find a, one calculates the eigenvectors of (1).
+%
 The largest one is the direction with maximal correlation, the second one the maximal correlation direction orthogonal to the first, and so on.
+%
 For each canonical dimension, the corresponding eigenvalue is the square of the cosine between the corresponding direction along X and the subspace spanned by Y.
@@
 If (1) is diagonal, for instance in the model presented above, the features of X are eigenvectors.
+%
 If all eigenvalues are either 0 or 1, since they correspond to the squared cosine of the angle between the corresponding feature and the subspace spanned by Y, all the features are either part of the Y subspace (eigenvalues of one) or orthogonal to it (for zero eigenvalues).
+%
 Thus, double regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
+%
 The closer we are to this hypothesis, the closer the results of CCA and double regression will be.
@@
 There are differences between CCA and double regression.
+%
 First, since our objective is not to characterise the proximity between X and Y, but just to extract causal features, we can dispense with the last step of CCA (diagonalisation of (1)).
+%
 Second, since we operate in a noisy environment, we use regularisation, introducing bias in the correlation calculations.
+%
 Finally, whereas CCA wants to leverage all correlations between X and Y, we filter some of them, through splitting and bagging.
+%
 In other words, whereas we are working from the same formulae as CCA, we use them in different ways, for different purposes
@@
 The model used for simulation is $Y=f(snr MEX+N)$, with X a random (dx,N) matrix, E a square binary diagonal of dimension dx with the nc last elements equal to 1 (this means that of the dx features of X, only the nc last have some influence on Y), M and N (dz,dx) and (dz,N) random matrices, and f a function from $\mathbb{R}^{dz}$ to $\mathbb{R}^{dy}$, that can be linear or non linear.
+%
 In the linear case, the model becomes $Y=F(MEX+N)$, with F a (dy,dz) matrix.
+%
 In the non linear case, F becomes $FSH$ where F and H are matrices of dimensions (dz,dz) and (dy,dz), and S a vector of dz sigmoid transformations.
+%
 Finally, $snr$ is a real number measuring the amount of noise.
@@
 Random matrices X, M, N, H and F are populated with centred gaussian random coefficients with unit variance, divided by $ \surd n$ where n is the smallest dimension of the matrix.
+%
 Since a random square matrix with unit gaussian coefficients of dimension n has eigenvalues between $ \surd n $ and $\surd n $, such  scaling allows for meaningful comparisons when dimensions dx, dy and dz vary.
@@
 We evaluate the performance of double regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
+%
 For feature extraction, the quality of fit is the number of disagreements between our estimator and the diagonal of E (ie, their Hamming distance, which should be minimised).
+%
 For regression, we use a test set of N additional measurements, and calculate the R-squared ratio of variances (which should be maximised).
+%
 Performance indicators are calculated by averaging the results of several experiments.
@@
 All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by double regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
+%
 The decrease is usually very steep until a Hamming distance under 2 is achieved.
+%
 Figure 1 presents a number of such curves for varying values of dx=dy=dz, nc=30 and N=1000.
@@
 The distance between $\hat E$ and E incorporates two different errors: false positives, where features not in E are extracted, and false negatives where features in E are not extracted.
+%
 For low values of the snr, and typical values of nc (much smaller than dx), false positives are high, but they fall faster as snr increases.
+%
 False negatives begin lower, but decrease more slowly (figure 2).
+%
 In other words, for each values of the parameters, one may determine a value of the signal to noise ratio, such that snr higher than this threshold allow for good feature extraction.
+%
 At this level, all the errors are false negatives.
@@
 Working from a fixed sample of 1000, and 15 active features, and varying values of dx, dy, dz (from 50 to 200) and snr (from 0.2 to 1), we observe that, all other factors being equal, the quality of feature extraction and regression decrease when dx increases (X has more inactive features), dz increase (more noise dimensions) and snr decreases (less signal, more noise).
+%
 The dimensionality of Y plays a much smaller role, except for very small values of dy (less than 10).
@@
 For feature extraction, double regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
+%
 The following graphs show the boundary between values of dx and dz that allow for features extractions (dx lower than boundary, or dz larger), for different values of dy (level curves on the graph), and different values of signal to noise ratio and number of active variables.
@@
 
-For values of ... in ..., an empirical formula can be provided. Double regression allows for good feature extraction if :
+For values of ...
+%
+in ..., an empirical formula can be provided.
+%
+Double regression allows for good feature extraction if :
 
@@
 \subsection{Regression }
-For regression, we want to understand how feature extraction helps predict Y from X. To this effect, we compare four approaches to regression:
+For regression, we want to understand how feature extraction helps predict Y from X.
+%
+To this effect, we compare four approaches to regression:
 \begin{itemize}
@@
 Over a large set of parameters, we observe that filtered regression is on average better than weighted, which is on average better than standard regression.
+%
 This demonstrates that double regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.
@@
 In this section, we perform the same tests as in the previous one, but the test set is taken from a different distribution of X than the training set.
+%
 In practice, we test the same trained predictors of Y over test samples generated from a distribution of X with a different covariance.
@@
 FC: ideally, we'd like to assess the multiple env performance of jrr, and compare it with others.
-Are multiple environments limited to regression analyses?
-How should environments be generated (not just variance of X, I expect)?
+%
+Are multiple environments limited to regression analyses? How should environments be generated (not just variance of X, I expect)?
 

### 2019-05-16 paper/paper.tex
Sampling notation
abstract

@@
 \begin{abstract}
-    To be written.
+Identify causes from observations is a core challenge to most empirical sciences.
+This endeavor is particularly challenging when (i) potential factors are difficult to manipulate separately and (ii) when observations are complex and/or multidimensional.
+To address this issue, we introduce in this paper the “back-and-forth regression” (BFR), a method that optimally identifies the most plausible causal factors from a set of covarying candidates.
+BFR consists in optimally (i) predicting the potential causes X from the observable Y, and subsequently (ii) identifying the only factors that contributed to those predictions.
+After detailing the proof of convergence, we show that BFR can systematically outperform least-square and related regression techniques, as well as cross-decomposition methods (e.g.
+canonical correlation analysis, partial least squares) on two distinct aspects: (i) causal identification and (ii) predictions outside independent and identically distributed datasets.
+We apply this method to empirical datasets and show that it retrieves plausible results.
+In particular, we apply BFR to neuroimaging recordings of 102 subjects presented with word sequences, and show, as expected, that the early brain responses appear to be specifically caused by low-level visual factors whereas late brain responses appear to be caused lexico-semantic factors, despite the fact that these two sets of factors are co-varying.
+We finish the paper by extending our approach to nonlinear cause-effects relationships.
+conclusion
 \end{abstract}
@@
 %
-Of all possible cause (features of X), only a few account for Y, let them be represented as EX, with E a binary diagonal matrix: all zeroes except a few ones on the diagonal, which correspond to active causal features of X.
+Of all possible causes (features of X), only a few account for Y, let them be represented as EX, with E a binary diagonal matrix: all zeroes except a few ones on the diagonal, which correspond to active causal features of X.
 %
@@
 
+    \node[] (sim1) at (0,-3.25) {\scalebox{0.85}{$X \sim P(X)$}};
+    \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
     \node[] (reg1) at (5,-3.25) {$\hat{E} = \overbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\underbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}_{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}^{\text{2) regression from } X \text{ to } \hat{X}} $};

### 2019-05-16 paper/paper.tex
odds and ends in paper

--- 2019-05-12 paper/paper.tex
@@
 \section{Experiments - simulated data}
-Experiments on simulated data serve three goals: compare double regression with other known techniques (CCA, OLS, PLS...) both as regression and feature extraction methods, provide evidence for our claims on stability, invariance to environment, and behavior as a causal detector, and understand the strengths and limitations of double regression, and directions for future improvement.
+Experiments on simulated data serve three purposes: compare double regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.
 
@@
 
+In the non linear case, sigmoid functions are...
 FC: insert comment about scaling sigmoids, and maybe something about uniform coeffs

### 2019-05-16 paper/paper.tex
One pass to abstract

--- 2019-05-15 paper/paper.tex
@@
 
-\title{Learning causal features with double regression}
+\title{Discovering causal factors with back-and-forth regression}
 
@@
 \begin{abstract}
-    To be written.
+Identifying causes from observations is at the core of science.
+This endeavor is particularly challenging when i) potential factors are difficult to manipulate individually and ii) observations are complex and multi-dimensional.
+To address this issue, we introduce ``back-and-forth regression'' (BFR), a method that optimally identifies the most plausible causal factors from a set of covarying candidates.
+BFR consists in i) predicting the potential causes $X$ from the observable $Y$, and subsequently ii) identifying the only factors that contributed to those predictions.
+After detailing the proof of convergence, we show that BFR outperforms least-squares and related regression techniques, as well as cross-decomposition methods (e.g.
+canonical correlation analysis, and partial least squares) on two causal identification and prediction out of distribution.
+We apply BFR to several empirical datasets to show that it retrieves plausible results.
+In particular, we apply BFR to neuroimaging recordings of 102 subjects presented with word sequences, and show, as expected, that the early brain responses appear to be specifically caused by low-level visual factors whereas late brain responses appear to be caused lexico-semantic factors, despite the fact that these factors covary.
+We finish extending BFR to nonlinear cause-effect relationships.
 \end{abstract}

### 2019-05-16 paper/paper.tex
break lines
fix stuff

--- 2019-05-16 paper/paper.tex
@@
 \begin{abstract}
-Identifying causes from observations is at the core of science.
-This endeavor is particularly challenging when i) potential factors are difficult to manipulate individually and ii) observations are complex and multi-dimensional.
-To address this issue, we introduce ``back-and-forth regression'' (BFR), a method that optimally identifies the most plausible causal factors from a set of covarying candidates.
-BFR consists in i) predicting the potential causes $X$ from the observable $Y$, and subsequently ii) identifying the only factors that contributed to those predictions.
-After detailing the proof of convergence, we show that BFR outperforms least-squares and related regression techniques, as well as cross-decomposition methods (e.g.
-canonical correlation analysis, and partial least squares) on two causal identification and prediction out of distribution.
-We apply BFR to several empirical datasets to show that it retrieves plausible results.
-In particular, we apply BFR to neuroimaging recordings of 102 subjects presented with word sequences, and show, as expected, that the early brain responses appear to be specifically caused by low-level visual factors whereas late brain responses appear to be caused lexico-semantic factors, despite the fact that these factors covary.
-We finish extending BFR to nonlinear cause-effect relationships.
+Identifying causes from observations is at the core of science. This endeavor
+is particularly challenging when i) potential factors are difficult to
+manipulate individually and ii) observations are complex and multi-dimensional.
+To address this issue, we introduce ``back-and-forth regression'' (BFR), a
+method that optimally identifies the most plausible causal factors from a set
+of covarying candidates. BFR consists of i) predicting the potential causes $X$
+from the observable $Y$, and subsequently ii) identifying the only factors that
+contributed to those predictions. After detailing the proof of convergence, we
+show that BFR outperforms least-squares and related regression techniques, as
+well as cross-decomposition methods (e.g. canonical correlation analysis, and
+partial least squares) on two tasks: causal identification and out-of-distribution
+prediction. We apply BFR to several empirical datasets to show that it
+retrieves plausible results. In particular, we apply BFR to neuroimaging
+recordings of 102 subjects presented with word sequences, and show, as
+expected, that the early brain responses appear to be specifically caused by
+low-level visual factors whereas late brain responses appear to be caused by
+lexico-semantic factors, despite the fact that these factors covary. We finish
+by extending BFR to nonlinear cause-effect relationships.
 \end{abstract}

### 2019-05-16 paper/paper.tex
modified asymptotic proof

@@
 \subsection{Asymptotic behaviour - linear case}
-As discussed in the introduction, our problem can be formulated (up to a redifinition of the features of X) as $Y=f(MEX+N)$, with F an unknown function.
+As discussed in the introduction, our problem can be formulated (up to a redifinition of the features of X) as $Y=f(EX+N)$, with F an unknown function.
 %
-If F is linearm we can rewrite it as $Y = F(MEX + N)$, with X an (N, dx) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
+If F is linear we can rewrite it as $Y = F(EX + N)$, with X an (n, dx) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
 %
-M an unknown full rank (dz,dx) transformation of EX, N a (dz,N) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dz) transformation corresponding to the measuring apparatus, and Y a (dx,dy) matrix of measured effects.
+N a (dx,n) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dx) transformation corresponding to the measuring apparatus, and Y a (n,dy) matrix of measured effects.
 
@@
 \begin{aligned}
-\min_G \left \| X-GY \right \|^2 &= \min_G \left \| X - GF(MEX+N)\right\|^2 \\
-&{}= \min_G \left \| (I-GFME)X\right\| ^2 + \left \| GFN\right \| ^2\\
-&{}\leq \min_G \left \| I-GFME\right\| ^2 \left \| X\right\| ^2 + k\left \| G\right \| ^2\\
+\min_G \left \| X-GY \right \|^2 &= \min_G \left \| X - GF(EX+N)\right\|^2 \\
+&{}= \min_G \left \| (I-GFE)X\right\| ^2 + \left \| GFN\right \| ^2\\
+&{}= \min_G \alpha \left \| I-GFE\right\| ^2  + \beta \left \| G\right \| ^2\\
 \end{aligned}
 \end{equation}
-with k a positive constant which depends on the variance of FN.
+with $\alpha$ and $\beta$ positive constants, the latter equal to the norm of FN.
+We are balancing two minimisation, with $/alpha$ close to zero, the solution is the null matrix, with $\beta$ close to zero, it is the pseudo inverse of $FE$.
 %
-This amounts to finding the minimal norm generalized inverse of FME, or its Moore-Penrose inverse $\hat  G= (FME)^{\dagger}$.
+
+In the general case, $\beta=\left \| FN\right \| ^2$, and the solution is $k (FE)^\dagger$, with $k= {\alpha\over{\alpha+\beta}}$.
 
@@
 \begin{aligned}
-arg \min_H \left \| (FME)^{\dagger}Y - HX \right \|^2 &= arg \min_H \left \| (FME)^{\dagger}F(MEX+N) - HX \right \|^2 \\
-&= arg \min_H \left \| ((FME)^{\dagger}(FME)-H)X \right \| ^2 + \left \| (FME)^{\dagger}FN \right \| ^2\\
-&= arg \min_H \left \| ((FME)^{\dagger}(FME)-H)X \right \| ^2\\
-&=(FME)^{\dagger}(FME)
+arg \min_H \left \| k(FE)^{\dagger}Y - HX \right \|^2 &= arg \min_H \left \| k(FE)^{\dagger}F(EX+N) - HX \right \|^2 \\
+&= arg \min_H \left \| (k(FE)^{\dagger}(FE)-H)X \right \| ^2 + \left \| k(FE)^{\dagger}FN \right \| ^2\\
+&= arg \min_H \left \| (k(FE)^{\dagger}(FE)-H)X \right \| ^2\\
+&=k(FE)^{\dagger}(FE)
 \end{aligned}
 \end{equation}
-Thus, in the linear case, double regression retrieves $(FME)^{\dagger}(FME)$.
+Thus, in the linear case, double regression retrieves $k(FE)^{\dagger}(FE)$.
+In many cases, if v s the variance of $FN$, k can be approximated as $k= {1\over{1+v}}$, it quantifies the amount of noise added to the causal process, and the difficulty of recovering the causal features.
 
@@
 %
-In our case, $Ker(FME)\supseteq Ker(E)$.
+In our case, $Ker(FE)\supseteq Ker(E)$.
 %
-If FM has full rank over the subspace spanned by the features selected by E (this is assumed for M and will happen for almost any random F), then $Ker(FME) = Ker(E)$,.
+If F has full rank over $Img(E)$, the subspace spanned by E, then $Ker(FE) = Ker(E)$,.
 %
-This means $I - \hat H$ is the projector over $Ker(E)$, the diagonal matrix that spans the unselected features, and $\hat H$ is the diagonal matrix corresponding to the selected features, that is E.
-%
-Under such hypotheses, $\hat H$ retrieves E, the features of X that have a causal influence on Y.
-%
-Since it is diagonal with eigenvalues well separated (0 and 1), it should do so in a fairly resilient way.
-
+This means $I - \hat H$ is the projector over $Ker(E)$, the diagonal matrix that spans the unselected features, and $\hat H$ is k times the diagonal matrix corresponding to the selected features, that is kE.
+Since it is diagonal with eigenvalues 0 and k, if k is large enough, it should allow us to retrieve causal features in a fairly robust way.
 \subsection{Asymptotic behaviour - non linear case}

### 2019-05-18 paper/paper.tex

@@
 
-In this paper, we introduce double regression, which combines direct and inverse approaches.
+In this paper, we introduce back and forth regression, which combines direct and inverse approaches.
 %
@@
 %
-The main purpose of double regression is feature extraction, but it can be turned into a regression method by filtering X on the selected features.
+The main purpose of back and forth regression is feature extraction, but it can be turned into a regression method by filtering X on the selected features.
 
-In the next sections, we describe double regression, prove that it asymptotically extracts E, both in the linear and non linear case, and that it can be understood as a special case of canonical component analysis (cca).
+In the next sections, we describe back and forth regression, prove that it asymptotically extracts E, both in the linear and non linear case, and that it can be understood as a special case of canonical component analysis (cca).
 %
@@
 
-\section{Double regression}
+\section{Back an forth regression}
 
@@
 
-Double regression consists in two steps.
+Back and forth regression consists in two steps.
 %
@@
 \begin{equation} \hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X\end{equation}
-Under good conditions (close to linear dependence between X and Y, and covariance matrices of X and Y well conditioned), this recovers a close approximation to E.
+Under good conditions (close to linear dependence between X and Y, and covariance matrices of X and Y well conditioned), this recovers a scaled approximation to E.
 
 In practice, since the covariance matrices of Y and X might exhibit multicollinearity, we use regularisation in both matrix inversions.
+This amounts to replacing $(X'X)^{-1}$ and $(Y'Y)^{-1}$ by $(X'X+\lambda)^{-1}$ and $(Y'Y+\mu)^{-1}$, and is done by using ridge regression, regularisation parameters being estimated by leave one out cross validation.
 %
-This is done by using ridge regression, and determining at each step the regularisation parameter by cross validation.
-%
-In the above formula, this amounts to replacing $(X'X)^{-1}$ and $(Y'Y)^{-1}$ by $(X'X+\lambda)^{-1}$ and $(Y'Y+\mu)^{-1}$.
-%
-Also, to prevent spurious correlations between the two regressions (should they be performed on the same sample), our original sample is randomly split into two halves, and each step is performed over one of them, effectively decorrelating the two steps.
+Also, to prevent spurious correlations between the two steps (should they be performed on the same sample), our original sample is randomly split into two halves, and each step is performed over one of them, effectively decorrelating the two steps.
 %
@@
 
-To select large elements of $\hat H$, the traditional approach in regression analysis consists in testing whether those coefficient differ from zero (or are larger from zero since we are only interested in large positive coefficients).
+To select large elements of $\hat H$, the traditional approach in regression analysis consists in testing whether its coefficients differ from zero.
+This is usually done via variants of Student t-test.
 %
-This is usually done via Student t-test of variants of it.
-%
-This will not work here, as double regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
+This will not work here, as back and forth regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
 %
@@
 %
-To use double regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
+To use back and forth regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
 
-Summarizing, double regression is performed as follows:
+Summarizing, back and forth regression is performed as follows:
 \begin{enumerate}
@@
 \subsection{Asymptotic behaviour - linear case}
-As discussed in the introduction, our problem can be formulated (up to a redifinition of the features of X) as $Y=f(EX+N)$, with F an unknown function.
+As discussed in the introduction, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(EX+N)$, with F an unknown function.
 %
@@
 
-Since N is centred, the first least square regression (X from Y), amounts to finding
+The two steps of back and forth regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth being Frobenius norms).
+
+The second step is straightforward, replacing Y and since N is centred, we have
+\begin{equation}
+\begin{aligned}
+\arg \min_H \left \| \hat GY - HX \right \|^2 &=\arg  \min_H \left \| \hat GF(EX+N) - HX \right \|^2 \\
+&=\arg \min_H \left \| (\hat G(FE)-H)X \right \| ^2 + \left \| \hat GFN \right \| ^2\\ 
+&= \arg \min_H \left \| (\hat G(FE)-H)X \right \| ^2\\
+&=\hat G (FE)
+\end{aligned}
+\end{equation}
+
+For the first step, we have 
 \begin{equation}
@@
 \min_G \left \| X-GY \right \|^2 &= \min_G \left \| X - GF(EX+N)\right\|^2 \\
-&{}= \min_G \left \| (I-GFE)X\right\| ^2 + \left \| GFN\right \| ^2\\
+&= \min_G \left \| (I-GFE)X\right\| ^2 + \left \| GFN\right \| ^2\\
+&= \min_G \sum_{i} \left \| (I-GFE)X_{i}\right\| ^2 + \left \| GFN_{i}\right \| ^2\\
 &{}= \min_G \alpha \left \| I-GFE\right\| ^2  + \beta \left \| G\right \| ^2\\
@@
 with $\alpha$ and $\beta$ positive constants, the latter equal to the norm of FN.
-We are balancing two minimisation, with $/alpha$ close to zero, the solution is the null matrix, with $\beta$ close to zero, it is the pseudo inverse of $FE$.
+We are balancing two minimisation, with $\alpha$ close to zero, the solution is the null matrix, with $\beta$ close to zero, it is the pseudo inverse of $FE$.
 %
+FC: lost here .
+I'd like to prove that we recover a scales version $(FE)^\dagger D$ with D a diagonal scaling over dimensions of X, according to signal noise ratio, covariance of X and covariance of N.
+This is what we sort of get from
+
+SVD of X, USV, so the minimum becomes
 
@@
 
-The second regression is from X to the predicted value of X in the above, or $\hat G Y$
-\begin{equation}
-\begin{aligned}
-arg \min_H \left \| k(FE)^{\dagger}Y - HX \right \|^2 &= arg \min_H \left \| k(FE)^{\dagger}F(EX+N) - HX \right \|^2 \\
-&= arg \min_H \left \| (k(FE)^{\dagger}(FE)-H)X \right \| ^2 + \left \| k(FE)^{\dagger}FN \right \| ^2\\
-&= arg \min_H \left \| (k(FE)^{\dagger}(FE)-H)X \right \| ^2\\
-&=k(FE)^{\dagger}(FE)
-\end{aligned}
-\end{equation}
-Thus, in the linear case, double regression retrieves $k(FE)^{\dagger}(FE)$.
+Thus, in the linear case, back and forth regression retrieves $k(FE)^{\dagger}(FE)$.
 In many cases, if v s the variance of $FN$, k can be approximated as $k= {1\over{1+v}}$, it quantifies the amount of noise added to the causal process, and the difficulty of recovering the causal features.
@@
 \section{Related works}
-\subsection{Double regression as a special case of canonical component analysis}
-Formula (1) shows that double regression can be seen as a special case of canonical component analysis (CCA).
+\subsection{back and forth regression as a special case of canonical component analysis}
+Formula (1) shows that back and forth regression can be seen as a special case of canonical component analysis (CCA).
 %
@@
 %
-Thus, double regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
+Thus, back and forth regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
 %
-The closer we are to this hypothesis, the closer the results of CCA and double regression will be.
+The closer we are to this hypothesis, the closer the results of CCA and back and forth regression will be.
 
-There are differences between CCA and double regression.
+There are differences between CCA and back and forth regression.
 %
@@
 \section{Experiments - simulated data}
-Experiments on simulated data serve three purposes: compare double regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.
+Experiments on simulated data serve three purposes: compare back and forth regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.
 
@@
 
-We evaluate the performance of double regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
+We evaluate the performance of back and forth regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
 %
@@
 \subsection{Feature extraction - linear case}
-All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by double regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
+All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by back and forth regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
 %
@@
 
-For feature extraction, double regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
+For feature extraction, back and forth regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
 %
@@
 %
-Double regression allows for good feature extraction if :
+back and forth regression allows for good feature extraction if :
 
@@
 %
-This demonstrates that double regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.
+This demonstrates that back and forth regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.
 
@@
 
-The rationale behind such tests is that whereas double regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.
+The rationale behind such tests is that whereas back and forth regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.
 

### 2019-05-19 code/synthetic_experiment/data.py
Synthetic testbed

@@ -1,0 +1,66 @@
+import numpy as np
+from scipy.linalg import sqrtm
+
+
+def rolling_covariance(rho, dim):
+    cov = np.zeros((dim, dim))
+    for i in range(cov.shape[0]):
+        for j in range(cov.shape[1]):
+            cov[i, j] = np.power(rho, np.abs(i - j) / dim)
+    return cov
+
+
+class Synthetic(object):
+    def __init__(self,            # number of samples
+                dim_x=50,         # number of features
+                dim_y=30,         # number of sensors
+                rho_x=0.5,        # correlation of features
+                rho_n=0.5,        # correlation of noise
+                nc=5,             # number of selected features
+                snr=1.0,          # signal-to-noise ratio
+                nonlinear=False): # number of selected features
+        # linear transformation
+        self.F = np.random.randn(dim_x, dim_y) / np.sqrt(dim_x)
+
+        # masking transformation
+        self.E = np.array([0] * (dim_x - nc) + [1] * (nc))
+        np.random.shuffle(self.E)
+        self.E = np.diag(self.E)
+
+        # features covariance
+        self.cov_X = rolling_covariance(rho_x, dim_x)
+        self.cols_X = np.random.permutation(dim_x)
+
+        # noise covariance
+        self.cov_N = rolling_covariance(rho_n, dim_x)
+        self.cols_N = np.random.permutation(dim_x)
+
+        self.dim_x = dim_x
+        self.dim_y = dim_y
+        self.nonlinear = nonlinear
+        self.snr = snr
+
+    def sample(self,
+               n_samples=1000,
+               in_domain=True):
+        X = np.random.randn(n_samples, self.dim_x)
+        if in_domain:
+            X = X @ sqrtm(self.cov_X)
+        X = X[:, self.cols_X]
+
+        N = np.random.randn(n_samples, self.dim_x)
+        if in_domain:
+            N = N @ sqrtm(self.cov_N)
+        N = N[:, self.cols_N]
+
+        # observed sensor data
+        Y = (self.snr * X @ self.E + N) @ self.F
+
+        if self.nonlinear:
+            Y = 1. / (1. + math.exp(-Y))
+
+        # return inputs, outputs, and solution
+        return X, Y
+
+    def solution(self):
+        return np.diag(self.E)

### 2019-05-19 code/synthetic_experiment/main.py

@@ -1,0 +1,85 @@
+from data import Synthetic
+from models import jrr, OLS
+
+import numpy as np
+import argparse
+
+
+def nmse(y_pred, y_true):
+    num = np.power(np.linalg.norm(y_pred - y_true), 2)
+    den = np.power(np.linalg.norm(y_true), 2)
+    return num / den
+
+
+def compute_false_positives(e_pred, e_true):
+    return np.mean(((e_pred == 1) * (e_true == 0) * 1.0))
+
+
+def compute_false_negatives(e_pred, e_true):
+    return np.mean(((e_pred == 0) * (e_true == 1) * 1.0))
+
+
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser(description='JRR synthetic experiment')
+    parser.add_argument('--model', type=str, default='ols')
+    parser.add_argument('--n_samples', type=int, default=1000)
+    parser.add_argument('--dim_x', type=int, default=50)
+    parser.add_argument('--dim_y', type=int, default=30)
+    parser.add_argument('--rho_x', type=float, default=0.5)
+    parser.add_argument('--rho_n', type=float, default=0.5)
+    parser.add_argument('--snr', type=float, default=0.1)
+    parser.add_argument('--nc', type=int, default=5)
+    parser.add_argument('--nonlinear', type=int, default=0)
+    args = parser.parse_args()
+
+    synthetic = Synthetic(args.dim_x,
+                          args.dim_y,
+                          args.rho_x,
+                          args.rho_n,
+                          args.nc,
+                          args.snr,
+                          args.nonlinear)
+
+    # training data
+    x_tr, y_tr = synthetic.sample(args.n_samples)
+
+    # testing data (in-domain)
+    x_te_in, y_te_in = synthetic.sample(args.n_samples * 10)
+
+    # testing data (out-domain)
+    x_te_out, y_te_out = synthetic.sample(args.n_samples * 10, in_domain=False)
+
+    if args.model == "ols":
+        model = OLS()
+    elif args.model == "jrr":
+        raise NotImplementedError
+    else:
+        raise NotImplementedError
+
+    # fit model on training data
+    model.fit(x_tr, y_tr)
+
+    error_in_all = nmse(model.predict(x_te_in), y_te_in)
+    error_out_all = nmse(model.predict(x_te_out), y_te_out)
+
+    # this is the binary mask E estimated by the model
+    mask = model.solution()
+
[…]

### 2019-05-19 code/synthetic_experiment/models.py

@@ -1,0 +1,78 @@
+from utils import sonquist_morgan
+import numpy as np
+
+def ridge_cv(X,
+             Y,
+             alphas=np.logspace(-5, 5, 20),
+             independent_alphas=True):
+    """
+    Similar to sklearn RidgeCV but
+    (1) can optimize a different alpha for each column of Y (independent_alphas=True)
+    (2) return leave-one-out Y_hat
+    """
+    if isinstance(alphas, (float, int)):
+        alphas = np.array([alphas, ], np.float64)
+    n, n_x = X.shape
+    n, n_y = Y.shape
+    # Decompose X
+    U, s, _ = np.linalg.svd(X, full_matrices=0)
+    v = s**2
+    UY = U.T @ Y
+
+    # For each alpha, solve leave-one-out error coefs
+    cv_duals = np.zeros((len(alphas), n, n_y))
+    cv_errors = np.zeros((len(alphas), n, n_y))
+    for alpha_idx, alpha in enumerate(alphas):
+        # Solve
+        w = ((v + alpha) ** -1) - alpha ** -1
+        c = U @ np.diag(w) @ UY + alpha ** -1 * Y
+        cv_duals[alpha_idx] = c
+
+        # compute diagonal of the matrix: dot(Q, dot(diag(v_prime), Q^T))
+        G_diag = (w * U ** 2).sum(axis=-1) + alpha ** -1
+        error = c / G_diag[:, np.newaxis]
+        cv_errors[alpha_idx] = error
+
+    # identify best alpha for each column of Y independently
+    if independent_alphas:
+        best_alphas = (cv_errors ** 2).mean(axis=1).argmin(axis=0)
+        duals = np.transpose([cv_duals[b, :, i]
+                              for i, b in enumerate(best_alphas)])
+        cv_errors = np.transpose([cv_errors[b, :, i]
+                                  for i, b in enumerate(best_alphas)])
+    else:
+        _cv_errors = cv_errors.reshape(len(alphas), -1)
+        best_alphas = (_cv_errors ** 2).mean(axis=1).argmin(axis=0)
+        duals = cv_duals[best_alphas]
+        cv_errors = cv_errors[best_alphas]
+
+    coefs = duals.T @ X
+    Y_hat = Y - cv_errors
+    return coefs, best_alphas, Y_hat
+
+
+def jrr(X, Y):
[…]

### 2019-05-19 code/synthetic_experiment/utils.py

@@ -1,0 +1,20 @@
+import numpy as np
+
+def sonquist_morgan(x):
+    z=np.sort(x)
+    n=z.size
+    m1=0
+    m2=np.sum(z)
+    mx=0
+    best=-1
+    for i in range(n-1):
+        m1+=z[i]
+        m2-=z[i]
+        ind=(i+1)*(n-i-1)*(m1/(i+1)-m2/(n-i-1))**2
+        if ind>mx :
+            mx=ind
+            best=z[i]
+    res=[0 for i in range(n)]
+    for i in range(n):
+        if x[i]>best: res[i] = 1
+    return np.array(res)

### 2019-05-19 code/synthetic_experiment/main.py
Oracle

@@ -1,8 +1,9 @@
 from data import Synthetic
-from models import jrr, OLS
+from models import jrr, OLS, Oracle
 
 import numpy as np
 import argparse
+import random
 
 
 def nmse(y_pred, y_true):
@@ -30,7 +31,11 @@ if __name__ == "__main__":
     parser.add_argument('--snr', type=float, default=0.1)
     parser.add_argument('--nc', type=int, default=5)
     parser.add_argument('--nonlinear', type=int, default=0)
+    parser.add_argument('--seed', type=int, default=0)
     args = parser.parse_args()
+
+    random.seed(args.seed)
+    np.random.seed(args.seed)
 
     synthetic = Synthetic(args.dim_x,
                           args.dim_y,
@@ -51,6 +56,8 @@ if __name__ == "__main__":
 
     if args.model == "ols":
         model = OLS()
+    elif args.model == "oracle":
+        model = Oracle(synthetic.solution())
     elif args.model == "jrr":
         raise NotImplementedError
     else:
@@ -68,9 +75,9 @@ if __name__ == "__main__":
     false_positives = compute_false_positives(mask, synthetic.solution())
     false_negatives = compute_false_negatives(mask, synthetic.solution())
 
+    # fit a basic model on the selected causes
     selected = np.nonzero(mask)[0]
 
-    # fit a basic model on the selected causes
     model = OLS()
     model.fit(x_tr[:, selected], y_tr)
 

### 2019-05-19 code/synthetic_experiment/models.py

@@ -56,10 +56,6 @@ def jrr(X, Y):
     H, _, _ = ridge_cv(X, X_hat)
     return np.diag(H)
 
-def ols(X, Y):
-    H, _, _ = ridge_cv(X, Y)
-    return np.diag(H)
-
 
 class OLS(object):
     def __init__(self):
@@ -67,6 +63,7 @@ class OLS(object):
 
     def fit(self, X, Y):
         self.coef = np.linalg.pinv(X.T @ X) @ X.T @ Y
+        return self
 
     def predict(self, X):
         return X @ self.coef
@@ -75,4 +72,17 @@ class OLS(object):
         return sonquist_morgan(np.power(self.coef, 2).sum(1))
 
 
+class Oracle(object):
+    def __init__(self, true_mask):
+        self.true_mask = true_mask 
+        self.selected = np.nonzero(self.true_mask)[0]
 
+    def fit(self, X, Y):
+        self.coef = OLS().fit(X[:, self.selected], Y).coef
+
+    def predict(self, X):
+        XS = X[:, self.selected]
+        return XS @ self.coef
+    
+    def solution(self):
+        return self.true_mask 

### 2019-05-19 code/synthetic_experiment/main.py
Simple JRR

@@ -1,5 +1,5 @@
 from data import Synthetic
-from models import jrr, OLS, Oracle
+from models import JRR, OLS, Oracle
 
 import numpy as np
 import argparse
@@ -22,7 +22,7 @@ def compute_false_negatives(e_pred, e_true):
 
 if __name__ == "__main__":
     parser = argparse.ArgumentParser(description='JRR synthetic experiment')
-    parser.add_argument('--model', type=str, default='ols')
+    parser.add_argument('--model', type=str, default='jrr')
     parser.add_argument('--n_samples', type=int, default=1000)
     parser.add_argument('--dim_x', type=int, default=50)
     parser.add_argument('--dim_y', type=int, default=30)
@@ -59,7 +59,7 @@ if __name__ == "__main__":
     elif args.model == "oracle":
         model = Oracle(synthetic.solution())
     elif args.model == "jrr":
-        raise NotImplementedError
+        model = JRR() 
     else:
         raise NotImplementedError
 

### 2019-05-19 code/synthetic_experiment/models.py

@@ -1,60 +1,60 @@
 from utils import sonquist_morgan
+from sklearn.linear_model import RidgeCV, LinearRegression
 import numpy as np
 
-def ridge_cv(X,
-             Y,
-             alphas=np.logspace(-5, 5, 20),
-             independent_alphas=True):
-    """
-    Similar to sklearn RidgeCV but
-    (1) can optimize a different alpha for each column of Y (independent_alphas=True)
-    (2) return leave-one-out Y_hat
-    """
-    if isinstance(alphas, (float, int)):
-        alphas = np.array([alphas, ], np.float64)
-    n, n_x = X.shape
-    n, n_y = Y.shape
-    # Decompose X
-    U, s, _ = np.linalg.svd(X, full_matrices=0)
-    v = s**2
-    UY = U.T @ Y
-
-    # For each alpha, solve leave-one-out error coefs
-    cv_duals = np.zeros((len(alphas), n, n_y))
-    cv_errors = np.zeros((len(alphas), n, n_y))
-    for alpha_idx, alpha in enumerate(alphas):
-        # Solve
-        w = ((v + alpha) ** -1) - alpha ** -1
-        c = U @ np.diag(w) @ UY + alpha ** -1 * Y
-        cv_duals[alpha_idx] = c
-
-        # compute diagonal of the matrix: dot(Q, dot(diag(v_prime), Q^T))
-        G_diag = (w * U ** 2).sum(axis=-1) + alpha ** -1
-        error = c / G_diag[:, np.newaxis]
-        cv_errors[alpha_idx] = error
-
-    # identify best alpha for each column of Y independently
-    if independent_alphas:
-        best_alphas = (cv_errors ** 2).mean(axis=1).argmin(axis=0)
-        duals = np.transpose([cv_duals[b, :, i]
-                              for i, b in enumerate(best_alphas)])
-        cv_errors = np.transpose([cv_errors[b, :, i]
-                                  for i, b in enumerate(best_alphas)])
+def basic_regression(X, Y, regularize=True):
+    if regularize:
+        alphas = np.logspace(-5, 5, 20) 
+        w = RidgeCV(fit_intercept=False, alphas=alphas).fit(X, Y).coef_
     else:
-        _cv_errors = cv_errors.reshape(len(alphas), -1)
-        best_alphas = (_cv_errors ** 2).mean(axis=1).argmin(axis=0)
-        duals = cv_duals[best_alphas]
-        cv_errors = cv_errors[best_alphas]
[…]

### 2019-05-19 code/synthetic_experiment/models.py
Simple.

@@ -2,9 +2,10 @@ from utils import sonquist_morgan
 from sklearn.linear_model import RidgeCV, LinearRegression
 import numpy as np
 
+
 def basic_regression(X, Y, regularize=True):
     if regularize:
-        alphas = np.logspace(-5, 5, 20) 
+        alphas = np.logspace(-5, 5, 20)
         w = RidgeCV(fit_intercept=False, alphas=alphas).fit(X, Y).coef_
     else:
         w = LinearRegression(fit_intercept=False).fit(X, Y).coef_
@@ -18,7 +19,7 @@ class JRR(object):
 
     def fit(self, X, Y):
         E = 0
-        for _ in range(self.n_splits): 
+        for _ in range(self.n_splits):
             perm = np.random.permutation(range(len(X)))
             G = basic_regression(Y[perm[0::2]], X[perm[0::2]])
             E += np.diag(basic_regression(X[perm[1::2]], Y[perm[1::2]] @ G))
@@ -33,30 +34,6 @@ class JRR(object):
         return self.E
 
 
-# class JRR(object):
-#     def __init__(self, G=None, H=None, n_splits=10):
-#         self.G = RidgeCV() if G is None else G
-#         self.H = RidgeCV() if H is None else H
-#         self.n_splits = n_splits
-# 
-#     def fit(self, X, Y):
-#         H = list()
-#         for _ in range(self.n_splits):
-#             p = np.random.permutation(range(len(X)))
-#             set1, set2 = p[::2], p[1::2]
-#             self.G.fit(Y[set1], X[set1])
-#             X_hat = self.G.predict(Y)
-#             H += [self.H.fit(X[set2], X_hat[set2]).coef_, ]
-# 
-#         # new step to allow predict Y from X
-#         self.E_ = np.diag(np.diag(np.mean(H, 0)))
-#         self.G.fit(X @ self.E_, Y)
-#         return self
-# 
-#     def predict(self, X):
-#         return self.G.predict(X @ self.E_)
-
-
 class OLS(object):
     def __init__(self):
         pass
@@ -67,7 +44,7 @@ class OLS(object):
 
     def predict(self, X):
         return X @ self.coef
-    
+
     def solution(self):
         return sonquist_morgan(np.power(self.coef, 2).sum(1))
 
@@ -81,7 +58,6 @@ class Oracle(object):
 
     def predict(self, X):
         return X @ self.true_mask @ self.coef
-    
+
     def solution(self):
         return np.diag(self.true_mask)
-

### 2019-05-20 code/synthetic_experiment/plot.py
Testing

@@ -97,10 +97,11 @@ if __name__ == "__main__":
                              competitor_mean,
                              xerr=jrr_variance,
                              yerr=competitor_variance)
+                plt.scatter(jrr_mean, competitior_mean, c="black")
 
             ax = plt.gca()
             ax.plot(ax.get_xlim(), ax.get_ylim(), ls="--", c=".8")
             plt.margins(0)
             plt.tight_layout(0, 0, 0)
-            plt.savefig(model + "_" + result + ".pdf")
+            plt.savefig("results/" + model + "_" + result + ".pdf")
             plt.show()

### 2019-05-20 code/synthetic_experiment/plot.py
Changes to plotting

@@ -10,13 +10,13 @@ stats = {
     "result_error_in_mask" : "error in-domain (masked)",
     "result_error_out_mask" : "error out-domain (masked)",
     "result_false_positives" : "false positives",
-    "result_false_negatives" : "false_negatives"
+    "result_false_negatives" : "false negatives"
 }
         
-#plt.rc('text', usetex=True)
-#plt.rc('text.latex', preamble=r'\usepackage{times}')
-#plt.rc('font', family='serif')
-#plt.rc('font', size=16)
+plt.rc('text', usetex=True)
+plt.rc('text.latex', preamble=r'\usepackage{times}')
+plt.rc('font', family='serif')
+plt.rc('font', size=10)
 
 if __name__ == "__main__":
     parser = argparse.ArgumentParser(description='JRR synthetic experiment')
@@ -79,13 +79,22 @@ if __name__ == "__main__":
 
     models.discard("jrr")
 
-    for r, result in enumerate(result_names):
-        for model in models:
-            ax = plt.figure(figsize=(4, 4))
-            plt.title(stats[result])
-            plt.xlabel(model.upper())
-            plt.ylabel("jrr".upper())
+    plot_rows = len(models)
+    plot_columns = len(stats)
+    plot_counter = 1
 
+    plt.figure(figsize=(len(stats) * 2 + 0.85, len(models) * 2))
+
+    for m, model in enumerate(models):
+        for r, result in enumerate(result_names):
+            plt.subplot(plot_rows, plot_columns, plot_counter)
+            plot_counter += 1
+
+            if r == 0:
+                plt.ylabel(model.upper())
+            if m == 0:
+                plt.title(stats[result], fontsize=10)
+            
             for eid in results:
                 competitor_mean = results[eid][model]["mean"][r]
                 competitor_variance = results[eid][model]["variance"][r]
@@ -93,15 +102,20 @@ if __name__ == "__main__":
                 jrr_mean = results[eid]["jrr"]["mean"][r]
                 jrr_variance = results[eid]["jrr"]["variance"][r]
 
-                plt.errorbar(jrr_mean,
-                             competitor_mean,
-                             xerr=jrr_variance,
-                             yerr=competitor_variance)
-                plt.scatter(jrr_mean, competitior_mean, c="black")
+                # plt.errorbar(jrr_mean,
+                #              competitor_mean,
+                #              xerr=jrr_variance,
[…]

### 2019-05-20 code/base.py
Cleaning

@@ -1,63 +1,0 @@
-import numpy as np
-from sklearn.preprocessing import scale
-
-
-def make_data(n: int = 1000,  # number of samples
-              nX: int = 100,  # dimensionality of X
-              nY: int = 10,  # dimensionality of Y
-              nM: int = 10,  # dimensionality of source
-              random_seed: int = None,
-              snr: float = 1e-1,  # signal to noise ratio
-              selected: float = .5,  # proportion of selected feature in E
-              E: np.ndarray = None,  # selecting matrix
-              M: np.ndarray = None,  # selecting matrix
-              F: np.ndarray = None,  # mixing matrix
-              Cx: np.ndarray = None,  # feature covariance
-              heteroscedastic: bool = False,  # noise type
-              ) -> (np.ndarray, np.ndarray, np.ndarray, np.ndarray):
-
-    """Y = F(MEX+N*snr)
-    returns X, Y, E, F, M
-    """
-    np.random.seed(random_seed)
-
-    if E is None:
-        E = np.zeros(nX)
-        # selected must be between 1 and nX-1 to ensure that there is at least
-        # one selected and one unselected X feature
-        selected = min(int(np.floor(selected*nX)) + 1, nX-1)
-        E[:selected] = 1
-        E = np.diag(E)
-    else:
-        nX, nX = E.shape
-
-    if M is None:
-        M = np.random.randn(nM, nX)
-        M /= np.sqrt(np.sum(M**2, 1, keepdims=True))
-    else:
-        nM = len(M)
-
-    if Cx is None:
-        Cx = np.random.randn(nX, nX)
-        Cx = Cx.dot(Cx.T) / nX  # sym pos-semidefin
-    else:
-        nX = len(Cx)
-    X = np.random.multivariate_normal(np.zeros(nX), Cx, n)
-
-    N = np.random.randn(n, nM)
-    if heteroscedastic:
-        Cn = np.random.randn(nM, nM)
-        Cn = Cn.dot(Cn.T) / nX  # sym pos-semidefin
-        N = np.random.multivariate_normal(np.zeros(nX), Cn, n)
-
-    if F is None:
-        F = np.random.randn(nY, nM)
-    else:
-        nY = len(F)
-
-    Y = ((X @ E.T @ M.T) * snr + N) @ F.T
-
-    X = scale(X)
-    Y = scale(Y)
-
-    return X, Y, E, F, M

### 2019-05-20 code/synthetic_experiment/main.py
More models

@@ -1,4 +1,4 @@
-from models import JRR, OLS, Oracle
+from models import JRR, OLS, Ridge, Oracle, PLS, Lasso
 from data import Synthetic
 
 import numpy as np
@@ -22,8 +22,8 @@ def compute_false_negatives(e_pred, e_true):
 
 if __name__ == "__main__":
     parser = argparse.ArgumentParser(description='JRR synthetic experiment')
-    parser.add_argument('--model', type=str, default='jrr')
-    parser.add_argument('--n_samples', type=int, default=1000)
+    parser.add_argument('--model', type=str, default='lasso')
+    parser.add_argument('--n_samples', type=int, default=100)
     parser.add_argument('--dim_x', type=int, default=50)
     parser.add_argument('--dim_y', type=int, default=30)
     parser.add_argument('--rho_x', type=float, default=0.5)
@@ -56,10 +56,16 @@ if __name__ == "__main__":
 
     if args.model == "ols":
         model = OLS()
+    elif args.model == "ridge":
+        model = Ridge()
     elif args.model == "oracle":
         model = Oracle(synthetic.solution())
     elif args.model == "jrr":
         model = JRR() 
+    elif args.model == "pls":
+        model = PLS() 
+    elif args.model == "lasso":
+        model = Lasso() 
     else:
         raise NotImplementedError
 

### 2019-05-20 code/synthetic_experiment/models.py

@@ -1,4 +1,7 @@
 from sklearn.linear_model import RidgeCV, LinearRegression
+from sklearn.cross_decomposition import PLSRegression
+from sklearn.linear_model import MultiTaskLassoCV
+from sklearn.model_selection import GridSearchCV
 import numpy as np
 
 def sonquist_morgan(x):
@@ -52,21 +55,6 @@ class JRR(object):
         return self.E
 
 
-class OLS(object):
-    def __init__(self):
-        pass
-
-    def fit(self, X, Y):
-        self.coef = basic_regression(X, Y)
-        return self
-
-    def predict(self, X):
-        return X @ self.coef
-
-    def solution(self):
-        return sonquist_morgan(np.power(self.coef, 2).sum(1))
-
-
 class Oracle(object):
     def __init__(self, true_mask):
         self.true_mask = np.diag(true_mask)
@@ -79,3 +67,42 @@ class Oracle(object):
 
     def solution(self):
         return np.diag(self.true_mask)
+
+
+class SKModel(object):
+    def __init__(self):
+        pass
+
+    def predict(self, X):
+        return X @ self.coef
+    
+    def solution(self):
+        return sonquist_morgan(np.power(self.coef, 2).sum(1))
+
+
+class OLS(SKModel):
+    def fit(self, X, Y):
+        self.coef = LinearRegression(fit_intercept=False).fit(X, Y).coef_.T
+        return self
+
+
+class Ridge(SKModel):
+    def fit(self, X, Y):
+        self.coef = RidgeCV(fit_intercept=False).fit(X, Y).coef_.T
+        return self
+
+
+class PLS(SKModel):
+    def fit(self, X, Y):
+        grid = { "n_components": list(range(1, X.shape[1] + 1)) }
+        pls = GridSearchCV(PLSRegression(), grid, n_jobs=-1, cv=5)
+        self.coef = pls.fit(X, Y).best_estimator_.coef_
+        return self
+
+class Lasso(SKModel):
+    def fit(self, X, Y):
+        self.coef = MultiTaskLassoCV(fit_intercept=False,
+                                     selection="random",
+                                     n_jobs=-1,
+                                     cv=5).fit(X, Y).coef_.T
+        return self

### 2019-05-20 code/synthetic_experiment/data.py
More fixes

@@ -12,13 +12,13 @@ def rolling_covariance(rho, dim):
 
 class Synthetic(object):
     def __init__(self,            # number of samples
-                dim_x=50,         # number of features
-                dim_y=30,         # number of sensors
-                rho_x=0.5,        # correlation of features
-                rho_n=0.5,        # correlation of noise
-                nc=5,             # number of selected features
-                snr=1.0,          # signal-to-noise ratio
-                nonlinear=False): # number of selected features
+                 dim_x=50,         # number of features
+                 dim_y=30,         # number of sensors
+                 rho_x=0.5,        # correlation of features
+                 rho_n=0.5,        # correlation of noise
+                 nc=5,             # number of selected features
+                 snr=1.0,          # signal-to-noise ratio
+                 nonlinear=False):  # number of selected features
         # linear transformation
         self.F = np.random.randn(dim_x, dim_y) / np.sqrt(dim_x)
 

### 2019-05-20 code/synthetic_experiment/main.py

@@ -1,9 +1,24 @@
-from models import JRR, OLS, Ridge, Oracle, PLS, Lasso
+# TODO: model Y = (XE + N)F + N' with second noise N'
+# TODO: uniform distributions
+
+from models import JRR, OLS, Ridge, Oracle, PLS, Lasso, RRR
+from sklearn.exceptions import ConvergenceWarning
 from data import Synthetic
 
 import numpy as np
+import warnings
 import argparse
 import random
+
+models = {
+    "OLS": OLS,
+    "Ridge": Ridge,
+    "Oracle": Oracle,
+    "JRR": JRR,
+    "PLS": PLS,
+    "RRR": RRR,
+    "Lasso": Lasso
+}
 
 
 def nmse(y_pred, y_true):
@@ -20,82 +35,89 @@ def compute_false_negatives(e_pred, e_true):
     return np.mean(((e_pred == 0) * (e_true == 1) * 1.0))
 
 
+def run_experiment(args):
+    for seed in range(args.n_seeds):
+        random.seed(seed)
+        np.random.seed(seed)
+
+        synthetic = Synthetic(args.dim_x,
+                              args.dim_y,
+                              args.rho_x,
+                              args.rho_n,
+                              args.nc,
+                              args.snr,
+                              args.nonlinear)
+
+        mask_true = synthetic.solution()
+
+        # training data
+        x_tr, y_tr = synthetic.sample(args.n_samples)
+
+        # testing data (in-domain)
+        x_te_in, y_te_in = synthetic.sample(args.n_samples * 10)
+
+        # testing data (out-domain)
+        x_te_out, y_te_out = synthetic.sample(args.n_samples * 10,
+                                              in_domain=False)
+
+        results = []
+
+        for m, Model in models.items():
+            if m == "Oracle":
+                model = Model(mask_true)
+            else:
+                model = Model()
+
+            # fit model on training data
+            model.fit(x_tr, y_tr)
+
+            error_in_all = nmse(model.predict(x_te_in), y_te_in)
+            error_out_all = nmse(model.predict(x_te_out), y_te_out)
+
+            # this is the binary mask E estimated by the model
+            mask = model.solution()
+
+            false_positives = compute_false_positives(mask, mask_true)
+            false_negatives = compute_false_negatives(mask, mask_true)
+
+            # fit a basic model on the selected causes
+            sel = np.nonzero(mask)[0]
+
+            model = OLS()
+            model.fit(x_tr[:, sel], y_tr)
+
[…]

### 2019-05-20 code/synthetic_experiment/models.py

@@ -2,25 +2,29 @@ from sklearn.linear_model import RidgeCV, LinearRegression
 from sklearn.cross_decomposition import PLSRegression
 from sklearn.linear_model import MultiTaskLassoCV
 from sklearn.model_selection import GridSearchCV
+from sklearn.base import BaseEstimator
+from sklearn.metrics import r2_score
 import numpy as np
 
+
 def sonquist_morgan(x):
-    z=np.sort(x)
-    n=z.size
-    m1=0
-    m2=np.sum(z)
-    mx=0
-    best=-1
-    for i in range(n-1):
-        m1+=z[i]
-        m2-=z[i]
-        ind=(i+1)*(n-i-1)*(m1/(i+1)-m2/(n-i-1))**2
-        if ind>mx :
-            mx=ind
-            best=z[i]
-    res=[0 for i in range(n)]
+    z = np.sort(x)
+    n = z.size
+    m1 = 0
+    m2 = np.sum(z)
+    mx = 0
+    best = -1
+    for i in range(n - 1):
+        m1 += z[i]
+        m2 -= z[i]
+        ind = (i + 1) * (n - i - 1) * (m1 / (i + 1) - m2 / (n - i - 1))**2
+        if ind > mx:
+            mx = ind
+            best = z[i]
+    res = [0 for i in range(n)]
     for i in range(n):
-        if x[i]>best: res[i] = 1
+        if x[i] > best:
+            res[i] = 1
     return np.array(res)
 
 
@@ -69,13 +73,13 @@ class Oracle(object):
         return np.diag(self.true_mask)
 
 
-class SKModel(object):
+class SKModel(BaseEstimator):
     def __init__(self):
         pass
 
     def predict(self, X):
         return X @ self.coef
-    
+
     def solution(self):
         return sonquist_morgan(np.power(self.coef, 2).sum(1))
 
@@ -94,15 +98,57 @@ class Ridge(SKModel):
 
 class PLS(SKModel):
     def fit(self, X, Y):
-        grid = { "n_components": list(range(1, X.shape[1] + 1)) }
+        grid = {"n_components": np.linspace(1, X.shape[1], 10).astype(int)}
         pls = GridSearchCV(PLSRegression(), grid, n_jobs=-1, cv=5)
         self.coef = pls.fit(X, Y).best_estimator_.coef_
         return self
 
+
 class Lasso(SKModel):
-    def fit(self, X, Y):
[…]

### 2019-05-20 code/synthetic_experiment/plot.py

@@ -5,14 +5,14 @@ from ast import literal_eval
 from matplotlib import pyplot as plt
 
 stats = {
-    "result_error_in_all" : "error in-domain",
-    "result_error_out_all" : "error out-domain",
-    "result_error_in_mask" : "error in-domain (masked)",
-    "result_error_out_mask" : "error out-domain (masked)",
-    "result_false_positives" : "false positives",
-    "result_false_negatives" : "false negatives"
+    "result_error_in_all": "error in-domain",
+    "result_error_out_all": "error out-domain",
+    "result_error_in_mask": "error in-domain (masked)",
+    "result_error_out_mask": "error out-domain (masked)",
+    "result_false_positives": "false positives",
+    "result_false_negatives": "false negatives"
 }
-        
+
 plt.rc('text', usetex=True)
 plt.rc('text.latex', preamble=r'\usepackage{times}')
 plt.rc('font', family='serif')
@@ -65,8 +65,8 @@ if __name__ == "__main__":
         for r, result in enumerate(result_names):
             results[line_str][this_model]["mean"][r] += result_dict[result]
             results[line_str][this_model]["power"][r] += result_dict[result] ** 2
-        
-        results[line_str][this_model]["counter"] += 1 
+
+        results[line_str][this_model]["counter"] += 1
 
     for eid in results:
         for model in results[eid]:
@@ -94,11 +94,11 @@ if __name__ == "__main__":
                 plt.ylabel(model.upper())
             if m == 0:
                 plt.title(stats[result], fontsize=10)
-            
+
             for eid in results:
                 competitor_mean = results[eid][model]["mean"][r]
                 competitor_variance = results[eid][model]["variance"][r]
-                
+
                 jrr_mean = results[eid]["jrr"]["mean"][r]
                 jrr_variance = results[eid]["jrr"]["variance"][r]
 

### 2019-05-20 code/synthetic_experiment/data.py
Fix

@@ -3,6 +3,9 @@ from scipy.linalg import sqrtm
 
 
 def rolling_covariance(rho, dim):
+    if rho == 0:
+        return np.eye(dim)
+
     cov = np.zeros((dim, dim))
     for i in range(cov.shape[0]):
         for j in range(cov.shape[1]):

### 2019-05-20 code/synthetic_experiment/data.py
Fixes

@@ -7,9 +7,15 @@ def rolling_covariance(rho, dim):
         return np.eye(dim)
 
     cov = np.zeros((dim, dim))
+    sign = int(np.sign(rho))
+    rho_abs = np.abs(rho)
+
     for i in range(cov.shape[0]):
         for j in range(cov.shape[1]):
-            cov[i, j] = np.power(rho, np.abs(i - j) / dim)
+            if i == j:
+                cov[j, j] = 1.0
+            else:
+                cov[i, j] = sign * np.power(rho_abs, np.abs(i - j) / dim)
     return cov
 
 

### 2019-05-20 code/synthetic_experiment/models.py

@@ -6,6 +6,7 @@ from sklearn.base import BaseEstimator
 from sklearn.metrics import r2_score
 import numpy as np
 
+N_JOBS = 1
 
 def sonquist_morgan(x):
     z = np.sort(x)
@@ -99,7 +100,7 @@ class Ridge(SKModel):
 class PLS(SKModel):
     def fit(self, X, Y):
         grid = {"n_components": np.linspace(1, X.shape[1], 10).astype(int)}
-        pls = GridSearchCV(PLSRegression(), grid, n_jobs=-1, cv=5)
+        pls = GridSearchCV(PLSRegression(), grid, n_jobs=N_JOBS, cv=5)
         self.coef = pls.fit(X, Y).best_estimator_.coef_
         return self
 
@@ -113,7 +114,7 @@ class Lasso(SKModel):
         self.coef = MultiTaskLassoCV(fit_intercept=False,
                                      selection="random",
                                      n_alphas=10,
-                                     n_jobs=-1,
+                                     n_jobs=N_JOBS,
                                      cv=5).fit(X[p], Y[p]).coef_.T
         return self
 
@@ -149,6 +150,6 @@ class RRR(SKModel):
             "reg": np.logspace(-3, 3, 10)
         }
 
-        rrr = GridSearchCV(RRROne(), grid, n_jobs=-1, cv=5)
+        rrr = GridSearchCV(RRROne(), grid, n_jobs=N_JOBS, cv=5)
         self.coef = rrr.fit(X, Y).best_estimator_.coef
         return self

### 2019-05-20 code/synthetic_experiment/data.py
New data

@@ -1,48 +1,29 @@
-import numpy as np
 from scipy.linalg import sqrtm
-
-
-def rolling_covariance(rho, dim):
-    if rho == 0:
-        return np.eye(dim)
-
-    cov = np.zeros((dim, dim))
-    sign = int(np.sign(rho))
-    rho_abs = np.abs(rho)
-
-    for i in range(cov.shape[0]):
-        for j in range(cov.shape[1]):
-            if i == j:
-                cov[j, j] = 1.0
-            else:
-                cov[i, j] = sign * np.power(rho_abs, np.abs(i - j) / dim)
-    return cov
+import numpy as np
+import math
 
 
 class Synthetic(object):
-    def __init__(self,            # number of samples
+    def __init__(self,             # number of samples
                  dim_x=50,         # number of features
                  dim_y=30,         # number of sensors
-                 rho_x=0.5,        # correlation of features
-                 rho_n=0.5,        # correlation of noise
                  nc=5,             # number of selected features
                  snr=1.0,          # signal-to-noise ratio
                  nonlinear=False):  # number of selected features
         # linear transformation
-        self.F = np.random.randn(dim_x, dim_y) / np.sqrt(dim_x)
+        self.F = np.random.randn(dim_x, dim_y) / math.sqrt(dim_x)
 
         # masking transformation
         self.E = np.array([0] * (dim_x - nc) + [1] * (nc))
-        np.random.shuffle(self.E)
         self.E = np.diag(self.E)
 
         # features covariance
-        self.cov_X = rolling_covariance(rho_x, dim_x)
-        self.cols_X = np.random.permutation(dim_x)
+        self.cov_X1 = np.random.randn(dim_x, dim_x) / math.sqrt(dim_x)
+        self.cov_X2 = np.random.randn(dim_x, dim_x) / math.sqrt(dim_x)
 
         # noise covariance
-        self.cov_N = rolling_covariance(rho_n, dim_x)
-        self.cols_N = np.random.permutation(dim_x)
+        self.cov_N1 = np.random.randn(dim_x, dim_x) / math.sqrt(dim_x)
+        self.cov_N2 = np.random.randn(dim_x, dim_x) / math.sqrt(dim_x)
 
         self.dim_x = dim_x
         self.dim_y = dim_y
@@ -52,15 +33,12 @@ class Synthetic(object):
     def sample(self,
                n_samples=1000,
                in_domain=True):
[…]

### 2019-05-20 code/synthetic_experiment/plot.py

@@ -20,7 +20,7 @@ plt.rc('font', size=10)
 
 if __name__ == "__main__":
     parser = argparse.ArgumentParser(description='JRR synthetic experiment')
-    parser.add_argument('--file', type=str, default='results.txt')
+    parser.add_argument('--file', type=str, default='results_05.txt')
     args = parser.parse_args()
 
     with open(args.file, "r") as f:
@@ -77,13 +77,15 @@ if __name__ == "__main__":
             results[eid][model].pop("power")
             results[eid][model].pop("counter")
 
-    models.discard("jrr")
+    models.discard("JRR")
 
     plot_rows = len(models)
     plot_columns = len(stats)
     plot_counter = 1
 
     plt.figure(figsize=(len(stats) * 2 + 0.85, len(models) * 2))
+
+    print(models)
 
     for m, model in enumerate(models):
         for r, result in enumerate(result_names):
@@ -96,17 +98,17 @@ if __name__ == "__main__":
                 plt.title(stats[result], fontsize=10)
 
             for eid in results:
-                competitor_mean = results[eid][model]["mean"][r]
-                competitor_variance = results[eid][model]["variance"][r]
+                    competitor_mean = results[eid][model]["mean"][r]
+                    competitor_variance = results[eid][model]["variance"][r]
 
-                jrr_mean = results[eid]["jrr"]["mean"][r]
-                jrr_variance = results[eid]["jrr"]["variance"][r]
+                    jrr_mean = results[eid]["JRR"]["mean"][r]
+                    jrr_variance = results[eid]["JRR"]["variance"][r]
 
-                # plt.errorbar(jrr_mean,
-                #              competitor_mean,
-                #              xerr=jrr_variance,
-                #              yerr=competitor_variance)
-                plt.scatter(jrr_mean, competitor_mean, c="black", alpha=0.5)
+                    # plt.errorbar(jrr_mean,
+                    #              competitor_mean,
+                    #              xerr=jrr_variance,
+                    #              yerr=competitor_variance)
+                    plt.scatter(jrr_mean, competitor_mean, c="black", alpha=0.5)
 
             ax = plt.gca()
             xl = ax.get_xlim()

### 2019-05-20 code/synthetic_experiment/main.py
Fix

@@ -42,8 +42,6 @@ def run_experiment(args):
 
         synthetic = Synthetic(args.dim_x,
                               args.dim_y,
-                              args.rho_x,
-                              args.rho_n,
                               args.nc,
                               args.snr,
                               args.nonlinear)

### 2019-05-20 code/synthetic_experiment/models.py
Sync

@@ -50,14 +50,14 @@ class JRR(object):
             G = basic_regression(Y[perm[0::2]], X[perm[0::2]])
             E += np.diag(basic_regression(X[perm[1::2]], Y[perm[1::2]] @ G))
 
-        self.E = sonquist_morgan(E / self.n_splits)
+        self.E = E / self.n_splits
         self.coef = basic_regression(X @ np.diag(self.E), Y)
 
     def predict(self, X):
         return X @ np.diag(self.E) @ self.coef
 
     def solution(self):
-        return self.E
+        return sonquist_morgan(self.E)
 
 
 class Oracle(object):
@@ -93,7 +93,8 @@ class OLS(SKModel):
 
 class Ridge(SKModel):
     def fit(self, X, Y):
-        self.coef = RidgeCV(fit_intercept=False).fit(X, Y).coef_.T
+        self.coef = RidgeCV(fit_intercept=False,
+                            alphas=np.logspace(-3, 3, 10)).fit(X, Y).coef_.T
         return self
 
 

### 2019-05-20 code/synthetic_experiment/plot.py

@@ -10,7 +10,8 @@ stats = {
     "result_error_in_mask": "error in-domain (masked)",
     "result_error_out_mask": "error out-domain (masked)",
     "result_false_positives": "false positives",
-    "result_false_negatives": "false negatives"
+    "result_false_negatives": "false negatives",
+    "result_auc": "AUC"
 }
 
 plt.rc('text', usetex=True)
@@ -98,6 +99,7 @@ if __name__ == "__main__":
                 plt.title(stats[result], fontsize=10)
 
             for eid in results:
+                if ("JRR" in results[eid]) and (model in results[eid]):
                     competitor_mean = results[eid][model]["mean"][r]
                     competitor_variance = results[eid][model]["variance"][r]
 

### 2019-05-20 code/synthetic_experiment/main.py
Sync done

@@ -2,11 +2,10 @@
 # TODO: uniform distributions
 
 from models import JRR, OLS, Ridge, Oracle, PLS, Lasso, RRR
-from sklearn.exceptions import ConvergenceWarning
+from sklearn.metrics import roc_auc_score
 from data import Synthetic
 
 import numpy as np
-import warnings
 import argparse
 import random
 
@@ -81,7 +80,7 @@ def run_experiment(args):
             # fit a basic model on the selected causes
             sel = np.nonzero(mask)[0]
 
-            model = OLS()
+            model = Ridge()
             model.fit(x_tr[:, sel], y_tr)
 
             error_in_mask = nmse(model.predict(x_te_in[:, sel]), y_te_in)
@@ -96,6 +95,7 @@ def run_experiment(args):
             result["result_error_out_mask"] = error_out_mask
             result["result_false_positives"] = false_positives
             result["result_false_negatives"] = false_negatives
+            result["result_auc"] = roc_auc_score(mask, mask_true) 
             results.append(result)
 
             print(results[-1])
@@ -108,14 +108,10 @@ if __name__ == "__main__":
     parser.add_argument('--n_samples', type=int, default=1000)
     parser.add_argument('--dim_x', type=int, default=100)
     parser.add_argument('--dim_y', type=int, default=100)
-    parser.add_argument('--rho_x', type=float, default=0.5)
-    parser.add_argument('--rho_n', type=float, default=0.5)
     parser.add_argument('--snr', type=float, default=0.1)
     parser.add_argument('--nc', type=int, default=5)
     parser.add_argument('--nonlinear', type=int, default=0)
     parser.add_argument('--n_seeds', type=int, default=10)
     args = parser.parse_args()
-
-    warnings.filterwarnings("ignore", category=ConvergenceWarning)
 
     run_experiment(args)

### 2019-05-20 code/synthetic_experiment/main.py
Fixed masks

@@ -20,6 +20,27 @@ models = {
 }
 
 
+def sonquist_morgan(x):
+    z = np.sort(x)
+    n = z.size
+    m1 = 0
+    m2 = np.sum(z)
+    mx = 0
+    best = -1
+    for i in range(n - 1):
+        m1 += z[i]
+        m2 -= z[i]
+        ind = (i + 1) * (n - i - 1) * (m1 / (i + 1) - m2 / (n - i - 1))**2
+        if ind > mx:
+            mx = ind
+            best = z[i]
+    res = [0 for i in range(n)]
+    for i in range(n):
+        if x[i] > best:
+            res[i] = 1
+    return np.array(res)
+
+
 def nmse(y_pred, y_true):
     num = np.power(np.linalg.norm(y_pred - y_true), 2)
     den = np.power(np.linalg.norm(y_true), 2)
@@ -78,7 +99,7 @@ def run_experiment(args):
             false_negatives = compute_false_negatives(mask, mask_true)
 
             # fit a basic model on the selected causes
-            sel = np.nonzero(mask)[0]
+            sel = np.nonzero(sonquist_morgan(mask))[0]
 
             model = Ridge()
             model.fit(x_tr[:, sel], y_tr)
@@ -95,7 +116,7 @@ def run_experiment(args):
             result["result_error_out_mask"] = error_out_mask
             result["result_false_positives"] = false_positives
             result["result_false_negatives"] = false_negatives
-            result["result_auc"] = roc_auc_score(mask, mask_true) 
+            result["result_auc"] = roc_auc_score(mask_true, mask) 
             results.append(result)
 
             print(results[-1])

### 2019-05-20 code/synthetic_experiment/models.py

@@ -8,26 +8,6 @@ import numpy as np
 
 N_JOBS = 1
 
-def sonquist_morgan(x):
-    z = np.sort(x)
-    n = z.size
-    m1 = 0
-    m2 = np.sum(z)
-    mx = 0
-    best = -1
-    for i in range(n - 1):
-        m1 += z[i]
-        m2 -= z[i]
-        ind = (i + 1) * (n - i - 1) * (m1 / (i + 1) - m2 / (n - i - 1))**2
-        if ind > mx:
-            mx = ind
-            best = z[i]
-    res = [0 for i in range(n)]
-    for i in range(n):
-        if x[i] > best:
-            res[i] = 1
-    return np.array(res)
-
 
 def basic_regression(X, Y, regularize=True):
     if regularize:
@@ -57,7 +37,7 @@ class JRR(object):
         return X @ np.diag(self.E) @ self.coef
 
     def solution(self):
-        return sonquist_morgan(self.E)
+        return self.E
 
 
 class Oracle(object):
@@ -82,7 +62,7 @@ class SKModel(BaseEstimator):
         return X @ self.coef
 
     def solution(self):
-        return sonquist_morgan(np.power(self.coef, 2).sum(1))
+        return np.power(self.coef, 2).sum(1)
 
 
 class OLS(SKModel):

### 2019-05-20 code/synthetic_experiment/main.py
Adding CCA

@@ -1,7 +1,7 @@
 # TODO: model Y = (XE + N)F + N' with second noise N'
 # TODO: uniform distributions
 
-from models import JRR, OLS, Ridge, Oracle, PLS, Lasso, RRR
+from models import JRR, OLS, Ridge, Oracle, PLS, Lasso, RRR, CCA
 from sklearn.exceptions import ConvergenceWarning
 from data import Synthetic
 
@@ -13,6 +13,7 @@ import random
 models = {
     "OLS": OLS,
     "Ridge": Ridge,
+    "CCA": CCA,
     "Oracle": Oracle,
     "JRR": JRR,
     "PLS": PLS,

### 2019-05-20 code/synthetic_experiment/models.py

@@ -4,6 +4,7 @@ from sklearn.linear_model import MultiTaskLassoCV
 from sklearn.model_selection import GridSearchCV
 from sklearn.base import BaseEstimator
 from sklearn.metrics import r2_score
+import scipy as sc
 import numpy as np
 
 N_JOBS = 1
@@ -94,7 +95,7 @@ class OLS(SKModel):
 class Ridge(SKModel):
     def fit(self, X, Y):
         self.coef = RidgeCV(fit_intercept=False,
-                            alphas=np.logspace(-3, 3, 10)).fit(X, Y).coef_.T
+                            alphas=np.logspace(-5, 5, 20)).fit(X, Y).coef_.T
         return self
 
 
@@ -120,7 +121,7 @@ class Lasso(SKModel):
         return self
 
 
-class RRROne(SKModel):
+class ReducedRankRegression(SKModel):
     """
     Adapted from: https://github.com/riscy/machine_learning_linear_models
     """
@@ -148,9 +149,54 @@ class RRR(SKModel):
     def fit(self, X, Y):
         grid = {
             "n_components": np.linspace(1, X.shape[1], 10).astype(int),
-            "reg": np.logspace(-3, 3, 10)
+            "reg": np.logspace(-5, 5, 20)
         }
 
-        rrr = GridSearchCV(RRROne(), grid, n_jobs=N_JOBS, cv=5)
+        rrr = GridSearchCV(ReducedRankRegression(), grid, n_jobs=N_JOBS, cv=5)
         self.coef = rrr.fit(X, Y).best_estimator_.coef
+        return self
+
+
+class CanonicalCorrelation(SKModel):
+    def __init__(self, n_components=-1, reg=0):
+        self.n_components = n_components
+        self.reg = reg
+
+    def fit(self, X, Y):
+        if self.n_components < 1:
+            self.n_components = min(X.shape[1], Y.shape[1])
+
+        CXX = (X.T @ X) / X.shape[0]
+        CYY = (Y.T @ Y) / Y.shape[0]
+        CXY = (X.T @ Y) / X.shape[0]
+
+        sqrtCX = sc.linalg.sqrtm(CXX + np.eye(CXX.shape[0]) * self.reg)
+        sqrtCY = sc.linalg.sqrtm(CYY + np.eye(CYY.shape[0]) * self.reg)
+    
+        isqrtCX = sc.linalg.inv(sqrtCX)
+        isqrtCY = sc.linalg.inv(sqrtCY)
[…]

### 2019-05-20 paper/meg_exps.tex
adding exp section in separate file

\subsection{The MOUS dataset}

Next, we apply our method to brain imaging data, on the ``Mother Of Unification
Studies'' (MOUS) dataset \cite{schoffelen2019204}. The dataset consists of
anonymized multimodal neuroimaging data that has been acquired from 204 healthy
human subjects, which participated in either a auditory or a visual version of a
language experiment. In this work, we use the subset corresponding to the visual
variant recorded using magneto-encephalography (MEG).

Subjects were presented a set of 120 sentences in Dutch, and a scrambled lists
of the same words for a controlled experiment. Each word was presented on a
screen for no less than 300 ms and no more than 1400ms, 351 ms on average.
Successive words were separated by an empty screen for 300ms, and successive
sentences were separated by an empty screen for a few (3-4) seconds.

The data contains the timestamp for each word presentation, allowing to align
time series to stimuli onsets for all words.

As a result, we obtain, for each of the 301 sensors and each word presented, a
time series of 67 real-values spanning $\left[-100ms, 1500ms\right]$ relative to
the stimulus onset. Each of the 102 subjects (we used data from 93 subjects) was
presented approximately 2700 words during the experiment.

\subsection{Application of the method}

### 2019-05-20 paper/paper.tex
adding exp section in separate file
theorem expanded

@@
 \usepackage[]{neurips_2019}
+\DeclareMathOperator{\Tr}{Tr}
 
@@
 %
-Of all possible causes (features of X), only a few account for Y, let them be represented as EX, with E a binary diagonal matrix: all zeroes except a few ones on the diagonal, which correspond to active causal features of X.
+Of all possible causes (features of X), only a few account for Y, let them be represented as EX, with E a binary diagonal matrix: a diagonal of zeroes and ones, which correspond to active causal features of X.
 %
@@
 %
-We first predict X from Y, as in the inverse approach, and retrieve a regularised least square estimate $\hat X$.
-%
-Then, we calculate a regularised least square estimator of $\hat X$ from X.
+We first predict X from Y, as in the inverse approach, and retrieve a robust least square estimate $\hat X$. Then, we calculate a robust least square estimator of $\hat X$ from X.
 %
@@
 
-In practice, since the covariance matrices of Y and X might exhibit multicollinearity, we use regularisation in both matrix inversions.
-This amounts to replacing $(X'X)^{-1}$ and $(Y'Y)^{-1}$ by $(X'X+\lambda)^{-1}$ and $(Y'Y+\mu)^{-1}$, and is done by using ridge regression, regularisation parameters being estimated by leave one out cross validation.
+In practice, the covariance matrices of Y and X might exhibit multicollinearity and be badly conditioned, which make their inversion unstable.
+To guard against this, we compute their singular value decompositions, and truncate their lowest eigenvalues so as to reach a specific condition number in the Frobenius norm.
+We prefer the truncated SVD approach over the more common regularisation (ridge regression) because it introduces less bias in the estimation.
+A small amount of regularisation is still used, for numerical stability purposes.
+
+If $X'X=U'DU$, U orthogonal and D diagonal, and D* is D with the lowest diagonal elements set to zero, this amounts to replacing $(X'X)^{-1}$ by $U(D*^{-1}+\epsilon I)U'$, with $\epsilon$ a positive number close to machine floating point accuracy.
+D is truncated so as to have a certain condition number that is determined by leave one out cross validation.
 %
@@
 %
-This is repeated over several (10) splits, and the resulting $\hat H$ are averaged.
+This is repeated over several (100) splits, and the resulting $\hat H$ are averaged.
 %
@@
 
-The two steps of back and forth regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth being Frobenius norms).
+The two steps of back and forth regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).
+
+For the first step, replacing Y, and since N is centred, we have 
+\begin{equation}
+\begin{aligned}
+\arg \min_G \left \| X-GY \right \|^2 &= \arg \min_G \left \| X - GF(EX+N)\right\|^2 \\
+&{}= \arg \min_G \left \| (I-GFE)X\right\| ^2  + \left \| GFN\right \| ^2\\
+&{}= \arg \min_G \Tr (X'(I-GFE)'(I-GFE)X) + \Tr (N'F'G'GFN)\\
+&{}= \arg \min_G \Tr ((I-GFE)'(I-GFE)XX') + \Tr (G'G(FN)(FN)')\\
+\end{aligned}
+\end{equation}
+
+As noise N gets close to zero, and if F and X have full rank, then the solution tends to $(FE)^\dagger$, the pseudo inverse of FE.
+
+If matrices X'X and (FN)'(FN) have low condition numbers, and denoting by VarX and VarN the variance of X and FN, we have $$\min_G \Tr ((I-GFE)'(I-GFE)XX') + \Tr (G'GFNN'F') \approx \left \| (I-GFE)\right\| ^2 VarX  + \left \| G\right \| ^2 VarN$$and the solution has the form $k(FE)^\dagger$, k a positive constant below one.
+To determine an approximate value of k, note that if r is the rank of FE (the number of ones in E) $$\left \| (I-GFE)\right\| ^2 = r (1-k)^2 + dx - r$$ and that under the hypothesis of F having a low condition number, the squared norm of $k(FE)^\dagger$ should be close to $k^2r$.
+We are therefore minimising a quantity close to $$ VarX (r (1-k)^2 + dx - r) + VarN k^2 r$$ Zeroing the derivative in k yields an approximate value of k, as $$k= \frac {VarX}{VarX + VarN}=\frac{1}{1+nsr}$$ nsr the noise to signal ratio, VarN divided by VarX.
+
+FC: something needed here about F and X'X not full rank or badly conditioned
 
@@
 
-For the first step, we have 
-\begin{equation}
-\begin{aligned}
-\min_G \left \| X-GY \right \|^2 &= \min_G \left \| X - GF(EX+N)\right\|^2 \\
-&= \min_G \left \| (I-GFE)X\right\| ^2 + \left \| GFN\right \| ^2\\
-&= \min_G \sum_{i} \left \| (I-GFE)X_{i}\right\| ^2 + \left \| GFN_{i}\right \| ^2\\
-&{}= \min_G \alpha \left \| I-GFE\right\| ^2  + \beta \left \| G\right \| ^2\\
-\end{aligned}
-\end{equation}
-with $\alpha$ and $\beta$ positive constants, the latter equal to the norm of FN.
-We are balancing two minimisation, with $\alpha$ close to zero, the solution is the null matrix, with $\beta$ close to zero, it is the pseudo inverse of $FE$.
-%
-FC: lost here .
-I'd like to prove that we recover a scales version $(FE)^\dagger D$ with D a diagonal scaling over dimensions of X, according to signal noise ratio, covariance of X and covariance of N.
-This is what we sort of get from
-
-SVD of X, USV, so the minimum becomes
-
-In the general case, $\beta=\left \| FN\right \| ^2$, and the solution is $k (FE)^\dagger$, with $k= {\alpha\over{\alpha+\beta}}$.
-
-Thus, in the linear case, back and forth regression retrieves $k(FE)^{\dagger}(FE)$.
-In many cases, if v s the variance of $FN$, k can be approximated as $k= {1\over{1+v}}$, it quantifies the amount of noise added to the causal process, and the difficulty of recovering the causal features.
+Thus, in the linear case, back and forth regression retrieves $\hat H = \frac{1}{1+nsr}(FE)^{\dagger}(FE)$.
 
@@
 Since it is diagonal with eigenvalues 0 and k, if k is large enough, it should allow us to retrieve causal features in a fairly robust way.
+
 \subsection{Asymptotic behaviour - non linear case}

### 2019-05-21 code/synthetic_experiment/main.py
Fixes

@@ -56,6 +56,14 @@ def compute_false_negatives(e_pred, e_true):
     return np.mean(((e_pred == 0) * (e_true == 1) * 1.0))
 
 
+def response_at_active(e_pred, e_true):
+    return np.mean(e_pred[np.where(e_true == 1)[0]])
+
+
+def response_at_inactive(e_pred, e_true):
+    return np.mean(e_pred[np.where(e_true == 0)[0]])
+
+
 def run_experiment(args):
     for seed in range(args.n_seeds):
         random.seed(seed)
@@ -95,12 +103,13 @@ def run_experiment(args):
 
             # this is the binary mask E estimated by the model
             mask = model.solution()
+            binary_mask = sonquist_morgan(mask)
 
-            false_positives = compute_false_positives(mask, mask_true)
-            false_negatives = compute_false_negatives(mask, mask_true)
+            false_positives = compute_false_positives(binary_mask, mask_true)
+            false_negatives = compute_false_negatives(binary_mask, mask_true)
 
             # fit a basic model on the selected causes
-            sel = np.nonzero(sonquist_morgan(mask))[0]
+            sel = np.nonzero(binary_mask)[0]
 
             model = Ridge()
             model.fit(x_tr[:, sel], y_tr)
@@ -117,6 +126,8 @@ def run_experiment(args):
             result["result_error_out_mask"] = error_out_mask
             result["result_false_positives"] = false_positives
             result["result_false_negatives"] = false_negatives
+            result["result_response_active"] = response_at_active(mask, mask_true) 
+            result["result_response_inactive"] = response_at_inactive(mask, mask_true) 
             result["result_auc"] = roc_auc_score(mask_true, mask) 
             results.append(result)
 

### 2019-05-21 code/synthetic_experiment/models.py

@@ -12,7 +12,7 @@ N_JOBS = 1
 
 def basic_regression(X, Y, regularize=True):
     if regularize:
-        alphas = np.logspace(-5, 5, 20)
+        alphas = np.logspace(-5, 5, 10)
         w = RidgeCV(fit_intercept=False, alphas=alphas).fit(X, Y).coef_
     else:
         w = LinearRegression(fit_intercept=False).fit(X, Y).coef_
@@ -75,7 +75,7 @@ class OLS(SKModel):
 class Ridge(SKModel):
     def fit(self, X, Y):
         self.coef = RidgeCV(fit_intercept=False,
-                            alphas=np.logspace(-5, 5, 20)).fit(X, Y).coef_.T
+                            alphas=np.logspace(-5, 5, 10)).fit(X, Y).coef_.T
         return self
 
 
@@ -129,7 +129,7 @@ class RRR(SKModel):
     def fit(self, X, Y):
         grid = {
             "n_components": np.linspace(1, X.shape[1], 10).astype(int),
-            "reg": np.logspace(-5, 5, 20)
+            "reg": np.logspace(-5, 5, 10)
         }
 
         rrr = GridSearchCV(ReducedRankRegression(), grid, n_jobs=N_JOBS, cv=5)
@@ -174,7 +174,7 @@ class CCA(SKModel):
         m = min(X.shape[1], Y.shape[1])
         grid = {
             "n_components": np.linspace(1, m, 10).astype(int),
-            "reg": np.logspace(-5, 5, 20)
+            "reg": np.logspace(-5, 5, 10)
         }
 
         cca = GridSearchCV(CanonicalCorrelation(), grid, n_jobs=N_JOBS, cv=5)

### 2019-05-21 code/synthetic_experiment/models.py
new models

@@ -9,6 +9,170 @@ import numpy as np
 
 N_JOBS = 1
 
+'''
+parameters for JRR_testing
+
+
+G_solvers = (
+    ('ols', RidgeCV(1e-6, False)),
+    ('ridgecv', RidgeCV(alphas, independent=True)),
+    ('rankcv', TrunkOLSCV(ranks=False, independent=True)),
+    ('rankmle', trunkated_ols)
+)
+
+H_solvers = (
+    ('ols', RidgeCV(1e-6, False)),
+    ('ridgecv', RidgeCV(alphas, independent=False)),
+    ('rankcv', TrunkOLSCV(ranks=False, independent=False)),
+    ('rankmle', trunkated_ols)
+)
+
+
+params = (
+    G_solvers,  # G
+    H_solvers,  # H
+    (('loo', 0), ('bag', 10)),  # bagging or leave one out
+)
+
+functions = dict()
+for (label_g, G), (label_h, H), (label_bag, bagging) in product(*params):
+    label = '_'.join((label_g, label_h, label_bag))
+    functions[label] = JRR(G=G, H=H, bagging=bagging)
+
+
+
+results = dict()
+for label, func in functions.items():
+    print(label)
+    E_hat = func(X, Y)
+    results[label] = E_hat
+'''
+
+
+
+def ridge_cv(X, Y, alphas, independent=False):
+    """
+    Similar to sklearn RidgeCV but
+    (1) can optimize a different alpha for each column of Y
+    (2) return leave-one-out Y_hat
+    """
+    if isinstance(alphas, (float, int)):
+        alphas = np.array([alphas, ], np.float64)
+    alphas = np.asarray(alphas)
+    n, n_x = X.shape
+    n, n_y = Y.shape
+    # Decompose X
+    U, s, _ = svd(X, full_matrices=0)
+    v = s**2
+    UY = U.T @ Y
+
+    # For each alpha, solve leave-one-out error coefs
+    cv_duals = np.zeros((len(alphas), n, n_y))
+    cv_errors = np.zeros((len(alphas), n, n_y))
+    for alpha_idx, alpha in enumerate(alphas):
+        # Solve
+        w = ((v + alpha) ** -1) - alpha ** -1
+        c = U @ np.diag(w) @ UY + alpha ** -1 * Y
+        cv_duals[alpha_idx] = c
+
+        # compute diagonal of the matrix: dot(Q, dot(diag(v_prime), Q^T))
[…]

### 2019-05-21 paper/paper.tex
wip: intro method

@@
 method that optimally identifies the most plausible causal factors from a set
-of covarying candidates. BFR consists of i) predicting the potential causes $X$
+of covarying candidates. BFR consists in i) predicting the potential causes $X$
 from the observable $Y$, and subsequently ii) identifying the only factors that
@@
 partial least squares) on two tasks: causal identification and out-of-distribution
-prediction. We apply BFR to several empirical datasets to show that it
-retrieves plausible results. In particular, we apply BFR to neuroimaging
-recordings of 102 subjects presented with word sequences, and show, as
-expected, that the early brain responses appear to be specifically caused by
-low-level visual factors whereas late brain responses appear to be caused by
-lexico-semantic factors, despite the fact that these factors covary. We finish
-by extending BFR to nonlinear cause-effect relationships.
+prediction. We apply BFR to neuroimaging recordings of 102 subjects reading word
+sequences. The results show, as expected, that the early brain responses to
+words appear to be specifically caused by low-level visual factors whereas late
+brain responses appear to be caused by lexical factors, despite the fact that
+these factors covary in the study. We finish by extending BFR to nonlinear cause-effect relationships.
 \end{abstract}
@@
 
-Causal discovery, telling causes from their effects, is an inverse problem.
+Natural sciences are tasked to find, from a set of potentially-covarying factors, the minimal subset that suffices to predict a set of potentially multidimensional observations.
 %
-Given experimental data (effects) and possible explanations (causes), we are tasked to find the causes that account for the observations, knowing that other non causal factors (contingent, random) are at work.
+This problem can be formalized as follow: we measure, via an apparatus F, a multidimensional signal Y, that is a function of possible causes X, and other unknown factors that can be understood as noise N: $Y=f(X,N)$.
+%
+Of all possible causes (features of X), only a subset may account for Y, let them be represented as EX, with E a binary diagonal matrix: a diagonal of zeroes and ones, which correspond to active causal features of X.
+%
+Under the assumption of centered and additive noise N, causal discovery thus consists in finding E in $Y=f(EX+N)$, with f and N being unknown.
 
-More precisely, we measure, via an apparatus F, a multidimensional signal Y, that is a function of possible causes X, and other, unknown, factors that can be understood as noise N: $Y=f(X,N)$.
 %
-Of all possible causes (features of X), only a few account for Y, let them be represented as EX, with E a binary diagonal matrix: a diagonal of zeroes and ones, which correspond to active causal features of X.
+Causal discovery is impeded by two challenges.
 %
-Let us assume that noise N is centred (adding, if need be, an intercept to the active causal features, to capture the bias in N), and additive (not a very strong assumption in the absence of further constraints on N).
+First, causal and non-causal factors may naturally co-vary.
+[For example, XXX.] (e.g.
+X'X is not an orthogonal matrix).
+When the number of co-varying factors is low, this issue is generally solved with factorial designs [cite the design of experiments, fisher 1935].
+[XXX medical example].
+As factorial design become hard to implement with numerous factors, the orthogonalization of co-varying factors is performed analytically, by estimating how each factor contributes to the variance of the observations (e.g analysis of variance, ANOVA).
+In this paper, we refer to this approach as 'forward modeling' as it models how factors X predict the observations Y.
+As illustrated in Fig X., forward model can be sub-optimal when observations are multidimensional and noisy, because it typically does not make use of the structure of observations (e.g.
+the model does not consider Y'Y).
 %
-Then, we can write our model of causal discovery as $Y=f(m(EX)+N)$, with f and m unknown functions.
+Second, causal discovery can be challenged by the multiplicity and complexity of the observations.
+[For example XXX].
+This issue has led many to reverse the problem and to focus on optimally predicting factors from complex and poorly controlled observations.
+In neuroimaging, for example, it is increasingly popular to use "decoding" methods: instead of testing whether the brain responds to particular factors, mathematical models are trained to maximally predict such factors from complex brain signals.
+We refer to this approach as 'backward modeling' as it models how the observations Y can be used to predict the factors X.
+However, backward can predict a non-causal factor as long as it correlates with a causal one.
+
 %
-Up to some rearrangement of the features of X, we can also suppose m linear.
-
-We end up with the following model $Y=f(MEX+N)$, where N is centred noise E a binary diagonal matrix, M a full rank matrix and f some unknown transformation.
+In sum, forward modeling approaches are well suited in disentangling the independent contribution of covarying factors, but can be suboptimal at combining multidimensional observations.
+By contrast, backward modeling approaches are well suited at combining multiple observations, but do not disentangle covarying factors [Figure X].
 %
-Given experimental measures of X and Y (as matrices), and our goal is to recover E, the active causal features of the process under study.
-
-*** description of the neuroscience problem
-
-Such problems have been attacked along two different lines.
-%
-In the direct approach, we use the above equation as a statistical model, try to express each feature of Y as a function of X, test whether the coefficients associated with specific features of X are different from zero, and recover E as the features of X active over some dimension of Y.
-%
-In the inverse approach, we try to express each feature of X as a function of Y, estimate its changes as Y varies, and recover E as the set of "non random" features of X.
-%
-In the presence of noise and a large number of features of X, Y and N, both approaches are limited because they fail to take advantage of the correlations between features of the variable they are modelling (or guard against their multicollinearity).
-%
-In the direct approach, features of Y are modelled separately.
-%
-In the inverse approaches, correlations between features of X are not accounted for.
-
-In this paper, we introduce back and forth regression, which combines direct and inverse approaches.
-%
-We first predict X from Y, as in the inverse approach, and retrieve a robust least square estimate $\hat X$. Then, we calculate a robust least square estimator of $\hat X$ from X.
+In this paper, we introduce the 'back-and-forth regression', which combines the benefit of forward and backward approaches.
+Back-and-forth regression consists in first predicting X from Y (as in the backward modeling) subsequently retrieving a robust least square estimate $\hat X$ and finally computing a least square estimator of $\hat X$ from X (as in forward modeling).
 %
@@
 %
-The main purpose of back and forth regression is feature extraction, but it can be turned into a regression method by filtering X on the selected features.
+The main purpose of back-and-forth regression is feature extraction, but it can be turned into a regression method by filtering X on the selected features.
 
-In the next sections, we describe back and forth regression, prove that it asymptotically extracts E, both in the linear and non linear case, and that it can be understood as a special case of canonical component analysis (cca).
+In the next sections, we (i) describe back-and-forth regression, (ii) prove that it asymptotically extracts E, (iii) both in the linear and non linear case, and that it can be understood as a special case of canonical component analysis (cca).
 %
-By experimenting with simulated data, we demonstrate its robustness and use cases, and show that it compares favourably with state of the art techniques for regression and feature extraction.
+By experimenting with simulated data, we demonstrate the robustness and typical use cases of back-and-forth regression, and show that it compares favourably with state of the art techniques for regression and feature extraction, both in terms of causal discovery and out-of-
 %
-Finally, we illustrate its use on an example from neuroscience.
+Finally, we illustrate its use on an example from neuroimaging.
 
@@
 
-\subsection{The method}
+\subsection{Algorithm}
 
-We are given two set of measurements over a sample of N individuals.
+We are given two set of measurements over a sample of n individuals.
+%
+% Francois can you use the Y in Real^{n, ny} format
 %
@@
 
-Under our hypotheses (centred noise, E a binary diagonal) and even under heavy noise, $\hat H$ is close to diagonal.
+Under our hypotheses (centered noise, E a binary diagonal) and even under heavy noise, $\hat H$ is close to diagonal.
 %
@@
 %
-This will not work here, as back and forth regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
+This will not work here, as back-and-forth regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
 %
@@
 
-This can be done by maximising the ratio of inter-group variance over total variance (ie minimising intra-group variance).
+This can be done by maximizing the ratio of inter-group variance over total variance (ie minimizing intra-group variance).
 %
@@
 %
-To use back and forth regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
+To use back-and-forth regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
 
-Summarizing, back and forth regression is performed as follows:
+Summarizing, back-and-forth regression is performed as follows:
 \begin{enumerate}
@@
 
-The two steps of back and forth regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).
+The two steps of back-and-forth regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).
 
@@
 
-Thus, in the linear case, back and forth regression retrieves $\hat H = \frac{1}{1+nsr}(FE)^{\dagger}(FE)$.
+Thus, in the linear case, back-and-forth regression retrieves $\hat H = \frac{1}{1+nsr}(FE)^{\dagger}(FE)$.
 
@@
 \section{Related works}
-\subsection{back and forth regression as a special case of canonical component analysis}
-Formula (1) shows that back and forth regression can be seen as a special case of canonical component analysis (CCA).
+\subsection{back-and-forth regression as a special case of canonical component analysis?}
+Formula (1) shows that back-and-forth regression may be related to canonical component analysis (CCA).
 %
-In CCA, we are given two matrixes X and Y (or size N,p and N,q) that describe different features measured on the same data sample, and characterise their proximity (that is, between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.
+In CCA, we are given two matrices X and Y (or size N,p and N,q) that describe different features measured on the same data sample, and characterize their proximity (that is, between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.
 
-This is done by finding vectors $a$ and $b$ such that $Xa$ and $Yb$ display maximal correlation, or by maximising $a'X'Yb$ over unit-normed $a$ and $b$.
+This is done by finding vectors $a$ and $b$ such that $Xa$ and $Yb$ display maximal correlation, or by maximizing $a'X'Yb$ over unit-normed $a$ and $b$.
 %
@@
 %
-Thus, back and forth regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
+Thus, back-and-forth regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
 %
-The closer we are to this hypothesis, the closer the results of CCA and back and forth regression will be.
+The closer we are to this hypothesis, the closer the results of CCA and back-and-forth regression will be.
 
-There are differences between CCA and back and forth regression.
+There are differences between CCA and back-and-forth regression.
 %
@@
 \section{Experiments - simulated data}
-Experiments on simulated data serve three purposes: compare back and forth regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.
+Experiments on simulated data serve three purposes: compare back-and-forth regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.
 
@@
 
-We evaluate the performance of back and forth regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
+We evaluate the performance of back-and-forth regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
 %
@@
 \subsection{Feature extraction - linear case}
-All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by back and forth regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
+All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by back-and-forth regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
 %
@@
 
-For feature extraction, back and forth regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
+For feature extraction, back-and-forth regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
 %
@@
 %
-back and forth regression allows for good feature extraction if :
+back-and-forth regression allows for good feature extraction if :
 
@@
 %
-This demonstrates that back and forth regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.
+This demonstrates that back-and-forth regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.
 
@@
 
-The rationale behind such tests is that whereas back and forth regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.
+The rationale behind such tests is that whereas back-and-forth regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.
 
@@
 
+\input{meg_exps.tex}
+
 \section{Conclusion}
+
+\section{Acknowlegements}
+Data were provided (in part) by the Donders Institute for Brain, Cognition and Behaviour.
 

### 2019-05-21 paper/paper.tex
wip: intro method

@@
 %
-To find a, one calculates the eigenvectors of (1).
+To find a, one calculates the eigenvectors of Eq. (1).
 %
@@
 
-If (1) is diagonal, for instance in the model presented above, the features of X are eigenvectors.
+If the result of Eq. (1) is diagonal, as in the model presented above, the features of X are eigenvectors.
 %
-If all eigenvalues are either 0 or 1, since they correspond to the squared cosine of the angle between the corresponding feature and the subspace spanned by Y, all the features are either part of the Y subspace (eigenvalues of one) or orthogonal to it (for zero eigenvalues).
+If all eigenvalues are either 0 or 1, then all the features are either part of the Y subspace (eigenvalues of one) or orthogonal to it (for zero eigenvalues) because they correspond to the squared cosine of the angle between the corresponding feature and the subspace spanned by Y, .
 %

### 2019-05-21 code/synthetic_experiment/models.py
new batch of models

@@ -342,14 +342,10 @@ def all_jrrs():
     alphas = np.logspace(-5, 5, 20)
 
     G_solvers = (
-        ('ols', MyRidgeCV(1e-6, False)),
         ('ridgecv', MyRidgeCV(alphas, independent=True)),
-        ('rankcv', TrunkOLSCV(ranks=False, independent=True)),
-        ('rankmle', trunkated_ols)
     )
 
     H_solvers = (
-        ('ols', MyRidgeCV(1e-6, False)),
         ('ridgecv', MyRidgeCV(alphas, independent=False)),
         ('rankcv', TrunkOLSCV(ranks=False, independent=False)),
         ('rankmle', trunkated_ols)
@@ -359,7 +355,7 @@ def all_jrrs():
     params = (
         G_solvers,  # G
         H_solvers,  # H
-        (('loo', 0), ('bag', 10)),  # bagging or leave one out
+        (('bag', 10)),  # bagging or leave one out
     )
 
     functions = dict()

### 2019-05-21 code/synthetic_experiment/models.py
new sweep params

@@ -355,7 +355,7 @@ def all_jrrs():
     params = (
         G_solvers,  # G
         H_solvers,  # H
-        (('bag', 10)),  # bagging or leave one out
+        (('bag', 100)),  # bagging or leave one out
     )
 
     functions = dict()

### 2019-05-21 paper/meg_exps.tex
figures experiments

@@
 
-Next, we apply our method to brain imaging data, on the ``Mother Of Unification
+Next, we apply our method to brain imaging data, on the ``Mother Of all Unification
 Studies'' (MOUS) dataset \cite{schoffelen2019204}. The dataset consists of
@@
 
-Subjects were presented a set of 120 sentences in Dutch, and a scrambled lists
+Subjects were presented a set of 120 sentences in Dutch, and scrambled lists
 of the same words for a controlled experiment. Each word was presented on a
@@
 
-As a result, we obtain, for each of the 301 sensors and each word presented, a
-time series of 67 real-values spanning $\left[-100ms, 1500ms\right]$ relative to
+As a result, we obtain, for each of the 301 channels (273 magnetometers on the
+subject and 28 compensation channels measuring the room) and each word
+presented, a time series of 67 real-values spanning $\left[-100ms, 1000ms\right]$ relative to
 the stimulus onset. Each of the 102 subjects (we used data from 93 subjects) was
@@
 
+We show, in Figure \ref{fig:megavg}, the average response across words of the 
+magnetometers for the first subject. Time $t=0$ corresponds to the stimulus onset.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth, trim=1.5cm 1cm 0.5cm 0cm, clip=True]{figure_meg.pdf}
+  \caption{Average magnetosensor response across words for one subject.}
+  \label{fig:megavg}
+\end{figure}
+
 \subsection{Application of the method}
+
+Using this dataset, we study which features of input words can elicit a strong
+MEG response in the brain.  We add to each word (i) visual features: word length
+(1 dim.), letter count (26 dim.), and (ii) lexical features : word frequency (1
+dim.) part-of-speech tag (one of ADJ, ADP, ADV, CONJ, DET, NOUN, NUM, PRON,
+PROPN, VERB; 10 dim.). Sentence-level word features are left for future study.
+
+This procedure yields an $\left(n \times 38\right)$-dimensional input matrix $X$,
+where $n$ is the number of words for the subject. Each feature in $X$ is centered and
+scaled for zero mean and unit variance.
+
+Then, we apply the BFR method independently on each of the 67 time slices of the
+MEG recordings. Each slice $Y_t$ is a $n \times 301$ matrix, also centered and
+scaled. We obtain a vector of 38 coefficients per slice.
+
+We show these coefficients as a function of time in the stacked plots, in Figure
+\ref{fig:megresult}.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth, trim=1.5cm 1cm 0.5cm 0cm,
+clip=True]{meg_result.pdf}
+  \caption{Average magnetosensor response across words for one subject.}
+  \label{fig:megresult}
+\end{figure}
+
+\subsection{Discussion of the results.}
+
+\subsubsection{Overview}

### 2019-05-21 code/synthetic_experiment/plot.py
Cleaning

@@ -16,14 +16,14 @@ stats = {
     "result_auc": "AUC"
 }
 
-plt.rc('text', usetex=True)
-plt.rc('text.latex', preamble=r'\usepackage{times}')
-plt.rc('font', family='serif')
+# plt.rc('text', usetex=True)
+# plt.rc('text.latex', preamble=r'\usepackage{times}')
+# plt.rc('font', family='serif')
 plt.rc('font', size=10)
 
 if __name__ == "__main__":
     parser = argparse.ArgumentParser(description='JRR synthetic experiment')
-    parser.add_argument('--file', type=str, default='results_05.txt')
+    parser.add_argument('--file', type=str, default='results_all_jrr.txt')
     args = parser.parse_args()
 
     with open(args.file, "r") as f:
@@ -92,27 +92,36 @@ if __name__ == "__main__":
 
     for m, model in enumerate(models):
         for r, result in enumerate(result_names):
+            result_mean = 0
+            result_variance = 0
+            result_counter = 0
             plt.subplot(plot_rows, plot_columns, plot_counter)
             plot_counter += 1
 
             if r == 0:
-                plt.ylabel(model.upper())
+                plt.ylabel(model.upper()[8:], fontsize=8)
             if m == 0:
                 plt.title(stats[result], fontsize=10)
 
             for eid in results:
-                if ("JRR" in results[eid]) and (model in results[eid]):
-                    competitor_mean = results[eid][model]["mean"][r]
-                    competitor_variance = results[eid][model]["variance"][r]
+                if model in results[eid]:
+                    result_mean += results[eid][model]["mean"][r]
+                    result_variance += results[eid][model]["variance"][r]
+                    result_counter += 1
+                    if "JRR" in results[eid]:
+                        competitor_mean = results[eid][model]["mean"][r]
+                        competitor_variance = results[eid][model]["variance"][r]
 
-                    jrr_mean = results[eid]["JRR"]["mean"][r]
-                    jrr_variance = results[eid]["JRR"]["variance"][r]
+                        jrr_mean = results[eid]["JRR"]["mean"][r]
+                        jrr_variance = results[eid]["JRR"]["variance"][r]
 
-                    # plt.errorbar(jrr_mean,
-                    #              competitor_mean,
-                    #              xerr=jrr_variance,
-                    #              yerr=competitor_variance)
-                    plt.scatter(jrr_mean, competitor_mean, c="black", alpha=0.5)
+                        # plt.errorbar(jrr_mean,
[…]

### 2019-05-21 code/synthetic_experiment/main.py
Ready to launch synthetic experiment

@@ -1,28 +1,25 @@
 # TODO: model Y = (XE + N)F + N' with second noise N'
 # TODO: uniform distributions
 
-from models import JRR, OLS, Ridge, Oracle, PLS, Lasso, RRR, CCA, all_jrrs
+from models import JRR, OLS, Ridge, Oracle, PLS, Lasso, RRR, CCA, JRR2
 from sklearn.metrics import roc_auc_score
 from data import Synthetic
-
 import numpy as np
 import argparse
 import random
 
 models = {
+    "JRR": JRR,
+    "JRR2": JRR2,
     "OLS": OLS,
     "Ridge": Ridge,
     "CCA": CCA,
     "Oracle": Oracle,
-    "JRR": JRR,
     "PLS": PLS,
     "RRR": RRR,
     "Lasso": Lasso
 }
 
-# TODO: deactivate after testing
-if True:
-    models = all_jrrs()
 
 def sonquist_morgan(x):
     z = np.sort(x)
@@ -94,7 +91,7 @@ def run_experiment(args):
 
         for m, Model in models.items():
             if "JRRTEST" in m:
-                model = Model 
+                model = Model
             elif m == "Oracle":
                 model = Model(mask_true)
             else:
@@ -131,9 +128,11 @@ def run_experiment(args):
             result["result_error_out_mask"] = error_out_mask
             result["result_false_positives"] = false_positives
             result["result_false_negatives"] = false_negatives
-            result["result_response_active"] = response_at_active(mask, mask_true) 
-            result["result_response_inactive"] = response_at_inactive(mask, mask_true) 
-            result["result_auc"] = roc_auc_score(mask_true, mask) 
+            result["result_response_active"] = response_at_active(
+                mask, mask_true)
+            result["result_response_inactive"] = response_at_inactive(
+                mask, mask_true)
+            result["result_auc"] = roc_auc_score(mask_true, mask)
             results.append(result)
 
             print(results[-1])

### 2019-05-21 code/synthetic_experiment/models.py

@@ -69,76 +69,9 @@ class MyRidgeCV(object):
         return ridge_cv(X, Y, self.alphas, self.independent)
 
 
-def trunkated_ols_cv(X, Y, ranks=True, independent=False):
-    """
-    solves (X'X)^-1 X'Y with rank optimal constrain on X
-    (1) can optimize a different alpha for each column of Y (independent=True)
-    (2) return leave-one-out Y_hat
-    """
-    independent = False
-    n, n_x = X.shape
-    n, n_y = Y.shape
-    # Decompose X
-    U, s, _ = np.linalg.svd(X, full_matrices=0)  # U = shape(n, n_x)
-    v = s**2
-
-    cv_duals = np.zeros((n_x, n, n_y))
-    cv_errors = np.inf * np.ones((n_x, n, n_y))
-
-    ranks = np.arange(n_x) if ranks is True else np.linspace(1, n_x-1, 30).astype(int)
-
-    for rank in ranks:
-        # Solve
-        # ridge cv: w = ((v + alpha) ** -1) - alpha ** -1
-        D = np.eye(n) - U[:, :rank] @ U[:, :rank].T  # n_samples x n_samples
-        c = D @ Y  # n_samples x dim_y
-        # ridge cv: c = U @ np.diag(w) @ UY + alpha ** -1 * Y
-
-        # compute diagonal of the matrix: dot(Q, dot(diag(v_prime), Q^T))
-        # G_diag = (w * U ** 2).sum(axis=-1) + alpha ** -1
-        G_diag = np.diag(D)
-        error = c / G_diag[:, np.newaxis]
-        cv_errors[rank] = error
-        cv_duals[rank] = U[:, :rank] @ np.diag(v[:rank]**-1) @ U[:, :rank].T @ Y # noqa
-
-    # identify best alpha for each column of Y independently
-    if independent:
-        best_ranks = (cv_errors ** 2).mean(axis=1).argmin(axis=0)
-        duals = np.transpose([cv_duals[b, :, i]
-                              for i, b in enumerate(best_ranks)])
-        cv_errors = np.transpose([cv_errors[b, :, i]
-                                  for i, b in enumerate(best_ranks)])
-    else:
-        _cv_errors = cv_errors.reshape(n_x, -1)
-        best_ranks = (_cv_errors ** 2).mean(axis=1).argmin(axis=0)
-        duals = cv_duals[best_ranks]
[…]

### 2019-05-21 paper/paper.tex
paper intro

@@
   Jean-Remi King\\
-  Facebook AI\\
+  CNRS\\
   \texttt{email} \\
@@
 To address this issue, we introduce ``back-and-forth regression'' (BFR), a
-method that optimally identifies the most plausible causal factors from a set
-of covarying candidates. BFR consists in i) predicting the potential causes $X$
-from the observable $Y$, and subsequently ii) identifying the only factors that
-contributed to those predictions. After detailing the proof of convergence, we
-show that BFR outperforms least-squares and related regression techniques, as
-well as cross-decomposition methods (e.g. canonical correlation analysis, and
-partial least squares) on two tasks: causal identification and out-of-distribution
-prediction. We apply BFR to neuroimaging recordings of 102 subjects reading word
-sequences. The results show, as expected, that the early brain responses to
-words appear to be specifically caused by low-level visual factors whereas late
-brain responses appear to be caused by lexical factors, despite the fact that
-these factors covary in the study. We finish by extending BFR to nonlinear cause-effect relationships.
+method designed to identify, from a set of co-varying factors, the factors
+that most plausibley account for multidimensional observations. After detailing
+the proof of convergence, we show that BFR outperforms least-squares and related
+regression techniques, as well as cross-decomposition methods (e.g. canonical
+correlation analysis, and partial least squares) on two tasks: causal
+identification and out-of-distribution prediction. Finally, we apply BFR to
+neuroimaging recordings of 102 subjects reading word sequences. The results
+show, as expected, that the early brain responses to words appear to be
+specifically caused by low-level visual factors whereas late brain responses
+appear to be caused by lexical factors, despite the fact that these factors
+covary in the study. XXX conlude / open
 \end{abstract}
@@
 
-Natural sciences are tasked to find, from a set of potentially-covarying factors, the minimal subset that suffices to predict a set of potentially multidimensional observations.
+Natural sciences are tasked to find, from a set of hypothetical factors, the minimal subset that suffices to reliably predict observations, especially in novel environments.
 %
-This problem can be formalized as follow: we measure, via an apparatus F, a multidimensional signal Y, that is a function of possible causes X, and other unknown factors that can be understood as noise N: $Y=f(X,N)$.
+This problem can be formalized as follow: we measure, via an apparatus F, a signal Y, that is a function of possible causes X, and other unknown causes that can be understood as noise N: $Y=F(X,N)$.
+Of all considered causes (features of X), only a subset account for Y, let them be represented as EX, with E a binary diagonal matrix: a diagonal of zeros and ones, which correspond to causal features of X.
 %
-Of all possible causes (features of X), only a subset may account for Y, let them be represented as EX, with E a binary diagonal matrix: a diagonal of zeroes and ones, which correspond to active causal features of X.
-%
-Under the assumption of centered and additive noise N, causal discovery thus consists in finding E in $Y=f(EX+N)$, with f and N being unknown.
+Under the assumption of centered and additive noise N, causal discovery thus consists in finding E in $Y=F(EX+N)$, with F and N being unknown.
 
 %
-Causal discovery is impeded by two challenges.
+Under this formalism, causal discovery is impeded by two challenges.
 %
-First, causal and non-causal factors may naturally co-vary.
-[For example, XXX.] (e.g.
-X'X is not an orthogonal matrix).
-When the number of co-varying factors is low, this issue is generally solved with factorial designs [cite the design of experiments, fisher 1935].
-[XXX medical example].
-As factorial design become hard to implement with numerous factors, the orthogonalization of co-varying factors is performed analytically, by estimating how each factor contributes to the variance of the observations (e.g analysis of variance, ANOVA).
-In this paper, we refer to this approach as 'forward modeling' as it models how factors X predict the observations Y.
-As illustrated in Fig X., forward model can be sub-optimal when observations are multidimensional and noisy, because it typically does not make use of the structure of observations (e.g.
-the model does not consider Y'Y).
+To illustrate these two challenges, let us consider the case of a neuroscientist who study the brain bases of reading.
+Our neuroscientist wants to determine whether the frequency of a word modulates brain activity (e.g.
+neurons may become more active when our retina is flashed with the word 'triskaidekaphobia' than when the word is 'car').
 %
-Second, causal discovery can be challenged by the multiplicity and complexity of the observations.
-[For example XXX].
-This issue has led many to reverse the problem and to focus on optimally predicting factors from complex and poorly controlled observations.
-In neuroimaging, for example, it is increasingly popular to use "decoding" methods: instead of testing whether the brain responds to particular factors, mathematical models are trained to maximally predict such factors from complex brain signals.
-We refer to this approach as 'backward modeling' as it models how the observations Y can be used to predict the factors X.
-However, backward can predict a non-causal factor as long as it correlates with a causal one.
-
+The first challenge to causal discovery is that causal and non-causal factors can co-vary.
+For example, the frequency of words in language correlates with their number of letters.
+When the number of co-varying factors is low, this issue can be solved with factorial designs \cite{fisher_1935}.
+Our neuroscientist may thus carefully select words matched for their numbers of letters, while letting their frequency vary.
+However, factorial design becomes increasingly hard to implement as the number of potential factor grows.
+As a fallback option, factors may be orthogonalized analytically, by estimating how each of them contributes to explaining the observations (e.g analysis of variance, ANOVA).
+In this paper, we refer to this approach as 'forward modeling' as it models how the factors X contribute to explain the observations Y.
 %
-In sum, forward modeling approaches are well suited in disentangling the independent contribution of covarying factors, but can be suboptimal at combining multidimensional observations.
-By contrast, backward modeling approaches are well suited at combining multiple observations, but do not disentangle covarying factors [Figure X].
+The second challenge to causal discovery is the multiplicity of the observations.
+For example, our neuroscientist may simultaneously collect noisy recordings of hundreds of thousands of neurons; it can thus be suboptimal to consider each neuron independently.
+More generally, with the rise of large and complex datasets, the complexity of observations has led many to invert the problem and to focus on optimally predicting putative factors from complex and poorly controlled observations.
+In neuroimaging, for example, it is increasingly popular to use "decoding" methods: instead of testing whether the brain responds to particular factors, a model is trained to maximally predict such factors from complex brain signals.
+We refer to this approach as 'backward modeling' as it models how the observations Y can be combined to predict the factors X.
+%
+As illustrated in Fig.
+1., both forward and backward modeling have competing benefits and drawbacks.
+Specifically, forward modeling efficiently disentangles the independent contribution of covarying factors, but does not combine multidimensional observations.
+By contrast, backward modeling approaches combine multiple observations, but do not disentangle covarying factors.
 %
@@
 %
-If the active features of X are represented as EX, with E a binary diagonal, this provides a good approximation of E, robust to noise and environment change.
+If the active features of X are represented as EX, with E a binary diagonal, back-and-forth regression provides a good approximation of E, robust to noise and environment change.
 %
-The main purpose of back-and-forth regression is feature extraction, but it can be turned into a regression method by filtering X on the selected features.
-
-In the next sections, we (i) describe back-and-forth regression, (ii) prove that it asymptotically extracts E, (iii) both in the linear and non linear case, and that it can be understood as a special case of canonical component analysis (cca).
+In the next sections, we describe back-and-forth regression, prove that it asymptotically extracts E, and show how it relates to and departs from other cross-decomposition and component analyses.
+We then show with synthetic experiments that back-and-forth regression outperforms state-of-the-art techniques for regression and feature extraction, both in terms of causal discovery and out-of-environment predictions.
+Finally, we apply back-and-forth regression to a 102 recordings of brain responses to words, and show that the first brain responses appear to be mainly caused by low-level visual features, such as the number of letters, while late brain responses appear to be mainly caused by high-level lexical features, such as word frequency.
 %
-By experimenting with simulated data, we demonstrate the robustness and typical use cases of back-and-forth regression, and show that it compares favourably with state of the art techniques for regression and feature extraction, both in terms of causal discovery and out-of-
-%
-Finally, we illustrate its use on an example from neuroimaging.
 

### 2019-05-21 paper/paper.tex
paper intro

@@
 This problem can be formalized as follow: we measure, via an apparatus F, a signal Y, that is a function of possible causes X, and other unknown causes that can be understood as noise N: $Y=F(X,N)$.
-Of all considered causes (features of X), only a subset account for Y, let them be represented as EX, with E a binary diagonal matrix: a diagonal of zeros and ones, which correspond to causal features of X.
+Of all considered causes (features of X), only a subset accounts for Y, let it be represented as EX, with E a binary diagonal matrix: a diagonal of zeros and ones, which selects causal features of X.
 %
@@
 %
+%
 The first challenge to causal discovery is that causal and non-causal factors can co-vary.
-For example, the frequency of words in language correlates with their number of letters.
-When the number of co-varying factors is low, this issue can be solved with factorial designs \cite{fisher_1935}.
-Our neuroscientist may thus carefully select words matched for their numbers of letters, while letting their frequency vary.
+For example, the frequency of words in language correlates with their length.
+Factor covariation can typically be addressed with factorial designs \cite{fisher_1935}.
+Our neuroscientist may thus carefully select words matched in terms of lengths, while letting their frequency vary.
 However, factorial design becomes increasingly hard to implement as the number of potential factor grows.
@@
 In this paper, we refer to this approach as 'forward modeling' as it models how the factors X contribute to explain the observations Y.
+%
 %
@@
 %
+%
 As illustrated in Fig.
@@
 By contrast, backward modeling approaches combine multiple observations, but do not disentangle covarying factors.
+%
 %
@@
 
-We are given two set of measurements over a sample of n individuals.
+We are given two sets of measurements over a sample of n individuals.
 %

### 2019-05-21 paper/paper.tex
intro

@@
 appear to be caused by lexical factors, despite the fact that these factors
-covary in the study. XXX conlude / open
+covary in the study.
+% conclude  open
 \end{abstract}
@@
 
-Natural sciences are tasked to find, from a set of hypothetical factors, the minimal subset that suffices to reliably predict observations, especially in novel environments.
+Natural sciences are tasked to find, from a set of hypothetical factors, the minimal subset that suffices to reliably predict novel observations potentially in novel environments.
 %
-This problem can be formalized as follow: we measure, via an apparatus F, a signal Y, that is a function of possible causes X, and other unknown causes that can be understood as noise N: $Y=F(X,N)$.
-Of all considered causes (features of X), only a subset accounts for Y, let it be represented as EX, with E a binary diagonal matrix: a diagonal of zeros and ones, which selects causal features of X.
+This problem can be formalized as follow: empirical studies consist in measuring, via an apparatus $F$, a signal $Y \in \mathbb{R}^{i, l}$, that is a function of possible causes $X \in \mathbb{R}^{i, j}$, and other unknown causes that can be understood as noise $N \in \mathbb{R}^{i, k}$: $Y=F(X,N)$.
+Of all considered causes (features of X), only a subset accounts for $Y$, let it be represented as $E \cdot X$, with $E \in \mathbb{R}^{j, j}$ a binary diagonal matrix, which selects causal features of $X$.
 %
-Under the assumption of centered and additive noise N, causal discovery thus consists in finding E in $Y=F(EX+N)$, with F and N being unknown.
+Under the restricted assumptions of centered and additive noise N and linear systems, causal discovery thus consists in finding $E$ in
+
+\begin{equation}
+    Y=(XEM+N)F
+\end{equation}
+with $M \in \mathbb{R}^{j, k}$ and $F \in \mathbb{R}^{j, k}$ being unknown.
+$M$ is a unknown projection matrix that ensures that the noise and the causal factors share a common space, and can thus be absorbed by $F$:
+\begin{equation}
+    Y=(XE+N^{*})F^{*}
+\end{equation}
+\par
 
@@
 %
-To illustrate these two challenges, let us consider the case of a neuroscientist who study the brain bases of reading.
+To illustrate these two challenges, let us consider the case of a neuroscientist who studies the brain bases of reading.
 Our neuroscientist wants to determine whether the frequency of a word modulates brain activity (e.g.
 neurons may become more active when our retina is flashed with the word 'triskaidekaphobia' than when the word is 'car').
+One issue is that word frequency tends to correlate with word length.\par
 %
 %
-The first challenge to causal discovery is that causal and non-causal factors can co-vary.
-For example, the frequency of words in language correlates with their length.
-Factor covariation can typically be addressed with factorial designs \cite{fisher_1935}.
-Our neuroscientist may thus carefully select words matched in terms of lengths, while letting their frequency vary.
-However, factorial design becomes increasingly hard to implement as the number of potential factor grows.
-As a fallback option, factors may be orthogonalized analytically, by estimating how each of them contributes to explaining the observations (e.g analysis of variance, ANOVA).
-In this paper, we refer to this approach as 'forward modeling' as it models how the factors X contribute to explain the observations Y.
+This is the first challenge to causal discovery: a set of putative factors, some causal, other non-causal, can be collinear.
+Collinearity in the putative factors $X$ is typically addressed with factorial designs \cite{fisher_1935}.
+For example, our neuroscientist may carefully select words matched in terms of length, while letting their frequency vary.
+However, factorial design becomes increasingly difficult to implement as the number of considered factor grows.
+As a fallback option, factors may be orthogonalized \textit{a posteriori}, by analytically estimating how each of them contributes to explaining the observations.
+Our neuroscientist will thus typically use ordinary least square (OLS) to model how brain activity can be specifically accounted for by word length and word frequency:
+\begin{equation}
+H = (X'X)^{-1} X'Y
+\end{equation}
+In this paper, we refer to this approach as 'forward modeling' as it models how the factors X contribute to explain the observations Y. \par
 %
@@
 The second challenge to causal discovery is the multiplicity of the observations.
-For example, our neuroscientist may simultaneously collect noisy recordings of hundreds of thousands of neurons; it can thus be suboptimal to consider each neuron independently.
+For example, our neuroscientist may simultaneously collect noisy recordings of hundreds of thousands of neurons.
+It can thus be suboptimal to consider each neuron independently, because they may be influenced by some unknown factors $N$.
 More generally, with the rise of large and complex datasets, the complexity of observations has led many to invert the problem and to focus on optimally predicting putative factors from complex and poorly controlled observations.
-In neuroimaging, for example, it is increasingly popular to use "decoding" methods: instead of testing whether the brain responds to particular factors, a model is trained to maximally predict such factors from complex brain signals.
-We refer to this approach as 'backward modeling' as it models how the observations Y can be combined to predict the factors X.
+In neuroimaging, for example, it is increasingly popular to use "decoding" methods: instead of testing whether the brain responds to particular factors, a model is trained to maximally predict factors from complex brain signals:
+\begin{equation}
+G = (Y'Y)^{-1} Y'X
+\end{equation}
+We refer to this approach as 'backward modeling' as it models how the observations Y can be combined to predict the factors X. \par
 %
@@
 1., both forward and backward modeling have competing benefits and drawbacks.
-Specifically, forward modeling efficiently disentangles the independent contribution of covarying factors, but does not combine multidimensional observations.
-By contrast, backward modeling approaches combine multiple observations, but do not disentangle covarying factors.
+Specifically, forward modeling efficiently disentangles the independent contribution of collinear factors, but does not combine multidimensional observations.
+By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors.
+\par
 %
 %
-In this paper, we introduce the 'back-and-forth regression', which combines the benefit of forward and backward approaches.
-Back-and-forth regression consists in first predicting X from Y (as in the backward modeling) subsequently retrieving a robust least square estimate $\hat X$ and finally computing a least square estimator of $\hat X$ from X (as in forward modeling).
+In this paper, we introduce the 'back-and-forth regression', which combines the benefits of forward and backward modeling.
+Back-and-forth regression consists in first predicting X from Y (as in the backward modeling) subsequently retrieving an estimate $\hat X$ and finally disentangling which of the considered factors X accounts for the variance in $\hat X$ (i.e.
+analogously to forward modeling).
+\par
 %
-If the active features of X are represented as EX, with E a binary diagonal, back-and-forth regression provides a good approximation of E, robust to noise and environment change.
-%
-In the next sections, we describe back-and-forth regression, prove that it asymptotically extracts E, and show how it relates to and departs from other cross-decomposition and component analyses.
+In the next sections, we detail back-and-forth regression, prove that it asymptotically extracts E, and show how it relates to and departs from other cross-decomposition and component analyses.
 We then show with synthetic experiments that back-and-forth regression outperforms state-of-the-art techniques for regression and feature extraction, both in terms of causal discovery and out-of-environment predictions.
-Finally, we apply back-and-forth regression to a 102 recordings of brain responses to words, and show that the first brain responses appear to be mainly caused by low-level visual features, such as the number of letters, while late brain responses appear to be mainly caused by high-level lexical features, such as word frequency.
+Finally, we apply back-and-forth regression to the brain activity of 102 subjects reading $\approx$ 2,000 words, and show that the first brain responses to these words appear to be mainly modulated by low-level visual features, such as the number of letters, while late brain responses appear to be mainly modulated by high-level lexical features, such as word frequency.
+\par
 %
@@
 \subsection{Algorithm}
-
-We are given two sets of measurements over a sample of n individuals.
-%
-% Francois can you use the Y in Real^{n, ny} format
-%
-Y are ny multidimensional effects, and X are nx multidimensional possible causes.
-%
-We assume that only some features of X influence Y, so that $Y=f(EX,N)$, E being a binary diagonal, N some unknown noise, and f an unknown function.
-%
-We try to recover E, the active causal features.
 

### 2019-05-22 paper/paper.tex
better theorem, still in the making

@@
 \subsection{Asymptotic behaviour - linear case}
-As discussed in the introduction, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(EX+N)$, with F an unknown function.
+As discussed in the introduction, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(EX+N)$, with f an unknown function.
 %
-If F is linear we can rewrite it as $Y = F(EX + N)$, with X an (n, dx) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
+If F is linear we can rewrite it as $Y = F(EX + N)$, with X an (dx, n) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
 %
-N a (dx,n) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dx) transformation corresponding to the measuring apparatus, and Y a (n,dy) matrix of measured effects.
+N a (dx,n) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dx) transformation corresponding to the measuring apparatus, and Y a (dy, n) matrix of measured effects.
 
@@
 \arg \min_G \left \| X-GY \right \|^2 &= \arg \min_G \left \| X - GF(EX+N)\right\|^2 \\
-&{}= \arg \min_G \left \| (I-GFE)X\right\| ^2  + \left \| GFN\right \| ^2\\
-&{}= \arg \min_G \Tr (X'(I-GFE)'(I-GFE)X) + \Tr (N'F'G'GFN)\\
-&{}= \arg \min_G \Tr ((I-GFE)'(I-GFE)XX') + \Tr (G'G(FN)(FN)')\\
+&{}= \arg \min_G \left \| X-GFEX\right\| ^2  + \left \| GFN\right \| ^2
 \end{aligned}
@@
 
-As noise N gets close to zero, and if F and X have full rank, then the solution tends to $(FE)^\dagger$, the pseudo inverse of FE.
+As N tends to zero, we are searching for the minimal norm solution of $\min_G \left \| X-GFEX\right\| ^2$, which is $X(FEX)^\dagger$ ($M^\dagger$ denoting the pseudo inverse of M).
+
+Let k be the number of active features of E, suppose for notational convenience that they are the first k features of X, and let $FEX$ be of rank k (which means F and X are of full rank on the space spanned by the active features of E).
+Let K be the (nx,k) diagonal matrix built by deleting the rightmost nx-k columns of E.
+Then, $K=(\frac{I_{k}}{0})$.
+we may write $E=KK'$, and $FEX=(FK)(K'X)$, with $FK$  and $K'X$ of full rank k.
+
+The rank factorisation property of the pseudo inverse (ref) says that, if $A=BC$, A (m,n), B(m,k) or rank k C(k,n) of rank k, then $A^\dagger=C^\dagger B^\dagger= C' (CC')^{-1}(B'B)^{-1}B'$.
+
+Replacing in the expression of the minimal norm solution yields 
+\begin{equation}
+\begin{aligned}
+X(FEX)^\dagger &=X((FK)(K'X))^\dagger \\
+&=X(K'X)'((K'X)(K'X)')^{-1}((FK)(FK)')^{-1}(FK)' \\
+&=(XX')K(K'XX'K)^{-1}(FKK'F')^{-1} K'F' 
+\end{aligned}
+\end{equation}
+
+Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{3} & \Sigma_{4}\end{array}\right)$, the top left block being (k,k), since $K=\left(\begin{array}{c}I \\\hline 0\end{array}\right)$, we have $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{3} \Sigma_{1}^{-1}\end{array}\right)=S$
+
+Thus we have $X(FEX)^\dagger =S (FKK'F')^{-1} K'F'$
+
+FC: this explains the negative diagonal elements, when X is negatively correlated. Need now to prove that this is S(FE)+
+
+FC: onward to the reduction, I assume that we get a scaled value of S(FE)+, this needs a better proof
 
@@
 We are therefore minimising a quantity close to $$ VarX (r (1-k)^2 + dx - r) + VarN k^2 r$$ Zeroing the derivative in k yields an approximate value of k, as $$k= \frac {VarX}{VarX + VarN}=\frac{1}{1+nsr}$$ nsr the noise to signal ratio, VarN divided by VarX.
-
-FC: something needed here about F and X'X not full rank or badly conditioned
 

### 2019-05-22 paper/paper.tex
cleanup

--- 2019-05-21 paper/paper.tex
@@
 
-\title{Discovering causal factors with back-and-forth regression}
+\title{Discovering causal factors with back-to-back regression}
 
@@
 manipulate individually and ii) observations are complex and multi-dimensional.
-To address this issue, we introduce ``back-and-forth regression'' (BFR), a
+To address this issue, we introduce ``back-to-back regression'' (BFR), a
 method designed to identify, from a set of co-varying factors, the factors
@@
 This problem can be formalized as follow: empirical studies consist in measuring, via an apparatus $F$, a signal $Y \in \mathbb{R}^{i, l}$, that is a function of possible causes $X \in \mathbb{R}^{i, j}$, and other unknown causes that can be understood as noise $N \in \mathbb{R}^{i, k}$: $Y=F(X,N)$.
-Of all considered causes (features of X), only a subset accounts for $Y$, let it be represented as $E \cdot X$, with $E \in \mathbb{R}^{j, j}$ a binary diagonal matrix, which selects causal features of $X$.
+Of all considered causes (features of $X$), only a subset accounts for $Y$, let it be represented as $E \cdot X$, with $E \in \mathbb{R}^{j, j}$ a binary diagonal matrix, which selects causal features of $X$.
 %
-Under the restricted assumptions of centered and additive noise N and linear systems, causal discovery thus consists in finding $E$ in
+Under the assumptions of centered and additive noise N, causal discovery thus consists in finding $E$ in
 
 \begin{equation}
-    Y=(XEM+N)F
+    Y=F(XE, N)
 \end{equation}
-with $M \in \mathbb{R}^{j, k}$ and $F \in \mathbb{R}^{j, k}$ being unknown.
-$M$ is a unknown projection matrix that ensures that the noise and the causal factors share a common space, and can thus be absorbed by $F$:
-\begin{equation}
-    Y=(XE+N^{*})F^{*}
-\end{equation}
-\par
 
@@
 %
-To illustrate these two challenges, let us consider the case of a neuroscientist who studies the brain bases of reading.
+% To illustrate these two challenges, let us consider the case of a neuroscientist who studies the brain bases of reading.
 Our neuroscientist wants to determine whether the frequency of a word modulates brain activity (e.g.
@@
 %
-%
-This is the first challenge to causal discovery: a set of putative factors, some causal, other non-causal, can be collinear.
-Collinearity in the putative factors $X$ is typically addressed with factorial designs \cite{fisher_1935}.
-For example, our neuroscientist may carefully select words matched in terms of length, while letting their frequency vary.
-However, factorial design becomes increasingly difficult to implement as the number of considered factor grows.
-As a fallback option, factors may be orthogonalized \textit{a posteriori}, by analytically estimating how each of them contributes to explaining the observations.
-Our neuroscientist will thus typically use ordinary least square (OLS) to model how brain activity can be specifically accounted for by word length and word frequency:
+% This is the first challenge to causal discovery
+The first challenge is factor collinearity: causal and non-causal factors may tend to covary with one another: i.e.
+$X'X$ is not diagonal.
+Collinearity in $X$ is typically addressed with factorial designs \cite{fisher_1935}.
+% For example, our neuroscientist may carefully select words matched in terms of length, while letting their frequency vary.
+However, factorial design becomes increasingly difficult to implement as the number of considered factors grows.
+As a fallback option, factors may be orthogonalized \textit{a posteriori}, by analytically estimating how each of them contributes to explaining the observations with ordinary least square (OLS):
+
+% Our neuroscientist will thus
 \begin{equation}
@@
 The second challenge to causal discovery is the multiplicity of the observations.
-For example, our neuroscientist may simultaneously collect noisy recordings of hundreds of thousands of neurons.
+% For example, our neuroscientist may simultaneously collect noisy recordings of hundreds of thousands of neurons.
 It can thus be suboptimal to consider each neuron independently, because they may be influenced by some unknown factors $N$.
-More generally, with the rise of large and complex datasets, the complexity of observations has led many to invert the problem and to focus on optimally predicting putative factors from complex and poorly controlled observations.
+With the rise of large datasets, the complexity of observations has led many to focus on optimally predicting putative factors from complex and poorly controlled observations.
 In neuroimaging, for example, it is increasingly popular to use "decoding" methods: instead of testing whether the brain responds to particular factors, a model is trained to maximally predict factors from complex brain signals:
@@
 1., both forward and backward modeling have competing benefits and drawbacks.
-Specifically, forward modeling efficiently disentangles the independent contribution of collinear factors, but does not combine multidimensional observations.
+Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations.
 By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors.
@@
 %
-In this paper, we introduce the 'back-and-forth regression', which combines the benefits of forward and backward modeling.
-Back-and-forth regression consists in first predicting X from Y (as in the backward modeling) subsequently retrieving an estimate $\hat X$ and finally disentangling which of the considered factors X accounts for the variance in $\hat X$ (i.e.
+In this paper, we introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling.
+Back-to-back regression consists in first predicting X from Y (as in the backward modeling) subsequently retrieving an estimate $\hat X$ and finally disentangling which of the considered factors X accounts for the variance in $\hat X$ (i.e.
 analogously to forward modeling).
@@
 %
-In the next sections, we detail back-and-forth regression, prove that it asymptotically extracts E, and show how it relates to and departs from other cross-decomposition and component analyses.
-We then show with synthetic experiments that back-and-forth regression outperforms state-of-the-art techniques for regression and feature extraction, both in terms of causal discovery and out-of-environment predictions.
-Finally, we apply back-and-forth regression to the brain activity of 102 subjects reading $\approx$ 2,000 words, and show that the first brain responses to these words appear to be mainly modulated by low-level visual features, such as the number of letters, while late brain responses appear to be mainly modulated by high-level lexical features, such as word frequency.
+In the next sections, we detail back-to-back regression, prove that it asymptotically extracts E, and show how it relates to and departs from other cross-decomposition and component analyses.
+We then show with synthetic experiments that back-to-back regression outperforms state-of-the-art techniques for regression and feature extraction, both in terms of causal discovery and out-of-environment predictions.
+Finally, we apply back-to-back regression to the multidimensional recordings of brain activity of 102 subjects reading $\approx$ 2,000 words, whose underlying features covary.
+Our results  show that the first brain responses to word presentation is mainly modulated by low-level visual features, such as the number and category of letters, while late brain responses appear to be mainly modulated by high-level lexical features, such as word frequency and part-of-speech.
 \par
@@
 
-Back and forth regression consists in two steps.
+Back-to-back regression consists of two steps.
 %
@@
 %
-This will not work here, as back-and-forth regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
+This will not work here, as back-to-back regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
 %
@@
 %
-To use back-and-forth regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
+To use back-to-back regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
 
-Summarizing, back-and-forth regression is performed as follows:
+Summarizing, back-to-back regression is performed as follows:
 \begin{enumerate}
@@
 
-The two steps of back-and-forth regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).
+The two steps of back-to-back regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).
 
@@
 
-Thus, in the linear case, back-and-forth regression retrieves $\hat H = \frac{1}{1+nsr}(FE)^{\dagger}(FE)$.
+Thus, in the linear case, back-to-back regression retrieves $\hat H = \frac{1}{1+nsr}(FE)^{\dagger}(FE)$.
 
@@
 \section{Related works}
-\subsection{back-and-forth regression as a special case of canonical component analysis?}
-Formula (1) shows that back-and-forth regression may be related to canonical component analysis (CCA).
+\subsection{back-to-back regression as a special case of canonical component analysis?}
+Formula (1) shows that back-to-back regression may be related to canonical component analysis (CCA).
 %
@@
 %
-Thus, back-and-forth regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
+Thus, back-to-back regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
 %
-The closer we are to this hypothesis, the closer the results of CCA and back-and-forth regression will be.
+The closer we are to this hypothesis, the closer the results of CCA and back-to-back regression will be.
 
-There are differences between CCA and back-and-forth regression.
+There are differences between CCA and back-to-back regression.
 %
@@
 \section{Experiments - simulated data}
-Experiments on simulated data serve three purposes: compare back-and-forth regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.
+Experiments on simulated data serve three purposes: compare back-to-back regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.
 
@@
 
-We evaluate the performance of back-and-forth regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
+We evaluate the performance of back-to-back regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
 %
@@
 \subsection{Feature extraction - linear case}
-All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by back-and-forth regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
+All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by back-to-back regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
 %
@@
 
-For feature extraction, back-and-forth regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
+For feature extraction, back-to-back regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
 %
@@
 %
-back-and-forth regression allows for good feature extraction if :
+back-to-back regression allows for good feature extraction if :
 
@@
 %
-This demonstrates that back-and-forth regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.
+This demonstrates that back-to-back regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.
 
@@
 
-The rationale behind such tests is that whereas back-and-forth regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.
+The rationale behind such tests is that whereas back-to-back regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.
 

### 2019-05-22 code/synthetic_experiment/main.py
jrr3

@@ -1,7 +1,7 @@
 # TODO: model Y = (XE + N)F + N' with second noise N'
 # TODO: uniform distributions
 
-from models import JRR, OLS, Ridge, Oracle, PLS, Lasso, RRR, CCA, JRR2
+from models import JRR, OLS, Ridge, Oracle, PLS, Lasso, RRR, CCA, JRR2, JRR3
 from sklearn.metrics import roc_auc_score
 from data import Synthetic
 import numpy as np
@@ -11,6 +11,7 @@ import random
 models = {
     "JRR": JRR,
     "JRR2": JRR2,
+    "JRR3": JRR3,
     "OLS": OLS,
     "Ridge": Ridge,
     "CCA": CCA,

### 2019-05-22 code/synthetic_experiment/models.py

@@ -266,6 +266,7 @@ class JRR(object):
     def solution(self):
         return self.E
 
+
 class JRR2(object):
     def __init__(self):
         alphas = np.logspace(-5, 5, 20)
@@ -283,6 +284,98 @@ class JRR2(object):
 
         self.E = E / self.n_splits
 
+        self.coef = basic_regression(X @ np.diag(self.E), Y)
+
+    def predict(self, X):
+        return X @ np.diag(self.E) @ self.coef
+
+    def solution(self):
+        return self.E
+
+
+def ridge_cv_Uv(X, Y, alphas, independent_alphas=False, Uv=None):
+    """
+    Similar to sklearn RidgeCV but
+    (1) can optimize a different alpha for each column of Y
+    (2) return leave-one-out Y_hat
+    """
+    from scipy.linalg import svd
+
+    if isinstance(alphas, (float, int)):
+        alphas = np.array([alphas, ], np.float64)
+    n, n_x = X.shape
+    n, n_y = Y.shape
+    # Decompose X
+    if Uv is None:
+        U, s, _ = svd(X, full_matrices=0)
+        v = s**2
+    else:
+        U, v = Uv
+    UY = U.T @ Y
+
+    # For each alpha, solve leave-one-out error coefs
+    cv_duals = np.zeros((len(alphas), n, n_y))
+    cv_errors = np.zeros((len(alphas), n, n_y))
+    for alpha_idx, alpha in enumerate(alphas):
+        # Solve
+        w = ((v + alpha) ** -1) - alpha ** -1
+        c = U @ np.diag(w) @ UY + alpha ** -1 * Y
+        cv_duals[alpha_idx] = c
+
+        # compute diagonal of the matrix: dot(Q, dot(diag(v_prime), Q^T))
+        G_diag = (w * U ** 2).sum(axis=-1) + alpha ** -1
+        error = c / G_diag[:, np.newaxis]
+        cv_errors[alpha_idx] = error
+
+    # identify best alpha for each column of Y independently
+    if independent_alphas:
+        best_alphas = (cv_errors ** 2).mean(axis=1).argmin(axis=0)
+        duals = np.transpose([cv_duals[b, :, i]
+                              for i, b in enumerate(best_alphas)])
+        cv_errors = np.transpose([cv_errors[b, :, i]
+                                  for i, b in enumerate(best_alphas)])
+    else:
[…]

### 2019-05-22 paper/paper.tex
intro fig

@@
 %
-This problem can be formalized as follow: empirical studies consist in measuring, via an apparatus $F$, a signal $Y \in \mathbb{R}^{i, l}$, that is a function of possible causes $X \in \mathbb{R}^{i, j}$, and other unknown causes that can be understood as noise $N \in \mathbb{R}^{i, k}$: $Y=F(X,N)$.
-Of all considered causes (features of $X$), only a subset accounts for $Y$, let it be represented as $E \cdot X$, with $E \in \mathbb{R}^{j, j}$ a binary diagonal matrix, which selects causal features of $X$.
+This problem can be formalized as follow: empirical studies consist in measuring, via an apparatus $F$, a signal $Y \in \mathbb{R}^{n, q}$, that is a function of possible causes $X \in \mathbb{R}^{n, d}$, and other unknown causes that can be understood as noise $N \in \mathbb{R}^{n, d}$: $Y=F(X,N)$.
+Of all considered causes (features of $X$), only a subset accounts for $Y$, let it be represented as $E \cdot X$, with $E \in \mathbb{D}^{d, d}$ a binary diagonal matrix, which selects causal features of $X$.
 %
@@
 
-    \node[] (labY) at (\posY, 2){observed effects};
-    \node[] (labX) at (\posX, 2.3){world};
-    \node[] (labX) at (\posX, 2){features};
-    \node[] (labE) at (\posE, 2.3){cause};
+    \node[] (labY) at (\posY, 2){Observations};
+    \node[] (labX) at (\posX, 2){Factors};
+    \node[] (labE) at (\posE, 2.3){Cause};
     \node[] (labE) at (\posE, 2){selection};
-    \node[] (labN) at (\posN, 2){noise};
-    \node[] (labF) at (\posF, 2.3){cause-effect};
+    \node[] (labN) at (\posN, 2){Noise};
+    \node[] (labF) at (\posF, 2.3){Cause-effect};
     \node[] (labF) at (\posF, 2){mapping};
@@
     \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
-    \node[] (reg1) at (5,-3.25) {$\hat{E} = \overbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\underbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}_{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}^{\text{2) regression from } X \text{ to } \hat{X}} $};
+    \node[] (reg1) at (5,-3.25) {$\hat{E} = \underbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}} $};
     \end{tikzpicture}
-    \caption{Summary of our method blah blah blah}
+    \caption{In its linear form, back-to-back regression identifies the subset of factors that suffices to account for multidimensional observations by piping two successive regressions from $Y$ to $X$ and from $X$ to $\hat X$}
     \label{fig:}
@@
 %
-If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we are maximising the Sonquist and Morgan criterion $$K = {p (n-p) \over n} {(s-r)^2 \over V}$$ and selecting the p features corresponding to the largest values.
+If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we are maximizing the Sonquist and Morgan criterion $$K = {p (n-p) \over n} {(s-r)^2 \over V}$$ and selecting the p features corresponding to the largest values.
 %
@@
 
-As noise N gets close to zero, and if F and X have full rank, then the solution tends to $(FE)^\dagger$, the pseudo inverse of FE.
+As N tends to zero, we are searching for the minimal norm solution of $\min_G \left \| X-GFEX\right\| ^2$, which is $X(FEX)^\dagger$ ($M^\dagger$ denoting the pseudo inverse of M).
+
+Let k be the number of active features of E, suppose for notational convenience that they are the first k features of X, and let $FEX$ be of rank k (which means F and X are of full rank on the space spanned by the active features of E).
+Let K be the (nx,k) diagonal matrix built by deleting the rightmost nx-k columns of E.
+Then, $K=(\frac{I_{k}}{0})$.
+we may write $E=KK'$, and $FEX=(FK)(K'X)$, with $FK$  and $K'X$ of full rank k.
+
+The rank factorisation property of the pseudo inverse (ref) says that, if $A=BC$, A (m,n), B(m,k) or rank k C(k,n) of rank k, then $A^\dagger=C^\dagger B^\dagger= C' (CC')^{-1}(B'B)^{-1}B'$.
+
+Replacing in the expression of the minimal norm solution yields
+\begin{equation}
+\begin{aligned}
+X(FEX)^\dagger &=X((FK)(K'X))^\dagger \\
+&=X(K'X)'((K'X)(K'X)')^{-1}((FK)(FK)')^{-1}(FK)' \\
+&=(XX')K(K'XX'K)^{-1}(FKK'F')^{-1} K'F'
+\end{aligned}
+\end{equation}
+
+Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{3} & \Sigma_{4}\end{array}\right)$, the top left block being (k,k), since $K=\left(\begin{array}{c}I \\\hline 0\end{array}\right)$, we have $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{3} \Sigma_{1}^{-1}\end{array}\right)=S$
+
+Thus we have $X(FEX)^\dagger =S (FKK'F')^{-1} K'F'$
+
+FC: this explains the negative diagonal elements, when X is negatively correlated. Need now to prove that this is S(FE)+
+
+FC: onward to the reduction, I assume that we get a scaled value of S(FE)+, this needs a better proof
 

### 2019-05-22 code/synthetic_experiment/main.py
Stuff

@@ -3,9 +3,11 @@
 
 from models import JRR, OLS, Ridge, Oracle, PLS, Lasso, RRR, CCA, JRR2
 from sklearn.metrics import roc_auc_score
+from sklearn.exceptions import ConvergenceWarning
 from data import Synthetic
 import numpy as np
 import argparse
+import warnings
 import random
 
 models = {
@@ -150,5 +152,8 @@ if __name__ == "__main__":
     parser.add_argument('--nonlinear', type=int, default=0)
     parser.add_argument('--n_seeds', type=int, default=10)
     args = parser.parse_args()
+
+    warnings.simplefilter(action='ignore', category=FutureWarning)
+    warnings.simplefilter(action='ignore', category=ConvergenceWarning)
 
     run_experiment(args)

### 2019-05-22 paper/paper.tex

@@
 %
-This problem can be formalized as follow: empirical studies consist in measuring, via an apparatus $F$, a signal $Y \in \mathbb{R}^{n, q}$, that is a function of possible causes $X \in \mathbb{R}^{n, d}$, and other unknown causes that can be understood as noise $N \in \mathbb{R}^{n, d}$: $Y=F(X,N)$.
-Of all considered causes (features of $X$), only a subset accounts for $Y$, let it be represented as $E \cdot X$, with $E \in \mathbb{D}^{d, d}$ a binary diagonal matrix, which selects causal features of $X$.
+This problem can be formalized as follow: empirical studies consist in measuring, via an apparatus $F$, a signal $Y \in \mathbb{R}^{i, l}$, that is a function of possible causes $X \in \mathbb{R}^{i, j}$, and other unknown causes that can be understood as noise $N \in \mathbb{R}^{i, k}$: $Y=F(X,N)$.
+Of all considered causes (features of $X$), only a subset accounts for $Y$, let it be represented as $E \cdot X$, with $E \in \mathbb{R}^{j, j}$ a binary diagonal matrix, which selects causal features of $X$.
 %
@@
 
-    \node[] (labY) at (\posY, 2){Observations};
-    \node[] (labX) at (\posX, 2){Factors};
-    \node[] (labE) at (\posE, 2.3){Cause};
+    \node[] (labY) at (\posY, 2){observed effects};
+    \node[] (labX) at (\posX, 2.3){world};
+    \node[] (labX) at (\posX, 2){features};
+    \node[] (labE) at (\posE, 2.3){cause};
     \node[] (labE) at (\posE, 2){selection};
-    \node[] (labN) at (\posN, 2){Noise};
-    \node[] (labF) at (\posF, 2.3){Cause-effect};
+    \node[] (labN) at (\posN, 2){noise};
+    \node[] (labF) at (\posF, 2.3){cause-effect};
     \node[] (labF) at (\posF, 2){mapping};
@@
     \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
-    \node[] (reg1) at (5,-3.25) {$\hat{E} = \underbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}} $};
+    \node[] (reg1) at (5,-3.25) {$\hat{E} = \overbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\underbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}_{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}^{\text{2) regression from } X \text{ to } \hat{X}} $};
     \end{tikzpicture}
-    \caption{In its linear form, back-to-back regression identifies the subset of factors that suffices to account for multidimensional observations by piping two successive regressions from $Y$ to $X$ and from $X$ to $\hat X$}
+    \caption{Summary of our method blah blah blah}
     \label{fig:}
@@
 %
-If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we are maximizing the Sonquist and Morgan criterion $$K = {p (n-p) \over n} {(s-r)^2 \over V}$$ and selecting the p features corresponding to the largest values.
+If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we are maximising the Sonquist and Morgan criterion $$K = {p (n-p) \over n} {(s-r)^2 \over V}$$ and selecting the p features corresponding to the largest values.
 %

### 2019-05-22 code/synthetic_experiment/main.py
fix jrr3

@@ -10,7 +10,7 @@ import random
 
 models = {
     "JRR": JRR,
-    "JRR2": JRR2,
+    # "JRR2": JRR2,
     "JRR3": JRR3,
     "OLS": OLS,
     "Ridge": Ridge,

### 2019-05-22 code/synthetic_experiment/models.py

@@ -244,7 +244,7 @@ class CCA(SKModel):
 class JRR(object):
     def __init__(self):
         alphas = np.logspace(-5, 5, 20)
-        self.G = MyRidgeCV(alphas, independent=True)
+        self.G = MyRidgeCV(alphas, independent=False)
         self.H = MyRidgeCV(alphas, independent=False)
         self.n_splits = 100
 
@@ -368,7 +368,7 @@ class JRR3(object):
             p = np.random.permutation(range(len(X)))
             train, test = p[::2], p[1::2]
             G, _, _ = ridge_cv_Uv(Y[train], X[train], self.alphas,
-                                  independent_alphas=True,
+                                  independent_alphas=False,
                                   Uv=(Y[train] @ Vs_inv_y, v_y))
             X_hat = Y[test] @ G.T
             Cxtxh.append(Xt[test].T @ X_hat)

### 2019-05-22 paper/paper.tex
math proof, pretty much done

@@
 
-\title{Discovering causal factors with back-to-back regression}
+\title{Discovering causal factors with back-and-forth regression}
 
@@
 manipulate individually and ii) observations are complex and multi-dimensional.
-To address this issue, we introduce ``back-to-back regression'' (BFR), a
+To address this issue, we introduce ``back-and-forth regression'' (BFR), a
 method designed to identify, from a set of co-varying factors, the factors
@@
 This problem can be formalized as follow: empirical studies consist in measuring, via an apparatus $F$, a signal $Y \in \mathbb{R}^{i, l}$, that is a function of possible causes $X \in \mathbb{R}^{i, j}$, and other unknown causes that can be understood as noise $N \in \mathbb{R}^{i, k}$: $Y=F(X,N)$.
-Of all considered causes (features of $X$), only a subset accounts for $Y$, let it be represented as $E \cdot X$, with $E \in \mathbb{R}^{j, j}$ a binary diagonal matrix, which selects causal features of $X$.
+Of all considered causes (features of X), only a subset accounts for $Y$, let it be represented as $E \cdot X$, with $E \in \mathbb{R}^{j, j}$ a binary diagonal matrix, which selects causal features of $X$.
 %
-Under the assumptions of centered and additive noise N, causal discovery thus consists in finding $E$ in
+Under the restricted assumptions of centered and additive noise N and linear systems, causal discovery thus consists in finding $E$ in
 
 \begin{equation}
-    Y=F(XE, N)
+    Y=(XEM+N)F
 \end{equation}
+with $M \in \mathbb{R}^{j, k}$ and $F \in \mathbb{R}^{j, k}$ being unknown.
+$M$ is a unknown projection matrix that ensures that the noise and the causal factors share a common space, and can thus be absorbed by $F$:
+\begin{equation}
+    Y=(XE+N^{*})F^{*}
+\end{equation}
+\par
 
@@
 %
-% To illustrate these two challenges, let us consider the case of a neuroscientist who studies the brain bases of reading.
+To illustrate these two challenges, let us consider the case of a neuroscientist who studies the brain bases of reading.
 Our neuroscientist wants to determine whether the frequency of a word modulates brain activity (e.g.
@@
 %
-% This is the first challenge to causal discovery
-The first challenge is factor collinearity: causal and non-causal factors may tend to covary with one another: i.e.
-$X'X$ is not diagonal.
-Collinearity in $X$ is typically addressed with factorial designs \cite{fisher_1935}.
-% For example, our neuroscientist may carefully select words matched in terms of length, while letting their frequency vary.
-However, factorial design becomes increasingly difficult to implement as the number of considered factors grows.
-As a fallback option, factors may be orthogonalized \textit{a posteriori}, by analytically estimating how each of them contributes to explaining the observations with ordinary least square (OLS):
-
-% Our neuroscientist will thus
+%
+This is the first challenge to causal discovery: a set of putative factors, some causal, other non-causal, can be collinear.
+Collinearity in the putative factors $X$ is typically addressed with factorial designs \cite{fisher_1935}.
+For example, our neuroscientist may carefully select words matched in terms of length, while letting their frequency vary.
+However, factorial design becomes increasingly difficult to implement as the number of considered factor grows.
+As a fallback option, factors may be orthogonalized \textit{a posteriori}, by analytically estimating how each of them contributes to explaining the observations.
+Our neuroscientist will thus typically use ordinary least square (OLS) to model how brain activity can be specifically accounted for by word length and word frequency:
 \begin{equation}
@@
 The second challenge to causal discovery is the multiplicity of the observations.
-% For example, our neuroscientist may simultaneously collect noisy recordings of hundreds of thousands of neurons.
+For example, our neuroscientist may simultaneously collect noisy recordings of hundreds of thousands of neurons.
 It can thus be suboptimal to consider each neuron independently, because they may be influenced by some unknown factors $N$.
-With the rise of large datasets, the complexity of observations has led many to focus on optimally predicting putative factors from complex and poorly controlled observations.
+More generally, with the rise of large and complex datasets, the complexity of observations has led many to invert the problem and to focus on optimally predicting putative factors from complex and poorly controlled observations.
 In neuroimaging, for example, it is increasingly popular to use "decoding" methods: instead of testing whether the brain responds to particular factors, a model is trained to maximally predict factors from complex brain signals:
@@
 1., both forward and backward modeling have competing benefits and drawbacks.
-Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations.
+Specifically, forward modeling efficiently disentangles the independent contribution of collinear factors, but does not combine multidimensional observations.
 By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors.
@@
 %
-In this paper, we introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling.
-Back-to-back regression consists in first predicting X from Y (as in the backward modeling) subsequently retrieving an estimate $\hat X$ and finally disentangling which of the considered factors X accounts for the variance in $\hat X$ (i.e.
+In this paper, we introduce the 'back-and-forth regression', which combines the benefits of forward and backward modeling.
+Back-and-forth regression consists in first predicting X from Y (as in the backward modeling) subsequently retrieving an estimate $\hat X$ and finally disentangling which of the considered factors X accounts for the variance in $\hat X$ (i.e.
 analogously to forward modeling).
@@
 %
-In the next sections, we detail back-to-back regression, prove that it asymptotically extracts E, and show how it relates to and departs from other cross-decomposition and component analyses.
-We then show with synthetic experiments that back-to-back regression outperforms state-of-the-art techniques for regression and feature extraction, both in terms of causal discovery and out-of-environment predictions.
-Finally, we apply back-to-back regression to the multidimensional recordings of brain activity of 102 subjects reading $\approx$ 2,000 words, whose underlying features covary.
-Our results  show that the first brain responses to word presentation is mainly modulated by low-level visual features, such as the number and category of letters, while late brain responses appear to be mainly modulated by high-level lexical features, such as word frequency and part-of-speech.
+In the next sections, we detail back-and-forth regression, prove that it asymptotically extracts E, and show how it relates to and departs from other cross-decomposition and component analyses.
+We then show with synthetic experiments that back-and-forth regression outperforms state-of-the-art techniques for regression and feature extraction, both in terms of causal discovery and out-of-environment predictions.
+Finally, we apply back-and-forth regression to the brain activity of 102 subjects reading $\approx$ 2,000 words, and show that the first brain responses to these words appear to be mainly modulated by low-level visual features, such as the number of letters, while late brain responses appear to be mainly modulated by high-level lexical features, such as word frequency.
 \par
@@
 
-Back-to-back regression consists of two steps.
+Back and forth regression consists in two steps.
 %
@@
 %
-This will not work here, as back-to-back regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
+This will not work here, as back-and-forth regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
 %
@@
 %
-To use back-to-back regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
+To use back-and-forth regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
 
-Summarizing, back-to-back regression is performed as follows:
+Summarizing, back-and-forth regression is performed as follows:
 \begin{enumerate}
@@
 
-The two steps of back-to-back regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).
+The two steps of back-and-forth regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).
 
@@
 X(FEX)^\dagger &=X((FK)(K'X))^\dagger \\
-&=X(K'X)'((K'X)(K'X)')^{-1}((FK)(FK)')^{-1}(FK)' \\
-&=(XX')K(K'XX'K)^{-1}(FKK'F')^{-1} K'F' 
-\end{aligned}
-\end{equation}
+&=X(K'X)'((K'X)(K'X)')^{-1}((FK)'(FK))^{-1}(FK)' \\
+&=(XX')K(K'XX'K)^{-1}(K'F'FK)^{-1} K'F' 
+\end{aligned}\end{equation}
 
-Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{3} & \Sigma_{4}\end{array}\right)$, the top left block being (k,k), since $K=\left(\begin{array}{c}I \\\hline 0\end{array}\right)$, we have $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{3} \Sigma_{1}^{-1}\end{array}\right)=S$
+Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{3} & \Sigma_{4}\end{array}\right)$, the top left block being (k,k), since $K=\left(\begin{array}{c}I \\\hline 0\end{array}\right)$, we have $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{3} \Sigma_{1}^{-1}\end{array}\right)$, call S the matrix obtained by adding Nx-k empty columns to this one (or, equivalently multiplying it by K').
 
-Thus we have $X(FEX)^\dagger =S (FKK'F')^{-1} K'F'$
+Let us also note that $FE=FKK'$, and so $(K'F'FK)^{-1} K'F'FKK'=I_{k}K'$. Therefore, we have $(K'F'FK)^{-1} K'F'=(FE)^\dagger$.
 
-FC: this explains the negative diagonal elements, when X is negatively correlated. Need now to prove that this is S(FE)+
+Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=X(FEX)^\dagger =S (FE)^\dagger$.
+If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (that is, if causal features do not act as confounders for inactive features), then S=E, and we are retrieving $E (FE)^\dagger$.
 
-FC: onward to the reduction, I assume that we get a scaled value of S(FE)+, this needs a better proof
+Under centred noise, and if (FN)'(FN) have low condition number (a weak assumption given that N is noise), we have $\left \| GFN\right \| ^2 \approx Var(FN) \left \| G\right \| ^2$, and minimisation of $\left \| X-GFEX\right\| ^2  + \left \| GFN\right \| ^2$ will recover of the minimum of the left norm, $\lambda S (FE)^\dagger$, $\lambda$ a positive number below 1.
 
-If matrices X'X and (FN)'(FN) have low condition numbers, and denoting by VarX and VarN the variance of X and FN, we have $$\min_G \Tr ((I-GFE)'(I-GFE)XX') + \Tr (G'GFNN'F') \approx \left \| (I-GFE)\right\| ^2 VarX  + \left \| G\right \| ^2 VarN$$and the solution has the form $k(FE)^\dagger$, k a positive constant below one.
-To determine an approximate value of k, note that if r is the rank of FE (the number of ones in E) $$\left \| (I-GFE)\right\| ^2 = r (1-k)^2 + dx - r$$ and that under the hypothesis of F having a low condition number, the squared norm of $k(FE)^\dagger$ should be close to $k^2r$.
-We are therefore minimising a quantity close to $$ VarX (r (1-k)^2 + dx - r) + VarN k^2 r$$ Zeroing the derivative in k yields an approximate value of k, as $$k= \frac {VarX}{VarX + VarN}=\frac{1}{1+nsr}$$ nsr the noise to signal ratio, VarN divided by VarX.
+Replacing, we have $\left \| X-GFEX\right\| ^2 = \left \| X-\lambda S (FE)^\dagger FEX\right\| = \left \| X(I-\lambda S)\right\|$.
+If the covariance of X is block diagonal (no confounding between active and inactive features of X), this is equal to $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ (denoting the variance of X over active and inactive features).
+
+For the right part of the sum, we have $GFN=\lambda S (FE)^\dagger FN= \lambda S (FE)^\dagger (FE+F(I-E))N= \lambda S (E +(FE)^\dagger F(I-E))N=\lambda E N $ (FC last part of proof to be done, need some lemma on (FE)+).
+The square norm is therefore $\lambda^2 Var(N) (k/nx)$.
+
+The total quantity to be minimised, under noise, but with non confounding features in X, F full rank over E, and FN decently conditioned, is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N) (k/nx)$, setting the partial derivative in $\lambda$ to zero, we obtain $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N) (k/nx)}$, and if we suppose that X has comparable variance over all features, this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).
+
+The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} E (FE)^\dagger$ if active and inactive features of X are not correlated
 
@@
 
-Thus, in the linear case, back-to-back regression retrieves $\hat H = \frac{1}{1+nsr}(FE)^{\dagger}(FE)$.
-
-For any matrix A, $A^\dagger A$ is the orthogonal projector over $Img(A')$, and $I-A^\dagger A$ the orthogonal projector over $Ker(A)$.
-%
-In our case, $Ker(FE)\supseteq Ker(E)$.
-%
-If F has full rank over $Img(E)$, the subspace spanned by E, then $Ker(FE) = Ker(E)$,.
-%
-This means $I - \hat H$ is the projector over $Ker(E)$, the diagonal matrix that spans the unselected features, and $\hat H$ is k times the diagonal matrix corresponding to the selected features, that is kE.
-Since it is diagonal with eigenvalues 0 and k, if k is large enough, it should allow us to retrieve causal features in a fairly robust way.
+Thus, in the linear case, back-and-forth regression retrieves $\hat H = \frac{1}{1+nsr} E (FE)^{\dagger}(FE) = \frac{1}{1+nsr} E$.
 
@@
 
-\subsection{Stability analysis}
-
 \section{Related works}
-\subsection{back-to-back regression as a special case of canonical component analysis?}
-Formula (1) shows that back-to-back regression may be related to canonical component analysis (CCA).
+\subsection{back-and-forth regression as a special case of canonical component analysis?}
+Formula (1) shows that back-and-forth regression may be related to canonical component analysis (CCA).
 %
@@
 %
-Thus, back-to-back regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
+Thus, back-and-forth regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
 %
-The closer we are to this hypothesis, the closer the results of CCA and back-to-back regression will be.
+The closer we are to this hypothesis, the closer the results of CCA and back-and-forth regression will be.
 
-There are differences between CCA and back-to-back regression.
+There are differences between CCA and back-and-forth regression.
 %
@@
 \section{Experiments - simulated data}
-Experiments on simulated data serve three purposes: compare back-to-back regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.
+Experiments on simulated data serve three purposes: compare back-and-forth regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.
 
@@
 
-We evaluate the performance of back-to-back regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
+We evaluate the performance of back-and-forth regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
 %
@@
 \subsection{Feature extraction - linear case}
-All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by back-to-back regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
+All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by back-and-forth regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
 %
@@
 
-For feature extraction, back-to-back regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
+For feature extraction, back-and-forth regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
 %
@@
 %
-back-to-back regression allows for good feature extraction if :
+back-and-forth regression allows for good feature extraction if :
 
@@
 %
-This demonstrates that back-to-back regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.
+This demonstrates that back-and-forth regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.
 
@@
 
-The rationale behind such tests is that whereas back-to-back regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.
+The rationale behind such tests is that whereas back-and-forth regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.
 

### 2019-05-22 paper/paper.tex

--- 2019-05-22 paper/paper.tex
@@
 X(FEX)^\dagger &=X((FK)(K'X))^\dagger \\
-&=X(K'X)'((K'X)(K'X)')^{-1}((FK)(FK)')^{-1}(FK)' \\
-&=(XX')K(K'XX'K)^{-1}(FKK'F')^{-1} K'F'
+&=X(K'X)'((K'X)(K'X)')^{-1}((FK)'(FK))^{-1}(FK)' \\
+&=(XX')K(K'XX'K)^{-1}(K'F'FK)^{-1} K'F' 
 \end{aligned}
@@
 
-Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{3} & \Sigma_{4}\end{array}\right)$, the top left block being (k,k), since $K=\left(\begin{array}{c}I \\\hline 0\end{array}\right)$, we have $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{3} \Sigma_{1}^{-1}\end{array}\right)=S$
+Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{3} & \Sigma_{4}\end{array}\right)$, the top left block being (k,k), since $K=\left(\begin{array}{c}I \\\hline 0\end{array}\right)$, we have $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{3} \Sigma_{1}^{-1}\end{array}\right)$, call S the matrix obtained by adding Nx-k empty columns to this one (or, equivalently multiplying it by K').
 
-Thus we have $X(FEX)^\dagger =S (FKK'F')^{-1} K'F'$
+Let us also note that $FE=FKK'$, and so $(K'F'FK)^{-1} K'F'FKK'=I_{k}K'$. Therefore, we have $(K'F'FK)^{-1} K'F'=(FE)^\dagger$.
 
-FC: this explains the negative diagonal elements, when X is negatively correlated. Need now to prove that this is S(FE)+
+Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=X(FEX)^\dagger =S (FE)^\dagger$.
+If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (that is, if causal features do not act as confounders for inactive features), then S=E, and we are retrieving $E (FE)^\dagger$.
 
-FC: onward to the reduction, I assume that we get a scaled value of S(FE)+, this needs a better proof
+Under centred noise, and if (FN)'(FN) have low condition number (a weak assumption given that N is noise), we have $\left \| GFN\right \| ^2 \approx Var(FN) \left \| G\right \| ^2$, and minimisation of $\left \| X-GFEX\right\| ^2  + \left \| GFN\right \| ^2$ will recover of the minimum of the left norm, $\lambda S (FE)^\dagger$, $\lambda$ a positive number below 1.
 
-If matrices X'X and (FN)'(FN) have low condition numbers, and denoting by VarX and VarN the variance of X and FN, we have $$\min_G \Tr ((I-GFE)'(I-GFE)XX') + \Tr (G'GFNN'F') \approx \left \| (I-GFE)\right\| ^2 VarX  + \left \| G\right \| ^2 VarN$$and the solution has the form $k(FE)^\dagger$, k a positive constant below one.
-To determine an approximate value of k, note that if r is the rank of FE (the number of ones in E) $$\left \| (I-GFE)\right\| ^2 = r (1-k)^2 + dx - r$$ and that under the hypothesis of F having a low condition number, the squared norm of $k(FE)^\dagger$ should be close to $k^2r$.
-We are therefore minimising a quantity close to $$ VarX (r (1-k)^2 + dx - r) + VarN k^2 r$$ Zeroing the derivative in k yields an approximate value of k, as $$k= \frac {VarX}{VarX + VarN}=\frac{1}{1+nsr}$$ nsr the noise to signal ratio, VarN divided by VarX.
+Replacing, we have $\left \| X-GFEX\right\| ^2 = \left \| X-\lambda S (FE)^\dagger FEX\right\| = \left \| X(I-\lambda S)\right\|$.
+If the covariance of X is block diagonal (no confounding between active and inactive features of X), this is equal to $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ (denoting the variance of X over active and inactive features).
+
+For the right part of the sum, we have $GFN=\lambda S (FE)^\dagger FN= \lambda S (FE)^\dagger (FE+F(I-E))N= \lambda S (E +(FE)^\dagger F(I-E))N=\lambda E N $ (FC last part of proof to be done, need some lemma on (FE)+).
+The square norm is therefore $\lambda^2 Var(N) (k/nx)$.
+
+The total quantity to be minimised, under noise, but with non confounding features in X, F full rank over E, and FN decently conditioned, is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N) (k/nx)$, setting the partial derivative in $\lambda$ to zero, we obtain $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N) (k/nx)}$, and if we suppose that X has comparable variance over all features, this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).
+
+The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} E (FE)^\dagger$ if active and inactive features of X are not correlated
 
@@
 
-Thus, in the linear case, back-to-back regression retrieves $\hat H = \frac{1}{1+nsr}(FE)^{\dagger}(FE)$.
-
-For any matrix A, $A^\dagger A$ is the orthogonal projector over $Img(A')$, and $I-A^\dagger A$ the orthogonal projector over $Ker(A)$.
-%
-In our case, $Ker(FE)\supseteq Ker(E)$.
-%
-If F has full rank over $Img(E)$, the subspace spanned by E, then $Ker(FE) = Ker(E)$,.
-%
-This means $I - \hat H$ is the projector over $Ker(E)$, the diagonal matrix that spans the unselected features, and $\hat H$ is k times the diagonal matrix corresponding to the selected features, that is kE.
-Since it is diagonal with eigenvalues 0 and k, if k is large enough, it should allow us to retrieve causal features in a fairly robust way.
+Thus, in the linear case, back-and-forth regression retrieves $\hat H = \frac{1}{1+nsr} E (FE)^{\dagger}(FE) = \frac{1}{1+nsr} E$.
 
 \subsection{Asymptotic behaviour - non linear case}
-
-\subsection{Stability analysis}
 

### 2019-05-22 paper/meg_exps.tex
wip: meg

@@
-\subsection{The MOUS dataset}
+\subsection{Dataset}
 
-Next, we apply our method to brain imaging data, on the ``Mother Of all Unification
-Studies'' (MOUS) dataset \cite{schoffelen2019204}. The dataset consists of
-anonymized multimodal neuroimaging data that has been acquired from 204 healthy
-human subjects, which participated in either a auditory or a visual version of a
-language experiment. In this work, we use the subset corresponding to the visual
-variant recorded using magneto-encephalography (MEG).
+Next, we apply our method to brain imaging data from the anonymized multimodal
+neuroimaging ``Mother Of all Unification Studies'' (MOUS) dataset \cite{schoffelen2019204}.
+The dataset contains magneto-encephalography (MEG) recordings of 102 healthy native-Dutch
+adults who participated in a reading task.
 
-Subjects were presented a set of 120 sentences in Dutch, and scrambled lists
-of the same words for a controlled experiment. Each word was presented on a
-screen for no less than 300 ms and no more than 1400ms, 351 ms on average.
-Successive words were separated by an empty screen for 300ms, and successive
+Subjects were exposed to a rapid serial visual presentation of Dutch words. The word lists
+consisted of 120 sentences, and scrambled lists of the same words. Each word was presented on the computer screen for 351ms on average (min: 300ms, max: 1400ms).
+Successive words were separated by a blank screen for 300ms, and successive
 sentences were separated by an empty screen for a few (3-4) seconds.
 
-The data contains the timestamp for each word presentation, allowing to align
-time series to stimuli onsets for all words.
+\subsection{MEG preprocessing}
 
-As a result, we obtain, for each of the 301 channels (273 magnetometers on the
-subject and 28 compensation channels measuring the room) and each word
-presented, a time series of 67 real-values spanning $\left[-100ms, 1000ms\right]$ relative to
-the stimulus onset. Each of the 102 subjects (we used data from 93 subjects) was
-presented approximately 2700 words during the experiment.
+The raw MEG data was bandpass-filtered between 0.1 and 40 Hz using MNE-Python default
+parameters \cite{mne}. Specifically, we used a zero-phase finite impulse response filter (FIR) with a hamming window and with transition bands of 0.1 Hz and 10 Hz for the low and high
+cut-off frequencies respectively.
 
-We show, in Figure \ref{fig:megavg}, the average response across words of the 
-magnetometers for the first subject. Time $t=0$ corresponds to the stimulus onset.
+The raw data was then segmented 100 ms before word onset and 1.000 sec after
+word onset.
+Time $t=0$ms corresponds to the word onset.
+Finally, each resulting segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and thus led a sampling frequency of 240 Hz.
+We show, in Figure \ref{fig:megavg}, the average response of the
+magnetometers evoked by the words presented to the first subject.
+
+For each word, each subject, and each time sample relative to word onset, we build an observation matrix $Y \in \mathbb{R}^{n, q}$ of $n\approx$ 2700 words by $q=301$ MEG channels (273 magnetometers and 28 compensation channels).
+Each of the columns of $Y$ is normalized to have a mean and a standard deviation of 0 and 1 respectively.
 
@@
 
-\subsection{Application of the method}
+\subsection{Word features}
+We annotated each word with 38 features: 26 corresponds to the number of each letter in the word (from 'a' to 'z', independently of their case), the total number of letters ('word length'), the frequency of the word in natural language (using the Zipf logarithmic scale \cite{marc_brysbaert}), and the part-of-speech tags (i.e.
+NOUN, PROPN (proper noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ (adjective), ADV (adverb), and NUM (numerical)) using the packages Spacy \cite{spacy} and WordFreq \cite{speer}.
+For clarity purposes, we refer to these features using two broad categories: visual (i.e.
+each letter and the total number of letters) and lexical (part-of-speech and word frequency).
+This procedure yields an $X \in \mathbb{R}^{n, d}$ matrix of $n\approx$ 2700 words by $d=38$ features for each subject.
+Each of the columns of $X$ is normalized to have a mean and a standard deviation of 0 and 1 respectively.
 
-Using this dataset, we study which features of input words can elicit a strong
-MEG response in the brain.  We add to each word (i) visual features: word length
-(1 dim.), letter count (26 dim.), and (ii) lexical features : word frequency (1
-dim.) part-of-speech tag (one of ADJ, ADP, ADV, CONJ, DET, NOUN, NUM, PRON,
-PROPN, VERB; 10 dim.). Sentence-level word features are left for future study.
+\subsection{Analyses}
+We aim to identify which factors of $X$ explain the MEG recordings $Y$ at each time sample and for each subject separately.
+To this aim, we implemented three types of analyses (forward, backward and back-to-back regressions), and tested whether their results were consistent across subjects, using a second-level statistics described below.
 
-This procedure yields an $\left(n \times 38\right)$-dimensional input matrix $X$,
+Forward regression, also known as 'encoding' in neuroimaging, here consisted in fitting a Ridge-regularized OLS from $X$ to $Y$:
+\begin{equation}
+H = (X^{T}X+\lambda I)^{-1} X^{T}Y
+\end{equation}
+where $\lambda$ is the regularization parameter fit with the leave-one-out cross-validation implemented in scikit-learn \cite{sklearn} from 20 possible value logarithmatically distributed between $10^{-5}$ and $10^5$.
+We aim to estimate the contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor.
+However, $H\in\mathbb{R}^{d, q}$ is a matrix of $d$ word features by $q$ MEG channels.
+We thus approximate the relative contribution of each feature by sum-squarring the H coefficients:
+\begin{equation}
+\hat e_d = \sum_{i=0}^{i=q} {h_d^i}^2
+\end{equation}
+
+Backward regression, also known as 'decoding' in neuroimaging, here consisted in fitting a Ridge-regularized OLS from $Y$ to $X$:
+\begin{equation}
+G = (Y^{T}Y+\lambda I)^{-1} Y^{T}X
+\end{equation}
+where $\lambda$ is the regularization parameter fit with the leave-one-out cross-validation implemented in scikit-learn \cite{sklearn} from 20 possible value logarithmically distributed between $10^{-5}$ and $10^5$.
+We aim to estimate the contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor.
+Using 10-fold cross-validation, we thus assessed the ability of this ridge regression to predict independent and identically distributed data:
+\begin{equation}
+\hat X_{test} = Y_{test}\cdot G_{train}
+\end{equation}
+and assessing, for each factor of X independently, the Pearson correlation coefficient $r$ between $X_{test}$ and $\hat X_{test}$ for each cross-validation split.
+
+Models were fit for each subject and for each time sample independently
+
+Using this dataset, we study which word features cause a brain response detectable
+via MEG.
+
+ $\left(n \times 38\right)$-dimensional input matrix $X$,
 where $n$ is the number of words for the subject. Each feature in $X$ is centered and

### 2019-05-22 paper/paper.tex
paper wip

--- 2019-05-22 paper/paper.tex
@@
     \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
-    \node[] (reg1) at (5,-3.25) {$\hat{E} = \overbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\underbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}_{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}^{\text{2) regression from } X \text{ to } \hat{X}} $};
+    \node[] (reg1) at (5,-3.25) {$\hat{E} = \underbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}} $};
     \end{tikzpicture}
-    \caption{Summary of our method blah blah blah}
+    \caption{In its linear form, back-to-back regression identifies the subset of factors that suffices to account for multidimensional observations by piping two successive regressions from $Y$ to $X$ and from $X$ to $\hat X$.
+}
     \label{fig:}

### 2019-05-22 paper/paper.tex

--- 2019-05-22 paper/paper.tex
@@
     \end{tikzpicture}
-    \caption{In its linear form, back-to-back regression identifies the subset of factors that suffices to account for multidimensional observations by piping two successive regressions from $Y$ to $X$ and from $X$ to $\hat X$}
+    \caption{In its linear form, back-to-back regression identifies the subset of factors that suffices to account for multidimensional observations by piping two successive regressions from $Y$ to $X$ and from $X$ to $\hat X$.
+}
     \label{fig:}

### 2019-05-22 paper/meg_exps.tex
wip: meg

@@
 For each word, each subject, and each time sample relative to word onset, we build an observation matrix $Y \in \mathbb{R}^{n, q}$ of $n\approx$ 2700 words by $q=301$ MEG channels (273 magnetometers and 28 compensation channels).
-Each of the columns of $Y$ is normalized to have a mean and a standard deviation of 0 and 1 respectively.
+Each of the columns of $Y$ is normalized to have zero mean and unit variance.
 
@@
 \subsection{Analyses}
-We aim to identify which factors of $X$ explain the MEG recordings $Y$ at each time sample and for each subject separately.
+We aimed to identify which factors of $X$ explain the MEG recordings $Y$ at each time sample and for each subject separately.
 To this aim, we implemented three types of analyses (forward, backward and back-to-back regressions), and tested whether their results were consistent across subjects, using a second-level statistics described below.
@@
 where $\lambda$ is the regularization parameter fit with the leave-one-out cross-validation implemented in scikit-learn \cite{sklearn} from 20 possible value logarithmatically distributed between $10^{-5}$ and $10^5$.
-We aim to estimate the contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor.
+We aimed to estimate the contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor.
 However, $H\in\mathbb{R}^{d, q}$ is a matrix of $d$ word features by $q$ MEG channels.
-We thus approximate the relative contribution of each feature by sum-squarring the H coefficients:
+We thus approximated the relative contribution of each feature by sum-squarring the H coefficients:
 \begin{equation}
@@
 Backward regression, also known as 'decoding' in neuroimaging, here consisted in fitting a Ridge-regularized OLS from $Y$ to $X$:
+
 \begin{equation}
@@
 \end{equation}
-where $\lambda$ is the regularization parameter fit with the leave-one-out cross-validation implemented in scikit-learn \cite{sklearn} from 20 possible value logarithmically distributed between $10^{-5}$ and $10^5$.
-We aim to estimate the contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor.
-Using 10-fold cross-validation, we thus assessed the ability of this ridge regression to predict independent and identically distributed data:
+
+where $\lambda$ is the regularization parameter fit with the leave-one-out cross-validation
+implemented in scikit-learn \cite{sklearn} from 20 possible value logarithmically distributed
+between $10^{-5}$ and $10^5$. We aimed to estimate the contribution $diag(\hat E)
+\in\mathbb{D}^{d} $ of each factor. Using 10-fold cross-validation, we thus assessed the ability
+of this ridge regression to predict independent and identically distributed data:
+
 \begin{equation}
@@
 \end{equation}
+
 and assessing, for each factor of X independently, the Pearson correlation coefficient $r$ between $X_{test}$ and $\hat X_{test}$ for each cross-validation split.
 
-Models were fit for each subject and for each time sample independently
+Finally, back-to-back regression was fit as decribed earlier using two additional stability tricks designed to minimize the impact of bagging.
+First the $G$ ridge regression was broken-down with SVD blablabla
+Second, $X$ was regularized using a maximal likelihood blablabla
+Third, no regularization
 
-Using this dataset, we study which word features cause a brain response detectable
-via MEG.
+Back-to-back regression directly leads to $\hat E$.
 
- $\left(n \times 38\right)$-dimensional input matrix $X$,
-where $n$ is the number of words for the subject. Each feature in $X$ is centered and
-scaled for zero mean and unit variance.
+\subsection{Second-level statistics}
 
-Then, we apply the BFR method independently on each of the 67 time slices of the
-MEG recordings. Each slice $Y_t$ is a $n \times 301$ matrix, also centered and
-scaled. We obtain a vector of 38 coefficients per slice.
+The three types of analyses were implemented for each subject and each time sample independently.
+To estimate the reliability of these estimates in the population, we tested, using one-sample t-test whether the distribution of the estimated contribution of each factor different from baseline ($t=-100$ms):
+
+\begin{equation}
+$ttest(\hat E_{t=-100} - \hat E_{\tau}$
+\end{equation}
+with $\tau \in $ 150ms and 400ms respectively
+
+\subsection{Results}
+
+We aimed to estimate
 

### 2019-05-22 paper/paper.tex
externalize intro

@@
 
-Natural sciences are tasked to find, from a set of hypothetical factors, the minimal subset that suffices to reliably predict novel observations potentially in novel environments.
-%
-This problem can be formalized as follow: empirical studies consist in measuring, via an apparatus $F$, a signal $Y \in \mathbb{R}^{n, q}$, that is a function of possible causes $X \in \mathbb{R}^{n, d}$, and other unknown causes that can be understood as noise $N \in \mathbb{R}^{n, d}$: $Y=F(X,N)$.
-Of all considered causes (features of $X$), only a subset accounts for $Y$, let it be represented as $E \cdot X$, with $E \in \mathbb{D}^{d, d}$ a binary diagonal matrix, which selects causal features of $X$.
-%
-Under the assumptions of centered and additive noise N, causal discovery thus consists in finding $E$ in
-
-\begin{equation}
-    Y=F(XE, N)
-\end{equation}
-
-%
-Under this formalism, causal discovery is impeded by two challenges.
-%
-% To illustrate these two challenges, let us consider the case of a neuroscientist who studies the brain bases of reading.
-Our neuroscientist wants to determine whether the frequency of a word modulates brain activity (e.g.
-neurons may become more active when our retina is flashed with the word 'triskaidekaphobia' than when the word is 'car').
-One issue is that word frequency tends to correlate with word length.\par
-%
-% This is the first challenge to causal discovery
-The first challenge is factor collinearity: causal and non-causal factors may tend to covary with one another: i.e.
-$X'X$ is not diagonal.
-Collinearity in $X$ is typically addressed with factorial designs \cite{fisher_1935}.
-% For example, our neuroscientist may carefully select words matched in terms of length, while letting their frequency vary.
-However, factorial design becomes increasingly difficult to implement as the number of considered factors grows.
-As a fallback option, factors may be orthogonalized \textit{a posteriori}, by analytically estimating how each of them contributes to explaining the observations with ordinary least square (OLS):
-
-% Our neuroscientist will thus
-\begin{equation}
-H = (X'X)^{-1} X'Y
-\end{equation}
-In this paper, we refer to this approach as 'forward modeling' as it models how the factors X contribute to explain the observations Y. \par
-%
-%
-The second challenge to causal discovery is the multiplicity of the observations.
-% For example, our neuroscientist may simultaneously collect noisy recordings of hundreds of thousands of neurons.
-It can thus be suboptimal to consider each neuron independently, because they may be influenced by some unknown factors $N$.
-With the rise of large datasets, the complexity of observations has led many to focus on optimally predicting putative factors from complex and poorly controlled observations.
-In neuroimaging, for example, it is increasingly popular to use "decoding" methods: instead of testing whether the brain responds to particular factors, a model is trained to maximally predict factors from complex brain signals:
-\begin{equation}
-G = (Y'Y)^{-1} Y'X
-\end{equation}
-We refer to this approach as 'backward modeling' as it models how the observations Y can be combined to predict the factors X. \par
-%
-%
-As illustrated in Fig.
-1., both forward and backward modeling have competing benefits and drawbacks.
-Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations.
-By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors.
-\par
-%
-%
-In this paper, we introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling.
-Back-to-back regression consists in first predicting X from Y (as in the backward modeling) subsequently retrieving an estimate $\hat X$ and finally disentangling which of the considered factors X accounts for the variance in $\hat X$ (i.e.
-analogously to forward modeling).
-\par
-%
-In the next sections, we detail back-to-back regression, prove that it asymptotically extracts E, and show how it relates to and departs from other cross-decomposition and component analyses.
-We then show with synthetic experiments that back-to-back regression outperforms state-of-the-art techniques for regression and feature extraction, both in terms of causal discovery and out-of-environment predictions.
-Finally, we apply back-to-back regression to the multidimensional recordings of brain activity of 102 subjects reading $\approx$ 2,000 words, whose underlying features covary.
-Our results  show that the first brain responses to word presentation is mainly modulated by low-level visual features, such as the number and category of letters, while late brain responses appear to be mainly modulated by high-level lexical features, such as word frequency and part-of-speech.
-\par
-%
+\input{introduction.tex}
 

### 2019-05-22 paper/paper.tex
small things

--- 2019-05-22 paper/paper.tex
@@
 method designed to identify, from a set of co-varying factors, the factors
-that most plausibley account for multidimensional observations. After detailing
+that most plausible account for multidimensional observations. After detailing
 the proof of convergence, we show that BFR outperforms least-squares and related
@@
 
-\section{Back an forth regression}
+\section{Back to back regression}
 
@@
     \end{tikzpicture}
-    \caption{In its linear form, back-to-back regression identifies the subset of factors that suffices to account for multidimensional observations by piping two successive regressions from $Y$ to $X$ and from $X$ to $\hat X$.
-}
+    \caption{In its linear form, back-to-back regression identifies the subset of factors that suffices to account for multidimensional observations by piping two successive regressions from $Y$ to $X$ and from $X$ to $\hat X$}
     \label{fig:}
@@
 
-Let k be the number of active features of E, suppose for notational convenience that they are the first k features of X, and let $FEX$ be of rank k (which means F and X are of full rank on the space spanned by the active features of E).
-Let K be the (nx,k) diagonal matrix built by deleting the rightmost nx-k columns of E.
-Then, $K=(\frac{I_{k}}{0})$.
-we may write $E=KK'$, and $FEX=(FK)(K'X)$, with $FK$  and $K'X$ of full rank k.
+Let k be the number of active features of E, suppose for notational convenience that they are the first k features of X, and let $FEX$ be of rank k (which means F and X have full rank on the space spanned by the active features of E).
+Let K be the $\mathbb{R}^{nx\times k}$ diagonal matrix obtained by deleting the rightmost nx-k columns of E.
+Then, $E=KK'$, and $FEX=(FK)(K'X)$, with $FK$  and $K'X$ of full rank k.
 
-The rank factorisation property of the pseudo inverse (ref) says that, if $A=BC$, A (m,n), B(m,k) or rank k C(k,n) of rank k, then $A^\dagger=C^\dagger B^\dagger= C' (CC')^{-1}(B'B)^{-1}B'$.
+The rank factorisation property of the pseudo inverse (ref) says that, if $A=BC$ with $A\in\mathbb{R}^{m\times n}$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$ of rank k, then $A^\dagger=C^\dagger B^\dagger$ and since C is full rank, $C^\dagger=C' (CC')^{-1}$.
 
@@
 X(FEX)^\dagger &=X((FK)(K'X))^\dagger \\
-&=X(K'X)'((K'X)(K'X)')^{-1}((FK)'(FK))^{-1}(FK)' \\
-&=(XX')K(K'XX'K)^{-1}(K'F'FK)^{-1} K'F' 
+&=X(K'X)'((K'X)(K'X)')^{-1}(FK)^\dagger \\
+&=(XX')K(K'XX'K)^{-1}(FK)^\dagger 
 \end{aligned}
@@
 
-Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{3} & \Sigma_{4}\end{array}\right)$, the top left block being (k,k), since $K=\left(\begin{array}{c}I \\\hline 0\end{array}\right)$, we have $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{3} \Sigma_{1}^{-1}\end{array}\right)$, call S the matrix obtained by adding Nx-k empty columns to this one (or, equivalently multiplying it by K').
+Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, the top left block being $k\times k$.
+Then, $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{2} \Sigma_{1}^{-1}\end{array}\right)$, call S the matrix obtained by adding Nx-k empty columns to this one (or, equivalently multiplying it by K').
 

### 2019-05-22 paper/introduction.tex

--- 2019-05-22 paper/paper.tex
@@
-\documentclass{article}
-
-\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx}
-\usepackage[hidelinks]{hyperref}
-\usepackage[]{neurips_2019}
-\DeclareMathOperator{\Tr}{Tr}
-
-\usetikzlibrary{calc}
-
-\tikzset{
-    ncbar angle/.initial=90,
-    ncbar/.style={
-        to path=(\tikztostart)
-        -- ($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)
-        -- ($(\tikztotarget)!($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztostart)$)
-        -- (\tikztotarget)
-    },
-    ncbar/.default=0.5cm,
-}
-
-\tikzset{round left paren/.style={ncbar=0.5cm,out=110,in=-110}}
-\tikzset{round right paren/.style={ncbar=0.5cm,out=70,in=-70}}
-
-\newcommand{\dlp}[1]{{\color{red} (DLP: #1)}}
-\newcommand{\mo}[1]{{\color{green} (MO: #1)}}
-\newcommand{\jrk}[1]{{\color{blue} (JRK: #1)}}
-\newcommand{\fc}[1]{{\color{orange} (FC: #1)}}
-
-\title{Discovering causal factors with back-to-back regression}
-
-\author{%
-  Jean-Remi King\\
-  CNRS\\
-  \texttt{email} \\
-  \And
-  Fran\c{c}ois Charton\\
-  Facebook AI\\
-  \texttt{fcharton@fb.com}\\
-  \And
-  David Lopez-Paz\\
-  Facebook AI\\
-  \texttt{dlp@fb.com}
-  \And
-  Maxime Oquab\\
-  Facebook AI\\
-  \texttt{qas@fb.com}
-}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-Identifying causes from observations is at the core of science. This endeavor
-is particularly challenging when i) potential factors are difficult to
-manipulate individually and ii) observations are complex and multi-dimensional.
-To address this issue, we introduce ``back-to-back regression'' (BFR), a
-method designed to identify, from a set of co-varying factors, the factors
-that most plausible account for multidimensional observations. After detailing
-the proof of convergence, we show that BFR outperforms least-squares and related
-regression techniques, as well as cross-decomposition methods (e.g. canonical
-correlation analysis, and partial least squares) on two tasks: causal
-identification and out-of-distribution prediction. Finally, we apply BFR to
-neuroimaging recordings of 102 subjects reading word sequences. The results
-show, as expected, that the early brain responses to words appear to be
-specifically caused by low-level visual factors whereas late brain responses
-appear to be caused by lexical factors, despite the fact that these factors
-covary in the study.
-% conclude  open
-\end{abstract}
-
-\section{Introduction}
-
 Natural sciences are tasked to find, from a set of hypothetical factors, the minimal subset that suffices to reliably predict novel observations potentially in novel environments.
@@
 %
-
-\section{Back to back regression}
-
-\begin{figure}[t!]
-    \centering
-    \begin{tikzpicture}
-    \newcommand\posY{0}
-    \newcommand\posX{3}
-    \newcommand\posE{5}
-    \newcommand\posN{7}
-    \newcommand\posF{9.5}
-
-    \node[thick, draw=black, minimum height=3cm, minimum width=2cm, fill=yellow!10] (Y) at (\posY, 0){};
-    \node[] (eq) at (1.5, 0){$=$};
-    \node[] (times) at (4, 0){$\times$};
-    \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=yellow!10] (X) at (\posX, 0){};
-    \node[thick, draw=black, minimum height=1cm, minimum width=1cm, fill=red!10] (E) at (\posE, 0){};
-
-    \draw[fill=white] (\posE - 0.5 + 0.0, 0.5 - 0.0) rectangle (\posE - 0.5 + 0.0 + 0.1, 0.5 - 0.0 - 0.1);
-    \draw[fill=black] (\posE - 0.5 + 0.1, 0.5 - 0.1) rectangle (\posE - 0.5 + 0.1 + 0.1, 0.5 - 0.1 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.2, 0.5 - 0.2) rectangle (\posE - 0.5 + 0.2 + 0.1, 0.5 - 0.2 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.3, 0.5 - 0.3) rectangle (\posE - 0.5 + 0.3 + 0.1, 0.5 - 0.3 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.4, 0.5 - 0.4) rectangle (\posE - 0.5 + 0.4 + 0.1, 0.5 - 0.4 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.5, 0.5 - 0.5) rectangle (\posE - 0.5 + 0.5 + 0.1, 0.5 - 0.5 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.6, 0.5 - 0.6) rectangle (\posE - 0.5 + 0.6 + 0.1, 0.5 - 0.6 - 0.1);
-    \draw[fill=black] (\posE - 0.5 + 0.7, 0.5 - 0.7) rectangle (\posE - 0.5 + 0.7 + 0.1, 0.5 - 0.7 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.8, 0.5 - 0.8) rectangle (\posE - 0.5 + 0.8 + 0.1, 0.5 - 0.8 - 0.1);
-    \draw[fill=black] (\posE - 0.5 + 0.9, 0.5 - 0.9) rectangle (\posE - 0.5 + 0.9 + 0.1, 0.5 - 0.9 - 0.1);
-
-    \node[] (plus) at (6, 0){$+$};
-    \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=blue!10] (N) at (\posN, 0){};
-    \node[thick, draw=black, minimum height=2cm, minimum width=1cm, fill=red!10] (F) at (\posF, 0){};
-    \draw [thick] (2.25, -1.5) to [round left paren ] (2.25, 1.5);
-    \draw [thick] (7.75, -1.5) to [round right paren ] (7.75, 1.5);
-
-    \node[] (times2) at (8.5, 0){$\times$};
-
-    \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{n \times q}$}};
-    \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{n \times d}$}};
-    \node[] (annE) at (\posE, -1.8){\scalebox{0.85}{$E \in \mathbb{D}^{d \times d}$}};
-    \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{n \times d}$}};
-    \node[] (annF) at (\posF, -1.8){\scalebox{0.85}{$F \in \mathbb{R}^{d \times q}$}};
-
-    \node[] (labY) at (\posY, 2){Observations};
-    \node[] (labX) at (\posX, 2){Factors};
-    \node[] (labE) at (\posE, 2.3){Cause};
-    \node[] (labE) at (\posE, 2){selection};
-    \node[] (labN) at (\posN, 2){Noise};
-    \node[] (labF) at (\posF, 2.3){Cause-effect};
-    \node[] (labF) at (\posF, 2){mapping};
-
-    \node[] (sim1) at (0,-3.25) {\scalebox{0.85}{$X \sim P(X)$}};
-    \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
-    \node[] (reg1) at (5,-3.25) {$\hat{E} = \underbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}} $};
-    \end{tikzpicture}
-    \caption{In its linear form, back-to-back regression identifies the subset of factors that suffices to account for multidimensional observations by piping two successive regressions from $Y$ to $X$ and from $X$ to $\hat X$}
-    \label{fig:}
-\end{figure}
-
-\subsection{Algorithm}
-
-Back-to-back regression consists of two steps.
-%
-First, we perform a least squares estimation of X from Y, that is we find $\hat G$ such that $\hat X=\hat G Y$ is the least square predictor of X.
-%
-This recovers an approximation of the causes of Y in X.
-%
-Then, we perform a least squares estimation of $\hat X$ from X, and determine $\hat H$ such that $\hat {\hat X}=\hat H X$ is the least square predictor of $\hat X$.
-%
-In matrix form, we have $\hat G=(Y'Y)^{-1} Y'X$, and
-\begin{equation} \hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X\end{equation}
-Under good conditions (close to linear dependence between X and Y, and covariance matrices of X and Y well conditioned), this recovers a scaled approximation to E.
-
-In practice, the covariance matrices of Y and X might exhibit multicollinearity and be badly conditioned, which make their inversion unstable.
-To guard against this, we compute their singular value decompositions, and truncate their lowest eigenvalues so as to reach a specific condition number in the Frobenius norm.
-We prefer the truncated SVD approach over the more common regularisation (ridge regression) because it introduces less bias in the estimation.
-A small amount of regularisation is still used, for numerical stability purposes.
-
-If $X'X=U'DU$, U orthogonal and D diagonal, and D* is D with the lowest diagonal elements set to zero, this amounts to replacing $(X'X)^{-1}$ by $U(D*^{-1}+\epsilon I)U'$, with $\epsilon$ a positive number close to machine floating point accuracy.
-D is truncated so as to have a certain condition number that is determined by leave one out cross validation.
-%
-Also, to prevent spurious correlations between the two steps (should they be performed on the same sample), our original sample is randomly split into two halves, and each step is performed over one of them, effectively decorrelating the two steps.
-%
-This is repeated over several (100) splits, and the resulting $\hat H$ are averaged.
-%
-This bagging compensates for the reduction in sample size caused by splitting.
-
-Under our hypotheses (centered noise, E a binary diagonal) and even under heavy noise, $\hat H$ is close to diagonal.
-%
-Yet, because of regularisation which adds positive bias to small eigenvalues, and noise which attenuates large eigenvalues, $\hat H$ is in fact a scaled and noisy approximation to E.
-%
-To recover a binary diagonal, as our estimate of E, we extract the diagonal of $\hat H$ and select its large elements.
-%
-The diagonal matrix $\hat E$, having large elements of H as its unit diagonal coefficients (the rest being zero) is our estimate of E, the active features of X.
-
-To select large elements of $\hat H$, the traditional approach in regression analysis consists in testing whether its coefficients differ from zero.
-This is usually done via variants of Student t-test.
-%
-This will not work here, as back-to-back regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
-%
-Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).
-
-This can be done by maximizing the ratio of inter-group variance over total variance (ie minimizing intra-group variance).
-%
-In a one dimensional two clusters setting, this amounts to sorting the data, and separating the p smallest from the n-p largest values so that inter group inertia is maximal.
-%
-If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we are maximizing the Sonquist and Morgan criterion $$K = {p (n-p) \over n} {(s-r)^2 \over V}$$ and selecting the p features corresponding to the largest values.
-%
-If diagonal elements of $\hat H$ are normally distributed, K follows a chi-square distribution with one degree of freedom.
-%
-At 95\% confidence level, a value of K superior to 3.84 makes our selection significant (for 99\% confidence, the corresponding value of K is 6.63) \citep{Kass_75}
-
-This extract from $\hat H$ a binary diagonal $\hat E$.
-%
-To use back-to-back regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
-
-Summarizing, back-to-back regression is performed as follows:
-\begin{enumerate}
-\item Randomly split the data sample.
-\item From the first half sample, using crossalidated ridge regression, find $\hat G$, the regularised least square regression of X from Y: $\hat G=(Y'Y)^{-1} Y'X$.
-\item From the second half sample, calculate $\hat X = \hat G Y$, and find $\hat H$, the regularized least square regression of $\hat X$ from X: $\hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X$.
-\item Repeat steps 1 to 3 and average the results as $\hat H$.
-\item Using the Sonquist and Morgan criterion, select the large diagonal elements of $\hat H$, set them to one and the rest of $\hat  H$ to zero.
-\item Use the resulting matrix $\hat E$ as an estimator of E, the active features in the causal process
-\item Alternatively, use $\hat E$ to predict Y from X, by performing a regularised least square regression of Y from $\hat E X$.
-%
-
-\end{enumerate}
-
-\subsection{Asymptotic behaviour - linear case}
-As discussed in the introduction, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(EX+N)$, with f an unknown function.
-%
-If F is linear we can rewrite it as $Y = F(EX + N)$, with X an (dx, n) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
-%
-N a (dx,n) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dx) transformation corresponding to the measuring apparatus, and Y a (dy, n) matrix of measured effects.
-
-The two steps of back-to-back regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).
-
-For the first step, replacing Y, and since N is centred, we have
-\begin{equation}
-\begin{aligned}
-\arg \min_G \left \| X-GY \right \|^2 &= \arg \min_G \left \| X - GF(EX+N)\right\|^2 \\
-&{}= \arg \min_G \left \| X-GFEX\right\| ^2  + \left \| GFN\right \| ^2
-\end{aligned}
-\end{equation}
-
-As N tends to zero, we are searching for the minimal norm solution of $\min_G \left \| X-GFEX\right\| ^2$, which is $X(FEX)^\dagger$ ($M^\dagger$ denoting the pseudo inverse of M).
-
-Let k be the number of active features of E, suppose for notational convenience that they are the first k features of X, and let $FEX$ be of rank k (which means F and X have full rank on the space spanned by the active features of E).
-Let K be the $\mathbb{R}^{nx\times k}$ diagonal matrix obtained by deleting the rightmost nx-k columns of E.
-Then, $E=KK'$, and $FEX=(FK)(K'X)$, with $FK$  and $K'X$ of full rank k.
-
-The rank factorisation property of the pseudo inverse (ref) says that, if $A=BC$ with $A\in\mathbb{R}^{m\times n}$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$ of rank k, then $A^\dagger=C^\dagger B^\dagger$ and since C is full rank, $C^\dagger=C' (CC')^{-1}$.
-
-Replacing in the expression of the minimal norm solution yields
-\begin{equation}
-\begin{aligned}
-X(FEX)^\dagger &=X((FK)(K'X))^\dagger \\
-&=X(K'X)'((K'X)(K'X)')^{-1}(FK)^\dagger \\
-&=(XX')K(K'XX'K)^{-1}(FK)^\dagger 
-\end{aligned}
-\end{equation}
-
-Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, the top left block being $k\times k$.
-Then, $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{2} \Sigma_{1}^{-1}\end{array}\right)$, call S the matrix obtained by adding Nx-k empty columns to this one (or, equivalently multiplying it by K').
-
-Let us also note that $FE=FKK'$, and so $(K'F'FK)^{-1} K'F'FKK'=I_{k}K'$. Therefore, we have $(K'F'FK)^{-1} K'F'=(FE)^\dagger$.
-
-Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=X(FEX)^\dagger =S (FE)^\dagger$.
-If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (that is, if causal features do not act as confounders for inactive features), then S=E, and we are retrieving $E (FE)^\dagger$.
-
-Under centred noise, and if (FN)'(FN) have low condition number (a weak assumption given that N is noise), we have $\left \| GFN\right \| ^2 \approx Var(FN) \left \| G\right \| ^2$, and minimisation of $\left \| X-GFEX\right\| ^2  + \left \| GFN\right \| ^2$ will recover of the minimum of the left norm, $\lambda S (FE)^\dagger$, $\lambda$ a positive number below 1.
-
-Replacing, we have $\left \| X-GFEX\right\| ^2 = \left \| X-\lambda S (FE)^\dagger FEX\right\| = \left \| X(I-\lambda S)\right\|$.
-If the covariance of X is block diagonal (no confounding between active and inactive features of X), this is equal to $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ (denoting the variance of X over active and inactive features).
-
-For the right part of the sum, we have $GFN=\lambda S (FE)^\dagger FN= \lambda S (FE)^\dagger (FE+F(I-E))N= \lambda S (E +(FE)^\dagger F(I-E))N=\lambda E N $ (FC last part of proof to be done, need some lemma on (FE)+).
-The square norm is therefore $\lambda^2 Var(N) (k/nx)$.
-
-The total quantity to be minimised, under noise, but with non confounding features in X, F full rank over E, and FN decently conditioned, is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N) (k/nx)$, setting the partial derivative in $\lambda$ to zero, we obtain $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N) (k/nx)}$, and if we suppose that X has comparable variance over all features, this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).
-
-The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} E (FE)^\dagger$ if active and inactive features of X are not correlated
-
-The second step is straightforward, replacing Y and since N is centred, we have
-\begin{equation}
-\begin{aligned}
-\arg \min_H \left \| \hat GY - HX \right \|^2 &=\arg  \min_H \left \| \hat GF(EX+N) - HX \right \|^2 \\
-&=\arg \min_H \left \| (\hat G(FE)-H)X \right \| ^2 + \left \| \hat GFN \right \| ^2\\
-&= \arg \min_H \left \| (\hat G(FE)-H)X \right \| ^2\\
-&=\hat G (FE)
-\end{aligned}
-\end{equation}
-
-Thus, in the linear case, back-and-forth regression retrieves $\hat H = \frac{1}{1+nsr} E (FE)^{\dagger}(FE) = \frac{1}{1+nsr} E$.
-
-\subsection{Asymptotic behaviour - non linear case}
-
-\section{Related works}
-\subsection{back-to-back regression as a special case of canonical component analysis?}
-Formula (1) shows that back-to-back regression may be related to canonical component analysis (CCA).
-%
-In CCA, we are given two matrices X and Y (or size N,p and N,q) that describe different features measured on the same data sample, and characterize their proximity (that is, between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.
-
-This is done by finding vectors $a$ and $b$ such that $Xa$ and $Yb$ display maximal correlation, or by maximizing $a'X'Yb$ over unit-normed $a$ and $b$.
-%
-To find a, one calculates the eigenvectors of Eq. (1).
-%
-The largest one is the direction with maximal correlation, the second one the maximal correlation direction orthogonal to the first, and so on.
-%
-For each canonical dimension, the corresponding eigenvalue is the square of the cosine between the corresponding direction along X and the subspace spanned by Y.
-
-If the result of Eq. (1) is diagonal, as in the model presented above, the features of X are eigenvectors.
-%
-If all eigenvalues are either 0 or 1, then all the features are either part of the Y subspace (eigenvalues of one) or orthogonal to it (for zero eigenvalues) because they correspond to the squared cosine of the angle between the corresponding feature and the subspace spanned by Y, .
-%
-Thus, back-to-back regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
-%
-The closer we are to this hypothesis, the closer the results of CCA and back-to-back regression will be.
-
-There are differences between CCA and back-to-back regression.
-%
-First, since our objective is not to characterise the proximity between X and Y, but just to extract causal features, we can dispense with the last step of CCA (diagonalisation of (1)).
-%
-Second, since we operate in a noisy environment, we use regularisation, introducing bias in the correlation calculations.
-%
-Finally, whereas CCA wants to leverage all correlations between X and Y, we filter some of them, through splitting and bagging.
-%
-In other words, whereas we are working from the same formulae as CCA, we use them in different ways, for different purposes
-
-\subsection{Related regression methods}
-FC: discuss similar methods of regression, that can be used for feature extraction
-And maybe have something about other causal discovery methods
-
-\section{Experiments - simulated data}
-Experiments on simulated data serve three purposes: compare back-to-back regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.
-
-The model used for simulation is $Y=f(snr MEX+N)$, with X a random (dx,N) matrix, E a square binary diagonal of dimension dx with the nc last elements equal to 1 (this means that of the dx features of X, only the nc last have some influence on Y), M and N (dz,dx) and (dz,N) random matrices, and f a function from $\mathbb{R}^{dz}$ to $\mathbb{R}^{dy}$, that can be linear or non linear.
-%
-In the linear case, the model becomes $Y=F(MEX+N)$, with F a (dy,dz) matrix.
-%
-In the non linear case, F becomes $FSH$ where F and H are matrices of dimensions (dz,dz) and (dy,dz), and S a vector of dz sigmoid transformations.
-%
-Finally, $snr$ is a real number measuring the amount of noise.
-
-Our simulation therefore depends upon 6 free parameters :
-\begin{itemize}
-\item dx : number of features of X
-\item nc: number of active features (in EX)
-\item dz: number of noise features
-\item snr: signal to noise ratio
-\item dy: number of features in Y
-\item N: size of sample
-\end{itemize}
-
-Random matrices X, M, N, H and F are populated with centred gaussian random coefficients with unit variance, divided by $ \surd n$ where n is the smallest dimension of the matrix.
-%
-Since a random square matrix with unit gaussian coefficients of dimension n has eigenvalues between $ \surd n $ and $\surd n $, such  scaling allows for meaningful comparisons when dimensions dx, dy and dz vary.
-
-In the non linear case, sigmoid functions are...
-FC: insert comment about scaling sigmoids, and maybe something about uniform coeffs
-
-We evaluate the performance of back-to-back regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
-%
-For feature extraction, the quality of fit is the number of disagreements between our estimator and the diagonal of E (ie, their Hamming distance, which should be minimised).
-%
-For regression, we use a test set of N additional measurements, and calculate the R-squared ratio of variances (which should be maximised).
-%
-Performance indicators are calculated by averaging the results of several experiments.
-
-\subsection{Feature extraction - linear case}
-All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by back-to-back regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
-%
-The decrease is usually very steep until a Hamming distance under 2 is achieved.
-%
-Figure 1 presents a number of such curves for varying values of dx=dy=dz, nc=30 and N=1000.
-
-Figure 1
-
-The distance between $\hat E$ and E incorporates two different errors: false positives, where features not in E are extracted, and false negatives where features in E are not extracted.
-%
-For low values of the snr, and typical values of nc (much smaller than dx), false positives are high, but they fall faster as snr increases.
-%
-False negatives begin lower, but decrease more slowly (figure 2).
-%
-In other words, for each values of the parameters, one may determine a value of the signal to noise ratio, such that snr higher than this threshold allow for good feature extraction.
-%
-At this level, all the errors are false negatives.
-
-Figure 2
-
-Working from a fixed sample of 1000, and 15 active features, and varying values of dx, dy, dz (from 50 to 200) and snr (from 0.2 to 1), we observe that, all other factors being equal, the quality of feature extraction and regression decrease when dx increases (X has more inactive features), dz increase (more noise dimensions) and snr decreases (less signal, more noise).
-%
-The dimensionality of Y plays a much smaller role, except for very small values of dy (less than 10).
-
-For feature extraction, back-to-back regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
-%
-The following graphs show the boundary between values of dx and dz that allow for features extractions (dx lower than boundary, or dz larger), for different values of dy (level curves on the graph), and different values of signal to noise ratio and number of active variables.
-
-FC: add graphs here, see whether it is better to have nc fixed, or a proportion of dx, note : part of this can be moved to appendix if lack of space
-
-For values of ...
-%
-in ..., an empirical formula can be provided.
-%
-back-to-back regression allows for good feature extraction if :
-
-\subsection{Feature extraction - nonlinear case}
-
-\subsection{Regression }
-For regression, we want to understand how feature extraction helps predict Y from X.
-%
-To this effect, we compare four approaches to regression:
-\begin{itemize}
-\item standard ridge regression of Y from X, where no features are extracted this is our baseline
-\item weighted ridge regression, of Y from $\hat H X$, where features of X are weighted according to the diagonal of $\hat H$
-\item filtered ridge regression, of Y from $\hat E X$, where active features are first extracted (as above), and Y is predicted from the extracted features only
-\item perfect filtered (oracle) ridge regression of Y from EX, where E is supposed known
-\end{itemize}
-Over a large set of parameters, we observe that filtered regression is on average better than weighted, which is on average better than standard regression.
-%
-This demonstrates that back-to-back regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.
-
-FC graphs go here
-
-The quality of regression is independent from dy (except for very small values), increase with snr, decreases with dx, and increases with dz if it is larger than dx
-FC graphs go here
-
-\subsection{Comparison with other methods}
-Maxime territory
-
-\subsection{Multiple environments}
-In this section, we perform the same tests as in the previous one, but the test set is taken from a different distribution of X than the training set.
-%
-In practice, we test the same trained predictors of Y over test samples generated from a distribution of X with a different covariance.
-
-The rationale behind such tests is that whereas back-to-back regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.
-
-FC: ideally, we'd like to assess the multiple env performance of jrr, and compare it with others.
-%
-Are multiple environments limited to regression analyses? How should environments be generated (not just variance of X, I expect)?
-
-\section{Experiments - real case}
-Jean Remi and Maxime territory
-
-\input{meg_exps.tex}
-
-\section{Conclusion}
-
-\section{Acknowlegements}
-Data were provided (in part) by the Donders Institute for Brain, Cognition and Behaviour.
-
-\clearpage
-\newpage
-
-\bibliographystyle{abbrvnat}
-\bibliography{paper}
-
-\section{Appendices}
-
-This should hold part of the tests, explanations on simulation, detailed results and stuff on meg data
-
-\end{document}

### 2019-05-22 paper/meg_exps.tex
externalize intro
cut in 80-spaced lines

@@
 Next, we apply our method to brain imaging data from the anonymized multimodal
-neuroimaging ``Mother Of all Unification Studies'' (MOUS) dataset \cite{schoffelen2019204}.
-The dataset contains magneto-encephalography (MEG) recordings of 102 healthy native-Dutch
-adults who participated in a reading task.
+neuroimaging ``Mother Of all Unification Studies'' (MOUS) dataset
+\cite{schoffelen2019204}.  The dataset contains magneto-encephalography (MEG)
+recordings of 102 healthy native-Dutch adults who participated in a reading
+task.
 
-Subjects were exposed to a rapid serial visual presentation of Dutch words. The word lists
-consisted of 120 sentences, and scrambled lists of the same words. Each word was presented on the computer screen for 351ms on average (min: 300ms, max: 1400ms).
-Successive words were separated by a blank screen for 300ms, and successive
-sentences were separated by an empty screen for a few (3-4) seconds.
+Subjects were exposed to a rapid serial visual presentation of Dutch words. The
+word lists consisted of 120 sentences, and scrambled lists of the same words.
+Each word was presented on the computer screen for 351ms on average (min: 300ms,
+max: 1400ms).  Successive words were separated by a blank screen for 300ms, and
+successive sentences were separated by an empty screen for a few (3-4) seconds.
 
@@
 
-The raw MEG data was bandpass-filtered between 0.1 and 40 Hz using MNE-Python default
-parameters \cite{mne}. Specifically, we used a zero-phase finite impulse response filter (FIR) with a hamming window and with transition bands of 0.1 Hz and 10 Hz for the low and high
-cut-off frequencies respectively.
+The raw MEG data was bandpass-filtered between 0.1 and 40 Hz using MNE-Python
+default parameters \cite{mne}. Specifically, we used a zero-phase finite impulse
+response filter (FIR) with a hamming window and with transition bands of 0.1 Hz
+and 10 Hz for the low and high cut-off frequencies respectively.
 
 The raw data was then segmented 100 ms before word onset and 1.000 sec after
-word onset.
-Time $t=0$ms corresponds to the word onset.
-Finally, each resulting segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and thus led a sampling frequency of 240 Hz.
-We show, in Figure \ref{fig:megavg}, the average response of the
-magnetometers evoked by the words presented to the first subject.
+word onset. Time $t=0$ms corresponds to the word onset. Finally, each resulting
+segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
+thus led a sampling frequency of 240 Hz. We show, in Figure \ref{fig:megavg},
+the average response of the magnetometers evoked by the words presented to the
+first subject.
 
-For each word, each subject, and each time sample relative to word onset, we build an observation matrix $Y \in \mathbb{R}^{n, q}$ of $n\approx$ 2700 words by $q=301$ MEG channels (273 magnetometers and 28 compensation channels).
-Each of the columns of $Y$ is normalized to have zero mean and unit variance.
+For each word, each subject, and each time sample relative to word onset, we
+build an observation matrix $Y \in \mathbb{R}^{n, q}$ of $n\approx$ 2700 words
+by $q=301$ MEG channels (273 magnetometers and 28 compensation channels). Each
+of the columns of $Y$ is normalized to have zero mean and unit variance.
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth, trim=1.5cm 1cm 0.5cm 0cm, clip=True]{figure_meg.pdf}
-  \caption{Average magnetosensor response across words for one subject.}
-  \label{fig:megavg}
-\end{figure}
+\begin{figure} \centering \includegraphics[width=\textwidth, trim=1.5cm 1cm
+0.5cm 0cm, clip=True]{figure_meg.pdf} \caption{Average magnetosensor response
+across words for one subject.} \label{fig:megavg} \end{figure}
 
-\subsection{Word features}
-We annotated each word with 38 features: 26 corresponds to the number of each letter in the word (from 'a' to 'z', independently of their case), the total number of letters ('word length'), the frequency of the word in natural language (using the Zipf logarithmic scale \cite{marc_brysbaert}), and the part-of-speech tags (i.e.
-NOUN, PROPN (proper noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ (adjective), ADV (adverb), and NUM (numerical)) using the packages Spacy \cite{spacy} and WordFreq \cite{speer}.
-For clarity purposes, we refer to these features using two broad categories: visual (i.e.
-each letter and the total number of letters) and lexical (part-of-speech and word frequency).
-This procedure yields an $X \in \mathbb{R}^{n, d}$ matrix of $n\approx$ 2700 words by $d=38$ features for each subject.
-Each of the columns of $X$ is normalized to have a mean and a standard deviation of 0 and 1 respectively.
+\subsection{Word features} We annotated each word with 38 features: 26
+corresponds to the number of each letter in the word (from 'a' to 'z',
+independently of their case), the total number of letters ('word length'), the
+frequency of the word in natural language (using the Zipf logarithmic scale
+\cite{marc_brysbaert}), and the part-of-speech tags (i.e. NOUN, PROPN (proper
+noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ
+(adjective), ADV (adverb), and NUM (numerical)) using the packages Spacy
+\cite{spacy} and WordFreq \cite{speer}. For clarity purposes, we refer to these
+features using two broad categories: visual (i.e. each letter and the total
+number of letters) and lexical (part-of-speech and word frequency).  This
+procedure yields an $X \in \mathbb{R}^{n, d}$ matrix of $n\approx$ 2700 words by
+$d=38$ features for each subject. Each of the columns of $X$ is normalized to
+have a mean and a standard deviation of 0 and 1 respectively.
 
-\subsection{Analyses}
-We aimed to identify which factors of $X$ explain the MEG recordings $Y$ at each time sample and for each subject separately.
-To this aim, we implemented three types of analyses (forward, backward and back-to-back regressions), and tested whether their results were consistent across subjects, using a second-level statistics described below.
+\subsection{Analyses} We aimed to identify which factors of $X$ explain the MEG
+recordings $Y$ at each time sample and for each subject separately. To this aim,
+we implemented three types of analyses (forward, backward and back-to-back
+regressions), and tested whether their results were consistent across subjects,
+using a second-level statistics described below.
 
-Forward regression, also known as 'encoding' in neuroimaging, here consisted in fitting a Ridge-regularized OLS from $X$ to $Y$:
-\begin{equation}
-H = (X^{T}X+\lambda I)^{-1} X^{T}Y
-\end{equation}
-where $\lambda$ is the regularization parameter fit with the leave-one-out cross-validation implemented in scikit-learn \cite{sklearn} from 20 possible value logarithmatically distributed between $10^{-5}$ and $10^5$.
-We aimed to estimate the contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor.
-However, $H\in\mathbb{R}^{d, q}$ is a matrix of $d$ word features by $q$ MEG channels.
-We thus approximated the relative contribution of each feature by sum-squarring the H coefficients:
-\begin{equation}
-\hat e_d = \sum_{i=0}^{i=q} {h_d^i}^2
-\end{equation}
+Forward regression, also known as 'encoding' in neuroimaging, here consisted in
+fitting a Ridge-regularized OLS from $X$ to $Y$: \begin{equation} H =
+(X^{T}X+\lambda I)^{-1} X^{T}Y \end{equation} where $\lambda$ is the
+regularization parameter fit with the leave-one-out cross-validation implemented
+in scikit-learn \cite{sklearn} from 20 possible value logarithmatically
+distributed between $10^{-5}$ and $10^5$. We aimed to estimate the contribution
+$diag(\hat E) \in\mathbb{D}^{d} $ of each factor. However, $H\in\mathbb{R}^{d,
+q}$ is a matrix of $d$ word features by $q$ MEG channels. We thus approximated
+the relative contribution of each feature by sum-squarring the H coefficients:
+\begin{equation} \hat e_d = \sum_{i=0}^{i=q} {h_d^i}^2 \end{equation}
 
-Backward regression, also known as 'decoding' in neuroimaging, here consisted in fitting a Ridge-regularized OLS from $Y$ to $X$:
+Backward regression, also known as 'decoding' in neuroimaging, here consisted in
+fitting a Ridge-regularized OLS from $Y$ to $X$:
 
-\begin{equation}
-G = (Y^{T}Y+\lambda I)^{-1} Y^{T}X
-\end{equation}
+\begin{equation} G = (Y^{T}Y+\lambda I)^{-1} Y^{T}X \end{equation}
 
-where $\lambda$ is the regularization parameter fit with the leave-one-out cross-validation
-implemented in scikit-learn \cite{sklearn} from 20 possible value logarithmically distributed
-between $10^{-5}$ and $10^5$. We aimed to estimate the contribution $diag(\hat E)
-\in\mathbb{D}^{d} $ of each factor. Using 10-fold cross-validation, we thus assessed the ability
-of this ridge regression to predict independent and identically distributed data:
+where $\lambda$ is the regularization parameter fit with the leave-one-out
+cross-validation implemented in scikit-learn \cite{sklearn} from 20 possible
+value logarithmically distributed between $10^{-5}$ and $10^5$. We aimed to
+estimate the contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor.
+Using 10-fold cross-validation, we thus assessed the ability of this ridge
+regression to predict independent and identically distributed data:
 
-\begin{equation}
-\hat X_{test} = Y_{test}\cdot G_{train}
-\end{equation}
+\begin{equation} \hat X_{test} = Y_{test}\cdot G_{train} \end{equation}
 
-and assessing, for each factor of X independently, the Pearson correlation coefficient $r$ between $X_{test}$ and $\hat X_{test}$ for each cross-validation split.
+and assessing, for each factor of X independently, the Pearson correlation
+coefficient $r$ between $X_{test}$ and $\hat X_{test}$ for each cross-validation
+split.
 
-Finally, back-to-back regression was fit as decribed earlier using two additional stability tricks designed to minimize the impact of bagging.
-First the $G$ ridge regression was broken-down with SVD blablabla
-Second, $X$ was regularized using a maximal likelihood blablabla
-Third, no regularization
+Finally, back-to-back regression was fit as decribed earlier using two
+additional stability tricks designed to minimize the impact of bagging.  First
+the $G$ ridge regression was broken-down with SVD blablabla Second, $X$ was
+regularized using a maximal likelihood blablabla Third, no regularization
 
@@
 
-The three types of analyses were implemented for each subject and each time sample independently.
-To estimate the reliability of these estimates in the population, we tested, using one-sample t-test whether the distribution of the estimated contribution of each factor different from baseline ($t=-100$ms):
+The three types of analyses were implemented for each subject and each time
+sample independently.  To estimate the reliability of these estimates in the
+population, we tested, using one-sample t-test whether the distribution of the
+estimated contribution of each factor different from baseline ($t=-100$ms):
 
-\begin{equation}
-$ttest(\hat E_{t=-100} - \hat E_{\tau}$
-\end{equation}
-with $\tau \in $ 150ms and 400ms respectively
+\begin{equation} ttest(\hat E_{t=-100} - \hat E_{\tau}) \end{equation} with
+$\tau \in $ 150ms and 400ms respectively
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth, trim=1.5cm 1cm 0.5cm 0cm,
-clip=True]{meg_result.pdf}
-  \caption{Average magnetosensor response across words for one subject.}
-  \label{fig:megresult}
-\end{figure}
+\begin{figure} \centering \includegraphics[width=\textwidth, trim=1.5cm 1cm
+0.5cm 0cm, clip=True]{meg_result.pdf} \caption{Average magnetosensor response
+across words for one subject.} \label{fig:megresult} \end{figure}
 

### 2019-05-22 code/synthetic_experiment/plot.py
Plot

@@ -1,136 +1,99 @@
 import argparse
 import numpy as np
-
 from ast import literal_eval
 from matplotlib import pyplot as plt
 
+plt.rcParams["font.family"] = "Times New Roman"
+
 stats = {
-    "result_error_in_all": "error in-domain",
-    "result_error_out_all": "error out-domain",
-    "result_error_in_mask": "error in-domain (masked)",
-    "result_error_out_mask": "error out-domain (masked)",
+    "result_error_in_all": "error in",
+    "result_error_out_all": "error out",
+    "result_error_in_mask": "error in (E)",
+    "result_error_out_mask": "error out (E)",
     "result_false_positives": "false positives",
     "result_false_negatives": "false negatives",
-    "result_response_active": "response at active",
-    "result_response_inactive": "response at inactive",
+    # "result_response_active": "response at active",
+    # "result_response_inactive": "response at inactive",
     "result_auc": "AUC"
 }
 
-# plt.rc('text', usetex=True)
-# plt.rc('text.latex', preamble=r'\usepackage{times}')
-# plt.rc('font', family='serif')
-plt.rc('font', size=10)
+models = {
+    "Ridge" : "C1",
+    "Lasso" : "C2",
+    "PLS" : "C3",
+    "CCA" : "C4",
+    "RRR" : "C5"
+}
 
 if __name__ == "__main__":
     parser = argparse.ArgumentParser(description='JRR synthetic experiment')
-    parser.add_argument('--file', type=str, default='results_all_jrr.txt')
+    parser.add_argument('--file', type=str, default='synthetic_result.txt')
     args = parser.parse_args()
 
     with open(args.file, "r") as f:
         lines = f.readlines()
 
-    result_names = []
-    first_result = literal_eval(lines[0])
-
-    for k in first_result.keys():
-        if "result" in k:
-            result_names.append(k)
-
     results = {}
-    models = set()
 
     for line in lines:
-        result_dict = {}
         line_dict = literal_eval(line)
-        line_str = ""
-        for k, v in line_dict.items():
-            if ("result" not in k) and ("model" not in k) and ("seed" not in k):
-                line_str += f"{k}={v}_"
-            elif k == "model":
-                this_model = v
-                models.add(v)
-            elif k == "seed":
-                this_seed = int(v)
-            else:
-                result_dict[k] = float(v)
 
-        line_str = line_str[:-1]
+        eid = "{}_{}_{}_{}_{}_{}_{}".format(line_dict["n_samples"],
[…]

### 2019-05-22 paper/synthetic.tex
Taking out synthetic.tex

--- 2019-05-22 paper/paper.tex
@@
-\documentclass{article}
-
-\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx}
-\usepackage[hidelinks]{hyperref}
-\usepackage[]{neurips_2019}
-\DeclareMathOperator{\Tr}{Tr}
-
-\usetikzlibrary{calc}
-
-\tikzset{
-    ncbar angle/.initial=90,
-    ncbar/.style={
-        to path=(\tikztostart)
-        -- ($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)
-        -- ($(\tikztotarget)!($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztostart)$)
-        -- (\tikztotarget)
-    },
-    ncbar/.default=0.5cm,
-}
-
-\tikzset{round left paren/.style={ncbar=0.5cm,out=110,in=-110}}
-\tikzset{round right paren/.style={ncbar=0.5cm,out=70,in=-70}}
-
-\newcommand{\dlp}[1]{{\color{red} (DLP: #1)}}
-\newcommand{\mo}[1]{{\color{green} (MO: #1)}}
-\newcommand{\jrk}[1]{{\color{blue} (JRK: #1)}}
-\newcommand{\fc}[1]{{\color{orange} (FC: #1)}}
-
-\title{Discovering causal factors with back-to-back regression}
-
-\author{%
-  Jean-Remi King\\
-  CNRS\\
-  \texttt{email} \\
-  \And
-  Fran\c{c}ois Charton\\
-  Facebook AI\\
-  \texttt{fcharton@fb.com}\\
-  \And
-  David Lopez-Paz\\
-  Facebook AI\\
-  \texttt{dlp@fb.com}
-  \And
-  Maxime Oquab\\
-  Facebook AI\\
-  \texttt{qas@fb.com}
-}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-Identifying causes from observations is at the core of science. This endeavor
-is particularly challenging when i) potential factors are difficult to
-manipulate individually and ii) observations are complex and multi-dimensional.
-To address this issue, we introduce ``back-to-back regression'' (BFR), a
-method designed to identify, from a set of co-varying factors, the factors
-that most plausibley account for multidimensional observations. After detailing
-the proof of convergence, we show that BFR outperforms least-squares and related
-regression techniques, as well as cross-decomposition methods (e.g. canonical
-correlation analysis, and partial least squares) on two tasks: causal
-identification and out-of-distribution prediction. Finally, we apply BFR to
-neuroimaging recordings of 102 subjects reading word sequences. The results
-show, as expected, that the early brain responses to words appear to be
-specifically caused by low-level visual factors whereas late brain responses
-appear to be caused by lexical factors, despite the fact that these factors
-covary in the study.
-% conclude  open
-\end{abstract}
-
-\section{Introduction}
-
-\input{introduction.tex}
-
-\section{Back an forth regression}
-
-\begin{figure}[t!]
-    \centering
-    \begin{tikzpicture}
-    \newcommand\posY{0}
-    \newcommand\posX{3}
-    \newcommand\posE{5}
-    \newcommand\posN{7}
-    \newcommand\posF{9.5}
-
-    \node[thick, draw=black, minimum height=3cm, minimum width=2cm, fill=yellow!10] (Y) at (\posY, 0){};
-    \node[] (eq) at (1.5, 0){$=$};
-    \node[] (times) at (4, 0){$\times$};
-    \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=yellow!10] (X) at (\posX, 0){};
-    \node[thick, draw=black, minimum height=1cm, minimum width=1cm, fill=red!10] (E) at (\posE, 0){};
-
-    \draw[fill=white] (\posE - 0.5 + 0.0, 0.5 - 0.0) rectangle (\posE - 0.5 + 0.0 + 0.1, 0.5 - 0.0 - 0.1);
-    \draw[fill=black] (\posE - 0.5 + 0.1, 0.5 - 0.1) rectangle (\posE - 0.5 + 0.1 + 0.1, 0.5 - 0.1 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.2, 0.5 - 0.2) rectangle (\posE - 0.5 + 0.2 + 0.1, 0.5 - 0.2 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.3, 0.5 - 0.3) rectangle (\posE - 0.5 + 0.3 + 0.1, 0.5 - 0.3 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.4, 0.5 - 0.4) rectangle (\posE - 0.5 + 0.4 + 0.1, 0.5 - 0.4 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.5, 0.5 - 0.5) rectangle (\posE - 0.5 + 0.5 + 0.1, 0.5 - 0.5 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.6, 0.5 - 0.6) rectangle (\posE - 0.5 + 0.6 + 0.1, 0.5 - 0.6 - 0.1);
-    \draw[fill=black] (\posE - 0.5 + 0.7, 0.5 - 0.7) rectangle (\posE - 0.5 + 0.7 + 0.1, 0.5 - 0.7 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.8, 0.5 - 0.8) rectangle (\posE - 0.5 + 0.8 + 0.1, 0.5 - 0.8 - 0.1);
-    \draw[fill=black] (\posE - 0.5 + 0.9, 0.5 - 0.9) rectangle (\posE - 0.5 + 0.9 + 0.1, 0.5 - 0.9 - 0.1);
-
-    \node[] (plus) at (6, 0){$+$};
-    \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=blue!10] (N) at (\posN, 0){};
-    \node[thick, draw=black, minimum height=2cm, minimum width=1cm, fill=red!10] (F) at (\posF, 0){};
-    \draw [thick] (2.25, -1.5) to [round left paren ] (2.25, 1.5);
-    \draw [thick] (7.75, -1.5) to [round right paren ] (7.75, 1.5);
-
-    \node[] (times2) at (8.5, 0){$\times$};
-
-    \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{n \times q}$}};
-    \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{n \times d}$}};
-    \node[] (annE) at (\posE, -1.8){\scalebox{0.85}{$E \in \mathbb{D}^{d \times d}$}};
-    \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{n \times d}$}};
-    \node[] (annF) at (\posF, -1.8){\scalebox{0.85}{$F \in \mathbb{R}^{d \times q}$}};
-
-    \node[] (labY) at (\posY, 2){Observations};
-    \node[] (labX) at (\posX, 2){Factors};
-    \node[] (labE) at (\posE, 2.3){Cause};
-    \node[] (labE) at (\posE, 2){selection};
-    \node[] (labN) at (\posN, 2){Noise};
-    \node[] (labF) at (\posF, 2.3){Cause-effect};
-    \node[] (labF) at (\posF, 2){mapping};
-
-    \node[] (sim1) at (0,-3.25) {\scalebox{0.85}{$X \sim P(X)$}};
-    \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
-    \node[] (reg1) at (5,-3.25) {$\hat{E} = \underbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}} $};
-    \end{tikzpicture}
-    \caption{In its linear form, back-to-back regression identifies the subset of factors that suffices to account for multidimensional observations by piping two successive regressions from $Y$ to $X$ and from $X$ to $\hat X$.
-}
-    \label{fig:}
-\end{figure}
-
-\subsection{Algorithm}
-
-Back-to-back regression consists of two steps.
-%
-First, we perform a least squares estimation of X from Y, that is we find $\hat G$ such that $\hat X=\hat G Y$ is the least square predictor of X.
-%
-This recovers an approximation of the causes of Y in X.
-%
-Then, we perform a least squares estimation of $\hat X$ from X, and determine $\hat H$ such that $\hat {\hat X}=\hat H X$ is the least square predictor of $\hat X$.
-%
-In matrix form, we have $\hat G=(Y'Y)^{-1} Y'X$, and
-\begin{equation} \hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X\end{equation}
-Under good conditions (close to linear dependence between X and Y, and covariance matrices of X and Y well conditioned), this recovers a scaled approximation to E.
-
-In practice, the covariance matrices of Y and X might exhibit multicollinearity and be badly conditioned, which make their inversion unstable.
-To guard against this, we compute their singular value decompositions, and truncate their lowest eigenvalues so as to reach a specific condition number in the Frobenius norm.
-We prefer the truncated SVD approach over the more common regularisation (ridge regression) because it introduces less bias in the estimation.
-A small amount of regularisation is still used, for numerical stability purposes.
-
-If $X'X=U'DU$, U orthogonal and D diagonal, and D* is D with the lowest diagonal elements set to zero, this amounts to replacing $(X'X)^{-1}$ by $U(D*^{-1}+\epsilon I)U'$, with $\epsilon$ a positive number close to machine floating point accuracy.
-D is truncated so as to have a certain condition number that is determined by leave one out cross validation.
-%
-Also, to prevent spurious correlations between the two steps (should they be performed on the same sample), our original sample is randomly split into two halves, and each step is performed over one of them, effectively decorrelating the two steps.
-%
-This is repeated over several (100) splits, and the resulting $\hat H$ are averaged.
-%
-This bagging compensates for the reduction in sample size caused by splitting.
-
-Under our hypotheses (centered noise, E a binary diagonal) and even under heavy noise, $\hat H$ is close to diagonal.
-%
-Yet, because of regularisation which adds positive bias to small eigenvalues, and noise which attenuates large eigenvalues, $\hat H$ is in fact a scaled and noisy approximation to E.
-%
-To recover a binary diagonal, as our estimate of E, we extract the diagonal of $\hat H$ and select its large elements.
-%
-The diagonal matrix $\hat E$, having large elements of H as its unit diagonal coefficients (the rest being zero) is our estimate of E, the active features of X.
-
-To select large elements of $\hat H$, the traditional approach in regression analysis consists in testing whether its coefficients differ from zero.
-This is usually done via variants of Student t-test.
-%
-This will not work here, as back-to-back regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
-%
-Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).
-
-This can be done by maximizing the ratio of inter-group variance over total variance (ie minimizing intra-group variance).
-%
-In a one dimensional two clusters setting, this amounts to sorting the data, and separating the p smallest from the n-p largest values so that inter group inertia is maximal.
-%
-If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we are maximizing the Sonquist and Morgan criterion $$K = {p (n-p) \over n} {(s-r)^2 \over V}$$ and selecting the p features corresponding to the largest values.
-%
-If diagonal elements of $\hat H$ are normally distributed, K follows a chi-square distribution with one degree of freedom.
-%
-At 95\% confidence level, a value of K superior to 3.84 makes our selection significant (for 99\% confidence, the corresponding value of K is 6.63) \citep{Kass_75}
-
-This extract from $\hat H$ a binary diagonal $\hat E$.
-%
-To use back-to-back regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
-
-Summarizing, back-to-back regression is performed as follows:
-\begin{enumerate}
-\item Randomly split the data sample.
-\item From the first half sample, using crossalidated ridge regression, find $\hat G$, the regularised least square regression of X from Y: $\hat G=(Y'Y)^{-1} Y'X$.
-\item From the second half sample, calculate $\hat X = \hat G Y$, and find $\hat H$, the regularized least square regression of $\hat X$ from X: $\hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X$.
-\item Repeat steps 1 to 3 and average the results as $\hat H$.
-\item Using the Sonquist and Morgan criterion, select the large diagonal elements of $\hat H$, set them to one and the rest of $\hat  H$ to zero.
-\item Use the resulting matrix $\hat E$ as an estimator of E, the active features in the causal process
-\item Alternatively, use $\hat E$ to predict Y from X, by performing a regularised least square regression of Y from $\hat E X$.
-%
-
-\end{enumerate}
-
-\subsection{Asymptotic behaviour - linear case}
-As discussed in the introduction, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(EX+N)$, with f an unknown function.
-%
-If F is linear we can rewrite it as $Y = F(EX + N)$, with X an (dx, n) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
-%
-N a (dx,n) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dx) transformation corresponding to the measuring apparatus, and Y a (dy, n) matrix of measured effects.
-
-The two steps of back-to-back regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).
-
-For the first step, replacing Y, and since N is centred, we have
-\begin{equation}
-\begin{aligned}
-\arg \min_G \left \| X-GY \right \|^2 &= \arg \min_G \left \| X - GF(EX+N)\right\|^2 \\
-&{}= \arg \min_G \left \| X-GFEX\right\| ^2  + \left \| GFN\right \| ^2
-\end{aligned}
-\end{equation}
-
-As N tends to zero, we are searching for the minimal norm solution of $\min_G \left \| X-GFEX\right\| ^2$, which is $X(FEX)^\dagger$ ($M^\dagger$ denoting the pseudo inverse of M).
-
-Let k be the number of active features of E, suppose for notational convenience that they are the first k features of X, and let $FEX$ be of rank k (which means F and X are of full rank on the space spanned by the active features of E).
-Let K be the (nx,k) diagonal matrix built by deleting the rightmost nx-k columns of E.
-Then, $K=(\frac{I_{k}}{0})$.
-we may write $E=KK'$, and $FEX=(FK)(K'X)$, with $FK$  and $K'X$ of full rank k.
-
-The rank factorisation property of the pseudo inverse (ref) says that, if $A=BC$, A (m,n), B(m,k) or rank k C(k,n) of rank k, then $A^\dagger=C^\dagger B^\dagger= C' (CC')^{-1}(B'B)^{-1}B'$.
-
-Replacing in the expression of the minimal norm solution yields
-\begin{equation}
-\begin{aligned}
-X(FEX)^\dagger &=X((FK)(K'X))^\dagger \\
-&=X(K'X)'((K'X)(K'X)')^{-1}((FK)'(FK))^{-1}(FK)' \\
-&=(XX')K(K'XX'K)^{-1}(K'F'FK)^{-1} K'F'
-\end{aligned}
-\end{equation}
-
-Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{3} & \Sigma_{4}\end{array}\right)$, the top left block being (k,k), since $K=\left(\begin{array}{c}I \\\hline 0\end{array}\right)$, we have $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{3} \Sigma_{1}^{-1}\end{array}\right)$, call S the matrix obtained by adding Nx-k empty columns to this one (or, equivalently multiplying it by K').
-
-Let us also note that $FE=FKK'$, and so $(K'F'FK)^{-1} K'F'FKK'=I_{k}K'$. Therefore, we have $(K'F'FK)^{-1} K'F'=(FE)^\dagger$.
-
-Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=X(FEX)^\dagger =S (FE)^\dagger$.
-If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (that is, if causal features do not act as confounders for inactive features), then S=E, and we are retrieving $E (FE)^\dagger$.
-
-Under centred noise, and if (FN)'(FN) have low condition number (a weak assumption given that N is noise), we have $\left \| GFN\right \| ^2 \approx Var(FN) \left \| G\right \| ^2$, and minimisation of $\left \| X-GFEX\right\| ^2  + \left \| GFN\right \| ^2$ will recover of the minimum of the left norm, $\lambda S (FE)^\dagger$, $\lambda$ a positive number below 1.
-
-Replacing, we have $\left \| X-GFEX\right\| ^2 = \left \| X-\lambda S (FE)^\dagger FEX\right\| = \left \| X(I-\lambda S)\right\|$.
-If the covariance of X is block diagonal (no confounding between active and inactive features of X), this is equal to $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ (denoting the variance of X over active and inactive features).
-
-For the right part of the sum, we have $GFN=\lambda S (FE)^\dagger FN= \lambda S (FE)^\dagger (FE+F(I-E))N= \lambda S (E +(FE)^\dagger F(I-E))N=\lambda E N $ (FC last part of proof to be done, need some lemma on (FE)+).
-The square norm is therefore $\lambda^2 Var(N) (k/nx)$.
-
-The total quantity to be minimised, under noise, but with non confounding features in X, F full rank over E, and FN decently conditioned, is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N) (k/nx)$, setting the partial derivative in $\lambda$ to zero, we obtain $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N) (k/nx)}$, and if we suppose that X has comparable variance over all features, this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).
-
-The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} E (FE)^\dagger$ if active and inactive features of X are not correlated
-
-The second step is straightforward, replacing Y and since N is centred, we have
-\begin{equation}
-\begin{aligned}
-\arg \min_H \left \| \hat GY - HX \right \|^2 &=\arg  \min_H \left \| \hat GF(EX+N) - HX \right \|^2 \\
-&=\arg \min_H \left \| (\hat G(FE)-H)X \right \| ^2 + \left \| \hat GFN \right \| ^2\\
-&= \arg \min_H \left \| (\hat G(FE)-H)X \right \| ^2\\
-&=\hat G (FE)
-\end{aligned}
-\end{equation}
-
-Thus, in the linear case, back-and-forth regression retrieves $\hat H = \frac{1}{1+nsr} E (FE)^{\dagger}(FE) = \frac{1}{1+nsr} E$.
-
-\subsection{Asymptotic behaviour - non linear case}
-
-\section{Related works}
-\subsection{back-to-back regression as a special case of canonical component analysis?}
-Formula (1) shows that back-to-back regression may be related to canonical component analysis (CCA).
-%
-In CCA, we are given two matrices X and Y (or size N,p and N,q) that describe different features measured on the same data sample, and characterize their proximity (that is, between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.
-
-This is done by finding vectors $a$ and $b$ such that $Xa$ and $Yb$ display maximal correlation, or by maximizing $a'X'Yb$ over unit-normed $a$ and $b$.
-%
-To find a, one calculates the eigenvectors of Eq. (1).
-%
-The largest one is the direction with maximal correlation, the second one the maximal correlation direction orthogonal to the first, and so on.
-%
-For each canonical dimension, the corresponding eigenvalue is the square of the cosine between the corresponding direction along X and the subspace spanned by Y.
-
-If the result of Eq. (1) is diagonal, as in the model presented above, the features of X are eigenvectors.
-%
-If all eigenvalues are either 0 or 1, then all the features are either part of the Y subspace (eigenvalues of one) or orthogonal to it (for zero eigenvalues) because they correspond to the squared cosine of the angle between the corresponding feature and the subspace spanned by Y, .
-%
-Thus, back-to-back regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
-%
-The closer we are to this hypothesis, the closer the results of CCA and back-to-back regression will be.
-
-There are differences between CCA and back-to-back regression.
-%
-First, since our objective is not to characterise the proximity between X and Y, but just to extract causal features, we can dispense with the last step of CCA (diagonalisation of (1)).
-%
-Second, since we operate in a noisy environment, we use regularisation, introducing bias in the correlation calculations.
-%
-Finally, whereas CCA wants to leverage all correlations between X and Y, we filter some of them, through splitting and bagging.
-%
-In other words, whereas we are working from the same formulae as CCA, we use them in different ways, for different purposes
-
-\subsection{Related regression methods}
-FC: discuss similar methods of regression, that can be used for feature extraction
-And maybe have something about other causal discovery methods
-
-\section{Experiments - simulated data}
 Experiments on simulated data serve three purposes: compare back-to-back regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.
@@
 Are multiple environments limited to regression analyses? How should environments be generated (not just variance of X, I expect)?
-
-\section{Experiments - real case}
-Jean Remi and Maxime territory
-
-\input{meg_exps.tex}
-
-\section{Conclusion}
-
-\section{Acknowlegements}
-Data were provided (in part) by the Donders Institute for Brain, Cognition and Behaviour.
-
-\clearpage
-\newpage
-
-\bibliographystyle{abbrvnat}
-\bibliography{paper}
-
-\section{Appendices}
-
-This should hold part of the tests, explanations on simulation, detailed results and stuff on meg data
-
-\end{document}

### 2019-05-22 code/synthetic_experiment/plot.py
Better plot

@@ -28,6 +28,7 @@ models = {
 if __name__ == "__main__":
     parser = argparse.ArgumentParser(description='JRR synthetic experiment')
     parser.add_argument('--file', type=str, default='synthetic_result.txt')
+    parser.add_argument('--nonlinear', type=int, default=1)
     args = parser.parse_args()
 
     with open(args.file, "r") as f:
@@ -37,6 +38,9 @@ if __name__ == "__main__":
 
     for line in lines:
         line_dict = literal_eval(line)
+
+        # if int(line_dict["nonlinear"]) != args.nonlinear:
+        #     continue
 
         eid = "{}_{}_{}_{}_{}_{}_{}".format(line_dict["n_samples"],
                                             line_dict["dim_x"],
@@ -71,12 +75,21 @@ if __name__ == "__main__":
                 plt.ylabel(model)
             data_x = []
             data_y = []
+            data_c = []
             if model != "JRR":
                 for eid in results:
-                    if model in results[eid]:
+                    if model in results[eid] and "JRR" in results[eid]:
                         data_x.append(np.mean(results[eid]["JRR"][stat]))
                         data_y.append(np.mean(results[eid][model][stat]))
-            plt.scatter(data_x, data_y, c="black", alpha=0.25, s=7)#, rasterized=True)
+
+                        is_nonlinear = int(eid.split("_")[5])
+
+                        if is_nonlinear == 0:
+                            data_c.append("C0")
+                        elif is_nonlinear == 1:
+                            data_c.append("C1")
+
+            plt.scatter(data_x, data_y, c=data_c, alpha=0.25, s=7)#, rasterized=True)
             xl = ax.get_xlim()
             yl = ax.get_ylim()
 

### 2019-05-22 paper/paper.tex
method written

--- 2019-05-22 paper/paper.tex
@@
 
+\subsection{Formalism}
+Via an apparatus F, we are measuring a multidimensional signal Y, which is a function of possible causes X and other non causal features N.
+Not all features of X exert causal influence on Y.
+Let EX (E a matrix of zero and one) the subset of causal factors.
+Our model can be formalised as $Y=f(EX,N)$.
+
+Without loss of generality, we may assume N to be centred(adding, when necessary, an intercept to the causal features), and redefine features and noise so that the model be written $Y=f(EX+N)$ (f a function, E, X and N matrices).
+In particular, the addition of "external noise" M, as in $Y=F(EX+N)+M$ can be viewed as a small change in F and N.
+
+We consider both the linear and non linear case.
+In the first one, our model is $$Y = F(EX+N)$$, with  $Y\in\mathbb{R}^{n\times ny}$, $F\in\mathbb{R}^{ny\times nx}$, $X\in\mathbb{R}^{nx\times n}$, $N\in\mathbb{R}^{nx\times n}$, and E a binary diagonal square matrice of size nx.
+
+In the non linear model, $F(EX+N)$ is defined as before, but Y is the result of a non linear transformation of it $$Y=\sigma (F(EX+N))$$
+
+Under this formalism, we are given samples of Y and X, and are tasked to recover E and use it to predict Y from X.
+
 \subsection{Algorithm}
@@
 \begin{equation} \hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X\end{equation}
-Under good conditions (close to linear dependence between X and Y, and covariance matrices of X and Y well conditioned), this recovers a scaled approximation to E.
 
-In practice, the covariance matrices of Y and X might exhibit multicollinearity and be badly conditioned, which make their inversion unstable.
-To guard against this, we compute their singular value decompositions, and truncate their lowest eigenvalues so as to reach a specific condition number in the Frobenius norm.
-We prefer the truncated SVD approach over the more common regularisation (ridge regression) because it introduces less bias in the estimation.
-A small amount of regularisation is still used, for numerical stability purposes.
+If, in the above formalism, F is invertible with respect to the causal features of X (ie the problem has a solution), and the relation is linear or close to linear, the diagonal of $\hat H$ will be a scaled approximation of the diagonal of E.
+That is to say, back to back regression will recover the causal features of the process considered, even in the presence of heavy noise N, and without needing to estimate the transformation F.
 
-If $X'X=U'DU$, U orthogonal and D diagonal, and D* is D with the lowest diagonal elements set to zero, this amounts to replacing $(X'X)^{-1}$ by $U(D*^{-1}+\epsilon I)U'$, with $\epsilon$ a positive number close to machine floating point accuracy.
-D is truncated so as to have a certain condition number that is determined by leave one out cross validation.
+In practical cases, the covariance matrices of Y and X will often be badly conditioned (because features of X are related, or measurement produces correlated features of Y).
+This makes the inversion of $X'X$ and $Y'Y$ sensitive to noise (either from N, from sampling or from computation rounding).
+To guard against high condition numbers, we have used two methods: truncated SVD (ref) and ridge regression (ref).
+
+In both, we compute the singular value decompositions of the covariance matrix, transform its diagonal part, and invert it.
+In truncated SVD, the lowest eigenvalues are set to zero, and the corresponding vectors eliminated from future calculations.
+In ridge regression, a constant factor is added to all eigenvalues.
+The regularisation factor and truncation threshold are chosen via leave one out cross validation.
+In practice, ridge regression tends to provide better stability but introduces a bias in the estimates.
+Truncated SVD is used to provide better results in low noise cases.
+
+Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data are detected in the first step and learnt in the second.
+To avoid this, we split our sample into two halves, and each regression is performed over one, effectively decorrelating the two steps.
 %
-Also, to prevent spurious correlations between the two steps (should they be performed on the same sample), our original sample is randomly split into two halves, and each step is performed over one of them, effectively decorrelating the two steps.
-%
-This is repeated over several (100) splits, and the resulting $\hat H$ are averaged.
-%
+To compensate for the reduction in sample size caused by the split, this process is repeated over many splits, and their results are averaged.
 This bagging compensates for the reduction in sample size caused by splitting.
 
-Under our hypotheses (centered noise, E a binary diagonal) and even under heavy noise, $\hat H$ is close to diagonal.
+As we shall see, the diagonal of $\hat H$ recovered by back to back regression is a scaled approximation of the diagonal of E.
+Noise tends reduce the large diagonal elements, whereas the bias introduced by regularisation tends to increase the lowest ones.
+
+To recover an estimate of E, we need a test to select the largest diagonal elements of $\hat H$.
 %
-Yet, because of regularisation which adds positive bias to small eigenvalues, and noise which attenuates large eigenvalues, $\hat H$ is in fact a scaled and noisy approximation to E.
+In regression analysis, a traditional approach to this problem consists in testing whether coefficients differ from zero, via variants of Student t-test.
 %
-To recover a binary diagonal, as our estimate of E, we extract the diagonal of $\hat H$ and select its large elements.
-%
-The diagonal matrix $\hat E$, having large elements of H as its unit diagonal coefficients (the rest being zero) is our estimate of E, the active features of X.
-
-To select large elements of $\hat H$, the traditional approach in regression analysis consists in testing whether its coefficients differ from zero.
-This is usually done via variants of Student t-test.
-%
-This will not work here, as back-to-back regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
+This will not succeed here.
+First because the conditions for these tests do not apply here (we have two regressions).
+Second, because regularisation adds bias on the diagonal, which will cause t-tests to succeed unduly.
 %
 Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).
+This can be done by maximizing the ratio of inter-group variance over total variance (ie minimizing intra-group variance), over all splits of the p largest data and the n-p smallest.
 
-This can be done by maximizing the ratio of inter-group variance over total variance (ie minimizing intra-group variance).
+If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we select the split that maximises the Sonquist and Morgan criterion $$K = \frac{p (n-p)}{n} \frac{(s-r)^2}{ V}$$.
 %
-In a one dimensional two clusters setting, this amounts to sorting the data, and separating the p smallest from the n-p largest values so that inter group inertia is maximal.
-%
-If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we are maximizing the Sonquist and Morgan criterion $$K = {p (n-p) \over n} {(s-r)^2 \over V}$$ and selecting the p features corresponding to the largest values.
-%
-If diagonal elements of $\hat H$ are normally distributed, K follows a chi-square distribution with one degree of freedom.
-%
-At 95\% confidence level, a value of K superior to 3.84 makes our selection significant (for 99\% confidence, the corresponding value of K is 6.63) \citep{Kass_75}
-
-This extract from $\hat H$ a binary diagonal $\hat E$.
+This recovers from $\hat H$ a binary diagonal $\hat E$.
 %
@@
 X(FEX)^\dagger &=X((FK)(K'X))^\dagger \\
-&=X(K'X)'((K'X)(K'X)')^{-1}((FK)'(FK))^{-1}(FK)' \\
-&=(XX')K(K'XX'K)^{-1}(K'F'FK)^{-1} K'F'
+&=X(K'X)'((K'X)(K'X)')^{-1}(FK)^\dagger \\
+&=(XX')K(K'XX'K)^{-1}(FK)^\dagger 
 \end{aligned}
@@
 
-Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{3} & \Sigma_{4}\end{array}\right)$, the top left block being (k,k), since $K=\left(\begin{array}{c}I \\\hline 0\end{array}\right)$, we have $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{3} \Sigma_{1}^{-1}\end{array}\right)$, call S the matrix obtained by adding Nx-k empty columns to this one (or, equivalently multiplying it by K').
+Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, the top left block being $k\times k$.
+Then, $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{2} \Sigma_{1}^{-1}\end{array}\right)$, call it S.
 
-Let us also note that $FE=FKK'$, and so $(K'F'FK)^{-1} K'F'FKK'=I_{k}K'$. Therefore, we have $(K'F'FK)^{-1} K'F'=(FE)^\dagger$.
+Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=X(FEX)^\dagger =S (FK)^\dagger$.
+If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $K (FK)^\dagger$.
 
-Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=X(FEX)^\dagger =S (FE)^\dagger$.
-If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (that is, if causal features do not act as confounders for inactive features), then S=E, and we are retrieving $E (FE)^\dagger$.
+Under centred noise, and if the covariance of FN has low condition number (a weak assumption given that N is noise), we have $\left \| GFN\right \| ^2 \approx Var(FN) \left \| G\right \| ^2$, and minimisation of $\left \| X-GFEX\right\| ^2  + \left \| GFN\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda S (FE)^\dagger$, $\lambda$ a positive number below one.
 
-Under centred noise, and if (FN)'(FN) have low condition number (a weak assumption given that N is noise), we have $\left \| GFN\right \| ^2 \approx Var(FN) \left \| G\right \| ^2$, and minimisation of $\left \| X-GFEX\right\| ^2  + \left \| GFN\right \| ^2$ will recover of the minimum of the left norm, $\lambda S (FE)^\dagger$, $\lambda$ a positive number below 1.
+Replacing in the left part of (ref), we have $X-GFEX=X- \lambda S (FK)^\dagger FKK'X= X - \lambda EX=X(I-\lambda E)$.
+Its square norm is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ ($Var(X_{.})$ denoting the variance of X over active and inactive features).
 
-Replacing, we have $\left \| X-GFEX\right\| ^2 = \left \| X-\lambda S (FE)^\dagger FEX\right\| = \left \| X(I-\lambda S)\right\|$.
-If the covariance of X is block diagonal (no confounding between active and inactive features of X), this is equal to $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ (denoting the variance of X over active and inactive features).
+For the right part, we have $GFN=\lambda S (FK)^\dagger FN= \lambda S (FK)^\dagger (FK+F(I-K))K'N= \lambda S EN=  \lambda EN$.
+(Since F is full rank over E, we have $(FK)^\dagger F(I-K)= 0$).
+The square norm is therefore $\lambda^2 Var(N_{active})$.
 
-For the right part of the sum, we have $GFN=\lambda S (FE)^\dagger FN= \lambda S (FE)^\dagger (FE+F(I-E))N= \lambda S (E +(FE)^\dagger F(I-E))N=\lambda E N $ (FC last part of proof to be done, need some lemma on (FE)+).
-The square norm is therefore $\lambda^2 Var(N) (k/nx)$.
+Thus, with centred and random noise, and with F full rank over E, we are minimising $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N_{active})$.
+Its least value is attained for $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N_{active})}$.
+Assuming X and N have the same variance over all features (this can always be guaranteed), active subscript can be dropped, and this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).
 
-The total quantity to be minimised, under noise, but with non confounding features in X, F full rank over E, and FN decently conditioned, is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N) (k/nx)$, setting the partial derivative in $\lambda$ to zero, we obtain $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N) (k/nx)}$, and if we suppose that X has comparable variance over all features, this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).
+The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} S (FK)^\dagger$.
 
-The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} E (FE)^\dagger$ if active and inactive features of X are not correlated
-
-The second step is straightforward, replacing Y and since N is centred, we have
+The second step is straightforward, replacing Y, we have
 \begin{equation}
@@
 
-Thus, in the linear case, back-and-forth regression retrieves $\hat H = \frac{1}{1+nsr} E (FE)^{\dagger}(FE) = \frac{1}{1+nsr} E$.
+Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} S (FK)^{\dagger}(FKK') = \frac{1}{1+nsr} E$.
 

### 2019-05-22 paper/introduction.tex

@@
-Natural sciences are tasked to find, from a set of hypothetical factors, the minimal subset that suffices to reliably predict novel observations potentially in novel environments.
-%
-This problem can be formalized as follow: empirical studies consist in measuring, via an apparatus $F$, a signal $Y \in \mathbb{R}^{n, q}$, that is a function of possible causes $X \in \mathbb{R}^{n, d}$, and other unknown causes that can be understood as noise $N \in \mathbb{R}^{n, d}$: $Y=F(X,N)$.
-Of all considered causes (features of $X$), only a subset accounts for $Y$, let it be represented as $E \cdot X$, with $E \in \mathbb{D}^{d, d}$ a binary diagonal matrix, which selects causal features of $X$.
-%
-Under the assumptions of centered and additive noise N, causal discovery thus consists in finding $E$ in
+Natural sciences are tasked to find, from a set of hypothetical factors, the minimal subset that suffices to reliably predict novel observations. This endeavor is impeded by two major challenges.
 
-\begin{equation}
-    Y=F(XE, N)
-\end{equation}
+First, causal and non-causal factors may be numerous and collinear.
+In physics, for example, a student may have difficulty identifying whether fusion is caused by a change in temperature or in pressure, as these two factors can be difficult to manipulate independently.
+This issue becomes increasingly pronounced as the number of potential factor increases.
+In neuroscience for example, it can be challenging to identify whether the frequency of a word presented on a subject' retina modulates brain activity.
+Indeed, word frequency strongly covaries with word length (short words are more frequent than long words) as well as with word category (determinants are more frequent than adverbs), the individual letters from which it is composed etc (e.g.
+words that contain "z" are rare) and so forth.
+As it can be particularly difficult to generate a set of words that control for all of these factors simultaneously, it is common to use forward modeling, i.e.
+to train a model to predict observations (e.g.
+brain activity) from a minimal combination of competing factors (e.g.
+word length, word frequency, etc), and investigate the resulting parameters of the best model.
 
-%
-Under this formalism, causal discovery is impeded by two challenges.
-%
-% To illustrate these two challenges, let us consider the case of a neuroscientist who studies the brain bases of reading.
-Our neuroscientist wants to determine whether the frequency of a word modulates brain activity (e.g.
-neurons may become more active when our retina is flashed with the word 'triskaidekaphobia' than when the word is 'car').
-One issue is that word frequency tends to correlate with word length.\par
-%
-% This is the first challenge to causal discovery
-The first challenge is factor collinearity: causal and non-causal factors may tend to covary with one another: i.e.
-$X'X$ is not diagonal.
-Collinearity in $X$ is typically addressed with factorial designs \cite{fisher_1935}.
-% For example, our neuroscientist may carefully select words matched in terms of length, while letting their frequency vary.
-However, factorial design becomes increasingly difficult to implement as the number of considered factors grows.
-As a fallback option, factors may be orthogonalized \textit{a posteriori}, by analytically estimating how each of them contributes to explaining the observations with ordinary least square (OLS):
+The second challenge to causal discovery is that observations are becoming increasingly large and complex.
+The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from large and complex observations.
+For example, brain activity is generally recorded with hundreds of sensors simultaneously.
+It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus.
 
-% Our neuroscientist will thus
-\begin{equation}
-H = (X'X)^{-1} X'Y
-\end{equation}
-In this paper, we refer to this approach as 'forward modeling' as it models how the factors X contribute to explain the observations Y. \par
-%
-%
-The second challenge to causal discovery is the multiplicity of the observations.
-% For example, our neuroscientist may simultaneously collect noisy recordings of hundreds of thousands of neurons.
-It can thus be suboptimal to consider each neuron independently, because they may be influenced by some unknown factors $N$.
-With the rise of large datasets, the complexity of observations has led many to focus on optimally predicting putative factors from complex and poorly controlled observations.
-In neuroimaging, for example, it is increasingly popular to use "decoding" methods: instead of testing whether the brain responds to particular factors, a model is trained to maximally predict factors from complex brain signals:
-\begin{equation}
-G = (Y'Y)^{-1} Y'X
-\end{equation}
-We refer to this approach as 'backward modeling' as it models how the observations Y can be combined to predict the factors X. \par
-%
-%
 As illustrated in Fig.
@@
 \par
-%
-%
+
 In this paper, we introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling.
-Back-to-back regression consists in first predicting X from Y (as in the backward modeling) subsequently retrieving an estimate $\hat X$ and finally disentangling which of the considered factors X accounts for the variance in $\hat X$ (i.e.
-analogously to forward modeling).
-\par
-%
-In the next sections, we detail back-to-back regression, prove that it asymptotically extracts E, and show how it relates to and departs from other cross-decomposition and component analyses.
-We then show with synthetic experiments that back-to-back regression outperforms state-of-the-art techniques for regression and feature extraction, both in terms of causal discovery and out-of-environment predictions.
-Finally, we apply back-to-back regression to the multidimensional recordings of brain activity of 102 subjects reading $\approx$ 2,000 words, whose underlying features covary.
-Our results  show that the first brain responses to word presentation is mainly modulated by low-level visual features, such as the number and category of letters, while late brain responses appear to be mainly modulated by high-level lexical features, such as word frequency and part-of-speech.
-\par
-%
+After detailing the method and proving its convergence, we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of causal discovery and out-of-environment predictions.
+Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features leads to remarkably distinguishable brain responses.

### 2019-05-22 paper/paper.tex

@@
 \section{Experiments - simulated data}
-Experiments on simulated data serve three purposes: compare back-to-back regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.
-
-The model used for simulation is $Y=f(snr MEX+N)$, with X a random (dx,N) matrix, E a square binary diagonal of dimension dx with the nc last elements equal to 1 (this means that of the dx features of X, only the nc last have some influence on Y), M and N (dz,dx) and (dz,N) random matrices, and f a function from $\mathbb{R}^{dz}$ to $\mathbb{R}^{dy}$, that can be linear or non linear.
-%
-In the linear case, the model becomes $Y=F(MEX+N)$, with F a (dy,dz) matrix.
-%
-In the non linear case, F becomes $FSH$ where F and H are matrices of dimensions (dz,dz) and (dy,dz), and S a vector of dz sigmoid transformations.
-%
-Finally, $snr$ is a real number measuring the amount of noise.
-
-Our simulation therefore depends upon 6 free parameters :
-\begin{itemize}
-\item dx : number of features of X
-\item nc: number of active features (in EX)
-\item dz: number of noise features
-\item snr: signal to noise ratio
-\item dy: number of features in Y
-\item N: size of sample
-\end{itemize}
-
-Random matrices X, M, N, H and F are populated with centred gaussian random coefficients with unit variance, divided by $ \surd n$ where n is the smallest dimension of the matrix.
-%
-Since a random square matrix with unit gaussian coefficients of dimension n has eigenvalues between $ \surd n $ and $\surd n $, such  scaling allows for meaningful comparisons when dimensions dx, dy and dz vary.
-
-In the non linear case, sigmoid functions are...
-FC: insert comment about scaling sigmoids, and maybe something about uniform coeffs
-
-We evaluate the performance of back-to-back regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
-%
-For feature extraction, the quality of fit is the number of disagreements between our estimator and the diagonal of E (ie, their Hamming distance, which should be minimised).
-%
-For regression, we use a test set of N additional measurements, and calculate the R-squared ratio of variances (which should be maximised).
-%
-Performance indicators are calculated by averaging the results of several experiments.
-
-\subsection{Feature extraction - linear case}
-All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by back-to-back regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
-%
-The decrease is usually very steep until a Hamming distance under 2 is achieved.
-%
-Figure 1 presents a number of such curves for varying values of dx=dy=dz, nc=30 and N=1000.
-
-Figure 1
-
-The distance between $\hat E$ and E incorporates two different errors: false positives, where features not in E are extracted, and false negatives where features in E are not extracted.
-%
-For low values of the snr, and typical values of nc (much smaller than dx), false positives are high, but they fall faster as snr increases.
-%
-False negatives begin lower, but decrease more slowly (figure 2).
-%
-In other words, for each values of the parameters, one may determine a value of the signal to noise ratio, such that snr higher than this threshold allow for good feature extraction.
-%
-At this level, all the errors are false negatives.
-
-Figure 2
-
-Working from a fixed sample of 1000, and 15 active features, and varying values of dx, dy, dz (from 50 to 200) and snr (from 0.2 to 1), we observe that, all other factors being equal, the quality of feature extraction and regression decrease when dx increases (X has more inactive features), dz increase (more noise dimensions) and snr decreases (less signal, more noise).
-%
-The dimensionality of Y plays a much smaller role, except for very small values of dy (less than 10).
-
-For feature extraction, back-to-back regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
-%
-The following graphs show the boundary between values of dx and dz that allow for features extractions (dx lower than boundary, or dz larger), for different values of dy (level curves on the graph), and different values of signal to noise ratio and number of active variables.
-
-FC: add graphs here, see whether it is better to have nc fixed, or a proportion of dx, note : part of this can be moved to appendix if lack of space
-
-For values of ...
-%
-in ..., an empirical formula can be provided.
-%
-back-to-back regression allows for good feature extraction if :
-
-\subsection{Feature extraction - nonlinear case}
-
-\subsection{Regression }
-For regression, we want to understand how feature extraction helps predict Y from X.
-%
-To this effect, we compare four approaches to regression:
-\begin{itemize}
-\item standard ridge regression of Y from X, where no features are extracted this is our baseline
-\item weighted ridge regression, of Y from $\hat H X$, where features of X are weighted according to the diagonal of $\hat H$
-\item filtered ridge regression, of Y from $\hat E X$, where active features are first extracted (as above), and Y is predicted from the extracted features only
-\item perfect filtered (oracle) ridge regression of Y from EX, where E is supposed known
-\end{itemize}
-Over a large set of parameters, we observe that filtered regression is on average better than weighted, which is on average better than standard regression.
-%
-This demonstrates that back-to-back regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.
-
-FC graphs go here
-
-The quality of regression is independent from dy (except for very small values), increase with snr, decreases with dx, and increases with dz if it is larger than dx
-FC graphs go here
-
-\subsection{Comparison with other methods}
-Maxime territory
-
-\subsection{Multiple environments}
-In this section, we perform the same tests as in the previous one, but the test set is taken from a different distribution of X than the training set.
-%
-In practice, we test the same trained predictors of Y over test samples generated from a distribution of X with a different covariance.
-
-The rationale behind such tests is that whereas back-to-back regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.
-
-FC: ideally, we'd like to assess the multiple env performance of jrr, and compare it with others.
-%
-Are multiple environments limited to regression analyses? How should environments be generated (not just variance of X, I expect)?
+\input{synthetic.tex}
 

### 2019-05-22 paper/synthetic.tex

@@
-Experiments on simulated data serve three purposes: compare back-to-back regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.
+We evaluate the performance of B2BR throughout a series of experiments on controlled synthetic data.
+%
+The purpose of these experiments is to evaluate the ability of B2BR in terms of prediction (inside and outside the training distribution), as well as a method to recover causal factors.
 
-The model used for simulation is $Y=f(snr MEX+N)$, with X a random (dx,N) matrix, E a square binary diagonal of dimension dx with the nc last elements equal to 1 (this means that of the dx features of X, only the nc last have some influence on Y), M and N (dz,dx) and (dz,N) random matrices, and f a function from $\mathbb{R}^{dz}$ to $\mathbb{R}^{dy}$, that can be linear or non linear.
+The data generating process for each experiment constructs $n$ training examples according to the model $Y = \sigma((\text{snr} \cdot XE + N)F)$.
 %
-In the linear case, the model becomes $Y=F(MEX+N)$, with F a (dy,dz) matrix.
+Here, 
+%\begin{itemize}
+%    \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
+%    \item $X \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_X)$,
+%    \item $N \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$,
+%    \item $E \in \mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
+%    \item $\Sigma_X = AA^\top$, where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
+%    \item $\Sigma_N = BB^\top$, where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
+%\end{itemize}
+    $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
+    $X \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_X)$,
+    $N \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$,
+    $E \in \mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
+    $\Sigma_X = AA^\top$ where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
+    $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
+the factor $\text{snr} \in (0, \infty)$, and $\sigma$ is an elementwise function.
+
+To simulate a wide range of experimental conditions, we vary $n \in \{500, 1000, 2000\}$, $d_x, d_y \in \{ 25, 50, 100, 150 \}$, $n_c \in \{ 5, 10 \}$, $\text{snr} \in \{ 0.2, 0.5, 0.75, 1 \}$, and $\sigma \in \{ \text{identity}, \text{sigmoid} \}$.
 %
-In the non linear case, F becomes $FSH$ where F and H are matrices of dimensions (dz,dz) and (dy,dz), and S a vector of dz sigmoid transformations.
+Each condition is simulated under $20$ different random seeds.
+
+We compare the performance of B2BR against five competing methods:
 %
-Finally, $snr$ is a real number measuring the amount of noise.
+Ridge regression \citep{}, multitask Lasso \citep{}, Partial Least Squares or PLS \citep[PLS]{}, Canonical Correlation Analysis \citep[CCA]{}, and Reduced Rank Regression \citep[RRR]{}.
+%
+Each of these methods estimates a matrix of coefficients $\hat{W} \in \mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$ using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
+%
+For B2BR, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
+%
+We employ $5$-fold cross-validation to select the optimal hyper-parameters of each method.
 
-Our simulation therefore depends upon 6 free parameters :
-\begin{itemize}
-\item dx : number of features of X
-\item nc: number of active features (in EX)
-\item dz: number of noise features
-\item snr: signal to noise ratio
-\item dy: number of features in Y
-\item N: size of sample
-\end{itemize}
+We evaluate seven metrics: 
+%
+test error in-distribution, test error out-distribution, test error-in distribution after weighting with $\hat{E}$, test error out-distribution after weighting with $\hat{E}$, Sonquist-Morgan false positives on $\hat{E}$, Sonquist-Morgan false negatives on $\hat{E}$, and the area under the ROC curve (AUC) between $\hat{E}$ and $E$.
+%
+The ``in-distribution'' metrics measure the test error of each predictor before and after weighting their features by the estimated mask $\hat{E}$.
+%
+The ``out-distribution'' metrics are similar, measuring the test error at samples drawn using two different random covariance matrices for both $X$ and $N$.
+%
+The false positives, false negatives, and AUC statistics measure the quality of identification of causal factors between $\hat{E}$ and the true, binary $E$.
 
-Random matrices X, M, N, H and F are populated with centred gaussian random coefficients with unit variance, divided by $ \surd n$ where n is the smallest dimension of the matrix.
+\begin{figure}[htpb]
+    \centering
+    \includegraphics[width=\textwidth]{synthetic.pdf}
+    \caption{Results of synthetic experiments.
+Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration, averaged over $20$ random seeds.
     %
-Since a random square matrix with unit gaussian coefficients of dimension n has eigenvalues between $ \surd n $ and $\surd n $, such  scaling allows for meaningful comparisons when dimensions dx, dy and dz vary.
+    Blue points indicate linear experimental conditions, while orange points indicate nonlinear conditions.}
+    \label{table:synthetic}
+\end{figure}
 
-In the non linear case, sigmoid functions are...
-FC: insert comment about scaling sigmoids, and maybe something about uniform coeffs
+Table~\ref{table:synthetic} summarizes the results for all experimental conditions, metrics, and methods...
 
-We evaluate the performance of back-to-back regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
-%
-For feature extraction, the quality of fit is the number of disagreements between our estimator and the diagonal of E (ie, their Hamming distance, which should be minimised).
-%
-For regression, we use a test set of N additional measurements, and calculate the R-squared ratio of variances (which should be maximised).
-%
-Performance indicators are calculated by averaging the results of several experiments.
-
-\subsection{Feature extraction - linear case}
-All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by back-to-back regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
-%
-The decrease is usually very steep until a Hamming distance under 2 is achieved.
-%
-Figure 1 presents a number of such curves for varying values of dx=dy=dz, nc=30 and N=1000.
-
-Figure 1
-
-The distance between $\hat E$ and E incorporates two different errors: false positives, where features not in E are extracted, and false negatives where features in E are not extracted.
-%
-For low values of the snr, and typical values of nc (much smaller than dx), false positives are high, but they fall faster as snr increases.
-%
-False negatives begin lower, but decrease more slowly (figure 2).
-%
-In other words, for each values of the parameters, one may determine a value of the signal to noise ratio, such that snr higher than this threshold allow for good feature extraction.
-%
-At this level, all the errors are false negatives.
-
-Figure 2
-
-Working from a fixed sample of 1000, and 15 active features, and varying values of dx, dy, dz (from 50 to 200) and snr (from 0.2 to 1), we observe that, all other factors being equal, the quality of feature extraction and regression decrease when dx increases (X has more inactive features), dz increase (more noise dimensions) and snr decreases (less signal, more noise).
-%
-The dimensionality of Y plays a much smaller role, except for very small values of dy (less than 10).
-
-For feature extraction, back-to-back regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
-%
-The following graphs show the boundary between values of dx and dz that allow for features extractions (dx lower than boundary, or dz larger), for different values of dy (level curves on the graph), and different values of signal to noise ratio and number of active variables.
-
-FC: add graphs here, see whether it is better to have nc fixed, or a proportion of dx, note : part of this can be moved to appendix if lack of space
-
-For values of ...
-%
-in ..., an empirical formula can be provided.
-%
-back-to-back regression allows for good feature extraction if :
-
-\subsection{Feature extraction - nonlinear case}
-
-\subsection{Regression }
-For regression, we want to understand how feature extraction helps predict Y from X.
-%
-To this effect, we compare four approaches to regression:
-\begin{itemize}
-\item standard ridge regression of Y from X, where no features are extracted this is our baseline
-\item weighted ridge regression, of Y from $\hat H X$, where features of X are weighted according to the diagonal of $\hat H$
-\item filtered ridge regression, of Y from $\hat E X$, where active features are first extracted (as above), and Y is predicted from the extracted features only
-\item perfect filtered (oracle) ridge regression of Y from EX, where E is supposed known
-\end{itemize}
-Over a large set of parameters, we observe that filtered regression is on average better than weighted, which is on average better than standard regression.
-%
-This demonstrates that back-to-back regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.
-
-FC graphs go here
-
-The quality of regression is independent from dy (except for very small values), increase with snr, decreases with dx, and increases with dz if it is larger than dx
-FC graphs go here
-
-\subsection{Comparison with other methods}
-Maxime territory
-
-\subsection{Multiple environments}
-In this section, we perform the same tests as in the previous one, but the test set is taken from a different distribution of X than the training set.
-%
-In practice, we test the same trained predictors of Y over test samples generated from a distribution of X with a different covariance.
-
-The rationale behind such tests is that whereas back-to-back regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.
-
-FC: ideally, we'd like to assess the multiple env performance of jrr, and compare it with others.
-%
-Are multiple environments limited to regression analyses? How should environments be generated (not just variance of X, I expect)?
+\clearpage
+\newpage

### 2019-05-22 paper/meg_exps.tex
exps

--- 2019-05-22 paper/meg_exps.tex
@@
+\newpage
 \subsection{Dataset}
@@
 Next, we apply our method to brain imaging data from the anonymized multimodal
-neuroimaging ``Mother Of all Unification Studies'' (MOUS) dataset \cite{schoffelen2019204}.
-The dataset contains magneto-encephalography (MEG) recordings of 102 healthy native-Dutch
-adults who participated in a reading task.
-
-Subjects were exposed to a rapid serial visual presentation of Dutch words. The word lists
-consisted of 120 sentences, and scrambled lists of the same words. Each word was presented on the computer screen for 351ms on average (min: 300ms, max: 1400ms).
-Successive words were separated by a blank screen for 300ms, and successive
-sentences were separated by an empty screen for a few (3-4) seconds.
+neuroimaging ``Mother Of all Unification Studies'' (MOUS) dataset
+(\cite{schoffelen2019204}).  The dataset contains magneto-encephalography (MEG)
+recordings of 102 healthy native-Dutch adults who participated in a reading
+task.
+%
+Subjects were exposed to a rapid serial visual presentation of Dutch words. The
+word lists consisted of 120 sentences, and scrambled lists of the same words.
+Each word was presented on the computer screen for 351ms on average (min: 300ms,
+max: 1400ms).  Successive words were separated by a blank screen for 300ms, and
+successive sentences were separated by an empty screen for a few (3-4) seconds.
 
@@
 
-The raw MEG data was bandpass-filtered between 0.1 and 40 Hz using MNE-Python default
-parameters \cite{mne}. Specifically, we used a zero-phase finite impulse response filter (FIR) with a hamming window and with transition bands of 0.1 Hz and 10 Hz for the low and high
-cut-off frequencies respectively.
+The raw MEG data was bandpass-filtered between 0.1 and 40 Hz using MNE-Python
+default parameters (\cite{gramfort2013meg, gramfort2014mne}). Specifically, we used a zero-phase finite impulse
+response filter (FIR) with a Hamming window and with transition bands of 0.1 Hz
+and 10 Hz for the low and high cut-off frequencies respectively.
 
 The raw data was then segmented 100 ms before word onset and 1.000 sec after
-word onset.
-Time $t=0$ms corresponds to the word onset.
-Finally, each resulting segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and thus led a sampling frequency of 240 Hz.
-We show, in Figure \ref{fig:megavg}, the average response of the
-magnetometers evoked by the words presented to the first subject.
+word onset. Time $t=0$ms corresponds to the word onset. Finally, each resulting
+segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
+thus led a sampling frequency of 240 Hz. We show, in Figure \ref{fig:megavg},
+the average response of the magnetometers evoked by the words presented to the
+first subject.
 
-For each word, each subject, and each time sample relative to word onset, we build an observation matrix $Y \in \mathbb{R}^{n, q}$ of $n\approx$ 2700 words by $q=301$ MEG channels (273 magnetometers and 28 compensation channels).
-Each of the columns of $Y$ is normalized to have zero mean and unit variance.
+For each subject and each time sample relative to word onset, we
+build an observation matrix $Y \in \mathbb{R}^{n, q}$ of $n\approx$ 2700 words
+by $q=301$ MEG channels (273 magnetometers and 28 compensation channels). Each
+of the columns of $Y$ is normalized to have zero mean and unit variance.
 
@@
   \centering
-  \includegraphics[width=\textwidth, trim=1.5cm 1cm 0.5cm 0cm, clip=True]{figure_meg.pdf}
-  \caption{Average magnetosensor response across words for one subject.}
+  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_sensors.pdf}
+  \caption{Magnetosensor response (femtoteslas fT) averaged across words for the first subject.
+The color coding corresponds to the positions of the sensors on the head as shown on the top-left diagram.
+The diverging curves correspond to artifacts in data collection.}
   \label{fig:megavg}
@@
 
-\subsection{Word features}
-We annotated each word with 38 features: 26 corresponds to the number of each letter in the word (from 'a' to 'z', independently of their case), the total number of letters ('word length'), the frequency of the word in natural language (using the Zipf logarithmic scale \cite{marc_brysbaert}), and the part-of-speech tags (i.e.
-NOUN, PROPN (proper noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ (adjective), ADV (adverb), and NUM (numerical)) using the packages Spacy \cite{spacy} and WordFreq \cite{speer}.
-For clarity purposes, we refer to these features using two broad categories: visual (i.e.
-each letter and the total number of letters) and lexical (part-of-speech and word frequency).
-This procedure yields an $X \in \mathbb{R}^{n, d}$ matrix of $n\approx$ 2700 words by $d=38$ features for each subject.
-Each of the columns of $X$ is normalized to have a mean and a standard deviation of 0 and 1 respectively.
+\subsection{Word features} We annotated each word with 38 features: 26
+correspond to the count of each letter in the word (from 'a' to 'z',
+independently of their case), the total number of letters ('word length'), the
+frequency of the word in natural language (using the Zipf logarithmic scale
+of \cite{van2014subtlex}), and the part-of-speech tags (i.e. NOUN, PROPN (proper
+noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ
+(adjective), ADV (adverb), and NUM (numerical)) using the packages Spacy
+(\cite{spacy2}) and WordFreq (\cite{speerwordfreq}). For clarity purposes, we refer to these
+features using two broad categories: visual (i.e. each letter and the total
+number of letters) and lexical (part-of-speech and word frequency).  This
+procedure yields an $X \in \mathbb{R}^{n, d}$ matrix of $n\approx$ 2700 words by
+$d=38$ features for each subject. Each of the columns of $X$ is normalized to
+have a mean and a standard deviation of 0 and 1 respectively.
 
-\subsection{Analyses}
-We aimed to identify which factors of $X$ explain the MEG recordings $Y$ at each time sample and for each subject separately.
-To this aim, we implemented three types of analyses (forward, backward and back-to-back regressions), and tested whether their results were consistent across subjects, using a second-level statistics described below.
+\subsection{Analyses} We attempt to identify which factors of $X$ explain the MEG
+recordings $Y$ at each time sample and for each subject separately. To this aim,
+we implement three types of analyses (forward, backward and back-to-back
+regressions). Corresponding results can be found in Figure \ref{fig:meg_twocurves}.
 
-Forward regression, also known as 'encoding' in neuroimaging, here consisted in fitting a Ridge-regularized OLS from $X$ to $Y$:
+\begin{figure}[t!]
+  \centering
+  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_twocurves.pdf}
+  \caption{We compare the encoding approach (columns 1 and 2, ridge regression $X \rightarrow Y$) and the decoding approach (column 3, ridge regression $Y \rightarrow X$) to our method (column 4, $X \rightarrow Y \rightarrow X$) on real MEG data from the MOUS dataset.
+Top row: the first subject.
+Bottom row: average across subjects.
+  We plot the contributions of the \textit{word frequency} and the \textit{word
+  length} over time. Our method can disambiguate them consistently with
+  previous neuroscientific work showing an early response ($\approx$ 150ms) to
+  length and a later response ($\approx$ 400ms) to frequency. See text for
+  details.}
+  \label{fig:meg_twocurves}
+\end{figure}
+
+\paragraph{Forward regression} also known as 'encoding' in neuroimaging, here consists of
+fitting a ridge-regularized ordinary least squares (OLS) regression \footnote{We use scikit-learn (\cite{sklearn}).} from $X$ to $Y$:
 \begin{equation}
@@
 \end{equation}
-where $\lambda$ is the regularization parameter fit with the leave-one-out cross-validation implemented in scikit-learn \cite{sklearn} from 20 possible value logarithmatically distributed between $10^{-5}$ and $10^5$.
-We aimed to estimate the contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor.
-However, $H\in\mathbb{R}^{d, q}$ is a matrix of $d$ word features by $q$ MEG channels.
-We thus approximated the relative contribution of each feature by sum-squarring the H coefficients:
+where $\lambda$ is the
+regularization parameter, cross-validated between $10^{-5}$ and $10^5$.
+
+We then estimate the contribution
+$diag(\hat E) \in\mathbb{D}^{d} $ of each factor. However, $H\in\mathbb{R}^{d,
+q}$ is a matrix of $d$ word features by $q$ MEG channels. We approximate
+the relative contribution of each feature with the sum-of-squares of the corresponding coefficients in $H$; for the $j$-th feature:
 \begin{equation}
-\hat e_d = \sum_{i=0}^{i=q} {h_d^i}^2
+  \hat e_j = \sum_{i=0}^{q} {h_{i,j}}^2
 \end{equation}
+\paragraph{Backward regression} or 'decoding', is similar but predicts $X$ from $Y$:
+\begin{equation} G = (Y^{T}Y+\lambda I)^{-1} Y^{T}X \end{equation}
+where $\lambda$ is cross-validated as for encoding. We estimate the
+contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor in the following
+way. We predict $X$ on a held-out split \textit{(train, test)}:
+% Using 10-fold cross-validation, we thus assessed the ability of this ridge
+% regression to predict independent and identically distributed data:
+\begin{equation} \hat X_{test} = Y_{test}\cdot G_{train} \end{equation}
+and assess, for each factor of X independently, the Pearson correlation
+coefficient $r$ between $X_{test}$ and $\hat X_{test}$. We repeat the procedure
+for 20 random splits and average the result.
 
-Backward regression, also known as 'decoding' in neuroimaging, here consisted in fitting a Ridge-regularized OLS from $Y$ to $X$:
-
-\begin{equation}
-G = (Y^{T}Y+\lambda I)^{-1} Y^{T}X
-\end{equation}
-
-where $\lambda$ is the regularization parameter fit with the leave-one-out cross-validation
-implemented in scikit-learn \cite{sklearn} from 20 possible value logarithmically distributed
-between $10^{-5}$ and $10^5$. We aimed to estimate the contribution $diag(\hat E)
-\in\mathbb{D}^{d} $ of each factor. Using 10-fold cross-validation, we thus assessed the ability
-of this ridge regression to predict independent and identically distributed data:
-
-\begin{equation}
-\hat X_{test} = Y_{test}\cdot G_{train}
-\end{equation}
-
-and assessing, for each factor of X independently, the Pearson correlation coefficient $r$ between $X_{test}$ and $\hat X_{test}$ for each cross-validation split.
-
-Finally, back-to-back regression was fit as decribed earlier using two additional stability tricks designed to minimize the impact of bagging.
-First the $G$ ridge regression was broken-down with SVD blablabla
-Second, $X$ was regularized using a maximal likelihood blablabla
-Third, no regularization
+% Finally,
+\paragraph{Back-to-back regression} was fit as described earlier using two
+additional stability tricks designed to minimize the impact of bagging.  First
+the $G$ ridge regression was broken-down with SVD blablabla Second, $X$ was
+regularized using a maximal likelihood blablabla Third, no regularization
 
@@
   \centering
-  \includegraphics[width=\textwidth, trim=1.5cm 1cm 0.5cm 0cm,
-clip=True]{meg_result.pdf}
+  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_result.pdf}
   \caption{Average magnetosensor response across words for one subject.}
@@
 \subsection{Discussion of the results.}
-
-\subsubsection{Overview}

### 2019-05-22 paper/meg_exps.tex
add paragraph meg stabilization

--- 2019-05-22 paper/meg_exps.tex
@@
 
-Finally, back-to-back regression was fit as decribed earlier using two additional stability tricks designed to minimize the impact of bagging.
-First the $G$ ridge regression was broken-down with SVD blablabla
-Second, $X$ was regularized using a maximal likelihood blablabla
-Third, no regularization
+Finally, we adapted back-to-back regression with two changes.
+First, to avoid getting a slightly positively-biased estimate of the factors contributing to the MEG signal, we swap the Tikhonov regularization of the first regression $G$ with a trunkated regularization.
+Trunkation was fitted by maximum likelihood estimate, as implemented in scikit-learn principal component analysis.
+Second, to minimize the variance increase induced by bagging, we computed the inverse regularized covariance matrices outside bagging:
+\begin{equation}
+G = (Y_^T Y+\lambda_Y I)^{-1} Y_i^T X_i
+\end{equation}
 
-Back-to-back regression directly leads to $\hat E$.
+\begin{equation}
+\hat X_j = Y_j \cdot G^T
+\end{equation}
+where i and j are partition of the data.
 
@@
 \begin{equation}
-$ttest(\hat E_{t=-100} - \hat E_{\tau}$
+ttest(\hat E_{t=-100} - \hat E_{\tau})
 \end{equation}

### 2019-05-22 paper/meg_exps.tex
meg exps

--- 2019-05-22 paper/meg_exps.tex
@@
 
-% Finally,
-\paragraph{Back-to-back regression} was fit as described earlier using two
-additional stability tricks designed to minimize the impact of bagging.  First
-the $G$ ridge regression was broken-down with SVD blablabla Second, $X$ was
-regularized using a maximal likelihood blablabla Third, no regularization
-
-Back-to-back regression directly leads to $\hat E$.
+\paragraph{Back-to-back regression} We adapted back-to-back regression with two
+changes. First, to avoid getting a slightly positively-biased estimate of the
+factors contributing to the MEG signal, we swap the Tikhonov regularization of
+the first regression $G$ with a trunkated regularization. Trunkation was fitted
+ by maximum likelihood estimate, as implemented in scikit-learn principal
+ component analysis. Second, to minimize the variance increase induced by
+ bagging, we computed the inverse regularized covariance matrices outside
+ bagging:
+\begin{equation}
+G = (Y_^T Y+\lambda_Y I)^{-1} Y_i^T X_i
+\end{equation}
 

### 2019-05-22 paper/paper.tex
Sync

@@
 
-\title{Discovering causal factors with back-to-back regression}
+\title{Measuring causal influence with\\ back-to-back regression}
 
@@
 
-    \node[] (labY) at (\posY, 2){Observations};
-    \node[] (labX) at (\posX, 2){Factors};
-    \node[] (labE) at (\posE, 2.3){Cause};
-    \node[] (labE) at (\posE, 2){selection};
-    \node[] (labN) at (\posN, 2){Noise};
-    \node[] (labF) at (\posF, 2.3){Cause-effect};
-    \node[] (labF) at (\posF, 2){mapping};
+    \node[] (labY) at (\posY, 3){Observations};
+    \node[] (labX) at (\posX, 3){Factors};
+    \node[] (labE) at (\posE, 3.3){Cause};
+    \node[] (labE) at (\posE, 3){selection};
+    \node[] (labN) at (\posN, 3){Noise};
+    \node[] (labF) at (\posF, 3.3){Cause-effect};
+    \node[] (labF) at (\posF, 3){mapping};
+
+    \node[] (labY) at (\posY, 2)   {\color{gray} MEG recordings};
+    \node[] (labX) at (\posX, 2)   {\color{gray} stimulus};
+    \node[] (labX) at (\posX, 2.3) {\color{gray} features};
+    \node[] (labE) at (\posE, 2.3) {\color{gray} causal};
+    \node[] (labE) at (\posE, 2)   {\color{gray} features};
+    \node[] (labN) at (\posN, 2)   {\color{gray} breathing\ldots};
+    \node[] (labF) at (\posN, 2.3) {\color{gray} bloodflow,};
+    \node[] (labF) at (\posF, 2)   {\color{gray} MEG helmet};
 
@@
     \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
-    \node[] (reg1) at (5,-3.25) {$\hat{E} = \underbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}} $};
+    \node[] (reg1) at (5.5,-3.25) {$\hat{E} = \text{Diag}(\underbrace{(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}})$};
     \end{tikzpicture}
-    \caption{In its linear form, back-to-back regression identifies the subset of factors that suffices to account for multidimensional observations by piping two successive regressions from $Y$ to $X$ and from $X$ to $\hat X$.
-}
+    \caption{Back-to-back regression identifies the subset of factors $E_{ii} = 1$ in $X$ that influence some observations $Y$ by 1) regressing from $Y$ to $X$ to obtain $\hat{X}$, and ii) returning the diagonal of the regression coefficients from $X$ to $\hat{X}$.}
     \label{fig:}
@@
 
-In the non linear model, $F(EX+N)$ is defined as before, but Y is the result of a non linear transformation of it $$Y=\sigma (F(EX+N))$$
+% In the non linear model, $F(EX+N)$ is defined as before, but Y is the result of a non linear transformation of it $$Y=\sigma (F(EX+N))$$
 
@@
 
-\subsection{Asymptotic behaviour - non linear case}
+% \subsection{Asymptotic behaviour - non linear case}
+
+\section{Experiments}
+\subsection{Synthetic data}
+
+\input{synthetic.tex}
+
+\subsection{Magnetoencephalography data}
+
+\input{meg_exps.tex}
 
@@
 
-\section{Experiments - simulated data}
-\input{synthetic.tex}
-
-\section{Experiments - real case}
-Jean Remi and Maxime territory
-
-\input{meg_exps.tex}
-
 \section{Conclusion}
 
-\section{Acknowlegements}
-Data were provided (in part) by the Donders Institute for Brain, Cognition and Behaviour.
+% \section{Acknowlegements}
+% Data were provided (in part) by the Donders Institute for Brain, Cognition and Behaviour.
 
@@
 
-\section{Appendices}
-
-This should hold part of the tests, explanations on simulation, detailed results and stuff on meg data
+%\section{Appendices}
+%
+%This should hold part of the tests, explanations on simulation, detailed results and stuff on meg data
 

### 2019-05-22 paper/meg_exps.tex
Sync
it builds

@@
 contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor in the following
-way. We predict $X$ on a held-out split \textit{(train, test)}:
+way. We predict $X$ on a held-out split \textit{test}:
 % Using 10-fold cross-validation, we thus assessed the ability of this ridge
@@
 factors contributing to the MEG signal, we swap the Tikhonov regularization of
-the first regression $G$ with a trunkated regularization. Trunkation was fitted
+the first regression $G$ with a truncated regularization. Truncation was fitted
  by maximum likelihood estimate, as implemented in scikit-learn principal
@@
 \begin{equation}
-G = (Y_^T Y+\lambda_Y I)^{-1} Y_i^T X_i
+  G = (Y^T Y+\lambda_Y I)^{-1} Y_{train}^T X_{train}
 \end{equation}
+\begin{equation}
+\hat X_{test} = Y_{test} \cdot G^T
+\end{equation} where $\left(train,test\right)$ are partitions of the data.
 

### 2019-05-22 paper/paper.tex
Up to proof

@@
 \usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx}
+\usepackage[ruled, linesnumbered]{algorithm2e}
+
 \usepackage[hidelinks]{hyperref}
@@
 \DeclareMathOperator{\Tr}{Tr}
+
+\SetKwComment{Comment}{$\triangleright$\ }{}
 
@@
     \caption{Back-to-back regression identifies the subset of factors $E_{ii} = 1$ in $X$ that influence some observations $Y$ by 1) regressing from $Y$ to $X$ to obtain $\hat{X}$, and ii) returning the diagonal of the regression coefficients from $X$ to $\hat{X}$.}
-    \label{fig:}
+    \label{fig:b2b}
 \end{figure}
 
-\subsection{Formalism}
-Via an apparatus F, we are measuring a multidimensional signal Y, which is a function of possible causes X and other non causal features N.
-Not all features of X exert causal influence on Y.
-Let EX (E a matrix of zero and one) the subset of causal factors.
-Our model can be formalised as $Y=f(EX,N)$.
+We consider the measurement of multivariate signal $Y$, generated from a set of variables $X$, via some unknown linear apparatus $F$.
+%
+Not all the variables in $X$ exert a causal influence on $Y$.
+%
+By considering a square binary diagonal matrix of \emph{causal influences} $E$, we denote by $XE$ the causal factors of $Y$.
+%
+In summary, our model can be formalized as:
+%
+\begin{equation}
+    Y = (XE + N)F + N',\label{eq:model}
+\end{equation}
+%
+where $N$ and $N'$ are some unobserved noise variables.
+%
+In practice, we observe $n$ samples $(X, Y)$ from the model.
+%
+This process, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.
 
-Without loss of generality, we may assume N to be centred(adding, when necessary, an intercept to the causal features), and redefine features and noise so that the model be written $Y=f(EX+N)$ (f a function, E, X and N matrices).
+Given the model in Equation~\eqref{eq:model}, the goal of Back-to-Back Regression (B2BR) is to estimate the matrix of causal influences $E$.
+
+%Without loss of generality, we may assume N to be centred(adding, when necessary, an intercept to the causal features), and redefine features and noise so that the model be written $Y=f(EX+N)$ (f a function, E, X and N matrices).
 In particular, the addition of "external noise" M, as in $Y=F(EX+N)+M$ can be viewed as a small change in F and N.
 
-We consider both the linear and non linear case.
+%We consider both the linear and non linear case.
 In the first one, our model is $$Y = F(EX+N)$$, with  $Y\in\mathbb{R}^{n\times ny}$, $F\in\mathbb{R}^{ny\times nx}$, $X\in\mathbb{R}^{nx\times n}$, $N\in\mathbb{R}^{nx\times n}$, and E a binary diagonal square matrice of size nx.
@@
 
-Under this formalism, we are given samples of Y and X, and are tasked to recover E and use it to predict Y from X.
+%Under this formalism, we are given samples of Y and X, and are tasked to recover E and use it to predict Y from X.
 
@@
 
-Back-to-back regression consists of two steps.
+Back-to-Back Regression (B2BR) consists in two steps.
 %
-First, we perform a least squares estimation of X from Y, that is we find $\hat G$ such that $\hat X=\hat G Y$ is the least square predictor of X.
+First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$, and construct the predictions $\hat X = \hat Y \hat G$.
 %
-This recovers an approximation of the causes of Y in X.
+This backward regression recovers an approximation of the causes of $Y$ in $X$.
 %
-Then, we perform a least squares estimation of $\hat X$ from X, and determine $\hat H$ such that $\hat {\hat X}=\hat H X$ is the least square predictor of $\hat X$.
+Second, we estimate the linear regression coefficients $\hat H$ from $X$ to $\hat X$.
 %
-In matrix form, we have $\hat G=(Y'Y)^{-1} Y'X$, and
-\begin{equation} \hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X\end{equation}
+The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{Diag}(E)$, is the desired estimate of the causal influence matrix $E$.
 
-If, in the above formalism, F is invertible with respect to the causal features of X (ie the problem has a solution), and the relation is linear or close to linear, the diagonal of $\hat H$ will be a scaled approximation of the diagonal of E.
+If using regularized least-squares, arguably the most commonly employed linear regression technique, the solution of B2BR has a closed form solution.
+%
+In particular, $\hat G = (Y^\top Y + \Lambda_Y)^{-1} Y^\top X$ and $\hat H=(X^\top X + \Lambda_X)^{-1} X^\top YG$, where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices containing regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.
+
+% If, in the above formalism, F is invertible with respect to the causal features of X (ie the problem has a solution), and the relation is linear or close to linear, the diagonal of $\hat H$ will be a scaled approximation of the diagonal of E.
 That is to say, back to back regression will recover the causal features of the process considered, even in the presence of heavy noise N, and without needing to estimate the transformation F.
 
-In practical cases, the covariance matrices of Y and X will often be badly conditioned (because features of X are related, or measurement produces correlated features of Y).
+% In practical cases, the covariance matrices of Y and X will often be badly conditioned (because features of X are related, or measurement produces correlated features of Y).
 This makes the inversion of $X'X$ and $Y'Y$ sensitive to noise (either from N, from sampling or from computation rounding).
@@
 
-In both, we compute the singular value decompositions of the covariance matrix, transform its diagonal part, and invert it.
+% In both, we compute the singular value decompositions of the covariance matrix, transform its diagonal part, and invert it.
 In truncated SVD, the lowest eigenvalues are set to zero, and the corresponding vectors eliminated from future calculations.
@@
 
-Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data are detected in the first step and learnt in the second.
-To avoid this, we split our sample into two halves, and each regression is performed over one, effectively decorrelating the two steps.
+Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
 %
-To compensate for the reduction in sample size caused by the split, this process is repeated over many splits, and their results are averaged.
-This bagging compensates for the reduction in sample size caused by splitting.
+To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
+%
+Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
+%
+To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split.
 
-As we shall see, the diagonal of $\hat H$ recovered by back to back regression is a scaled approximation of the diagonal of E.
-Noise tends reduce the large diagonal elements, whereas the bias introduced by regularisation tends to increase the lowest ones.
+We summarize the B2BR procedure in Algorithm~\ref{algorithm:b2br}.
+%
+The rest of this section provides a theoretical guarantee on the correctness of B2BR, and an optional post-processing step to binarize the estimated causal influence matrix.
 
-To recover an estimate of E, we need a test to select the largest diagonal elements of $\hat H$.
-%
-In regression analysis, a traditional approach to this problem consists in testing whether coefficients differ from zero, via variants of Student t-test.
-%
-This will not succeed here.
-First because the conditions for these tests do not apply here (we have two regressions).
-Second, because regularisation adds bias on the diagonal, which will cause t-tests to succeed unduly.
-%
-Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).
-This can be done by maximizing the ratio of inter-group variance over total variance (ie minimizing intra-group variance), over all splits of the p largest data and the n-p smallest.
+\begin{algorithm}[H]
+    %\SetAlgoLined
+    \KwIn{input data $X \in \mathbb{R}^{n \times d_x}$, output data $Y \in \mathbb{R}^{n\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
+    \KwOut{estimate of causal influences $\hat{E} \in \mathbb{D}^{d_x \times d_x}$.}
+    $\hat{E} \leftarrow 0$\;
+    \For{$i = 1, \ldots, m$}{
+        $(X, Y) \leftarrow \text{ShuffleRows}((X, Y))$\;
+        $(X_1, Y_1), (X_2, Y_2) \leftarrow \text{SplitRowsInHalf}((X, Y))$\;
+        $\hat{G} = \text{LinearRegression}(Y_1, X_1)$ \Comment*[r]{$\hat G = (Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1$}
+        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$\;
+        $E \leftarrow E + \text{Diag}(H)$\;
+    }
+    \Return{$\hat{E} / m$}
+    \caption{Back-to-back regression.}
+    \label{algorithm:b2br}
+\end{algorithm}
 
-If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we select the split that maximises the Sonquist and Morgan criterion $$K = \frac{p (n-p)}{n} \frac{(s-r)^2}{ V}$$.
-%
-This recovers from $\hat H$ a binary diagonal $\hat E$.
-%
-To use back-to-back regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
-
-Summarizing, back-to-back regression is performed as follows:
-\begin{enumerate}
-\item Randomly split the data sample.
-\item From the first half sample, using crossalidated ridge regression, find $\hat G$, the regularised least square regression of X from Y: $\hat G=(Y'Y)^{-1} Y'X$.
-\item From the second half sample, calculate $\hat X = \hat G Y$, and find $\hat H$, the regularized least square regression of $\hat X$ from X: $\hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X$.
-\item Repeat steps 1 to 3 and average the results as $\hat H$.
-\item Using the Sonquist and Morgan criterion, select the large diagonal elements of $\hat H$, set them to one and the rest of $\hat  H$ to zero.
-\item Use the resulting matrix $\hat E$ as an estimator of E, the active features in the causal process
-\item Alternatively, use $\hat E$ to predict Y from X, by performing a regularised least square regression of Y from $\hat E X$.
-%
-
-\end{enumerate}
-
-\subsection{Asymptotic behaviour - linear case}
+\subsection{Theoretical guarantees}
 As discussed in the introduction, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(EX+N)$, with f an unknown function.
@@
 
+As we shall see, the diagonal of $\hat H$ recovered by back to back regression is a scaled approximation of the diagonal of E.
+Noise tends reduce the large diagonal elements, whereas the bias introduced by regularisation tends to increase the lowest ones.
 % \subsection{Asymptotic behaviour - non linear case}
+
+\subsection{Use cases}
+
+To recover an estimate of E, we need a test to select the largest diagonal elements of $\hat H$.
+%
+In regression analysis, a traditional approach to this problem consists in testing whether coefficients differ from zero, via variants of Student t-test.
+%
+This will not succeed here.
+First because the conditions for these tests do not apply here (we have two regressions).
+Second, because regularisation adds bias on the diagonal, which will cause t-tests to succeed unduly.
+%
+Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).
+This can be done by maximizing the ratio of inter-group variance over total variance (ie minimizing intra-group variance), over all splits of the p largest data and the n-p smallest.
+
+If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we select the split that maximises the Sonquist and Morgan criterion $$K = \frac{p (n-p)}{n} \frac{(s-r)^2}{ V}$$.
+%
+This recovers from $\hat H$ a binary diagonal $\hat E$.
+%
+To use back-to-back regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
 

### 2019-05-22 paper/paper.tex
Proof stuff

@@
 %
+While the triplet of variables $X$, $N$, and $N'$ are independent, we allow each of them to have a general covariance matrix.
+%
 In practice, we observe $n$ samples $(X, Y)$ from the model.
@@
 \subsection{Theoretical guarantees}
-As discussed in the introduction, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(EX+N)$, with f an unknown function.
+% As discussed in the introduction, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(EX+N)$, with f an unknown function.
+% %
+% If F is linear we can rewrite it as $Y = F(EX + N)$, with X an (dx, n) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
+% %
+% N a (dx,n) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dx) transformation corresponding to the measuring apparatus, and Y a (dy, n) matrix of measured effects.
 % 
-If F is linear we can rewrite it as $Y = F(EX + N)$, with X an (dx, n) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
+% The two steps of back-to-back regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).
+
+Assuming centered noise $N$,
+\begin{align*}
+    \arg \min_G \left \| YG - X \right \|^2 &= \arg \min_G \left \| (EX + N)FG \right\|^2 \\
+                                            &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
+\end{align*}
 %
-N a (dx,n) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dx) transformation corresponding to the measuring apparatus, and Y a (dy, n) matrix of measured effects.
+As the noise $N$ tends to zero, the solution to the previous problem is $\hat G = (X E F)^\dagger X$, where ``$\dagger$'' denotes the Moore-Penrose pseudo-inverse.
 
-The two steps of back-to-back regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).
+Next, let $k$ be the number of ones in the causal influence matrix $E$.
+%
+Without loss of generality, assume that these are the first $k$ variables in $X$.
+%
+Assume that $XEF$ is of rank $k$, which means that $F$ and $X$ have full rank on the space spanned by the active elements in $E$.
+%
+Let $K \in \mathbb{R}^{d_x \times k}$ be the diagonal matrix obtained by deleting the right-most $d_x - k$ columns of $E$.
+%
+Then, $E = KK^\top$, and $XEF = (XK)(K^\top F)$, with both $XK$ and $K^\top$ of rank $k$.
 
-For the first step, replacing Y, and since N is centred, we have
-\begin{equation}
-\begin{aligned}
-\arg \min_G \left \| X-GY \right \|^2 &= \arg \min_G \left \| X - GF(EX+N)\right\|^2 \\
-&{}= \arg \min_G \left \| X-GFEX\right\| ^2  + \left \| GFN\right \| ^2
-\end{aligned}
-\end{equation}
-
-As N tends to zero, we are searching for the minimal norm solution of $\min_G \left \| X-GFEX\right\| ^2$, which is $X(FEX)^\dagger$ ($M^\dagger$ denoting the pseudo inverse of M).
-
-Let k be the number of active features of E, suppose for notational convenience that they are the first k features of X, and let $FEX$ be of rank k (which means F and X have full rank on the space spanned by the active features of E).
-Let K be the $\mathbb{R}^{nx\times k}$ diagonal matrix obtained by deleting the rightmost nx-k columns of E.
-Then, $E=KK'$, and $FEX=(FK)(K'X)$, with $FK$  and $K'X$ of full rank k.
-
-The rank factorisation property of the pseudo inverse (ref) says that, if $A=BC$ with $A\in\mathbb{R}^{m\times n}$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$ of rank k, then $A^\dagger=C^\dagger B^\dagger$ and since C is full rank, $C^\dagger=C' (CC')^{-1}$.
-
-Replacing in the expression of the minimal norm solution yields
+Recall the rank factorisation property of the pseudo-inverse \citep{todo}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $C$ is full rank, $C^\dagger=C' (CC')^{-1}$.
+%
+Using this property on the minimal norm solution $\hat G$ yields
 \begin{equation}

### 2019-05-22 paper/paper.tex
Sonquist

@@
 
-\subsection{Use cases}
+\subsection{Post-processing of B2BR}
 
-To recover an estimate of E, we need a test to select the largest diagonal elements of $\hat H$.
+On the one hand, we have leveraged B2BR to obtain the estimate $\hat{E}$, which is a diagonal matrix with real entries.
 %
-In regression analysis, a traditional approach to this problem consists in testing whether coefficients differ from zero, via variants of Student t-test.
+On the other hand, the true causal influence matrix $E$ is a binary matrix, which hard-selects the causal factors of $Y$ from $X$.
 %
-This will not succeed here.
-First because the conditions for these tests do not apply here (we have two regressions).
-Second, because regularisation adds bias on the diagonal, which will cause t-tests to succeed unduly.
+In this section, we provide a recipe to binarize $\hat{E}$ to estimate the collection of causal factors.
+
+In regression analysis, the traditional approach to this problem consists in testing whether the regression coefficients differ from zero.
 %
-Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).
-This can be done by maximizing the ratio of inter-group variance over total variance (ie minimizing intra-group variance), over all splits of the p largest data and the n-p smallest.
-
-If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we select the split that maximises the Sonquist and Morgan criterion $$K = \frac{p (n-p)}{n} \frac{(s-r)^2}{ V}$$.
+However, this test will not succeed here, since the estimate $\hat{E}$ is obtained from a double regression, and any employed regularization will add a bias to the diagonal of $\hat{E}$.
 %
-This recovers from $\hat H$ a binary diagonal $\hat E$.
+Instead, we treat the binarization of $\hat{E}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''.
 %
-To use back-to-back regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
+More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
+%
+Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
+%
+To binarize $\hat{E}$, set to one all the diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the diagonal entries.
 

### 2019-05-22 paper/paper.tex
backwards theorem

--- 2019-05-22 paper/paper.tex
@@
 \usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx}
-\usepackage[ruled, linesnumbered]{algorithm2e}
-
 \usepackage[hidelinks]{hyperref}
@@
 \DeclareMathOperator{\Tr}{Tr}
-
-\SetKwComment{Comment}{$\triangleright$\ }{}
 
@@
 
-\title{Measuring causal influence with\\ back-to-back regression}
+\title{Discovering causal factors with back-to-back regression}
 
@@
 
-    \node[] (labY) at (\posY, 3){Observations};
-    \node[] (labX) at (\posX, 3){Factors};
-    \node[] (labE) at (\posE, 3.3){Cause};
-    \node[] (labE) at (\posE, 3){selection};
-    \node[] (labN) at (\posN, 3){Noise};
-    \node[] (labF) at (\posF, 3.3){Cause-effect};
-    \node[] (labF) at (\posF, 3){mapping};
-
-    \node[] (labY) at (\posY, 2)   {\color{gray} MEG recordings};
-    \node[] (labX) at (\posX, 2)   {\color{gray} stimulus};
-    \node[] (labX) at (\posX, 2.3) {\color{gray} features};
-    \node[] (labE) at (\posE, 2.3) {\color{gray} causal};
-    \node[] (labE) at (\posE, 2)   {\color{gray} features};
-    \node[] (labN) at (\posN, 2)   {\color{gray} breathing\ldots};
-    \node[] (labF) at (\posN, 2.3) {\color{gray} bloodflow,};
-    \node[] (labF) at (\posF, 2)   {\color{gray} MEG helmet};
+    \node[] (labY) at (\posY, 2){Observations};
+    \node[] (labX) at (\posX, 2){Factors};
+    \node[] (labE) at (\posE, 2.3){Cause};
+    \node[] (labE) at (\posE, 2){selection};
+    \node[] (labN) at (\posN, 2){Noise};
+    \node[] (labF) at (\posF, 2.3){Cause-effect};
+    \node[] (labF) at (\posF, 2){mapping};
 
@@
     \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
-    \node[] (reg1) at (5.5,-3.25) {$\hat{E} = \text{Diag}(\underbrace{(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}})$};
+    \node[] (reg1) at (5,-3.25) {$\hat{E} = \underbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}} $};
     \end{tikzpicture}
-    \caption{Back-to-back regression identifies the subset of factors $E_{ii} = 1$ in $X$ that influence some observations $Y$ by 1) regressing from $Y$ to $X$ to obtain $\hat{X}$, and ii) returning the diagonal of the regression coefficients from $X$ to $\hat{X}$.}
-    \label{fig:b2b}
+    \caption{In its linear form, back-to-back regression identifies the subset of factors that suffices to account for multidimensional observations by piping two successive regressions from $Y$ to $X$ and from $X$ to $\hat X$.
+}
+    \label{fig:}
 \end{figure}
 
-We consider the measurement of multivariate signal $Y$, generated from a set of variables $X$, via some unknown linear apparatus $F$.
-%
-Not all the variables in $X$ exert a causal influence on $Y$.
-%
-By considering a square binary diagonal matrix of \emph{causal influences} $E$, we denote by $XE$ the causal factors of $Y$.
-%
-In summary, our model can be formalized as:
-%
-\begin{equation}
-    Y = (XE + N)F + N',\label{eq:model}
-\end{equation}
-%
-where $N$ and $N'$ are some unobserved noise variables.
-%
-In practice, we observe $n$ samples $(X, Y)$ from the model.
-%
-This process, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.
+\subsection{Formalism}
+Via an apparatus F, we are measuring a multidimensional signal Y, which is a function of possible causes X and other non causal features N.
+Not all features of X exert causal influence on Y.
+Let $XE$ (E a matrix of zero and one) the subset of causal factors.
+Our model can be formalised as $Y=f(XE,N)$.
 
-Given the model in Equation~\eqref{eq:model}, the goal of Back-to-Back Regression (B2BR) is to estimate the matrix of causal influences $E$.
+Without loss of generality, we may assume N to be centred(adding, when necessary, an intercept to the causal features), and redefine features and noise so that the model be written $Y=f(XE+N)$ (f a function, E, X and N matrices).
+In particular, the addition of "external noise" M, as in $Y=(XE+N)F+M$ can be viewed as a small change in F and N.
 
-%Without loss of generality, we may assume N to be centred(adding, when necessary, an intercept to the causal features), and redefine features and noise so that the model be written $Y=f(EX+N)$ (f a function, E, X and N matrices).
-In particular, the addition of "external noise" M, as in $Y=F(EX+N)+M$ can be viewed as a small change in F and N.
+We consider both the linear and non linear case.
+In the first one, our model is $$Y = (XE+N)F$$with  $Y\in\mathbb{R}^{n\times n_{y}}$, $F\in\mathbb{R}^{n_{x}\times n_{y}}$, $X\in\mathbb{R}^{n\times n_{x}}$, $N\in\mathbb{R}^{n\times n_{x}}$, and E a binary diagonal square matrix of size $n_{x}$.
 
-%We consider both the linear and non linear case.
-In the first one, our model is $$Y = F(EX+N)$$, with  $Y\in\mathbb{R}^{n\times ny}$, $F\in\mathbb{R}^{ny\times nx}$, $X\in\mathbb{R}^{nx\times n}$, $N\in\mathbb{R}^{nx\times n}$, and E a binary diagonal square matrice of size nx.
+In the non linear model, $(XE+N)F$ is defined as before, but Y is the result of a non linear transformation of it $$Y=\sigma ((XE+N)F)$$
 
-% In the non linear model, $F(EX+N)$ is defined as before, but Y is the result of a non linear transformation of it $$Y=\sigma (F(EX+N))$$
-
-%Under this formalism, we are given samples of Y and X, and are tasked to recover E and use it to predict Y from X.
+Under this formalism, we are given samples of Y and X, and are tasked to recover E and use it to predict Y from X.
 
@@
 
-Back-to-Back Regression (B2BR) consists in two steps.
+Back-to-back regression consists of two steps.
 %
-First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$, and construct the predictions $\hat X = \hat Y \hat G$.
+First, we perform a least squares estimation of X from Y, that is we find $\hat G$ such that $\hat X=Y\hat G$ is the least square predictor of X.
 %
-This backward regression recovers an approximation of the causes of $Y$ in $X$.
+This recovers an approximation of the causes of Y in X.
 %
-Second, we estimate the linear regression coefficients $\hat H$ from $X$ to $\hat X$.
+Then, we perform a least squares estimation of $\hat X$ from X, and determine $\hat H$ such that $\hat {\hat X}=\hat H X$ is the least square predictor of $\hat X$.
 %
-The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{Diag}(E)$, is the desired estimate of the causal influence matrix $E$.
+In matrix form, we have $\hat G=(Y'Y)^{-1} Y'X$, and
+\begin{equation} \hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X\end{equation}
 
-If using regularized least-squares, arguably the most commonly employed linear regression technique, the solution of B2BR has a closed form solution.
-%
-In particular, $\hat G = (Y^\top Y + \Lambda_Y)^{-1} Y^\top X$ and $\hat H=(X^\top X + \Lambda_X)^{-1} X^\top YG$, where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices containing regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.
-
-% If, in the above formalism, F is invertible with respect to the causal features of X (ie the problem has a solution), and the relation is linear or close to linear, the diagonal of $\hat H$ will be a scaled approximation of the diagonal of E.
+If, in the above formalism, F is invertible with respect to the causal features of X (ie the problem has a solution), and the relation is linear or close to linear, the diagonal of $\hat H$ will be a scaled approximation of the diagonal of E.
 That is to say, back to back regression will recover the causal features of the process considered, even in the presence of heavy noise N, and without needing to estimate the transformation F.
 
-% In practical cases, the covariance matrices of Y and X will often be badly conditioned (because features of X are related, or measurement produces correlated features of Y).
+In practical cases, the covariance matrices of Y and X will often be badly conditioned (because features of X are related, or measurement produces correlated features of Y).
 This makes the inversion of $X'X$ and $Y'Y$ sensitive to noise (either from N, from sampling or from computation rounding).
@@
 
-% In both, we compute the singular value decompositions of the covariance matrix, transform its diagonal part, and invert it.
+In both, we compute the singular value decompositions of the covariance matrix, transform its diagonal part, and invert it.
 In truncated SVD, the lowest eigenvalues are set to zero, and the corresponding vectors eliminated from future calculations.
@@
 
-Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
+Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data are detected in the first step and learnt in the second.
+To avoid this, we split our sample into two halves, and each regression is performed over one, effectively decorrelating the two steps.
 %
-To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
+To compensate for the reduction in sample size caused by the split, this process is repeated over many splits, and their results are averaged.
+This bagging compensates for the reduction in sample size caused by splitting.
+
+As we shall see, the diagonal of $\hat H$ recovered by back to back regression is a scaled approximation of the diagonal of E.
+Noise tends reduce the large diagonal elements, whereas the bias introduced by regularisation tends to increase the lowest ones.
+
+To recover an estimate of E, we need a test to select the largest diagonal elements of $\hat H$.
 %
-Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
+In regression analysis, a traditional approach to this problem consists in testing whether coefficients differ from zero, via variants of Student t-test.
 %
-To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split.
+This will not succeed here.
+First because the conditions for these tests do not apply here (we have two regressions).
+Second, because regularisation adds bias on the diagonal, which will cause t-tests to succeed unduly.
+%
+Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).
+This can be done by maximizing the ratio of inter-group variance over total variance (ie minimizing intra-group variance), over all splits of the p largest data and the n-p smallest.
 
-We summarize the B2BR procedure in Algorithm~\ref{algorithm:b2br}.
+If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we select the split that maximises the Sonquist and Morgan criterion $$K = \frac{p (n-p)}{n} \frac{(s-r)^2}{ V}$$.
 %
-The rest of this section provides a theoretical guarantee on the correctness of B2BR, and an optional post-processing step to binarize the estimated causal influence matrix.
+This recovers from $\hat H$ a binary diagonal $\hat E$.
+%
+To use back-to-back regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $X \hat E $.
 
-\begin{algorithm}[H]
-    %\SetAlgoLined
-    \KwIn{input data $X \in \mathbb{R}^{n \times d_x}$, output data $Y \in \mathbb{R}^{n\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
-    \KwOut{estimate of causal influences $\hat{E} \in \mathbb{D}^{d_x \times d_x}$.}
-    $\hat{E} \leftarrow 0$\;
-    \For{$i = 1, \ldots, m$}{
-        $(X, Y) \leftarrow \text{ShuffleRows}((X, Y))$\;
-        $(X_1, Y_1), (X_2, Y_2) \leftarrow \text{SplitRowsInHalf}((X, Y))$\;
-        $\hat{G} = \text{LinearRegression}(Y_1, X_1)$ \Comment*[r]{$\hat G = (Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1$}
-        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$\;
-        $E \leftarrow E + \text{Diag}(H)$\;
-    }
-    \Return{$\hat{E} / m$}
-    \caption{Back-to-back regression.}
-    \label{algorithm:b2br}
-\end{algorithm}
+Summarizing, back-to-back regression is performed as follows:
+\begin{enumerate}
+\item Randomly split the data sample.
+\item From the first half sample, using cross-validated ridge regression, find $\hat G$, the regularised least square regression of X from Y: $\hat G=(Y'Y)^{-1} Y'X$.
+\item From the second half sample, calculate $\hat X = Y \hat G$, and find $\hat H$, the regularised least square regression of $\hat X$ from X: $\hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X$.
+\item Repeat steps 1 to 3 and average the results as $\hat H$.
+\item Using the Sonquist and Morgan criterion, select the large diagonal elements of $\hat H$, set them to one and the rest of $\hat  H$ to zero.
+\item Use the resulting matrix $\hat E$ as an estimator of E, the active features in the causal process
+\item Alternatively, use $\hat E$ to predict Y from X, by performing a regularised least square regression of Y from $X \hat E$.
+%
 
-\subsection{Theoretical guarantees}
-As discussed in the introduction, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(EX+N)$, with f an unknown function.
+\end{enumerate}
+
+\subsection{Asymptotic behaviour - linear case}
+As discussed above, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(XE+N)$, with f an unknown function.
 %
-If F is linear we can rewrite it as $Y = F(EX + N)$, with X an (dx, n) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
+If F is linear we can rewrite it as $Y = (XE + N)F$, with $X\in\mathbb{R}^{n\times n_{x}}$ a matrix of possible causes, E a  binary diagonal $n_{x}$ square matrix. selecting active features of X.
 %
-N a (dx,n) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dx) transformation corresponding to the measuring apparatus, and Y a (dy, n) matrix of measured effects.
+$N\in\mathbb{R}^{n\times n_{x}}$, a homoscedastic noise matrix, $F\in\mathbb{R}^{n_{x}\times n_{y}}$ an unknown transformation corresponding to the measuring apparatus, and $Y\in\mathbb{R}^{n\times n_{y}}$ a matrix of measured effects.
 
-The two steps of back-to-back regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).
+The two steps of back-to-back regression amounts to finding $\hat G=\arg \min_G \left \| X-YG \right \|^2$ and $\hat H =\arg \min_H \left \| \hat YG - XH \right \|^2$ (all matrix norms henceforth are Frobenius norms).
 
@@
 \begin{aligned}
-\arg \min_G \left \| X-GY \right \|^2 &= \arg \min_G \left \| X - GF(EX+N)\right\|^2 \\
-&{}= \arg \min_G \left \| X-GFEX\right\| ^2  + \left \| GFN\right \| ^2
+\arg \min_G \left \| X-YG \right \|^2 &= \arg \min_G \left \| X - (XE+N)FG\right\|^2 \\
+&{}= \arg \min_G \left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2
 \end{aligned}
@@
 
-As N tends to zero, we are searching for the minimal norm solution of $\min_G \left \| X-GFEX\right\| ^2$, which is $X(FEX)^\dagger$ ($M^\dagger$ denoting the pseudo inverse of M).
+As N tends to zero, we are searching for the minimal norm solution of $\min_G \left \| X-XEFG\right\| ^2$, which is $(XEF)^\dagger X$ ($M^\dagger$ denoting the pseudo inverse of M).
 
-Let k be the number of active features of E, suppose for notational convenience that they are the first k features of X, and let $FEX$ be of rank k (which means F and X have full rank on the space spanned by the active features of E).
-Let K be the $\mathbb{R}^{nx\times k}$ diagonal matrix obtained by deleting the rightmost nx-k columns of E.
-Then, $E=KK'$, and $FEX=(FK)(K'X)$, with $FK$  and $K'X$ of full rank k.
+Let k be the number of active features of E, suppose for notational convenience that they are the first k features of X, and let $XEF$ have rank k (which means F and X have full rank on the space spanned by the active features of E).
+Let K be the $\mathbb{R}^{k\times N_{x}}$ diagonal matrix obtained by deleting the bottom $n_{x}-k$ lines of E.
+Then, $E=K'K$, and $XEF=(XK')(KF)$, with $KF$  and $XK'$ of full rank k.
 
@@
 \begin{aligned}
-X(FEX)^\dagger &=X((FK)(K'X))^\dagger \\
-&=X(K'X)'((K'X)(K'X)')^{-1}(FK)^\dagger \\
-&=(XX')K(K'XX'K)^{-1}(FK)^\dagger 
+(XEF)^\dagger X&=((XK')(KF))^\dagger X\\
+&=(FK)^\dagger (XK')'((XK')(XK')')^{-1} X\\
+&=(FK)^\dagger KX' (K'XX'K)^{-1} X
 \end{aligned}
@@
 
-As we shall see, the diagonal of $\hat H$ recovered by back to back regression is a scaled approximation of the diagonal of E.
-Noise tends reduce the large diagonal elements, whereas the bias introduced by regularisation tends to increase the lowest ones.
-% \subsection{Asymptotic behaviour - non linear case}
-
-\subsection{Use cases}
-
-To recover an estimate of E, we need a test to select the largest diagonal elements of $\hat H$.
-%
-In regression analysis, a traditional approach to this problem consists in testing whether coefficients differ from zero, via variants of Student t-test.
-%
-This will not succeed here.
-First because the conditions for these tests do not apply here (we have two regressions).
-Second, because regularisation adds bias on the diagonal, which will cause t-tests to succeed unduly.
-%
-Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).
-This can be done by maximizing the ratio of inter-group variance over total variance (ie minimizing intra-group variance), over all splits of the p largest data and the n-p smallest.
-
-If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we select the split that maximises the Sonquist and Morgan criterion $$K = \frac{p (n-p)}{n} \frac{(s-r)^2}{ V}$$.
-%
-This recovers from $\hat H$ a binary diagonal $\hat E$.
-%
-To use back-to-back regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.
-
-\section{Experiments}
-\subsection{Synthetic data}
-
-\input{synthetic.tex}
-
-\subsection{Magnetoencephalography data}
-
-\input{meg_exps.tex}
+\subsection{Asymptotic behaviour - non linear case}
 
 \section{Related works}
-\subsection{back-to-back regression as a special case of canonical component analysis?}
+\subsection{Back-to-back regression as a special case of canonical component analysis}
 Formula (1) shows that back-to-back regression may be related to canonical component analysis (CCA).
@@
 
+\section{Experiments - simulated data}
+\input{synthetic.tex}
+
+\section{Experiments - real case}
+Jean Remi and Maxime territory
+
+\input{meg_exps.tex}
+
 \section{Conclusion}
 
-% \section{Acknowlegements}
-% Data were provided (in part) by the Donders Institute for Brain, Cognition and Behaviour.
+\section{Acknowlegements}
+Data were provided (in part) by the Donders Institute for Brain, Cognition and Behaviour.
 
@@
 
-%\section{Appendices}
-%
-%This should hold part of the tests, explanations on simulation, detailed results and stuff on meg data
+\section{Appendices}
+
+This should hold part of the tests, explanations on simulation, detailed results and stuff on meg data
 

### 2019-05-22 paper/paper.tex

--- 2019-05-22 paper/paper.tex
@@
 \section{Related works}
-\subsection{back-to-back regression as a special case of canonical component analysis?}
+\subsection{Back-to-back regression as a special case of canonical component analysis}
 Formula (1) shows that back-to-back regression may be related to canonical component analysis (CCA).

### 2019-05-22 paper/meg_exps.tex
Fix
Merging...

@@
-\newpage
-\subsection{Dataset}
+%\newpage
+%\subsection{Dataset}
 
@@
 neuroimaging ``Mother Of all Unification Studies'' (MOUS) dataset
-(\cite{schoffelen2019204}).  The dataset contains magneto-encephalography (MEG)
+\citep{schoffelen2019204}.  The dataset contains magneto-encephalography (MEG)
 recordings of 102 healthy native-Dutch adults who participated in a reading
@@
 
-\subsection{MEG preprocessing}
+\subsubsection{MEG preprocessing}
 
-The raw MEG data was bandpass-filtered between 0.1 and 40 Hz using MNE-Python
-default parameters (\cite{gramfort2013meg, gramfort2014mne}). Specifically, we used a zero-phase finite impulse
-response filter (FIR) with a Hamming window and with transition bands of 0.1 Hz
-and 10 Hz for the low and high cut-off frequencies respectively.
+The raw MEG data was bandpass-filtered between 0.1 and 40Hz using MNE-Python
+default parameters \citep{gramfort2013meg, gramfort2014mne}. Specifically, we used a zero-phase finite impulse
+response filter (FIR) with a Hamming window and with transition bands of 0.1Hz
+and 10Hz for the low and high cut-off frequencies respectively.
 
-The raw data was then segmented 100 ms before word onset and 1.000 sec after
+The raw data was then segmented 100ms before word onset and 1.000 sec after
 word onset. Time $t=0$ms corresponds to the word onset. Finally, each resulting
 segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
-thus led a sampling frequency of 240 Hz. We show, in Figure \ref{fig:megavg},
+thus led a sampling frequency of 240Hz. Figure \ref{fig:megavg} shows
 the average response of the magnetometers evoked by the words presented to the
@@
 
-\subsection{Word features} We annotated each word with 38 features: 26
+\subsubsection{Word features} We annotated each word with 38 features: 26
 correspond to the count of each letter in the word (from 'a' to 'z',
@@
 frequency of the word in natural language (using the Zipf logarithmic scale
-of \cite{van2014subtlex}), and the part-of-speech tags (i.e. NOUN, PROPN (proper
+of \citep{van2014subtlex}), and the part-of-speech tags (i.e. NOUN, PROPN (proper
 noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ
 (adjective), ADV (adverb), and NUM (numerical)) using the packages Spacy
-(\cite{spacy2}) and WordFreq (\cite{speerwordfreq}). For clarity purposes, we refer to these
+\citep{spacy2} and WordFreq \citep{speerwordfreq}. For clarity purposes, we refer to these
 features using two broad categories: visual (i.e. each letter and the total
@@
 
-\subsection{Analyses} We attempt to identify which factors of $X$ explain the MEG
+\subsubsection{Methods}
+
+We attempt to identify which factors of $X$ explain the MEG
 recordings $Y$ at each time sample and for each subject separately. To this aim,
-we implement three types of analyses (forward, backward and back-to-back
+we compare three methods (forward, backward and back-to-back
 regressions). Corresponding results can be found in Figure \ref{fig:meg_twocurves}.
@@
   previous neuroscientific work showing an early response ($\approx$ 150ms) to
-  length and a later response ($\approx$ 400ms) to frequency. See text for
-  details.}
+  length and a later response ($\approx$ 400ms) to frequency.}
   \label{fig:meg_twocurves}
@@
 
-\paragraph{Forward regression} also known as 'encoding' in neuroimaging, here consists of
-fitting a ridge-regularized ordinary least squares (OLS) regression \footnote{We use scikit-learn (\cite{sklearn}).} from $X$ to $Y$:
-\begin{equation}
-  H = (X^{T}X+\lambda I)^{-1} X^{T}Y
-\end{equation}
-where $\lambda$ is the
-regularization parameter, cross-validated between $10^{-5}$ and $10^5$.
+Forward regression performs a regularized least-squares regression $A \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| A_{i, :} \|^2$.
+%
+Backward regression performs a regularized least-squares regression $B \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YB$, using a validation sample.
+%
 
-We then estimate the contribution
-$diag(\hat E) \in\mathbb{D}^{d} $ of each factor. However, $H\in\mathbb{R}^{d,
-q}$ is a matrix of $d$ word features by $q$ MEG channels. We approximate
-the relative contribution of each feature with the sum-of-squares of the corresponding coefficients in $H$; for the $j$-th feature:
-\begin{equation}
-  \hat e_j = \sum_{i=0}^{q} {h_{i,j}}^2
-\end{equation}
-\paragraph{Backward regression} or 'decoding', is similar but predicts $X$ from $Y$:
-\begin{equation} G = (Y^{T}Y+\lambda I)^{-1} Y^{T}X \end{equation}
-where $\lambda$ is cross-validated as for encoding. We estimate the
-contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor in the following
-way. We predict $X$ on a held-out split \textit{test}:
-% Using 10-fold cross-validation, we thus assessed the ability of this ridge
-% regression to predict independent and identically distributed data:
-\begin{equation} \hat X_{test} = Y_{test}\cdot G_{train} \end{equation}
-and assess, for each factor of X independently, the Pearson correlation
-coefficient $r$ between $X_{test}$ and $\hat X_{test}$. We repeat the procedure
-for 20 random splits and average the result.
+% \paragraph{Forward regression} also known as 'encoding' in neuroimaging, here consists of
+% fitting a ridge-regularized ordinary least squares (OLS) regression \footnote{We use scikit-learn \citep{sklearn}.} from $X$ to $Y$:
+% \begin{equation}
+%   H = (X^{T}X+\lambda I)^{-1} X^{T}Y
+% \end{equation}
+% where $\lambda$ is the
+% regularization parameter, cross-validated between $10^{-5}$ and $10^5$.
+% 
+% We then estimate the contribution
+% $diag(\hat E) \in\mathbb{D}^{d} $ of each factor. However, $H\in\mathbb{R}^{d,
+% q}$ is a matrix of $d$ word features by $q$ MEG channels. We approximate
+% the relative contribution of each feature with the sum-of-squares of the corresponding coefficients in $H$; for the $j$-th feature:
+% \begin{equation}
+%   \hat e_j = \sum_{i=0}^{q} {h_{i,j}}^2
+% \end{equation}
+% \paragraph{Backward regression} or 'decoding', is similar but predicts $X$ from $Y$:
+% \begin{equation} G = (Y^{T}Y+\lambda I)^{-1} Y^{T}X \end{equation}
+% where $\lambda$ is cross-validated as for encoding. We estimate the
+% contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor in the following
+% way. We predict $X$ on a held-out split \textit{test}:
+% % Using 10-fold cross-validation, we thus assessed the ability of this ridge
+% % regression to predict independent and identically distributed data:
+% \begin{equation} \hat X_{test} = Y_{test}\cdot G_{train} \end{equation}
+% and assess, for each factor of X independently, the Pearson correlation
+% coefficient $r$ between $X_{test}$ and $\hat X_{test}$. We repeat the procedure
+% for 20 random splits and average the result.
 
@@
 
-\subsection{Second-level statistics}
+\subsubsection{Second-level statistics}
 
@@
 sample independently.  To estimate the reliability of these estimates in the
-population, we tested, using one-sample t-test whether the distribution of the
-estimated contribution of each factor different from baseline ($t=-100$ms):
+population, we tested, using one-sample $t$-test whether the distribution of
+the estimated contribution of each factor different from baseline ($t=-100$ms),
+as $\text{t-test}(\hat E_{t=-100} - \hat E_{\tau})$, with $\tau \in $ 150ms and
+400ms respectively
 
-\begin{equation} ttest(\hat E_{t=-100} - \hat E_{\tau}) \end{equation} with
-$\tau \in $ 150ms and 400ms respectively
-
-\subsection{Results}
+\subsubsection{Results}
 
@@
 \end{figure}
-
-\subsection{Discussion of the results.}

### 2019-05-22 paper/paper.tex
Merging...

--- 2019-05-22 paper/paper.tex
@@
 %
-In particular, $\hat G = (Y^\top Y + \Lambda_Y)^{-1} Y^\top X$ and $\hat H=(X^\top X + \Lambda_X)^{-1} X^\top YG$, where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices containing regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.
+In particular, $\hat G = (Y^\top Y + \Lambda_Y)^{-1} Y^\top X$ and $\hat H=(X^\top X + \Lambda_X)^{-1} X^\top Y \hat G$, where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices containing regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.
 
@@
 To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split.
+%
+After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$ as a final B2BR prediction rule.
 
@@
         $\hat{G} = \text{LinearRegression}(Y_1, X_1)$ \Comment*[r]{$\hat G = (Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1$}
-        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$\;
-        $E \leftarrow E + \text{Diag}(H)$\;
+        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$ \Comment*[r]{$\hat H=(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2 \hat G$}
+        $\hat{E} \leftarrow \hat{E} + \text{Diag}(H)$\;
     }
-    \Return{$\hat{E} / m$}
+    $\hat{E} \leftarrow \hat{E} / m$\;
+    $\hat{W} \leftarrow \text{LinearRegression}(X \hat{E}, Y)$\;
+    \Return{$\hat{E}$, $\hat{W}$}
     \caption{Back-to-back regression.}
@@
 
-\subsection{Post-processing of B2BR}
+\subsection{Binarizing $\hat{E}$}
 
@@
 %
-Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
+Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{todo} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
 %
@@
 \section{Experiments}
+
+We perform two experiments to evaluate the performance of BB2R: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography dataset.
+%
+We use scikit-learn to implement all of our simulations \citep{sklearn}.
+
 \subsection{Synthetic data}
@@
 
-\subsection{Related regression methods}
-FC: discuss similar methods of regression, that can be used for feature extraction
-And maybe have something about other causal discovery methods
-
 \section{Conclusion}

### 2019-05-22 paper/meg_exps.tex
meg_exps"

@@
 
-We aimed to estimate
-
-We show these coefficients as a function of time in the stacked plots, in Figure
-\ref{fig:megresult}.
+We show all the feature coefficients as a function of time in the stacked plots,
+Figure \ref{fig:megresult}. Our method confirms previous findings in
+neuroscience literature. In particular, \cite{kutas2011thirty} find that word
+frequency correlates with a brain wave around 400ms. Similarly,
+\cite{pegado2014timing} show that letter strings correlate with early brain
+responses. Both these phenomena can be observed in Figures
+\label{fig:meg_twocurves} and \ref{fig:megresult}, where we show that our
+method is able to separate them more clearly than the forward regression
+approach.
 
@@
   \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_result.pdf}
-  \caption{Average magnetosensor response across words for one subject.}
+  \caption{We plot the coefficients corresponding to each of the features,
+  describing their importance over time after stimuli onsets.
+We can observe that visual features start peaking around 150ms, while lexical features peak around 400ms.
+We show a similar plot for the Ridge regression baseline (forward) in Figure \ref{fig:ridgebaselineresult}.}
   \label{fig:megresult}
 \end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/ridgecv_baseline_result.pdf}
+  \caption{Ridge regression baseline. All the coefficients appear to be
+  linearly correlated, unlike our method in Figure \ref{fig:megresult}.}
+  \label{fig:ridgebaselineresult}
+\end{figure}

### 2019-05-22 paper/meg_exps.tex
meg_exps

@@
 \end{equation} where $\left(train,test\right)$ are partitions of the data.
-
-\subsubsection{Second-level statistics}
-
-The three types of analyses were implemented for each subject and each time
-sample independently.  To estimate the reliability of these estimates in the
-population, we tested, using one-sample $t$-test whether the distribution of
-the estimated contribution of each factor different from baseline ($t=-100$ms),
-as $\text{t-test}(\hat E_{t=-100} - \hat E_{\tau})$, with $\tau \in $ 150ms and
-400ms respectively
+%
+% \subsubsection{Second-level statistics}
+%
+% The three types of analyses were implemented for each subject and each time
+% sample independently.  To estimate the reliability of these estimates in the
+% population, we tested, using one-sample $t$-test whether the distribution of
+% the estimated contribution of each factor different from baseline ($t=-100$ms),
+% as $\text{t-test}(\hat E_{t=-100} - \hat E_{\tau})$, with $\tau \in $ 150ms and
+% 400ms respectively
 

### 2019-05-22 paper/meg_exps.tex
small fix meg_exps

@@
 responses. Both these phenomena can be observed in Figures
-\label{fig:meg_twocurves} and \ref{fig:megresult}, where we show that our
+\ref{fig:meg_twocurves} and \ref{fig:megresult}, where we show that our
 method is able to separate them more clearly than the forward regression

### 2019-05-22 paper/meg_exps.tex
Sync

--- 2019-05-22 paper/meg_exps.tex
@@
 recordings $Y$ at each time sample and for each subject separately. To this aim,
-we compare three methods (forward, backward and back-to-back
-regressions). Corresponding results can be found in Figure \ref{fig:meg_twocurves}.
+we compare three methods: forward regression, backward regression, and back-to-back regression.
+Corresponding results can be found in Figure \ref{fig:meg_twocurves}.
 
@@
 %
+For back-to-back regression, we observed that two modifications improved its performance on MEG data.
+%
+First, we replace the Tikhonov regularization in the first regression by truncated SVD (performed as a maximum likelihood estimation of principal components).
+%
+Second, we invert the data covariances using all the data, and use these common inverses across B2BR repetitions.
 
@@
 
-\paragraph{Back-to-back regression} We adapted back-to-back regression with two
-changes. First, to avoid getting a slightly positively-biased estimate of the
-factors contributing to the MEG signal, we swap the Tikhonov regularization of
-the first regression $G$ with a truncated regularization. Truncation was fitted
- by maximum likelihood estimate, as implemented in scikit-learn principal
- component analysis. Second, to minimize the variance increase induced by
- bagging, we computed the inverse regularized covariance matrices outside
- bagging:
-\begin{equation}
-  G = (Y^T Y+\lambda_Y I)^{-1} Y_{train}^T X_{train}
-\end{equation}
-\begin{equation}
-\hat X_{test} = Y_{test} \cdot G^T
-\end{equation} where $\left(train,test\right)$ are partitions of the data.
+% \paragraph{Back-to-back regression} We adapted back-to-back regression with two
+% changes. First, to avoid getting a slightly positively-biased estimate of the
+% factors contributing to the MEG signal, we swap the Tikhonov regularization of
+% the first regression $G$ with a truncated regularization. Truncation was fitted
+%  by maximum likelihood estimate, as implemented in scikit-learn principal
+%  component analysis. Second, to minimize the variance increase induced by
+%  bagging, we computed the inverse regularized covariance matrices outside
+%  bagging:
+% \begin{equation}
+%   G = (Y^T Y+\lambda_Y I)^{-1} Y_{train}^T X_{train}
+% \end{equation}
+% \begin{equation}
+% \hat X_{test} = Y_{test} \cdot G^T
+% \end{equation} where $\left(train,test\right)$ are partitions of the data.
 
@@
 
-We show all the feature coefficients as a function of time in the stacked plots,
-Figure \ref{fig:megresult}. Our method confirms previous findings in
-neuroscience literature. In particular, \cite{kutas2011thirty} find that word
-frequency correlates with a brain wave around 400ms. Similarly,
-\cite{pegado2014timing} show that letter strings correlate with early brain
-responses. Both these phenomena can be observed in Figures
-\label{fig:meg_twocurves} and \ref{fig:megresult}, where we show that our
-method is able to separate them more clearly than the forward regression
-approach.
+We aimed to estimate
+
+We show these coefficients as a function of time in the stacked plots, in Figure
+\ref{fig:megresult}.
 
@@
   \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_result.pdf}
-  \caption{We plot the coefficients corresponding to each of the features,
-  describing their importance over time after stimuli onsets.
-We can observe that visual features start peaking around 150ms, while lexical features peak around 400ms.
-We show a similar plot for the Ridge regression baseline (forward) in Figure \ref{fig:ridgebaselineresult}.}
+  \caption{Average magnetosensor response across words for one subject.}
   \label{fig:megresult}
 \end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/ridgecv_baseline_result.pdf}
-  \caption{Ridge regression baseline. All the coefficients appear to be
-  linearly correlated, unlike our method in Figure \ref{fig:megresult}.}
-  \label{fig:ridgebaselineresult}
-\end{figure}

### 2019-05-22 paper/paper.tex
Sync

@@
 
-\section{Related works}
-\subsection{Back-to-back regression as a special case of canonical component analysis}
+\section{Related work}
 Formula (1) shows that back-to-back regression may be related to canonical component analysis (CCA).

### 2019-05-22 paper/meg_exps.tex

@@
 
+%<<<<<<< HEAD
 % \paragraph{Back-to-back regression} We adapted back-to-back regression with two
@@
 400ms respectively
+% =======
+% \paragraph{Back-to-back regression} We adapted back-to-back regression with two
+% changes. First, to avoid getting a slightly positively-biased estimate of the
+% factors contributing to the MEG signal, we swap the Tikhonov regularization of
+% the first regression $G$ with a truncated regularization. Truncation was fitted
+%  by maximum likelihood estimate, as implemented in scikit-learn principal
+%  component analysis. Second, to minimize the variance increase induced by
+%  bagging, we computed the inverse regularized covariance matrices outside
+%  bagging:
+% \begin{equation}
+%   G = (Y^T Y+\lambda_Y I)^{-1} Y_{train}^T X_{train}
+% \end{equation}
+% \begin{equation}
+% \hat X_{test} = Y_{test} \cdot G^T
+% \end{equation} where $\left(train,test\right)$ are partitions of the data.
+% %
+% % \subsubsection{Second-level statistics}
+% %
+% % The three types of analyses were implemented for each subject and each time
+% % sample independently.  To estimate the reliability of these estimates in the
+% % population, we tested, using one-sample $t$-test whether the distribution of
+% % the estimated contribution of each factor different from baseline ($t=-100$ms),
+% % as $\text{t-test}(\hat E_{t=-100} - \hat E_{\tau})$, with $\tau \in $ 150ms and
+% % 400ms respectively
+% >>>>>>> 1775190a9737244f84b8ace529198a443029ad8d
 
@@
 
-We aimed to estimate
-
-We show these coefficients as a function of time in the stacked plots, in Figure
-\ref{fig:megresult}.
+We show all the feature coefficients as a function of time in the stacked plots,
+Figure \ref{fig:megresult}. Our method confirms previous findings in
+neuroscience literature. In particular, \cite{kutas2011thirty} find that word
+frequency correlates with a brain wave around 400ms. Similarly,
+\cite{pegado2014timing} show that letter strings correlate with early brain
+responses. Both these phenomena can be observed in Figures
+\ref{fig:meg_twocurves} and \ref{fig:megresult}, where we show that our
+method is able to separate them more clearly than the forward regression
+approach.
 

### 2019-05-22 paper/conclusion.tex
conclusion tex

In this work, we explored the back-to-back (B2B) regression, a novel method for
identifying causes from observations building on two successive
multidimensional regressions; from one domain to the other, and then back.
We analyzed the asymptotic behavior of this approach, proving its convergence,
and compared it to several baselines in controlled synthetic experiments. We
showcased our approach on a recent brain imaging dataset, analyzing the timing
of brain responses and their connection to features of words, and obtained
results consistent with prior work in neuroscience literature, confirming the
viability of our method in the case of real data.

### 2019-05-22 paper/paper.tex
Going over intro.

@@
 
-If using regularized least-squares, arguably the most commonly employed linear regression technique, the solution of B2BR has a closed form solution.
+If using regularized least-squares, arguably the most commonly employed linear regression technique, the solution of B2BR has a closed form solution:
+\begin{align}
+    \hat G &= (Y^\top Y + \Lambda_Y)^{-1} Y^\top X,\label{eq:solG}\\
+    \hat H &=(X^\top X + \Lambda_X)^{-1} X^\top Y \hat G,\label{eq:solH}
+\end{align}
 %
-In particular, $\hat G = (Y^\top Y + \Lambda_Y)^{-1} Y^\top X$ and $\hat H=(X^\top X + \Lambda_X)^{-1} X^\top Y \hat G$, where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices containing regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.
+where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices containing regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.
 
@@
 \section{Related work}
-Formula (1) shows that back-to-back regression may be related to canonical component analysis (CCA).
+
+Back-to-back regression is closely related to Canonical Correlation Analysis \citep[CCA]{todo}.
 %
-In CCA, we are given two matrices X and Y (or size N,p and N,q) that describe different features measured on the same data sample, and characterize their proximity (that is, between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.
+While CCA computes the eigenvector decomposition of \eqref{eq:solH}, while we compute its diagonal.
+%
+If \eqref{eq:solH} is diagonal, as in our model \eqref{eq:model}, the variables in $X$ are the eigenvectors.
+%
+If all the associated eigenvalues are either $0$ or $1$, then all the variables are either part of the $Y$ subspace, or orthogonal to it, respectively.
+%
+This is because these eigenvalues correspond to the squared cosine of the angle between the corresponding variable in $X$ and the $Y$ subspace.
+%
+Thus, back-to-back regression is equivalent to CCA under the constraint that the subspaces spanned by $X$ and $Y$ are perpendicular, $X$ is spanned by vectors of $Y$ and vectors orthogonal to $Y$, and that the parallel and orthogonal components of $X$ are spanned by disjoint sets of features of $X$.
 
-This is done by finding vectors $a$ and $b$ such that $Xa$ and $Yb$ display maximal correlation, or by maximizing $a'X'Yb$ over unit-normed $a$ and $b$.
+There are, however, differences between CCA and back-to-back regression.
 %
-To find a, one calculates the eigenvectors of Eq. (1).
+First, since our objective is not to characterise the proximity between $X$ and $Y$, but to measure causal influences, we can disregard the diagonalization of \eqref{eq:solH} performed by CCA. 
 %
-The largest one is the direction with maximal correlation, the second one the maximal correlation direction orthogonal to the first, and so on.
-%
-For each canonical dimension, the corresponding eigenvalue is the square of the cosine between the corresponding direction along X and the subspace spanned by Y.
-
-If the result of Eq. (1) is diagonal, as in the model presented above, the features of X are eigenvectors.
-%
-If all eigenvalues are either 0 or 1, then all the features are either part of the Y subspace (eigenvalues of one) or orthogonal to it (for zero eigenvalues) because they correspond to the squared cosine of the angle between the corresponding feature and the subspace spanned by Y, .
-%
-Thus, back-to-back regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
-%
-The closer we are to this hypothesis, the closer the results of CCA and back-to-back regression will be.
-
-There are differences between CCA and back-to-back regression.
-%
-First, since our objective is not to characterise the proximity between X and Y, but just to extract causal features, we can dispense with the last step of CCA (diagonalisation of (1)).
-%
-Second, since we operate in a noisy environment, we use regularisation, introducing bias in the correlation calculations.
-%
-Finally, whereas CCA wants to leverage all correlations between X and Y, we filter some of them, through splitting and bagging.
-%
-In other words, whereas we are working from the same formulae as CCA, we use them in different ways, for different purposes
+Second, whereas CCA leverages all correlations between $X$ and $Y$, we filter some of them, through splitting and bagging.
 
 \section{Conclusion}
+
+\input{conclusion}
 

### 2019-05-22 paper/paper.tex
Intro

@@
 manipulate individually and ii) observations are complex and multi-dimensional.
-To address this issue, we introduce ``back-to-back regression'' (BFR), a
-method designed to identify, from a set of co-varying factors, the factors
-that most plausible account for multidimensional observations. After detailing
-the proof of convergence, we show that BFR outperforms least-squares and related
+To address this issue, we introduce ``Back-to-Back Regression'' (B2BR), a
+method designed to measure, from a set of co-varying factors, the causal influences 
+that most plausible account for multidimensional observations. After proving the consistency of B2BR,
+we show that our method outperforms least-squares and related
 regression techniques, as well as cross-decomposition methods (e.g. canonical
 correlation analysis, and partial least squares) on two tasks: causal
-identification and out-of-distribution prediction. Finally, we apply BFR to
+identification and out-of-distribution prediction. Finally, we apply B2BR to
 neuroimaging recordings of 102 subjects reading word sequences. The results

### 2019-05-22 paper/paper.tex
backwards done

--- 2019-05-22 paper/paper.tex
@@
 %
-In particular, $\hat G = (Y^\top Y + \Lambda_Y)^{-1} Y^\top X$ and $\hat H=(X^\top X + \Lambda_X)^{-1} X^\top Y \hat G$, where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices containing regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.
+In particular, $\hat G = (Y^\top Y + \Lambda_Y)^{-1} Y^\top X$ and $\hat H=(X^\top X + \Lambda_X)^{-1} X^\top YG$, where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices containing regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.
 
@@
 To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split.
-%
-After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$ as a final B2BR prediction rule.
 
@@
         $\hat{G} = \text{LinearRegression}(Y_1, X_1)$ \Comment*[r]{$\hat G = (Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1$}
-        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$ \Comment*[r]{$\hat H=(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2 \hat G$}
-        $\hat{E} \leftarrow \hat{E} + \text{Diag}(H)$\;
+        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$\;
+        $E \leftarrow E + \text{Diag}(H)$\;
     }
-    $\hat{E} \leftarrow \hat{E} / m$\;
-    $\hat{W} \leftarrow \text{LinearRegression}(X \hat{E}, Y)$\;
-    \Return{$\hat{E}$, $\hat{W}$}
+    \Return{$\hat{E} / m$}
     \caption{Back-to-back regression.}
@@
 \begin{align*}
-    \arg \min_G \left \| YG - X \right \|^2 &= \arg \min_G \left \| (EX + N)FG \right\|^2 \\
+    \arg \min_G \left \| YG - X \right \|^2 &= \arg \min_G \left \| X - (XE + N)FG \right\|^2 \\
                                             &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
@@
 %
-Let $K \in \mathbb{R}^{d_x \times k}$ be the diagonal matrix obtained by deleting the right-most $d_x - k$ columns of $E$.
+Let $K \in \mathbb{R}^{k \times n_{x}}$ be the diagonal matrix obtained by deleting the bottom $dn_{x} - k$ lines of $E$.
 %
-Then, $E = KK^\top$, and $XEF = (XK)(K^\top F)$, with both $XK$ and $K^\top$ of rank $k$.
+Then, $E = K^\top K$, and $XEF = (XK^\top)(KF)$, with both $XK^\top$ and $KF$ of rank $k$.
 
-Recall the rank factorisation property of the pseudo-inverse \citep{todo}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $C$ is full rank, $C^\dagger=C' (CC')^{-1}$.
+Recall the rank factorisation property of the pseudo-inverse \citep{todo}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
 %
@@
 \begin{aligned}
-(XEF)^\dagger X&=((XK')(KF))^\dagger X\\
-&=(FK)^\dagger (XK')'((XK')(XK')')^{-1} X\\
-&=(FK)^\dagger KX' (K'XX'K)^{-1} X
+(XEF)^\dagger X&=((XK^\top)(KF))^\dagger X\\
+&=(KF)^\dagger ((XK^\top)^\top (XK^\top))^{-1} (XK^\top)^\top X\\
+&=(KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X
 \end{aligned}
@@
 
-Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, the top left block being $k\times k$.
-Then, $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{2} \Sigma_{1}^{-1}\end{array}\right)$, call it S.
+Let $X^\top X = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, the top left block being $k\times k$.
+Then, $(KX^\top X K^\top)=\Sigma_{1}$ and $(KX^\top XK^\top)^{-1} KX^\top X=\left(\begin{array}{cc}I_{k} &  \Sigma_{1}^{-1} \Sigma_{2}\end{array}\right)$, call it S.
 
-Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=X(FEX)^\dagger =S (FK)^\dagger$.
-If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $K (FK)^\dagger$.
+Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=(XEF)^\dagger X=(KF)^\dagger S$.
+If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $(KF)^\dagger K$.
 
-Under centred noise, and if the covariance of FN has low condition number (a weak assumption given that N is noise), we have $\left \| GFN\right \| ^2 \approx Var(FN) \left \| G\right \| ^2$, and minimisation of $\left \| X-GFEX\right\| ^2  + \left \| GFN\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda S (FE)^\dagger$, $\lambda$ a positive number below one.
+Under centred noise, and if the covariance of NF has low condition number (a weak assumption given that N is noise), we have $\left \| NFG\right \| ^2 \approx Var(NF) \left \| G\right \| ^2$, and minimisation of $\left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda (KF)^\dagger S$, $\lambda$ a positive number below one.
 
-Replacing in the left part of (ref), we have $X-GFEX=X- \lambda S (FK)^\dagger FKK'X= X - \lambda EX=X(I-\lambda E)$.
+Replacing in the left part of (ref), we have $X-XEFG=X- XK^\top KF \lambda (KF)^\dagger S= X - \lambda XE=X(I-\lambda E)$.
 Its square norm is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ ($Var(X_{.})$ denoting the variance of X over active and inactive features).
 
-For the right part, we have $GFN=\lambda S (FK)^\dagger FN= \lambda S (FK)^\dagger (FK+F(I-K))K'N= \lambda S EN=  \lambda EN$.
-(Since F is full rank over E, we have $(FK)^\dagger F(I-K)= 0$).
+For the right part, we have $NFG=NF \lambda (KF)^\dagger S = NK^\top (KF+(I-K)F) \lambda  (KF)^\dagger S = \lambda NES=  \lambda NE$.
+(Since F is full rank over E, we have $(KF)^\dagger (I-K)F= 0$).
 The square norm is therefore $\lambda^2 Var(N_{active})$.
@@
 
-The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} S (FK)^\dagger$.
+The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} S (KF)^\dagger$.
 
@@
 \begin{aligned}
-\arg \min_H \left \| \hat GY - HX \right \|^2 &=\arg  \min_H \left \| \hat GF(EX+N) - HX \right \|^2 \\
-&=\arg \min_H \left \| (\hat G(FE)-H)X \right \| ^2 + \left \| \hat GFN \right \| ^2\\
-&= \arg \min_H \left \| (\hat G(FE)-H)X \right \| ^2\\
-&=\hat G (FE)
+\arg \min_H \left \| \hat YG - XH \right \|^2 &=\arg  \min_H \left \| (XE + N)F \hat G - XH \right \|^2 \\
+&=\arg \min_H \left \| X(EF \hat G - H) \right \| ^2 + \left \| NF\hat G \right \| ^2\\
+&= \arg \min_H \left \| X(EF \hat G - H) \right \| ^2\\
+&= EF \hat G 
 \end{aligned}
@@
 
-Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} S (FK)^{\dagger}(FKK') = \frac{1}{1+nsr} E$.
+Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} (K^\top K F)(KF)^{\dagger} S = \frac{1}{1+nsr} E$.
 
-As we shall see, the diagonal of $\hat H$ recovered by back to back regression is a scaled approximation of the diagonal of E.
+%The diagonal of $\hat H$ recovered by back to back regression is a scaled approximation of the diagonal of E.
 Noise tends reduce the large diagonal elements, whereas the bias introduced by regularisation tends to increase the lowest ones.
@@
 
-\subsection{Binarizing $\hat{E}$}
+\subsection{Post-processing of B2BR}
 
@@
 %
-Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{todo} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
+Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
 %
@@
 \section{Experiments}
-
-We perform two experiments to evaluate the performance of BB2R: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography dataset.
-%
-We use scikit-learn to implement all of our simulations \citep{sklearn}.
-
 \subsection{Synthetic data}
@@
 
-\section{Related work}
+\section{Related works}
+\subsection{Back-to-back regression as a special case of canonical component analysis}
 Formula (1) shows that back-to-back regression may be related to canonical component analysis (CCA).
@@
 
+\subsection{Related regression methods}
+FC: discuss similar methods of regression, that can be used for feature extraction
+And maybe have something about other causal discovery methods
+
 \section{Conclusion}

### 2019-05-22 paper/conclusion.tex

@@
-In this work, we explored the back-to-back (B2B) regression, a novel method for
-identifying causes from observations building on two successive
-multidimensional regressions; from one domain to the other, and then back.
-We analyzed the asymptotic behavior of this approach, proving its convergence,
-and compared it to several baselines in controlled synthetic experiments. We
-showcased our approach on a recent brain imaging dataset, analyzing the timing
-of brain responses and their connection to features of words, and obtained
-results consistent with prior work in neuroscience literature, confirming the
-viability of our method in the case of real data.
+In this work, we proposed Back-to-Back Regression (B2BR), a method to 
+measure the causal influence of a potential set of variables generating some observations.
+%
+B2BR performs two successive multidimensional regressions: one from the output domain, and another one from the input domain.
+%
+We provided a theoretical guarantee about the consistency of B2BR, and compared it to several baselines in controlled synthetic experiments.
+%
+We also applied B2BR to a recent brain imaging dataset, analyzing the timing
+of brain responses and their connection to word features.
+%
+We obtained results consistent with prior work in neuroscience literature, confirming the efficacy of B2BR for real data analysis.

### 2019-05-22 paper/introduction.tex

@@
 This issue becomes increasingly pronounced as the number of potential factor increases.
-In neuroscience for example, it can be challenging to identify whether the frequency of a word presented on a subject' retina modulates brain activity.
+In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity.
 Indeed, word frequency strongly covaries with word length (short words are more frequent than long words) as well as with word category (determinants are more frequent than adverbs), the individual letters from which it is composed etc (e.g.
@@
 
-The second challenge to causal discovery is that observations are becoming increasingly large and complex.
+The second challenge to measuring causal influence is that observations are becoming increasingly large and complex.
 The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from large and complex observations.
@@
 It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus.
-
-As illustrated in Fig.
-1., both forward and backward modeling have competing benefits and drawbacks.
+%
+Both forward and backward modeling have competing benefits and drawbacks.
 Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations.
@@
 In this paper, we introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling.
-After detailing the method and proving its convergence, we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of causal discovery and out-of-environment predictions.
+After detailing the method and proving its convergence, we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating causal influence and out-of-environment predictions.
 Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features leads to remarkably distinguishable brain responses.

### 2019-05-22 paper/synthetic.tex

@@
 
-Table~\ref{table:synthetic} summarizes the results for all experimental conditions, metrics, and methods...
-
-\clearpage
-\newpage
+Table~\ref{table:synthetic} summarizes the results for all experimental conditions, metrics, and methods.
+%
+B2BR is the method obtaining best test error out-distribution, lowest false positive and false negative discovery about $E$, and overall the best AUC when it comes to detecting causal influences.

### 2019-05-23 paper/introduction.tex
Added citation placeholders.

@@
 This issue becomes increasingly pronounced as the number of potential factor increases.
-In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity.
+In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity [CITE].
 Indeed, word frequency strongly covaries with word length (short words are more frequent than long words) as well as with word category (determinants are more frequent than adverbs), the individual letters from which it is composed etc (e.g.
@@
 brain activity) from a minimal combination of competing factors (e.g.
-word length, word frequency, etc), and investigate the resulting parameters of the best model.
+word length, word frequency, etc), and investigate the resulting parameters of the best model [CITE].
 
@@
 For example, brain activity is generally recorded with hundreds of sensors simultaneously.
-It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus.
+It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus [CITE].
 %
@@
 Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations.
-By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors.
-\par
+By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors [CITE].
 
-In this paper, we introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling.
-After detailing the method and proving its convergence, we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating causal influence and out-of-environment predictions.
-Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features leads to remarkably distinguishable brain responses.
+We introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}) After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating causal influence and out-of-environment predictions (Section~\ref{sec:experiment_synthetic}).
+Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features leads to distinguishable brain responses (Section~\ref{sec:experiment_real}).

### 2019-05-23 paper/paper.tex
Added citation placeholders.

@@
+% TODO: Talk about cause-effect methods, trace method, half-sibling
+% TODO: Cite Roweis component analysis methods paper
+% TODO: review abstract
+% TODO: review conclusion
+
 \documentclass{article}
@@
 \tikzset{round right paren/.style={ncbar=0.5cm,out=70,in=-70}}
-
-\newcommand{\dlp}[1]{{\color{red} (DLP: #1)}}
-\newcommand{\mo}[1]{{\color{green} (MO: #1)}}
-\newcommand{\jrk}[1]{{\color{blue} (JRK: #1)}}
-\newcommand{\fc}[1]{{\color{orange} (FC: #1)}}
 
@@
 \section{Back to back regression}
+\label{sec:algorithm}
 
@@
 This process, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.
-
-Given the model in Equation~\eqref{eq:model}, the goal of Back-to-Back Regression (B2BR) is to estimate the matrix of causal influences $E$.
+%
+Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back Regression (B2BR) is to estimate the matrix of causal influences $E$.
 
@@
 
-If using regularized least-squares, arguably the most commonly employed linear regression technique, the solution of B2BR has a closed form solution:
+If using regularized least-squares, arguably the most commonly employed linear regression technique [CITE], the solution of B2BR has a closed form solution:
 \begin{align}
@@
 %
-where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices containing regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.
+where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices of regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.
 
@@
 %
-To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split.
+To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split [CITE BOOTSTRAP].
 %
@@
 \subsection{Theoretical guarantees}
+\label{sec:theorem}
 % As discussed in the introduction, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(EX+N)$, with f an unknown function.
@@
 
-Recall the rank factorisation property of the pseudo-inverse \citep{todo}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
+Recall the rank factorisation property of the pseudo-inverse [CITE]: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
 %
@@
 
-In regression analysis, the traditional approach to this problem consists in testing whether the regression coefficients differ from zero.
+In regression analysis, the traditional approach to this problem employs a $t$-test to check whether the regression coefficients differ from zero [CITE].
 %
@@
 %
-Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{todo} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
+Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan [CITE] criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
 %
@@
 \subsection{Synthetic data}
+\label{sec:experiment_synthetic}
 
@@
 \subsection{Magnetoencephalography data}
+\label{sec:experiment_real}
 
@@
 
-Back-to-back regression is closely related to Canonical Correlation Analysis \citep[CCA]{todo}.
+Back-to-back regression is closely related to Canonical Correlation Analysis [CITE].
 %

### 2019-05-23 paper/synthetic.tex
Added citation placeholders.

@@
 %
-Ridge regression \citep{}, multitask Lasso \citep{}, Partial Least Squares or PLS \citep[PLS]{}, Canonical Correlation Analysis \citep[CCA]{}, and Reduced Rank Regression \citep[RRR]{}.
+Ridge regression [CITE], multitask Lasso [CITE], Partial Least Squares or PLS [CITE], Canonical Correlation Analysis or CCA [CITE], and Reduced Rank Regression or RRR [CITE].
 %

### 2019-05-23 code/synthetic_experiment/plot.py
plot synthetic

@@ -54,7 +54,7 @@ if __name__ == "__main__":
         if eid not in results:
             results[eid] = {}
 
-        if model not in results[eid]: 
+        if model not in results[eid]:
             results[eid][model] = {}
 
         for stat in stats:
@@ -62,7 +62,6 @@ if __name__ == "__main__":
                 results[eid][model][stat] = []
             results[eid][model][stat].append(line_dict[stat])
 
-    
     plt.figure(figsize=(8, 5))
     counter = 0
     for m, model in enumerate(models):
@@ -100,7 +99,7 @@ if __name__ == "__main__":
 
             ax.set_xlim(min_, max_)
             ax.set_ylim(min_, max_)
-            
+
             if m != (len(models) - 1):
                 plt.xticks([])
                 plt.yticks([])
@@ -110,3 +109,70 @@ if __name__ == "__main__":
     plt.tight_layout(0, 0, 0)
     plt.savefig(args.file + ".pdf")
     # plt.show()
+
+    # Figure 2
+    # Gather results
+    import pandas as pd
+    summary = list()
+    for m, model in enumerate(models):
+        for s, stat in enumerate(stats):
+            data_jrr = []
+            data_other = []
+            data_nl = []
+            if model != "JRR":
+                for eid in results:
+                    if model in results[eid]:
+                        data_jrr.append(np.mean(results[eid]["JRR"][stat]))
+                        data_other.append(np.mean(results[eid][model][stat]))
+                        is_nonlinear = int(eid.split("_")[5])
+                        data_nl.append(is_nonlinear)
+            # Compare jrr metric with model metric
+            diff = np.array(data_other) - np.array(data_jrr)
+            for d, nl in zip(diff, data_nl):
+                summary.append(dict(model=model, stat=stat,
+                                    effect=d, nonlinear=nl))
+
+    summary = pd.DataFrame(summary)
+    fig, axes = plt.subplots(1, 7, figsize=(8, 2), sharex=True)
+    for stat, ax in zip(stats, axes):
+        effect = summary.query('stat==@stat')
+
+        x = 0
+        colors = dict(Ridge='C0', Lasso='C1', PLS='C2', CCA='C3', RRR='C4')
+        for (model, nonlinear), d in effect.groupby(['model', 'nonlinear']):
+            x += 1
+            # plot mearn
+            m = d.effect.mean()
+            ax.bar(x, m,
[…]

### 2019-05-23 paper/paper.tex

@@
 % TODO: Cite Roweis component analysis methods paper
+% TODO: read quantifying causal influences
 % TODO: review abstract
@@
 
-\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx}
+\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref}
 \usepackage[ruled, linesnumbered]{algorithm2e}
-
-\usepackage[hidelinks]{hyperref}
 \usepackage[]{neurips_2019}
-\DeclareMathOperator{\Tr}{Tr}
 
@@
 
-If using regularized least-squares, arguably the most commonly employed linear regression technique [CITE], the solution of B2BR has a closed form solution:
+If using regularized least-squares, arguably the most commonly employed linear regression technique \citep{hoerl1959optimum, rifkin2007notes}, the solution of B2BR has a closed form solution:
 \begin{align}
@@
 %
-To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split [CITE BOOTSTRAP].
+To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split \citep{efron1992bootstrap}.
 %
@@
 
-Recall the rank factorisation property of the pseudo-inverse [CITE]: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
+Recall the rank factorisation property of the pseudo-inverse \citep{ben2003generalize}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
 %
@@
 
-In regression analysis, the traditional approach to this problem employs a $t$-test to check whether the regression coefficients differ from zero [CITE].
+In regression analysis, the traditional approach to this problem employs a $t$-test to check whether the regression coefficients differ from zero \citep{student1908probable}.
 %
@@
 %
-Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan [CITE] criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
+Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{sonquist_morgan} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
 %
@@
 
-Back-to-back regression is closely related to Canonical Correlation Analysis [CITE].
+Back-to-back regression is closely related to Canonical Correlation Analysis \citep{cca_hotelling}. 
 %
@@
 
-\bibliographystyle{abbrvnat}
+\bibliographystyle{plain}
 \bibliography{paper}

### 2019-05-23 paper/synthetic.tex

@@
 %
-Ridge regression [CITE], multitask Lasso [CITE], Partial Least Squares or PLS [CITE], Canonical Correlation Analysis or CCA [CITE], and Reduced Rank Regression or RRR [CITE].
+Ridge regression \citep{hoerl1959optimum}, multitask Lasso \citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls, tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and Reduced Rank Regression or RRR \citep{Izenman_rrr}.
 %

### 2019-05-23 paper/meg_exps.tex
meg results

@@
 %
-For back-to-back regression, we observed that two modifications improved its performance on MEG data.
-%
-First, we replace the Tikhonov regularization in the first regression by truncated SVD (performed as a maximum likelihood estimation of principal components).
-%
-Second, we invert the data covariances using all the data, and use these common inverses across B2BR repetitions.
+Finally, we used the back-to-back regression as described earlier, using a bagging of 10 alphas logarithmically distributed between -5 and 5.
 
@@
 population, we tested, using one-sample $t$-test whether the distribution of
-the estimated contribution of each factor different from baseline ($t=-100$ms),
-as $\text{t-test}(\hat E_{t=-100} - \hat E_{\tau})$, with $\tau \in $ 150ms and
-400ms respectively
+the estimated contribution of each factor different from baseline ($t=0$ms),
+as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau \in $ 150ms and
+400ms respectively. We use the resulting p-values as proxy to quantify the
+performance of each model.
 % =======
@@
 % % population, we tested, using one-sample $t$-test whether the distribution of
-% % the estimated contribution of each factor different from baseline ($t=-100$ms),
-% % as $\text{t-test}(\hat E_{t=-100} - \hat E_{\tau})$, with $\tau \in $ 150ms and
+% % the estimated contribution of each factor different from baseline ($t=0$ms),
+% % as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau \in $ 150ms and
 % % 400ms respectively
@@
 \subsubsection{Results}
+It is a well established fact that the number of letters in a string flashed
+onto the retina modulates brain activity from 150ms after stimulus onset
+\cite{pegado2014timing}. Similarly, word frequency is known to modulate brain
+activity around 400ms \cite{kutas2011thirty}. Consequently, we aimed to compare
+how sensitive each modeling approach in detecting these two types of brain
+responses in a large neuroimaging dataset.
 
-We show all the feature coefficients as a function of time in the stacked plots,
-Figure \ref{fig:megresult}. Our method confirms previous findings in
-neuroscience literature. In particular, \cite{kutas2011thirty} find that word
-frequency correlates with a brain wave around 400ms. Similarly,
-\cite{pegado2014timing} show that letter strings correlate with early brain
-responses. Both these phenomena can be observed in Figures
-\ref{fig:meg_twocurves} and \ref{fig:megresult}, where we show that our
-method is able to separate them more clearly than the forward regression
-approach.
+In Figure \ref{fig:meg_twocurves} we show for each model the effects obtained in a
+representative subject, as well as the average effects obtained across subjects.
+The results show that all methods significantly detect that both word length and
+word frequency modulate brain responses after the stimulus onset. However
+forward regression detected word length effects (XXX) and word frequency (XXX)
+effects less reliably than B2BR (length=XXX, frequency=XXX). B2BR appeared to
+be less sensitive than the backward regression (mean XXX). However, it is
+important to highlight that backward models may mistakenly confound the two
+factors and thus double the signal-to-noise ratio. This possibility is
+corroborated by the remarkably similar dynamics between the two curves.
+
+Finally we show in Figure \ref{fig:megresult} that B2BR is sufficiently sensitive
+to reliably detect a multiplicity of subtle features, such as individual letters
+and part-of-speech. This methods thus promises to improve the detectability of
+subtle effects.
 

### 2019-05-23 paper/meg_exps.tex
meg results

@@
 the estimated contribution of each factor different from baseline ($t=0$ms),
-as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau \in $ 150ms and
-400ms respectively. We use the resulting p-values as proxy to quantify the
-performance of each model.
+as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau \in $ ranging from
+150ms to 1s after word onset. We use the resulting p-values as proxy to quantify
+the performance of each model. These results are depicted in Figure
+\ref{fig:meg_pval}. For concision purposes, we only report this metric averaged
+between 150ms and 1s after word onset.
 % =======
@@
 \subsubsection{Results}
-It is a well established fact that the number of letters in a string flashed
-onto the retina modulates brain activity from 150ms after stimulus onset
-\cite{pegado2014timing}. Similarly, word frequency is known to modulate brain
-activity around 400ms \cite{kutas2011thirty}. Consequently, we aimed to compare
-how sensitive each modeling approach in detecting these two types of brain
-responses in a large neuroimaging dataset.
+The number of letters in a string flashed onto the retina is known to modulate
+brain activity from 150ms after stimulus onset \cite{pegado2014timing}.
+Similarly, word frequency is known to modulate brain activity around 400ms \cite{kutas2011thirty}. Consequently, we aimed to compare how sensitive each
+model is in detecting these two types of brain responses in a large
+neuroimaging dataset, where word length and word frequency were collinear.
 
-In Figure \ref{fig:meg_twocurves} we show for each model the effects obtained in a
-representative subject, as well as the average effects obtained across subjects.
-The results show that all methods significantly detect that both word length and
+In Figure \ref{fig:meg_twocurves} we show, for each model, the effects obtained
+in a representative subject, as well as the average effects across subjects.
+The results show that all models significantly detect that both word length and
 word frequency modulate brain responses after the stimulus onset. However
-forward regression detected word length effects (XXX) and word frequency (XXX)
-effects less reliably than B2BR (length=XXX, frequency=XXX). B2BR appeared to
-be less sensitive than the backward regression (mean XXX). However, it is
-important to highlight that backward models may mistakenly confound the two
-factors and thus double the signal-to-noise ratio. This possibility is
-corroborated by the remarkably similar dynamics between the two curves.
+forward regression detected word length ($t-value=10.6$, $p=10^{-16}$) and
+word frequency ($t-value=11.1$, $p=10^-17$) less reliably than B2B (length:
+$t-value=18.7$, $p=10^{-31}, frequency:$t-value=15.9$, $p=10^{-26}$). B2B
+was less sensitive than the backward regression (length: $t-value=33.6$,
+$p=10^{-50}$, $t-value=31.7$, frequency:$p=10^{-48}$). However, as detailed in
+the introduction, backward models confound collinear factors. Consequently,
+the high sensitivity of our backward regression is likely to come from a loss of specificity. This possibility is corroborated by the fact that the dynamics
+associated with word length and word frequency are remarkably similar in the
+backward model.
 
-Finally we show in Figure \ref{fig:megresult} that B2BR is sufficiently sensitive
-to reliably detect a multiplicity of subtle features, such as individual letters
-and part-of-speech. This methods thus promises to improve the detectability of
-subtle effects.
+Finally we investigate in Figure \ref{fig:megresult} whether B2B was sufficiently sensitive to reliably detect a multiplicity of subtle features, such as individual letters and part-of-speech.
+All of these features appeared to be reliably detected \cite{fig:megbar}.
+B2B thus promises to improve the detectability of subtle effects.
 

### 2019-05-23 paper/paper.tex

@@
 manipulate individually and ii) observations are complex and multi-dimensional.
-To address this issue, we introduce ``Back-to-Back Regression'' (B2BR), a
+To address this issue, we introduce ``Back-to-Back'' regression (B2B), a
 method designed to measure, from a set of co-varying factors, the causal influences
-that most plausible account for multidimensional observations. After proving the consistency of B2BR,
-we show that our method outperforms least-squares and related
-regression techniques, as well as cross-decomposition methods (e.g. canonical
-correlation analysis, and partial least squares) on two tasks: causal
-identification and out-of-distribution prediction. Finally, we apply B2BR to
+that most plausibly account for multidimensional observations. After proving the consistency of B2B, we show that our method outperforms least-squares and related
+regression and cross-decomposition  techniques (e.g. canonical
+correlation analysis and partial least squares) on two tasks: causal
+identification and out-of-distribution prediction. Finally, we apply B2B to
 neuroimaging recordings of 102 subjects reading word sequences. The results
-show, as expected, that the early brain responses to words appear to be
-specifically caused by low-level visual factors whereas late brain responses
-appear to be caused by lexical factors, despite the fact that these factors
-covary in the study.
+show that the early and late brain responses caused by low- and high-level
+word features are more easily detected with B2B than with standard forward
+regression whereas techniques.
 % conclude  open
@@
 %
-Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back Regression (B2BR) is to estimate the matrix of causal influences $E$.
+Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back Regression (B2B) is to estimate the matrix of causal influences $E$.
 
@@
 
-Back-to-Back Regression (B2BR) consists in two steps.
+Back-to-Back Regression (B2B) consists in two steps.
 %
@@
 
-If using regularized least-squares, arguably the most commonly employed linear regression technique \citep{hoerl1959optimum, rifkin2007notes}, the solution of B2BR has a closed form solution:
+If using regularized least-squares, arguably the most commonly employed linear regression technique \citep{hoerl1959optimum, rifkin2007notes}, the solution of B2B has a closed form solution:
 \begin{align}
@@
 %
-To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split \citep{efron1992bootstrap}.
+To compensate for the reduction in sample size caused by the split, B2B is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split \citep{efron1992bootstrap}.
 %
-After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$ as a final B2BR prediction rule.
+After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$ as a final B2B prediction rule.
 
-We summarize the B2BR procedure in Algorithm~\ref{algorithm:b2br}.
+We summarize the B2B procedure in Algorithm~\ref{algorithm:b2br}.
 %
-The rest of this section provides a theoretical guarantee on the correctness of B2BR, and an optional post-processing step to binarize the estimated causal influence matrix.
+The rest of this section provides a theoretical guarantee on the correctness of B2B, and an optional post-processing step to binarize the estimated causal influence matrix.
 
@@
 
-On the one hand, we have leveraged B2BR to obtain the estimate $\hat{E}$, which is a diagonal matrix with real entries.
+On the one hand, we have leveraged B2B to obtain the estimate $\hat{E}$, which is a diagonal matrix with real entries.
 %

### 2019-05-23 paper/paper.tex
Changes

--- 2019-05-23 paper/paper.tex
@@
 
-\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref}
+\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref, amsthm}
 \usepackage[ruled, linesnumbered]{algorithm2e}
 \usepackage[]{neurips_2019}
+
+\newtheorem{theorem}{Theorem}
 
@@
 \begin{equation}
-    Y = (XE + N)F + N',\label{eq:model}
+    Y = (XE + N)F,\label{eq:model}
 \end{equation}
 %
-where $N$ and $N'$ are some unobserved noise variables.
+where $N$ is a centered, unobserved noise variable.
 %
-While the triplet of variables $X$, $N$, and $N'$ are independent, we allow each of them to have a general covariance matrix.
+While the triplet of variables $X$ and $N$ are independent, we allow each of them to have a general covariance matrix.
 %
@@
 
-Assuming centered noise $N$,
+This section provides a consistency proof for B2BR in the case of low noise.
+
+\begin{theorem}[Low-noise consistency]
+    Consider the B2BR model from Equation~\ref{eq:model}.
+    %
+    Assume that $F$ and $X$ are full-rank on the subspace spanned by active entries in $E$. 
+    %
+    Then, as the variance of $N$ tends to zero and the sample size tends to infinity, B2BR recovers $E$.
+\end{theorem}
+\begin{proof}
+For centered noise $N$, the first regression satisfies:
 \begin{align*}
-    \arg \min_G \left \| YG - X \right \|^2 &= \arg \min_G \left \| X - (XE + N)FG \right\|^2 \\
-                                            &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
+    \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
+                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2].
 \end{align*}
 %
-As the noise $N$ tends to zero, the solution to the previous problem is $\hat G = (X E F)^\dagger X$, where ``$\dagger$'' denotes the Moore-Penrose pseudo-inverse.
+The solution to this first regression is $\hat{G} = [(XEF)^\top (XEF) + (NF)^\top (NF)]^{-1} (XEF)^\top X$.
 
-Next, let $k$ be the number of ones in the causal influence matrix $E$.
+The second regression is: 
 %
-Without loss of generality, assume that these are the first $k$ variables in $X$.
+\begin{align*}
+    \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
+    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
+    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]
+\end{align*}
 %
-Assume that $XEF$ is of rank $k$, which means that $F$ and $X$ have full rank on the space spanned by the active elements in $E$.
+The solution to this second regression is $H = EF\hat{G}$.
 %
-Let $K \in \mathbb{R}^{k \times n_{x}}$ be the diagonal matrix obtained by deleting the bottom $dn_{x} - k$ lines of $E$.
+Therefore, our estimate of $E$ is:
+\begin{equation}
+    \hat{E} = EF[(XEF)^\top (XEF) + (NF)^\top (NF)]^{-1} (XEF)^\top X.
+    \label{eq:estimate}
+\end{equation}
 %
-Then, $E = K^\top K$, and $XEF = (XK^\top)(KF)$, with both $XK^\top$ and $KF$ of rank $k$.
+By the definition of the Moore-Penrose, if $(NF)^\top (NF)$ is full-rank with $\| (NF)^\top (NF) \| \to 0$, it follows that $\hat{E} = EF(XEF)^\dagger X$.
+%
+Next, assume that the true causal influence matrix contains $k$ ones arranged as $E = \left(\begin{array}{cc} I_k & 0 \\ 0 & 0 \end{array}\right)$, and let $E = K^\top K$, where $K\in \mathbb{R}^{k \times d_x}$.
+%
+    Also, let the covariance matrix of $X$ be $X^\top X = \left(\begin{array}{cc}\Sigma_{1} & \Sigma_{2} \\ \Sigma_{3} & \Sigma_{4}\end{array}\right)$, where $\Sigma_1 \in \mathbb{R}^{k\times k}$.
+%
+Then,
+\begin{align*}
+    \hat{E} &= EF(XEF)^\dagger X = K^\top K F(XK^\top K F)^\dagger X \stackrel{i)}{=} K^\top (K F) (KF)^\dagger (XK^\top)^\dagger X\\
+            &\stackrel{ii)}{=} K^\top K(XK^\top)^\dagger X = E (XK^\top)^\dagger X \stackrel{iii)}{=} E (KX^\top X K^\top)^{-1} KX^\top X\\
+            &= E (KX^\top X K^\top)^{-1} (\Sigma_1 \,|\, \Sigma_2)  = E \Sigma_1^{-1} (\Sigma_1 \,|\, \Sigma_2)  
+            = \left(\begin{array}{cc} I_k & \Sigma_1^{-1} \Sigma_2 \\ 0 & 0 \end{array}\right),
+\end{align*}
+where the equalities follow due to i) a full-rank factorization of $XEF$, ii) $KF$ being full-rank yields $(KF)(KF)^\dagger = K$, and iii) $X K^\top$ being full-rank yields $(XK^\top)^\dagger = (KX^\top X K^\top)^{-1} KX^\top$.
+%
+Therefore, $\text{Diag}(\hat{E})$ is the desired influence causal matrix $E$. 
+\end{proof}
 
-Recall the rank factorisation property of the pseudo-inverse \citep{ben2003generalize}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
+We observe that, even in the presence of strong noise, the expression of our estimate \eqref{eq:estimate} involves the left-most multiplication with the true causal influence matrix $E$.
 %
-Using this property on the minimal norm solution $\hat G$ yields
-\begin{equation}
-\begin{aligned}
-(XEF)^\dagger X&=((XK^\top)(KF))^\dagger X\\
-&=(KF)^\dagger ((XK^\top)^\top (XK^\top))^{-1} (XK^\top)^\top X\\
-&=(KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X
-\end{aligned}
-\end{equation}
+This means that our estimate will, in expectation, contain zeros for all features that do not causally influence $Y$.
+%
+When the variance of $N$ increases, we of course pay a price: since the norm $\| X - XEF\hat{G} \| = \| X - X\hat{H} \|$ increases, this means that the values of $\hat{E}$ associated to the causal factors will be dampened.
 
-Let $X^\top X = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, the top left block being $k\times k$.
-Then, $(KX^\top X K^\top)=\Sigma_{1}$ and $(KX^\top XK^\top)^{-1} KX^\top X=\left(\begin{array}{cc}I_{k} &  \Sigma_{1}^{-1} \Sigma_{2}\end{array}\right)$, call it S.
-
-Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=(XEF)^\dagger X=(KF)^\dagger S$.
-If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $(KF)^\dagger K$.
-
-Under centred noise, and if the covariance of NF has low condition number (a weak assumption given that N is noise), we have $\left \| NFG\right \| ^2 \approx Var(NF) \left \| G\right \| ^2$, and minimisation of $\left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda (KF)^\dagger S$, $\lambda$ a positive number below one.
-
-Replacing in the left part of (ref), we have $X-XEFG=X- XK^\top KF \lambda (KF)^\dagger S= X - \lambda XE=X(I-\lambda E)$.
+%%% Next, let $k$ be the number of ones in the binary diagonal causal influence matrix $E$.
+%%% %
+%%% That is, $\text{tr}(E) = k$.
+%%% %
+%%% Without loss of generality, assume that these are the first $k$ variables in $X$.
+%%% %
+%%% Assume that $XEF$ is of rank $k$, which means that $F$ and $X$ have full rank on the space spanned by the active elements in $E$.
+%%% %
+%%% Let $K \in \mathbb{R}^{k \times d_{x}}$ be the diagonal matrix obtained by deleting the bottom $d_{x} - k$ rows of $E$.
+%%% %
+%%% %
+%%% First, we factorize $(XEF)^\dagger = ((XK^\top)(KF))^\dagger = (KF)^\dagger (XK^\top)^\dagger$.
+%%% %
+%%% Second, since $XK^\top$ is full rank, we write $(XK^\top)^\dagger = ((XK^\top)^\top (XK^\top))^{-1}(XK^\top)^\top$.
+%%% %
+%%% Then, $(XEF)^\dagger X = (KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X$.
+%%% 
+%%% %Recall the rank factorisation property of the pseudo-inverse \citep{ben2003generalized}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
+%%% %
+%%% % Using this property on the minimal norm solution $\hat G$ yields
+%%% % \begin{equation}
+%%% % \begin{aligned}
+%%% % (XEF)^\dagger X&=((XK^\top)(KF))^\dagger X\\
+%%% % &=(KF)^\dagger ((XK^\top)^\top (XK^\top))^{-1} (XK^\top)^\top X\\
+%%% % &=(KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X
+%%% % \end{aligned}
+%%% % \end{equation}
+%%% 
+%%% Write the covariance of $X$ as $X^\top X = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, with the block $\Sigma_1$ being $k\times k$.
+Then, $(KX^\top X K^\top)=\Sigma_{1}$ and $(KX^\top XK^\top)^{-1} KX^\top X=\left(\begin{array}{c|c}I_{k} &  \Sigma_{1}^{-1} \Sigma_{2}\end{array}\right) =: S$.
+%%% 
+%%% 
+%%% Under centred noise, and if the covariance of NF has low condition number (a weak assumption given that N is noise), we have $\left \| NFG\right \| ^2 \approx Var(NF) \left \| G\right \| ^2$, and minimisation of $\left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda (KF)^\dagger S$, $\lambda$ a positive number below one.
+%%% 
+%%% Replacing in the left part of (ref), we have $X-XEFG=X- XK^\top KF \lambda (KF)^\dagger S= X - \lambda XE=X(I-\lambda E)$.
 Its square norm is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ ($Var(X_{.})$ denoting the variance of X over active and inactive features).
-
-For the right part, we have $NFG=NF \lambda (KF)^\dagger S = NK^\top (KF+(I-K)F) \lambda  (KF)^\dagger S = \lambda NES=  \lambda NE$.
+%%% 
+%%% For the right part, we have $NFG=NF \lambda (KF)^\dagger S = NK^\top (KF+(I-K)F) \lambda  (KF)^\dagger S = \lambda NES=  \lambda NE$.
 (Since F is full rank over E, we have $(KF)^\dagger (I-K)F= 0$).
 The square norm is therefore $\lambda^2 Var(N_{active})$.
-
-Thus, with centred and random noise, and with F full rank over E, we are minimising $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N_{active})$.
+%%% 
+%%% Thus, with centred and random noise, and with F full rank over E, we are minimising $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N_{active})$.
 Its least value is attained for $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N_{active})}$.
 Assuming X and N have the same variance over all features (this can always be guaranteed), active subscript can be dropped, and this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).
-
-The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} S (KF)^\dagger$.
-
-The second step is straightforward, replacing Y, we have
-\begin{equation}
-\begin{aligned}
-\arg \min_H \left \| \hat YG - XH \right \|^2 &=\arg  \min_H \left \| (XE + N)F \hat G - XH \right \|^2 \\
-&=\arg \min_H \left \| X(EF \hat G - H) \right \| ^2 + \left \| NF\hat G \right \| ^2\\
-&= \arg \min_H \left \| X(EF \hat G - H) \right \| ^2\\
-&= EF \hat G 
-\end{aligned}
-\end{equation}
-
-Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} (K^\top K F)(KF)^{\dagger} S = \frac{1}{1+nsr} E$.
+%%% 
+%%% The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} (KF)^\dagger S$.
+%%% 
+%%% 
+%%% Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} (K^\top K F)(KF)^{\dagger} S = \frac{1}{1+nsr} E$.
+%%% 
+%%% ADD AGAIN:
+%%% 
+%%% Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=(XEF)^\dagger X=(KF)^\dagger S$.
+If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $(KF)^\dagger K$.
 

### 2019-05-23 paper/introduction.tex

@@
 First, causal and non-causal factors may be numerous and collinear.
-In physics, for example, a student may have difficulty identifying whether fusion is caused by a change in temperature or in pressure, as these two factors can be difficult to manipulate independently.
+In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or in pressure, as these two factors can be difficult to manipulate independently.
 This issue becomes increasingly pronounced as the number of potential factor increases.
-In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity [CITE].
-Indeed, word frequency strongly covaries with word length (short words are more frequent than long words) as well as with word category (determinants are more frequent than adverbs), the individual letters from which it is composed etc (e.g.
-words that contain "z" are rare) and so forth.
-As it can be particularly difficult to generate a set of words that control for all of these factors simultaneously, it is common to use forward modeling, i.e.
+In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity as this factor covaries with numerous other factors such as word length (short words are more frequent than long words) word category (determinants are more frequent than adverbs), individual letters etc (e.g.
+words that contain "z" are rare) \cite{kutas2011thirty,pegado2014timing}.
+Generating a set of words that control for all of these factors simultaneously can be challenging; it is thus common to use forward modeling, i.e.
 to train a model to predict observations (e.g.
 brain activity) from a minimal combination of competing factors (e.g.
-word length, word frequency, etc), and investigate the resulting parameters of the best model [CITE].
+word length, word frequency, etc, e.g.
+\cite{huth2016natural}), and investigate the resulting parameters of the best model.
 
-The second challenge to measuring causal influence is that observations are becoming increasingly large and complex.
+The second challenge to measuring causal influence is that observations can be large and complex.
 The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from large and complex observations.
 For example, brain activity is generally recorded with hundreds of sensors simultaneously.
-It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus [CITE].
-%
+It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \cite{cichy2014resolving, king2016brain, kriegeskorte2008representational}.
+
 Both forward and backward modeling have competing benefits and drawbacks.
 Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations.
-By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors [CITE].
+By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart_neuroimage_2015, king2018encoding}.
 

### 2019-05-23 paper/meg_exps.tex

@@
 
-% \paragraph{Forward regression} also known as 'encoding' in neuroimaging, here consists of
-% fitting a ridge-regularized ordinary least squares (OLS) regression \footnote{We use scikit-learn \citep{sklearn}.} from $X$ to $Y$:
-% \begin{equation}
-%   H = (X^{T}X+\lambda I)^{-1} X^{T}Y
-% \end{equation}
-% where $\lambda$ is the
-% regularization parameter, cross-validated between $10^{-5}$ and $10^5$.
-%
-% We then estimate the contribution
-% $diag(\hat E) \in\mathbb{D}^{d} $ of each factor. However, $H\in\mathbb{R}^{d,
-% q}$ is a matrix of $d$ word features by $q$ MEG channels. We approximate
-% the relative contribution of each feature with the sum-of-squares of the corresponding coefficients in $H$; for the $j$-th feature:
-% \begin{equation}
-%   \hat e_j = \sum_{i=0}^{q} {h_{i,j}}^2
-% \end{equation}
-% \paragraph{Backward regression} or 'decoding', is similar but predicts $X$ from $Y$:
-% \begin{equation} G = (Y^{T}Y+\lambda I)^{-1} Y^{T}X \end{equation}
-% where $\lambda$ is cross-validated as for encoding. We estimate the
-% contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor in the following
-% way. We predict $X$ on a held-out split \textit{test}:
-% % Using 10-fold cross-validation, we thus assessed the ability of this ridge
-% % regression to predict independent and identically distributed data:
-% \begin{equation} \hat X_{test} = Y_{test}\cdot G_{train} \end{equation}
-% and assess, for each factor of X independently, the Pearson correlation
-% coefficient $r$ between $X_{test}$ and $\hat X_{test}$. We repeat the procedure
-% for 20 random splits and average the result.
-
-%<<<<<<< HEAD
-% \paragraph{Back-to-back regression} We adapted back-to-back regression with two
-% changes. First, to avoid getting a slightly positively-biased estimate of the
-% factors contributing to the MEG signal, we swap the Tikhonov regularization of
-% the first regression $G$ with a truncated regularization. Truncation was fitted
-%  by maximum likelihood estimate, as implemented in scikit-learn principal
-%  component analysis. Second, to minimize the variance increase induced by
-%  bagging, we computed the inverse regularized covariance matrices outside
-%  bagging:
-% \begin{equation}
-%   G = (Y^T Y+\lambda_Y I)^{-1} Y_{train}^T X_{train}
-% \end{equation}
-% \begin{equation}
-% \hat X_{test} = Y_{test} \cdot G^T
-% \end{equation} where $\left(train,test\right)$ are partitions of the data.
-
 \subsubsection{Second-level statistics}
@@
 between 150ms and 1s after word onset.
-% =======
-% \paragraph{Back-to-back regression} We adapted back-to-back regression with two
-% changes. First, to avoid getting a slightly positively-biased estimate of the
-% factors contributing to the MEG signal, we swap the Tikhonov regularization of
-% the first regression $G$ with a truncated regularization. Truncation was fitted
-%  by maximum likelihood estimate, as implemented in scikit-learn principal
-%  component analysis. Second, to minimize the variance increase induced by
-%  bagging, we computed the inverse regularized covariance matrices outside
-%  bagging:
-% \begin{equation}
-%   G = (Y^T Y+\lambda_Y I)^{-1} Y_{train}^T X_{train}
-% \end{equation}
-% \begin{equation}
-% \hat X_{test} = Y_{test} \cdot G^T
-% \end{equation} where $\left(train,test\right)$ are partitions of the data.
-% %
-% % \subsubsection{Second-level statistics}
-% %
-% % The three types of analyses were implemented for each subject and each time
-% % sample independently.  To estimate the reliability of these estimates in the
-% % population, we tested, using one-sample $t$-test whether the distribution of
-% % the estimated contribution of each factor different from baseline ($t=0$ms),
-% % as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau \in $ 150ms and
-% % 400ms respectively
-% >>>>>>> 1775190a9737244f84b8ace529198a443029ad8d
 
@@
 brain activity from 150ms after stimulus onset \cite{pegado2014timing}.
-Similarly, word frequency is known to modulate brain activity around 400ms \cite{kutas2011thirty}. Consequently, we aimed to compare how sensitive each
+Similarly, word frequency is known to modulate brain activity around 400ms
+\cite{kutas2011thirty}. Consequently, we aimed to compare how sensitive each
 model is in detecting these two types of brain responses in a large
@@
 word frequency ($t-value=11.1$, $p=10^-17$) less reliably than B2B (length:
-$t-value=18.7$, $p=10^{-31}, frequency:$t-value=15.9$, $p=10^{-26}$). B2B
+$t-value=18.7$, $p=10^{-31}$, frequency:$t-value=15.9$, $p=10^{-26}$). B2B
 was less sensitive than the backward regression (length: $t-value=33.6$,

### 2019-05-23 paper/synthetic.tex

@@
 
-The data generating process for each experiment constructs $n$ training examples according to the model $Y = \sigma((\text{snr} \cdot XE + N)F)$.
+The data generating process for each experiment constructs $n$ training examples according to the model $Y = (\text{snr} \cdot XE + N)F$.
 %
@@
     $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
-the factor $\text{snr} \in (0, \infty)$, and $\sigma$ is an elementwise function.
+and the factor $\text{snr} \in (0, \infty)$.
 
-To simulate a wide range of experimental conditions, we vary $n \in \{500, 1000, 2000\}$, $d_x, d_y \in \{ 25, 50, 100, 150 \}$, $n_c \in \{ 5, 10 \}$, $\text{snr} \in \{ 0.2, 0.5, 0.75, 1 \}$, and $\sigma \in \{ \text{identity}, \text{sigmoid} \}$.
+To simulate a wide range of experimental conditions, we vary $n \in \{500, 1000, 2000\}$, $d_x, d_y \in \{ 25, 50, 100, 150 \}$, $n_c \in \{ 5, 10 \}$, $\text{snr} \in \{ 0.2, 0.5, 0.75, 1 \}$. 
 %
@@
     \caption{Results of synthetic experiments.
-Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration, averaged over $20$ random seeds.
-    %
-    Blue points indicate linear experimental conditions, while orange points indicate nonlinear conditions.}
+Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration, averaged over $20$ random seeds.}
     \label{table:synthetic}

### 2019-05-23 paper/introduction.tex
meg

@@
 In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity as this factor covaries with numerous other factors such as word length (short words are more frequent than long words) word category (determinants are more frequent than adverbs), individual letters etc (e.g.
-words that contain "z" are rare) \cite{kutas2011thirty,pegado2014timing}.
+words that contain "z" are rare) \citep{kutas2011thirty,pegado2014timing}.
 Generating a set of words that control for all of these factors simultaneously can be challenging; it is thus common to use forward modeling, i.e.
@@
 word length, word frequency, etc, e.g.
-\cite{huth2016natural}), and investigate the resulting parameters of the best model.
+\citep{huth2016natural}), and investigate the resulting parameters of the best model.
 
@@
 For example, brain activity is generally recorded with hundreds of sensors simultaneously.
-It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \cite{cichy2014resolving, king2016brain, kriegeskorte2008representational}.
+It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving, king2016brain, kriegeskorte2008representational}.
 
@@
 Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations.
-By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart_neuroimage_2015, king2018encoding}.
+By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \citep{weichwald2015causal, hebart_neuroimage_2015, king2018encoding}.
 

### 2019-05-23 paper/meg_exps.tex
meg

@@
 The color coding corresponds to the positions of the sensors on the head as shown on the top-left diagram.
-The diverging curves correspond to artifacts in data collection.}
+The diverging curves correspond to artifacts in the MEG data.}
   \label{fig:megavg}
@@
 noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ
-(adjective), ADV (adverb), and NUM (numerical)) using the packages Spacy
-\citep{spacy2} and WordFreq \citep{speerwordfreq}. For clarity purposes, we refer to these
+(adjective), ADV (adverb), and NUM (numerical)) using Spacy \citep{spacy2} and
+WordFreq \citep{speerwordfreq}. For clarity purposes, we refer to these
 features using two broad categories: visual (i.e. each letter and the total
@@
   \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_twocurves.pdf}
-  \caption{We compare the encoding approach (columns 1 and 2, ridge regression $X \rightarrow Y$) and the decoding approach (column 3, ridge regression $Y \rightarrow X$) to our method (column 4, $X \rightarrow Y \rightarrow X$) on real MEG data from the MOUS dataset.
+  \caption{We applied the forward regression (a.k.a encoding coefficients), the backward regression (a.k.a decoding scores) and B2B on real MEG data.
 Top row: the first subject.

### 2019-05-23 paper/introduction.tex
refs

@@
 brain activity) from a minimal combination of competing factors (e.g.
-word length, word frequency, etc, e.g.
+word length, word frequency, e.g.
 \citep{huth2016natural}), and investigate the resulting parameters of the best model.
@@
 Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations.
-By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \citep{weichwald2015causal, hebart_neuroimage_2015, king2018encoding}.
+By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \citep{weichwald2015causal, hebart2018deconstructing, king2018encoding}.
 

### 2019-05-23 paper/paper.tex
meg
refs

--- 2019-05-23 paper/paper.tex
@@
 
-\title{Measuring causal influence with\\ back-to-back regression}
+\title{Measuring causal influence with\\ back-to-back regression: the linear case}
 
@@
 To address this issue, we introduce ``Back-to-Back'' regression (B2B), a
-method designed to measure, from a set of co-varying factors, the causal influences
-that most plausibly account for multidimensional observations. After proving the consistency of B2B, we show that our method outperforms least-squares and related
-regression and cross-decomposition  techniques (e.g. canonical
-correlation analysis and partial least squares) on two tasks: causal
+method designed to measure, from a set of co-varying factors, the causal
+influences that most plausibly account for multidimensional observations. After
+proving the consistency of B2B, we show that our method outperforms
+least-squares and related regression and cross-decomposition techniques (e.g.
+canonical correlation analysis and partial least squares) on two tasks: causal
 identification and out-of-distribution prediction. Finally, we apply B2B to
@@
 show that the early and late brain responses caused by low- and high-level
-word features are more easily detected with B2B than with standard forward
-regression whereas techniques.
+word features respectively are more easily detected with B2B than with standard forward regression.
 % conclude  open
@@
 
-    \node[] (labY) at (\posY, 3){Observations};
-    \node[] (labX) at (\posX, 3){Factors};
-    \node[] (labE) at (\posE, 3.3){Cause};
-    \node[] (labE) at (\posE, 3){selection};
-    \node[] (labN) at (\posN, 3){Noise};
-    \node[] (labF) at (\posF, 3.3){Cause-effect};
-    \node[] (labF) at (\posF, 3){mapping};
+    \node[] (labY) at (\posY, 2){Observations};
+    \node[] (labX) at (\posX, 2){Factors};
+    \node[] (labE) at (\posE, 2){Cause};
+    \node[] (labE) at (\posE, 1.7){selection};
+    \node[] (labN) at (\posN, 2){Noise};
+    \node[] (labF) at (\posF, 2){Cause-effect};
+    \node[] (labF) at (\posF, 1.7){mapping};
 
-    \node[] (labY) at (\posY, 2)   {\color{gray} MEG recordings};
-    \node[] (labX) at (\posX, 2)   {\color{gray} stimulus};
-    \node[] (labX) at (\posX, 2.3) {\color{gray} features};
-    \node[] (labE) at (\posE, 2.3) {\color{gray} causal};
-    \node[] (labE) at (\posE, 2)   {\color{gray} features};
-    \node[] (labN) at (\posN, 2)   {\color{gray} breathing\ldots};
-    \node[] (labF) at (\posN, 2.3) {\color{gray} bloodflow,};
-    \node[] (labF) at (\posF, 2)   {\color{gray} MEG helmet};
+    % \node[] (labY) at (\posY, 2)   {\color{gray} MEG recordings};
+    % \node[] (labX) at (\posX, 2)   {\color{gray} stimulus};
+    % \node[] (labX) at (\posX, 2.3) {\color{gray} features};
+    % \node[] (labE) at (\posE, 2.3) {\color{gray} causal};
+    % \node[] (labE) at (\posE, 2)   {\color{gray} features};
+    % \node[] (labN) at (\posN, 2)   {\color{gray} breathing\ldots};
+    % \node[] (labF) at (\posN, 2.3) {\color{gray} bloodflow,};
+    % \node[] (labF) at (\posF, 2)   {\color{gray} MEG helmet};
 

### 2019-05-23 paper/introduction.tex
Updates

--- 2019-05-23 paper/introduction.tex
@@
 Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations.
-By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart_neuroimage_2015, king2018encoding}.
+By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}.
 

### 2019-05-23 paper/meg_exps.tex
Updates

--- 2019-05-23 paper/meg_exps.tex
@@
 response filter (FIR) with a Hamming window and with transition bands of 0.1Hz
-and 10Hz for the low and high cut-off frequencies respectively.
+and 10Hz for the low and high cut-off frequencies.
 
@@
   \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/ridgecv_baseline_result.pdf}
-  \caption{Ridge regression baseline. All the coefficients appear to be
-  linearly correlated, unlike our method in Figure \ref{fig:megresult}.}
+  \caption{Regression baseline. Coefficients appear to be
+  linearly correlated, unlike our method Fig. \ref{fig:megresult}.}
   \label{fig:ridgebaselineresult}

### 2019-05-23 paper/paper.tex
Updates

--- 2019-05-23 paper/paper.tex
@@
-% TODO: Talk about cause-effect methods, trace method, half-sibling
-% TODO: Cite Roweis component analysis methods paper
-% TODO: read quantifying causal influences
-% TODO: review abstract
-% TODO: review conclusion
-
 \documentclass{article}
@@
 %
-To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split \citep{efron1992bootstrap}.
+To compensate for the reduction in sample size caused by the split, B2B is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split \citep{efron1992bootstrap}.
 %
-After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$ as a final B2BR prediction rule.
+After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$.
 
-We summarize the B2BR procedure in Algorithm~\ref{algorithm:b2br}.
+We summarize the B2B procedure in Algorithm~\ref{algorithm:b2br}.
 %
-The rest of this section provides a theoretical guarantee on the correctness of B2BR, and an optional post-processing step to binarize the estimated causal influence matrix.
+The rest of this section provides a theoretical guarantee on the correctness of B2B, and an optional post-processing step to binarize the estimated causal influence matrix.
 
@@
 
-%%% Next, let $k$ be the number of ones in the binary diagonal causal influence matrix $E$.
-%%% %
-%%% That is, $\text{tr}(E) = k$.
-%%% %
-%%% Without loss of generality, assume that these are the first $k$ variables in $X$.
-%%% %
-%%% Assume that $XEF$ is of rank $k$, which means that $F$ and $X$ have full rank on the space spanned by the active elements in $E$.
-%%% %
-%%% Let $K \in \mathbb{R}^{k \times d_{x}}$ be the diagonal matrix obtained by deleting the bottom $d_{x} - k$ rows of $E$.
-%%% %
-%%% %
-%%% First, we factorize $(XEF)^\dagger = ((XK^\top)(KF))^\dagger = (KF)^\dagger (XK^\top)^\dagger$.
-%%% %
-%%% Second, since $XK^\top$ is full rank, we write $(XK^\top)^\dagger = ((XK^\top)^\top (XK^\top))^{-1}(XK^\top)^\top$.
-%%% %
-%%% Then, $(XEF)^\dagger X = (KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X$.
+%%%%%% Next, let $k$ be the number of ones in the binary diagonal causal influence matrix $E$.
+%%%%%% %
+%%%%%% That is, $\text{tr}(E) = k$.
+%%%%%% %
+%%%%%% Without loss of generality, assume that these are the first $k$ variables in $X$.
+%%%%%% %
+%%%%%% Assume that $XEF$ is of rank $k$, which means that $F$ and $X$ have full rank on the space spanned by the active elements in $E$.
+%%%%%% %
+%%%%%% Let $K \in \mathbb{R}^{k \times d_{x}}$ be the diagonal matrix obtained by deleting the bottom $d_{x} - k$ rows of $E$.
+%%%%%% %
+%%%%%% %
+%%%%%% First, we factorize $(XEF)^\dagger = ((XK^\top)(KF))^\dagger = (KF)^\dagger (XK^\top)^\dagger$.
+%%%%%% %
+%%%%%% Second, since $XK^\top$ is full rank, we write $(XK^\top)^\dagger = ((XK^\top)^\top (XK^\top))^{-1}(XK^\top)^\top$.
+%%%%%% %
+%%%%%% Then, $(XEF)^\dagger X = (KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X$.
+%%%%%% 
+%%%%%% %Recall the rank factorisation property of the pseudo-inverse \citep{ben2003generalized}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
+%%%%%% %
+%%%%%% % Using this property on the minimal norm solution $\hat G$ yields
+%%%%%% % \begin{equation}
+%%%%%% % \begin{aligned}
+%%%%%% % (XEF)^\dagger X&=((XK^\top)(KF))^\dagger X\\
+%%%%%% % &=(KF)^\dagger ((XK^\top)^\top (XK^\top))^{-1} (XK^\top)^\top X\\
+%%%%%% % &=(KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X
+%%%%%% % \end{aligned}
+%%%%%% % \end{equation}
+%%%%%% 
+%%%%%% Write the covariance of $X$ as $X^\top X = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, with the block $\Sigma_1$ being $k\times k$.
+Then, $(KX^\top X K^\top)=\Sigma_{1}$ and $(KX^\top XK^\top)^{-1} KX^\top X=\left(\begin{array}{c|c}I_{k} &  \Sigma_{1}^{-1} \Sigma_{2}\end{array}\right) =: S$.
+%%%%%% 
+%%%%%% 
+%%%%%% Under centred noise, and if the covariance of NF has low condition number (a weak assumption given that N is noise), we have $\left \| NFG\right \| ^2 \approx Var(NF) \left \| G\right \| ^2$, and minimisation of $\left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda (KF)^\dagger S$, $\lambda$ a positive number below one.
+%%%%%% 
+%%%%%% Replacing in the left part of (ref), we have $X-XEFG=X- XK^\top KF \lambda (KF)^\dagger S= X - \lambda XE=X(I-\lambda E)$.
+Its square norm is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ ($Var(X_{.})$ denoting the variance of X over active and inactive features).
+%%%%%% 
+%%%%%% For the right part, we have $NFG=NF \lambda (KF)^\dagger S = NK^\top (KF+(I-K)F) \lambda  (KF)^\dagger S = \lambda NES=  \lambda NE$.
+(Since F is full rank over E, we have $(KF)^\dagger (I-K)F= 0$).
+The square norm is therefore $\lambda^2 Var(N_{active})$.
+%%%%%% 
+%%%%%% Thus, with centred and random noise, and with F full rank over E, we are minimising $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N_{active})$.
+Its least value is attained for $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N_{active})}$.
+Assuming X and N have the same variance over all features (this can always be guaranteed), active subscript can be dropped, and this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).
+%%%%%% 
+%%%%%% The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} (KF)^\dagger S$.
+%%%%%% 
+%%%%%% 
+%%%%%% Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} (K^\top K F)(KF)^{\dagger} S = \frac{1}{1+nsr} E$.
+%%%%%% 
+%%%%%% ADD AGAIN:
+%%%%%% 
+%%%%%% Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=(XEF)^\dagger X=(KF)^\dagger S$.
+If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $(KF)^\dagger K$.
+%%%=======
 %%%
-%%% %Recall the rank factorisation property of the pseudo-inverse \citep{ben2003generalized}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
-%%% %
-%%% % Using this property on the minimal norm solution $\hat G$ yields
-%%% % \begin{equation}
-%%% % \begin{aligned}
-%%% % (XEF)^\dagger X&=((XK^\top)(KF))^\dagger X\\
-%%% % &=(KF)^\dagger ((XK^\top)^\top (XK^\top))^{-1} (XK^\top)^\top X\\
-%%% % &=(KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X
-%%% % \end{aligned}
-%%% % \end{equation}
+%%%Let $X^\top X = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, the top left block being $k\times k$.
+Then, $(KX^\top X K^\top)=\Sigma_{1}$ and $(KX^\top XK^\top)^{-1} KX^\top X=\left(\begin{array}{cc}I_{k} &  \Sigma_{1}^{-1} \Sigma_{2}\end{array}\right)$, call it S.
 %%%
-%%% Write the covariance of $X$ as $X^\top X = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, with the block $\Sigma_1$ being $k\times k$.
-Then, $(KX^\top X K^\top)=\Sigma_{1}$ and $(KX^\top XK^\top)^{-1} KX^\top X=\left(\begin{array}{c|c}I_{k} &  \Sigma_{1}^{-1} \Sigma_{2}\end{array}\right) =: S$.
+%%%Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=(XEF)^\dagger X=(KF)^\dagger S$.
+If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $(KF)^\dagger K$.
 %%%
+%%%Under centred noise, and if the covariance of NF has low condition number (a weak assumption given that N is noise), we have $\left \| NFG\right \| ^2 \approx Var(NF) \left \| G\right \| ^2$, and minimisation of $\left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda (KF)^\dagger S$, $\lambda$ a positive number below one.
 %%%
-%%% Under centred noise, and if the covariance of NF has low condition number (a weak assumption given that N is noise), we have $\left \| NFG\right \| ^2 \approx Var(NF) \left \| G\right \| ^2$, and minimisation of $\left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda (KF)^\dagger S$, $\lambda$ a positive number below one.
-%%% 
-%%% Replacing in the left part of (ref), we have $X-XEFG=X- XK^\top KF \lambda (KF)^\dagger S= X - \lambda XE=X(I-\lambda E)$.
+%%%Replacing in the left part of (ref), we have $X-XEFG=X- XK^\top KF \lambda (KF)^\dagger S= X - \lambda XE=X(I-\lambda E)$.
 Its square norm is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ ($Var(X_{.})$ denoting the variance of X over active and inactive features).
 %%%
-%%% For the right part, we have $NFG=NF \lambda (KF)^\dagger S = NK^\top (KF+(I-K)F) \lambda  (KF)^\dagger S = \lambda NES=  \lambda NE$.
+%%%For the right part, we have $NFG=NF \lambda (KF)^\dagger S = NK^\top (KF+(I-K)F) \lambda  (KF)^\dagger S = \lambda NES=  \lambda NE$.
 (Since F is full rank over E, we have $(KF)^\dagger (I-K)F= 0$).
@@
 %%%
-%%% Thus, with centred and random noise, and with F full rank over E, we are minimising $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N_{active})$.
+%%%Thus, with centred and random noise, and with F full rank over E, we are minimising $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N_{active})$.
 Its least value is attained for $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N_{active})}$.
@@
 %%%
-%%% The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} (KF)^\dagger S$.
+%%%The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} S (KF)^\dagger$.
 %%%
+%%%The second step is straightforward, replacing Y, we have
+%%%\begin{equation}
+%%%\begin{aligned}
+%%%\arg \min_H \left \| \hat YG - XH \right \|^2 &=\arg  \min_H \left \| (XE + N)F \hat G - XH \right \|^2 \\
+%%%&=\arg \min_H \left \| X(EF \hat G - H) \right \| ^2 + \left \| NF\hat G \right \| ^2\\
+%%%&= \arg \min_H \left \| X(EF \hat G - H) \right \| ^2\\
+%%%&= EF \hat G
+%%%\end{aligned}
+%%%\end{equation}
 %%%
-%%% Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} (K^\top K F)(KF)^{\dagger} S = \frac{1}{1+nsr} E$.
-%%% 
-%%% ADD AGAIN:
-%%% 
-%%% Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=(XEF)^\dagger X=(KF)^\dagger S$.
-If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $(KF)^\dagger K$.
+%%%Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} (K^\top K F)(KF)^{\dagger} S = \frac{1}{1+nsr} E$.
+%%%>>>>>>> 4d110522fbb65b4da60327da2627241785a06113
 
@@
 \section{Related work}
+% scholkopf2016modeling
 
@@
 
+B2BR is also related, but different, from usual causal discovery algorithms \citep{peters2017elements} that hypothesize on the direction of causality, or from measures of causal influence requiring time information \citep{granger1969investigating, janzing2013quantifying}.
+
 \section{Conclusion}

### 2019-05-23 paper/synthetic.tex
Updates

@@
     \includegraphics[width=\textwidth]{synthetic.pdf}
-    \caption{Results of synthetic experiments.
-Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration, averaged over $20$ random seeds.}
+    \caption{Results of synthetic experiments. Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration, $20$ random seeds.}
     \label{table:synthetic}

### 2019-05-23 paper/meg_exps.tex
supplementary

--- 2019-05-23 paper/meg_exps.tex
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_sensors.pdf}
-  \caption{Magnetosensor response (femtoteslas fT) averaged across words for the first subject.
-The color coding corresponds to the positions of the sensors on the head as shown on the top-left diagram.
-The diverging curves correspond to artifacts in the MEG data.}
-  \label{fig:megavg}
-\end{figure}
-
 \subsubsection{Word features} We annotated each word with 38 features: 26
@@
 
-We attempt to identify which factors of $X$ explain the MEG
-recordings $Y$ at each time sample and for each subject separately. To this aim,
+We attempt to identify which word features ($X$) is reliably associated with the MEG recordings $Y$ at each time sample and for each subject separately. To this aim,
 we compare three methods: forward regression, backward regression, and back-to-back regression.
@@
   We plot the contributions of the \textit{word frequency} and the \textit{word
-  length} over time. Our method can disambiguate them consistently with
-  previous neuroscientific work showing an early response ($\approx$ 150ms) to
-  length and a later response ($\approx$ 400ms) to frequency.}
+  length} over time. B2B can disambiguate these two factors consistently with
+  previous neuroscientific work \cite{kutas2011thirty}\citep{pegado2014timing}
+  showing an early response ($\approx$ 150ms) to length and a later response
+  ($\approx$ 400ms) to frequency.}
   \label{fig:meg_twocurves}
@@
 
-Forward regression performs a regularized least-squares regression $A \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| A_{i, :} \|^2$.
+Forward regression was performed with an l2-regularized least-squares regression $H \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| H_{i, :} \|^2$.
 %
-Backward regression performs a regularized least-squares regression $B \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YB$, using a validation sample.
+Backward regression was performed with an l2-regularized least-squares regression $G \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YG$, using a validation sample.
 %
-Finally, we used the back-to-back regression as described earlier, using a bagging of 10 alphas logarithmically distributed between -5 and 5.
+Finally, we used the back-to-back regression was performed using two l2-regularized
+least squares regressions $G$ and $H$ as described in section \ref{sec:algorithm}, using a bagging of 10 alphas logarithmically distributed between -5 and 5.
 
@@
 as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau \in $ ranging from
-150ms to 1s after word onset. We use the resulting p-values as proxy to quantify
-the performance of each model. These results are depicted in Figure
+150ms to 1s after word onset. We use the resulting t and p-values as proxy to
+quantify the performance of each model. These results are depicted in Figure
 \ref{fig:meg_pval}. For concision purposes, we only report this metric averaged
@@
 word frequency modulate brain responses after the stimulus onset. However
-forward regression detected word length ($t-value=10.6$, $p=10^{-16}$) and
-word frequency ($t-value=11.1$, $p=10^-17$) less reliably than B2B (length:
-$t-value=18.7$, $p=10^{-31}$, frequency:$t-value=15.9$, $p=10^{-26}$). B2B
-was less sensitive than the backward regression (length: $t-value=33.6$,
-$p=10^{-50}$, $t-value=31.7$, frequency:$p=10^{-48}$). However, as detailed in
+forward regression detected word length ($t$-value$=10.6$, $p=10^{-16}$) and
+word frequency ($t$-value$=11.1$, $p=10^-17$) less reliably than B2B (length:
+$t$-value$=18.7$, $p=10^{-31}$, frequency:$t$-value$=15.9$, $p=10^{-26}$). B2B
+was less sensitive than the backward regression (length: $t$-value$=33.6$,
+$p=10^{-50}$, $t$-value$=31.7$, frequency:$p=10^{-48}$). However, as detailed in
 the introduction, backward models confound collinear factors. Consequently,

### 2019-05-23 paper/supplementary.tex
supplementary

\begin{figure}
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_sensors.pdf}
  \caption{Magnetosensor response (femtoteslas fT) averaged across words for the first subject. The color coding corresponds to the positions of the sensors on the head as shown on the top-left diagram. The diverging curves correspond to artifacts in the MEG data.}
  \label{fig:megavg}
\end{figure}

### 2019-05-23 paper/introduction.tex

--- 2019-05-23 paper/introduction.tex
@@
 Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations.
-By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \citep{weichwald2015causal, hebart2018deconstructing, king2018encoding}.
+By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}.
 

### 2019-05-23 paper/meg_exps.tex
supmat

@@
 segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
-thus led a sampling frequency of 240Hz. Figure \ref{fig:megavg} shows
-the average response of the magnetometers evoked by the words presented to the
-first subject.
+thus led a sampling frequency of 240Hz. Figure 1 in the supplementary material
+shows the average response of the magnetometers evoked by the words presented
+to the first subject.
 
@@
 We can observe that visual features start peaking around 150ms, while lexical features peak around 400ms.
-We show a similar plot for the Ridge regression baseline (forward) in Figure \ref{fig:ridgebaselineresult}.}
+We show a similar plot for the Ridge regression baseline (forward) in the supplementary material.}
   \label{fig:megresult}
 \end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/ridgecv_baseline_result.pdf}
-  \caption{Regression baseline. Coefficients appear to be
-  linearly correlated, unlike our method Fig. \ref{fig:megresult}.}
-  \label{fig:ridgebaselineresult}
-\end{figure}

### 2019-05-23 paper/meg_exps.tex
meg

@@
 segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
-thus led a sampling frequency of 240Hz. Figure 1 in the supplementary material
-shows the average response of the magnetometers evoked by the words presented
-to the first subject.
+thus led a sampling frequency of 240Hz. Figure \ref{fig:megavg} shows
+the average response of the magnetometers evoked by the words presented to the
+first subject.
 
@@
   \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_twocurves.pdf}
-  \caption{We applied the forward regression (a.k.a encoding coefficients), the backward regression (a.k.a decoding scores) and B2B on real MEG data.
+  \caption{We applied forward regression (a.k.a encoding coefficients), backward regression (a.k.a decoding scores) and B2B regression on 301-sensor MEG recordings of brain activity, while subjects were flashed with individual words every $\approx$650ms.
 Top row: the first subject.
 Bottom row: average across subjects.
-  We plot the contributions of the \textit{word frequency} and the \textit{word
-  length} over time. B2B can disambiguate these two factors consistently with
-  previous neuroscientific work \cite{kutas2011thirty}\citep{pegado2014timing}
-  showing an early response ($\approx$ 150ms) to length and a later response
-  ($\approx$ 400ms) to frequency.}
+Each line represent the association between the MEG signals and two word features as a function of time relative to word onset.
+Forward regression makes prediction for each MEG channel independently; detecting an effect thus requires an additional sum-square operation, which can dimish sensitivity.
+Backward regression operates across all channels, but does not learn to disambiguate the two word features.
+B2B efficiently disambiguates these two factors consistently with previous neuroscientific work \cite{kutas2011thirty}\citep{pegado2014timing}
+  showing early responses ($\approx$ 150ms) to word length and later responses
+  ($\approx$ 400ms) to word frequency.}
   \label{fig:meg_twocurves}
@@
 \subsubsection{Results}
-The number of letters in a string flashed onto the retina is known to modulate
-brain activity from 150ms after stimulus onset \cite{pegado2014timing}.
+When human subjects read individual words, the number of letters is known to modulate
+brain activity from 150ms after word onset \citep{pegado2014timing}.
 Similarly, word frequency is known to modulate brain activity around 400ms
-\cite{kutas2011thirty}. Consequently, we aimed to compare how sensitive each
-model is in detecting these two types of brain responses in a large
-neuroimaging dataset, where word length and word frequency were collinear.
+\citep{kutas2011thirty}.
+Consequently, we compared the ability of forward, backward and B2B regressions to specifically detect these two types of brain responses in a large neuroimaging dataset, where word length and word frequency are collinear.
 
-In Figure \ref{fig:meg_twocurves} we show, for each model, the effects obtained
+Figure \ref{fig:meg_twocurves} shows, for each model, the effects obtained
 in a representative subject, as well as the average effects across subjects.
 The results show that all models significantly detect that both word length and
-word frequency modulate brain responses after the stimulus onset. However
+word frequency modulate brain responses after word onset. However,
 forward regression detected word length ($t$-value$=10.6$, $p=10^{-16}$) and
 word frequency ($t$-value$=11.1$, $p=10^-17$) less reliably than B2B (length:
-$t$-value$=18.7$, $p=10^{-31}$, frequency:$t$-value$=15.9$, $p=10^{-26}$). B2B
+$t$-value$=18.7$, $p=10^{-31}$, frequency: $t$-value$=15.9$, $p=10^{-26}$). B2B
 was less sensitive than the backward regression (length: $t$-value$=33.6$,
-$p=10^{-50}$, $t$-value$=31.7$, frequency:$p=10^{-48}$). However, as detailed in
-the introduction, backward models confound collinear factors. Consequently,
-the high sensitivity of our backward regression is likely to come from a loss of specificity. This possibility is corroborated by the fact that the dynamics
-associated with word length and word frequency are remarkably similar in the
-backward model.
+$p=10^{-50}$, frequency: $t$-value$=31.7$, $p=10^{-48}$).
+However, as backward models confound collinear factors, such high sensitivity likely comes from a loss of specificity.
+This possibility is corroborated by the fact that the effects of word length and word frequency identified with backward regression present similar dynamics.
 
 Finally we investigate in Figure \ref{fig:megresult} whether B2B was sufficiently sensitive to reliably detect a multiplicity of subtle features, such as individual letters and part-of-speech.
-All of these features appeared to be reliably detected \cite{fig:megbar}.
+The vast majority the visual and lexical features appeared to be reliably detected \ref{fig:megbar}.
 B2B thus promises to improve the detectability of subtle effects.
@@
   \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_result.pdf}
-  \caption{We plot the coefficients corresponding to each of the features,
-  describing their importance over time after stimuli onsets.
-We can observe that visual features start peaking around 150ms, while lexical features peak around 400ms.
-We show a similar plot for the Ridge regression baseline (forward) in the supplementary material.}
+  \caption{Cumulative estimates of word features as a function of time relative to word onset (chance level = 0).
+The dark blue area (word length) and dark red area (word frequency) correspond to the same data as those of the B2B average across subjects presented in \label{fig:meg_twocurves}.
+A variety of visual features peak around 150ms, while a variety of lexical features tend to peak around 400ms.}
   \label{fig:megresult}

### 2019-05-23 paper/supplementary.tex

@@
-\begin{figure}
+\documentclass{article}
+
+\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref, amsthm}
+\usepackage[ruled, linesnumbered]{algorithm2e}
+\usepackage[]{neurips_2019}
+
+\title{Measuring causal influence with\\ back-to-back regression: the linear case - supplementary material}
+
+\begin{document}
+
+\maketitle
+
+\begin{figure}[h]
   \centering
@@
 \end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/ridgecv_baseline_result.pdf}
+  \caption{Regression baseline. Coefficients appear to be
+  linearly correlated, unlike our method in the main text.}
+  \label{fig:ridgebaselineresult}
+\end{figure}
+
+\end{document}

### 2019-05-23 paper/introduction.tex
intro

@@
 First, causal and non-causal factors may be numerous and collinear.
-In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or in pressure, as these two factors can be difficult to manipulate independently.
+In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or a change in pressure, as these two factors may, at first, be difficult to manipulate independently.
 This issue becomes increasingly pronounced as the number of potential factor increases.
-In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity as this factor covaries with numerous other factors such as word length (short words are more frequent than long words) word category (determinants are more frequent than adverbs), individual letters etc (e.g.
-words that contain "z" are rare) \citep{kutas2011thirty,pegado2014timing}.
+In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity because word frequency covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs), their individual letters (e.g.
+words that contain "z" are rare) and so forth \citep{kutas2011thirty,pegado2014timing}.
 Generating a set of words that control for all of these factors simultaneously can be challenging; it is thus common to use forward modeling, i.e.
@@
 word length, word frequency, e.g.
-\citep{huth2016natural}), and investigate the resulting parameters of the best model.
+\citep{huth2016natural}), and investigate, in the model, the estimated contribution of each factor \citep{friston1994statistical}.
 
@@
 For example, brain activity is generally recorded with hundreds of sensors simultaneously.
-It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving, king2016brain, kriegeskorte2008representational}.
+It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving, king2016brain, kriegeskorte2008representational, norman2006beyond}.
 
@@
 By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}.
+There also exist cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA), but they are notoriously difficult to interpretat \citep{lebart1995statistique}.
 

### 2019-05-23 paper/supplementary.tex
supplementary

@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm]{figures/pvalues_vertical.pdf}
+  \caption{We show the average value and the standard deviation of the coefficients obtained with out method during the trials.
+the $\star$  symbol corresponds to statistically significant (t-test) nonzero values.
+We notice in particular that the letters q, x, y are under-represented, as expected.
+Similarly there were only few proper nouns (PROPN) in text corpora.
+}
+  \label{fig:ridgebaselineresult}
+\end{figure}
+
 \end{document}

### 2019-05-23 paper/introduction.tex
a

@@
 By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}.
-There also exist cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA), but they are notoriously difficult to interpretat \citep{lebart1995statistique}.
+There also exist cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA), but they are notoriously difficult to interpretet \citep{lebart1995statistique}.
 
-We introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}) After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating causal influence and out-of-environment predictions (Section~\ref{sec:experiment_synthetic}).
+We introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}) After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating i) causal influence and ii) out-of-environment predictions (Section~\ref{sec:experiment_synthetic}).
 Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features leads to distinguishable brain responses (Section~\ref{sec:experiment_real}).

### 2019-05-23 paper/paper.tex
Scholkopf
a

@@
 %
-First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$, and construct the predictions $\hat X = \hat Y \hat G$.
+First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$, and construct the predictions $\hat X = Y \hat G$.
 %
-This backward regression recovers an approximation of the causes of $Y$ in $X$.
+This backward regression recovers the correlations between $Y$ and each factor of $X$.
 %
@@
 
-This section provides a consistency proof for B2BR in the case of low noise.
+This section provides a consistency proof for B2B in the case of low noise.
 
 \begin{theorem}[Low-noise consistency]
-    Consider the B2BR model from Equation~\ref{eq:model}.
+    Consider the B2B model from Equation~\ref{eq:model}.
     %
@@
     %
-    Then, as the variance of $N$ tends to zero and the sample size tends to infinity, B2BR recovers $E$.
+    Then, as the variance of $N$ tends to zero and the sample size tends to infinity, B2B recovers $E$.
 \end{theorem}
@@
 %
-In this section, we provide a recipe to binarize $\hat{E}$ to estimate the collection of causal factors.
+In this section, we provide a recipe to binarize $\hat{E}$ and estimate the collection of causal factors.
 
@@
 \section{Related work}
-% scholkopf2016modeling
 
@@
 
-B2BR is also related, but different, from usual causal discovery algorithms \citep{peters2017elements} that hypothesize on the direction of causality, or from measures of causal influence requiring time information \citep{granger1969investigating, janzing2013quantifying}.
+B2B is also related, but different, from usual causal discovery algorithms \citep{peters2017elements} that hypothesize on the direction of causality, or from measures of causal influence requiring time information \citep{granger1969investigating, janzing2013quantifying}.
+%
+B2B seems to have a similar flavour to half-sibling regression \citep{scholkopf2016modeling}, in the sense that we observe a large number of variables corrupted by the same source of noise; we leave the study of the precise link between these two models for future work.
 
@@
 
+% \input{supplementary.tex}
+
 %\section{Appendices}

### 2019-05-23 paper/synthetic.tex
a

@@
-We evaluate the performance of B2BR throughout a series of experiments on controlled synthetic data.
+We evaluate the performance of B2B throughout a series of experiments on controlled synthetic data.
 %
-The purpose of these experiments is to evaluate the ability of B2BR in terms of prediction (inside and outside the training distribution), as well as a method to recover causal factors.
+The purpose of these experiments is to evaluate the ability of B2B in terms of prediction (inside and outside the training distribution), as well as a method to recover causal factors.
 
-The data generating process for each experiment constructs $n$ training examples according to the model $Y = (\text{snr} \cdot XE + N)F$.
+The data generating process for each experiment constructs $n$ training examples according to the model $Y = (\text{snr} \cdot XE + N)F$, where $snr$ is a scalar that modulates the signal-to-noise ratio.
 %
@@
 
-We compare the performance of B2BR against five competing methods:
+We compare the performance of B2B against five competing methods:
 %
@@
 %
-For B2BR, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
+For B2B, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
 %
@@
 %
-B2BR is the method obtaining best test error out-distribution, lowest false positive and false negative discovery about $E$, and overall the best AUC when it comes to detecting causal influences.
+B2B is the method obtaining best test error out-distribution, lowest false positive and false negative discovery about $E$, and overall the best AUC when it comes to detecting causal influences.

### 2019-05-23 paper/meg_exps.tex
almost done fellas

@@
 quantify the performance of each model. These results are depicted in Figure
-\ref{fig:meg_pval}. For concision purposes, we only report this metric averaged
+4 of the Supplementary Material. For concision purposes, we only report this metric averaged
 between 150ms and 1s after word onset.
@@
 Finally we investigate in Figure \ref{fig:megresult} whether B2B was sufficiently sensitive to reliably detect a multiplicity of subtle features, such as individual letters and part-of-speech.
-The vast majority the visual and lexical features appeared to be reliably detected \ref{fig:megbar}.
+The vast majority the visual and lexical features appeared to be reliably detected (see Figure 3 in Supplementary Material).
 B2B thus promises to improve the detectability of subtle effects.
@@
   \caption{Cumulative estimates of word features as a function of time relative to word onset (chance level = 0).
-The dark blue area (word length) and dark red area (word frequency) correspond to the same data as those of the B2B average across subjects presented in \label{fig:meg_twocurves}.
+The dark blue area (word length) and dark red area (word frequency) correspond to the same data as those of the B2B average across subjects presented in \ref{fig:meg_twocurves}.
 A variety of visual features peak around 150ms, while a variety of lexical features tend to peak around 400ms.}

### 2019-05-23 paper/paper.tex
almost done fellas

@@
 
-We perform two experiments to evaluate the performance of BB2R: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography dataset.
+We perform two experiments to evaluate the performance of B2BR: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography dataset.
 %

### 2019-05-23 paper/supplementary.tex
almost done fellas

@@
 We notice in particular that the letters q, x, y are under-represented, as expected.
-Similarly there were only few proper nouns (PROPN) in text corpora.
+Similarly there were only few proper nouns (PROPN) in the word lists shown to subjects.
 }
@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm]{figures/pvalues.pdf}
+  \caption{We apply a statistical t-test to measure if coefficients were significantly different from the baseline value (at word onset, $t=0$ms).
+  We compute the negative logarithm of the p-value (\textit{higher is better}: nonzero effects are more easily detected). We plot the average across features, over time.
+  Full line: B2BR, dashed line: Forward regression. We see that our method more easily detects the influence of causes of MEG response.}
+\end{figure}
+
 \end{document}

### 2019-05-23 paper/paper.tex
aaa

--- 2019-05-23 paper/paper.tex
@@
 
-Back-to-back regression is closely related to Canonical Correlation Analysis \citep{cca_hotelling}.
+In the context of neuroimaging, the first regression of B2B relates to the first step of Representational Similarity Analysis (RSA) \citep{kriegeskorte2008representational}.
+Indeed, RSA analyzes the similarity of brain responses associated with specific conditions (e.g.
+distinct images), by (1) fitting one-against-all classifiers on each condition and (2) testing whether these classifiers can discriminate all other conditions.
+The resulting confusion matrix can then be analyzed in an unsupervised manner to reveal that some conditions lead to similar brain activity patterns.
+B2B differs from RSA in that (1) it uses regressions instead of classifications, and can thus generalize to new contexts and (2)
+it is fully supervised.
+
+More generally, B2B is related to CCA \citep{cca_hotelling} in that it can be thought of as a cross-decomposition of $X$ and $Y$.
+B2B departs from CCA in four main aspect.
+First, CCA performs an additional eigenvector decomposition what would be the $H$ coefficients in B2B \eqref{eq:solH}.
+Second, CCA does not typically optimize two distinct regularization parameters.
+Third CCA does not use bagging between $G$ and $H$.
+Finally, CCA does not diagonalize nor binarize the $H$ matrix to estimate $\hat E$.
+More generally, CCA aims to find maximally correlated components between $X$ and $Y$, and does not assume any causal direction.
+By contrast, B2B assumes that $X$ causes $Y$.
+Together, these differences may explain why B2B can reliably outperform CCA on estimating causal influences and predicting $Y$ in novel environments \ref{table:synthetic}.
 %
-While CCA computes the eigenvector decomposition of \eqref{eq:solH}, while we compute its diagonal.
-%
-If \eqref{eq:solH} is diagonal, as in our model \eqref{eq:model}, the variables in $X$ are the eigenvectors.
-%
-If all the associated eigenvalues are either $0$ or $1$, then all the variables are either part of the $Y$ subspace, or orthogonal to it, respectively.
-%
-This is because these eigenvalues correspond to the squared cosine of the angle between the corresponding variable in $X$ and the $Y$ subspace.
-%
-Thus, back-to-back regression is equivalent to CCA under the constraint that the subspaces spanned by $X$ and $Y$ are perpendicular, $X$ is spanned by vectors of $Y$ and vectors orthogonal to $Y$, and that the parallel and orthogonal components of $X$ are spanned by disjoint sets of features of $X$.
+% If all the associated eigenvalues are either $0$ or $1$, then all the variables are either part of the $Y$ subspace, or orthogonal to it, respectively.
+% %
+% This is because these eigenvalues correspond to the squared cosine of the angle between the corresponding variable in $X$ and the $Y$ subspace.
+% %
+% Thus, B2B is equivalent to CCA under the constraint that the subspaces spanned by $X$ and $Y$ are perpendicular, $X$ is spanned by vectors of $Y$ and vectors orthogonal to $Y$, and that the parallel and orthogonal components of $X$ are spanned by disjoint sets of features of $X$.
 
-There are, however, differences between CCA and back-to-back regression.
-%
-First, since our objective is not to characterise the proximity between $X$ and $Y$, but to measure causal influences, we can disregard the diagonalization of \eqref{eq:solH} performed by CCA.
-%
-Second, whereas CCA leverages all correlations between $X$ and $Y$, we filter some of them, through splitting and bagging.
+% There are, however, differences between CCA and back-to-back regression.
+% %
+% First, since our objective is not to characterise the proximity between $X$ and $Y$, but to measure causal influences, we can disregard the diagonalization of \eqref{eq:solH} performed by CCA.
+% %
+% Second, whereas CCA leverages all correlations between $X$ and $Y$, we filter some of them, through splitting and bagging.
 
@@
 \section{Conclusion}
-
 \input{conclusion}

### 2019-05-23 paper/conclusion.tex
conclusion

--- 2019-05-22 paper/conclusion.tex
@@
-In this work, we explored the back-to-back (B2B) regression, a novel method for
-identifying causes from observations building on two successive
-multidimensional regressions; from one domain to the other, and then back.
-We analyzed the asymptotic behavior of this approach, proving its convergence,
-and compared it to several baselines in controlled synthetic experiments. We
-showcased our approach on a recent brain imaging dataset, analyzing the timing
-of brain responses and their connection to features of words, and obtained
-results consistent with prior work in neuroscience literature, confirming the
-viability of our method in the case of real data.
+In this work, we proposed Back-to-Back (B2B) regression, a linear method to
+measure the causal influence of a potential set of variables generating multidimensional observations.
+%
+B2B performs two successive multidimensional regressions: one from the output domain, and another one from the input domain.
+%
+We provided a theoretical guarantee about the consistency of B2B, and compared it to several baselines in controlled synthetic experiments.
+%
+We also applied B2B to a recent brain imaging dataset, analyzing the timing
+of brain responses and their connection to word features.
+%
+We obtained results consistent with prior work in neuroscience literature, confirming the efficacy of B2B for real data analysis.
+%

### 2019-05-23 paper/introduction.tex
typos

--- 2019-05-23 paper/introduction.tex
@@
 In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or a change in pressure, as these two factors may, at first, be difficult to manipulate independently.
-This issue becomes increasingly pronounced as the number of potential factor increases.
+This issue becomes increasingly pronounced as the number of potential factors increases.
 In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity because word frequency covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs), their individual letters (e.g.
@@
 By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}.
-There also exist cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA), but they are notoriously difficult to interpretat \citep{lebart1995statistique}.
+There also exist cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA), but they are notoriously difficult to interpret \citep{lebart1995statistique}.
 
-We introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}) After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating causal influence and out-of-environment predictions (Section~\ref{sec:experiment_synthetic}).
-Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features leads to distinguishable brain responses (Section~\ref{sec:experiment_real}).
+We introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}).
+After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating i) causal influence and ii) out-of-environment predictions (Section~\ref{sec:experiment_synthetic}).
+Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features lead to distinguishable brain responses (Section~\ref{sec:experiment_real}).

### 2019-05-23 paper/paper.tex
typos

@@
 
-Back-to-Back Regression (B2B) consists in two steps.
+Back-to-Back Regression (B2B) consists of two steps.
 %
@@
 %
-By the definition of the Moore-Penrose, if $(NF)^\top (NF)$ is full-rank with $\| (NF)^\top (NF) \| \to 0$, it follows that $\hat{E} = EF(XEF)^\dagger X$.
+By definition of the Moore-Penrose inverse (also known as pseudo-inverse), if $(NF)^\top (NF)$ is full-rank with $\|~(NF)^\top~(NF)~\|~\to~0$, it follows that $\hat{E} = EF(XEF)^\dagger X$.
 %
@@
 %
-Therefore, $\text{Diag}(\hat{E})$ is the desired influence causal matrix $E$.
+Therefore, $\text{Diag}(\hat{E})$ is the desired causal influence matrix $E$.
 \end{proof}

### 2019-05-23 paper/introduction.tex
Typos

@@
 In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity because word frequency covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs), their individual letters (e.g.
-words that contain "z" are rare) and so forth \citep{kutas2011thirty,pegado2014timing}.
+words that contain `z' are rare) and so forth \citep{kutas2011thirty,pegado2014timing}.
 Generating a set of words that control for all of these factors simultaneously can be challenging; it is thus common to use forward modeling, i.e.
@@
 For example, brain activity is generally recorded with hundreds of sensors simultaneously.
-It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving, king2016brain, kriegeskorte2008representational, norman2006beyond}.
+It is thus common to use `decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving, king2016brain, kriegeskorte2008representational, norman2006beyond}.
 
@@
 
-We introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}).
+We introduce the `back-to-back regression', which combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}).
 After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating i) causal influence and ii) out-of-environment predictions (Section~\ref{sec:experiment_synthetic}).

### 2019-05-23 paper/meg_exps.tex
Typos

@@
 For each subject and each time sample relative to word onset, we
-build an observation matrix $Y \in \mathbb{R}^{n, q}$ of $n\approx$ 2700 words
-by $q=301$ MEG channels (273 magnetometers and 28 compensation channels). Each
+build an observation matrix $Y \in \mathbb{R}^{n \times d_y}$ of $n\approx$ 2700 words
+by $d_y=301$ MEG channels (273 magnetometers and 28 compensation channels). Each
 of the columns of $Y$ is normalized to have zero mean and unit variance.
@@
 number of letters) and lexical (part-of-speech and word frequency). This
-procedure yields an $X \in \mathbb{R}^{n, d}$ matrix of $n\approx$ 2700 words by
-$d=38$ features for each subject. Each of the columns of $X$ is normalized to
+procedure yields an $X \in \mathbb{R}^{n \times d_x}$ matrix of $n\approx$ 2700 words by
+$d_x=38$ features for each subject. Each of the columns of $X$ is normalized to
 have a mean and a standard deviation of 0 and 1 respectively.
@@
 
-Forward regression was performed with an l2-regularized least-squares regression $H \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| H_{i, :} \|^2$.
+Forward regression was performed with an $\ell_2$-regularized least-squares regression $H \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| H_{i, :} \|^2$.
 %
-Backward regression was performed with an l2-regularized least-squares regression $G \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YG$, using a validation sample.
+Backward regression was performed with an $\ell_2$-regularized least-squares regression $G \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YG$, using a validation sample.
 %
-Finally, we used the back-to-back regression was performed using two l2-regularized
+Finally, we used the back-to-back regression was performed using two $\ell_2$-regularized
 least squares regressions $G$ and $H$ as described in section \ref{sec:algorithm}, using a bagging of 10 alphas logarithmically distributed between -5 and 5.

### 2019-05-23 paper/paper.tex
Typos

@@
 
-    \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{n \times q}$}};
-    \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{n \times d}$}};
-    \node[] (annE) at (\posE, -1.8){\scalebox{0.85}{$E \in \mathbb{D}^{d \times d}$}};
-    \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{n \times d}$}};
-    \node[] (annF) at (\posF, -1.8){\scalebox{0.85}{$F \in \mathbb{R}^{d \times q}$}};
+    \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{n \times d_y}$}};
+    \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{n \times d_x}$}};
+    \node[] (annE) at (\posE, -1.8){\scalebox{0.85}{$E \in \mathbb{D}^{d_x \times d_x}$}};
+    \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{n \times d_x}$}};
+    \node[] (annF) at (\posF, -1.8){\scalebox{0.85}{$F \in \mathbb{R}^{d_x \times d_y}$}};
 
@@
 %
-The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{Diag}(E)$, is the desired estimate of the causal influence matrix $E$.
+The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{Diag}(\hat{H})$, is the desired estimate of the causal influence matrix $E$.
 
@@
         $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$ \Comment*[r]{$\hat H=(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2 \hat G$}
-        $\hat{E} \leftarrow \hat{E} + \text{Diag}(H)$\;
+        $\hat{E} \leftarrow \hat{E} + \text{Diag}(\hat{H})$\;
     }
@@
 %
-In this section, we provide a recipe to binarize $\hat{E}$ and estimate the collection of causal factors.
+In this section, we provide one of the many possible recipes to binarize $\hat{E}$ and estimate the collection of causal factors.
 
@@
 %
-To binarize $\hat{E}$, set to one all the diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the diagonal entries.
+To binarize $\hat{E}$, set to one all the $p$ diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the $d_x-p$ diagonal entries.
 
@@
 
-We perform two experiments to evaluate the performance of B2BR: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography dataset.
+We perform two experiments to evaluate the performance of B2B: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography dataset.
 %

### 2019-05-23 paper/synthetic.tex
Typos

@@
 
-The data generating process for each experiment constructs $n$ training examples according to the model $Y = (\text{snr} \cdot XE + N)F$, where $snr$ is a scalar that modulates the signal-to-noise ratio.
+The data generating process for each experiment constructs $n$ training examples according to the model $Y = (\text{snr} \cdot XE + N)F$, where $\text{snr}$ is a scalar that modulates the signal-to-noise ratio.
 %
@@
 %
-The ``in-distribution'' metrics measure the test error of each predictor before and after weighting their features by the estimated mask $\hat{E}$.
+The ``in-distribution'' metrics measure the test error of each predictor before and after weighting their features by the estimated (continous!) mask $\hat{E}$.
 %
@@
     \includegraphics[width=\textwidth]{synthetic.pdf}
-    \caption{Results of synthetic experiments. Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration, $20$ random seeds.}
+    \caption{Results of synthetic experiments.
+Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration averaged over $20$ random seeds.}
     \label{table:synthetic}
@@
 %
-B2B is the method obtaining best test error out-distribution, lowest false positive and false negative discovery about $E$, and overall the best AUC when it comes to detecting causal influences.
+B2B is the method obtaining best test errors, lowest false positive and false negative discovery about $E$, and overall the best AUC when it comes to detecting causal influences.

### 2019-05-23 paper/synthetic.tex
typo

@@
 %
-The ``in-distribution'' metrics measure the test error of each predictor before and after weighting their features by the estimated (continous!) mask $\hat{E}$.
+The ``in-distribution'' metrics measure the test error of each predictor before and after weighting their features by the estimated (continuous!) mask $\hat{E}$.
 %

### 2019-05-23 paper/meg_exps.tex
typos

@@
 
-We attempt to identify which word features ($X$) is reliably associated with the MEG recordings $Y$ at each time sample and for each subject separately. To this aim,
+We attempt to identify which word features ($X$) are reliably associated with the MEG recordings $Y$ at each time sample and for each subject separately. To this aim,
 we compare three methods: forward regression, backward regression, and back-to-back regression.
@@
 Finally, we used the back-to-back regression was performed using two $\ell_2$-regularized
-least squares regressions $G$ and $H$ as described in section \ref{sec:algorithm}, using a bagging of 10 alphas logarithmically distributed between -5 and 5.
+least squares regressions $G$ and $H$ as described in section \ref{sec:algorithm}, using a bagging factor of 10, and alpha cross-validated between $10^{-5}$ and $10^{5}$.
 
@@
 sample independently. To estimate the reliability of these estimates in the
-population, we tested, using one-sample $t$-test whether the distribution of
-the estimated contribution of each factor different from baseline ($t=0$ms),
-as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau \in $ ranging from
+population, we tested, using a one-sample $t$-test whether the distribution of
+the estimated contribution of each factor differed from baseline ($t=0$ms),
+as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau$ ranging from
 150ms to 1s after word onset. We use the resulting t and p-values as proxy to
 quantify the performance of each model. These results are depicted in Figure
-4 of the Supplementary Material. For concision purposes, we only report this metric averaged
+3 of the Supplementary Material. For concision purposes, we only report this metric averaged
 between 150ms and 1s after word onset.
@@
 Finally we investigate in Figure \ref{fig:megresult} whether B2B was sufficiently sensitive to reliably detect a multiplicity of subtle features, such as individual letters and part-of-speech.
-The vast majority the visual and lexical features appeared to be reliably detected (see Figure 3 in Supplementary Material).
+The vast majority of the visual and lexical features appeared to be reliably detected (see Figure 4 in Supplementary Material).
 B2B thus promises to improve the detectability of subtle effects.

### 2019-05-23 paper/paper.tex
typos

@@
 More generally, B2B is related to CCA \citep{cca_hotelling} in that it can be thought of as a cross-decomposition of $X$ and $Y$.
-B2B departs from CCA in four main aspect.
+B2B departs from CCA in four main aspects.
 First, CCA performs an additional eigenvector decomposition what would be the $H$ coefficients in B2B \eqref{eq:solH}.
 Second, CCA does not typically optimize two distinct regularization parameters.
-Third CCA does not use bagging between $G$ and $H$.
+Third, CCA does not use bagging between $G$ and $H$.
 Finally, CCA does not diagonalize nor binarize the $H$ matrix to estimate $\hat E$.
@@
 
-B2B is also related, but different, from usual causal discovery algorithms \citep{peters2017elements} that hypothesize on the direction of causality, or from measures of causal influence requiring time information \citep{granger1969investigating, janzing2013quantifying}.
+B2B is also related to, but different from usual causal discovery algorithms \citep{peters2017elements} that hypothesize on the direction of causality, or from measures of causal influence requiring time information \citep{granger1969investigating, janzing2013quantifying}.
 %

### 2019-05-23 paper/introduction.tex
jr

@@
 This issue becomes increasingly pronounced as the number of potential factors increases.
-In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity because word frequency covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs), their individual letters (e.g.
+In neuroscience, for example, identifying whether the frequency of a word presented on a subject's retina modulates brain activity can be surprisingly difficult.
+Indeed, word frequency covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs), their individual letters (e.g.
 words that contain `z' are rare) and so forth \citep{kutas2011thirty,pegado2014timing}.
-Generating a set of words that control for all of these factors simultaneously can be challenging; it is thus common to use forward modeling, i.e.
-to train a model to predict observations (e.g.
+Instead of selecting a set of words that control for all of these factors simultaneously, it is thus common to use forward modeling, i.e.
+to train a mathematical model to predict observations (e.g.
 brain activity) from a minimal combination of competing factors (e.g.
@@
 The second challenge to measuring causal influence is that observations can be large and complex.
-The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from large and complex observations.
-For example, brain activity is generally recorded with hundreds of sensors simultaneously.
+The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from multidimensional observations.
+For example, brain activity is often recorded with hundreds or thousands of sensors simultaneously.
 It is thus common to use `decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving, king2016brain, kriegeskorte2008representational, norman2006beyond}.
@@
 By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}.
-There also exist cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA), but they are notoriously difficult to interpret \citep{lebart1995statistique}.
+There also exist cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA), but their resulting coefficients are notoriously difficult to interpret \citep{lebart1995statistique}.
 

### 2019-05-23 paper/meg_exps.tex
jr

@@
 
-The raw data was then segmented 100ms before word onset and 1.000 sec after
-word onset. Time $t=0$ms corresponds to the word onset. Finally, each resulting
+The raw data was then segmented 100ms before word onset and 1s after
+word onset ($t=0$ms corresponds to word onset). Finally, each resulting
 segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
@@
 
-We attempt to identify which word features ($X$) are reliably associated with the MEG recordings $Y$ at each time sample and for each subject separately. To this aim,
+Our goal is to identify which word features ($X$) lead to a reliable modulation of brain activity as measured with MEG recordings ($Y$) at each time sample and for each subject separately.
+To this aim,
 we compare three methods: forward regression, backward regression, and back-to-back regression.
@@
 Bottom row: average across subjects.
-Each line represent the association between the MEG signals and two word features as a function of time relative to word onset.
-Forward regression makes prediction for each MEG channel independently; detecting an effect thus requires an additional sum-square operation, which can dimish sensitivity.
+Each line represent the link between word features $X$ and MEG signals $Y$
+  as a function of time relative to word onset ($x$ axis).
+The $y$-axis varies for each model (forward: H coefficients; backward: pearson correlation coefficients; B2B: $\hat E$).
+Forward regression makes prediction for each MEG channel independently; detecting an effect thus requires an additional average of the norm and can diminish sensitivity.
 Backward regression operates across all channels, but does not learn to disambiguate the two word features.
-B2B efficiently disambiguates these two factors consistently with previous neuroscientific work \cite{kutas2011thirty}\citep{pegado2014timing}
+B2B disambiguates these two factors consistently with previous neuroscientific work \cite{kutas2011thirty}\citep{pegado2014timing}
   showing early responses ($\approx$ 150ms) to word length and later responses
@@
 $p=10^{-50}$, frequency: $t$-value$=31.7$, $p=10^{-48}$).
-However, as backward models confound collinear factors, such high sensitivity likely comes from a loss of specificity.
-This possibility is corroborated by the fact that the effects of word length and word frequency identified with backward regression present similar dynamics.
+However, as backward models confound collinear factors, this high sensitivity likely comes from a loss of specificity.
+This possibility is corroborated by the fact that the effects of word length and word frequency identified with backward regression present largely similar dynamics.
 

### 2019-05-23 paper/supplementary.tex
Typos
typos
jr

--- 2019-05-23 paper/supplementary.tex
@@
   \centering
-  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/ridgecv_baseline_result.pdf}
-  \caption{Regression baseline. Coefficients appear to be
-  linearly correlated, unlike our method in the main text.}
+  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm]{figures/pvalues.pdf}
+  \caption{We apply a statistical t-test to measure if coefficients were significantly different from baseline value (at word onset, $t=0$ms).
+  We compute the negative logarithm of the p-value (\textit{higher is better}: nonzero effects are more easily detected).
+We plot the average across features, over time.
+Red: word length, blue: word frequency.
+  Full line: B2B, dashed line: Forward regression. We see that our method more easily detects the influence of causes of MEG response.}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm]{figures/pvalues_vertical.pdf}
+  \caption{We show the average value and the standard deviation across subjects of the coefficients obtained with our method during the trials.
+the $\star$  symbol corresponds to statistically significant (t-test) nonzero values.}
   \label{fig:ridgebaselineresult}

### 2019-05-23 paper/synthetic.tex
jr

@@
     \caption{Results of synthetic experiments.
-Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration averaged over $20$ random seeds.}
+Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration averaged over $20$ random seeds.
+The prediction error on $Y$ within (in) and outside (out) of the training distribution are weakly but consistantly larger for all of the models as compared to B2B with and without the Sonquist binarization (E).
+The AUC, measuring the ability of each model to separate causal from noncausal factors, is consistantly larger in B2B as compared to the other models.}
     \label{table:synthetic}

### 2019-05-23 paper/introduction.tex
typo

@@
 For example, brain activity is often recorded with hundreds or thousands of sensors simultaneously.
-It is thus common to use `decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving, king2016brain, kriegeskorte2008representational, norman2006beyond}.
+It is thus common to use `decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving,  kriegeskorte2008representational, norman2006beyond}.
 

### 2019-05-23 paper/supplementary.tex
new theorem

--- 2019-05-23 paper/supplementary.tex
@@
 
+\clearpage
+\newpage
+A stronger theorem can be proven, that includes provision for noise
+
+THEOREM
+
+    Consider the B2B model from Equation~\ref{eq:model}.
+    %
+    Assume that $F$ and $X$ are full-rank on the subspace spanned by active entries in $E$.
+    %
+    Then, the solution of B2B, $\hat H$, will minimise 
+    %
+    $\min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2$
+
+PROOF
+
+Let $\hat G$ and $\hat H$ be the solutions of the first and second step of B2B, $\hat H=EF\hat G$ and $\hat G$ minimises 
+\begin{equation}
+	\left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
+    \label{eq:doublenorm}
+\end{equation}
+
+Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero.
+
+Let $Z=F^\dagger EF\hat G$, we have $FZ = FF^\dagger EF  \hat G= EF\hat G$ since F is full rank over E. and since E is a contraction, we have 
+$$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \left \| X - XEF\hat G\right \| ^2  + \left \| NEF\hat G\right \| ^2 \leq \left \| X - XEF\hat G\right \| ^2  + \left \| NF\hat G\right \| ^2$$
+
+Since $\hat G$ attains the minimum value of \ref{eq:doublenorm}, we must have $EF\hat G = F\Hat G$.
+
+Therefore, $\left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 = \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2$ are minimised by $\hat G$ and $\hat H$.
+Noting that $\hat H = E\hat H$ completes the proof.
+
+The solution of $\min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2$ is $\hat H= (E X^\top XE +EN^\top NE) ^\dagger EXX^\top$.
+
+If the covariance matrices of $NE$ and $XE$ (the restriction of N and X to the active causal features) can be co-diagonalised, with orthogonal transformation $U$, then the diagonal elements of $\hat H$ will be those of $U^\top  \frac{\sigma_X}{\sigma_X + \sigma_N} U $.
+
+In any case, the $d_x-k$ last diagonal elements will be zero.
+All others will be bounded by the maximum and minimum value of $\frac{\sigma_X}{\sigma_X + \sigma_N}$ over all eigenvalues of the covariance matrices of $XE$ and $NE$.
+If the eigenvalues are sorted from largest to smallest, diagonal elements of $\hat H$ will be between $\frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$ and $\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}}$.
+
 \end{document}

### 2019-05-23 paper/paper.tex

@@
 
-    % \node[] (labY) at (\posY, 2)   {\color{gray} MEG recordings};
-    % \node[] (labX) at (\posX, 2)   {\color{gray} stimulus};
-    % \node[] (labX) at (\posX, 2.3) {\color{gray} features};
-    % \node[] (labE) at (\posE, 2.3) {\color{gray} causal};
-    % \node[] (labE) at (\posE, 2)   {\color{gray} features};
-    % \node[] (labN) at (\posN, 2)   {\color{gray} breathing\ldots};
-    % \node[] (labF) at (\posN, 2.3) {\color{gray} bloodflow,};
-    % \node[] (labF) at (\posF, 2)   {\color{gray} MEG helmet};
-
     \node[] (sim1) at (0,-3.25) {\scalebox{0.85}{$X \sim P(X)$}};
@@
 
-%Without loss of generality, we may assume N to be centred(adding, when necessary, an intercept to the causal features), and redefine features and noise so that the model be written $Y=f(EX+N)$ (f a function, E, X and N matrices).
-In particular, the addition of "external noise" M, as in $Y=F(EX+N)+M$ can be viewed as a small change in F and N.
-
-%We consider both the linear and non linear case.
-In the first one, our model is $$Y = F(EX+N)$$, with  $Y\in\mathbb{R}^{n\times ny}$, $F\in\mathbb{R}^{ny\times nx}$, $X\in\mathbb{R}^{nx\times n}$, $N\in\mathbb{R}^{nx\times n}$, and E a binary diagonal square matrice of size nx.
-
-% In the non linear model, $F(EX+N)$ is defined as before, but Y is the result of a non linear transformation of it $$Y=\sigma (F(EX+N))$$
-
-%Under this formalism, we are given samples of Y and X, and are tasked to recover E and use it to predict Y from X.
-
 \subsection{Algorithm}
@@
 where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices of regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.
-
-% If, in the above formalism, F is invertible with respect to the causal features of X (ie the problem has a solution), and the relation is linear or close to linear, the diagonal of $\hat H$ will be a scaled approximation of the diagonal of E.
-That is to say, back to back regression will recover the causal features of the process considered, even in the presence of heavy noise N, and without needing to estimate the transformation F.
-
-% In practical cases, the covariance matrices of Y and X will often be badly conditioned (because features of X are related, or measurement produces correlated features of Y).
-This makes the inversion of $X'X$ and $Y'Y$ sensitive to noise (either from N, from sampling or from computation rounding).
-To guard against high condition numbers, we have used two methods: truncated SVD (ref) and ridge regression (ref).
-
-% In both, we compute the singular value decompositions of the covariance matrix, transform its diagonal part, and invert it.
-In truncated SVD, the lowest eigenvalues are set to zero, and the corresponding vectors eliminated from future calculations.
-In ridge regression, a constant factor is added to all eigenvalues.
-The regularisation factor and truncation threshold are chosen via leave one out cross validation.
-In practice, ridge regression tends to provide better stability but introduces a bias in the estimates.
-Truncated SVD is used to provide better results in low noise cases.
 
@@
 \label{sec:theorem}
-% As discussed in the introduction, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(EX+N)$, with f an unknown function.
-% %
-% If F is linear we can rewrite it as $Y = F(EX + N)$, with X an (dx, n) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
-% %
-% N a (dx,n) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dx) transformation corresponding to the measuring apparatus, and Y a (dy, n) matrix of measured effects.
-%
-% The two steps of back-to-back regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).
 
@@
 
-%%%%%% Next, let $k$ be the number of ones in the binary diagonal causal influence matrix $E$.
-%%%%%% %
-%%%%%% That is, $\text{tr}(E) = k$.
-%%%%%% %
-%%%%%% Without loss of generality, assume that these are the first $k$ variables in $X$.
-%%%%%% %
-%%%%%% Assume that $XEF$ is of rank $k$, which means that $F$ and $X$ have full rank on the space spanned by the active elements in $E$.
-%%%%%% %
-%%%%%% Let $K \in \mathbb{R}^{k \times d_{x}}$ be the diagonal matrix obtained by deleting the bottom $d_{x} - k$ rows of $E$.
-%%%%%% %
-%%%%%% %
-%%%%%% First, we factorize $(XEF)^\dagger = ((XK^\top)(KF))^\dagger = (KF)^\dagger (XK^\top)^\dagger$.
-%%%%%% %
-%%%%%% Second, since $XK^\top$ is full rank, we write $(XK^\top)^\dagger = ((XK^\top)^\top (XK^\top))^{-1}(XK^\top)^\top$.
-%%%%%% %
-%%%%%% Then, $(XEF)^\dagger X = (KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X$.
-%%%%%%
-%%%%%% %Recall the rank factorisation property of the pseudo-inverse \citep{ben2003generalized}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
-%%%%%% %
-%%%%%% % Using this property on the minimal norm solution $\hat G$ yields
-%%%%%% % \begin{equation}
-%%%%%% % \begin{aligned}
-%%%%%% % (XEF)^\dagger X&=((XK^\top)(KF))^\dagger X\\
-%%%%%% % &=(KF)^\dagger ((XK^\top)^\top (XK^\top))^{-1} (XK^\top)^\top X\\
-%%%%%% % &=(KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X
-%%%%%% % \end{aligned}
-%%%%%% % \end{equation}
-%%%%%%
-%%%%%% Write the covariance of $X$ as $X^\top X = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, with the block $\Sigma_1$ being $k\times k$.
-Then, $(KX^\top X K^\top)=\Sigma_{1}$ and $(KX^\top XK^\top)^{-1} KX^\top X=\left(\begin{array}{c|c}I_{k} &  \Sigma_{1}^{-1} \Sigma_{2}\end{array}\right) =: S$.
-%%%%%%
-%%%%%%
-%%%%%% Under centred noise, and if the covariance of NF has low condition number (a weak assumption given that N is noise), we have $\left \| NFG\right \| ^2 \approx Var(NF) \left \| G\right \| ^2$, and minimisation of $\left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda (KF)^\dagger S$, $\lambda$ a positive number below one.
-%%%%%%
-%%%%%% Replacing in the left part of (ref), we have $X-XEFG=X- XK^\top KF \lambda (KF)^\dagger S= X - \lambda XE=X(I-\lambda E)$.
-Its square norm is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ ($Var(X_{.})$ denoting the variance of X over active and inactive features).
-%%%%%%
-%%%%%% For the right part, we have $NFG=NF \lambda (KF)^\dagger S = NK^\top (KF+(I-K)F) \lambda  (KF)^\dagger S = \lambda NES=  \lambda NE$.
-(Since F is full rank over E, we have $(KF)^\dagger (I-K)F= 0$).
-The square norm is therefore $\lambda^2 Var(N_{active})$.
-%%%%%%
-%%%%%% Thus, with centred and random noise, and with F full rank over E, we are minimising $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N_{active})$.
-Its least value is attained for $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N_{active})}$.
-Assuming X and N have the same variance over all features (this can always be guaranteed), active subscript can be dropped, and this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).
-%%%%%%
-%%%%%% The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} (KF)^\dagger S$.
-%%%%%%
-%%%%%%
-%%%%%% Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} (K^\top K F)(KF)^{\dagger} S = \frac{1}{1+nsr} E$.
-%%%%%%
-%%%%%% ADD AGAIN:
-%%%%%%
-%%%%%% Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=(XEF)^\dagger X=(KF)^\dagger S$.
-If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $(KF)^\dagger K$.
-%%%=======
-%%%
-%%%Let $X^\top X = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, the top left block being $k\times k$.
-Then, $(KX^\top X K^\top)=\Sigma_{1}$ and $(KX^\top XK^\top)^{-1} KX^\top X=\left(\begin{array}{cc}I_{k} &  \Sigma_{1}^{-1} \Sigma_{2}\end{array}\right)$, call it S.
-%%%
-%%%Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=(XEF)^\dagger X=(KF)^\dagger S$.
-If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $(KF)^\dagger K$.
-%%%
-%%%Under centred noise, and if the covariance of NF has low condition number (a weak assumption given that N is noise), we have $\left \| NFG\right \| ^2 \approx Var(NF) \left \| G\right \| ^2$, and minimisation of $\left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda (KF)^\dagger S$, $\lambda$ a positive number below one.
-%%%
-%%%Replacing in the left part of (ref), we have $X-XEFG=X- XK^\top KF \lambda (KF)^\dagger S= X - \lambda XE=X(I-\lambda E)$.
-Its square norm is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ ($Var(X_{.})$ denoting the variance of X over active and inactive features).
-%%%
-%%%For the right part, we have $NFG=NF \lambda (KF)^\dagger S = NK^\top (KF+(I-K)F) \lambda  (KF)^\dagger S = \lambda NES=  \lambda NE$.
-(Since F is full rank over E, we have $(KF)^\dagger (I-K)F= 0$).
-The square norm is therefore $\lambda^2 Var(N_{active})$.
-%%%
-%%%Thus, with centred and random noise, and with F full rank over E, we are minimising $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N_{active})$.
-Its least value is attained for $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N_{active})}$.
-Assuming X and N have the same variance over all features (this can always be guaranteed), active subscript can be dropped, and this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).
-%%%
-%%%The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} S (KF)^\dagger$.
-%%%
-%%%The second step is straightforward, replacing Y, we have
-%%%\begin{equation}
-%%%\begin{aligned}
-%%%\arg \min_H \left \| \hat YG - XH \right \|^2 &=\arg  \min_H \left \| (XE + N)F \hat G - XH \right \|^2 \\
-%%%&=\arg \min_H \left \| X(EF \hat G - H) \right \| ^2 + \left \| NF\hat G \right \| ^2\\
-%%%&= \arg \min_H \left \| X(EF \hat G - H) \right \| ^2\\
-%%%&= EF \hat G
-%%%\end{aligned}
-%%%\end{equation}
-%%%
-%%%Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} (K^\top K F)(KF)^{\dagger} S = \frac{1}{1+nsr} E$.
-%%%>>>>>>> 4d110522fbb65b4da60327da2627241785a06113
-
-%The diagonal of $\hat H$ recovered by back to back regression is a scaled approximation of the diagonal of E.
-Noise tends reduce the large diagonal elements, whereas the bias introduced by regularisation tends to increase the lowest ones.
-% \subsection{Asymptotic behaviour - non linear case}
-
 \subsection{Binarizing $\hat{E}$}
@@
 
-More generally, B2B is related to CCA \citep{cca_hotelling} in that it can be thought of as a cross-decomposition of $X$ and $Y$.
-B2B departs from CCA in four main aspects.
-First, CCA performs an additional eigenvector decomposition what would be the $H$ coefficients in B2B \eqref{eq:solH}.
+More generally, B2B is related to CCA \citep{cca_hotelling}, but departs from it in four main aspects.
+First, CCA performs an additional eigenvector decomposition of what would be the $H$ coefficients in B2B \eqref{eq:solH}.
 Second, CCA does not typically optimize two distinct regularization parameters.
@@
 By contrast, B2B assumes that $X$ causes $Y$.
-Together, these differences may explain why B2B can reliably outperform CCA on estimating causal influences and predicting $Y$ in novel environments \ref{table:synthetic}.
-%
-% If all the associated eigenvalues are either $0$ or $1$, then all the variables are either part of the $Y$ subspace, or orthogonal to it, respectively.
-% %
-% This is because these eigenvalues correspond to the squared cosine of the angle between the corresponding variable in $X$ and the $Y$ subspace.
-% %
-% Thus, B2B is equivalent to CCA under the constraint that the subspaces spanned by $X$ and $Y$ are perpendicular, $X$ is spanned by vectors of $Y$ and vectors orthogonal to $Y$, and that the parallel and orthogonal components of $X$ are spanned by disjoint sets of features of $X$.
-
-% There are, however, differences between CCA and back-to-back regression.
-% %
-% First, since our objective is not to characterise the proximity between $X$ and $Y$, but to measure causal influences, we can disregard the diagonalization of \eqref{eq:solH} performed by CCA.
-% %
-% Second, whereas CCA leverages all correlations between $X$ and $Y$, we filter some of them, through splitting and bagging.
+Together, these differences may explain why B2B can reliably outperform CCA on estimating causal influences and predicting $Y$ in novel environments (Fig.
+\ref{table:synthetic}).
 
@@
 \input{conclusion}
-
-% \section{Acknowlegements}
-% Data were provided (in part) by the Donders Institute for Brain, Cognition and Behaviour.
 

### 2019-05-23 paper/synthetic.tex

--- 2019-05-23 paper/synthetic.tex
@@
     \caption{Results of synthetic experiments.
-Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration averaged over $20$ random seeds.}
+Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration averaged over $20$ random seeds.
+The prediction error on $Y$ within (in) and outside (out) of the training distribution are weakly but consistently larger for all of the models as compared to B2B with and without the Sonquist binarization (E).
+The AUC, measuring the ability of each model to separate causal from noncausal factors, is consistently larger in B2B as compared to the other models.}
     \label{table:synthetic}

### 2019-05-23 paper/meg_exps.tex
back to 8 pages

@@
 
-\subsubsection{MEG preprocessing}
+\subsubsection{MEG preprocessing and feature generation}
 
@@
 response filter (FIR) with a Hamming window and with transition bands of 0.1Hz
-and 10Hz for the low and high cut-off frequencies.
-
-The raw data was then segmented 100ms before word onset and 1s after
+and 10Hz for the low and high cut-off frequencies. The raw data was then segmented 100ms before word onset and 1s after
 word onset ($t=0$ms corresponds to word onset). Finally, each resulting
@@
 shows the average response of the magnetometers evoked by the words presented
-to the first subject.
-
-For each subject and each time sample relative to word onset, we
+to the first subject. For each subject and each time sample relative to word onset, we
 build an observation matrix $Y \in \mathbb{R}^{n \times d_y}$ of $n\approx$ 2700 words
@@
 
-\subsubsection{Word features} We annotated each word with 38 features: 26
+We annotated each word with 38 features: 26
 correspond to the count of each letter in the word (from 'a' to 'z',
@@
 
-\subsubsection{Methods}
+\subsubsection{Models and statistics}
 
@@
 Forward regression was performed with an $\ell_2$-regularized least-squares regression $H \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| H_{i, :} \|^2$.
-%
 Backward regression was performed with an $\ell_2$-regularized least-squares regression $G \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YG$, using a validation sample.
-%
 Finally, we used the back-to-back regression was performed using two $\ell_2$-regularized
 least squares regressions $G$ and $H$ as described in section \ref{sec:algorithm}, using a bagging factor of 10, and alpha cross-validated between $10^{-5}$ and $10^{5}$.
-
-\subsubsection{Second-level statistics}
 

### 2019-05-23 paper/supplementary.tex

@@
   \centering
-  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/ridgecv_baseline_result.pdf}
-  \caption{Regression baseline. Coefficients appear to be
-  linearly correlated, unlike our method in the main text.}
-  \label{fig:ridgebaselineresult}
+  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm]{figures/pvalues.pdf}
+  \caption{We apply a statistical t-test to measure if coefficients were significantly different from baseline value (at word onset, $t=0$ms).
+  We compute the negative logarithm of the p-value (\textit{higher is better}: nonzero effects are more easily detected).
+We plot the average across features, over time.
+Red: word length, blue: word frequency.
+  Full line: B2B, dashed line: Forward regression. We see that our method more easily detects the influence of causes of MEG response.}
 \end{figure}
@@
   \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm]{figures/pvalues_vertical.pdf}
-  \caption{We show the average value and the standard deviation of the coefficients obtained with out method during the trials.
-the $\star$  symbol corresponds to statistically significant (t-test) nonzero values.
-We notice in particular that the letters q, x, y are under-represented, as expected.
-Similarly there were only few proper nouns (PROPN) in the word lists shown to subjects.
-}
+  \caption{We show the average value and the standard deviation across subjects of the coefficients obtained with our method during the trials.
+the $\star$  symbol corresponds to statistically significant (t-test) nonzero values.}
   \label{fig:ridgebaselineresult}

### 2019-05-24 paper/supplementary.tex
hide francois last theorem because does not render, feel free to fix
second theorem

--- 2019-05-23 paper/supplementary.tex
@@
 \usepackage[]{neurips_2019}
+\newtheorem{theorem}{Theorem}
 
@@
 
+\clearpage
+ \newpage
+ \begin{theorem}[B2B consistency - general case]
+
+     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full rank noise.
+     %
+     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimise
+     %
+     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $E\hat H = \hat H$
+\end{theorem}
+\begin{proof}
+
+ Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B, we have
+ \begin{align*}
+    \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
+                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
+                                                        &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
+     \label{eq:doublenorm}
+\end{align*}
+\begin{align*}
+    \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
+    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
+    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
+    &= EF \hat G
+\end{align*}
+
+Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero.
+
+ Let $Z=F^\dagger EF\hat G$.
+We have $FZ = FF^\dagger EF  \hat G= EF\hat G$ ($FF^\dagger E =E$ since $F$ is full rank on $Img(E)$).
+Since E is a contraction, we have $ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$.
+Therefore,
+ $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
+
+But as $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, the above inequality is an equality, $Z=\hat G$ and $EF\hat G = F\Hat G$.
+
+Therefore, $\left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 = \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 = \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2$ is minimised by $\hat G$ and $\hat H$.
+
+Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$. This completes the proof.
+\end{proof}
+
+So, $\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top$.
+
+Assuming, without loss of generality, that the active features are the $k$ first, we have
+
+$X^\top X = \left(\begin{array}{lccl}\Sigma_{X1} & \Sigma_{X2} \\ \Sigma_{X2} & \Sigma_{X3}\end{array}\right)$, $N^\top N = \left(\begin{array}{lccl}\Sigma_{N1} & \Sigma_{N2} \\ \Sigma_{N2} & \Sigma_{N3}\end{array}\right)$ and
+
+$\hat H = \left(\begin{array}{cc}(\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1} & (\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X2} \\0 & 0\end{array}\right)$
+
+Because $\hat H$ is built from covariance matrices of $X$ and $N$, all its diagonal elements are positive, with only the k first different from zero.
+Their average value is $\frac{1}{k} Tr(\hat H)$.
+When noise tends to zero, they approach one, and $Diag(\hat H) = E$.
+
+In any case, positive diagonal elements will be bounded by the maximum and minimum value of $\frac{\sigma_X}{\sigma_X + \sigma_N}$ over all eigenvalues of the covariance matrices of $XE$ and $NE$, that is, denoting $\sigma_{X1_1}$, $\sigma_{X1_k}$, $\sigma_{N1_1}$ and $\sigma_{N1_k}$ the largest and smallest eigenvalues of $\Sigma_{X1}$ and $\Sigma_{N1}$, all positive diagonal values of $\hat H$ are bounded by $\frac{\sigma_{X1_1}}{\sigma_{X1_1} +\sigma_{N1_k}}$ and $\frac{\sigma_{X1_k}}{\sigma_{X1_k} +\sigma_{N1_1}}$.
+
+On average, and since $X$ and $N$ are uncorrelated, the average value of positive diagonal terms of $\hat H$ should be close to $\frac{Var(X)}{Var(X)+Var(N)}$
+
 \end{document}

### 2019-05-24 paper/supplementary.tex
theorem 2

@@
 
-Because $\hat H$ is built from covariance matrices of $X$ and $N$, all its diagonal elements are positive, with only the k first different from zero.
-Their average value is $\frac{1}{k} Tr(\hat H)$.
-When noise tends to zero, they approach one, and $Diag(\hat H) = E$.
+$Diag_k (\hat H) = Diag((\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1}) = Diag((I+\Sigma_{X1}^{-1}\Sigma_{N1})^{-1})$
 
-In any case, positive diagonal elements will be bounded by the maximum and minimum value of $\frac{\sigma_X}{\sigma_X + \sigma_N}$ over all eigenvalues of the covariance matrices of $XE$ and $NE$, that is, denoting $\sigma_{X1_1}$, $\sigma_{X1_k}$, $\sigma_{N1_1}$ and $\sigma_{N1_k}$ the largest and smallest eigenvalues of $\Sigma_{X1}$ and $\Sigma_{N1}$, all positive diagonal values of $\hat H$ are bounded by $\frac{\sigma_{X1_1}}{\sigma_{X1_1} +\sigma_{N1_k}}$ and $\frac{\sigma_{X1_k}}{\sigma_{X1_k} +\sigma_{N1_1}}$.
+In the absence of noise, we have $Diag(\hat H) = Diag(E)$.
+Else the k first diagonal elements of $\hat H$ are all positive, and bounded by $\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}}$ and $\frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$, where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X1}$ and $\Sigma_{N1}$.
 
-On average, and since $X$ and $N$ are uncorrelated, the average value of positive diagonal terms of $\hat H$ should be close to $\frac{Var(X)}{Var(X)+Var(N)}$
+The average value of the coefficients of $Diag(\hat H)$ is the trace of $\hat H$ divided by k.
+On average, since $X$ and $N$ are not correlated, $\Sigma_{X1}^{-1}\Sigma_{N1}$ should have a low condition numbers, and $\frac{Var(X)}{Var(X)+Var(N)}$ provide a good estimate of $\frac{1}{k}Tr(\hat H)$.
 

### 2019-05-24 paper/supplementary.tex
Some typos

@@
 
+\appendix
+
 \maketitle
+
+\section{Additional Figures}
 
@@
 \newpage
- \begin{theorem}[B2B consistency - general case]
 
-     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full rank noise.
+\section{Additional Theorems}
+
+\setcounter{theorem}{1}
+ \begin{theorem}[B2B consistency, general case]
+
+     Consider the B2B model from Equation $Y = (XE + N)F$, with centered, full-rank noise $N$.
      %
-     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimise
+     If $F$ and $X$ are full-rank on $\text{Img}(E)$, then, the solution of B2B, $\hat H$, will minimise
      %
@@
                                                         &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
-                                                        &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
+                                                        &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2,
      \label{eq:doublenorm}
@@
     &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
-    &= EF \hat G
+    &= EF \hat G.
 \end{align*}
-
-Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero.
-
- Let $Z=F^\dagger EF\hat G$.
-We have $FZ = FF^\dagger EF  \hat G= EF\hat G$ ($FF^\dagger E =E$ since $F$ is full rank on $Img(E)$).
-Since E is a contraction, we have $ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$.
-Therefore,
- $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
+%
+First, we show that $EF\hat G = F\hat G$.
+%
+Let $Z=F^\dagger EF\hat G$. We have $FZ = FF^\dagger EF  \hat G= EF\hat G$.
+Then, $FF^\dagger E =E$ since $F$ is full rank on $\text{Img}(E)$.
+%
+Since $E$ is a contraction, we have $ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$. Therefore, 
+ $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2.$$
 
@@
 
-$Diag_k (\hat H) = Diag((\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1}) = Diag((I+\Sigma_{X1}^{-1}\Sigma_{N1})^{-1})$
+    $\text{Diag}_k (\hat H) = \text{Diag}((\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1}) = \text{Diag}((I+\Sigma_{X1}^{-1}\Sigma_{N1})^{-1})$
 
-In the absence of noise, we have $Diag(\hat H) = Diag(E)$.
+In the absence of noise, we have $\text{Diag}(\hat H) = \text{Diag}(E)$.
 Else the k first diagonal elements of $\hat H$ are all positive, and bounded by $\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}}$ and $\frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$, where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X1}$ and $\Sigma_{N1}$.
 
-The average value of the coefficients of $Diag(\hat H)$ is the trace of $\hat H$ divided by k.
+The average value of the coefficients of $\text{Diag}(\hat H)$ is the trace of $\hat H$ divided by k.
 On average, since $X$ and $N$ are not correlated, $\Sigma_{X1}^{-1}\Sigma_{N1}$ should have a low condition numbers, and $\frac{Var(X)}{Var(X)+Var(N)}$ provide a good estimate of $\frac{1}{k}Tr(\hat H)$.

### 2019-05-27 paper/paper.tex
Final version
final post extension

@@
 
-This section provides a consistency proof for B2B in the case of low noise.
+\begin{theorem}[B2B consistency - general case]
 
-\begin{theorem}[Low-noise consistency]
-    Consider the B2B model from Equation~\ref{eq:model}.
+     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full rank noise.
      %
-    Assume that $F$ and $X$ are full-rank on the subspace spanned by active entries in $E$.
+     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimise
      %
-    Then, as the variance of $N$ tends to zero and the sample size tends to infinity, B2B recovers $E$.
+     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $E\hat H = \hat H$
 \end{theorem}
 \begin{proof}
-For centered noise $N$, the first regression satisfies:
+
+ Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B, we have
  \begin{align*}
-    \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
-                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2].
+    \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
+%                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
+                                                        &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
+     \label{eq:doublenorm}
 \end{align*}
-%
-The solution to this first regression is $\hat{G} = [(XEF)^\top (XEF) + (NF)^\top (NF)]^{-1} (XEF)^\top X$.
+\begin{align*}
+    \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
+    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
+%    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
+    &= EF \hat G
+\end{align*}
 
-The second regression is:
-%
-\begin{align*}
-    \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
-    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
-    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]
-\end{align*}
-%
-The solution to this second regression is $H = EF\hat{G}$.
-%
-Therefore, our estimate of $E$ is:
-\begin{equation}
-    \hat{E} = EF[(XEF)^\top (XEF) + (NF)^\top (NF)]^{-1} (XEF)^\top X.
-    \label{eq:estimate}
-\end{equation}
-%
-By definition of the Moore-Penrose inverse (also known as pseudo-inverse), if $(NF)^\top (NF)$ is full-rank with $\|~(NF)^\top~(NF)~\|~\to~0$, it follows that $\hat{E} = EF(XEF)^\dagger X$.
-%
-Next, assume that the true causal influence matrix contains $k$ ones arranged as $E = \left(\begin{array}{cc} I_k & 0 \\ 0 & 0 \end{array}\right)$, and let $E = K^\top K$, where $K\in \mathbb{R}^{k \times d_x}$.
-%
-    Also, let the covariance matrix of $X$ be $X^\top X = \left(\begin{array}{cc}\Sigma_{1} & \Sigma_{2} \\ \Sigma_{3} & \Sigma_{4}\end{array}\right)$, where $\Sigma_1 \in \mathbb{R}^{k\times k}$.
-%
-Then,
-\begin{align*}
-    \hat{E} &= EF(XEF)^\dagger X = K^\top K F(XK^\top K F)^\dagger X \stackrel{i)}{=} K^\top (K F) (KF)^\dagger (XK^\top)^\dagger X\\
-            &\stackrel{ii)}{=} K^\top K(XK^\top)^\dagger X = E (XK^\top)^\dagger X \stackrel{iii)}{=} E (KX^\top X K^\top)^{-1} KX^\top X\\
-            &= E (KX^\top X K^\top)^{-1} (\Sigma_1 \,|\, \Sigma_2)  = E \Sigma_1^{-1} (\Sigma_1 \,|\, \Sigma_2)
-            = \left(\begin{array}{cc} I_k & \Sigma_1^{-1} \Sigma_2 \\ 0 & 0 \end{array}\right),
-\end{align*}
-where the equalities follow due to i) a full-rank factorization of $XEF$, ii) $KF$ being full-rank yields $(KF)(KF)^\dagger = K$, and iii) $X K^\top$ being full-rank yields $(XK^\top)^\dagger = (KX^\top X K^\top)^{-1} KX^\top$.
-%
-Therefore, $\text{Diag}(\hat{E})$ is the desired causal influence matrix $E$.
+Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero.
+Let $Z=F^\dagger EF\hat G$.
+We have $FZ = FF^\dagger EF  \hat G= EF\hat G$ ($FF^\dagger E =E$ since $F$ is full rank on $Img(E)$).
+Since E is a contraction, we have $ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$.
+Therefore,
+ $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
+
+But as $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, the above inequality is an equality, $Z=\hat G$ and $EF\hat G = F\Hat G$.
+Therefore, $\left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 = \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 = \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2$ is minimised by $\hat G$ and $\hat H$.
+Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$.
+This completes the proof.
 \end{proof}
 
-We observe that, even in the presence of strong noise, the expression of our estimate \eqref{eq:estimate} involves the left-most multiplication with the true causal influence matrix $E$.
-%
-This means that our estimate will, in expectation, contain zeros for all features that do not causally influence $Y$.
-%
-When the variance of $N$ increases, we of course pay a price: since the norm $\| X - XEF\hat{G} \| = \| X - X\hat{H} \|$ increases, this means that the values of $\hat{E}$ associated to the causal factors will be dampened.
+So, $\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top$.
+
+Assuming, without loss of generality, that the active features are the $k$ first, we have
+
+$X^\top X = \left(\begin{array}{lccl}\Sigma_{X1} & \Sigma_{X2} \\ \Sigma_{X2} & \Sigma_{X3}\end{array}\right)$, $N^\top N = \left(\begin{array}{lccl}\Sigma_{N1} & \Sigma_{N2} \\ \Sigma_{N2} & \Sigma_{N3}\end{array}\right)$ and
+
+$\hat H = \left(\begin{array}{cc}(\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1} & (\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X2} \\0 & 0\end{array}\right)$
+
+$Diag_k (\hat H) = Diag((\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1}) = Diag((I+\Sigma_{X1}^{-1}\Sigma_{N1})^{-1})$
+
+In the absence of noise, we have $Diag(\hat H) = Diag(E)$.
+Else the k first diagonal elements of $\hat H$ are all positive, and bounded by $\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}}$ and $\frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$, where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X1}$ and $\Sigma_{N1}$.
+
+The average value of the coefficients of $Diag(\hat H)$ is the trace of $\hat H$ divided by k.
+On average, since $X$ and $N$ are not correlated, $\Sigma_{X1}^{-1}\Sigma_{N1}$ should have a low condition numbers, and $\frac{Var(X)}{Var(X)+Var(N)}$ provide a good estimate of $\frac{1}{k}Tr(\hat H)$.
 
@@
 %
-We use scikit-learn to implement all of our simulations \citep{sklearn}.
+We use scikit-learn \citep{sklearn}.
 

### 2019-05-27 paper/supplementary.tex
final post extension

--- 2019-05-24 paper/supplementary.tex
@@
 
-\clearpage
- \newpage
- \begin{theorem}[B2B consistency - general case]
-
-     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full rank noise.
-     %
-     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimise
-     %
-     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $E\hat H = \hat H$
-\end{theorem}
-\begin{proof}
-
- Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B, we have
- \begin{align*}
-    \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
-                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
-                                                        &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
-     \label{eq:doublenorm}
-\end{align*}
-\begin{align*}
-    \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
-    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
-    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
-    &= EF \hat G
-\end{align*}
-
-Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero.
-
- Let $Z=F^\dagger EF\hat G$.
-We have $FZ = FF^\dagger EF  \hat G= EF\hat G$ ($FF^\dagger E =E$ since $F$ is full rank on $Img(E)$).
-Since E is a contraction, we have $ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$.
-Therefore,
- $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
-
-But as $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, the above inequality is an equality, $Z=\hat G$ and $EF\hat G = F\Hat G$.
-
-Therefore, $\left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 = \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 = \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2$ is minimised by $\hat G$ and $\hat H$.
-
-Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$. This completes the proof.
-\end{proof}
-
-So, $\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top$.
-
-Assuming, without loss of generality, that the active features are the $k$ first, we have
-
-$X^\top X = \left(\begin{array}{lccl}\Sigma_{X1} & \Sigma_{X2} \\ \Sigma_{X2} & \Sigma_{X3}\end{array}\right)$, $N^\top N = \left(\begin{array}{lccl}\Sigma_{N1} & \Sigma_{N2} \\ \Sigma_{N2} & \Sigma_{N3}\end{array}\right)$ and
-
-$\hat H = \left(\begin{array}{cc}(\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1} & (\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X2} \\0 & 0\end{array}\right)$
-
-$Diag_k (\hat H) = Diag((\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1}) = Diag((I+\Sigma_{X1}^{-1}\Sigma_{N1})^{-1})$
-
-In the absence of noise, we have $Diag(\hat H) = Diag(E)$.
-Else the k first diagonal elements of $\hat H$ are all positive, and bounded by $\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}}$ and $\frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$, where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X1}$ and $\Sigma_{N1}$.
-
-The average value of the coefficients of $Diag(\hat H)$ is the trace of $\hat H$ divided by k.
-On average, since $X$ and $N$ are not correlated, $\Sigma_{X1}^{-1}\Sigma_{N1}$ should have a low condition numbers, and $\frac{Var(X)}{Var(X)+Var(N)}$ provide a good estimate of $\frac{1}{k}Tr(\hat H)$.
-
 \end{document}

### 2019-07-26 paper/paper.tex
typos in paper, ref in bib

@@
 %
-Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{sonquist_morgan} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
+Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{sonquist_morgan} \citep{kass75} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
 %

### 2019-07-26 paper/synthetic.tex
typos in paper, ref in bib

@@
     \caption{Results of synthetic experiments.
-Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration averaged over $20$ random seeds.
+Each dot depicts the average value of a metric for B2B ($x$-axis) against a competitor ($y$-axis) for each experiment configuration averaged over $20$ random seeds.
 The prediction error on $Y$ within (in) and outside (out) of the training distribution are weakly but consistently larger for all of the models as compared to B2B with and without the Sonquist binarization (E).
@@
 
-Table~\ref{table:synthetic} summarizes the results for all experimental conditions, metrics, and methods.
+Figure~\ref{table:synthetic} summarizes the results for all experimental conditions, metrics, and methods.
 %

### 2019-07-28 paper/supplementary.tex
ending document
theorem, long proof

@@
 
+\appendix
+
 \maketitle
+
+\section{Theorem - detailed proof}
+\label{sec:theorem}
+
+\begin{theorem}[B2B consistency - general case]
+
+     Consider the B2B model from Equation $$Y = (XE + N)F$$ $N$ centred and full rank noise.
+
+     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimise
+
+     $$\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$$ and satisfy
+
+     $$E\hat H = \hat H$$
+\end{theorem}
+\begin{proof}
+
+ Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B.
+
+ Since $\hat G$ is the least square estimator of $X$ from $Y$
+ \begin{align*}
+    \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] 
+\end{align*}
+Replacing $Y$ by its model definition $Y = (XE+N)F$, we have
+ \begin{align*}    
+    \hat G &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2] =\arg \min_G \mathbb{E}[\left \| X - XEFG + NFG \right\|^2]
+  \end{align*}
+  Since $N$ is centered and independent of $X$, we have 
+  \begin{align}  
+    	  \hat G &=  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
+     \label{eq:Gdoublenorm}
+\end{align}
+
+Samely, for $\hat H$, we have
+\begin{align*}
+    \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
+    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
+    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]
+ \end{align*}
+ a positive quantity which reaches a minimum (zero) for 
+ \begin{align}
+    \hat H = EF \hat G
+    \label{eq:Hdoublenom}
+\end{align}
+
+Let us now prove that $EF\hat G = F\hat G$.
+
+Let $F^\dagger$ be the pseudo inverse of $F$, and $Z=F^\dagger EF\hat G$, we have $FZ = FF^\dagger EF \hat G$
+
+Since $F$ is full rank on $Img(E)$, we have $FF^\dagger E =E$, and $FZ = EF\hat G$
+
+As $E$ is a binary diagonal matrix, it is an orthogonal projection and therefore a contraction, thus
+ $$ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$$ and
+ $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
+
+But since $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, we also have
+$$\left \| X - XEF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 \leq \left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2$$
+
+Summarizing the above,
+$$\left \| X - XEF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
+$$\left \| X - XEF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2$$
+$$\left \| NF\hat G\right \| ^2 =  \| NEF\hat G \| ^2$$
+
+$N$ being full rank, this yields $EF\hat G = F\Hat G$.
+
+Replacing into $\eqref{eq:Gdoublenorm}$, and setting $H = EFG$, we have
+\begin{align*}
+	\hat G &=  \arg \min_G  \left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 \\
+	&=   \arg \min_G \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 \\
+	\hat H &=  \arg \min_H \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2
+	\label{eq:4}	
+\end{align*}
+
+Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$, since $E$, a binary diagonal matrix, is involutive. This completes the proof.
+\end{proof}
+
+\section{Additional Figures}
 

### 2019-09-24 code/meg_experiment/base.py
iclr meg and synthetic update

@@ -3,7 +3,10 @@ import numpy as np
 import os.path as op
 from scipy.linalg import svd
 from scipy.stats import pearsonr
+
+from sklearn.base import clone
 from sklearn.preprocessing import scale
+from sklearn.model_selection import ShuffleSplit
 from common import (add_part_of_speech, add_word_frequency, add_word_length,
                     read_log, get_log_times, data_path)
 
@@ -43,20 +46,27 @@ def fetch_data(log_file):
     df = add_word_frequency(epochs.metadata[['word', ]])
     df = add_word_length(df)
     df = add_part_of_speech(df)
+    df['word_func'] = False
+    df.loc[df.query('(ADP+CONJ+DET+PRON)==True').index,
+           'word_func'] = True
     word = df.word.apply(lambda s: s.lower())
     for letter in 'abcdefghijklmnopqrstuvwxyz':
         df[letter] = word.str.count(letter) > 1
+    
+    df['word_dummy'] = (scale(df['word_length']) + 
+                        scale(df['word_freq']) + 
+                        np.random.randn(len(df)))
 
     meg = epochs.get_data()
     n_samples, n_chans, n_times = meg.shape
     for t in range(n_times):
         meg[:, :, t] = scale(meg[:, :, t])
 
-    columns = ['word_freq', 'word_length']
-    names = [k for k in df.keys() if k not in ('word', 'pos')]
+    names = ['word_length', 'word_freq', 'word_func', 'word_dummy']
+
     features = scale(df[names].values)
 
-    return meg, epochs.times, features, names
+    return meg, epochs.times, features, names, df.word
 
 
 def ridge_cv(X, Y, alphas, independent_alphas=False, Uv=None):
@@ -110,14 +120,13 @@ def ridge_cv(X, Y, alphas, independent_alphas=False, Uv=None):
     return coefs, best_alphas, Y_hat
 
 
-def fit_knockout(X, Y, alphas, knockout=True, independent_alphas=False,
+def fit_knockout(X, Y, alphas, independent_alphas=False,
                  pairwise=False):
     """Fit model using all but knockout X features"""
     n_x, n_y = X.shape[1], Y.shape[1]
-    knockout = np.arange(n_x)[:, None] if knockout is True else knockout
 
     K = list()
-    for xi in knockout:
+    for xi in np.arange(n_x)[:, None]:
         not_xi = np.setdiff1d(np.arange(n_x), xi)
 
[…]

### 2019-09-24 code/synthetic_experiment/rebuttal/base.py
rebuttal synthetic WIP

@@ -1,0 +1,305 @@
+import mne
+import numpy as np
+import os.path as op
+from scipy.linalg import svd
+from scipy.stats import pearsonr
+
+from sklearn.base import clone
+from sklearn.preprocessing import scale
+from sklearn.model_selection import ShuffleSplit
+
+
+def ridge_cv(X, Y, alphas, independent_alphas=False, Uv=None):
+    """ Similar to sklearn RidgeCV but
+   (1) can optimize a different alpha for each column of Y
+   (2) return leave-one-out Y_hat
+   """
+    if isinstance(alphas, (float, int)):
+        alphas = np.array([alphas, ], np.float64)
+    if Y.ndim == 1:
+        Y = Y[:, None]
+    n, n_x = X.shape
+    n, n_y = Y.shape
+    # Decompose X
+    if Uv is None:
+        U, s, _ = svd(X, full_matrices=0)
+        v = s**2
+    else:
+        U, v = Uv
+    UY = U.T @ Y
+
+    # For each alpha, solve leave-one-out error coefs
+    cv_duals = np.zeros((len(alphas), n, n_y))
+    cv_errors = np.zeros((len(alphas), n, n_y))
+    for alpha_idx, alpha in enumerate(alphas):
+        # Solve
+        w = ((v + alpha) ** -1) - alpha ** -1
+        c = U @ np.diag(w) @ UY + alpha ** -1 * Y
+        cv_duals[alpha_idx] = c
+
+        # compute diagonal of the matrix: dot(Q, dot(diag(v_prime), Q^T))
+        G_diag = (w * U ** 2).sum(axis=-1) + alpha ** -1
+        error = c / G_diag[:, np.newaxis]
+        cv_errors[alpha_idx] = error
+
+    # identify best alpha for each column of Y independently
+    if independent_alphas:
+        best_alphas = (cv_errors ** 2).mean(axis=1).argmin(axis=0)
+        duals = np.transpose([cv_duals[b, :, i]
+                              for i, b in enumerate(best_alphas)])
+        cv_errors = np.transpose([cv_errors[b, :, i]
+                                  for i, b in enumerate(best_alphas)])
+    else:
+        _cv_errors = cv_errors.reshape(len(alphas), -1)
+        best_alphas = (_cv_errors ** 2).mean(axis=1).argmin(axis=0)
+        duals = cv_duals[best_alphas]
+        cv_errors = cv_errors[best_alphas]
+
[…]

### 2019-09-24 paper/introduction.tex
wip paper

@@
 In neuroscience, for example, identifying whether the frequency of a word presented on a subject's retina modulates brain activity can be surprisingly difficult.
-Indeed, word frequency covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs), their individual letters (e.g.
-words that contain `z' are rare) and so forth \citep{kutas2011thirty,pegado2014timing}.
+Indeed, the frequency of words in natural language covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs) and so forth \citep{kutas2011thirty,pegado2014timing}.
 Instead of selecting a set of words that control for all of these factors simultaneously, it is thus common to use forward modeling, i.e.
-to train a mathematical model to predict observations (e.g.
+to train a model to predict observations (e.g.
 brain activity) from a minimal combination of competing factors (e.g.
@@
 For example, brain activity is often recorded with hundreds or thousands of sensors simultaneously.
-It is thus common to use `decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving,  kriegeskorte2008representational, norman2006beyond}.
+As multiple sensors may be affected by common noise sources, it is common to use `decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving,  kriegeskorte2008representational, norman2006beyond}.
 
@@
 By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}.
-There also exist cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA), but their resulting coefficients are notoriously difficult to interpret \citep{lebart1995statistique}.
+To combine the benefits of forward and backward modeling, several authors have proposed to use cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA) \citep{dechevigne2018CCA}.
+CCA and PLS aim to find, from two sets of data $X$ and $Y$, the components $v$ and $w$ were $Xw$ and $Yv$ are maximially correlated or maximally covarying respectively.
+Because CCA and PLS are are based on a generalized eigen decomposition, their resulting coefficients are mixing the features of $X$ and $Y$ in a way that makes them notoriously difficult to interpret \citep{lebart1995statistique}.
 
-We introduce the `back-to-back regression', which combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}).
-After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating i) causal influence and ii) out-of-environment predictions (Section~\ref{sec:experiment_synthetic}).
+Here, we introduce the `back-to-back regression', which not only combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}), but also provide a fast, robust and interpretable causal coefficients.
+After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques in identifying causal influence (Section~\ref{sec:experiment_synthetic}).
 Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features lead to distinguishable brain responses (Section~\ref{sec:experiment_real}).

### 2019-09-24 paper/meg_exps.tex
wip paper

@@
 
-\subsubsection{MEG preprocessing and feature generation}
+\subsubsection{MEG preprocessing}
 
@@
 segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
-thus led a sampling frequency of 240Hz. Figure 1 in the supplementary material
-shows the average response of the magnetometers evoked by the words presented
-to the first subject. For each subject and each time sample relative to word onset, we
+thus led a sampling frequency of 240Hz.
+For each subject and each time sample relative to word onset, we
 build an observation matrix $Y \in \mathbb{R}^{n \times d_y}$ of $n\approx$ 2700 words
@@
 
-We annotated each word with 38 features: 26
-correspond to the count of each letter in the word (from 'a' to 'z',
-independently of their case), the total number of letters ('word length'), the
-frequency of the word in natural language (using the Zipf logarithmic scale
-of \citep{van2014subtlex}), and the part-of-speech tags (i.e. NOUN, PROPN (proper
-noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ
-(adjective), ADV (adverb), and NUM (numerical)) using Spacy \citep{spacy2} and
-WordFreq \citep{speerwordfreq}. For clarity purposes, we refer to these
-features using two broad categories: visual (i.e. each letter and the total
-number of letters) and lexical (part-of-speech and word frequency). This
-procedure yields an $X \in \mathbb{R}^{n \times d_x}$ matrix of $n\approx$ 2700 words by
-$d_x=38$ features for each subject. Each of the columns of $X$ is normalized to
+\subsubsection{Feature definition}
+We annotated each word with three collinear features.
+%
+"Word Length' refers to the total number of letters, and is expected to specifically cause a variation in the early evoked MEG responses (i.e.
+from 100 ms after stimulus onset), because it variably engages different retinotopically-tuned visual areas in the primary visual cortex.
+%
+'Word Frequency', as derived with WordFreq \citep{speerwordfreq}, indexes how frequently each word appears in Dutch (using the Zipf logarithmic scale of \citep{van2014subtlex}).
+This feature is expected to specifically cause a variation in the late evoked MEG responses (i.e.
+from 400 ms), because it variably engages semantic processes in the temporal cortices \citep{kutas2011thirty}.
+%
+Finally, 'Word Function' indicates whether each word is a content word (i.e.
+a noun, a verb, an adjective or an adverb) or a function word (i.e.
+a preposition, a conjunction, a determinant, a pronoun or a numeral), and was derived from Spacy's part of speech tagger \citep{spacy2}.
+To our knowledge, this feature has not been thouroughly investigated with MEG.
+Its causal contribution to reading processes in the brain is therefore largely unknown.
+%
+To verify that B2B and other models would not inadequately identify non-causal features, we added a dummy feature, constructed from a noisy combination of "Word Length" and "Word Frequency":
+\begin{equation}
+  dummy = z(length) + z(frequency) + z(\mathcal{N})
+\end{equation}
+where $z$ transforms features into zero-mean and unit variance vectors.
+
+This procedure yields an $X \in \mathbb{R}^{n \times d_x}$ matrix of $n\approx$ 2700 words by
+$d_x=4$ features for each subject. Each of the columns of $X$ is normalized to
 have a mean and a standard deviation of 0 and 1 respectively.
@@
 
-Our goal is to identify which word features ($X$) lead to a reliable modulation of brain activity as measured with MEG recordings ($Y$) at each time sample and for each subject separately.
-To this aim,
-we compare three methods: forward regression, backward regression, and back-to-back regression.
+We compare B2B to four standard models: forward regression, backward regression, CCA and PLS, as implemented in scikit-learn, and optimized with nested cross validation over 20 $\lambda$ l2 regularization parameters logarithmically spaced between $10^-4$ and $10^4$ or 1 to 4 caconical components.
+
+We used the feature importance described in XXX algorithms XXX to assess the extent to which each feature $X_i$ specifically improves the prediction of held-out $Y$ data.
+
 Corresponding results can be found in Figure \ref{fig:meg_twocurves}.
@@
 
-Forward regression was performed with an $\ell_2$-regularized least-squares regression $H \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| H_{i, :} \|^2$.
-Backward regression was performed with an $\ell_2$-regularized least-squares regression $G \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YG$, using a validation sample.
+%Forward regression was performed with an $\ell_2$-regularized least-squares regression $H \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| H_{i, :} \|^2$.
+%Backward regression was performed with an $\ell_2$-regularized least-squares regression $G \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YG$, using a validation sample.
 Finally, we used the back-to-back regression was performed using two $\ell_2$-regularized
-least squares regressions $G$ and $H$ as described in section \ref{sec:algorithm}, using a bagging factor of 10, and alpha cross-validated between $10^{-5}$ and $10^{5}$.
+%least squares regressions $G$ and $H$ as described in section \ref{sec:algorithm}, using a bagging factor of 10, and alpha cross-validated between $10^{-5}$ and $10^{5}$.
 
-The three types of analyses were implemented for each subject and each time
-sample independently. To estimate the reliability of these estimates in the
-population, we tested, using a one-sample $t$-test whether the distribution of
-the estimated contribution of each factor differed from baseline ($t=0$ms),
-as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau$ ranging from
-150ms to 1s after word onset. We use the resulting t and p-values as proxy to
-quantify the performance of each model. These results are depicted in Figure
-3 of the Supplementary Material. For concision purposes, we only report this metric averaged
-between 150ms and 1s after word onset.
+%The three types of analyses were implemented for each subject and each time
+%sample independently. To estimate the reliability of these estimates in the
+%population, we tested, using a one-sample $t$-test whether the distribution of
+%the estimated contribution of each factor differed from baseline ($t=0$ms),
+%as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau$ ranging from
+%150ms to 1s after word onset. We use the resulting t and p-values as proxy to
+%quantify the performance of each model. These results are depicted in Figure
+%3 of the Supplementary Material. For concision purposes, we only report this metric averaged
+%between 150ms and 1s after word onset.
 

### 2019-09-24 paper/paper.tex
ICLR format.
wip paper

--- 2019-05-27 paper/paper.tex
@@
 \usepackage[ruled, linesnumbered]{algorithm2e}
-\usepackage[]{neurips_2019}
+\usepackage{iclr2020_conference,times}
 
@@
   CNRS\\
-  \texttt{email} \\
+  \texttt{jeanremi.king@gmail.com} \\
   \And
@@
 \begin{abstract}
-Identifying causes from observations is at the core of science. This endeavor
-is particularly challenging when i) potential factors are difficult to
-manipulate individually and ii) observations are complex and multi-dimensional.
+Identifying causes from observations can be particularly challenging when i) potential factors are difficult to manipulate individually and ii) observations are complex and multi-dimensional.
 To address this issue, we introduce ``Back-to-Back'' regression (B2B), a
-method designed to measure, from a set of co-varying factors, the causal
+method designed to efficiently measure, from a set of co-varying factors, the causal
 influences that most plausibly account for multidimensional observations. After
-proving the consistency of B2B, we show that our method outperforms
-least-squares and related regression and cross-decomposition techniques (e.g.
-canonical correlation analysis and partial least squares) on two tasks: causal
-identification and out-of-distribution prediction. Finally, we apply B2B to
+proving the consistency of B2B and its links to other linear approaches, we show that our method outperforms
+least-squares regression and cross-decomposition techniques (e.g.
+canonical correlation analysis and partial least squares) on causal
+identification. Finally, we apply B2B to
 neuroimaging recordings of 102 subjects reading word sequences. The results
 show that the early and late brain responses caused by low- and high-level
-word features respectively are more easily detected with B2B than with standard forward regression.
+word features respectively are more reliably detected with B2B than with  other standard techniques.
 % conclude  open
@@
 
-\section{Back to back regression}
+\section{Back-to-Back regression}
 \label{sec:algorithm}
@@
 While the triplet of variables $X$ and $N$ are independent, we allow each of them to have a general covariance matrix.
+% what is a general covariance matrix?
 %
@@
 %
-To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
+To avoid this issue, we split our sample $(X, Y)$ at random into two subsets $(X_1, Y_1)$ and $(X_2, Y_2)$.
 %
@@
 %
-After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$.
+To accelerate this ensembling procedure, we implemented an efficient leave-one-out cross-validation scheme as detailed in \citep{RifkinLippert2007} % http://cbcl.mit.edu/publications/ps/MIT-CSAIL-TR-2007-025.pdf
+, as follow:
+%
+
+\begin{equation}
+\hat{Y} = (\Sigma_X G Y - diag(\Sigma_X G) Y) / diag(I - \Sigma_X G)
+\end{equation}
+
+where $\Sigma_X$ is the $X$ kernel matrix and where $G$ is computed with an eigen decomposition of $X$:
+\begin{equation}
+  \begin{aligned}
+  \Sigma_X = Q V Q^T \\
+  G = Q (V + \alpha*I)^{-1} Q^T\\
+  \end{aligned}
+\end{equation}
+where $Q$, $V$ and $\alpha$ are the eigen vectors, eigen values and regularization, respectively.
+
+%
+% After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$.
 
@@
 
+% JR: shall we update algorithm with LOO trick?
 \begin{algorithm}[H]
@@
      %
-     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimise
+     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimize
      %
@@
 The average value of the coefficients of $Diag(\hat H)$ is the trace of $\hat H$ divided by k.
-On average, since $X$ and $N$ are not correlated, $\Sigma_{X1}^{-1}\Sigma_{N1}$ should have a low condition numbers, and $\frac{Var(X)}{Var(X)+Var(N)}$ provide a good estimate of $\frac{1}{k}Tr(\hat H)$.
-
-\subsection{Binarizing $\hat{E}$}
-
-On the one hand, we have leveraged B2B to obtain the estimate $\hat{E}$, which is a diagonal matrix with real entries.
-%
-On the other hand, the true causal influence matrix $E$ is a binary matrix, which hard-selects the causal factors of $Y$ from $X$.
-%
-In this section, we provide one of the many possible recipes to binarize $\hat{E}$ and estimate the collection of causal factors.
-
-In regression analysis, the traditional approach to this problem employs a $t$-test to check whether the regression coefficients differ from zero \citep{student1908probable}.
-%
-However, this test will not succeed here, since the estimate $\hat{E}$ is obtained from a double regression, and any employed regularization will add a bias to the diagonal of $\hat{E}$.
-%
-Instead, we treat the binarization of $\hat{E}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''.
-%
-More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
-%
-Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{sonquist_morgan} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
-%
-To binarize $\hat{E}$, set to one all the $p$ diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the $d_x-p$ diagonal entries.
+On average, since $X$ and $N$ are not correlated by definition, $\Sigma_{X1}^{-1}\Sigma_{N1}$ should have a low condition numbers, and $\frac{Var(X)}{Var(X)+Var(N)}$ provide a good estimate of $\frac{1}{k}Tr(\hat H)$.
 
@@
 
-We perform two experiments to evaluate the performance of B2B: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography dataset.
+We perform two experiments to evaluate B2B: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography dataset.
 %
-We use scikit-learn \citep{sklearn}.
+We use scikit-learn's CCA, PLS, and RidgeCV objects to compare B2B against the standard baselines \citep{sklearn}.
 
@@
 In the context of neuroimaging, the first regression of B2B relates to the first step of Representational Similarity Analysis (RSA) \citep{kriegeskorte2008representational}.
-Indeed, RSA analyzes the similarity of brain responses associated with specific conditions (e.g.
+Indeed, RSA analyzes the similarity of brain responses associated with specific categorical conditions (e.g.
 distinct images), by (1) fitting one-against-all classifiers on each condition and (2) testing whether these classifiers can discriminate all other conditions.
@@
 
+% THIS NEEDS TO BE UPDATED: see de chevigner's latest paper
 More generally, B2B is related to CCA \citep{cca_hotelling}, but departs from it in four main aspects.
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{iclr2020_conference}
 \bibliography{paper}

### 2019-09-24 paper/synthetic.tex
wip paper

@@
 
-We compare the performance of B2B against five competing methods:
+We compare the performance of B2B against four competing methods, all implemented in scikit-learn \citep{scikit}:
 %
+% To be updated
+
+\subsection{Baseline models}
+
+Forward regression consists of an $l2$-regularized "ridge" regression from the putative causes $X$ to the observations $Y$:
+\begin{equation}
+  H = (X^T X +\lambda I)^{-1} X^T Y
+\end{equation}
+
+Backward regression consists of an $l2$-regularized "ridge" regression from $Y$ to $X$:
+\begin{equation}
+  G = (Y^T Y +\lambda I)^{-1} Y^T X
+\end{equation}
+
+CCA finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$ such that $X$ and $Y$ are maximally correlated in a latent $Z$ space:
+\begin{equation}
+  max corr(XH^T, YG^T)
+\end{equation}
+
+% To be checked
+PLS finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$ such that $X$ and $Y$ are maximally covarying in a latent $Z$ space:
+\begin{equation}
+  max cov(XH^T, YG^T)
+\end{equation}
+
+We employ $5$-fold cross-validation to select the optimal number of components for CCA and PLS.
+Regressions were l2-regularized with a $\lambda$ regularization parameters fitted with the efficient leave-one-out procedure implemented in scikit-learn RidgeCV \citep{scikit}.
+
+\subsection{Evaluating Causal Discovery from models' coefficients}
+
+B2B leads to unbiased (i.e.
+zeros-centered) scalar coefficients for non-causal features.
+In contrast, the Forward, Backward, CCA and PLS models lead to a loading vector $H_i$ per feature $i$ (or one vector $G^i$ for the backward model).
+To transform such vector into an estimated causal contribution $\hat E$, we take the sum of square coefficients:
+
+\begin{equation}
+  \hat E_i = \sum_j {H^j_i}^2
+\end{equation}
+
+To estimate whether models accurately identify causal factors, we compute the area-under-the-curve (AUC) across factors $AUC(E, \hat E)$.
+%\begin{equation}
+%  AUC = 1 - \sum_1^n (E_k - E_{k-1}) ( \hat{E}_k + \hat{E}_{k-1}) / 2
+%\end{equation}
+
+\subsection{Evaluating Causal Discovery with held-out prediction reliability}
+
+In most cases, $E$ is not known and AUC can thus not be estimated.
+
+To address this issue, we assess the ability of each model to reliably predict independent and identically distributed data from $Y$, given all of the $X$ features versus all but one feature: $X_k$ (i.e.
+'knock-out X').
+This procedure results in two correlation metrics $R_{full}$ and $R_{knockout}$, whose difference $\Delta R=R_{full}-R_{knockout}$ indicates how much $X_i$ improves the prediction of $Y$ above and beyond what can be predicted from all other features $X_k$.
+
+For, the Forward Model, this feature importance is assessed as follow:
+
+\begin{algorithm}[H]
+    %\SetAlgoLined
+    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$,
+          $X_{test} \in \mathbb{R}^{n' \times d_x}$,
+          $Y_{train} \in \mathbb{R}^{n\times d_y}$,
+          $Y_{test} \in \mathbb{R}^{n'\times d_y}$,
+          }
+    \KwOut{estimate of prediction improvement $\Delta{R} \in \mathbb{D}^{d_x, d_y}$.}
+    $H = \text{LinearRegression}(X_{train}, Y_{train})$\;
+    $R_{full} = \text{corr}(X_{test} H, Y_{test})$\;
+
+    \For{$i = 1, \ldots, d_x$}{
+        $K = Id$\;
+        $K[i] \leftarrow 0$\;
+        $H_k = \text{LinearRegression}(X_{train} K, Y_{train})$\;
+        $R_{k} = \text{corr}(X_{test} {H_k}, Y_{test})$\;
+        $\Delta R_i = R_{full} - R_{k}$\;
+    }
+    \Return{$\Delta R$}
+    \caption{Forward feature importance.}
+    \label{algorithm:fwd_fi}
+\end{algorithm}
+
+\iffalse
+\begin{enumerate}
+  \item Fit $H$ given $X_{train}$ and $Y_{train}$
+  \item For each feature $i$:
+  \begin{itemize}
+    \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
+    \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train}$
+  \end{itemize}
+\end{enumerate}
+\fi
+
+\begin{algorithm}[H]
+    %\SetAlgoLined
+    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$,
+          $X_{test} \in \mathbb{R}^{n' \times d_x}$,
+          $Y_{train} \in \mathbb{R}^{n\times d_y}$,
+          $Y_{test} \in \mathbb{R}^{n'\times d_y}$,
+          }
+    \KwOut{estimate of prediction improvement $\Delta{R} \in \mathbb{D}^{d_x, d_z}$.}
+    $H, G = \text{CCA}(X_{train}, Y_{train})$\;
+    $R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;
+
+    \For{$i = 1, \ldots, d_x$}{
+        $K = Id$\;
+        $K[i] \leftarrow 0$\;
+        $H_k, G_k = \text{CCA}(X_{train} K, Y_{train})$\;
+        $R_{k} = \text{corr}(X_{test} H_k, Y_{test} G_k)$\;
+        $\Delta R_i = R_{full} - R_{k}$\;
+    }
+    \Return{$\Delta R$}
+    \caption{CCA and PLS feature importance.}
+    \label{algorithm:cdp_fi}
+\end{algorithm}
+
+\iffalse
+For CCA and PLS, feature importance is assessed as follow:
+\begin{enumerate}
+  \item Fit $H$ and $G$ given $X_{train}$ and $Y_{train}$
+  \item For each feature $i$:
+  \begin{itemize}
+    \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
+    \item Fit $H^i_k$ and $G_k$ given $X_{train} K$ and $Y_{train}$
+  \end{itemize}
+  \end{enumerate}
+\fi
+
+For the Backward Model, feature importance cannot be assessed. We thus simply report the model performance $R_i = corr(YG_i, X_i)$
+
+For, B2B feature importance is assessed as follow:
+
+\begin{algorithm}[H]
+    %\SetAlgoLined
+    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$,
+          $X_{test} \in \mathbb{R}^{n' \times d_x}$,
+          $Y_{train} \in \mathbb{R}^{n\times d_y}$,
+          $Y_{test} \in \mathbb{R}^{n'\times d_y}$,
+          }
+    \KwOut{estimate of prediction improvement $\Delta{R} \in \mathbb{D}^{d_x}$.}
+    $H, G = \text{B2B}(X_{train}, Y_{train})$\;
+    $R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;
+
+    \For{$i = 1, \ldots, d_x$}{
+        $K = Id$\;
+        $K[i] \leftarrow 0$\;
+        $H_k = \text{LinearRegression}(X_{train} K, Y_{train} G_i)$\;
+        $R_{k} = \text{corr}(X_{test} H_k, Y_{test} G_i)$\;
+        $\Delta R_i = R_{full} - R_{k}$\;
+    }
+    \Return{$\Delta R$}
+    \caption{B2B feature importance.}
+    \label{algorithm:cdp_fi}
+\end{algorithm}
+
+\iffalse
+\begin{enumerate}
+
+  \item Fit $G$ given $X_{train}$ and $Y_{train}$
+  \item Fit $H$ given $X_{train}$ and $Y_{train} G^T$
+  \item For each feature $i$:
+  \begin{itemize}
+    \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
+    \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train} G^T$
+  \end{itemize}
+
+  \item Compute $R_{full}$ given $Y_{test} G^T$ and $X_{test}$
+  \item For all $i \in d_x$:
+  \begin{itemize}
+     \item Compute $R^i_{knockout}$ given $Y_{test} G^T$ and $X_{test} K {H^i_k}^T$
+     \item Compute $R^i_{delta}=R_{full} - R^i_{knockout}$
+   \end{itemize}
+\end{enumerate}
+\fi
+
+\iffalse
 Ridge regression \citep{hoerl1959optimum}, multitask Lasso \citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls, tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and Reduced Rank Regression or RRR \citep{Izenman_rrr}.
@@
 %
-We employ $5$-fold cross-validation to select the optimal hyper-parameters of each method.
 
@@
 \end{figure}
+\fi
 

### 2019-09-24 paper/synthetic.tex
no changes, just line length set to 80

--- 2019-07-26 paper/synthetic.tex
@@
-We evaluate the performance of B2B throughout a series of experiments on controlled synthetic data.
+We evaluate the performance of B2B throughout a series of experiments on
+controlled synthetic data.
 %
-The purpose of these experiments is to evaluate the ability of B2B in terms of prediction (inside and outside the training distribution), as well as a method to recover causal factors.
+The purpose of these experiments is to evaluate the ability of B2B in terms of
+prediction (inside and outside the training distribution), as well as a method
+to recover causal factors.
 
-The data generating process for each experiment constructs $n$ training examples according to the model $Y = (\text{snr} \cdot XE + N)F$, where $\text{snr}$ is a scalar that modulates the signal-to-noise ratio.
+The data generating process for each experiment constructs $n$ training examples
+according to the model $Y = (\text{snr} \cdot XE + N)F$, where $\text{snr}$ is a
+scalar that modulates the signal-to-noise ratio.
 %
 Here,
-%\begin{itemize}
-%    \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
-%    \item $X \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_X)$,
-%    \item $N \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$,
-%    \item $E \in \mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
-%    \item $\Sigma_X = AA^\top$, where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
-%    \item $\Sigma_N = BB^\top$, where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
-%\end{itemize}
-    $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
-    $X \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_X)$,
-    $N \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$,
-    $E \in \mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
-    $\Sigma_X = AA^\top$ where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
-    $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
-and the factor $\text{snr} \in (0, \infty)$.
+%\begin{itemize} \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries
+%drawn from $\mathcal{N}(0, d_x^{-1})$, \item $X \in \mathbb{R}^{n \times d_x}$
+%contains rows drawn from $\mathcal{N}(0, \Sigma_X)$, \item $N \in \mathbb{R}^{n
+%\times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, \item $E \in
+%\mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
+%\item $\Sigma_X = AA^\top$, where $A \in \mathbb{R}^{d_x \times d_x}$ contains
+%entries drawn from $\mathcal{N}(0, d_x^{-1})$, \item $\Sigma_N = BB^\top$,
+%where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
+%$\mathcal{N}(0, d_x^{-1})$, \end{itemize}
+    $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from
+$\mathcal{N}(0, d_x^{-1})$, $X \in \mathbb{R}^{n \times d_x}$ contains rows
+drawn from $\mathcal{N}(0, \Sigma_X)$, $N \in \mathbb{R}^{n \times d_x}$
+contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, $E \in \mathbb{R}^{d_x
+\times d_x}$ is a binary diagonal matrix containing $n_c$ ones, $\Sigma_X =
+AA^\top$ where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
+$\mathcal{N}(0, d_x^{-1})$, $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x
+\times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$, and the
+factor $\text{snr} \in (0, \infty)$.
 
-To simulate a wide range of experimental conditions, we vary $n \in \{500, 1000, 2000\}$, $d_x, d_y \in \{ 25, 50, 100, 150 \}$, $n_c \in \{ 5, 10 \}$, $\text{snr} \in \{ 0.2, 0.5, 0.75, 1 \}$.
+To simulate a wide range of experimental conditions, we vary $n \in \{500, 1000,
+2000\}$, $d_x, d_y \in \{ 25, 50, 100, 150 \}$, $n_c \in \{ 5, 10 \}$,
+$\text{snr} \in \{ 0.2, 0.5, 0.75, 1 \}$.
 %
@@
 
-We compare the performance of B2B against five competing methods:
+We compare the performance of B2B against four competing methods, all
+implemented in scikit-learn \citep{scikit}:
 %
-Ridge regression \citep{hoerl1959optimum}, multitask Lasso \citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls, tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and Reduced Rank Regression or RRR \citep{Izenman_rrr}.
+% To be updated
+
+\subsection{Baseline models}
+
+Forward regression consists of an $l2$-regularized "ridge" regression from the
+putative causes $X$ to the observations $Y$: \begin{equation} H = (X^T X
++\lambda I)^{-1} X^T Y \end{equation}
+
+Backward regression consists of an $l2$-regularized "ridge" regression from $Y$
+to $X$: \begin{equation} G = (Y^T Y +\lambda I)^{-1} Y^T X \end{equation}
+
+CCA finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$ such that
+$X$ and $Y$ are maximally correlated in a latent $Z$ space: \begin{equation} max
+corr(XH^T, YG^T) \end{equation}
+
+% To be checked
+PLS finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$ such that
+$X$ and $Y$ are maximally covarying in a latent $Z$ space: \begin{equation} max
+cov(XH^T, YG^T) \end{equation}
+
+We employ $5$-fold cross-validation to select the optimal number of components
+for CCA and PLS. Regressions were l2-regularized with a $\lambda$ regularization
+parameters fitted with the efficient leave-one-out procedure implemented in
+scikit-learn RidgeCV \citep{scikit}.
+
+\subsection{Evaluating Causal Discovery from models' coefficients}
+
+B2B leads to unbiased (i.e. zeros-centered) scalar coefficients for non-causal
+features. In contrast, the Forward, Backward, CCA and PLS models lead to a
+loading vector $H_i$ per feature $i$ (or one vector $G^i$ for the backward
+model). To transform such vector into an estimated causal contribution $\hat E$,
+we take the sum of square coefficients:
+
+\begin{equation} \hat E_i = \sum_j {H^j_i}^2 \end{equation}
+
+To estimate whether models accurately identify causal factors, we compute the
+area-under-the-curve (AUC) across factors $AUC(E, \hat E)$.
+%\begin{equation} AUC = 1 - \sum_1^n (E_k - E_{k-1}) ( \hat{E}_k +
+%\hat{E}_{k-1}) / 2 \end{equation}
+
+\subsection{Evaluating Causal Discovery with held-out prediction reliability}
+
+In most cases, $E$ is not known and AUC can thus not be estimated.
+
+To address this issue, we assess the ability of each model to reliably predict
+independent and identically distributed data from $Y$, given all of the $X$
+features versus all but one feature: $X_k$ (i.e. 'knock-out X'). This procedure
+results in two correlation metrics $R_{full}$ and $R_{knockout}$, whose
+difference $\Delta R=R_{full}-R_{knockout}$ indicates how much $X_i$ improves
+the prediction of $Y$ above and beyond what can be predicted from all other
+features $X_k$.
+
+For, the Forward Model, this feature importance is assessed as follow:
+
+\begin{algorithm}[H]
+    %\SetAlgoLined
+    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
+\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
+\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
+$\Delta{R} \in \mathbb{D}^{d_x, d_y}$.} $H = \text{LinearRegression}(X_{train},
+Y_{train})$\; $R_{full} = \text{corr}(X_{test} H, Y_{test})$\;
+
+    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $H_k =
+\text{LinearRegression}(X_{train} K, Y_{train})$\; $R_{k} = \text{corr}(X_{test}
+{H_k}, Y_{test})$\; $\Delta R_i = R_{full} - R_{k}$\; } \Return{$\Delta R$}
+\caption{Forward feature importance.} \label{algorithm:fwd_fi} \end{algorithm}
+
+\iffalse \begin{enumerate} \item Fit $H$ given $X_{train}$ and $Y_{train}$ \item
+For each feature $i$: \begin{itemize} \item Define $K$, and identity matrix
+whose row $i$ has been zeroed-out \item Fit $H^i_k$ given $X_{train} K$ and
+$Y_{train}$ \end{itemize} \end{enumerate} \fi
+
+\begin{algorithm}[H]
+    %\SetAlgoLined
+    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
+\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
+\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
+$\Delta{R} \in \mathbb{D}^{d_x, d_z}$.} $H, G = \text{CCA}(X_{train},
+Y_{train})$\; $R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;
+
+    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $H_k, G_k =
+\text{CCA}(X_{train} K, Y_{train})$\; $R_{k} = \text{corr}(X_{test} H_k,
+Y_{test} G_k)$\; $\Delta R_i = R_{full} - R_{k}$\; } \Return{$\Delta R$}
+\caption{CCA and PLS feature importance.} \label{algorithm:cdp_fi}
+\end{algorithm}
+
+\iffalse For CCA and PLS, feature importance is assessed as follow:
+\begin{enumerate} \item Fit $H$ and $G$ given $X_{train}$ and $Y_{train}$ \item
+For each feature $i$: \begin{itemize} \item Define $K$, and identity matrix
+whose row $i$ has been zeroed-out \item Fit $H^i_k$ and $G_k$ given $X_{train}
+K$ and $Y_{train}$ \end{itemize} \end{enumerate} \fi
+
+For the Backward Model, feature importance cannot be assessed. We thus simply
+report the model performance $R_i = corr(YG_i, X_i)$
+
+For, B2B feature importance is assessed as follow:
+
+\begin{algorithm}[H]
+    %\SetAlgoLined
+    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
+\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
+\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
+$\Delta{R} \in \mathbb{D}^{d_x}$.} $H, G = \text{B2B}(X_{train}, Y_{train})$\;
+$R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;
+
+    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $H_k =
+\text{LinearRegression}(X_{train} K, Y_{train} G_i)$\; $R_{k} =
+\text{corr}(X_{test} H_k, Y_{test} G_i)$\; $\Delta R_i = R_{full} - R_{k}$\; }
+\Return{$\Delta R$} \caption{B2B feature importance.} \label{algorithm:cdp_fi}
+\end{algorithm}
+
+\iffalse \begin{enumerate}
+
+  \item Fit $G$ given $X_{train}$ and $Y_{train}$ \item Fit $H$ given
+$X_{train}$ and $Y_{train} G^T$ \item For each feature $i$: \begin{itemize}
+\item Define $K$, and identity matrix whose row $i$ has been zeroed-out \item
+Fit $H^i_k$ given $X_{train} K$ and $Y_{train} G^T$ \end{itemize}
+
+  \item Compute $R_{full}$ given $Y_{test} G^T$ and $X_{test}$ \item For all $i
+\in d_x$: \begin{itemize} \item Compute $R^i_{knockout}$ given $Y_{test} G^T$
+and $X_{test} K {H^i_k}^T$ \item Compute $R^i_{delta}=R_{full} - R^i_{knockout}$
+\end{itemize} \end{enumerate} \fi
+
+\iffalse Ridge regression \citep{hoerl1959optimum}, multitask Lasso
+\citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls,
+tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and
+Reduced Rank Regression or RRR \citep{Izenman_rrr}.
 %
-Each of these methods estimates a matrix of coefficients $\hat{W} \in \mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$ using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
+Each of these methods estimates a matrix of coefficients $\hat{W} \in
+\mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$
+using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
 %
@@
 %
-test error in-distribution, test error out-distribution, test error-in distribution after weighting with $\hat{E}$, test error out-distribution after weighting with $\hat{E}$, Sonquist-Morgan false positives on $\hat{E}$, Sonquist-Morgan false negatives on $\hat{E}$, and the area under the ROC curve (AUC) between $\hat{E}$ and $E$.
+test error in-distribution, test error out-distribution, test error-in
+distribution after weighting with $\hat{E}$, test error out-distribution after
+weighting with $\hat{E}$, Sonquist-Morgan false positives on $\hat{E}$,
+Sonquist-Morgan false negatives on $\hat{E}$, and the area under the ROC curve
+(AUC) between $\hat{E}$ and $E$.
 %
-The ``in-distribution'' metrics measure the test error of each predictor before and after weighting their features by the estimated (continuous!) mask $\hat{E}$.
+The ``in-distribution'' metrics measure the test error of each predictor before
+and after weighting their features by the estimated (continuous!) mask
+$\hat{E}$.
 %
-The ``out-distribution'' metrics are similar, measuring the test error at samples drawn using two different random covariance matrices for both $X$ and $N$.
+The ``out-distribution'' metrics are similar, measuring the test error at
+samples drawn using two different random covariance matrices for both $X$ and
+$N$.
 %
-The false positives, false negatives, and AUC statistics measure the quality of identification of causal factors between $\hat{E}$ and the true, binary $E$.
+The false positives, false negatives, and AUC statistics measure the quality of
+identification of causal factors between $\hat{E}$ and the true, binary $E$.
 
-\begin{figure}[htpb]
-    \centering
-    \includegraphics[width=\textwidth]{synthetic.pdf}
-    \caption{Results of synthetic experiments.
-Each dot depicts the average value of a metric for B2B ($x$-axis) against a competitor ($y$-axis) for each experiment configuration averaged over $20$ random seeds.
-The prediction error on $Y$ within (in) and outside (out) of the training distribution are weakly but consistently larger for all of the models as compared to B2B with and without the Sonquist binarization (E).
-The AUC, measuring the ability of each model to separate causal from noncausal factors, is consistently larger in B2B as compared to the other models.}
-    \label{table:synthetic}
-\end{figure}
+\begin{figure}[htpb] \centering
+\includegraphics[width=\textwidth]{synthetic.pdf} \caption{Results of synthetic
+experiments. Each dot depicts the average value of a metric for B2B ($x$-axis)
+against a competitor ($y$-axis) for each experiment configuration averaged over
+$20$ random seeds. The prediction error on $Y$ within (in) and outside (out) of
+the training distribution are weakly but consistently larger for all of the
+models as compared to B2B with and without the Sonquist binarization (E). The
+AUC, measuring the ability of each model to separate causal from noncausal
+factors, is consistently larger in B2B as compared to the other models.}
+\label{table:synthetic} \end{figure} \fi
 
-Figure~\ref{table:synthetic} summarizes the results for all experimental conditions, metrics, and methods.
+Figure~\ref{table:synthetic} summarizes the results for all experimental
+conditions, metrics, and methods.
 %
-B2B is the method obtaining best test errors, lowest false positive and false negative discovery about $E$, and overall the best AUC when it comes to detecting causal influences.
+B2B is the method obtaining best test errors, lowest false positive and false
+negative discovery about $E$, and overall the best AUC when it comes to
+detecting causal influences.

### 2019-09-24 paper/synthetic.tex
modifs in synthetic

@@
 
-The data generating process for each experiment constructs $n$ training examples
+The data generating process for each experiment constructs $n=1000$ training examples
 according to the model $Y = (\text{snr} \cdot XE + N)F$, where $\text{snr}$ is a
@@
 
-To simulate a wide range of experimental conditions, we vary $n \in \{500, 1000,
-2000\}$, $d_x, d_y \in \{ 25, 50, 100, 150 \}$, $n_c \in \{ 5, 10 \}$,
-$\text{snr} \in \{ 0.2, 0.5, 0.75, 1 \}$.
+To simulate a wide range of experimental conditions, we vary $d_x, d_y \in \{ 10, 15, 25, 39, 63, 100 \}$, $n_c \in \{ 1, 2, 6, 15, 39, 100 \}$,
+$\text{snr} \in \{ 0.001, 0.015, 0.25, 3.98, 63, 1000 \}$. We discard the cases where $n_c > d_x$. We limit $d_x, d_y$ to 100 in order to keep the running time under 2 hours for each condition.
 %
-Each condition is simulated under $20$ different random seeds.
+
+Each condition is simulated under $5$ different random seeds.
 
@@
 
-CCA finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$ such that
-$X$ and $Y$ are maximally correlated in a latent $Z$ space: \begin{equation} max
-corr(XH^T, YG^T) \end{equation}
+CCA finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$
+% such that
+s.t.
+$X$ and $Y$ are maximally correlated in a latent $Z$ space:
+% \begin{equation} maxcorr(XH^T, YG^T) \end{equation}
+\begin{equation} G,H = \argmax_{G,H} corr(XH^T, YG^T) \end{equation}
 
 % To be checked
-PLS finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$ such that
-$X$ and $Y$ are maximally covarying in a latent $Z$ space: \begin{equation} max
-cov(XH^T, YG^T) \end{equation}
+PLS finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$
+% such that
+s.t.
+$X$ and $Y$ are maximally covarying in a latent $Z$ space:
+% \begin{equation} maxcov(XH^T, YG^T) \end{equation}
+\begin{equation} G,H = \argmax_{G,H} cov(XH^T, YG^T) \end{equation}
 
 We employ $5$-fold cross-validation to select the optimal number of components
-for CCA and PLS. Regressions were l2-regularized with a $\lambda$ regularization
+for CCA and PLS. Regressions were $\ell2$-regularized with a $\lambda$ regularization
 parameters fitted with the efficient leave-one-out procedure implemented in
@@
 %\hat{E}_{k-1}) / 2 \end{equation}
+
+The AUC allows evaluating the capacity of models at detecting the causal
+importance of factors when ground truth labels are available, as is the case in
+this setup.
+
+We report AUC results in Figure~\ref{fig:auc_plots}, and compare favorably to all baselines.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.8\linewidth]{figures/AUC.png}
+  \caption{We show boxplots for the AUC metric obtained in our synthetic experiments. Higher is better. B2B compares favorably to baselines.}
+  \label{fig:auc_plots}
+\end{figure}
 
@@
 
-\iffalse \begin{enumerate} \item Fit $H$ given $X_{train}$ and $Y_{train}$ \item
-For each feature $i$: \begin{itemize} \item Define $K$, and identity matrix
-whose row $i$ has been zeroed-out \item Fit $H^i_k$ given $X_{train} K$ and
-$Y_{train}$ \end{itemize} \end{enumerate} \fi
+\iffalse
+\begin{enumerate}
+  \item Fit $H$ given $X_{train}$ and $Y_{train}$
+  \item For each feature $i$:
+  \begin{itemize}
+    \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
+    \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train}$
+  \end{itemize}
+  \end{enumerate}
+\fi
 
@@
 
-\iffalse For CCA and PLS, feature importance is assessed as follow:
-\begin{enumerate} \item Fit $H$ and $G$ given $X_{train}$ and $Y_{train}$ \item
-For each feature $i$: \begin{itemize} \item Define $K$, and identity matrix
-whose row $i$ has been zeroed-out \item Fit $H^i_k$ and $G_k$ given $X_{train}
-K$ and $Y_{train}$ \end{itemize} \end{enumerate} \fi
+\iffalse
+For CCA and PLS, feature importance is assessed as follow:
+\begin{enumerate}
+  \item Fit $H$ and $G$ given $X_{train}$ and $Y_{train}$
+  \item For each feature $i$:
+  \begin{itemize}
+    \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
+    \item Fit $H^i_k$ and $G_k$ given $X_{train} K$ and $Y_{train}$
+  \end{itemize}
+\end{enumerate}
+\fi
 
@@
 
-\iffalse \begin{enumerate}
+\iffalse
+\begin{enumerate}
 
-  \item Fit $G$ given $X_{train}$ and $Y_{train}$ \item Fit $H$ given
-$X_{train}$ and $Y_{train} G^T$ \item For each feature $i$: \begin{itemize}
-\item Define $K$, and identity matrix whose row $i$ has been zeroed-out \item
-Fit $H^i_k$ given $X_{train} K$ and $Y_{train} G^T$ \end{itemize}
+  \item Fit $G$ given $X_{train}$ and $Y_{train}$
+  \item Fit $H$ given $X_{train}$ and $Y_{train} G^T$
+  \item For each feature $i$:
+  \begin{itemize}
+    \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
+    \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train} G^T$
+  \end{itemize}
 
-  \item Compute $R_{full}$ given $Y_{test} G^T$ and $X_{test}$ \item For all $i
-\in d_x$: \begin{itemize} \item Compute $R^i_{knockout}$ given $Y_{test} G^T$
-and $X_{test} K {H^i_k}^T$ \item Compute $R^i_{delta}=R_{full} - R^i_{knockout}$
-\end{itemize} \end{enumerate} \fi
+  \item Compute $R_{full}$ given $Y_{test} G^T$ and $X_{test}$
+  \item For all $i \in d_x$:
+  \begin{itemize}
+    \item Compute $R^i_{knockout}$ given $Y_{test} G^T$ and $X_{test} K {H^i_k}^T$
+    \item Compute $R^i_{delta}=R_{full} - R^i_{knockout}$
+  \end{itemize}
+  \end{enumerate}
+\fi
 
@@
 detecting causal influences.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.7\linewidth]{figures/delta_b2b.png}
+  \caption{Boxplots "boxen" of feature importance values relative to B2B on synthetic experiments. Boxes below the dashed line mean B2B is better.}
+  \label{}
+\end{figure}

### 2019-09-25 paper/meg_exps.tex
wip

@@
 recordings of 102 healthy native-Dutch adults who participated in a reading
-task.
+task. Twelve subjects were excluded from the analysis because of corrupted file headers.
 %
@@
 segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
-thus led a sampling frequency of 240Hz.
+thus led a sampling frequency of 240Hz. The average responses across words is displayed in \ref{fig:meg_evoked}
 For each subject and each time sample relative to word onset, we
@@
 
+\begin{figure}[t!]
+  \centering
+  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_evoked.pdf}
+  \caption{102 subjects read $\approx$ 2,700 words while their brain activity was recorded with MEG.
+Top.
+Average brain response to words relative to their onset (gradiometers), as viewed from above the head.
+Bottom.
+Each line represents a MEG sensor (magnetometer), color-coded by spatial position.
+Posterior responses, typical of primary visual cortex activity, peak around 100 ms after word onset and are followed by an anterior propagation of activity typical of semantic processing in the associative cortices.
+Vertical lines indicate moments of above topographies.
+  }
+  \label{fig:meg_evoked}
+\end{figure}
+
 \subsubsection{Feature definition}
-We annotated each word with three collinear features.
+
+When human subjects read words, the number of letters is known to modulate
+brain activity from 100ms after word onset \citep{pegado2014timing}.
+Similarly, word frequency is known to modulate brain activity around 400ms
+\citep{kutas2011thirty}. In the present study, we consider these two well-known features, as well as an untested feature ('Word Function') and dummy variable.
 %
 "Word Length' refers to the total number of letters, and is expected to specifically cause a variation in the early evoked MEG responses (i.e.
-from 100 ms after stimulus onset), because it variably engages different retinotopically-tuned visual areas in the primary visual cortex.
+from 100 ms after stimulus onset) elicited by the retinotopically-tuned visual cortices.
 %
@@
 To our knowledge, this feature has not been thouroughly investigated with MEG.
-Its causal contribution to reading processes in the brain is therefore largely unknown.
+Its causal contribution to reading processes in the brain is thus largely unknown.
 %
@@
 \begin{equation}
-  dummy = z(length) + z(frequency) + z(\mathcal{N})
+  dummy = z(length) + z(frequency) + \mathcal{N}
 \end{equation}
-where $z$ transforms features into zero-mean and unit variance vectors.
+where $z$ transforms features into zero-mean and unit variance vectors and $\mathcal{N}$ is random vectors sampling a zero-mean unit-variance Gaussian distribution.
 
@@
 
-We compare B2B to four standard models: forward regression, backward regression, CCA and PLS, as implemented in scikit-learn, and optimized with nested cross validation over 20 $\lambda$ l2 regularization parameters logarithmically spaced between $10^-4$ and $10^4$ or 1 to 4 caconical components.
+We compare B2B to four standard models: forward regression, backward regression, CCA and PLS, as implemented in scikit-learn, and optimized with nested cross validation over twenty l2 regularization parameters logarithmically spaced between $10^-4$ and $10^4$ or 1 to 4 caconical components.
 
-We used the feature importance described in XXX algorithms XXX to assess the extent to which each feature $X_i$ specifically improves the prediction of held-out $Y$ data.
+We used the feature importance described in \ref{algorithm:b2b_fi} to assess the extent to which each feature $X_i$ specifically improves the prediction of held-out $Y$ data.
 
-Corresponding results can be found in Figure \ref{fig:meg_twocurves}.
+Each model was implemented for each subject and each time sample independently. Pairwise comparison across models were performed using a wilcoxon test on the $\Delta R$ averaged across time.
+
+Corresponding results can be found in Figure \ref{fig:meg_results}. Significant pairwise differences across models are highlighted with the top horizontal bars in \ref{fig:meg_results}.
 
@@
   \centering
-  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_twocurves.pdf}
-  \caption{We applied forward regression (a.k.a encoding coefficients), backward regression (a.k.a decoding scores) and B2B regression on 301-sensor MEG recordings of brain activity, while subjects were flashed with individual words every $\approx$650ms.
-Top row: the first subject.
-Bottom row: average across subjects.
-Each line represent the link between word features $X$ and MEG signals $Y$
-  as a function of time relative to word onset ($x$ axis).
-The $y$-axis varies for each model (forward: H coefficients; backward: pearson correlation coefficients; B2B: $\hat E$).
-Forward regression makes prediction for each MEG channel independently; detecting an effect thus requires an additional average of the norm and can diminish sensitivity.
-Backward regression operates across all channels, but does not learn to disambiguate the two word features.
-B2B disambiguates these two factors consistently with previous neuroscientific work \cite{kutas2011thirty}\citep{pegado2014timing}
-  showing early responses ($\approx$ 150ms) to word length and later responses
-  ($\approx$ 400ms) to word frequency.}
-  \label{fig:meg_twocurves}
+  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg.pdf}
+  \caption{Forward regression (Blue), PLS (orange), CCA (green) and B2B (red) are compared on their ability to reliably predict single-trial MEG signals evoked by words.
+Left.
+Average improvement of correlation coefficient $\Delta R$ obtained on held-out data when each of the four features "Word Length", "Word Frequency", "Word Function" and "Dummy Variable" was provided to the model \ref{algorithm:b2b_fi}.
+Error bars indicate standard error of the mean (SEM) across subjects.
+Right.
+Average $\Delta R$ across time.
+Each dot represents a subject.
+The top horizontal lines indicate significant differences between pairwise comparison across models, color coded by the winning model.
+B2B outperforms all models on meaningful features, at the exception of CCA for the "word length".}
+  \label{fig:meg_results}
 \end{figure}
 
-%Forward regression was performed with an $\ell_2$-regularized least-squares regression $H \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| H_{i, :} \|^2$.
-%Backward regression was performed with an $\ell_2$-regularized least-squares regression $G \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YG$, using a validation sample.
-Finally, we used the back-to-back regression was performed using two $\ell_2$-regularized
-%least squares regressions $G$ and $H$ as described in section \ref{sec:algorithm}, using a bagging factor of 10, and alpha cross-validated between $10^{-5}$ and $10^{5}$.
+\subsubsection{Results}
+We compared the ability of Forward regression, Backward regression, CCA, PLS and B2B to estimate the causal contribution of four distinct but collinear features on brain evoked responses to words.
 
-%The three types of analyses were implemented for each subject and each time
-%sample independently. To estimate the reliability of these estimates in the
-%population, we tested, using a one-sample $t$-test whether the distribution of
-%the estimated contribution of each factor differed from baseline ($t=0$ms),
-%as $\text{t-test}(\hat E_{t=0} - \hat E_{\tau})$, with $\tau$ ranging from
-%150ms to 1s after word onset. We use the resulting t and p-values as proxy to
-%quantify the performance of each model. These results are depicted in Figure
-%3 of the Supplementary Material. For concision purposes, we only report this metric averaged
-%between 150ms and 1s after word onset.
+First, \ref{fig:meg_supp} shows that the Backward model decodes the dummy variable well above chance level.
+In addition, "Word Length" and  "Word Frequency" present a similar decoding time course, even though these features are known to influence early and late MEG responses respectively \citep{kutas2011thirty}.
+These results illustrate the well-known fact that backward modelling cannot be used to estimate the causal contribution of collinear features.
 
-\subsubsection{Results}
-When human subjects read individual words, the number of letters is known to modulate
-brain activity from 150ms after word onset \citep{pegado2014timing}.
-Similarly, word frequency is known to modulate brain activity around 400ms
-\citep{kutas2011thirty}.
-Consequently, we compared the ability of forward, backward and B2B regressions to specifically detect these two types of brain responses in a large neuroimaging dataset, where word length and word frequency are collinear.
+We thus focus on valid modeling (Forward Regression, PLS, CCA, and B2B) and estimate the change of correlation coefficients $\Delta R$ induced by the introduction of each feature into the model \ref{algorithm:b2b_fi}.
+As expected, none of the models significantly predicted the 'Dummy Variable' to be a causal factors (i.e.
+$\Delta R < 0$ for all models, all $p > .089$)
 
-Figure \ref{fig:meg_twocurves} shows, for each model, the effects obtained
-in a representative subject, as well as the average effects across subjects.
+Figure \ref{fig:meg_results} shows, for each model, the effects obtained across time (left) and subjects (right).
+
 The results show that all models significantly detect that both word length and
@@
 B2B thus promises to improve the detectability of subtle effects.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_result.pdf}
-  \caption{Cumulative estimates of word features as a function of time relative to word onset (chance level = 0).
-The dark blue area (word length) and dark red area (word frequency) correspond to the same data as those of the B2B average across subjects presented in \ref{fig:meg_twocurves}.
-A variety of visual features peak around 150ms, while a variety of lexical features tend to peak around 400ms.}
-  \label{fig:megresult}
-\end{figure}

### 2019-09-25 paper/supplementary.tex
wip

@@
 
-\begin{figure}[h]
+\begin{figure}[t!]
   \centering
-  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_sensors.pdf}
-  \caption{Magnetosensor response (femtoteslas fT) averaged across words for the first subject.
-The color coding corresponds to the positions of the sensors on the head as shown on the top-left diagram.
-The diverging curves correspond to artifacts in the MEG data.}
-  \label{fig:megavg}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm]{figures/pvalues.pdf}
-  \caption{We apply a statistical t-test to measure if coefficients were significantly different from baseline value (at word onset, $t=0$ms).
-  We compute the negative logarithm of the p-value (\textit{higher is better}: nonzero effects are more easily detected).
-We plot the average across features, over time.
-Red: word length, blue: word frequency.
-  Full line: B2B, dashed line: Forward regression. We see that our method more easily detects the influence of causes of MEG response.}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm]{figures/pvalues_vertical.pdf}
-  \caption{We show the average value and the standard deviation across subjects of the coefficients obtained with our method during the trials.
-the $\star$  symbol corresponds to statistically significant (t-test) nonzero values.}
-  \label{fig:ridgebaselineresult}
+  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_supp.pdf}
+  \caption{}
+  \label{fig:meg_supp}
 \end{figure}

### 2019-09-25 paper/synthetic.tex
wip
added plots

@@
 
-To simulate a wide range of experimental conditions, we vary $d_x, d_y \in \{ 10, 15, 25, 39, 63, 100 \}$, $n_c \in \{ 1, 2, 6, 15, 39, 100 \}$,
-$\text{snr} \in \{ 0.001, 0.015, 0.25, 3.98, 63, 1000 \}$. We discard the cases where $n_c > d_x$. We limit $d_x, d_y$ to 100 in order to keep the running time under 2 hours for each condition.
+To simulate a wide range of experimental conditions, we sample 10 values in log-space for $d_x, d_y \in \left[ 10, 100 \right]$, $n_c \in \left[ 3, 63 \right]$,
+$\text{snr} \in \left[ 0.001, 10 \right]$. We discard the cases where $n_c > d_x$, limit $d_x, d_y$ to 100 to keep the running time under 2 hours for each condition, and average over 5 random seeds.
 %
-
-Each condition is simulated under $5$ different random seeds.
+% Each condition is simulated under $5$ different random seeds.
 
@@
 
-We report AUC results in Figure~\ref{fig:auc_plots}, and compare favorably to all baselines.
+We report AUC results in Figures~\ref{fig:auc_plots}~(left) and ~\ref{fig:percondition}~(top), and compare favorably to all baselines.
 
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{figures/AUC.png}
-  \caption{We show boxplots for the AUC metric obtained in our synthetic experiments. Higher is better. B2B compares favorably to baselines.}
-  \label{fig:auc_plots}
+  \includegraphics[width=\linewidth]{figures/AUC_R.png}
+  \vspace{-4ex}
+  \caption{Synthetic experiments.
+Distribution of differences in AUC (left) and Feature Importance (right) metrics between our method and the baselines.
+Mass under the dashed line corresponds to B2B outperforming the baseline.
+\label{fig:auc_plots}}
+  \vspace{2ex}
+  \includegraphics[width=\linewidth]{figures/AUC_conditions.png}
+  \includegraphics[width=\linewidth]{figures/R_conditions.png}
+  \vspace{-4ex}
+  \caption{Synthetic experiments. Average AUC (top) and Feature Importance R (bottom) metrics as a function of experimental conditions. Higher is better. \label{fig:percondition}}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \label{}
 \end{figure}
@@
 \text{corr}(X_{test} H_k, Y_{test} G_i)$\; $\Delta R_i = R_{full} - R_{k}$\; }
-\Return{$\Delta R$} \caption{B2B feature importance.} \label{algorithm:cdp_fi}
+\Return{$\Delta R$} \caption{B2B feature importance.} \label{algorithm:b2b_fi}
 \end{algorithm}
@@
 detecting causal influences.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.7\linewidth]{figures/delta_b2b.png}
-  \caption{Boxplots "boxen" of feature importance values relative to B2B on synthetic experiments. Boxes below the dashed line mean B2B is better.}
-  \label{}
-\end{figure}

### 2019-09-25 paper/synthetic.tex
figures synthetic

@@
   \caption{Synthetic experiments.
-Distribution of differences in AUC (left) and Feature Importance (right) metrics between our method and the baselines.
+Distribution of differences in AUC (left) and Feature Importance $\Delta R$ (right) metrics between our method and the baselines.
 Mass under the dashed line corresponds to B2B outperforming the baseline.
@@
   \vspace{2ex}
-  \includegraphics[width=\linewidth]{figures/AUC_conditions.png}
-  \includegraphics[width=\linewidth]{figures/R_conditions.png}
+  \includegraphics[width=\linewidth]{figures/AUC_conditions}
+  \includegraphics[width=\linewidth]{figures/R_conditions}
   \vspace{-4ex}
-  \caption{Synthetic experiments. Average AUC (top) and Feature Importance R (bottom) metrics as a function of experimental conditions. Higher is better. \label{fig:percondition}}
+  \caption{Synthetic experiments. Average AUC (top) and Feature Importance $\Delta R$ (bottom) metrics as a function of experimental conditions. Higher is better. \label{fig:percondition}}
 \end{figure}
@@
 independent and identically distributed data from $Y$, given all of the $X$
-features versus all but one feature: $X_k$ (i.e. 'knock-out X'). This procedure
+features versus all-but-one feature: $X_k$ (i.e. 'knock-out X'). This procedure
 results in two correlation metrics $R_{full}$ and $R_{knockout}$, whose
 difference $\Delta R=R_{full}-R_{knockout}$ indicates how much $X_i$ improves
-the prediction of $Y$ above and beyond what can be predicted from all other
+the prediction of $Y$% above and
+beyond what can be predicted from all other
 features $X_k$.

### 2019-09-25 paper/appendix.tex
synthetic

--- 2019-09-24 paper/synthetic.tex
@@
-We evaluate the performance of B2B throughout a series of experiments on
-controlled synthetic data.
-%
-The purpose of these experiments is to evaluate the ability of B2B in terms of
-prediction (inside and outside the training distribution), as well as a method
-to recover causal factors.
+\newpage
+\appendix
+\section{Appendix}
+\subsection{Feature importance}
 
-The data generating process for each experiment constructs $n=1000$ training examples
-according to the model $Y = (\text{snr} \cdot XE + N)F$, where $\text{snr}$ is a
-scalar that modulates the signal-to-noise ratio.
-%
-Here,
-%\begin{itemize} \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries
-%drawn from $\mathcal{N}(0, d_x^{-1})$, \item $X \in \mathbb{R}^{n \times d_x}$
-%contains rows drawn from $\mathcal{N}(0, \Sigma_X)$, \item $N \in \mathbb{R}^{n
-%\times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, \item $E \in
-%\mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
-%\item $\Sigma_X = AA^\top$, where $A \in \mathbb{R}^{d_x \times d_x}$ contains
-%entries drawn from $\mathcal{N}(0, d_x^{-1})$, \item $\Sigma_N = BB^\top$,
-%where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
-%$\mathcal{N}(0, d_x^{-1})$, \end{itemize}
-    $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from
-$\mathcal{N}(0, d_x^{-1})$, $X \in \mathbb{R}^{n \times d_x}$ contains rows
-drawn from $\mathcal{N}(0, \Sigma_X)$, $N \in \mathbb{R}^{n \times d_x}$
-contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, $E \in \mathbb{R}^{d_x
-\times d_x}$ is a binary diagonal matrix containing $n_c$ ones, $\Sigma_X =
-AA^\top$ where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
-$\mathcal{N}(0, d_x^{-1})$, $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x
-\times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$, and the
-factor $\text{snr} \in (0, \infty)$.
-
-To simulate a wide range of experimental conditions, we vary $d_x, d_y \in \{ 10, 15, 25, 39, 63, 100 \}$, $n_c \in \{ 1, 2, 6, 15, 39, 100 \}$,
-$\text{snr} \in \{ 0.001, 0.015, 0.25, 3.98, 63, 1000 \}$. We discard the cases where $n_c > d_x$. We limit $d_x, d_y$ to 100 in order to keep the running time under 2 hours for each condition.
-%
-
-Each condition is simulated under $5$ different random seeds.
-
-We compare the performance of B2B against four competing methods, all
-implemented in scikit-learn \citep{scikit}:
-%
-% To be updated
-
-\subsection{Baseline models}
-
-Forward regression consists of an $l2$-regularized "ridge" regression from the
-putative causes $X$ to the observations $Y$: \begin{equation} H = (X^T X
-+\lambda I)^{-1} X^T Y \end{equation}
-
-Backward regression consists of an $l2$-regularized "ridge" regression from $Y$
-to $X$: \begin{equation} G = (Y^T Y +\lambda I)^{-1} Y^T X \end{equation}
-
-CCA finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$
-% such that
-s.t.
-$X$ and $Y$ are maximally correlated in a latent $Z$ space:
-% \begin{equation} maxcorr(XH^T, YG^T) \end{equation}
-\begin{equation} G,H = \argmax_{G,H} corr(XH^T, YG^T) \end{equation}
-
-% To be checked
-PLS finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$
-% such that
-s.t.
-$X$ and $Y$ are maximally covarying in a latent $Z$ space:
-% \begin{equation} maxcov(XH^T, YG^T) \end{equation}
-\begin{equation} G,H = \argmax_{G,H} cov(XH^T, YG^T) \end{equation}
-
-We employ $5$-fold cross-validation to select the optimal number of components
-for CCA and PLS. Regressions were $\ell2$-regularized with a $\lambda$ regularization
-parameters fitted with the efficient leave-one-out procedure implemented in
-scikit-learn RidgeCV \citep{scikit}.
-
-\subsection{Evaluating Causal Discovery from models' coefficients}
-
-B2B leads to unbiased (i.e. zeros-centered) scalar coefficients for non-causal
-features. In contrast, the Forward, Backward, CCA and PLS models lead to a
-loading vector $H_i$ per feature $i$ (or one vector $G^i$ for the backward
-model). To transform such vector into an estimated causal contribution $\hat E$,
-we take the sum of square coefficients:
-
-\begin{equation} \hat E_i = \sum_j {H^j_i}^2 \end{equation}
-
-To estimate whether models accurately identify causal factors, we compute the
-area-under-the-curve (AUC) across factors $AUC(E, \hat E)$.
-%\begin{equation} AUC = 1 - \sum_1^n (E_k - E_{k-1}) ( \hat{E}_k +
-%\hat{E}_{k-1}) / 2 \end{equation}
-
-The AUC allows evaluating the capacity of models at detecting the causal
-importance of factors when ground truth labels are available, as is the case in
-this setup.
-
-We report AUC results in Figure~\ref{fig:auc_plots}, and compare favorably to all baselines.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.8\linewidth]{figures/AUC.png}
-  \caption{We show boxplots for the AUC metric obtained in our synthetic experiments. Higher is better. B2B compares favorably to baselines.}
-  \label{fig:auc_plots}
-\end{figure}
-
-\subsection{Evaluating Causal Discovery with held-out prediction reliability}
-
-In most cases, $E$ is not known and AUC can thus not be estimated.
-
-To address this issue, we assess the ability of each model to reliably predict
-independent and identically distributed data from $Y$, given all of the $X$
-features versus all but one feature: $X_k$ (i.e. 'knock-out X'). This procedure
-results in two correlation metrics $R_{full}$ and $R_{knockout}$, whose
-difference $\Delta R=R_{full}-R_{knockout}$ indicates how much $X_i$ improves
-the prediction of $Y$ above and beyond what can be predicted from all other
-features $X_k$.
-
-For, the Forward Model, this feature importance is assessed as follow:
+For the Forward Model, the feature importance is assessed as follows:
 
@@
 
-\iffalse
-\begin{enumerate}
-  \item Fit $H$ given $X_{train}$ and $Y_{train}$
-  \item For each feature $i$:
-  \begin{itemize}
-    \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
-    \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train}$
-  \end{itemize}
-  \end{enumerate}
-\fi
+% \iffalse
+% \begin{enumerate}
+%   \item Fit $H$ given $X_{train}$ and $Y_{train}$
+%   \item For each feature $i$:
+%   \begin{itemize}
+%     \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
+%     \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train}$
+%   \end{itemize}
+%   \end{enumerate}
+% \fi
+
+For the CCA and PLS models, the feature importance is assessed as follows:
 
@@
 
-\iffalse
-For CCA and PLS, feature importance is assessed as follow:
-\begin{enumerate}
-  \item Fit $H$ and $G$ given $X_{train}$ and $Y_{train}$
-  \item For each feature $i$:
-  \begin{itemize}
-    \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
-    \item Fit $H^i_k$ and $G_k$ given $X_{train} K$ and $Y_{train}$
-  \end{itemize}
-\end{enumerate}
-\fi
-
-For the Backward Model, feature importance cannot be assessed. We thus simply
-report the model performance $R_i = corr(YG_i, X_i)$
-
-For, B2B feature importance is assessed as follow:
-
-\begin{algorithm}[H]
-    %\SetAlgoLined
-    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
-\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
-\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
-$\Delta{R} \in \mathbb{D}^{d_x}$.} $H, G = \text{B2B}(X_{train}, Y_{train})$\;
-$R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;
-
-    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $H_k =
-\text{LinearRegression}(X_{train} K, Y_{train} G_i)$\; $R_{k} =
-\text{corr}(X_{test} H_k, Y_{test} G_i)$\; $\Delta R_i = R_{full} - R_{k}$\; }
-\Return{$\Delta R$} \caption{B2B feature importance.} \label{algorithm:cdp_fi}
-\end{algorithm}
-
-\iffalse
-\begin{enumerate}
-
-  \item Fit $G$ given $X_{train}$ and $Y_{train}$
-  \item Fit $H$ given $X_{train}$ and $Y_{train} G^T$
-  \item For each feature $i$:
-  \begin{itemize}
-    \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
-    \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train} G^T$
-  \end{itemize}
-
-  \item Compute $R_{full}$ given $Y_{test} G^T$ and $X_{test}$
-  \item For all $i \in d_x$:
-  \begin{itemize}
-    \item Compute $R^i_{knockout}$ given $Y_{test} G^T$ and $X_{test} K {H^i_k}^T$
-    \item Compute $R^i_{delta}=R_{full} - R^i_{knockout}$
-  \end{itemize}
-  \end{enumerate}
-\fi
-
-\iffalse Ridge regression \citep{hoerl1959optimum}, multitask Lasso
-\citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls,
-tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and
-Reduced Rank Regression or RRR \citep{Izenman_rrr}.
-%
-Each of these methods estimates a matrix of coefficients $\hat{W} \in
-\mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$
-using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
-%
-For B2B, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
-%
-
-We evaluate seven metrics:
-%
-test error in-distribution, test error out-distribution, test error-in
-distribution after weighting with $\hat{E}$, test error out-distribution after
-weighting with $\hat{E}$, Sonquist-Morgan false positives on $\hat{E}$,
-Sonquist-Morgan false negatives on $\hat{E}$, and the area under the ROC curve
-(AUC) between $\hat{E}$ and $E$.
-%
-The ``in-distribution'' metrics measure the test error of each predictor before
-and after weighting their features by the estimated (continuous!) mask
-$\hat{E}$.
-%
-The ``out-distribution'' metrics are similar, measuring the test error at
-samples drawn using two different random covariance matrices for both $X$ and
-$N$.
-%
-The false positives, false negatives, and AUC statistics measure the quality of
-identification of causal factors between $\hat{E}$ and the true, binary $E$.
-
-\begin{figure}[htpb] \centering
-\includegraphics[width=\textwidth]{synthetic.pdf} \caption{Results of synthetic
-experiments. Each dot depicts the average value of a metric for B2B ($x$-axis)
-against a competitor ($y$-axis) for each experiment configuration averaged over
-$20$ random seeds. The prediction error on $Y$ within (in) and outside (out) of
-the training distribution are weakly but consistently larger for all of the
-models as compared to B2B with and without the Sonquist binarization (E). The
-AUC, measuring the ability of each model to separate causal from noncausal
-factors, is consistently larger in B2B as compared to the other models.}
-\label{table:synthetic} \end{figure} \fi
-
-Figure~\ref{table:synthetic} summarizes the results for all experimental
-conditions, metrics, and methods.
-%
-B2B is the method obtaining best test errors, lowest false positive and false
-negative discovery about $E$, and overall the best AUC when it comes to
-detecting causal influences.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.7\linewidth]{figures/delta_b2b.png}
-  \caption{Boxplots "boxen" of feature importance values relative to B2B on synthetic experiments. Boxes below the dashed line mean B2B is better.}
-  \label{}
-\end{figure}
+For the Backward Model, feature importance cannot be assessed because there is no prediction.

### 2019-09-25 paper/synthetic.tex
synthetic

@@
 
-\begin{figure}
+\begin{figure}[ht]
   \centering
@@
   \caption{Synthetic experiments.
-Distribution of differences in AUC (left) and Feature Importance $\Delta R$ (right) metrics between our method and the baselines.
+Distribution (over conditions) of differences in AUC (left) and Feature Importance $\Delta R$ (right) metrics between our method and the baselines.
 Mass under the dashed line corresponds to B2B outperforming the baseline.
@@
 
-\begin{figure}
-  \centering
-  \label{}
-\end{figure}
-
 \subsection{Evaluating Causal Discovery with held-out prediction reliability}
@@
 independent and identically distributed data from $Y$, given all of the $X$
-features versus all-but-one feature: $X_k$ (i.e. 'knock-out X'). This procedure
+features versus all-but-ones feature $X_{-i}$ (i.e. 'knock-out X'). This procedure
 results in two correlation metrics $R_{full}$ and $R_{knockout}$, whose
-difference $\Delta R=R_{full}-R_{knockout}$ indicates how much $X_i$ improves
-the prediction of $Y$% above and
-beyond what can be predicted from all other
-features $X_k$.
+difference $\Delta R_i = R_{full}-R_{knockout}$ indicates how much each $X_i$
+improves the prediction of $Y$.
+%beyond what can be predicted from all other features $X_{-i}$.
+In our figures, $\Delta R$ is the average of $\Delta R_i$. A higher score means that for prediction, the model relies on individual features rather than combinations of features.
 
-For, the Forward Model, this feature importance is assessed as follow:
-
-\begin{algorithm}[H]
-    %\SetAlgoLined
-    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
-\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
-\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
-$\Delta{R} \in \mathbb{D}^{d_x, d_y}$.} $H = \text{LinearRegression}(X_{train},
-Y_{train})$\; $R_{full} = \text{corr}(X_{test} H, Y_{test})$\;
-
-    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $H_k =
-\text{LinearRegression}(X_{train} K, Y_{train})$\; $R_{k} = \text{corr}(X_{test}
-{H_k}, Y_{test})$\; $\Delta R_i = R_{full} - R_{k}$\; } \Return{$\Delta R$}
-\caption{Forward feature importance.} \label{algorithm:fwd_fi} \end{algorithm}
-
-\iffalse
-\begin{enumerate}
-  \item Fit $H$ given $X_{train}$ and $Y_{train}$
-  \item For each feature $i$:
-  \begin{itemize}
-    \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
-    \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train}$
-  \end{itemize}
-  \end{enumerate}
-\fi
-
-\begin{algorithm}[H]
-    %\SetAlgoLined
-    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
-\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
-\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
-$\Delta{R} \in \mathbb{D}^{d_x, d_z}$.} $H, G = \text{CCA}(X_{train},
-Y_{train})$\; $R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;
-
-    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $H_k, G_k =
-\text{CCA}(X_{train} K, Y_{train})$\; $R_{k} = \text{corr}(X_{test} H_k,
-Y_{test} G_k)$\; $\Delta R_i = R_{full} - R_{k}$\; } \Return{$\Delta R$}
-\caption{CCA and PLS feature importance.} \label{algorithm:cdp_fi}
-\end{algorithm}
-
-\iffalse
-For CCA and PLS, feature importance is assessed as follow:
-\begin{enumerate}
-  \item Fit $H$ and $G$ given $X_{train}$ and $Y_{train}$
-  \item For each feature $i$:
-  \begin{itemize}
-    \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
-    \item Fit $H^i_k$ and $G_k$ given $X_{train} K$ and $Y_{train}$
-  \end{itemize}
-\end{enumerate}
-\fi
-
-For the Backward Model, feature importance cannot be assessed. We thus simply
-report the model performance $R_i = corr(YG_i, X_i)$
-
-For, B2B feature importance is assessed as follow:
+For B2B, feature importance is assessed as follows:
 
@@
 
-\iffalse
-\begin{enumerate}
+Feature importance can be computed similarly for the CCA, PLS and Forward baselines, and we provide similar algorithm boxes in the appendix.
+For the Backward Model, feature importance cannot be assessed as there is no prediction.
 
-  \item Fit $G$ given $X_{train}$ and $Y_{train}$
-  \item Fit $H$ given $X_{train}$ and $Y_{train} G^T$
-  \item For each feature $i$:
-  \begin{itemize}
-    \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
-    \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train} G^T$
-  \end{itemize}
+We show in Figures~\ref{fig:auc_plots} (right) and~\ref{fig:percondition} (bottom) that our method compares favorably to baselines.
 
-  \item Compute $R_{full}$ given $Y_{test} G^T$ and $X_{test}$
-  \item For all $i \in d_x$:
-  \begin{itemize}
-    \item Compute $R^i_{knockout}$ given $Y_{test} G^T$ and $X_{test} K {H^i_k}^T$
-    \item Compute $R^i_{delta}=R_{full} - R^i_{knockout}$
-  \end{itemize}
-  \end{enumerate}
-\fi
+%
+% \iffalse
+% For CCA and PLS, feature importance is assessed as follow:
+% \begin{enumerate}
+%   \item Fit $H$ and $G$ given $X_{train}$ and $Y_{train}$
+%   \item For each feature $i$:
+%   \begin{itemize}
+%     \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
+%     \item Fit $H^i_k$ and $G_k$ given $X_{train} K$ and $Y_{train}$
+%   \end{itemize}
+% \end{enumerate}
+% \fi
 
-\iffalse Ridge regression \citep{hoerl1959optimum}, multitask Lasso
-\citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls,
-tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and
-Reduced Rank Regression or RRR \citep{Izenman_rrr}.
+For the Backward Model, feature importance cannot be assessed because there is no prediction.
+% We thus simply
+% report the model performance $R_i = corr(YG_i, X_i)$.
+
+% \iffalse
+% \begin{enumerate}
 %
-Each of these methods estimates a matrix of coefficients $\hat{W} \in
-\mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$
-using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
+%   \item Fit $G$ given $X_{train}$ and $Y_{train}$
+%   \item Fit $H$ given $X_{train}$ and $Y_{train} G^T$
+%   \item For each feature $i$:
+%   \begin{itemize}
+%     \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
+%     \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train} G^T$
+%   \end{itemize}
 %
-For B2B, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
+%   \item Compute $R_{full}$ given $Y_{test} G^T$ and $X_{test}$
+%   \item For all $i \in d_x$:
+%   \begin{itemize}
+%     \item Compute $R^i_{knockout}$ given $Y_{test} G^T$ and $X_{test} K {H^i_k}^T$
+%     \item Compute $R^i_{delta}=R_{full} - R^i_{knockout}$
+%   \end{itemize}
+%   \end{enumerate}
+% \fi
+
+% \iffalse Ridge regression \citep{hoerl1959optimum}, multitask Lasso
+% \citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls,
+% tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and
+% Reduced Rank Regression or RRR \citep{Izenman_rrr}.
+% %
+% Each of these methods estimates a matrix of coefficients $\hat{W} \in
+% \mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$
+% using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
+% %
+% For B2B, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
+% %
 %
-
-We evaluate seven metrics:
 %
-test error in-distribution, test error out-distribution, test error-in
-distribution after weighting with $\hat{E}$, test error out-distribution after
-weighting with $\hat{E}$, Sonquist-Morgan false positives on $\hat{E}$,
-Sonquist-Morgan false negatives on $\hat{E}$, and the area under the ROC curve
-(AUC) between $\hat{E}$ and $E$.
+% We evaluate seven metrics:
+% %
+% test error in-distribution, test error out-distribution, test error-in
+% distribution after weighting with $\hat{E}$, test error out-distribution after
+% weighting with $\hat{E}$, Sonquist-Morgan false positives on $\hat{E}$,
+% Sonquist-Morgan false negatives on $\hat{E}$, and the area under the ROC curve
+% (AUC) between $\hat{E}$ and $E$.
+% %
+% The ``in-distribution'' metrics measure the test error of each predictor before
+% and after weighting their features by the estimated (continuous!) mask
+% $\hat{E}$.
+% %
+% The ``out-distribution'' metrics are similar, measuring the test error at
+% samples drawn using two different random covariance matrices for both $X$ and
+% $N$.
+% %
+% The false positives, false negatives, and AUC statistics measure the quality of
+% identification of causal factors between $\hat{E}$ and the true, binary $E$.
 %
-The ``in-distribution'' metrics measure the test error of each predictor before
-and after weighting their features by the estimated (continuous!) mask
-$\hat{E}$.
-%
-The ``out-distribution'' metrics are similar, measuring the test error at
-samples drawn using two different random covariance matrices for both $X$ and
-$N$.
-%
-The false positives, false negatives, and AUC statistics measure the quality of
-identification of causal factors between $\hat{E}$ and the true, binary $E$.
-
-\begin{figure}[htpb] \centering
-\includegraphics[width=\textwidth]{synthetic.pdf} \caption{Results of synthetic
-experiments. Each dot depicts the average value of a metric for B2B ($x$-axis)
-against a competitor ($y$-axis) for each experiment configuration averaged over
-$20$ random seeds. The prediction error on $Y$ within (in) and outside (out) of
-the training distribution are weakly but consistently larger for all of the
-models as compared to B2B with and without the Sonquist binarization (E). The
-AUC, measuring the ability of each model to separate causal from noncausal
-factors, is consistently larger in B2B as compared to the other models.}
-\label{table:synthetic} \end{figure} \fi
+% \begin{figure}[htpb] \centering
+% \includegraphics[width=\textwidth]{synthetic.pdf} \caption{Results of synthetic
+% experiments. Each dot depicts the average value of a metric for B2B ($x$-axis)
+% against a competitor ($y$-axis) for each experiment configuration averaged over
+% $20$ random seeds. The prediction error on $Y$ within (in) and outside (out) of
+% the training distribution are weakly but consistently larger for all of the
+% models as compared to B2B with and without the Sonquist binarization (E). The
+% AUC, measuring the ability of each model to separate causal from noncausal
+% factors, is consistently larger in B2B as compared to the other models.}
+% \label{table:synthetic} \end{figure} \fi
 

### 2019-09-25 paper/paper.tex
modifs in synthetic
synthetic
pushing things in appendix

@@
 \usepackage{iclr2020_conference,times}
+\DeclareMathOperator*{\argmax}{argmax} % thin space, limits underneath in displays
 
 \newtheorem{theorem}{Theorem}
+\newtheorem*{theorem*}{Theorem}
 
@@
 \end{theorem}
+
 \begin{proof}
+  See Appendix \ref{appendix:theorem_proof}.
+\end{proof}
 
- Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B, we have
- \begin{align*}
-    \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
-%                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
-                                                        &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
-     \label{eq:doublenorm}
-\end{align*}
-\begin{align*}
-    \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
-    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
-%    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
-    &= EF \hat G
-\end{align*}
-
-Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero.
+% \begin{proof}
+%
+%  Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B, we have
+%  \begin{align*}
+%     \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
+% %                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
+%                                                         &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
+%      \label{eq:doublenorm}
+% \end{align*}
+% \begin{align*}
+%     \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
+%     &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
+% %    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
+%     &= EF \hat G
+% \end{align*}
+%
+% Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero.
 Let $Z=F^\dagger EF\hat G$.
@@
 Therefore,
- $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
-
-But as $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, the above inequality is an equality, $Z=\hat G$ and $EF\hat G = F\Hat G$.
+%  $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
+%
+% But as $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, the above inequality is an equality, $Z=\hat G$ and $EF\hat G = F\Hat G$.
 Therefore, $\left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 = \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 = \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2$ is minimised by $\hat G$ and $\hat H$.
@@
 This completes the proof.
-\end{proof}
+% \end{proof}
 
@@
 
+\input{appendix}
+
 \end{document}

### 2019-09-25 paper/appendix.tex

@@
 \section{Appendix}
+
+\subsection{Proof of consistency theorem}
+\label{appendix:theorem_proof}
+
+Proof of the theorem in \ref{sec:theorem}:
+
+\begin{theorem*}[B2B consistency - general case]
+
+     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full rank noise.
+     %
+     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimize
+     %
+     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $E\hat H = \hat H$
+\end{theorem*}
+
+\begin{proof}
+
+ Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B, we have
+ \begin{align*}
+    \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
+%                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
+                                                        &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
+     \label{eq:doublenorm}
+\end{align*}
+\begin{align*}
+    \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
+    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
+%    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
+    &= EF \hat G
+\end{align*}
+
+Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero.
+Let $Z=F^\dagger EF\hat G$.
+We have $FZ = FF^\dagger EF  \hat G= EF\hat G$ ($FF^\dagger E =E$ since $F$ is full rank on $Img(E)$).
+Since E is a contraction, we have $ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$.
+Therefore,
+ $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
+
+But as $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, the above inequality is an equality, $Z=\hat G$ and $EF\hat G = F\Hat G$.
+Therefore, $\left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 = \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 = \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2$ is minimised by $\hat G$ and $\hat H$.
+Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$.
+This completes the proof.
+\end{proof}
+
+\newpage
 \subsection{Feature importance}
+\label{appendix:feature_importance}
 

### 2019-09-25 paper/meg_exps.tex

@@
 \begin{figure}[t!]
-  \centering
+  \begin{minipage}[c]{0.6\textwidth}
     \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_evoked.pdf}
-  \caption{102 subjects read $\approx$ 2,700 words while their brain activity was recorded with MEG.
+  \end{minipage}\hfill
+  \begin{minipage}[c]{0.4\textwidth}
+    \caption{Ninety subjects read approximately 2,700 words while their brain activity was recorded with MEG.
 Top.
-Average brain response to words relative to their onset (gradiometers), as viewed from above the head.
+Average brain response to words (word onset at t=0 ms), as viewed from above the head (red= higher gradient of magnetic flux).
 Bottom.
-Each line represents a MEG sensor (magnetometer), color-coded by spatial position.
+Each line represents magnetometer, color-coded by spatial position.
 Posterior responses, typical of primary visual cortex activity, peak around 100 ms after word onset and are followed by an anterior propagation of activity typical of semantic processing in the associative cortices.
-Vertical lines indicate moments of above topographies.
     }
     \label{fig:meg_evoked}
+  \end{minipage}
 \end{figure}
@@
 
-When human subjects read words, the number of letters is known to modulate
-brain activity from 100ms after word onset \citep{pegado2014timing}.
-Similarly, word frequency is known to modulate brain activity around 400ms
-\citep{kutas2011thirty}. In the present study, we consider these two well-known features, as well as an untested feature ('Word Function') and dummy variable.
+We are here interested in identifying the features of word that modulate (i.e. cause) brain activity. We consider four distinct, but colinear features.
 %
-"Word Length' refers to the total number of letters, and is expected to specifically cause a variation in the early evoked MEG responses (i.e.
-from 100 ms after stimulus onset) elicited by the retinotopically-tuned visual cortices.
+"Word Length' refers to the total number of letters.
+Word Length is expected to specifically cause a variation in the early evoked MEG responses (i.e.
+from 100 ms after stimulus onset) elicited by the retinotopically-tuned visual cortices (e.g.
+\citep{pegado2014timing}.).
 %
 'Word Frequency', as derived with WordFreq \citep{speerwordfreq}, indexes how frequently each word appears in Dutch (using the Zipf logarithmic scale of \citep{van2014subtlex}).
-This feature is expected to specifically cause a variation in the late evoked MEG responses (i.e.
+Word Frequency is expected to specifically cause a variation in the late evoked MEG responses (i.e.
 from 400 ms), because it variably engages semantic processes in the temporal cortices \citep{kutas2011thirty}.
 %
-Finally, 'Word Function' indicates whether each word is a content word (i.e.
+'Word Function' indicates whether each word is a content word (i.e.
 a noun, a verb, an adjective or an adverb) or a function word (i.e.
@@
 To our knowledge, this feature has not been thouroughly investigated with MEG.
-Its causal contribution to reading processes in the brain is thus largely unknown.
+Its causal contribution to reading processes in the brain thus remains unclear.
 %
-To verify that B2B and other models would not inadequately identify non-causal features, we added a dummy feature, constructed from a noisy combination of "Word Length" and "Word Frequency":
+Finally, to verify that B2B and other methods would not inadequately identify non-causal features, we added a dummy feature, constructed from a noisy combination of "Word Length" and "Word Frequency":
 \begin{equation}
@@
 \end{equation}
-where $z$ transforms features into zero-mean and unit variance vectors and $\mathcal{N}$ is random vectors sampling a zero-mean unit-variance Gaussian distribution.
+where $z$ transforms features into zero-mean and unit variance vectors and $\mathcal{N}$ is a random vector sampling a zero-mean unit-variance Gaussian distribution.
 
@@
 
-We compare B2B to four standard models: forward regression, backward regression, CCA and PLS, as implemented in scikit-learn, and optimized with nested cross validation over twenty l2 regularization parameters logarithmically spaced between $10^-4$ and $10^4$ or 1 to 4 caconical components.
+We compare B2B to four standard methods: Forward regression, Backward regression, CCA and PLS, as implemented in scikit-learn \citep{scikit}, and optimized with nested cross validation over twenty l2 regularization parameters logarithmically spaced between $10^-4$ and $10^4$ (for regression methods) or 1 to 4 canonical components (for cross-decomposition methods).
 
-We used the feature importance described in \ref{algorithm:b2b_fi} to assess the extent to which each feature $X_i$ specifically improves the prediction of held-out $Y$ data.
+We used the feature importance described in Algorithm \ref{algorithm:b2b_fi} to assess the extent to which each feature $X_i$ specifically improves the prediction of held-out $Y$ data.
 
-Each model was implemented for each subject and each time sample independently. Pairwise comparison across models were performed using a wilcoxon test on the $\Delta R$ averaged across time.
+Each model was implemented for each subject and each time sample independently.
+Pairwise comparison between models were performed using a Wilcoxon test across subjects (n=90) using the average $\Delta R$ across time.
 
-Corresponding results can be found in Figure \ref{fig:meg_results}. Significant pairwise differences across models are highlighted with the top horizontal bars in \ref{fig:meg_results}.
+The chapter corresponding results can be found in Figure \ref{fig:meg_results}.
+Significant ($p<0.05$) pairwise differences across models are highlighted with the top horizontal bars in \ref{fig:meg_results}.
 
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg.pdf}
-  \caption{Forward regression (Blue), PLS (orange), CCA (green) and B2B (red) are compared on their ability to reliably predict single-trial MEG signals evoked by words.
+\begin{wrapfigure}{r}{.5\textwidth}
+  \begin{center}
+    \includegraphics[width=0.48\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg.pdf}
+
+    \label{fig:meg_results}
+  \end{center}
+  \caption{Multiple models (color-coded) are compared on their ability to reliably predict single-trial MEG signals evoked by words.
 Left.
-Average improvement of correlation coefficient $\Delta R$ obtained on held-out data when each of the four features "Word Length", "Word Frequency", "Word Function" and "Dummy Variable" was provided to the model \ref{algorithm:b2b_fi}.
+Average improvement of correlation coefficient $\Delta R$ for each pf the four features (rows).
 Error bars indicate standard error of the mean (SEM) across subjects.
 Right.
-Average $\Delta R$ across time.
-Each dot represents a subject.
-The top horizontal lines indicate significant differences between pairwise comparison across models, color coded by the winning model.
-B2B outperforms all models on meaningful features, at the exception of CCA for the "word length".}
-  \label{fig:meg_results}
-\end{figure}
+Average $\Delta R$ across time for each subject (dot).
+Top horizontal lines indicate when B2B significantly outperform other methods.}
+\end{wrapfigure}
 
@@
 
-First, \ref{fig:meg_supp} shows that the Backward model decodes the dummy variable well above chance level.
-In addition, "Word Length" and  "Word Frequency" present a similar decoding time course, even though these features are known to influence early and late MEG responses respectively \citep{kutas2011thirty}.
+Supplementary Figure \ref{fig:meg_supp} shows that the Backward model decodes the dummy variable well above chance level.
+In addition, "Word Length" and "Word Frequency" present a similar decoding time course, even though these features are known to specifically influence early and late MEG responses respectively \citep{kutas2011thirty}.
 These results illustrate the well-known fact that backward modelling cannot be used to estimate the causal contribution of collinear features.
 
-We thus focus on valid modeling (Forward Regression, PLS, CCA, and B2B) and estimate the change of correlation coefficients $\Delta R$ induced by the introduction of each feature into the model \ref{algorithm:b2b_fi}.
+We thus focus on the valid methods (i.e.
+Forward Regression, PLS, CCA, and B2B) and estimate the change of correlation coefficients $\Delta R$ induced by the introduction of each feature into the model \ref{algorithm:b2b_fi}.
 As expected, none of the models significantly predicted the 'Dummy Variable' to be a causal factors (i.e.
@@
 
-The results show that all models significantly detect that both word length and
-word frequency modulate brain responses after word onset. However,
-forward regression detected word length ($t$-value$=10.6$, $p=10^{-16}$) and
-word frequency ($t$-value$=11.1$, $p=10^-17$) less reliably than B2B (length:
-$t$-value$=18.7$, $p=10^{-31}$, frequency: $t$-value$=15.9$, $p=10^{-26}$). B2B
-was less sensitive than the backward regression (length: $t$-value$=33.6$,
-$p=10^{-50}$, frequency: $t$-value$=31.7$, $p=10^{-48}$).
-However, as backward models confound collinear factors, this high sensitivity likely comes from a loss of specificity.
-This possibility is corroborated by the fact that the effects of word length and word frequency identified with backward regression present largely similar dynamics.
+On average across time, "Word Length" and "Word Frequency" improved the prediction performance of all methods $\Delta R>0$ (all $p<0.0001$).
+As expected, the time course of "Word Length" and "Word Frequency" rose from $\approx$ 100 ms and from $\approx$ 400 ms respectively.
+Furthermore, 'Word Function' improved the prediction performance of all models (all $p < 0.0002$) except PLS ($p=0.7989$).
+Overall, these results confirm that "Word Length", "Word Frequency" and "Word Function" each causally influence specific periods of brain responses to words.
 
-Finally we investigate in Figure \ref{fig:megresult} whether B2B was sufficiently sensitive to reliably detect a multiplicity of subtle features, such as individual letters and part-of-speech.
-The vast majority of the visual and lexical features appeared to be reliably detected (see Figure 4 in Supplementary Material).
-B2B thus promises to improve the detectability of subtle effects.
+To assess which model would be more sensitive to these causal discoveries, we performed pairwise comparisons between models and across subjects (Figure \ref{fig:meg} right).
+For "Word Length" B2B outperforms all models (all $p < 0.00001$) but CCA ($p=0.0678$).
+For "Word Frequency", B2B outperforms all models (all $p < 0.0006$).
+For "Word Function", B2B outperforms all models (all $p < 0.0015$).
+Overall, these results suggest that B2B reliably outperform standard methods, especially when the effects are difficult to detect.

### 2019-09-25 paper/appendix.tex
supp
pushing things in appendix

@@
 
+For B2B, feature importance is assessed as follows:
+
+\begin{algorithm}[H]
+    %\SetAlgoLined
+    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
+\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
+\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
+$\Delta{R} \in \mathbb{D}^{d_x}$.} $H, G = \text{B2B}(X_{train}, Y_{train})$\;
+$R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;
+
+    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $H_k =
+\text{LinearRegression}(X_{train} K, Y_{train} G_i)$\; $R_{k} =
+\text{corr}(X_{test} H_k, Y_{test} G_i)$\; $\Delta R_i = R_{full} - R_{k}$\; }
+\Return{$\Delta R$} \caption{B2B feature importance.} \label{algorithm:b2b_fi}
+\end{algorithm}
+
 For the Forward Model, the feature importance is assessed as follows:
@@
 For the Backward Model, feature importance cannot be assessed because there is no prediction.
+
+\section{Additional Figures}
+
+\begin{figure}[t!]
+  \centering
+  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_supp.pdf}
+  \caption{}
+  \label{fig:meg_supp}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/AUC_R.png}
+  \caption{Synthetic experiments.
+Distribution (over conditions) of differences in AUC (left) and Feature Importance $\Delta R$ (right) metrics between our method and the baselines.
+Mass under the dashed line corresponds to B2B outperforming the baseline.
+\label{fig:auc_plots}}
+\end{figure}

### 2019-09-25 paper/meg_exps.tex
meg correction

@@
 segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
-thus led a sampling frequency of 240Hz. The average responses across words is displayed in \ref{fig:meg_evoked}
+thus led a sampling frequency of 240Hz. The average responses across words is displayed in Figure \ref{fig:meg_evoked}.
 For each subject and each time sample relative to word onset, we
@@
 
-We are here interested in identifying the features of word that modulate (i.e. cause) brain activity. We consider four distinct, but colinear features.
+We aim to identify the word features that cause a variation in brain responses. We consider four distinct but colinear features.
 %
-"Word Length' refers to the total number of letters.
+First, 'Word Length' refers to the total number of letters.
 Word Length is expected to specifically cause a variation in the early evoked MEG responses (i.e.
@@
 %
-'Word Frequency', as derived with WordFreq \citep{speerwordfreq}, indexes how frequently each word appears in Dutch (using the Zipf logarithmic scale of \citep{van2014subtlex}).
+Second, 'Word Frequency' indexes how frequently each word appears in Dutch and was derived with the the Zipf logarithmic scale of \citep{van2014subtlex} provided by the WordFreq package \citep{speerwordfreq}.
 Word Frequency is expected to specifically cause a variation in the late evoked MEG responses (i.e.
@@
 %
-'Word Function' indicates whether each word is a content word (i.e.
+Third, 'Word Function' indicates whether each word is a content word (i.e.
 a noun, a verb, an adjective or an adverb) or a function word (i.e.
@@
 %
-Finally, to verify that B2B and other methods would not inadequately identify non-causal features, we added a dummy feature, constructed from a noisy combination of "Word Length" and "Word Frequency":
-\begin{equation}
-  dummy = z(length) + z(frequency) + \mathcal{N}
-\end{equation}
-where $z$ transforms features into zero-mean and unit variance vectors and $\mathcal{N}$ is a random vector sampling a zero-mean unit-variance Gaussian distribution.
+Finally, to verify that B2B and other methods would not inadequately identify non-causal features, we added a dummy feature, constructed from a noisy combination of Word Length and Word Frequency:
+$dummy = z(length) + z(frequency) + \mathcal{N}$, where $z$ normalizes features and $\mathcal{N}$ is a random vector sampling Gaussian distribution (all terms thus have a zero-mean and a unit-variance).
 
@@
 
-We compare B2B to four standard methods: Forward regression, Backward regression, CCA and PLS, as implemented in scikit-learn \citep{scikit}, and optimized with nested cross validation over twenty l2 regularization parameters logarithmically spaced between $10^-4$ and $10^4$ (for regression methods) or 1 to 4 canonical components (for cross-decomposition methods).
+We compare B2B to four standard methods: Forward regression, Backward regression, CCA and PLS, as implemented in scikit-learn \citep{scikit}, and optimized with nested cross validation over twenty $l2$ regularization parameters logarithmically spaced between $10^{-4}$ and $10^4$ (for regression methods) or 1 to 4 canonical components (for cross-decomposition methods).
 
-We used the feature importance described in Algorithm \ref{algorithm:b2b_fi} to assess the extent to which each feature $X_i$ specifically improves the prediction of held-out $Y$ data.
+We used the feature importance described in Algorithm \ref{algorithm:b2b_fi} to assess the extent to which each feature $X_i$ specifically improves the prediction of held-out $Y$ data, using a 5-fold cross-validation.
 
@@
 
-The chapter corresponding results can be found in Figure \ref{fig:meg_results}.
-Significant ($p<0.05$) pairwise differences across models are highlighted with the top horizontal bars in \ref{fig:meg_results}.
+The corresponding results can be found in Figure \ref{fig:meg_results}.
 
@@
 Supplementary Figure \ref{fig:meg_supp} shows that the Backward model decodes the dummy variable well above chance level.
-In addition, "Word Length" and "Word Frequency" present a similar decoding time course, even though these features are known to specifically influence early and late MEG responses respectively \citep{kutas2011thirty}.
-These results illustrate the well-known fact that backward modelling cannot be used to estimate the causal contribution of collinear features.
+In addition, the Backward model reveal a similar decoding time course for Word Length and Word Frequency, even though these features are known to specifically influence early and late MEG responses respectively \citep{kutas2011thirty}.
+These results illustrate that backward modelling cannot be used to estimate the causal contribution of collinear features.
 
-We thus focus on the valid methods (i.e.
-Forward Regression, PLS, CCA, and B2B) and estimate the change of correlation coefficients $\Delta R$ induced by the introduction of each feature into the model \ref{algorithm:b2b_fi}.
-As expected, none of the models significantly predicted the 'Dummy Variable' to be a causal factors (i.e.
-$\Delta R < 0$ for all models, all $p > .089$)
+We thus focus on the four remaining methods (i.e.
+Forward Regression, PLS, CCA, and B2B) and estimate their $\Delta R$ (i.e.
+the improvement of Y prediction induced by the introduction of a given feature into the model \ref{algorithm:b2b_fi}).
+Contrary to the Backward Model, none of the models predicted the Dummy Variable to improve the $Y$ prediction: all $\Delta R < 0$ (all $p > .089$).
 
@@
 
-On average across time, "Word Length" and "Word Frequency" improved the prediction performance of all methods $\Delta R>0$ (all $p<0.0001$).
-As expected, the time course of "Word Length" and "Word Frequency" rose from $\approx$ 100 ms and from $\approx$ 400 ms respectively.
-Furthermore, 'Word Function' improved the prediction performance of all models (all $p < 0.0002$) except PLS ($p=0.7989$).
-Overall, these results confirm that "Word Length", "Word Frequency" and "Word Function" each causally influence specific periods of brain responses to words.
+Word Length and Word Frequency improved the prediction performance of all methods: $\Delta R>0$ for all models (all $p<0.0001$).
+As expected, the time course associated with Word Length and Word Frequency rose from $\approx$ 100 ms and from $\approx$ 400 ms respectively.
+Furthermore, Word Function improved the prediction performance of all models (all $p < 0.0002$) except PLS ($p=0.7989$).
+Overall, these results confirm that Word Length, Word Frequency and Word Function causally influence specific periods of brain responses to words.
 
-To assess which model would be more sensitive to these causal discoveries, we performed pairwise comparisons between models and across subjects (Figure \ref{fig:meg} right).
-For "Word Length" B2B outperforms all models (all $p < 0.00001$) but CCA ($p=0.0678$).
-For "Word Frequency", B2B outperforms all models (all $p < 0.0006$).
+To assess which model would be most sensitive to these causal discoveries, we compared B2B to other models across subjects (Figure \ref{fig:meg} right).
+For Word Length B2B outperforms all models (all $p < 0.00001$) but CCA ($p=0.0678$).
+For Word Frequency, B2B outperforms all models (all $p < 0.0006$).
 For "Word Function", B2B outperforms all models (all $p < 0.0015$).
-Overall, these results suggest that B2B reliably outperform standard methods, especially when the effects are difficult to detect.
+Overall, these results show that B2B reliably outperform standard methods, especially when the effects are difficult to detect.

### 2019-09-25 paper/synthetic.tex

@@
 we take the sum of square coefficients:
-
-\begin{equation} \hat E_i = \sum_j {H^j_i}^2 \end{equation}
+% \begin{equation}
+  $\hat E_i = \sum_j {H^j_i}^2 $
+% \end{equation}
 
@@
 %\hat{E}_{k-1}) / 2 \end{equation}
-
 The AUC allows evaluating the capacity of models at detecting the causal
@@
 
-We report AUC results in Figures~\ref{fig:auc_plots}~(left) and ~\ref{fig:percondition}~(top), and compare favorably to all baselines.
+We report AUC results in Figures~\ref{fig:auc_plots}~(left) and ~\ref{fig:percondition}~(top, in Appendix), and compare favorably to all baselines.
 
-\begin{figure}[ht]
+\begin{figure}[t]
   \centering
-  \includegraphics[width=\linewidth]{figures/AUC_R.png}
-  \vspace{-4ex}
-  \caption{Synthetic experiments.
-Distribution (over conditions) of differences in AUC (left) and Feature Importance $\Delta R$ (right) metrics between our method and the baselines.
-Mass under the dashed line corresponds to B2B outperforming the baseline.
-\label{fig:auc_plots}}
-  \vspace{2ex}
   \includegraphics[width=\linewidth]{figures/AUC_conditions}
@@
   \vspace{-4ex}
-  \caption{Synthetic experiments. Average AUC (top) and Feature Importance $\Delta R$ (bottom) metrics as a function of experimental conditions. Higher is better. \label{fig:percondition}}
+  \caption{Synthetic experiments.
+Average AUC (top) and Feature Importance $\Delta R$ (bottom) when varying experimental conditions individually.
+Higher is better.
+B2B compares favorably in all cases.
+\label{fig:percondition}}
 \end{figure}
@@
 
-For B2B, feature importance is assessed as follows:
-
-\begin{algorithm}[H]
-    %\SetAlgoLined
-    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
-\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
-\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
-$\Delta{R} \in \mathbb{D}^{d_x}$.} $H, G = \text{B2B}(X_{train}, Y_{train})$\;
-$R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;
-
-    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $H_k =
-\text{LinearRegression}(X_{train} K, Y_{train} G_i)$\; $R_{k} =
-\text{corr}(X_{test} H_k, Y_{test} G_i)$\; $\Delta R_i = R_{full} - R_{k}$\; }
-\Return{$\Delta R$} \caption{B2B feature importance.} \label{algorithm:b2b_fi}
-\end{algorithm}
-
-Feature importance can be computed similarly for the CCA, PLS and Forward baselines, and we provide similar algorithm boxes in the appendix.
+We show in Appendix~\ref{appendix:feature_importance} pseudo-code to assess feature importance for our algorithm as well as baselines.
 For the Backward Model, feature importance cannot be assessed as there is no prediction.
 
-We show in Figures~\ref{fig:auc_plots} (right) and~\ref{fig:percondition} (bottom) that our method compares favorably to baselines.
+% Feature importance can be computed similarly for the CCA, PLS and Forward baselines (see Appendix \ref{appendix:feature_importance}).
+For the Backward Model, feature importance cannot be assessed as there is no prediction.
+
+We show in Figures~\ref{fig:percondition} (bottom) and \ref{fig:auc_plots} (right, in Appendix) that our method compares favorably to baselines.
 
@@
 % \fi
-
-For the Backward Model, feature importance cannot be assessed because there is no prediction.
+% For the Backward Model, feature importance cannot be assessed because there is no prediction.
 % We thus simply
@@
 
-Figure~\ref{table:synthetic} summarizes the results for all experimental
-conditions, metrics, and methods.
-%
-B2B is the method obtaining best test errors, lowest false positive and false
-negative discovery about $E$, and overall the best AUC when it comes to
-detecting causal influences.
+% Figure~\ref{table:synthetic} summarizes the results for all experimental
+% conditions, metrics, and methods.
+% %
+% B2B is the method obtaining best test errors, lowest false positive and false
+% negative discovery about $E$, and overall the best AUC when it comes to
+% detecting causal influences.

### 2019-09-25 paper/paper.tex
wrapfig
wip

@@
 \usepackage[ruled, linesnumbered]{algorithm2e}
-\usepackage{iclr2020_conference,times}
+\usepackage{iclr2020_conference,times,wrapfig}
+
 \DeclareMathOperator*{\argmax}{argmax} % thin space, limits underneath in displays
@@
 \end{theorem}
-
+%
 \begin{proof}
@@
 \end{proof}
-
+%
 % \begin{proof}
@@
 % \end{proof}
-
-So, $\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top$.
+%
+\begin{equation}
+  \text{So, }\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top.
+\end{equation}
+% So, $\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top$.
 
@@
 
-$X^\top X = \left(\begin{array}{lccl}\Sigma_{X1} & \Sigma_{X2} \\ \Sigma_{X2} & \Sigma_{X3}\end{array}\right)$, $N^\top N = \left(\begin{array}{lccl}\Sigma_{N1} & \Sigma_{N2} \\ \Sigma_{N2} & \Sigma_{N3}\end{array}\right)$ and
+% $$
+\begin{equation}
+  X^\top X = \left(\begin{array}{lccl}\Sigma_{X1} & \Sigma_{X2} \\ \Sigma_{X2} & \Sigma_{X3}\end{array}\right),\qquad\qquad N^\top N = \left(\begin{array}{lccl}\Sigma_{N1} & \Sigma_{N2} \\ \Sigma_{N2} & \Sigma_{N3}\end{array}\right),
+\end{equation}
+% $$
+% and
+\begin{equation}
+  \hat H = \left(\begin{array}{cc}(\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1} & (\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X2} \\0 & 0\end{array}\right),
+\end{equation}
+\begin{equation}
+  Diag_k (\hat H) = Diag((\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1}) = Diag((I+\Sigma_{X1}^{-1}\Sigma_{N1})^{-1}).
+\end{equation}
+%
+% $$
+% $$
+% $$
+% $$
+%
+In the absence of noise, we have $Diag(\hat H) = Diag(E)$.
+Otherwise the $k$ first diagonal elements of $\hat H$ are all positive, and bounded by $\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}}$ and $\frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$, where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X1}$ and $\Sigma_{N1}$.
 
-$\hat H = \left(\begin{array}{cc}(\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1} & (\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X2} \\0 & 0\end{array}\right)$
-
-$Diag_k (\hat H) = Diag((\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1}) = Diag((I+\Sigma_{X1}^{-1}\Sigma_{N1})^{-1})$
-
-In the absence of noise, we have $Diag(\hat H) = Diag(E)$.
-Else the k first diagonal elements of $\hat H$ are all positive, and bounded by $\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}}$ and $\frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$, where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X1}$ and $\Sigma_{N1}$.
-
-The average value of the coefficients of $Diag(\hat H)$ is the trace of $\hat H$ divided by k.
-On average, since $X$ and $N$ are not correlated by definition, $\Sigma_{X1}^{-1}\Sigma_{N1}$ should have a low condition numbers, and $\frac{Var(X)}{Var(X)+Var(N)}$ provide a good estimate of $\frac{1}{k}Tr(\hat H)$.
+The average value of the coefficients of $Diag(\hat H)$ is the trace of $\hat H$ divided by $k$.
+On average, since $X$ and $N$ are not correlated by definition, $\Sigma_{X1}^{-1}\Sigma_{N1}$ should have a low condition number, and $\frac{Var(X)}{Var(X)+Var(N)}$ provides a good estimate of $\frac{1}{k}Tr(\hat H)$.
 

### 2019-09-25 paper/paper.tex
a

--- 2019-09-25 paper/paper.tex
@@
 %
-To accelerate this ensembling procedure, we implemented an efficient leave-one-out cross-validation scheme as detailed in \citep{RifkinLippert2007} % http://cbcl.mit.edu/publications/ps/MIT-CSAIL-TR-2007-025.pdf
-, as follow:
+To accelerate this ensembling procedure, we implemented an efficient leave-one-out cross-validation scheme as detailed in \citep{rifkin2007notes} % http://cbcl.mit.edu/publications/ps/MIT-CSAIL-TR-2007-025.pdf
+, as follows:
 %
@@
   \Sigma_X = Q V Q^T \\
-  G = Q (V + \alpha*I)^{-1} Q^T\\
+  G = Q (V + \lambda*I)^{-1} Q^T\\
   \end{aligned}
 \end{equation}
-where $Q$, $V$ and $\alpha$ are the eigen vectors, eigen values and regularization, respectively.
+where $Q$, $V$ and $\lambda$ are the eigen vectors, eigen values and regularization, respectively.
 
@@
 
-In the context of neuroimaging, the first regression of B2B relates to the first step of Representational Similarity Analysis (RSA) \citep{kriegeskorte2008representational}.
-Indeed, RSA analyzes the similarity of brain responses associated with specific categorical conditions (e.g.
+B2B present multiple major advantages as compared to other
+B2B provides a single scalar values $\hat E$ tending towards 1
+
+A variety of statistical methods applied to neuroimaging data have been proposed to clarify what is being represented in brain activity - i.e.
+what feature cause specifically cause brain activity.
+One of the popular method is Representational Similarity Analysis (RSA) \citep{kriegeskorte2008representational}, and consists in analyzing the similarity of brain responses associated with specific categorical conditions (e.g.
 distinct images), by (1) fitting one-against-all classifiers on each condition and (2) testing whether these classifiers can discriminate all other conditions.
 The resulting confusion matrix can then be analyzed in an unsupervised manner to reveal that some conditions lead to similar brain activity patterns.
-B2B differs from RSA in that (1) it uses regressions instead of classifications, and can thus generalize to new contexts and (2)
-it is fully supervised.
+B2B differs from RSA in that (1) it uses regressions instead of classifications, and can thus generalize to new items and new contexts and (2) it is fully supervised.
 
-% THIS NEEDS TO BE UPDATED: see de chevigner's latest paper
-More generally, B2B is related to CCA \citep{cca_hotelling}, but departs from it in four main aspects.
-First, CCA performs an additional eigenvector decomposition of what would be the $H$ coefficients in B2B \eqref{eq:solH}.
-Second, CCA does not typically optimize two distinct regularization parameters.
-Third, CCA does not use bagging between $G$ and $H$.
-Finally, CCA does not diagonalize nor binarize the $H$ matrix to estimate $\hat E$.
-More generally, CCA aims to find maximally correlated components between $X$ and $Y$, and does not assume any causal direction.
-By contrast, B2B assumes that $X$ causes $Y$.
-Together, these differences may explain why B2B can reliably outperform CCA on estimating causal influences and predicting $Y$ in novel environments (Fig.
+Furthermore, B2B closely relates to CCA \citep{cca_hotelling}, but departs from it in four main aspects.
+First, B2B is not symmetric between $X$ and $Y$: it aims to identify specific causal features by first optimizing over the decoders $G$ and then over $H$.
+By contrast, CCA is symmetric between $X$ and $Y$, and aims to find $G$ and $H$ such that they project $X$ and $Y$ on maximally correlated dimensions.
+Second, CCA is based a eigenvector decomposition of $XH$ and $YG$ - the corresponding canonical components are thus mixing the $X$ features in way that limit interpretability.
+Third, CCA does not typically optimize two distinct regularization parameters.
+Finally, CCA does not use different data splits to estimate $G$ and $H$.
+Together, these differences may explain why B2B reliably outperform CCA on estimating causal influences (Fig.
 \ref{table:synthetic}).
+Extensions of CCA have been explored in neuroimaging (see \citep{de2019multiway} for brief review), but typically do not address the issue of causal detection.
 

### 2019-09-25 paper/paper.tex
a

@@
 
-B2B present multiple major advantages as compared to other
-B2B provides a single scalar values $\hat E$ tending towards 1
+Forward and cross-decomposition (CCA, PLS) models are regularly used to identify the causal contribution of collinear features onto multi-dimensional observations (e.g.
+\citep{naselaris2011encoding}).
+These approaches typically lead to multiple coefficients for each features (i.e.
+one per dimension of $Y$ or one per component respectively).
+Furthermore, these  coefficients can be difficult to summarize into a single 'causal' estimation.
+By contrast, B2B quick (Fig.
+\ref{fig:duration}) leads to a single unbiased scalar values $\hat E$ tending towards 1 and 0 for causal and non-causal features.
 
-A variety of statistical methods applied to neuroimaging data have been proposed to clarify what is being represented in brain activity - i.e.
-what feature cause specifically cause brain activity.
-One of the popular method is Representational Similarity Analysis (RSA) \citep{kriegeskorte2008representational}, and consists in analyzing the similarity of brain responses associated with specific categorical conditions (e.g.
+A variety of other statistical methods applied to neuroimaging data have been proposed to clarify what is being represented in brain responses - i.e.
+what feature specifically cause brain activity.
+One of the popular linear method is Representational Similarity Analysis (RSA) \citep{kriegeskorte2008representational}, and consists in analyzing the similarity of brain responses associated with specific categorical conditions (e.g.
 distinct images), by (1) fitting one-against-all classifiers on each condition and (2) testing whether these classifiers can discriminate all other conditions.
-The resulting confusion matrix can then be analyzed in an unsupervised manner to reveal that some conditions lead to similar brain activity patterns.
+The resulting confusion matrix is then analyzed in an unsupervised manner to reveal which conditions lead to similar brain activity patterns.
 B2B differs from RSA in that (1) it uses regressions instead of classifications, and can thus generalize to new items and new contexts and (2) it is fully supervised.
@@
 Furthermore, B2B closely relates to CCA \citep{cca_hotelling}, but departs from it in four main aspects.
-First, B2B is not symmetric between $X$ and $Y$: it aims to identify specific causal features by first optimizing over the decoders $G$ and then over $H$.
+First, B2B is not symmetric between $X$ and $Y$: it aims to identify specific causal features.
 By contrast, CCA is symmetric between $X$ and $Y$, and aims to find $G$ and $H$ such that they project $X$ and $Y$ on maximally correlated dimensions.
-Second, CCA is based a eigenvector decomposition of $XH$ and $YG$ - the corresponding canonical components are thus mixing the $X$ features in way that limit interpretability.
+Second, CCA is based an eigenvector decomposition of $XH$ and $YG$ - the corresponding canonical components are thus mixing the $X$ features in way that limit interpretability.
 Third, CCA does not typically optimize two distinct regularization parameters.
-Finally, CCA does not use different data splits to estimate $G$ and $H$.
+Finally, CCA does not use different data splits to estimate $G$ and $H$, and is thus prone to overfitting.
 Together, these differences may explain why B2B reliably outperform CCA on estimating causal influences (Fig.
@@
 
-B2B is also related to, but different from usual causal discovery algorithms \citep{peters2017elements} that hypothesize on the direction of causality, or from measures of causal influence requiring time information \citep{granger1969investigating, janzing2013quantifying}.
-%
-B2B seems to have a similar flavour to half-sibling regression \citep{scholkopf2016modeling}, in the sense that we observe a large number of variables corrupted by the same source of noise; we leave the study of the precise link between these two models for future work.
+Finally, B2B is also related to, but different from causal discovery algorithms used outside the neuroimaging field.
+These methods are often nonlinear \citep{peters2017elements} and base their causal estimates on temporal delays \citep{granger1969investigating, janzing2013quantifying}.
+For example, B2B relates to tthe half-sibling regression \citep{scholkopf2016modeling}, in the sense that we observe a large number of variables corrupted by the same source of noise; we leave the extension of B2B to nonlinear time-series for future work.
 

### 2019-09-25 paper/meg_exps.tex
moved figure with vspace

@@
 
-The corresponding results can be found in Figure \ref{fig:meg_results}.
+Corresponding results are shown in Figure~\ref{fig:meg_results}.
 
 \begin{wrapfigure}{r}{.5\textwidth}
+  \vspace{-12ex}
   \begin{center}
@@
 Top horizontal lines indicate when B2B significantly outperform other methods.}
+  \vspace{-9ex}
 \end{wrapfigure}
@@
 As expected, the time course associated with Word Length and Word Frequency rose from $\approx$ 100 ms and from $\approx$ 400 ms respectively.
-Furthermore, Word Function improved the prediction performance of all models (all $p < 0.0002$) except PLS ($p=0.7989$).
+Furthermore, Word Function improved the prediction performance of all models (all $p < 0.0002$) except for PLS~($p=0.7989$).
 Overall, these results confirm that Word Length, Word Frequency and Word Function causally influence specific periods of brain responses to words.

### 2019-09-25 paper/introduction.tex
wip

@@
 By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}.
-To combine the benefits of forward and backward modeling, several authors have proposed to use cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA) \citep{dechevigne2018CCA}.
-CCA and PLS aim to find, from two sets of data $X$ and $Y$, the components $v$ and $w$ were $Xw$ and $Yv$ are maximially correlated or maximally covarying respectively.
+To combine the benefits of forward and backward modeling, several authors have proposed to use cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA) \citep{de2019multiway}.
+CCA and PLS aim to find, from two sets of data $X$ and $Y$, the components $H$ and $G$ were $XH$ and $YG$ are maximially correlated or maximally covarying respectively.
 Because CCA and PLS are are based on a generalized eigen decomposition, their resulting coefficients are mixing the features of $X$ and $Y$ in a way that makes them notoriously difficult to interpret \citep{lebart1995statistique}.
 
-Here, we introduce the `back-to-back regression', which not only combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}), but also provide a fast, robust and interpretable causal coefficients.
+Here, we introduce the `back-to-back regression' (B2B), which not only combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}), but also provide robust and interpretable causal coefficients.
 After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques in identifying causal influence (Section~\ref{sec:experiment_synthetic}).
-Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features lead to distinguishable brain responses (Section~\ref{sec:experiment_real}).
+Finally, we apply B2B to a large neuroimaging dataset and reveal that distinct but collinear word features lead to distinguishable brain representations (Section~\ref{sec:experiment_real}).

### 2019-09-25 paper/paper.tex
wip

@@
 neuroimaging recordings of 102 subjects reading word sequences. The results
-show that the early and late brain responses caused by low- and high-level
-word features respectively are more reliably detected with B2B than with  other standard techniques.
-% conclude  open
+show that the early and late brain representations, caused by low- and high-level
+word features respectively, are more reliably detected with B2B than with other standard techniques.
+
 \end{abstract}
@@
 %
-In summary, our model can be formalized as:
+In summary, the problem can be formalized as:
 %

### 2019-09-25 paper/meg_exps.tex
modifs

@@
 
-We compare B2B to four standard methods: Forward regression, Backward regression, CCA and PLS, as implemented in scikit-learn \citep{scikit}, and optimized with nested cross validation over twenty $l2$ regularization parameters logarithmically spaced between $10^{-4}$ and $10^4$ (for regression methods) or 1 to 4 canonical components (for cross-decomposition methods).
+We compare B2B to four standard methods: Forward regression, Backward regression, CCA and PLS, as implemented in scikit-learn \citep{sklearn}, and optimized with nested cross validation over twenty $l2$ regularization parameters logarithmically spaced between $10^{-4}$ and $10^4$ (for regression methods) or 1 to 4 canonical components (for cross-decomposition methods).
 
@@
 Left.
-Average improvement of correlation coefficient $\Delta R$ for each pf the four features (rows).
+Average improvement of correlation coefficient $\Delta R$ for each of the four features (rows).
 Error bars indicate standard error of the mean (SEM) across subjects.
@@
 Average $\Delta R$ across time for each subject (dot).
-Top horizontal lines indicate when B2B significantly outperform other methods.}
+Top horizontal lines indicate when B2B significantly outperforms other methods.}
   \vspace{-9ex}
@@
 
-To assess which model would be most sensitive to these causal discoveries, we compared B2B to other models across subjects (Figure \ref{fig:meg} right).
+To assess which model would be most sensitive to these causal discoveries, we compared B2B to other models across subjects (Figure \ref{fig:meg_results} right).
 For Word Length B2B outperforms all models (all $p < 0.00001$) but CCA ($p=0.0678$).

### 2019-09-25 paper/synthetic.tex
modifs

@@
 We compare the performance of B2B against four competing methods, all
-implemented in scikit-learn \citep{scikit}:
+implemented in scikit-learn \citep{sklearn}:
 %
@@
 parameters fitted with the efficient leave-one-out procedure implemented in
-scikit-learn RidgeCV \citep{scikit}.
+scikit-learn RidgeCV \citep{sklearn}.
 
@@
 
-We report AUC results in Figures~\ref{fig:auc_plots}~(left) and ~\ref{fig:percondition}~(top, in Appendix), and compare favorably to all baselines.
+We report AUC results in Figures~\ref{fig:percondition}~(top) and ~\ref{fig:auc_plots}~(left, in Appendix), and compare favorably to all baselines.
 

### 2019-09-25 paper/paper.tex
modifs
a

@@
 
-We consider the measurement of multivariate signal $Y$, generated from a set of variables $X$, via some unknown linear apparatus $F$.
+We consider the measurement of multivariate signal $Y$, generated from a set of putative causes $X$, via some unknown linear apparatus $F$.
 %
@@
 %
-While the triplet of variables $X$ and $N$ are independent, we allow each of them to have a general covariance matrix.
+While the triplet of variables $X$ and $N$ are independent, we allow each of them to have any form of covariance.
 % what is a general covariance matrix?
@@
 %
-This process, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.
+This problem space, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.
 %
@@
 %
-To avoid this issue, we split our sample $(X, Y)$ at random into two subsets $(X_1, Y_1)$ and $(X_2, Y_2)$.
+To avoid this issue, we split our sample $(X, Y)$ into two splits $(X_1, Y_1)$ and $(X_2, Y_2)$.
 %
@@
 \begin{equation}
-\hat{Y} = (\Sigma_X G Y - diag(\Sigma_X G) Y) / diag(I - \Sigma_X G)
+\hat{Y} = (\Sigma_X G Y - diag(\Sigma_X G) Y) \;/\; diag(I - \Sigma_X G) \qquad \text{(element-wise division)}
 \end{equation}
@@
 %
-The rest of this section provides a theoretical guarantee on the correctness of B2B, and an optional post-processing step to binarize the estimated causal influence matrix.
+The rest of this section provides a theoretical guarantee on the correctness of B2B.
 
@@
 
-We perform two experiments to evaluate B2B: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography dataset.
+We perform two experiments to evaluate B2B: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography (MEG) dataset.
 %

### 2019-09-25 paper/synthetic.tex
a

--- 2019-09-25 paper/synthetic.tex
@@
 The purpose of these experiments is to evaluate the ability of B2B in terms of
-prediction (inside and outside the training distribution), as well as a method
+prediction of independent and identically distributed data, as well as a method
 to recover causal factors.

### 2019-09-25 paper/paper.tex

--- 2019-09-25 paper/paper.tex
@@
 Furthermore, B2B closely relates to CCA \citep{cca_hotelling}, but departs from it in four main aspects.
-First, B2B is not symmetric between $X$ and $Y$: it aims to identify specific causal features.
+First, B2B is not symmetric between $X$ and $Y$: it aims to identify specific causal features by first optimizing over the decoders $G$ and then over $H$.
 By contrast, CCA is symmetric between $X$ and $Y$, and aims to find $G$ and $H$ such that they project $X$ and $Y$ on maximally correlated dimensions.
-Second, CCA is based an eigenvector decomposition of $XH$ and $YG$ - the corresponding canonical components are thus mixing the $X$ features in way that limit interpretability.
+Second, CCA is based a eigenvector decomposition of $XH$ and $YG$ - the corresponding canonical components are thus mixing the $X$ features in way that limit interpretability.
 Third, CCA does not typically optimize two distinct regularization parameters.
-Finally, CCA does not use different data splits to estimate $G$ and $H$, and is thus prone to overfitting.
-Together, these differences may explain why B2B reliably outperform CCA on estimating causal influences (Fig.
-\ref{table:synthetic}).
+Finally, CCA does not use different data splits to estimate $G$ and $H$.
+Together, these differences may explain why B2B reliably outperform CCA on estimating causal influences (Figs.
+\ref{fig:percondition} and \ref{fig:auc_plots}).
 Extensions of CCA have been explored in neuroimaging (see \citep{de2019multiway} for brief review), but typically do not address the issue of causal detection.

### 2019-09-25 paper/meg_exps.tex
a

--- 2019-09-25 paper/meg_exps.tex
@@
 
-We used the feature importance described in Algorithm \ref{algorithm:b2b_fi} to assess the extent to which each feature $X_i$ specifically improves the prediction of held-out $Y$ data, using a 5-fold cross-validation.
+We used the feature importance described in Algorithm \ref{algorithm:b2b_fi} to assess the extent to which each feature $X_i$ specifically improves the prediction of held-out $Y$ data, using a 5-fold cross-validation (with shuffled trials to homogeneize the distributions between the training and testing splits).
 

### 2019-09-25 paper/meg_exps.tex
a

@@
   \begin{minipage}[c]{0.4\textwidth}
-    \caption{Ninety subjects read approximately 2,700 words while their brain activity was recorded with MEG.
+    \caption{Ninety subjects each read approximately 2,700 words while their brain activity was recorded with MEG.
 Top.
@@
 Bottom.
-Each line represents magnetometer, color-coded by spatial position.
+Each line represents a magnetometer, color-coded by its spatial position.
 Posterior responses, typical of primary visual cortex activity, peak around 100 ms after word onset and are followed by an anterior propagation of activity typical of semantic processing in the associative cortices.
@@
 $dummy = z(length) + z(frequency) + \mathcal{N}$, where $z$ normalizes features and $\mathcal{N}$ is a random vector sampling Gaussian distribution (all terms thus have a zero-mean and a unit-variance).
-
 This procedure yields an $X \in \mathbb{R}^{n \times d_x}$ matrix of $n\approx$ 2700 words by
@@
 
-We compare B2B to four standard methods: Forward regression, Backward regression, CCA and PLS, as implemented in scikit-learn \citep{sklearn}, and optimized with nested cross validation over twenty $l2$ regularization parameters logarithmically spaced between $10^{-4}$ and $10^4$ (for regression methods) or 1 to 4 canonical components (for cross-decomposition methods).
+We compare B2B to four standard methods: Forward regression, Backward regression, CCA and PLS, as implemented in scikit-learn \citep{sklearn}, and optimized with nested cross-validation over twenty $l2$ regularization parameters logarithmically spaced between $10^{-4}$ and $10^4$ (for regression methods) or 1 to 4 canonical components (for cross-decomposition methods).
 
@@
 Each model was implemented for each subject and each time sample independently.
-Pairwise comparison between models were performed using a Wilcoxon test across subjects (n=90) using the average $\Delta R$ across time.
-
-Corresponding results are shown in Figure~\ref{fig:meg_results}.
+Pairwise comparison between models were performed using a two-sided Wilcoxon test across subjects (n=90) using the average $\Delta R$ across time.
+Corresponding effect sizes are shown in Figure~\ref{fig:meg_results}, and p-values are reported below.
 
@@
 Supplementary Figure \ref{fig:meg_supp} shows that the Backward model decodes the dummy variable well above chance level.
-In addition, the Backward model reveal a similar decoding time course for Word Length and Word Frequency, even though these features are known to specifically influence early and late MEG responses respectively \citep{kutas2011thirty}.
+In addition, the Backward model reveals a similar decoding time course for Word Length and Word Frequency, even though these features are known to specifically influence early and late MEG responses respectively \citep{kutas2011thirty}.
 These results illustrate that backward modelling cannot be used to estimate the causal contribution of collinear features.
@@
 Forward Regression, PLS, CCA, and B2B) and estimate their $\Delta R$ (i.e.
-the improvement of Y prediction induced by the introduction of a given feature into the model \ref{algorithm:b2b_fi}).
+the improvement of Y prediction induced by the introduction of a given feature into the model,as described in Algorithm \ref{algorithm:b2b_fi}).
 Contrary to the Backward Model, none of the models predicted the Dummy Variable to improve the $Y$ prediction: all $\Delta R < 0$ (all $p > .089$).
@@
 For "Word Function", B2B outperforms all models (all $p < 0.0015$).
-Overall, these results show that B2B reliably outperform standard methods, especially when the effects are difficult to detect.
+Overall, these results show that B2B reliably outperforms standard methods, especially when the effects are difficult to detect.

### 2019-09-25 paper/synthetic.tex
a

@@
 
-In most cases, $E$ is not known and AUC can thus not be estimated.
-
-To address this issue, we assess the ability of each model to reliably predict
+In most cases, $E$ is not known and AUC can thus not be estimated. To address this issue, we assess the ability of each model to reliably predict
 independent and identically distributed data from $Y$, given all of the $X$
@@
 We show in Appendix~\ref{appendix:feature_importance} pseudo-code to assess feature importance for our algorithm as well as baselines.
-For the Backward Model, feature importance cannot be assessed as there is no prediction.
+For the Backward Model, feature importance cannot be assessed as the $X$ colinearity is never taken into account.
 
@@
 
-We show in Figures~\ref{fig:percondition} (bottom) and \ref{fig:auc_plots} (right, in Appendix) that our method compares favorably to baselines.
+We show in Figures~\ref{fig:percondition} (bottom) and \ref{fig:auc_plots} (right, in Appendix) that our method outperforms baselines.
 

### 2019-09-25 paper/paper.tex
a

@@
 one per dimension of $Y$ or one per component respectively).
-Furthermore, these  coefficients can be difficult to summarize into a single 'causal' estimation.
-By contrast, B2B quick (Fig.
-\ref{fig:duration}) leads to a single unbiased scalar values $\hat E$ tending towards 1 and 0 for causal and non-causal features.
+Furthermore, these  coefficients can be difficult to summarize into a single causal estimate.
+By contrast, B2B quickly (Fig.
+\ref{fig:duration}) leads to a single unbiased scalar values $\hat E$ tending towards 1 and 0 for causal and non-causal features respectively.
 
 A variety of other statistical methods applied to neuroimaging data have been proposed to clarify what is being represented in brain responses - i.e.
-what feature specifically cause brain activity.
+what feature causes specific brain activity.
 One of the popular linear method is Representational Similarity Analysis (RSA) \citep{kriegeskorte2008representational}, and consists in analyzing the similarity of brain responses associated with specific categorical conditions (e.g.
@@
 By contrast, CCA is symmetric between $X$ and $Y$, and aims to find $G$ and $H$ such that they project $X$ and $Y$ on maximally correlated dimensions.
-Second, CCA is based a eigenvector decomposition of $XH$ and $YG$ - the corresponding canonical components are thus mixing the $X$ features in way that limit interpretability.
+Second, CCA is based an eigen decomposition of $XH$ and $YG$ - the corresponding canonical components are thus mixing the $X$ features in way that limit interpretability.
 Third, CCA does not typically optimize two distinct regularization parameters.
@@
 \ref{fig:percondition} and \ref{fig:auc_plots}).
-Extensions of CCA have been explored in neuroimaging (see \citep{de2019multiway} for brief review), but typically do not address the issue of causal detection.
+Extensions of CCA have been explored in neuroimaging (see \citep{de2019multiway} for brief review), but typically do not address the issue of causal discovery.
 

### 2019-09-25 paper/appendix.tex
duration plot
a

@@
   \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_supp.pdf}
-  \caption{}
+  \caption{Average $\Delta R$ for each $Y$ dimension (Forward model), canonical components (PLS and CCA) or feature (B2B) across subjects.
+Formally, Backward model cannot have a $\Delta R$ because it never demix multiple $X$ features.
+In such cases, we thus simply report the decoding score.}
   \label{fig:meg_supp}
@@
 \end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.7\linewidth]{figures/duration}
+  \caption{Wall-clock run-time for our method B2B and for the baselines. We see that B2B runs much faster than baselines. \label{fig:duration}}
+\end{figure}

### 2019-11-12 code/meg_experiment/models.py
models clean

@@ -1,5 +1,6 @@
 import numpy as np
 from scipy import linalg
+from copy import deepcopy
 from scipy.stats import pearsonr
 from sklearn.cross_decomposition import CCA as SkCCA
 from sklearn.cross_decomposition import PLSRegression as SkPLS
@@ -7,12 +8,18 @@ from sklearn.model_selection import GridSearchCV
 from sklearn.model_selection import ShuffleSplit
 from sklearn.metrics import r2_score
 from sklearn.base import BaseEstimator, RegressorMixin
+import warnings
+warnings.simplefilter("ignore")
+
+alphas = np.logspace(-4, 4, 20)
+components = np.linspace(0., 1., 20)
 
 
 def r_score(X, Y, multioutput='uniform_average'):
     """column-wise correlation coefficients"""
 
-    assert multioutput in ('raw', 'uniform_average', 'variance_weighted')
+    assert multioutput in ('raw', 'raw_values', 'uniform_average',
+                           'variance_weighted')
     if X.ndim == 1:
         X = X[:, None]
     if Y.ndim == 1:
@@ -34,7 +41,8 @@ def r_score(X, Y, multioutput='uniform_average'):
 
 def rn_score(X, Y, scoring='r', multioutput='uniform_average'):
     assert scoring in ('r', 'r2')
-    assert multioutput in ('raw', 'uniform_average', 'variance_weighted')
+    assert multioutput in ('raw', 'raw_values',
+                           'uniform_average', 'variance_weighted')
 
     if scoring == 'r':
         return r_score(X, Y, multioutput=multioutput)
@@ -44,11 +52,14 @@ def rn_score(X, Y, scoring='r', multioutput='uniform_average'):
 
 
 class B2B(BaseEstimator, RegressorMixin):
-    def __init__(self, alphas=np.logspace(-4, 4, 20),
-                 independent_alphas=True, ensemble=None):
+    def __init__(self, alphas=alphas,
+                 independent_alphas=True, ensemble=None,
+                 scoring='r'):
         self.alphas = alphas
         self.independent_alphas = independent_alphas
         self.ensemble = ensemble
+        self.scoring = scoring
+        self.__name__ = 'B2B'
 
     def fit(self, X, Y):
 
@@ -104,15 +115,16 @@ class B2B(BaseEstimator, RegressorMixin):
         self.E_ = np.diag(self.H_)
         return self
 
[…]

### 2019-11-12 paper/appendix.tex
a
sonquist, new proof, end of 2.2

@@
 
-\begin{theorem*}[B2B consistency - general case]
+\begin{theorem}[B2B consistency - general case]
 
-     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full rank noise.
-     %
-     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimize
-     %
-     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $E\hat H = \hat H$
-\end{theorem*}
+     Consider the B2B model from Equation $$Y = (XE + N)F$$ $N$ centred and full rank noise.
 
+     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimise
+
+     $$\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$$ and satisfy
+
+     $$E\hat H = \hat H$$
+\end{theorem}
 \begin{proof}
 
- Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B, we have
+ Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B.
+
+ Since $\hat G$ is the least square estimator of $X$ from $Y$
  \begin{align*}
-    \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
-%                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
-                                                        &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
-     \label{eq:doublenorm}
+    \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2]
 \end{align*}
+Replacing $Y$ by its model definition $Y = (XE+N)F$, we have
+ \begin{align*}
+    \hat G &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2] =\arg \min_G \mathbb{E}[\left \| X - XEFG + NFG \right\|^2]
+  \end{align*}
+  Since $N$ is centered and independent of $X$, we have
+  \begin{align}
+    	  \hat G &=  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
+     \label{eq:Gdoublenorm}
+\end{align}
+
+Samely, for $\hat H$, we have
 \begin{align*}
@@
     &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
-%    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
-    &= EF \hat G
+    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]
+ \end{align*}
+ a positive quantity which reaches a minimum (zero) for
+ \begin{align}
+    \hat H = EF \hat G
+    \label{eq:Hdoublenom}
+\end{align}
+
+Let us now prove that $EF\hat G = F\hat G$.
+
+Let $F^\dagger$ be the pseudo inverse of $F$, and $Z=F^\dagger EF\hat G$, we have $FZ = FF^\dagger EF \hat G$
+
+Since $F$ is full rank on $Img(E)$, we have $FF^\dagger E =E$, and $FZ = EF\hat G$
+
+As $E$ is a binary diagonal matrix, it is an orthogonal projection and therefore a contraction, thus
+ $$ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$$ and
+ $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
+
+But since $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, we also have
+$$\left \| X - XEF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 \leq \left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2$$
+
+Summarizing the above,
+$$\left \| X - XEF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
+$$\left \| X - XEF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2$$
+$$\left \| NF\hat G\right \| ^2 =  \| NEF\hat G \| ^2$$
+
+$N$ being full rank, this yields $EF\hat G = F\Hat G$.
+
+Replacing into $\eqref{eq:Gdoublenorm}$, and setting $H = EFG$, we have
+\begin{align*}
+	\hat G &=  \arg \min_G  \left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 \\
+	&=   \arg \min_G \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 \\
+	\hat H &=  \arg \min_H \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2
+	\label{eq:4}
 \end{align*}
 
-Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero.
-Let $Z=F^\dagger EF\hat G$.
-We have $FZ = FF^\dagger EF  \hat G= EF\hat G$ ($FF^\dagger E =E$ since $F$ is full rank on $Img(E)$).
-Since E is a contraction, we have $ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$.
-Therefore,
- $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
-
-But as $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, the above inequality is an equality, $Z=\hat G$ and $EF\hat G = F\Hat G$.
-Therefore, $\left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 = \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 = \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2$ is minimised by $\hat G$ and $\hat H$.
-Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$.
-This completes the proof.
+Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$, since $E$, a binary diagonal matrix, is involutive. This completes the proof.
 \end{proof}
@@
 
+\subsection{Binarizing $\hat{H}$ - the Sonquist-Morgan criterion}
+
+Back to back regression provides us with $Diag(\hat{H})$, a vector of real numbers that estimate the causal influence of features in $X$.
+In order to derive the true causal influence matrix $E$ , a binary matrix, we need to binarize  $\hat{H}$, i.e.
+test each feature for equality with zero.
+In regression analysis, the traditional approach to this problem employs a $t$-test to check whether the regression coefficients differ from zero \citep{student1908probable}.
+This test will not succeed here: $\hat{H}$ is obtained from a double regression with regularization, and will not follow the t-distribution.
+
+Instead, we treat the binarization of $\hat{H}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''.
+More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
+%
+Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{sonquist_morgan} \citep{kass75} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
+%
+To binarize $\hat{H}$, set to one all the $p$ diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the $d_x-p$ diagonal entries.
+
 \section{Additional Figures}
@@
   \caption{Average $\Delta R$ for each $Y$ dimension (Forward model), canonical components (PLS and CCA) or feature (B2B) across subjects.
-Formally, Backward model cannot have a $\Delta R$ because it never demix multiple $X$ features.
+Formally, Backward model cannot have a $\Delta R$ because it never unmixes the multiple $X$ features.
 In such cases, we thus simply report the decoding score.}

### 2019-11-12 paper/paper.tex
sonquist, new proof, end of 2.2

@@
 \end{proof}
-%
-% \begin{proof}
-%
-%  Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B, we have
-%  \begin{align*}
-%     \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
-% %                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
-%                                                         &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
-%      \label{eq:doublenorm}
-% \end{align*}
-% \begin{align*}
-%     \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
-%     &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
-% %    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
-%     &= EF \hat G
-% \end{align*}
-%
-% Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero.
-Let $Z=F^\dagger EF\hat G$.
-We have $FZ = FF^\dagger EF  \hat G= EF\hat G$ ($FF^\dagger E =E$ since $F$ is full rank on $Img(E)$).
-Since E is a contraction, we have $ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$.
-Therefore,
-%  $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
-%
-% But as $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, the above inequality is an equality, $Z=\hat G$ and $EF\hat G = F\Hat G$.
-Therefore, $\left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 = \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 = \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2$ is minimised by $\hat G$ and $\hat H$.
-Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$.
-This completes the proof.
-% \end{proof}
-%
+
+Since  $E\hat H = \hat H$, we have
 \begin{equation}
-  \text{So, }\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top.
+  \hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top.
 \end{equation}
@@
 
-Assuming, without loss of generality, that the active features are the $k$ first, we have
+Assuming, without loss of generality, that the active features in $E$ are the $k$ first, and rewriting $X=(X_1,X_2)$ and $N=(N_1,N_2)$ ($X_1$ and $N_1$ containing the $k$ first features), we have ($\Sigma_{A B}$ denoting the covariance of $A$ and $B$)
 
@@
 \begin{equation}
-  X^\top X = \left(\begin{array}{lccl}\Sigma_{X1} & \Sigma_{X2} \\ \Sigma_{X2} & \Sigma_{X3}\end{array}\right),\qquad\qquad N^\top N = \left(\begin{array}{lccl}\Sigma_{N1} & \Sigma_{N2} \\ \Sigma_{N2} & \Sigma_{N3}\end{array}\right),
+  X^\top X = \left(\begin{array}{lccl}\Sigma_{X_1 X_1} & \Sigma_{X_1 X_2} \\ \Sigma_{X_1 X_2} & \Sigma_{X_2 X_2}\end{array}\right),\qquad\qquad N^\top N = \left(\begin{array}{lccl}\Sigma_{N_1 N_1} & \Sigma_{N_1 N_2} \\ \Sigma_{N_1 N_2} & \Sigma_{N_2 N_2}\end{array}\right),
 \end{equation}
@@
 \begin{equation}
-  \hat H = \left(\begin{array}{cc}(\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1} & (\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X2} \\0 & 0\end{array}\right),
+  \hat H = \left(\begin{array}{cc}(\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1} & (\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_2} \\0 & 0\end{array}\right),
 \end{equation}
 \begin{equation}
-  Diag_k (\hat H) = Diag((\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1}) = Diag((I+\Sigma_{X1}^{-1}\Sigma_{N1})^{-1}).
+  Diag_k (\hat H) = Diag((\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1}) = Diag((I+\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1})^{-1}).
 \end{equation}
@@
 %
-In the absence of noise, we have $Diag(\hat H) = Diag(E)$.
-Otherwise the $k$ first diagonal elements of $\hat H$ are all positive, and bounded by $\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}}$ and $\frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$, where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X1}$ and $\Sigma_{N1}$.
+In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so $Diag_k(\hat H)=I$, and $$Diag(\hat H) = Diag(E)$$
 
-The average value of the coefficients of $Diag(\hat H)$ is the trace of $\hat H$ divided by $k$.
-On average, since $X$ and $N$ are not correlated by definition, $\Sigma_{X1}^{-1}\Sigma_{N1}$ should have a low condition number, and $\frac{Var(X)}{Var(X)+Var(N)}$ provides a good estimate of $\frac{1}{k}Tr(\hat H)$.
+Otherwise the $k$ first diagonal elements of $\hat H$ are all positive (since we assume that $F$ has full rank over $Img(E)$, they cannot be zero), lower than $1$, and bounded by 
+$$\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}} \text{ and }  \frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$$where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X_1 X_1}$ and $\Sigma_{N_1 N_1}$.
+In real cases, noise will be added to the diagonal elements of $\hat H$, by non-linearities and sampling error.
+
+The average value of the coefficients of $Diag_k(\hat H)$ is the trace of $\hat H$ divided by $k$.
+Since $X$ and $N$ are not correlated, $\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1}$ should have a low condition number, and $\frac{Var(X)}{Var(X)+Var(N)}$ provides a good estimate of $\frac{1}{k}Tr(\hat H)$, that could be used to test whether a feature of $X$ has causal influence.
+The Sonquist and Morgan criterion defined in the appendix is another possibility.
 

### 2019-11-12 paper/appendix.tex
new figs

--- 2019-09-25 paper/appendix.tex
@@
 
-    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $H_k =
-\text{LinearRegression}(X_{train} K, Y_{train} G_i)$\; $R_{k} =
-\text{corr}(X_{test} H_k, Y_{test} G_i)$\; $\Delta R_i = R_{full} - R_{k}$\; }
+    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\;
+    $R_{k} =
+\text{corr}(X_{test} K H, Y_{test} G_i)$\; $\Delta R_i = R_{full} - R_{k}$\; }
 \Return{$\Delta R$} \caption{B2B feature importance.} \label{algorithm:b2b_fi}
@@
 \in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
-$\Delta{R} \in \mathbb{D}^{d_x, d_y}$.} $H = \text{LinearRegression}(X_{train},
-Y_{train})$\; $R_{full} = \text{corr}(X_{test} H, Y_{test})$\;
+$\Delta{R} \in \mathbb{D}^{d_x, d_y}$.}
 
-    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $H_k =
-\text{LinearRegression}(X_{train} K, Y_{train})$\; $R_{k} = \text{corr}(X_{test}
-{H_k}, Y_{test})$\; $\Delta R_i = R_{full} - R_{k}$\; } \Return{$\Delta R$}
+$H = \text{LinearRegression}(X_{train}, Y_{train})\;
+R_{full} = \text{corr}(X_{test} K, Y_{test})$\;
+
+    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $R_{k} = \text{corr}(X_{test}
+K H, Y_{test})$\; $\Delta R_i = R_{full} - R_{k}$\; } \Return{$\Delta R$}
 \caption{Forward feature importance.} \label{algorithm:fwd_fi} \end{algorithm}
@@
 
-    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $H_k, G_k =
-\text{CCA}(X_{train} K, Y_{train})$\; $R_{k} = \text{corr}(X_{test} H_k,
-Y_{test} G_k)$\; $\Delta R_i = R_{full} - R_{k}$\; } \Return{$\Delta R$}
+    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $R_{k} = \text{corr}(X_{test} K H,
+Y_{test} G)$\; $\Delta R_i = R_{full} - R_{k}$\; } \Return{$\Delta R$}
 \caption{CCA and PLS feature importance.} \label{algorithm:cdp_fi}
@@
 
-\begin{figure}[t!]
+% \begin{figure}[t!]
+%   \centering
+%   \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_supp.pdf}
+%   \caption{Average $\Delta R$ for each $Y$ dimension (Forward model), canonical components (PLS and CCA) or feature (B2B) across subjects.
+Formally, Backward model cannot have a $\Delta R$ because it never unmixes the multiple $X$ features.
+In such cases, we thus simply report the decoding score.}
+%   \label{fig:meg_supp}
+% \end{figure}
+
+\begin{figure}
   \centering
-  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_supp.pdf}
-  \caption{}
-  \label{fig:meg_supp}
+  \includegraphics[width=\linewidth]{figures/delta_auc.pdf}
+  \caption{Synthetic experiments.
+Distribution (over conditions) of AUC (top) and Feature Importance $\Delta R$ (bottom) metrics between our method (y-axis) and the baselines (x-axis).
+Each dot is a distinct synthetic experiment.
+Dots below the diagonal indicates that B2B outperform the tested model.
+\label{fig:auc_plots}}
 \end{figure}
@@
   \centering
-  \includegraphics[width=\linewidth]{figures/AUC_R.png}
-  \caption{Synthetic experiments.
-Distribution (over conditions) of differences in AUC (left) and Feature Importance $\Delta R$ (right) metrics between our method and the baselines.
-Mass under the dashed line corresponds to B2B outperforming the baseline.
-\label{fig:auc_plots}}
+  \includegraphics[width=0.7\linewidth]{figures/duration}
+  \caption{Wall-clock run-time for our method B2B and for the baselines. Each dot is a distinct synthetic experiment. B2B runs much faster than cross-decomposition baselines. \label{fig:duration}}
 \end{figure}
+
+\subsection{Robustness on increasing number of factors}
+
+To test whether each of the methods robustly scale to an increasingly large number of potential causes $X$, we enhance the four ad-hoc features (word length, word frequency, word function, dummy variable) with another ten features.
+These additional features corresponds to the first dimensions of word embedding as provided by Spacy \citep{spacy2}.
+The results shown in Figure \ref{fig:embbeding}, show that the ad-hoc features remain unchanged (and are actually improved) in the B2B.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.7\linewidth]{figures/compare_embeddings.pdf}
+  \caption{Comparison of $\Delta R$ when the models are tested on four variables (top) and when the models are tested on an these four variables as well as another 10 word-embedding features (bottom).
+These results illustrate that, unlike Regularized CCA, B2B remains robust even when the number of tested factors increases.
+  \label{fig:embeddings}}
+\end{figure}

### 2019-11-12 paper/meg_exps.tex
fix figure
new figs

@@
 \begin{wrapfigure}{r}{.5\textwidth}
-  \vspace{-12ex}
+  \vspace{-11ex}
   \begin{center}
@@
 Right.
-Average $\Delta R$ across time for each subject (dot).
-Top horizontal lines indicate when B2B significantly outperforms other methods.}
-  \vspace{-9ex}
+Average $\Delta R$ across time for each subject (dots).
+Top horizontal lines indicate when B2B significantly outperforms other methods (red) and vice versa.}
+  \vspace{-6ex}
 \end{wrapfigure}
@@
 
-Supplementary Figure \ref{fig:meg_supp} shows that the Backward model decodes the dummy variable well above chance level.
-In addition, the Backward model reveals a similar decoding time course for Word Length and Word Frequency, even though these features are known to specifically influence early and late MEG responses respectively \citep{kutas2011thirty}.
+As expected, the Backward model reveals a similar decoding time course for Word Length and Word Frequency, even though these features are known to specifically influence early and late MEG responses respectively \citep{kutas2011thirty}.
+In addition, the same decoding time course was observed for the dummy variable.
 These results illustrate that backward modelling cannot be used to estimate the causal contribution of collinear features.

### 2019-11-12 paper/appendix.tex

--- 2019-11-12 paper/appendix.tex
@@
 
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_supp.pdf}
-  \caption{Average $\Delta R$ for each $Y$ dimension (Forward model), canonical components (PLS and CCA) or feature (B2B) across subjects.
+% \begin{figure}[t!]
+%   \centering
+%   \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_supp.pdf}
+%   \caption{Average $\Delta R$ for each $Y$ dimension (Forward model), canonical components (PLS and CCA) or feature (B2B) across subjects.
 Formally, Backward model cannot have a $\Delta R$ because it never unmixes the multiple $X$ features.
 In such cases, we thus simply report the decoding score.}
-  \label{fig:meg_supp}
-\end{figure}
+%   \label{fig:meg_supp}
+% \end{figure}
 
@@
   \centering
-  \includegraphics[width=\linewidth]{figures/AUC_R.png}
+  \includegraphics[width=\linewidth]{figures/delta_auc.pdf}
   \caption{Synthetic experiments.
-Distribution (over conditions) of differences in AUC (left) and Feature Importance $\Delta R$ (right) metrics between our method and the baselines.
-Mass under the dashed line corresponds to B2B outperforming the baseline.
+Distribution (over conditions) of AUC (top) and Feature Importance $\Delta R$ (bottom) metrics between our method (y-axis) and the baselines (x-axis).
+Each dot is a distinct synthetic experiment.
+Dots below the diagonal indicates that B2B outperform the tested model.
 \label{fig:auc_plots}}
@@
   \includegraphics[width=0.7\linewidth]{figures/duration}
-  \caption{Wall-clock run-time for our method B2B and for the baselines. We see that B2B runs much faster than baselines. \label{fig:duration}}
+  \caption{Wall-clock run-time for our method B2B and for the baselines. Each dot is a distinct synthetic experiment. B2B runs much faster than cross-decomposition baselines. \label{fig:duration}}
 \end{figure}
+
+\subsection{Robustness on increasing number of factors}
+
+To test whether each of the methods robustly scale to an increasingly large number of potential causes $X$, we enhance the four ad-hoc features (word length, word frequency, word function, dummy variable) with another ten features.
+These additional features corresponds to the first dimensions of word embedding as provided by Spacy \citep{spacy2}.
+The results shown in Figure \ref{fig:embbeding}, show that the ad-hoc features remain unchanged (and are actually improved) in the B2B.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.7\linewidth]{figures/compare_embeddings.pdf}
+  \caption{Comparison of $\Delta R$ when the models are tested on four variables (top) and when the models are tested on an these four variables as well as another 10 word-embedding features (bottom).
+These results illustrate that, unlike Regularized CCA, B2B remains robust even when the number of tested factors increases.
+  \label{fig:embeddings}}
+\end{figure}

### 2019-11-12 paper/paper.tex

@@
 
+Note that the absence, in our model, of measurement noise over $Y$, is not restrictive, so long $F$ has full-rank (or over $Img(F)$, the subspace spanned by the columns of $F$).
+In fact, adding $M$ as measurement noise on $Y$, we have $$Y=(XE+N)F+M$$ On $Img(F)$,
+rewriting $M = MF^{-1}F$, we have
+$$Y = (XE+N)F+M = (XE+N+MF^{-1})F = (XE+N')F$$
+
 \subsection{Algorithm}
@@
 In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so $Diag_k(\hat H)=I$, and $$Diag(\hat H) = Diag(E)$$ 
-
 Otherwise the $k$ first diagonal elements of $\hat H$ are all positive (since we assume that $F$ has full rank over $Img(E)$, they cannot be zero), lower than $1$, and bounded by 
 $$\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}} \text{ and }  \frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$$where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X_1 X_1}$ and $\Sigma_{N_1 N_1}$.
-In real cases, noise will be added to the diagonal elements of $\hat H$, by non-linearities and sampling error.
 
-The average value of the coefficients of $Diag_k(\hat H)$ is the trace of $\hat H$ divided by $k$.
-Since $X$ and $N$ are not correlated, $\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1}$ should have a low condition number, and $\frac{Var(X)}{Var(X)+Var(N)}$ provides a good estimate of $\frac{1}{k}Tr(\hat H)$, that could be used to test whether a feature of $X$ has causal influence.
-The Sonquist and Morgan criterion defined in the appendix is another possibility.
+In practical cases, non linearities and sampling error will add noise to the diagonal elements of $\hat H$.
+To test diagonal elements for equality with zero, we note that the average value of the non-zero coefficients of $Diag(\hat H)$ is the trace of $\hat H$ divided by $k$.
+Since $X$ and $N$ are not correlated, $\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1}$ should have a low condition number, which means $\frac{Var(X)}{Var(X)+Var(N)}$ should provide a good estimate of $\frac{1}{k}Tr(\hat H)$.
+This can be used to test whether a feature of $X$ has causal influence.
+An alternative is the Sonquist-Morgan criterion presented in the appendix.
 

### 2019-11-12 paper/introduction.tex
correct alex comments

@@
 For example, brain activity is often recorded with hundreds or thousands of sensors simultaneously.
-As multiple sensors may be affected by common noise sources, it is common to use `decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving,  kriegeskorte2008representational, norman2006beyond}.
+As multiple sensors may be affected by common noise sources, it is common to use backward regression, by, for example, fitting a support vector machine across multiple sensors to decode the category of a stimulus \citep{cichy2014resolving,  kriegeskorte2008representational, norman2006beyond}.
 
@@
 Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations.
-By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}.
+By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \citep{weichwald2015causal, hebart2018deconstructing, king2018encoding}.
 To combine the benefits of forward and backward modeling, several authors have proposed to use cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA) \citep{de2019multiway}.
 CCA and PLS aim to find, from two sets of data $X$ and $Y$, the components $H$ and $G$ were $XH$ and $YG$ are maximially correlated or maximally covarying respectively.
-Because CCA and PLS are are based on a generalized eigen decomposition, their resulting coefficients are mixing the features of $X$ and $Y$ in a way that makes them notoriously difficult to interpret \citep{lebart1995statistique}.
+Because CCA and PLS are based on a generalized eigen decomposition, their resulting coefficients are mixing the features of $X$ and $Y$ in a way that makes them notoriously difficult to interpret \citep{lebart1995statistique}.
 

### 2019-11-12 paper/paper.tex
correct alex comments

@@
     \newcommand\posN{7}
-    \newcommand\posF{9.5}
+    \newcommand\posF{10}
 
@@
     \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=blue!10] (N) at (\posN, 0){};
-    \node[thick, draw=black, minimum height=2cm, minimum width=1cm, fill=red!10] (F) at (\posF, 0){};
+    \node[thick, draw=black, minimum height=1cm, minimum width=2cm, fill=red!10] (F) at (\posF, 0){};
     \draw [thick] (2.25, -1.5) to [round left paren ] (2.25, 1.5);
@@
     \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
-    \node[] (reg1) at (5.5,-3.25) {$\hat{E} = \text{Diag}(\underbrace{(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}})$};
+    \node[] (reg1) at (5.5,-3.25) {$\hat{E} = \text{diag}(\underbrace{(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}})$};
     \end{tikzpicture}
-    \caption{Back-to-back regression identifies the subset of factors $E_{ii} = 1$ in $X$ that influence some observations $Y$ by 1) regressing from $Y$ to $X$ to obtain $\hat{X}$, and ii) returning the diagonal of the regression coefficients from $X$ to $\hat{X}$.}
+    \caption{Back-to-back regression identifies the subset of factors $E_{ii} = 1$ in $X$ that influence some observations $Y$ by 1) regressing from $Y$ to $X$ to obtain $\hat{X}$, and 2) returning the diagonal of the regression coefficients from $X$ to $\hat{X}$.}
     \label{fig:b2b}
@@
 
-We consider the measurement of multivariate signal $Y$, generated from a set of putative causes $X$, via some unknown linear apparatus $F$.
+We consider the measurement of multivariate signal $Y \in \mathbb{R}^{n \times d_y}$, generated from a set of putative causes $X \in \mathbb{R}^{n \times d_x}$, via some unknown linear apparatus $F \in \mathbb{R}^{d_x \times d_y}$.
 %
@@
 %
-By considering a square binary diagonal matrix of \emph{causal influences} $E$, we denote by $XE$ the causal factors of $Y$.
+By considering a square binary diagonal matrix of \emph{causal influences} $E \in \mathbb{D}^{d_x \times d_x}$, we denote by $XE$ the causal factors of $Y$.
 %
@@
 \begin{equation}
-    Y = (XE + N)F,\label{eq:model}
+    y_i = (x_i E + n_i) F
+    \label{eq:model}
 \end{equation}
 %
-where $N$ is a centered, unobserved noise variable.
+where $i$ is a given sample, and $n$ is a centered, unobserved noise variable.
 %
@@
 %
-The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{Diag}(\hat{H})$, is the desired estimate of the causal influence matrix $E$.
+The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{diag}(\hat{H})$, is the desired estimate of the causal influence matrix $E$.
 
@@
 %
-To compensate for the reduction in sample size caused by the split, B2B is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split \citep{efron1992bootstrap}.
+To compensate for the reduction in sample size caused by the split, B2B is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split \citep{breiman1996bagging}.
 %
@@
 \begin{equation}
-\hat{Y} = (\Sigma_X G Y - diag(\Sigma_X G) Y) \;/\; diag(I - \Sigma_X G) \qquad \text{(element-wise division)}
+\hat{Y} = (\Sigma_X G Y - \text{diag}(\Sigma_X G) Y) \;/\; \text{diag}(I - \Sigma_X G) \qquad \text{(element-wise division)}
 \end{equation}
@@
   \Sigma_X = Q V Q^T \\
-  G = Q (V + \lambda*I)^{-1} Q^T\\
+  G = Q (V + \lambda I)^{-1} Q^T\\
   \end{aligned}
@@
         $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$ \Comment*[r]{$\hat H=(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2 \hat G$}
-        $\hat{E} \leftarrow \hat{E} + \text{Diag}(\hat{H})$\;
+        $\hat{E} \leftarrow \hat{E} + \text{diag}(\hat{H})$\;
     }
@@
      %
-     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimize
+     Let $Img(M)$ refers to the image of the matrix $M$. If $F$ and $X$ are full-rank on the $Img(E)$, then, the solution of B2B, $\hat H$, will minimize
      %
@@
 
-Assuming, without loss of generality, that the active features in $E$ are the $k$ first, and rewriting $X=(X_1,X_2)$ and $N=(N_1,N_2)$ ($X_1$ and $N_1$ containing the $k$ first features), we have ($\Sigma_{A B}$ denoting the covariance of $A$ and $B$)
+Assuming, without loss of generality, that the active features in $E$ are the $k \in \mathbb{Z}: k \in [0, d_x]$ first features, and rewriting $X=(X_1,X_2)$ and $N=(N_1,N_2)$ ($X_1$ and $N_1$ containing the $k$ first features), we have ($\Sigma_{A B}$ denoting the covariance of $A$ and $B$)
 
@@
 \begin{equation}
-  Diag_k (\hat H) = Diag((\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1}) = Diag((I+\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1})^{-1}).
+  \text{diag}_k (\hat H) = \text{diag}((\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1}) = \text{diag}((I+\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1})^{-1}).
 \end{equation}
@@
 %
-In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so $Diag_k(\hat H)=I$, and $$Diag(\hat H) = Diag(E)$$ 
+In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so $\text{diag}_k(\hat H)=I$, and $$\text{diag}(\hat H) = \text{diag}(E)$$
 Otherwise the $k$ first diagonal elements of $\hat H$ are all positive (since we assume that $F$ has full rank over $Img(E)$, they cannot be zero), lower than $1$, and bounded by
@@
 In practical cases, non linearities and sampling error will add noise to the diagonal elements of $\hat H$.
-To test diagonal elements for equality with zero, we note that the average value of the non-zero coefficients of $Diag(\hat H)$ is the trace of $\hat H$ divided by $k$.
+To test diagonal elements for equality with zero, we note that the average value of the non-zero coefficients of $\text{diag}(\hat H)$ is the trace of $\hat H$ divided by $k$.
 Since $X$ and $N$ are not correlated, $\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1}$ should have a low condition number, which means $\frac{Var(X)}{Var(X)+Var(N)}$ should provide a good estimate of $\frac{1}{k}Tr(\hat H)$.
@@
 
+% I find a bit surprising to mention cauaslity literare in ML at the end. I would have expected this in the introduction to convince reviewer that know your stuff.
+
 Finally, B2B is also related to, but different from causal discovery algorithms used outside the neuroimaging field.
 These methods are often nonlinear \citep{peters2017elements} and base their causal estimates on temporal delays \citep{granger1969investigating, janzing2013quantifying}.
-For example, B2B relates to tthe half-sibling regression \citep{scholkopf2016modeling}, in the sense that we observe a large number of variables corrupted by the same source of noise; we leave the extension of B2B to nonlinear time-series for future work.
+For example, B2B relates to the half-sibling regression \citep{scholkopf2016modeling}, in the sense that we observe a large number of variables corrupted by the same source of noise; we leave the extension of B2B to nonlinear time-series for future work.
 

### 2019-11-12 paper/synthetic.tex
new figs
correct alex comments

@@
 The data generating process for each experiment constructs $n=1000$ training examples
-according to the model $Y = (\text{snr} \cdot XE + N)F$, where $\text{snr}$ is a
+according to the model $Y = (\text{h} XE + N)F$, where $\text{h}$ is a
 scalar that modulates the signal-to-noise ratio.
@@
 %\begin{itemize} \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries
-%drawn from $\mathcal{N}(0, d_x^{-1})$, \item $X \in \mathbb{R}^{n \times d_x}$
+%drawn from $\mathcal{N}(0, \sigma^2)$, \item $X \in \mathbb{R}^{n \times d_x}$
 %contains rows drawn from $\mathcal{N}(0, \Sigma_X)$, \item $N \in \mathbb{R}^{n
@@
 %\item $\Sigma_X = AA^\top$, where $A \in \mathbb{R}^{d_x \times d_x}$ contains
-%entries drawn from $\mathcal{N}(0, d_x^{-1})$, \item $\Sigma_N = BB^\top$,
+%entries drawn from $\mathcal{N}(0, \sigma^2)$, \item $\Sigma_N = BB^\top$,
 %where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
-%$\mathcal{N}(0, d_x^{-1})$, \end{itemize}
+%$\mathcal{N}(0, \sigma^2)$, \end{itemize}
     $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from
-$\mathcal{N}(0, d_x^{-1})$, $X \in \mathbb{R}^{n \times d_x}$ contains rows
+$\mathcal{N}(0, \sigma^2)$ where $\sigma^2$ is inversely proportional to $d_x$,
+$X \in \mathbb{R}^{n \times d_x}$ contains rows
 drawn from $\mathcal{N}(0, \Sigma_X)$, $N \in \mathbb{R}^{n \times d_x}$
@@
 AA^\top$ where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
-$\mathcal{N}(0, d_x^{-1})$, $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x
-\times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$, and the
-factor $\text{snr} \in (0, \infty)$.
+$\mathcal{N}(0, \sigma^2)$, $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x
+\times d_x}$ contains entries drawn from $\mathcal{N}(0, \sigma^2)$, and the
+factor $\text{h} \in \mathcal{R}_+$.
 
 To simulate a wide range of experimental conditions, we sample 10 values in log-space for $d_x, d_y \in \left[ 10, 100 \right]$, $n_c \in \left[ 3, 63 \right]$,
-$\text{snr} \in \left[ 0.001, 10 \right]$. We discard the cases where $n_c > d_x$, limit $d_x, d_y$ to 100 to keep the running time under 2 hours for each condition, and average over 5 random seeds.
+$\text{h} \in \left[ 0.001, 10 \right]$. We discard the cases where $n_c > d_x$, limit $d_x, d_y$ to 100 to keep the running time under 2 hours for each condition, and average over 5 random seeds.
 %
@@
 % \begin{equation} maxcov(XH^T, YG^T) \end{equation}
-\begin{equation} G,H = \argmax_{G,H} cov(XH^T, YG^T) \end{equation}
+\begin{equation} G,H = \argmax_{G,H} \text{cov}(XH^T, YG^T) \end{equation}
 
@@
   \centering
-  \includegraphics[width=\linewidth]{figures/AUC_conditions}
-  \includegraphics[width=\linewidth]{figures/R_conditions}
+  \includegraphics[width=\linewidth]{figures/auc_condition.pdf}
+  \includegraphics[width=\linewidth]{figures/r_in_condition.pdf}
   \vspace{-4ex}

### 2019-11-12 paper/meg_exps.tex
o

@@
 
-We compare B2B to four standard methods: Forward regression, Backward regression, CCA and PLS, as implemented in scikit-learn \citep{sklearn}, and optimized with nested cross-validation over twenty $l2$ regularization parameters logarithmically spaced between $10^{-4}$ and $10^4$ (for regression methods) or 1 to 4 canonical components (for cross-decomposition methods).
+We compare B2B to four standard methods: Forward regression, Backward regression, CCA and PLS, as implemented in scikit-learn \citep{sklearn} and \citep{bilenko2016pyrcca}, and optimized with nested cross-validation over twenty $l2$ regularization parameters logarithmically spaced between $10^{-4}$ and $10^4$ (for regression and CCA methods) or 1 to 4 canonical components (for PLS).
 

### 2019-11-12 paper/paper.tex
o

@@
 %
-We use scikit-learn's CCA, PLS, and RidgeCV objects to compare B2B against the standard baselines \citep{sklearn}.
+We use scikit-learn's Partial Least Regression (PLS) and RidgeCV \citep{sklearn} as well as pyrcca' regularized canonical compnent analysis (RegCCA, \citep{bilenko2016pyrcca}) objects to compare B2B against the standard baselines.
 

### 2019-11-13 paper/introduction.tex
fix

@@
 First, causal and non-causal factors may be numerous and collinear.
-In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or a change in pressure, as these two factors may, at first, be difficult to manipulate independently.
+%In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or a change in pressure, as these two factors may, at first, be difficult to manipulate independently.
 This issue becomes increasingly pronounced as the number of potential factors increases.
@@
 
-Here, we introduce the `back-to-back regression' (B2B), which not only combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}), but also provide robust and interpretable causal coefficients.
-After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques in identifying causal influence (Section~\ref{sec:experiment_synthetic}).
+Here, we introduce the `back-to-back regression' (B2B), which not only combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}), but also provide robust, interpretable and unidimensional causal coefficients for each tested factor.
+
+The present paper focuses on the restricted issue of disentangling the causal influence of linearly correlated predictors ($X$) onto multivariate observations ($Y$).
+This issue thus differs from other causal discovery algorithms based on temporal-delays and/or nonlinear interaction in systems where the directionality of causation (from X to Y or vice versa) is unknown (e.g.
+\citep{peters2017elements, granger1969investigating, janzing2013quantifying, scholkopf2016modeling}.
+
+After detailing B2B method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques in identifying causal influence (Section~\ref{sec:experiment_synthetic}).
 Finally, we apply B2B to a large neuroimaging dataset and reveal that distinct but collinear word features lead to distinguishable brain representations (Section~\ref{sec:experiment_real}).

### 2019-11-13 paper/paper.tex
fix

--- 2019-11-12 paper/paper.tex
@@
 %
-We use scikit-learn's CCA, PLS, and RidgeCV objects to compare B2B against the standard baselines \citep{sklearn}.
+We use scikit-learn's PLS and RidgeCV \citep{sklearn} as well as pyrcca' regularized canonical compnent analysis (RegCCA, \citep{bilenko2016pyrcca}) objects to compare B2B against the standard baselines.
 
@@
 % I find a bit surprising to mention cauaslity literare in ML at the end. I would have expected this in the introduction to convince reviewer that know your stuff.
-
-Finally, B2B is also related to, but different from causal discovery algorithms used outside the neuroimaging field.
-These methods are often nonlinear \citep{peters2017elements} and base their causal estimates on temporal delays \citep{granger1969investigating, janzing2013quantifying}.
-For example, B2B relates to the half-sibling regression \citep{scholkopf2016modeling}, in the sense that we observe a large number of variables corrupted by the same source of noise; we leave the extension of B2B to nonlinear time-series for future work.
+CONCLUSION SENTENCE
 

### 2019-11-13 paper/paper.tex
small change

@@
 
-Note that the absence, in our model, of measurement noise over $Y$, is not restrictive, so long $F$ has full-rank (or over $Img(F)$, the subspace spanned by the columns of $F$).
-In fact, adding $M$ as measurement noise on $Y$, we have $$Y=(XE+N)F+M$$ On $Img(F)$,
-rewriting $M = MF^{-1}F$, we have
+Note that the absence, in our model, of measurement noise over $Y$, is not restrictive. Starting from a model with measurement noise $M$, $$Y=(XE+N)F+M$$ On $Img(F)$,
+we can rewrite $M = MF^{-1}F$, and we have
 $$Y = (XE+N)F+M = (XE+N+MF^{-1})F = (XE+N')F$$
+If $F$ has full rank, the
 
@@
 % I find a bit surprising to mention cauaslity literare in ML at the end. I would have expected this in the introduction to convince reviewer that know your stuff.
-CONCLUSION SENTENCE
 

### 2019-11-13 paper/paper.tex
small change

@@
 $$Y = (XE+N)F+M = (XE+N+MF^{-1})F = (XE+N')F$$
-If $F$ has full rank, the
+If $F$ has full rank, or in $Img(F)$, the two models are equivalent.
 

### 2019-11-13 paper/paper.tex
fix

@@
 %
-The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{diag}(\hat{H})$, is the desired estimate of the causal influence matrix $E$.
+The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{diag}(\hat{H})$, is the desired estimate of the causal influence matrix $E$, as detailed in the \ref{appendix:theorem_proof}.
 
-If using regularized least-squares, arguably the most commonly employed linear regression technique \citep{hoerl1959optimum, rifkin2007notes}, the solution of B2B has a closed form solution:
+If using l2-regularized least-squares \citep{hoerl1959optimum, rifkin2007notes}, B2B has a closed form solution:
 \begin{align}

### 2019-11-13 paper/synthetic.tex
o
fix

@@
 We compare the performance of B2B against four competing methods, all
-implemented in scikit-learn \citep{sklearn}:
+implemented in scikit-learn \citep{sklearn} and pyrcca \citep{bilenko2016pyrcca}:
 %
@@
 Forward regression consists of an $l2$-regularized "ridge" regression from the
-putative causes $X$ to the observations $Y$: \begin{equation} H = (X^T X
+putative causes $X$ to the observations $Y$: \begin{equation} H_{fwd} = (X^T X
 +\lambda I)^{-1} X^T Y \end{equation}
@@
 Backward regression consists of an $l2$-regularized "ridge" regression from $Y$
-to $X$: \begin{equation} G = (Y^T Y +\lambda I)^{-1} Y^T X \end{equation}
+to $X$: \begin{equation} G_{bwd} = (Y^T Y +\lambda I)^{-1} Y^T X \end{equation}
 
-CCA finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$
+CCA finds $G_{cca}\in\mathbb{R}^{d_z, d_y}$ and $H_{cca}\in\mathbb{R}^{d_z, d_x}$
 % such that
@@
 % \begin{equation} maxcorr(XH^T, YG^T) \end{equation}
-\begin{equation} G,H = \argmax_{G,H} corr(XH^T, YG^T) \end{equation}
+\begin{equation} G_{cca},H_{cca} = \argmax_{G,H} corr(XH^T, YG^T) \end{equation}
 
 % To be checked
-PLS finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$
+PLS finds $G_{pls}\in\mathbb{R}^{d_z, d_y}$ and $H_{pls}\in\mathbb{R}^{d_z, d_x}$
 % such that
@@
 % \begin{equation} maxcov(XH^T, YG^T) \end{equation}
-\begin{equation} G,H = \argmax_{G,H} \text{cov}(XH^T, YG^T) \end{equation}
+\begin{equation} G_{pls},H_{pls} = \argmax_{G,H} \text{cov}(XH^T, YG^T) \end{equation}
 

### 2019-11-13 paper/appendix.tex
fix
o

@@
 
-\subsection{Robustness on increasing number of factors}
+\subsection{Robustness to increasing number of factors}
 
-To test whether each of the methods robustly scale to an increasingly large number of potential causes $X$, we enhance the four ad-hoc features (word length, word frequency, word function, dummy variable) with another ten features.
+To test whether each of the methods robustly scales to an increasingly large number of potential causes $X$, we enhanced the four ad-hoc features (word length, word frequency, word function, dummy variable) with another ten features.
 These additional features corresponds to the first dimensions of word embedding as provided by Spacy \citep{spacy2}.
-The results shown in Figure \ref{fig:embbeding}, show that the ad-hoc features remain unchanged (and are actually improved) in the B2B.
+The results shown in Figure \ref{fig:embbedings}, show that the feature importance of ad-hoc features as derived by B2B remain unchanged (and are actually improved).
 
@@
   \centering
-  \includegraphics[width=0.7\linewidth]{figures/compare_embeddings.pdf}
+  \includegraphics[width=0.5\linewidth]{figures/compare_embeddings.pdf}
   \caption{Comparison of $\Delta R$ when the models are tested on four variables (top) and when the models are tested on an these four variables as well as another 10 word-embedding features (bottom).

### 2019-11-13 paper/introduction.tex
o

@@
 
-The present paper focuses on the restricted issue of disentangling the causal influence of linearly correlated predictors ($X$) onto multivariate observations ($Y$).
-This issue thus differs from other causal discovery algorithms based on temporal-delays and/or nonlinear interaction in systems where the directionality of causation (from X to Y or vice versa) is unknown (e.g.
+The present paper focuses on the restricted issue of disentangling the causal influence of linearly correlated predictors ($X$) onto noisy multivariate observations ($Y$).
+The present approach thus differs from other causal discovery algorithms based on temporal-delays and/or nonlinear interaction in systems where the directionality of causation (from X to Y or vice versa) is unknown (e.g.
 \citep{peters2017elements, granger1969investigating, janzing2013quantifying, scholkopf2016modeling}.

### 2019-11-13 paper/paper.tex
small change

@@
 In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so $\text{diag}_k(\hat H)=I$, and $$\text{diag}(\hat H) = \text{diag}(E)$$
-Otherwise the $k$ first diagonal elements of $\hat H$ are all positive (since we assume that $F$ has full rank over $Img(E)$, they cannot be zero), lower than $1$, and bounded by
-$$\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}} \text{ and }  \frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$$where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X_1 X_1}$ and $\Sigma_{N_1 N_1}$.
+Otherwise the $k$ first diagonal elements of $\hat H$ are all positive, since we assume that $F$ has full rank over $Img(E)$. In addition, they are bounded:
+$$\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}} \leq \text{diag}_k(\hat{H})\leq  \frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$$
+where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X_1 X_1}$ and $\Sigma_{N_1 N_1}$.
+This allows us to recover $E$ by identifying the non-zero elements on the diagonal of $\hat{H}$.
 
-In practical cases, non linearities and sampling error will add noise to the diagonal elements of $\hat H$.
-To test diagonal elements for equality with zero, we note that the average value of the non-zero coefficients of $\text{diag}(\hat H)$ is the trace of $\hat H$ divided by $k$.
-Since $X$ and $N$ are not correlated, $\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1}$ should have a low condition number, which means $\frac{Var(X)}{Var(X)+Var(N)}$ should provide a good estimate of $\frac{1}{k}Tr(\hat H)$.
-This can be used to test whether a feature of $X$ has causal influence.
-An alternative is the Sonquist-Morgan criterion presented in the appendix.
+In practical cases, non-linearities and sampling error will add noise to the diagonal elements of $\hat H$.
+We need to test whether diagonal elements are different from zero.
+The average value $\mu$ of non-zero coefficients of $\text{diag}(\hat H)$ is the trace of $\hat H$ divided by $k$.
+$\mu$ can be estimated as follow :
+\begin{equation}
+$\mu = \frac{Var(X)}{Var(X)+Var(N)}$
+\end{equation}
+This provides a method to identify causal features
+%Since $X$ and $N$ are not correlated, $\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1}$ should have a low condition number, which means  should provide a good estimate of $\frac{1}{k}Tr(\hat H)$.
+
+To test whether a feature of $X$ has causal influence. An alternative is the Sonquist-Morgan criterion presented in the appendix.
 

### 2019-11-13 paper/appendix.tex
appen

@@
 
-     Consider the B2B model from Equation $$Y = (XE + N)F$$ $N$ centred and full rank noise.
+     Consider the B2B model from equation \ref{eq:model} $$Y = (XE + N)F$$ with $N$ centred and full rank noise.
 
-     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimise
+     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$ minimizes
 
-     $$\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$$ and satisfy
+     $$\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$$ and satisfies
 

### 2019-11-13 paper/appendix.tex
clarification

@@
 \newpage
+\subsection{Modeling measurement noise}
+
+Equation \ref{eq:model} does not explicitly contain a measurement noise term. Yet, in moast experimental cases, the problem is best described as:
+\begin{equation}
+  Y = (XE+N)F+M
+  \label{eq:model_noisy_measure}
+\end{equation}
+  with $M\in \mathbb{R}^{n \times d_y}$.
+
+This equation is actually equivalent to Equation \ref{eq:model} given our hypotheses. Indeed, we can rewrite $M = MF^{-1}F$ over $Img(F)$, which leads to:
+$$Y = (XE+N)F+M = (XE+N+MF^{-1})F = (XE+N')F$$
+
+Consequently, assuming that $F$ is full rank on $Img(XE)$, B2B yields the same solutions to equations \ref{eq:model} and \ref{model_noisy_measure}.
+
 \subsection{Feature importance}

### 2019-11-13 paper/paper.tex
fix
clarification

@@
 %
-where $i$ is a given sample, and $n$ is a centered, unobserved noise variable.
+where $i$ is a given sample, and $n_i$ is a sample-specific noise drawen from a centered distribution.
 %
@@
 Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back Regression (B2B) is to estimate the matrix of causal influences $E$.
-
-Note that the absence, in our model, of measurement noise over $Y$, is not restrictive. Starting from a model with measurement noise $M$, $$Y=(XE+N)F+M$$ On $Img(F)$,
-we can rewrite $M = MF^{-1}F$, and we have
-$$Y = (XE+N)F+M = (XE+N+MF^{-1})F = (XE+N')F$$
-If $F$ has full rank, or in $Img(F)$, the two models are equivalent.
 

### 2019-11-13 paper/appendix.tex
clarification thresholding

@@
 
-\subsection{Binarizing $\hat{H}$ - the Sonquist-Morgan criterion}
+\subsection{Recovering E}
 
-Back to back regression provides us with $Diag(\hat{H})$, a vector of real numbers that estimate the causal influence of features in $X$.
-In order to derive the true causal influence matrix $E$ , a binary matrix, we need to binarize  $\hat{H}$, i.e.
-test each feature for equality with zero.
-In regression analysis, the traditional approach to this problem employs a $t$-test to check whether the regression coefficients differ from zero \citep{student1908probable}.
-This test will not succeed here: $\hat{H}$ is obtained from a double regression with regularization, and will not follow the t-distribution.
+Three distinct thresholding rules can be used to binarize $\text{diag(\hat H)}$ and tbus explicitly recover causal features under noise. (In the MEG experiment, we implement the third one).
 
-Instead, we treat the binarization of $\hat{H}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''.
-More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
+First, given known signal-to-noise ratio, one can derive the bound above which a feature should considered to be causal.
+Indeed, Equation $\ref:diagk$ implies that the $k$ first diagonal elements of $\hat H$ are bounded:  % we assume that $F$ has full rank over $Img(E)$.
+
+$$0 \leq \frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}} \leq \text{diag}_k(\hat{H})\leq  \frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$$
+where $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X_1 X_1}$ and $\Sigma_{N_1 N_1}$.
+
+The average value $\mu$ of non-zero coefficients of $\text{diag}(\hat H)$ is the trace of $\hat H$ divided by $k$. In practive $\mu$ can be estimated as :
+\begin{equation}
+\mu = \frac{Var(X)}{Var(X)+Var(N)}
+\end{equation}
+
+The decision threshold between causal and non-causal elements is thus a fraction $\mu$. This threshold is to be adjusted depending on the type I and type II risks.
+
+Second, $\text{diag{\hat H}}$ can be binarized with the Sonquist-Morgan criterion \citep{sonquist_morgan}, a clustering procedure separating small and large values.
+This procedure maximizes the ratio of inter-group variance ands minimizes the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
 %
-Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{sonquist_morgan} \citep{kass75} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
+Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, Sonquist-Morgan criterion maximizes \citep{sonquist_morgan} \citep{kass75}:
+\begin{equation}
+  \frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}
+\end{equation}
 %
-To binarize $\hat{H}$, set to one all the $p$ diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the $d_x-p$ diagonal entries.
+Finally, second-order statistics across multiple datasets can be used to identify the elements of $\text{diag}(\hat H)$ that are sgnificantly different from 0.
+This procedure is detailed in the methods.
 

### 2019-11-13 paper/paper.tex
clarification thresholding

@@
   \text{diag}_k (\hat H) = \text{diag}((\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1}) = \text{diag}((I+\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1})^{-1}).
+  \label{eq:diagk}
 \end{equation}
@@
 %
-In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so $\text{diag}_k(\hat H)=I$, and $$\text{diag}(\hat H) = \text{diag}(E)$$
-Otherwise the $k$ first diagonal elements of $\hat H$ are all positive, since we assume that $F$ has full rank over $Img(E)$. In addition, they are bounded:
-$$\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}} \leq \text{diag}_k(\hat{H})\leq  \frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$$
-where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X_1 X_1}$ and $\Sigma_{N_1 N_1}$.
-This allows us to recover $E$ by identifying the non-zero elements on the diagonal of $\hat{H}$.
+In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so $\text{diag}_k(\hat H)=I$, and $$\text{diag}(\hat H) = \text{diag}(E)$$ Therefore, we recover $E$ from $\hat H$.
 
-In practical cases, non-linearities and sampling error will add noise to the diagonal elements of $\hat H$.
-We need to test whether diagonal elements are different from zero.
-The average value $\mu$ of non-zero coefficients of $\text{diag}(\hat H)$ is the trace of $\hat H$ divided by $k$.
-$\mu$ can be estimated as follow :
-\begin{equation}
-$\mu = \frac{Var(X)}{Var(X)+Var(N)}$
-\end{equation}
-This provides a method to identify causal features
-%Since $X$ and $N$ are not correlated, $\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1}$ should have a low condition number, which means  should provide a good estimate of $\frac{1}{k}Tr(\hat H)$.
-
-To test whether a feature of $X$ has causal influence. An alternative is the Sonquist-Morgan criterion presented in the appendix.
+Otherwise, with noise, the causal factors of $E$ correspond to the positive elements of $\test{diag}(\hatH)$. The methods to recover them are presented in Appendix XXXX$$$
 

### 2019-11-13 paper/paper.tex
small fixes

--- 2019-11-13 paper/paper.tex
@@
 \begin{equation}
-$\mu = \frac{Var(X)}{Var(X)+Var(N)}$
+\mu = \frac{Var(X)}{Var(X)+Var(N)}
 \end{equation}
@@
 %
-We use scikit-learn's PLS and RidgeCV \citep{sklearn} as well as pyrcca' regularized canonical compnent analysis (RegCCA, \citep{bilenko2016pyrcca}) objects to compare B2B against the standard baselines.
+We use scikit-learn's PLS and RidgeCV \citep{sklearn} as well as pyrcca's regularized canonical component analysis (RegCCA, \citep{bilenko2016pyrcca}) objects to compare B2B against the standard baselines.
 

### 2019-11-13 paper/meg_exps.tex
fix

@@
 
-\begin{wrapfigure}{r}{.5\textwidth}
-  \vspace{-11ex}
-  \begin{center}
-    \includegraphics[width=0.48\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg.pdf}
-
-    \label{fig:meg_results}
-  \end{center}
-  \caption{Multiple models (color-coded) are compared on their ability to reliably predict single-trial MEG signals evoked by words.
-Left.
-Average improvement of correlation coefficient $\Delta R$ for each of the four features (rows).
-Error bars indicate standard error of the mean (SEM) across subjects.
-Right.
-Average $\Delta R$ across time for each subject (dots).
-Top horizontal lines indicate when B2B significantly outperforms other methods (red) and vice versa.}
-  \vspace{-6ex}
-\end{wrapfigure}
-
 \subsubsection{Results}
@@
 Overall, these results show that B2B reliably outperforms standard methods, especially when the effects are difficult to detect.
+
+\begin{wrapfigure}{r}{.6\textwidth}
+  \vspace{-12ex}
+  \begin{center}
+    \includegraphics[width=0.6\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg.pdf}
+  \end{center}
+  \vspace{-5ex}
+  \caption{Multiple models (color-coded) are compared on their ability to reliably predict single-trial MEG signals evoked by words.
+Left.
+Average improvement of correlation coefficient $\Delta R$ for each of the four features (rows).
+Error bars indicate standard error of the mean (SEM) across subjects.
+Right.
+Average $\Delta R$ across time for each subject (dots).
+Top horizontal lines indicate when B2B significantly outperforms other methods (red) and vice versa.
+\label{fig:meg_results}
+}
+\end{wrapfigure}

### 2019-11-13 paper/appendix.tex
small fixes

--- 2019-11-13 paper/appendix.tex
@@
 
-     Consider the B2B model from equation \ref{eq:model} $$Y = (XE + N)F$$ with $N$ centred and full rank noise.
+     Consider the B2B model from Equation $$Y = (XE + N)F$$ $N$ centred and full rank noise.
 
-     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$ minimizes
+     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimise
 
-     $$\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$$ and satisfies
+     $$\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$$ and satisfy
 
@@
 Instead, we treat the binarization of $\hat{H}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''.
-More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
+More specifically, we propose to maximize the inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
 %

### 2019-11-13 paper/appendix.tex
wip

--- 2019-11-13 paper/appendix.tex
@@
 
-Three distinct thresholding rules can be used to binarize $\text{diag(\hat H)}$ and tbus explicitly recover causal features under noise. (In the MEG experiment, we implement the third one).
+In case of noise, B2B yields non binary $\hat E$. Three thresholding rules can be used to binarize its values thus explicitly recover causal features.
 
-First, given known signal-to-noise ratio, one can derive the bound above which a feature should considered to be causal.
+First, given known signal-to-noise ratio,the threshold above which a feature should considered to be causal can be derived analytically.
 Indeed, Equation $\ref:diagk$ implies that the $k$ first diagonal elements of $\hat H$ are bounded:  % we assume that $F$ has full rank over $Img(E)$.
@@
 
-The average value $\mu$ of non-zero coefficients of $\text{diag}(\hat H)$ is the trace of $\hat H$ divided by $k$. In practive $\mu$ can be estimated as :
+The average value $\mu$ of non-zero coefficients of $\text{diag}(\hat H)$ is the trace of $\hat H$ divided by $k$, and can be computed as
 \begin{equation}
@@
 
-The decision threshold between causal and non-causal elements is thus a fraction $\mu$. This threshold is to be adjusted depending on the type I and type II risks.
+The decision threshold between causal and non-causal elements is thus a fraction $\mu$, whose proportion arbitrarily depends on the necessity to favor type I and type II errors.
+In practice, we cannot use this procedure for our MEG study, because signal-to-noise ratio is unknown.
 
-Second, $\text{diag{\hat H}}$ can be binarized with the Sonquist-Morgan criterion \citep{sonquist_morgan}, a clustering procedure separating small and large values.
-This procedure maximizes the ratio of inter-group variance ands minimizes the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
-%
-Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, Sonquist-Morgan criterion maximizes \citep{sonquist_morgan} \citep{kass75}:
+Second, $\text{diag{\hat H}}$ can be binarized with the Sonquist-Morgan criterion \citep{sonquist_morgan}, a non-parametric clustering procedure separating small and large values in a given set.
+This procedure maximizes the ratio of inter-group variance while minimizing the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
+Let $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, Sonquist-Morgan criterion maximizes \citep{kass75}:
 \begin{equation}
@@
 \end{equation}
+
+This procedure assumes that there exists at least one causal and at least one non-causal feature.
 %
-Finally, second-order statistics across multiple datasets can be used to identify the elements of $\text{diag}(\hat H)$ that are sgnificantly different from 0.
-This procedure is detailed in the methods.
+Third, second-order statistics across multiple datasets can be used to identify the elements of $\text{diag}(\hat H)$ that are sgnificantly different from 0.
+This procedure is detailed in the method section of our MEG experiment.
+
+Overall, these three procedure thus varies in their additional assumptions: i.e.
+(1) a known signal-to-noise ratio, (2) the existence of both causal and non-causal factors or (3) independent repetitions of the experiment.
 

### 2019-11-13 paper/paper.tex
discussion

@@
 
-Forward and cross-decomposition (CCA, PLS) models are regularly used to identify the causal contribution of collinear features onto multi-dimensional observations (e.g.
+Forward and cross-decomposition models have been used to identify the causal contribution of collinear features onto multi-dimensional observations (e.g.
 \citep{naselaris2011encoding}).
@@
 
-Furthermore, B2B closely relates to CCA \citep{cca_hotelling}, but departs from it in four main aspects.
-First, B2B is not symmetric between $X$ and $Y$: it aims to identify specific causal features by first optimizing over the decoders $G$ and then over $H$.
+Finally, CCA has been used in neuroimaging for a variety of purposes such as denoising and subject alignment \citep{cca_hotelling, de2019multiway}.
+While CCA relates to B2B, these two methods diverge in several ways.
+First, CCA and B2B have different objectives: CCA aims to find the potentially numerous and poorly interpretable components where X and Y are maximally correlated, whereas B2B aims to recover the causal factors from X to Y.
+Second, B2B is not symmetric between $X$ and $Y$: it aims to identify specific causal features by first optimizing over the decoders $G$ and then over $H$.
 By contrast, CCA is symmetric between $X$ and $Y$, and aims to find $G$ and $H$ such that they project $X$ and $Y$ on maximally correlated dimensions.
-Second, CCA is based an eigen decomposition of $XH$ and $YG$ - the corresponding canonical components are thus mixing the $X$ features in way that limit interpretability.
-Third, CCA does not typically optimize two distinct regularization parameters.
+Third, CCA is based an eigen decomposition of $XH$ and $YG$ - the corresponding canonical components are thus mixing the $X$ features in way that limit interpretability and potentially dilute the impact of each feature onto multiple components.
+In contrast B2B assesses each feature $X^j$ on a single $Y$ component specifically selected to maximize signal-to-noise ratio of that feature $j$.
+Fourth, and unlike B2B, CCA does not separately optimize two distinct regularization parameters for $G$ and $H$.
 Finally, CCA does not use different data splits to estimate $G$ and $H$.
-Together, these differences may explain why B2B reliably outperform CCA on estimating causal influences (Figs.
-\ref{fig:percondition} and \ref{fig:auc_plots}).
-Extensions of CCA have been explored in neuroimaging (see \citep{de2019multiway} for brief review), but typically do not address the issue of causal discovery.
+Together, these differences may explain why B2B reliably outperform CCA on estimating causal influences (Figs. \ref{fig:percondition} and \ref{fig:auc_plots}).
 

### 2019-11-14 paper/appendix.tex
fixes

@@
 
-Consequently, assuming that $F$ is full rank on $Img(XE)$, B2B yields the same solutions to equations \ref{eq:model} and \ref{model_noisy_measure}.
+Consequently, assuming that $F$ is full rank on $Img(XE)$, B2B yields the same solutions to equations \ref{eq:model} and \ref{eq:model_noisy_measure}.
 
@@
 \subsection{Recovering E}
+\label{recovering}
 
@@
 First, given known signal-to-noise ratio,the threshold above which a feature should considered to be causal can be derived analytically.
-Indeed, Equation $\ref:diagk$ implies that the $k$ first diagonal elements of $\hat H$ are bounded:  % we assume that $F$ has full rank over $Img(E)$.
+Indeed, Equation $\ref{eq:diagk}$ implies that the $k$ first diagonal elements of $\hat H$ are bounded:  % we assume that $F$ has full rank over $Img(E)$.
 
@@
 
-Second, $\text{diag{\hat H}}$ can be binarized with the Sonquist-Morgan criterion \citep{sonquist_morgan}, a non-parametric clustering procedure separating small and large values in a given set.
+Second, $\text{diag}(\hat H)$ can be binarized with the Sonquist-Morgan criterion \citep{sonquist_morgan}, a non-parametric clustering procedure separating small and large values in a given set.
 This procedure maximizes the ratio of inter-group variance while minimizing the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
@@
 These additional features corresponds to the first dimensions of word embedding as provided by Spacy \citep{spacy2}.
-The results shown in Figure \ref{fig:embbedings}, show that the feature importance of ad-hoc features as derived by B2B remain unchanged (and are actually improved).
+The results shown in Figure \ref{fig:embeddings}, show that the feature importance of ad-hoc features as derived by B2B remain unchanged and are actually improved.
 

### 2019-11-14 paper/introduction.tex
fixes

@@
 The present paper focuses on the restricted issue of disentangling the causal influence of linearly correlated predictors ($X$) onto noisy multivariate observations ($Y$).
-The present approach thus differs from other causal discovery algorithms based on temporal-delays and/or nonlinear interaction in systems where the directionality of causation (from X to Y or vice versa) is unknown (e.g.
+The present approach thus differs from other causal discovery algorithms based on temporal-delays and/or nonlinear interactions in systems where the directionality of causation (from X to Y or vice versa) is unknown (e.g.
 \citep{peters2017elements, granger1969investigating, janzing2013quantifying, scholkopf2016modeling}.

### 2019-11-14 paper/paper.tex
fixes

@@
 
-Assuming, without loss of generality, that the active features in $E$ are the $k \in \mathbb{Z}: k \in [0, d_x]$ first features, and rewriting $X=(X_1,X_2)$ and $N=(N_1,N_2)$ ($X_1$ and $N_1$ containing the $k$ first features), we have ($\Sigma_{A B}$ denoting the covariance of $A$ and $B$)
+Assuming, without loss of generality, that the active features in $E$ are the $k \in \mathbb{Z}: k \in [0, d_x]$ first features, and rewriting $X=(X_1,X_2)$ and $N=(N_1,N_2)$ ($X_1$ and $N_1$ containing the $k$ first features), we have:
 
@@
 \end{equation}
-% $$
+
+where $\Sigma_{A B}$ is the covariance of $A$ and $B$, and:
 % and
 \begin{equation}
-  \hat H = \left(\begin{array}{cc}(\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1} & (\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_2} \\0 & 0\end{array}\right),
+  \hat H = \left(\begin{array}{cc}(\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1} & (\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_2} \\0 & 0\end{array}\right)
 \end{equation}
 \begin{equation}
-  \text{diag}_k (\hat H) = \text{diag}((\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1}) = \text{diag}((I+\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1})^{-1}).
+  \text{diag}_k (\hat H) = \text{diag}((\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1}) = \text{diag}((I+\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1})^{-1})
   \label{eq:diagk}
@@
 
-Otherwise, with noise, the causal factors of $E$ correspond to the positive elements of $\test{diag}(\hatH)$. The methods to recover them are presented in Appendix XXXX$$$
+In the presence of noise, the causal factors of $E$ correspond to the positive elements of $\text{diag}(\hat H)$. The methods to recover them are presented in the Appendix \ref{recovering}.
 
@@
 
-We perform two experiments to evaluate B2B: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography (MEG) dataset.
+We perform two sets of experiments to evaluate B2B: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography (MEG) dataset.
 %

### 2020-03-04 paper/paper.tex
hard rewrap, causal terminology

@@
 \usepackage[ruled, linesnumbered]{algorithm2e}
-\usepackage{iclr2020_conference,times,wrapfig}
+\usepackage{times,wrapfig}
+\usepackage{authblk}
+\usepackage{apacite}
+\usepackage{natbib}
+\usepackage{biblatex}
 
@@
 
-\title{Measuring causal influence with\\ back-to-back regression: the linear case}
+\title{Discriminating causes \\from multivariate observations:\\the back-to-back regression}
 
-\author{%
-  Jean-Remi King\\
-  CNRS\\
-  \texttt{jeanremi.king@gmail.com} \\
-  \And
-  Fran\c{c}ois Charton\\
-  Facebook AI\\
-  \texttt{fcharton@fb.com}\\
-  \And
-  David Lopez-Paz\\
-  Facebook AI\\
-  \texttt{dlp@fb.com}
-  \And
-  Maxime Oquab\\
-  Facebook AI\\
-  \texttt{qas@fb.com}
-}
+\author[1, 2]{Jean-Rémi King}
+\author[2]{Fran\c{c}ois Charton}
+\author[2]{David Lopez-Paz}
+\author[2]{Maxime Oquab}
+\affil[1]{Laboratoire des systèmes perceptifs, PSL University, CNRS}
+\affil[2]{Facebook AI}
+\date{}
+
+% \author{%
+%   Jean-Rémi King\\
+%   CNRS\\
+%   \texttt{jeanremi.king@gmail.com} \\
+%   \And
+%   Fran\c{c}ois Charton\\
+%   Facebook AI\\
+%   \texttt{fcharton@fb.com}\\
+%   \And
+%   David Lopez-Paz\\
+%   Facebook AI\\
+%   \texttt{dlp@fb.com}
+%   \And
+%   Maxime Oquab\\
+%   Facebook AI\\
+%   \texttt{qas@fb.com}
+% }
 
@@
 \begin{abstract}
-Identifying causes from observations can be particularly challenging when i) potential factors are difficult to manipulate individually and ii) observations are complex and multi-dimensional.
-To address this issue, we introduce ``Back-to-Back'' regression (B2B), a
-method designed to efficiently measure, from a set of co-varying factors, the causal
-influences that most plausibly account for multidimensional observations. After
-proving the consistency of B2B and its links to other linear approaches, we show that our method outperforms
-least-squares regression and cross-decomposition techniques (e.g.
-canonical correlation analysis and partial least squares) on causal
-identification. Finally, we apply B2B to
-neuroimaging recordings of 102 subjects reading word sequences. The results
-show that the early and late brain representations, caused by low- and high-level
-word features respectively, are more reliably detected with B2B than with other standard techniques.
+Identifying causes solely from observations can be particularly challenging when
+i) potential factors are difficult to manipulate individually and ii)
+observations are complex and multi-dimensional. To address this issue, we
+introduce ``Back-to-Back'' regression (B2B), a method designed to efficiently
+measure, from a set of colinear factors, those that most plausibly account for
+multidimensional observations. After proving the consistency of B2B and its
+links to other linear approaches, we show that our method outperforms
+least-squares regression and cross-decomposition techniques (e.g. canonical
+correlation analysis and partial least squares) on causal identification.
+Finally, we apply B2B to neuroimaging recordings of 102 subjects reading word
+sequences. The results show that the early and late brain representations,
+caused by low- and high-level word features respectively, are more reliably
+detected with B2B than with other standard techniques.
 
@@
 
-\begin{figure}[t!]
-    \centering
-    \begin{tikzpicture}
-    \newcommand\posY{0}
-    \newcommand\posX{3}
-    \newcommand\posE{5}
-    \newcommand\posN{7}
-    \newcommand\posF{10}
+\include{fig_problem.tex}
 
-    \node[thick, draw=black, minimum height=3cm, minimum width=2cm, fill=yellow!10] (Y) at (\posY, 0){};
-    \node[] (eq) at (1.5, 0){$=$};
-    \node[] (times) at (4, 0){$\times$};
-    \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=yellow!10] (X) at (\posX, 0){};
-    \node[thick, draw=black, minimum height=1cm, minimum width=1cm, fill=red!10] (E) at (\posE, 0){};
-
-    \draw[fill=white] (\posE - 0.5 + 0.0, 0.5 - 0.0) rectangle (\posE - 0.5 + 0.0 + 0.1, 0.5 - 0.0 - 0.1);
-    \draw[fill=black] (\posE - 0.5 + 0.1, 0.5 - 0.1) rectangle (\posE - 0.5 + 0.1 + 0.1, 0.5 - 0.1 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.2, 0.5 - 0.2) rectangle (\posE - 0.5 + 0.2 + 0.1, 0.5 - 0.2 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.3, 0.5 - 0.3) rectangle (\posE - 0.5 + 0.3 + 0.1, 0.5 - 0.3 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.4, 0.5 - 0.4) rectangle (\posE - 0.5 + 0.4 + 0.1, 0.5 - 0.4 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.5, 0.5 - 0.5) rectangle (\posE - 0.5 + 0.5 + 0.1, 0.5 - 0.5 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.6, 0.5 - 0.6) rectangle (\posE - 0.5 + 0.6 + 0.1, 0.5 - 0.6 - 0.1);
-    \draw[fill=black] (\posE - 0.5 + 0.7, 0.5 - 0.7) rectangle (\posE - 0.5 + 0.7 + 0.1, 0.5 - 0.7 - 0.1);
-    \draw[fill=white] (\posE - 0.5 + 0.8, 0.5 - 0.8) rectangle (\posE - 0.5 + 0.8 + 0.1, 0.5 - 0.8 - 0.1);
-    \draw[fill=black] (\posE - 0.5 + 0.9, 0.5 - 0.9) rectangle (\posE - 0.5 + 0.9 + 0.1, 0.5 - 0.9 - 0.1);
-
-    \node[] (plus) at (6, 0){$+$};
-    \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=blue!10] (N) at (\posN, 0){};
-    \node[thick, draw=black, minimum height=1cm, minimum width=2cm, fill=red!10] (F) at (\posF, 0){};
-    \draw [thick] (2.25, -1.5) to [round left paren ] (2.25, 1.5);
-    \draw [thick] (7.75, -1.5) to [round right paren ] (7.75, 1.5);
-
-    \node[] (times2) at (8.5, 0){$\times$};
-
-    \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{n \times d_y}$}};
-    \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{n \times d_x}$}};
-    \node[] (annE) at (\posE, -1.8){\scalebox{0.85}{$E \in \mathbb{D}^{d_x \times d_x}$}};
-    \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{n \times d_x}$}};
-    \node[] (annF) at (\posF, -1.8){\scalebox{0.85}{$F \in \mathbb{R}^{d_x \times d_y}$}};
-
-    \node[] (labY) at (\posY, 2){Observations};
-    \node[] (labX) at (\posX, 2){Factors};
-    \node[] (labE) at (\posE, 2){Cause};
-    \node[] (labE) at (\posE, 1.7){selection};
-    \node[] (labN) at (\posN, 2){Noise};
-    \node[] (labF) at (\posF, 2){Cause-effect};
-    \node[] (labF) at (\posF, 1.7){mapping};
-
-    \node[] (sim1) at (0,-3.25) {\scalebox{0.85}{$X \sim P(X)$}};
-    \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
-    \node[] (reg1) at (5.5,-3.25) {$\hat{E} = \text{diag}(\underbrace{(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}})$};
-    \end{tikzpicture}
-    \caption{Back-to-back regression identifies the subset of factors $E_{ii} = 1$ in $X$ that influence some observations $Y$ by 1) regressing from $Y$ to $X$ to obtain $\hat{X}$, and 2) returning the diagonal of the regression coefficients from $X$ to $\hat{X}$.}
-    \label{fig:b2b}
-\end{figure}
-
-We consider the measurement of multivariate signal $Y \in \mathbb{R}^{n \times d_y}$, generated from a set of putative causes $X \in \mathbb{R}^{n \times d_x}$, via some unknown linear apparatus $F \in \mathbb{R}^{d_x \times d_y}$.
+We consider the measurement of multivariate signal $Y \in \mathbb{R}^{n \times
+d_y}$, generated from a set of putative causes $X \in \mathbb{R}^{n \times
+d_x}$, via some unknown linear apparatus $F \in \mathbb{R}^{d_x \times d_y}$.
 %
@@
 %
-By considering a square binary diagonal matrix of \emph{causal influences} $E \in \mathbb{D}^{d_x \times d_x}$, we denote by $XE$ the causal factors of $Y$.
+By considering a square binary diagonal matrix of \emph{causal influences} $E
+\in \mathbb{D}^{d_x \times d_x}$, we denote by $XE$ the causal factors of $Y$.
 %
@@
 %
-where $i$ is a given sample, and $n_i$ is a sample-specific noise drawen from a centered distribution.
+where $i$ is a given sample, and $n_i$ is a sample-specific noise drawen from a
+centered distribution.
 %
-While the triplet of variables $X$ and $N$ are independent, we allow each of them to have any form of covariance.
+While the triplet of variables $X$ and $N$ are independent, we allow each of
+them to have any form of covariance.
 % what is a general covariance matrix?
@@
 %
-This problem space, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.
+This problem space, along with the sizes of all variables involved, is
+illustrated in Figure~\ref{fig:b2b}.
 %
-Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back Regression (B2B) is to estimate the matrix of causal influences $E$.
+Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back
+Regression (B2B) is to estimate the matrix of causal influences $E$.
 
@@
 %
-First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$, and construct the predictions $\hat X = Y \hat G$.
+First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$,
+and construct the predictions $\hat X = Y \hat G$.
 %
-This backward regression recovers the correlations between $Y$ and each factor of $X$.
+This backward regression recovers the correlations between $Y$ and each factor
+of $X$.
 %
-Second, we estimate the linear regression coefficients $\hat H$ from $X$ to $\hat X$.
+Second, we estimate the linear regression coefficients $\hat H$ from $X$ to
+$\hat X$.
 %
-The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{diag}(\hat{H})$, is the desired estimate of the causal influence matrix $E$, as detailed in the \ref{appendix:theorem_proof}.
+The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} =
+\text{diag}(\hat{H})$, is the desired estimate of the causal influence matrix
+$E$, as detailed in the \ref{appendix:theorem_proof}.
 
-If using l2-regularized least-squares \citep{hoerl1959optimum, rifkin2007notes}, B2B has a closed form solution:
+If using l2-regularized least-squares \citep{hoerl1959optimum, rifkin2007notes},
+B2B has a closed form solution:
 \begin{align}
@@
 %
-where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices of regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.
+where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices of regularization
+parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are
+ill-conditioned.
 
-Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
+Performing two regressions over the same data sample can result in overfitting,
+as spurious correlations in the data absorbed by the first regression will be
+leveraged by the second one.
 %
-To avoid this issue, we split our sample $(X, Y)$ into two splits $(X_1, Y_1)$ and $(X_2, Y_2)$.
+To avoid this issue, we split our sample $(X, Y)$ into two splits $(X_1, Y_1)$
+and $(X_2, Y_2)$.
 %
-Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
+Then, the first regression is performed using $(X_1, Y_1)$, and the second
+regression is performed using $(X_2, Y_2)$.
 %
-To compensate for the reduction in sample size caused by the split, B2B is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split \citep{breiman1996bagging}.
+To compensate for the reduction in sample size caused by the split, B2B is
+repeated over many random splits, and the final estimate $\hat E$ of the causal
+influence matrix is the average over the estimates associated to each split
+\citep{breiman1996bagging}.
 %
-To accelerate this ensembling procedure, we implemented an efficient leave-one-out cross-validation scheme as detailed in \citep{rifkin2007notes} % http://cbcl.mit.edu/publications/ps/MIT-CSAIL-TR-2007-025.pdf
-, as follows:
+To accelerate this ensembling procedure, we implemented an efficient
+leave-one-out cross-validation scheme as detailed in \citep{rifkin2007notes} %
+http://cbcl.mit.edu/publications/ps/MIT-CSAIL-TR-2007-025.pdf, as follows:
 %
@@
 
-where $\Sigma_X$ is the $X$ kernel matrix and where $G$ is computed with an eigen decomposition of $X$:
+where $\Sigma_X$ is the $X$ kernel matrix and where $G$ is computed with an
+eigen decomposition of $X$:
 \begin{equation}
@@
 \end{equation}
-where $Q$, $V$ and $\lambda$ are the eigen vectors, eigen values and regularization, respectively.
+where $Q$, $V$ and $\lambda$ are the eigen vectors, eigen values and
+regularization, respectively.
 
@@
 %
-The rest of this section provides a theoretical guarantee on the correctness of B2B.
+The rest of this section provides a theoretical guarantee on the correctness of
+B2B.
 
@@
 
-     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full rank noise.
+     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full
+     rank noise.
      %
-     Let $Img(M)$ refers to the image of the matrix $M$. If $F$ and $X$ are full-rank on the $Img(E)$, then, the solution of B2B, $\hat H$, will minimize
+     Let $Img(M)$ refers to the image of the matrix $M$. If $F$ and $X$ are
+     full-rank on the $Img(E)$, then, the solution of B2B, $\hat H$, will
+     minimize
      %
@@
 
-Assuming, without loss of generality, that the active features in $E$ are the $k \in \mathbb{Z}: k \in [0, d_x]$ first features, and rewriting $X=(X_1,X_2)$ and $N=(N_1,N_2)$ ($X_1$ and $N_1$ containing the $k$ first features), we have:
+Assuming, without loss of generality, that the active features in $E$ are the $k
+\in \mathbb{Z}: k \in [0, d_x]$ first features, and rewriting $X=(X_1,X_2)$ and
+$N=(N_1,N_2)$ ($X_1$ and $N_1$ containing the $k$ first features), we have:
 
@@
 %
-In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so $\text{diag}_k(\hat H)=I$, and $$\text{diag}(\hat H) = \text{diag}(E)$$ Therefore, we recover $E$ from $\hat H$.
+In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so
+$\text{diag}_k(\hat H)=I$, and $$\text{diag}(\hat H) = \text{diag}(E)$$
+Therefore, we recover $E$ from $\hat H$.
 
-In the presence of noise, the causal factors of $E$ correspond to the positive elements of $\text{diag}(\hat H)$. The methods to recover them are presented in the Appendix \ref{recovering}.
+In the presence of noise, the causal factors of $E$ correspond to the positive
+elements of $\text{diag}(\hat H)$. The methods to recover them are presented in
+the Appendix \ref{recovering}.
 
@@
 
-We perform two sets of experiments to evaluate B2B: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography (MEG) dataset.
+We perform two sets of experiments to evaluate B2B: one on controlled synthetic
+data, and a second one on a real, large-scale magneto-encephalography (MEG)
+dataset.
 %
-We use scikit-learn's PLS and RidgeCV \citep{sklearn} as well as pyrcca's regularized canonical component analysis (RegCCA, \citep{bilenko2016pyrcca}) objects to compare B2B against the standard baselines.
+We use scikit-learn's PLS and RidgeCV \citep{sklearn} as well as pyrcca's
+regularized canonical component analysis (RegCCA, \citep{bilenko2016pyrcca})
+objects to compare B2B against the standard baselines.
 
@@
 
-Forward and cross-decomposition models have been used to identify the causal contribution of collinear features onto multi-dimensional observations (e.g.
-\citep{naselaris2011encoding}).
-These approaches typically lead to multiple coefficients for each features (i.e.
-one per dimension of $Y$ or one per component respectively).
-Furthermore, these coefficients can be difficult to summarize into a single causal estimate.
-By contrast, B2B quickly (Fig.
-\ref{fig:duration}) leads to a single unbiased scalar values $\hat E$ tending towards 1 and 0 for causal and non-causal features respectively.
+Forward and cross-decomposition models have been used to identify the causal
+contribution of collinear features onto multi-dimensional observations (e.g.
+\citep{naselaris2011encoding}). These approaches typically lead to multiple
+coefficients for each features (i.e. one per dimension of $Y$ or one per
+component respectively). Furthermore, these coefficients can be difficult to
+summarize into a single causal estimate. By contrast, B2B quickly (Fig.
+\ref{fig:duration}) leads to a single unbiased scalar values $\hat E$ tending
+towards 1 and 0 for causal and non-causal features respectively.
 
-A variety of other statistical methods applied to neuroimaging data have been proposed to clarify what is being represented in brain responses - i.e.
-what feature causes specific brain activity.
-One of the popular linear method is Representational Similarity Analysis (RSA) \citep{kriegeskorte2008representational}, and consists in analyzing the similarity of brain responses associated with specific categorical conditions (e.g.
-distinct images), by (1) fitting one-against-all classifiers on each condition and (2) testing whether these classifiers can discriminate all other conditions.
-The resulting confusion matrix is then analyzed in an unsupervised manner to reveal which conditions lead to similar brain activity patterns.
-B2B differs from RSA in that (1) it uses regressions instead of classifications, and can thus generalize to new items and new contexts and (2) it is fully supervised.
+A variety of other statistical methods applied to neuroimaging data have been
+proposed to clarify what is being represented in brain responses - i.e. what
+feature causes specific brain activity. One of the popular linear method is
+Representational Similarity Analysis (RSA)
+\citep{kriegeskorte2008representational}, and consists in analyzing the
+similarity of brain responses associated with specific categorical conditions
+(e.g. distinct images), by (1) fitting one-against-all classifiers on each
+condition and (2) testing whether these classifiers can discriminate all other
+conditions. The resulting confusion matrix is then analyzed in an unsupervised
+manner to reveal which conditions lead to similar brain activity patterns. B2B
+differs from RSA in that (1) it uses regressions instead of classifications, and
+can thus generalize to new items and new contexts and (2) it is fully
+supervised.
 
-Finally, CCA has been used in neuroimaging for a variety of purposes such as denoising and subject alignment \citep{cca_hotelling, de2019multiway}.
-While CCA relates to B2B, these two methods diverge in several ways.
-First, CCA and B2B have different objectives: CCA aims to find the potentially numerous and poorly interpretable components where X and Y are maximally correlated, whereas B2B aims to recover the causal factors from X to Y.
-Second, B2B is not symmetric between $X$ and $Y$: it aims to identify specific causal features by first optimizing over the decoders $G$ and then over $H$.
-By contrast, CCA is symmetric between $X$ and $Y$, and aims to find $G$ and $H$ such that they project $X$ and $Y$ on maximally correlated dimensions.
-Third, CCA is based an eigen decomposition of $XH$ and $YG$ - the corresponding canonical components are thus mixing the $X$ features in way that limit interpretability and potentially dilute the impact of each feature onto multiple components.
-In contrast B2B assesses each feature $X^j$ on a single $Y$ component specifically selected to maximize signal-to-noise ratio of that feature $j$.
-Fourth, and unlike B2B, CCA does not separately optimize two distinct regularization parameters for $G$ and $H$.
-Finally, CCA does not use different data splits to estimate $G$ and $H$.
-Together, these differences may explain why B2B reliably outperform CCA on estimating causal influences (Figs. \ref{fig:percondition} and \ref{fig:auc_plots}).
+Finally, CCA has been used in neuroimaging for a variety of purposes such as
+denoising and subject alignment \citep{cca_hotelling, de2019multiway}. While CCA
+relates to B2B, these two methods diverge in several ways. First, CCA and B2B
+have different objectives: CCA aims to find the potentially numerous and poorly
+interpretable components where X and Y are maximally correlated, whereas B2B
+aims to recover the causal factors from X to Y. Second, B2B is not symmetric
+between $X$ and $Y$: it aims to identify specific causal features by first
+optimizing over the decoders $G$ and then over $H$. By contrast, CCA is
+symmetric between $X$ and $Y$, and aims to find $G$ and $H$ such that they
+project $X$ and $Y$ on maximally correlated dimensions. Third, CCA is based an
+eigen decomposition of $XH$ and $YG$ - the corresponding canonical components
+are thus mixing the $X$ features in way that limit interpretability and
+potentially dilute the impact of each feature onto multiple components. In
+contrast B2B assesses each feature $X^j$ on a single $Y$ component specifically
+selected to maximize signal-to-noise ratio of that feature $j$. Fourth, and
+unlike B2B, CCA does not separately optimize two distinct regularization
+parameters for $G$ and $H$. Finally, CCA does not use different data splits to
+estimate $G$ and $H$. Together, these differences may explain why B2B reliably
+outperform CCA on estimating causal influences (Figs. \ref{fig:percondition} and
+\ref{fig:auc_plots}).
 

### 2020-03-04 paper/__latexindent_temp.tex
clean up

@article{cardinali2013data,
  title={Data assimilation: Observation influence diagnostic of a data assimilation system},
  author={Cardinali, C},
  journal={Shinfield Park, England: European Centre for Medium-Range Weather Forecasts: Research Department},
  year={2013}
}

### 2020-03-04 paper/appendix.tex
clean up

@@
 
-     Consider the B2B model from equation \ref{eq:model} $$Y = (XE + N)F$$ with $N$ centred and full rank noise.
+     Consider the B2B model from equation \ref{eq:model} $$Y = (XE + N)F$$ with
+     $N$ centered and full rank noise.
 
-     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$ minimizes
+     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat
+     H$ minimizes
 
@@
 
- Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B.
+ Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions
+ of B2B.
 
@@
 
-Samely, for $\hat H$, we have
+In the same way, for $\hat H$, we have
 \begin{align*}
@@
 
-Let $F^\dagger$ be the pseudo inverse of $F$, and $Z=F^\dagger EF\hat G$, we have $FZ = FF^\dagger EF \hat G$
+Let $F^\dagger$ be the pseudo inverse of $F$, and $Z=F^\dagger EF\hat G$, we
+have $FZ = FF^\dagger EF \hat G$
 
-Since $F$ is full rank on $Img(E)$, we have $FF^\dagger E =E$, and $FZ = EF\hat G$
+Since $F$ is full rank on $Img(E)$, we have $FF^\dagger E =E$, and $FZ = EF\hat
+G$
 
-As $E$ is a binary diagonal matrix, it is an orthogonal projection and therefore a contraction, thus
+As $E$ is a binary diagonal matrix, it is an orthogonal projection and therefore
+a contraction, thus
  $$ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$$ and
@@
 
-Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$, since $E$, a binary diagonal matrix, is involutive. This completes the proof.
+Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$, since $E$, a binary
+diagonal matrix, is involutive. This completes the proof.
 \end{proof}
@@
 
-Equation \ref{eq:model} does not explicitly contain a measurement noise term. Yet, in moast experimental cases, the problem is best described as:
+Equation \ref{eq:model} does not explicitly contain a measurement noise term.
+Yet, in most experimental cases, the problem is best described as:
 \begin{equation}
@@
 
-This equation is actually equivalent to Equation \ref{eq:model} given our hypotheses. Indeed, we can rewrite $M = MF^{-1}F$ over $Img(F)$, which leads to:
+This equation is actually equivalent to Equation \ref{eq:model} given our
+hypotheses. Indeed, we can rewrite $M = MF^{-1}F$ over $Img(F)$, which leads to:
 $$Y = (XE+N)F+M = (XE+N+MF^{-1})F = (XE+N')F$$
 
-Consequently, assuming that $F$ is full rank on $Img(XE)$, B2B yields the same solutions to equations \ref{eq:model} and \ref{eq:model_noisy_measure}.
+Consequently, assuming that $F$ is full rank on $Img(XE)$, B2B yields the same
+solutions to equations \ref{eq:model} and \ref{eq:model_noisy_measure}.
 
@@
 
-For the Backward Model, feature importance cannot be assessed because there is no prediction.
+For the Backward Model, feature importance cannot be assessed because there is
+no prediction.
 
@@
 
-In case of noise, B2B yields non binary $\hat E$. Three thresholding rules can be used to binarize its values thus explicitly recover causal features.
+In case of noise, B2B yields non binary $\hat E$. Three thresholding rules can
+be used to binarize its values thus explicitly recover "causal" features.
 
-First, given known signal-to-noise ratio,the threshold above which a feature should considered to be causal can be derived analytically.
-Indeed, Equation $\ref{eq:diagk}$ implies that the $k$ first diagonal elements of $\hat H$ are bounded:  % we assume that $F$ has full rank over $Img(E)$.
+First, given known signal-to-noise ratio, the threshold above which a feature
+should considered to be "causal" can be derived analytically. Indeed, Equation
+$\ref{eq:diagk}$ implies that the $k$ first diagonal elements of $\hat H$ are
+bounded:  % we assume that $F$ has full rank over $Img(E)$.
 
-$$0 \leq \frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}} \leq \text{diag}_k(\hat{H})\leq  \frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$$
-where $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X_1 X_1}$ and $\Sigma_{N_1 N_1}$.
+$$0 \leq \frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}} \leq
+\text{diag}_k(\hat{H})\leq  \frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$$
+where $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote
+the largest and smallest eigenvalues of $\Sigma_{X_1 X_1}$ and $\Sigma_{N_1
+N_1}$.
 
-The average value $\mu$ of non-zero coefficients of $\text{diag}(\hat H)$ is the trace of $\hat H$ divided by $k$, and can be computed as
+The average value $\mu$ of non-zero coefficients of $\text{diag}(\hat H)$ is the
+trace of $\hat H$ divided by $k$, and can be computed as
 \begin{equation}
@@
 
-The decision threshold between causal and non-causal elements is thus a fraction $\mu$, whose proportion arbitrarily depends on the necessity to favor type I and type II errors.
-In practice, we cannot use this procedure for our MEG study, because signal-to-noise ratio is unknown.
+The decision threshold between "causal" and "non-causal" elements is thus a fraction
+$\mu$, whose proportion arbitrarily depends on the necessity to favor type I and
+type II errors. In practice, we cannot use this procedure for our MEG study,
+because signal-to-noise ratio is unknown.
 
-Second, $\text{diag}(\hat H)$ can be binarized with the Sonquist-Morgan criterion \citep{sonquist_morgan}, a non-parametric clustering procedure separating small and large values in a given set.
-This procedure maximizes the ratio of inter-group variance while minimizing the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
-Let $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, Sonquist-Morgan criterion maximizes \citep{kass75}:
+Second, $\text{diag}(\hat H)$ can be binarized with the Sonquist-Morgan
+criterion \citep{sonquist_morgan}, a non-parametric clustering procedure
+separating small and large values in a given set. This procedure maximizes the
+ratio of inter-group variance while minimizing the intra-group variance, over
+all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest
+values. Let $m_0$ and $m_1$ be the average values of the two clusters, $p$ and
+$d_x-p$ their size, and $v$ the total variance of the sample, Sonquist-Morgan
+criterion maximizes \citep{kass75}:
 \begin{equation}
@@
 
-This procedure assumes that there exists at least one causal and at least one non-causal feature.
+This procedure assumes that there exists at least one causal and at least one
+non-causal feature.
 %
-Third, second-order statistics across multiple datasets can be used to identify the elements of $\text{diag}(\hat H)$ that are sgnificantly different from 0.
+Third, second-order statistics across multiple datasets can be used to identify
+the elements of $\text{diag}(\hat H)$ that are significantly different from 0.
 This procedure is detailed in the method section of our MEG experiment.
 
-Overall, these three procedure thus varies in their additional assumptions: i.e.
-(1) a known signal-to-noise ratio, (2) the existence of both causal and non-causal factors or (3) independent repetitions of the experiment.
+Overall, these three procedures thus vary in their additional assumptions: i.e.
+(1) a known signal-to-noise ratio, (2) the existence of both causal and
+non-causal factors or (3) independent repetitions of the experiment.
 
@@
   \includegraphics[width=\linewidth]{figures/delta_auc.pdf}
-  \caption{Synthetic experiments.
-Distribution (over conditions) of AUC (top) and Feature Importance $\Delta R$ (bottom) metrics between our method (y-axis) and the baselines (x-axis).
-Each dot is a distinct synthetic experiment.
-Dots below the diagonal indicates that B2B outperform the tested model.
-\label{fig:auc_plots}}
+  \caption{Synthetic experiments. Distribution (over conditions) of AUC (top) and Feature
+  Importance $\Delta R$ (bottom) metrics between our method (y-axis) and the baselines
+  (x-axis). Each dot is a distinct synthetic experiment. Dots below the diagonal indicates
+  that B2B outperform the tested model. \label{fig:auc_plots}}
 \end{figure}
@@
   \includegraphics[width=0.7\linewidth]{figures/duration}
-  \caption{Wall-clock run-time for our method B2B and for the baselines. Each dot is a distinct synthetic experiment. B2B runs much faster than cross-decomposition baselines. \label{fig:duration}}
+  \caption{Wall-clock run-time for our method B2B and for the baselines. Each dot is a
+  distinct synthetic experiment. B2B runs much faster than cross-decomposition baselines.
+  \label{fig:duration}}
 \end{figure}
@@
 
-To test whether each of the methods robustly scales to an increasingly large number of potential causes $X$, we enhanced the four ad-hoc features (word length, word frequency, word function, dummy variable) with another ten features.
-These additional features corresponds to the first dimensions of word embedding as provided by Spacy \citep{spacy2}.
-The results shown in Figure \ref{fig:embeddings}, show that the feature importance of ad-hoc features as derived by B2B remain unchanged and are actually improved.
+To test whether each of the methods robustly scales to an increasingly
+large number of potential causes $X$, we enhanced the four ad-hoc features
+(word length, word frequency, word function, dummy variable) with another
+ten features. These additional features corresponds to the first dimensions
+of word embedding as provided by Spacy \citep{spacy2}. The results shown
+in Figure \ref{fig:embeddings}, show that the feature importance of ad-hoc
+features as derived by B2B remain unchanged and are actually improved.
 
@@
   \includegraphics[width=0.5\linewidth]{figures/compare_embeddings.pdf}
-  \caption{Comparison of $\Delta R$ when the models are tested on four variables (top) and when the models are tested on an these four variables as well as another 10 word-embedding features (bottom).
-These results illustrate that, unlike Regularized CCA, B2B remains robust even when the number of tested factors increases.
+  \caption{Comparison of $\Delta R$ when the models are tested on four
+  variables (top) and when the models are tested on an these four variables
+  as well as another 10 word-embedding features (bottom). These results
+  illustrate that, unlike Regularized CCA, B2B remains robust even when
+  the number of tested factors increases.
   \label{fig:embeddings}}

### 2020-03-04 paper/elsarticle-template/elsarticle-template.tex
clean up

\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\modulolinenumbers[5]

\journal{Journal of \LaTeX\ Templates}

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
%\bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{frontmatter}

\title{Elsevier \LaTeX\ template\tnoteref{mytitlenote}}
\tnotetext[mytitlenote]{Fully documented templates are available in the elsarticle package on \href{http://www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}{CTAN}.}

%% Group authors per affiliation:
\author{Elsevier\fnref{myfootnote}}
\address{Radarweg 29, Amsterdam}
\fntext[myfootnote]{Since 1880.}

%% or include affiliations in footnotes:
\author[mymainaddress,mysecondaryaddress]{Elsevier Inc}
\ead[url]{www.elsevier.com}

\author[mysecondaryaddress]{Global Customer Service\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
\ead{support@elsevier.com}

\address[mymainaddress]{1600 John F Kennedy Boulevard, Philadelphia}
\address[mysecondaryaddress]{360 Park Avenue South, New York}

\begin{abstract}
This template helps you to create a properly formatted \LaTeX\ manuscript.
\end{abstract}

\begin{keyword}
\texttt{elsarticle.cls}\sep \LaTeX\sep Elsevier \sep template
\MSC[2010] 00-01\sep  99-00
\end{keyword}

\end{frontmatter}

\linenumbers

\section{The Elsevier article class}

\paragraph{Installation} If the document class \emph{elsarticle} is not available on your computer, you can download and install the system package \emph{texlive-publishers} (Linux) or install the \LaTeX\ package \emph{elsarticle} using the package manager of your \TeX\ installation, which is typically \TeX\ Live or Mik\TeX.

\paragraph{Usage} Once the package is properly installed, you can use the document class \emph{elsarticle} to create a manuscript. Please make sure that your manuscript follows the guidelines in the Guide for Authors of the relevant journal. It is not necessary to typeset your manuscript in exactly the same way as an article, unless you are submitting to a camera-ready copy (CRC) journal.

\paragraph{Functionality} The Elsevier article class is based on the standard article class and supports almost all of the functionality of that class. In addition, it features commands and options to format the
\begin{itemize}
\item document style
\item baselineskip
\item front matter
\item keywords and MSC codes
\item theorems, definitions and proofs
\item lables of enumerations
\item citation style and labeling.
\end{itemize}

\section{Front matter}

The author names and affiliations could be formatted in two ways:
\begin{enumerate}[(1)]
\item Group the authors per affiliation.
\item Use footnotes to indicate the affiliations.
\end{enumerate}
See the front matter of this document for examples. You are recommended to conform your choice to the journal you are submitting to.

\section{Bibliography styles}

There are various bibliography styles available. You can select the style of your choice in the preamble of this document. These styles are Elsevier styles based on standard styles like Harvard and Vancouver. Please use Bib\TeX\ to generate your bibliography and include DOIs whenever available.

Here are two sample references: \citep{Feynman1963118,Dirac1953888}.

\section*{References}

\bibliography{mybibfile}

\end{document}

### 2020-03-04 paper/iclr2020_conference.tex
clean up

\documentclass{article} % For LaTeX2e
\usepackage{iclr2020_conference,times}

% Optional math commands from https://github.com/goodfeli/dlbook_notation.
\input{math_commands.tex}

\usepackage{hyperref}
\usepackage{url}

\title{Formatting Instructions for ICLR 2020 \\ Conference Submissions}

% Authors must not appear in the submitted version. They should be hidden
% as long as the \iclrfinalcopy macro remains commented out below.
% Non-anonymous submissions will be rejected without review.

\author{Antiquus S.~Hippocampus, Natalia Cerebro \& Amelie P. Amygdale \thanks{ Use footnote for providing further information
about author (webpage, alternative address)---\emph{not} for acknowledging
funding agencies.  Funding acknowledgements go at the end of the paper.} \\
Department of Computer Science\\
Cranberry-Lemon University\\
Pittsburgh, PA 15213, USA \\
\texttt{\{hippo,brain,jen\}@cs.cranberry-lemon.edu} \\
\And
Ji Q. Ren \& Yevgeny LeNet \\
Department of Computational Neuroscience \\
University of the Witwatersrand \\
Joburg, South Africa \\
\texttt{\{robot,net\}@wits.ac.za} \\
\AND
Coauthor \\
Affiliation \\
Address \\
\texttt{email}
}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

%\iclrfinalcopy % Uncomment for camera-ready version, but NOT for submission.
\begin{document}

\maketitle

\begin{abstract}
The abstract paragraph should be indented 1/2~inch (3~picas) on both left and
right-hand margins. Use 10~point type, with a vertical spacing of 11~points.
The word \textsc{Abstract} must be centered, in small caps, and in point size 12. Two
line spaces precede the abstract. The abstract must be limited to one
paragraph.
\end{abstract}

\section{Submission of conference papers to ICLR 2020}

ICLR requires electronic submissions, processed by
\url{https://openreview.net/}. See ICLR's website for more instructions.

If your paper is ultimately accepted, the statement {\tt
  {\textbackslash}iclrfinalcopy} should be inserted to adjust the
format to the camera ready requirements.

The format for the submissions is a variant of the NIPS format.
Please read carefully the instructions below, and follow them
faithfully.

\subsection{Style}

Papers to be submitted to ICLR 2020 must be prepared according to the
instructions presented here.

%% Please note that we have introduced automatic line number generation
%% into the style file for \LaTeXe. This is to help reviewers
%% refer to specific lines of the paper when they make their comments. Please do
%% NOT refer to these line numbers in your paper as they will be removed from the
%% style file for the final version of accepted papers.

Authors are required to use the ICLR \LaTeX{} style files obtainable at the
ICLR website. Please make sure you use the current files and
not previous versions. Tweaking the style files may be grounds for rejection.

\subsection{Retrieval of style files}

The style files for ICLR and other conference information are available online at:
\begin{center}
   \url{http://www.iclr.cc/}
\end{center}
The file \verb+iclr2020_conference.pdf+ contains these
instructions and illustrates the
various formatting requirements your ICLR paper must satisfy.
Submissions must be made using \LaTeX{} and the style files
\verb+iclr2020_conference.sty+ and \verb+iclr2020_conference.bst+ (to be used with \LaTeX{}2e). The file
\verb+iclr2020_conference.tex+ may be used as a ``shell'' for writing your paper. All you
have to do is replace the author, title, abstract, and text of the paper with
your own.

The formatting instructions contained in these style files are summarized in
sections \ref{gen_inst}, \ref{headings}, and \ref{others} below.

\section{General formatting instructions}
\label{gen_inst}

The text must be confined within a rectangle 5.5~inches (33~picas) wide and
9~inches (54~picas) long. The left margin is 1.5~inch (9~picas).
Use 10~point type with a vertical spacing of 11~points. Times New Roman is the
preferred typeface throughout. Paragraphs are separated by 1/2~line space,
with no indentation.

Paper title is 17~point, in small caps and left-aligned.
All pages should start at 1~inch (6~picas) from the top of the page.

Authors' names are
set in boldface, and each name is placed above its corresponding
address. The lead author's name is to be listed first, and
the co-authors' names are set to follow. Authors sharing the
same address can be on the same line.

Please pay special attention to the instructions in section \ref{others}
regarding figures, tables, acknowledgments, and references.

The recommended paper length is 8 pages, with unlimited additional pages for
citations.  There will be a strict upper limit of 10 pages for the main text.
Reviewers will be instructed to apply a higher standard to papers in excess of 8
pages.  Authors may use as many pages of appendices (after the bibliography) as
they wish, but reviewers are not required to read these.

\section{Headings: first level}
\label{headings}

First level headings are in small caps,
flush left and in point size 12. One line space before the first level
heading and 1/2~line space after the first level heading.

\subsection{Headings: second level}

Second level headings are in small caps,
flush left and in point size 10. One line space before the second level
heading and 1/2~line space after the second level heading.

\subsubsection{Headings: third level}

Third level headings are in small caps,
flush left and in point size 10. One line space before the third level
heading and 1/2~line space after the third level heading.

\section{Citations, figures, tables, references}
\label{others}

These instructions apply to everyone, regardless of the formatter being used.

\subsection{Citations within the text}

Citations within the text should be based on the \texttt{natbib} package
and include the authors' last names and year (with the ``et~al.'' construct
for more than two authors). When the authors or the publication are
included in the sentence, the citation should not be in parenthesis using \verb|\citet{}| (as
in ``See \citet{Hinton06} for more information.''). Otherwise, the citation
should be in parenthesis using \verb|\citep{}| (as in ``Deep learning shows promise to make progress
towards AI~\citep{Bengio+chapter2007}.'').

The corresponding references are to be listed in alphabetical order of
authors, in the \textsc{References} section. As to the format of the
references themselves, any style is acceptable as long as it is used
consistently.

\subsection{Footnotes}

Indicate footnotes with a number\footnote{Sample of the first footnote} in the
text. Place the footnotes at the bottom of the page on which they appear.
Precede the footnote with a horizontal rule of 2~inches
(12~picas).\footnote{Sample of the second footnote}

\subsection{Figures}

All artwork must be neat, clean, and legible. Lines should be dark
enough for purposes of reproduction; art work should not be
hand-drawn. The figure number and caption always appear after the
figure. Place one line space before the figure caption, and one line
space after the figure. The figure caption is lower case (except for
first word and proper nouns); figures are numbered consecutively.

Make sure the figure caption does not get separated from the figure.
Leave sufficient space to avoid splitting the figure and figure caption.

You may use color figures.
However, it is best for the
figure captions and the paper body to make sense if the paper is printed
either in black/white or in color.
\begin{figure}[h]
\begin{center}
%\framebox[4.0in]{$\;$}
\fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
\end{center}
\caption{Sample figure caption.}
\end{figure}

\subsection{Tables}

All tables must be centered, neat, clean and legible. Do not use hand-drawn
tables. The table number and title always appear before the table. See
Table~\ref{sample-table}.

Place one line space before the table title, one line space after the table
title, and one line space after the table. The table title must be lower case
(except for first word and proper nouns); tables are numbered consecutively.

\begin{table}[t]
\caption{Sample table title}
\label{sample-table}
\begin{center}
\begin{tabular}{ll}
\multicolumn{1}{c}{\bf PART}  &\multicolumn{1}{c}{\bf DESCRIPTION}
\\ \hline \\
Dendrite         &Input terminal \\
Axon             &Output terminal \\
Soma             &Cell body (contains cell nucleus) \\
\end{tabular}
\end{center}
\end{table}

\section{Default Notation}

In an attempt to encourage standardized notation, we have included the
notation file from the textbook, \textit{Deep Learning}
\citep{goodfellow2016deep} available at
\url{https://github.com/goodfeli/dlbook_notation/}.  Use of this style
is not required and can be disabled by commenting out
\texttt{math\_commands.tex}.

\centerline{\bf Numbers and Arrays}
\bgroup
\def\arraystretch{1.5}
\begin{tabular}{p{1in}p{3.25in}}
$\displaystyle a$ & A scalar (integer or real)\\
$\displaystyle \va$ & A vector\\
$\displaystyle \mA$ & A matrix\\
$\displaystyle \tA$ & A tensor\\
$\displaystyle \mI_n$ & Identity matrix with $n$ rows and $n$ columns\\
$\displaystyle \mI$ & Identity matrix with dimensionality implied by context\\
$\displaystyle \ve^{(i)}$ & Standard basis vector $[0,\dots,0,1,0,\dots,0]$ with a 1 at position $i$\\
$\displaystyle \text{diag}(\va)$ & A square, diagonal matrix with diagonal entries given by $\va$\\
$\displaystyle \ra$ & A scalar random variable\\
$\displaystyle \rva$ & A vector-valued random variable\\
$\displaystyle \rmA$ & A matrix-valued random variable\\
\end{tabular}
\egroup
\vspace{0.25cm}

\centerline{\bf Sets and Graphs}
\bgroup
\def\arraystretch{1.5}

\begin{tabular}{p{1.25in}p{3.25in}}
$\displaystyle \sA$ & A set\\
$\displaystyle \R$ & The set of real numbers \\
$\displaystyle \{0, 1\}$ & The set containing 0 and 1 \\
$\displaystyle \{0, 1, \dots, n \}$ & The set of all integers between $0$ and $n$\\
$\displaystyle [a, b]$ & The real interval including $a$ and $b$\\
$\displaystyle (a, b]$ & The real interval excluding $a$ but including $b$\\
$\displaystyle \sA \backslash \sB$ & Set subtraction, i.e., the set containing the elements of $\sA$ that are not in $\sB$\\
$\displaystyle \gG$ & A graph\\
$\displaystyle \parents_\gG(\ervx_i)$ & The parents of $\ervx_i$ in $\gG$
\end{tabular}
\vspace{0.25cm}

\centerline{\bf Indexing}
\bgroup
\def\arraystretch{1.5}

\begin{tabular}{p{1.25in}p{3.25in}}
$\displaystyle \eva_i$ & Element $i$ of vector $\va$, with indexing starting at 1 \\
$\displaystyle \eva_{-i}$ & All elements of vector $\va$ except for element $i$ \\
$\displaystyle \emA_{i,j}$ & Element $i, j$ of matrix $\mA$ \\
$\displaystyle \mA_{i, :}$ & Row $i$ of matrix $\mA$ \\
$\displaystyle \mA_{:, i}$ & Column $i$ of matrix $\mA$ \\
$\displaystyle \etA_{i, j, k}$ & Element $(i, j, k)$ of a 3-D tensor $\tA$\\
$\displaystyle \tA_{:, :, i}$ & 2-D slice of a 3-D tensor\\
$\displaystyle \erva_i$ & Element $i$ of the random vector $\rva$ \\
\end{tabular}
\egroup
\vspace{0.25cm}

\centerline{\bf Calculus}
\bgroup
\def\arraystretch{1.5}
\begin{tabular}{p{1.25in}p{3.25in}}
% NOTE: the [2ex] on the next line adds extra height to that row of the table.
% Without that command, the fraction on the first line is too tall and collides
% with the fraction on the second line.
$\displaystyle\frac{d y} {d x}$ & Derivative of $y$ with respect to $x$\\ [2ex]
$\displaystyle \frac{\partial y} {\partial x} $ & Partial derivative of $y$ with respect to $x$ \\
$\displaystyle \nabla_\vx y $ & Gradient of $y$ with respect to $\vx$ \\
$\displaystyle \nabla_\mX y $ & Matrix derivatives of $y$ with respect to $\mX$ \\
$\displaystyle \nabla_\tX y $ & Tensor containing derivatives of $y$ with respect to $\tX$ \\
$\displaystyle \frac{\partial f}{\partial \vx} $ & Jacobian matrix $\mJ \in \R^{m\times n}$ of $f: \R^n \rightarrow \R^m$\\
$\displaystyle \nabla_\vx^2 f(\vx)\text{ or }\mH( f)(\vx)$ & The Hessian matrix of $f$ at input point $\vx$\\
$\displaystyle \int f(\vx) d\vx $ & Definite integral over the entire domain of $\vx$ \\
$\displaystyle \int_\sS f(\vx) d\vx$ & Definite integral with respect to $\vx$ over the set $\sS$ \\
\end{tabular}
\egroup
\vspace{0.25cm}

\centerline{\bf Probability and Information Theory}
\bgroup
\def\arraystretch{1.5}
\begin{tabular}{p{1.25in}p{3.25in}}
$\displaystyle P(\ra)$ & A probability distribution over a discrete variable\\
$\displaystyle p(\ra)$ & A probability distribution over a continuous variable, or over
a variable whose type has not been specified\\
$\displaystyle \ra \sim P$ & Random variable $\ra$ has distribution $P$\\% so thing on left of \sim should always be a random variable, with name beginning with \r
$\displaystyle  \E_{\rx\sim P} [ f(x) ]\text{ or } \E f(x)$ & Expectation of $f(x)$ with respect to $P(\rx)$ \\
$\displaystyle \Var(f(x)) $ &  Variance of $f(x)$ under $P(\rx)$ \\
$\displaystyle \Cov(f(x),g(x)) $ & Covariance of $f(x)$ and $g(x)$ under $P(\rx)$\\
$\displaystyle H(\rx) $ & Shannon entropy of the random variable $\rx$\\
$\displaystyle \KL ( P \Vert Q ) $ & Kullback-Leibler divergence of P and Q \\
$\displaystyle \mathcal{N} ( \vx ; \vmu , \mSigma)$ & Gaussian distribution %
over $\vx$ with mean $\vmu$ and covariance $\mSigma$ \\
\end{tabular}
\egroup
\vspace{0.25cm}

\centerline{\bf Functions}
\bgroup
\def\arraystretch{1.5}
\begin{tabular}{p{1.25in}p{3.25in}}
$\displaystyle f: \sA \rightarrow \sB$ & The function $f$ with domain $\sA$ and range $\sB$\\
$\displaystyle f \circ g $ & Composition of the functions $f$ and $g$ \\
  $\displaystyle f(\vx ; \vtheta) $ & A function of $\vx$ parametrized by $\vtheta$.
  (Sometimes we write $f(\vx)$ and omit the argument $\vtheta$ to lighten notation) \\
$\displaystyle \log x$ & Natural logarithm of $x$ \\
$\displaystyle \sigma(x)$ & Logistic sigmoid, $\displaystyle \frac{1} {1 + \exp(-x)}$ \\
$\displaystyle \zeta(x)$ & Softplus, $\log(1 + \exp(x))$ \\
$\displaystyle || \vx ||_p $ & $\normlp$ norm of $\vx$ \\
$\displaystyle || \vx || $ & $\normltwo$ norm of $\vx$ \\
$\displaystyle x^+$ & Positive part of $x$, i.e., $\max(0,x)$\\
$\displaystyle \1_\mathrm{condition}$ & is 1 if the condition is true, 0 otherwise\\
\end{tabular}
\egroup
\vspace{0.25cm}

\section{Final instructions}
Do not change any aspects of the formatting parameters in the style files.
In particular, do not modify the width or length of the rectangle the text
should fit into, and do not change font sizes (except perhaps in the
\textsc{References} section; see below). Please note that pages should be
numbered.

\section{Preparing PostScript or PDF files}

Please prepare PostScript or PDF files with paper size ``US Letter'', and
not, for example, ``A4''. The -t
letter option on dvips will produce US Letter files.

Consider directly generating PDF files using \verb+pdflatex+
(especially if you are a MiKTeX user).
PDF figures must be substituted for EPS figures, however.

Otherwise, please generate your PostScript and PDF files with the following commands:
\begin{verbatim}
dvips mypaper.dvi -t letter -Ppdf -G0 -o mypaper.ps
ps2pdf mypaper.ps mypaper.pdf
\end{verbatim}

\subsection{Margins in LaTeX}

Most of the margin problems come from figures positioned by hand using
\verb+\special+ or other commands. We suggest using the command
\verb+\includegraphics+
from the graphicx package. Always specify the figure width as a multiple of
the line width as in the example below using .eps graphics
\begin{verbatim}
   \usepackage[dvips]{graphicx} ...
   \includegraphics[width=0.8\linewidth]{myfile.eps}
\end{verbatim}
or % Apr 2009 addition
\begin{verbatim}
   \usepackage[pdftex]{graphicx} ...
   \includegraphics[width=0.8\linewidth]{myfile.pdf}
\end{verbatim}
for .pdf graphics.
See section~4.4 in the graphics bundle documentation (\url{http://www.ctan.org/tex-archive/macros/latex/required/graphics/grfguide.ps})

A number of width problems arise when LaTeX cannot properly hyphenate a
line. Please give LaTeX hyphenation hints using the \verb+\-+ command.

\subsubsection*{Author Contributions}
If you'd like to, you may include  a section for author contributions as is done
in many journals. This is optional and at the discretion of the authors.

\subsubsection*{Acknowledgments}
Use unnumbered third level headings for the acknowledgments. All
acknowledgments, including those to funding agencies, go at the end of the paper.

\bibliography{iclr2020_conference}
\bibliographystyle{iclr2020_conference}

\appendix
\section{Appendix}
You may include other additional sections here.

\end{document}

### 2020-03-04 paper/introduction.tex
clean up

@@
-Natural sciences are tasked to find, from a set of hypothetical factors, the minimal subset that suffices to reliably predict novel observations. This endeavor is impeded by two major challenges.
+Natural sciences are tasked to find, from a set of hypothetical factors, the
+minimal subset that suffices to reliably predict novel observations. This
+endeavor is impeded by two major challenges.
 
 First, causal and non-causal factors may be numerous and collinear.
-%In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or a change in pressure, as these two factors may, at first, be difficult to manipulate independently.
-This issue becomes increasingly pronounced as the number of potential factors increases.
-In neuroscience, for example, identifying whether the frequency of a word presented on a subject's retina modulates brain activity can be surprisingly difficult.
-Indeed, the frequency of words in natural language covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs) and so forth \citep{kutas2011thirty,pegado2014timing}.
-Instead of selecting a set of words that control for all of these factors simultaneously, it is thus common to use forward modeling, i.e.
-to train a model to predict observations (e.g.
-brain activity) from a minimal combination of competing factors (e.g.
-word length, word frequency, e.g.
-\citep{huth2016natural}), and investigate, in the model, the estimated contribution of each factor \citep{friston1994statistical}.
+% In physics,
+% for example, one may be challenged to identify whether fusion is caused by a
+% change in temperature or a change in pressure, as these two factors may, at
+% first, be difficult to manipulate independently. This issue becomes increasingly
+% pronounced as the number of potential factors increases.
+In neuroscience, for
+example, it can be challenging to identify whether word frequency modulates
+brain activity during reading. Indeed, the
+frequency of words in natural language covaries with other factors such as their
+length (short words are more frequent than long words) and their categories
+(determinants are more frequent than adverbs)
+\citep{kutas2011thirty,pegado2014timing}. Instead of selecting a set of words
+that controls for all of these factors simultaneously, it is thus common to use
+a \emph{forward} "encoding model", i.e. to fit a linear regression to predict observations
+(e.g. brain activity) from a minimal combination of competing factors (e.g.
+word length, word frequency), and analytically investigate,
+the estimated contribution of each factor from the model's coefficients
+\citep{friston1994statistical,naselaris2011encoding,weichwald2015causal,
+king2018encoding,huth2016natural}.
 
-The second challenge to measuring causal influence is that observations can be large and complex.
-The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from multidimensional observations.
-For example, brain activity is often recorded with hundreds or thousands of sensors simultaneously.
-As multiple sensors may be affected by common noise sources, it is common to use backward regression, by, for example, fitting a support vector machine across multiple sensors to decode the category of a stimulus \citep{cichy2014resolving,  kriegeskorte2008representational, norman2006beyond}.
+The second challenge to measuring causal influence is that observations can be
+multidimensional. The relationship between causes and effects is thus often
+considered in a \emph{backward} manner, by training models to maximally predict causes
+from multidimensional observations. For example, brain activity is often
+recorded with hundreds or thousands of simultaneous measurements via functional
+Magnetic Resonance Imaging, magneto-encephalography (MEG) or multiple electro-
+physiological probes \citep{friston1994statistical,steinmetz2018challenges}.
+As simultaneous measurements may be affected by common noise sources, it is
+common to use backward
+modeling, by, for example, fitting a support vector machine across multiple
+sensors to decode the category of a stimulus \citep{norman2006beyond,cichy2014resolving,
+kriegeskorte2008representational}.
 
-Both forward and backward modeling have competing benefits and drawbacks.
-Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations.
-By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \citep{weichwald2015causal, hebart2018deconstructing, king2018encoding}.
-To combine the benefits of forward and backward modeling, several authors have proposed to use cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA) \citep{de2019multiway}.
-CCA and PLS aim to find, from two sets of data $X$ and $Y$, the components $H$ and $G$ were $XH$ and $YG$ are maximially correlated or maximally covarying respectively.
-Because CCA and PLS are based on a generalized eigen decomposition, their resulting coefficients are mixing the features of $X$ and $Y$ in a way that makes them notoriously difficult to interpret \citep{lebart1995statistique}.
+Both \emph{forward} and \emph{backward} modeling have competing benefits and drawbacks.
+Specifically, forward modeling disentangles the independent contribution of
+collinear factors, but does not combine multidimensional observations. By
+contrast, backward modeling combines multiple observations, but does not
+disentangle collinear factors \citep{weichwald2015causal,
+hebart2018deconstructing, king2018encoding}. To combine some of the benefits of forward
+and backward modeling, several authors have proposed to use cross-decomposition
+techniques such as Partial Least Squares (PLS) and Canonical Correlation
+Analysis (CCA) \citep{de2019multiway}. CCA and PLS aim to find, from two sets of
+data $X$ and $Y$, the components $H$ and $G$ were $XH$ and $YG$ are maximally
+correlated or maximally covarying respectively. Because CCA and PLS are based on
+a generalized eigen decomposition, their resulting coefficients are mixing the
+features of $X$ and $Y$ in a way that makes them notoriously difficult to
+interpret \citep{lebart1995statistique}.
 
-Here, we introduce the `back-to-back regression' (B2B), which not only combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}), but also provide robust, interpretable and unidimensional causal coefficients for each tested factor.
+Here, we introduce the `back-to-back regression' (B2B), which not only combines
+the benefits of forward and backward modeling (Section~\ref{sec:algorithm}), but
+also provides robust, interpretable, unidimensional and unbiased coefficients for
+each of tested factor.
 
-The present paper focuses on the restricted issue of disentangling the causal influence of linearly correlated predictors ($X$) onto noisy multivariate observations ($Y$).
-The present approach thus differs from other causal discovery algorithms based on temporal-delays and/or nonlinear interactions in systems where the directionality of causation (from X to Y or vice versa) is unknown (e.g.
-\citep{peters2017elements, granger1969investigating, janzing2013quantifying, scholkopf2016modeling}.
+The present paper focuses on the restricted issue of disentangling the
+influence of linearly correlated predictors ($X$) onto noisy multivariate
+observations ($Y$). The present approach thus differs from other causal
+discovery algorithms based on temporal-delays and/or nonlinear interactions in
+systems where the directionality of causation (from X to Y or vice versa) is
+unknown (e.g. \citep{peters2017elements, granger1969investigating,
+janzing2013quantifying, scholkopf2016modeling}.
 
-After detailing B2B method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques in identifying causal influence (Section~\ref{sec:experiment_synthetic}).
-Finally, we apply B2B to a large neuroimaging dataset and reveal that distinct but collinear word features lead to distinguishable brain representations (Section~\ref{sec:experiment_real}).
+After detailing B2B method and proving its convergence
+(Section~\ref{sec:theorem}), we show with synthetic data that it outperforms
+state-of-the-art forward, backward and cross-decomposition techniques in
+disentangling causal factors (Section~\ref{sec:experiment_synthetic}). Finally,
+we apply B2B to a large neuroimaging dataset and reveal that distinct but
+collinear word features lead to distinguishable brain representations
+(Section~\ref{sec:experiment_real}).

### 2020-03-04 paper/meg_exps.tex
clean up

@@
 The raw MEG data was bandpass-filtered between 0.1 and 40Hz using MNE-Python
-default parameters \citep{gramfort2013meg, gramfort2014mne}. Specifically, we used a zero-phase finite impulse
+default parameters \citep{gramfort2013meg, gramfort2014mne}. Specifically, we
+used a zero-phase finite impulse
 response filter (FIR) with a Hamming window and with transition bands of 0.1Hz
-and 10Hz for the low and high cut-off frequencies. The raw data was then segmented 100ms before word onset and 1s after
+and 10Hz for the low and high cut-off frequencies. The raw data was then
+segmented 100ms before word onset and 1s after
 word onset ($t=0$ms corresponds to word onset). Finally, each resulting
 segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
-thus led a sampling frequency of 240Hz. The average responses across words is displayed in Figure \ref{fig:meg_evoked}.
+thus led a sampling frequency of 240Hz. The average responses across words is
+displayed in Figure \ref{fig:meg_evoked}.
 For each subject and each time sample relative to word onset, we
-build an observation matrix $Y \in \mathbb{R}^{n \times d_y}$ of $n\approx$ 2700 words
+build an observation matrix $Y \in \mathbb{R}^{n \times d_y}$ of $n\approx$ 2,700 words
 by $d_y=301$ MEG channels (273 magnetometers and 28 compensation channels). Each
@@
   \begin{minipage}[c]{0.4\textwidth}
-    \caption{Ninety subjects each read approximately 2,700 words while their brain activity was recorded with MEG.
-Top.
-Average brain response to words (word onset at t=0 ms), as viewed from above the head (red= higher gradient of magnetic flux).
-Bottom.
-Each line represents a magnetometer, color-coded by its spatial position.
-Posterior responses, typical of primary visual cortex activity, peak around 100 ms after word onset and are followed by an anterior propagation of activity typical of semantic processing in the associative cortices.
+    \caption{A hundred subjects read approximately 2,700 words while their
+    brain activity was recorded with MEG. Top. Average brain response to words
+    (word onset at t=0 ms), as viewed from above the head (red= higher gradient
+    of magnetic flux). Bottom. Each line represents a magnetometer, color-coded
+    by its spatial position. Posterior responses, typical of primary visual
+    cortex activity, peak around 100 ms after word onset and are followed by
+    an anterior propagation of activity typical of semantic processing in the
+    associative cortices.
     }
@@
 
-We aim to identify the word features that cause a variation in brain responses. We consider four distinct but colinear features.
+We aim to identify the word features that cause a variation in brain responses.
+We consider four distinct but collinear features.
 %
-First, 'Word Length' refers to the total number of letters.
-Word Length is expected to specifically cause a variation in the early evoked MEG responses (i.e.
-from 100 ms after stimulus onset) elicited by the retinotopically-tuned visual cortices (e.g.
-\citep{pegado2014timing}.).
+First, 'Word Length' refers to the total number of letters. Word Length is
+expected to specifically cause a variation in the early evoked MEG responses
+(i.e. from 100 ms after stimulus onset) elicited by the retinotopically-tuned
+visual cortices (e.g. \citep{pegado2014timing}.).
 %
-Second, 'Word Frequency' indexes how frequently each word appears in Dutch and was derived with the the Zipf logarithmic scale of \citep{van2014subtlex} provided by the WordFreq package \citep{speerwordfreq}.
-Word Frequency is expected to specifically cause a variation in the late evoked MEG responses (i.e.
-from 400 ms), because it variably engages semantic processes in the temporal cortices \citep{kutas2011thirty}.
+Second, 'Word Frequency' indexes how frequently each word appears in Dutch and
+was derived with the the Zipf logarithmic scale of \citep{van2014subtlex}
+provided by the WordFreq package \citep{speerwordfreq}. Word Frequency is
+expected to specifically cause a variation in the late evoked MEG responses
+(i.e. from 400 ms), because it variably engages semantic processes in the
+temporal cortices \citep{kutas2011thirty}.
 %
-Third, 'Word Function' indicates whether each word is a content word (i.e.
-a noun, a verb, an adjective or an adverb) or a function word (i.e.
-a preposition, a conjunction, a determinant, a pronoun or a numeral), and was derived from Spacy's part of speech tagger \citep{spacy2}.
-To our knowledge, this feature has not been thouroughly investigated with MEG.
-Its causal contribution to reading processes in the brain thus remains unclear.
+Third, 'Word Function' indicates whether each word is a content word (i.e. a
+noun, a verb, an adjective or an adverb) or a function word (i.e. a preposition,
+a conjunction, a determinant, a pronoun or a numeral), and was derived from
+Spacy's part of speech tagger \citep{spacy2}. To our knowledge, this feature has
+not been thoroughly investigated with MEG. Its causal contribution to reading
+processes in the brain thus remains unclear.
 %
-Finally, to verify that B2B and other methods would not inadequately identify non-causal features, we added a dummy feature, constructed from a noisy combination of Word Length and Word Frequency:
-$dummy = z(length) + z(frequency) + \mathcal{N}$, where $z$ normalizes features and $\mathcal{N}$ is a random vector sampling Gaussian distribution (all terms thus have a zero-mean and a unit-variance).
-This procedure yields an $X \in \mathbb{R}^{n \times d_x}$ matrix of $n\approx$ 2700 words by
-$d_x=4$ features for each subject. Each of the columns of $X$ is normalized to
-have a mean and a standard deviation of 0 and 1 respectively.
+Finally, to verify that B2B and other methods would not inadequately identify
+non-causal features, we added a dummy feature, constructed from a noisy
+combination of Word Length and Word Frequency: $dummy = z(length) + z(frequency)
++ \mathcal{N}$, where $z$ normalizes features and $\mathcal{N}$ is a random
+vector sampling Gaussian distribution (all terms thus have a zero-mean and a
+unit-variance). This procedure yields an $X \in \mathbb{R}^{n \times d_x}$
+matrix of $n\approx$ 2,700 words by $d_x=4$ features for each subject. Each of
+the columns of $X$ is normalized to have a mean and a standard deviation of 0
+and 1 respectively.
 
@@
 
-We compare B2B to four standard methods: Forward regression, Backward regression, CCA and PLS, as implemented in scikit-learn \citep{sklearn} and \citep{bilenko2016pyrcca}, and optimized with nested cross-validation over twenty $l2$ regularization parameters logarithmically spaced between $10^{-4}$ and $10^4$ (for regression and CCA methods) or 1 to 4 canonical components (for PLS).
+We compare B2B to four standard methods: Forward regression, Backward
+regression, CCA and PLS, as implemented in scikit-learn \citep{sklearn} and
+\citep{bilenko2016pyrcca}, and optimized with nested cross-validation over
+twenty $l2$ regularization parameters logarithmically spaced between $10^{-4}$
+and $10^4$ (for regression and CCA methods) or 1 to 4 canonical components (for
+PLS).
 
-We used the feature importance described in Algorithm \ref{algorithm:b2b_fi} to assess the extent to which each feature $X_i$ specifically improves the prediction of held-out $Y$ data, using a 5-fold cross-validation (with shuffled trials to homogeneize the distributions between the training and testing splits).
+We used the feature importance described in Algorithm \ref{algorithm:b2b_fi} to
+assess the extent to which each feature $X_i$ specifically improves the
+prediction of held-out $Y$ data, using a five-fold cross-validation (with shuffled
+trials to homogeneize the distributions between the training and testing
+splits).
 
 Each model was implemented for each subject and each time sample independently.
-Pairwise comparison between models were performed using a two-sided Wilcoxon test across subjects (n=90) using the average $\Delta R$ across time.
-Corresponding effect sizes are shown in Figure~\ref{fig:meg_results}, and p-values are reported below.
+Pairwise comparison between models were performed using a two-sided Wilcoxon
+test across subjects (n=90) using the average $\Delta R$ across time.
+Corresponding effect sizes are shown in Figure~\ref{fig:meg_results}, and
+p-values are reported below.
 
 \subsubsection{Results}
-We compared the ability of Forward regression, Backward regression, CCA, PLS and B2B to estimate the causal contribution of four distinct but collinear features on brain evoked responses to words.
-
-As expected, the Backward model reveals a similar decoding time course for Word Length and Word Frequency, even though these features are known to specifically influence early and late MEG responses respectively \citep{kutas2011thirty}.
-In addition, the same decoding time course was observed for the dummy variable.
-These results illustrate that backward modelling cannot be used to estimate the causal contribution of collinear features.
-
-We thus focus on the four remaining methods (i.e.
-Forward Regression, PLS, CCA, and B2B) and estimate their $\Delta R$ (i.e.
-the improvement of Y prediction induced by the introduction of a given feature into the model,as described in Algorithm \ref{algorithm:b2b_fi}).
-Contrary to the Backward Model, none of the models predicted the Dummy Variable to improve the $Y$ prediction: all $\Delta R < 0$ (all $p > .089$).
-
-Figure \ref{fig:meg_results} shows, for each model, the effects obtained across time (left) and subjects (right).
-
-Word Length and Word Frequency improved the prediction performance of all methods: $\Delta R>0$ for all models (all $p<0.0001$).
-As expected, the time course associated with Word Length and Word Frequency rose from $\approx$ 100 ms and from $\approx$ 400 ms respectively.
-Furthermore, Word Function improved the prediction performance of all models (all $p < 0.0002$) except for PLS~($p=0.7989$).
-Overall, these results confirm that Word Length, Word Frequency and Word Function causally influence specific periods of brain responses to words.
-
-To assess which model would be most sensitive to these causal discoveries, we compared B2B to other models across subjects (Figure \ref{fig:meg_results} right).
-For Word Length B2B outperforms all models (all $p < 0.00001$) but CCA ($p=0.0678$).
-For Word Frequency, B2B outperforms all models (all $p < 0.0006$).
-For "Word Function", B2B outperforms all models (all $p < 0.0015$).
-Overall, these results show that B2B reliably outperforms standard methods, especially when the effects are difficult to detect.
+We compared the ability of Forward regression, Backward regression, CCA, PLS and
+B2B to estimate the causal contribution of four distinct but collinear features
+on brain evoked responses to words.
 
 \begin{wrapfigure}{r}{.6\textwidth}
-  \vspace{-12ex}
+  %\vspace{-12ex}
   \begin{center}
@@
   \end{center}
-  \vspace{-5ex}
-  \caption{Multiple models (color-coded) are compared on their ability to reliably predict single-trial MEG signals evoked by words.
-Left.
-Average improvement of correlation coefficient $\Delta R$ for each of the four features (rows).
-Error bars indicate standard error of the mean (SEM) across subjects.
-Right.
-Average $\Delta R$ across time for each subject (dots).
-Top horizontal lines indicate when B2B significantly outperforms other methods (red) and vice versa.
-\label{fig:meg_results}
+  %\vspace{-5ex}
+  \caption{Multiple models (color-coded) are compared on their ability to
+  reliably predict single-trial MEG signals evoked by words. Left. Average
+  improvement of correlation coefficient $\Delta R$ for each of the four
+  features (rows). Error bars indicate standard error of the mean (SEM) across
+  subjects. Right. Average $\Delta R$ across time for each subject (dots). Top
+  horizontal lines indicate when B2B significantly outperforms other methods
+  (red) and vice versa. \label{fig:meg_results}
 }
 \end{wrapfigure}
+
+As expected, the Backward model reveals a similar decoding time course for Word
+Length and Word Frequency, even though these features are known to specifically
+influence early and late MEG responses respectively \citep{kutas2011thirty}. In
+addition, the same decoding time course was observed for the dummy variable.
+These results illustrate that backward modeling cannot be used to estimate the
+causal contribution of collinear features.
+
+We thus focus on the four remaining methods (i.e. Forward Regression, PLS, CCA,
+and B2B) and estimate their $\Delta R$ (i.e. the improvement of Y prediction
+induced by the introduction of a given feature into the model,as described in
+Algorithm \ref{algorithm:b2b_fi}). Contrary to the Backward Model, none of the
+models predicted the Dummy Variable to improve the $Y$ prediction: all $\Delta R
+< 0$ (all $p > .089$).
+
+Figure \ref{fig:meg_results} shows, for each model, the effects obtained across
+time (left) and subjects (right).
+
+Word Length and Word Frequency improved the prediction performance of all
+methods: $\Delta R>0$ for all models (all $p<0.0001$). As expected, the time
+course associated with Word Length and Word Frequency rose from $\approx$ 100 ms
+and from $\approx$ 400 ms respectively. Furthermore, Word Function improved the
+prediction performance of all models (all $p < 0.0002$) except for
+PLS~($p=0.7989$). Overall, these results confirm that Word Length, Word
+Frequency and Word Function causally influence specific periods of brain
+responses to words.
+
+To assess which model would be most sensitive to these causal discoveries, we
+compared B2B to other models across subjects (Figure \ref{fig:meg_results}
+right). For Word Length B2B outperforms all models (all $p < 0.00001$) but CCA
+($p=0.0678$). For Word Frequency, B2B outperforms all models (all $p < 0.0006$).
+For "Word Function", B2B outperforms all models (all $p < 0.0015$). Overall,
+these results show that B2B reliably outperforms standard methods, especially
+when the effects are difficult to detect.

### 2020-03-04 paper/paper.tex
clean up

@@
-\documentclass{article}
+\documentclass[preprint,12pt,3p]{elsarticle}
 
-\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref, amsthm}
+\usepackage{lineno,hyperref}
+\modulolinenumbers[5]
+
+\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, amsthm}
 \usepackage[ruled, linesnumbered]{algorithm2e}
 \usepackage{times,wrapfig}
-\usepackage{authblk}
-\usepackage{apacite}
-\usepackage{natbib}
-\usepackage{biblatex}
+\usepackage{cleveref}
 
@@
 
-\title{Discriminating causes \\from multivariate observations:\\the back-to-back regression}
+\journal{Neuroimage}
 
-\author[1, 2]{Jean-Rémi King}
+% \bibliographystyle{elsarticle-num}
+\bibliographystyle{plain}
+
+\begin{document}
+\begin{frontmatter}
+
+\title{Discriminating the Linear Influences of Collinear Factors \\from Multivariate Observations: the Back-to-Back Regression}
+
+\author[1,2]{Jean-Rémi King\corref{email}}
+\cortext[email]{corresponding author: jeanremi@fb.com}
+
 \author[2]{Fran\c{c}ois Charton}
@@
 \author[2]{Maxime Oquab}
-\affil[1]{Laboratoire des systèmes perceptifs, PSL University, CNRS}
-\affil[2]{Facebook AI}
-\date{}
-
-% \author{%
-%   Jean-Rémi King\\
-%   CNRS\\
-%   \texttt{jeanremi.king@gmail.com} \\
-%   \And
-%   Fran\c{c}ois Charton\\
-%   Facebook AI\\
-%   \texttt{fcharton@fb.com}\\
-%   \And
-%   David Lopez-Paz\\
-%   Facebook AI\\
-%   \texttt{dlp@fb.com}
-%   \And
-%   Maxime Oquab\\
-%   Facebook AI\\
-%   \texttt{qas@fb.com}
-% }
-
-\begin{document}
-
-\maketitle
+\address[1]{Laboratoire des systèmes perceptifs, PSL University, CNRS}
+\address[2]{Facebook AI}
 
@@
 i) potential factors are difficult to manipulate individually and ii)
-observations are complex and multi-dimensional. To address this issue, we
-introduce ``Back-to-Back'' regression (B2B), a method designed to efficiently
-measure, from a set of colinear factors, those that most plausibly account for
-multidimensional observations. After proving the consistency of B2B and its
-links to other linear approaches, we show that our method outperforms
+observations are multi-dimensional. To address this issue, we
+introduce ``Back-to-Back'' regression (B2B), a linear method designed to efficiently
+measure, from a set of collinear factors, those that most plausibly account for
+multidimensional observations. First, we prove the consistency of B2B, its links to
+other linear approaches, and show how it provides a robust, unbiased and
+interpretable scalar estimate for each factor.
+Second, we use a variety of simulated data to show that B2B outperforms
 least-squares regression and cross-decomposition techniques (e.g. canonical
-correlation analysis and partial least squares) on causal identification.
-Finally, we apply B2B to neuroimaging recordings of 102 subjects reading word
-sequences. The results show that the early and late brain representations,
-caused by low- and high-level word features respectively, are more reliably
-detected with B2B than with other standard techniques.
+correlation analysis and partial least squares) on causal identification when
+both the factors and the observations are collinear.
+Finally, we apply B2B to magneto-encephalography of 102 subjects
+recorded during a reading task to test whether our method appropriately disentangles
+the respective contribution of word length and word frequency - two collinear
+factors known to cause early and late brain responses respectively. The results
+show that these two factors are better disentangled with B2B than with
+other standard techniques. We finally discuss the limits of B2B.
+\end{abstract}
 
-\end{abstract}
+\begin{keyword}
+  Cross-Decomposition \sep
+  Feature Discovery \sep
+  Magnetoencephalography \sep
+  Decoding \sep
+  Encoding \sep
+  Reading \sep
+  N400
+\end{keyword}
+
+\end{frontmatter}
+
+\linenumbers
 
@@
 
+\begin{figure}[t!]
   \include{fig_problem.tex}
+  \label{fig:b2b}
+\end{figure}
 
 We consider the measurement of multivariate signal $Y \in \mathbb{R}^{n \times
-d_y}$, generated from a set of putative causes $X \in \mathbb{R}^{n \times
-d_x}$, via some unknown linear apparatus $F \in \mathbb{R}^{d_x \times d_y}$.
+d_y}$ (the dependent variables), generated from a set of putative causes $X \in \mathbb{R}^{n \times
+d_x}$ (the independent variables), via some unknown linear apparatus $F \in \mathbb{R}^{d_x \times d_y}$.
 %
@@
 %
-where $i$ is a given sample, and $n_i$ is a sample-specific noise drawen from a
+where $i$ is a given sample, and $n_i$ is a sample-specific noise drawn from a
 centered distribution.
@@
 %
-Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back
-Regression (B2B) is to estimate the matrix of causal influences $E$.
+Given the model in Equation~\cref{eq:model}, the goal of Back-to-Back
+Regression (B2B) is to estimate the matrix of $E$, i.e. to identify the factors
+that most reliably account for the multivariate observations.
 
@@
 To accelerate this ensembling procedure, we implemented an efficient
-leave-one-out cross-validation scheme as detailed in \citep{rifkin2007notes} %
-http://cbcl.mit.edu/publications/ps/MIT-CSAIL-TR-2007-025.pdf, as follows:
+leave-one-out cross-validation scheme as detailed in \citep{rifkin2007notes}
+% http://cbcl.mit.edu/publications/ps/MIT-CSAIL-TR-2007-025.pdf,
+as follows:
 %
@@
 
-     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full
+     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centered and full
      rank noise.
@@
 %
-We use scikit-learn's PLS and RidgeCV \citep{sklearn} as well as pyrcca's
+We use scikit-learn's PLS and RidgeCV \citep{sklearn} as well as Pyrcca's
 regularized canonical component analysis (RegCCA, \citep{bilenko2016pyrcca})
@@
 
-% I find a bit surprising to mention cauaslity literare in ML at the end. I would have expected this in the introduction to convince reviewer that know your stuff.
+% I find a bit surprising to mention causality literature in ML at the end. I
+% would have expected this in the introduction to convince reviewer that know
+% your stuff.
 
 \section{Conclusion}
-\input{conclusion}
+In this work, we proposed Back-to-Back (B2B) regression, a linear method to
+disentangle confounded factors from multidimensional observations.
+%
+B2B repeatedly performs two successive multidimensional regressions on
+independent subsets of the data: the first regression is applied on the output
+domain (as in backward decoding), whereas the second regression is applied on
+the input domain (as in forward encoding).
+%
+We provided a theoretical guarantee about the consistency of B2B, and compared
+it to several baselines in controlled synthetic experiments.
+%
+We also applied B2B to a recent brain imaging dataset, analyzing the timing of
+brain responses and their connection to word features.
+%
+We obtained results consistent with prior work in neuroscience literature,
+confirming the reliability of B2B for real data analysis.
+%
 
@@
 
-\bibliographystyle{iclr2020_conference}
 \bibliography{paper}
 
-% \input{supplementary.tex}
-
-%\section{Appendices}
-%
-%This should hold part of the tests, explanations on simulation, detailed results and stuff on meg data
-
+\section{Appendices}
 \input{appendix}

### 2020-03-04 paper/synthetic.tex
clean up

@@
 
-We employ $5$-fold cross-validation to select the optimal number of components
+We employ five-fold cross-validation to select the optimal number of components
 for CCA and PLS. Regressions were $\ell2$-regularized with a $\lambda$ regularization
@@
 We show in Appendix~\ref{appendix:feature_importance} pseudo-code to assess feature importance for our algorithm as well as baselines.
-For the Backward Model, feature importance cannot be assessed as the $X$ colinearity is never taken into account.
+For the Backward Model, feature importance cannot be assessed as the $X$ collinearity is never taken into account.
 

### 2020-03-04 code/_tmp.py

@@ -1,0 +1,381 @@
+import numpy as np
+from scipy import linalg
+from scipy.stats import pearsonr
+from sklearn.cross_decomposition import CCA as SkCCA
+from sklearn.cross_decomposition import PLSRegression as SkPLS
+from sklearn.model_selection import GridSearchCV
+from sklearn.model_selection import ShuffleSplit
+from sklearn.metrics import r2_score
+from sklearn.base import BaseEstimator, RegressorMixin
+
+
+def r_score(X, Y, multioutput='uniform_average'):
+    """column-wise correlation coefficients"""
+
+    assert multioutput in ('raw', 'uniform_average', 'variance_weighted')
+    if X.ndim == 1:
+        X = X[:, None]
+    if Y.ndim == 1:
+        Y = Y[:, None]
+    assert len(X) >= 2
+    np.testing.assert_equal(X.shape, Y.shape)
+
+    R = np.zeros(X.shape[1])
+    for idx, (x, y) in enumerate(zip(X.T, Y.T)):
+        R[idx] = pearsonr(x, y)[0]
+
+    if multioutput == 'uniform_average':
+        R = R.mean()
+    elif multioutput == 'variance_weighted':
+        std = np.r_[X, Y].std(0)
+        R = np.average(R, weights=std)
+    return R
+
+
+def rn_score(X, Y, scoring='r', multioutput='uniform_average'):
+    assert scoring in ('r', 'r2')
+    assert multioutput in ('raw', 'uniform_average', 'variance_weighted')
+
+    if scoring == 'r':
+        return r_score(X, Y, multioutput=multioutput)
+
+    elif scoring == 'r2':
+        return r2_score(X, Y, multioutput=multioutput)
+
+
+class B2B(BaseEstimator, RegressorMixin):
+    def __init__(self, alphas=np.logspace(-4, 4, 20),
+                 independent_alphas=True, ensemble=None):
+        self.alphas = alphas
+        self.independent_alphas = independent_alphas
+        self.ensemble = ensemble
+
+    def fit(self, X, Y):
+
+        self.G_ = list()
+        self.H_ = list()
+
+        # Prepare ensembling
+        if self.ensemble is None:
+            ensemble = [(range(len(X)), range(len(X))), ]
+        else:
+            if isinstance(self.ensemble, int):
+                ensemble = ShuffleSplit(self.ensemble)
+            else:
+                ensemble = self.ensemble
[…]

### 2020-03-04 code/synth_exp.py

@@ -1,0 +1,96 @@
+from models import JRR, OLS, Ridge, Oracle, PLS, Lasso, RRR, CCA
+from models_k import JRR_k, Ridge_k, CCA_k
+from sklearn.metrics import roc_auc_score
+from data import Synthetic
+import numpy as np
+import argparse
+import random
+
+
+models = {
+    # "JRR": JRR,
+    # "OLS": OLS,
+    # "Ridge": Ridge,
+    # "CCA": CCA,
+    # "Oracle": Oracle,
+    # "PLS": PLS,
+    # "RRR": RRR,
+    # "Lasso": Lasso,
+    "JRR_k": JRR_k,
+    "Ridge_k": Ridge_k,
+    "CCA_k": CCA_k
+}
+
+
+def nmse(y_pred, y_true):
+    num = np.power(np.linalg.norm(y_pred - y_true), 2)
+    den = np.power(np.linalg.norm(y_true), 2)
+    return num / den
+
+
+def run_experiment(args):
+    for seed in range(args.n_seeds):
+        random.seed(seed)
+        np.random.seed(seed)
+
+        synthetic = Synthetic(args.dim_x,
+                              args.dim_y,
+                              args.nc,
+                              args.snr,
+                              args.nonlinear)
+
+        # true binary E
+        mask_true = synthetic.solution()
+
+        # training data
+        x_tr, y_tr = synthetic.sample(args.n_samples)
+
+        # testing data (in-domain)
+        x_te_in, y_te_in = synthetic.sample(args.n_samples * 10)
+
+        # testing data (out-domain)
+        x_te_out, y_te_out = synthetic.sample(args.n_samples * 10,
+                                              in_domain=False)
+
+        results = []
+
+        for m, Model in models.items():
+            if m == "Oracle":
+                model = Model(mask_true)
+            else:
+                model = Model()
+
+            # fit model on training data
+            model.fit(x_tr, y_tr)
+
+            error_in_all = nmse(model.predict(x_te_in), y_te_in)
+            error_out_all = nmse(model.predict(x_te_out), y_te_out)
+
+            # this is the binary mask E estimated by the model
+            mask = model.solution()
+
+            result = dict(vars(args))
+            result["model"] = m
+            result["seed"] = seed
+            result["result_error_in_all"] = error_in_all
+            result["result_error_out_all"] = error_out_all
+            result["result_auc"] = roc_auc_score(mask_true, mask)
+            results.append(result)
+
[…]

### 2020-03-04 code/synthetic_iclr/_models_k.py

@@ -1,0 +1,403 @@
+import numpy as np
+from scipy.linalg import svd
+from scipy.stats import pearsonr
+from sklearn.preprocessing import scale
+from sklearn.cross_decomposition import PLSRegression, CCA
+from sklearn.model_selection import GridSearchCV
+from sklearn.model_selection import ShuffleSplit
+from sklearn.base import clone
+
+
+N_JOBS = 1
+
+
+def ridge_cv(X, Y, alphas, independent_alphas=False, Uv=None):
+    """ Similar to sklearn RidgeCV but
+   (1) can optimize a different alpha for each column of Y
+   (2) return leave-one-out Y_hat
+   """
+    if isinstance(alphas, (float, int)):
+        alphas = np.array([
+            alphas,
+        ], np.float64)
+    if Y.ndim == 1:
+        Y = Y[:, None]
+    n, n_x = X.shape
+    n, n_y = Y.shape
+    # Decompose X
+    if Uv is None:
+        U, s, _ = svd(X, full_matrices=0)
+        v = s**2
+    else:
+        U, v = Uv
+    UY = U.T @ Y
+
+    # For each alpha, solve leave-one-out error coefs
+    cv_duals = np.zeros((len(alphas), n, n_y))
+    cv_errors = np.zeros((len(alphas), n, n_y))
+    for alpha_idx, alpha in enumerate(alphas):
+        # Solve
+        w = ((v + alpha)**-1) - alpha**-1
+        c = U @ np.diag(w) @ UY + alpha**-1 * Y
+        cv_duals[alpha_idx] = c
+
+        # compute diagonal of the matrix: dot(Q, dot(diag(v_prime), Q^T))
+        G_diag = (w * U**2).sum(axis=-1) + alpha**-1
+        error = c / G_diag[:, np.newaxis]
+        cv_errors[alpha_idx] = error
+
+    # identify best alpha for each column of Y independently
+    if independent_alphas:
+        best_alphas = (cv_errors**2).mean(axis=1).argmin(axis=0)
+        duals = np.transpose(
+            [cv_duals[b, :, i] for i, b in enumerate(best_alphas)])
+        cv_errors = np.transpose(
+            [cv_errors[b, :, i] for i, b in enumerate(best_alphas)])
+    else:
+        _cv_errors = cv_errors.reshape(len(alphas), -1)
+        best_alphas = (_cv_errors**2).mean(axis=1).argmin(axis=0)
[…]

### 2020-03-04 code/synthetic_iclr/base.py

@@ -1,0 +1,64 @@
+import mne
+import numpy as np
+import os.path as op
+from sklearn.preprocessing import scale
+from common import (add_part_of_speech, add_word_frequency, add_word_length,
+                    read_log, get_log_times, data_path)
+
+
+def fetch_data(log_file):
+    # file name
+    raw_fname = op.join(data_path, log_file['meg_file'])
+    log_fname = op.join(data_path, log_file['log_file'])
+
+    # preprocess meg data
+    raw = mne.io.read_raw_ctf(raw_fname, preload=True)
+    raw.filter(.1, 40.)
+    events = mne.find_events(raw)
+
+    # preprocess annotations
+    log = read_log(log_fname)
+
+    # link meg and annotations
+    log = get_log_times(log, events,  raw.info['sfreq'])
+
+    # Segment into word-locked epochs
+    log_events = np.c_[log.meg_sample, np.ones((len(log), 2), int)]
+    _, idx = np.unique(log_events[:, 0], return_index=True)
+    picks = mne.pick_types(raw.info, meg=True, eeg=False,
+                           stim=False, eog=False, ecg=False)
+    epochs = mne.Epochs(
+        raw,
+        events=log_events[idx], metadata=log.iloc[idx],
+        tmin=-.100, tmax=1.5, decim=20,
+        picks=picks, preload=True,
+    )
+    epochs = epochs['condition=="word"']
+
+    epochs = epochs.apply_baseline(None, 0).crop(-.2, 1.)
+
+    # Preprocessing features
+    df = add_word_frequency(epochs.metadata[['word', ]])
+    df = add_word_length(df)
+    df = add_part_of_speech(df)
+    df['word_func'] = False
+    df.loc[df.query('(ADP+CONJ+DET+PRON)==True').index,
+           'word_func'] = True
+    word = df.word.apply(lambda s: s.lower())
+    for letter in 'abcdefghijklmnopqrstuvwxyz':
+        df[letter] = word.str.count(letter) > 1
+
+    df['word_dummy'] = (scale(df['word_length']) +
+                        scale(df['word_freq']) +
+                        np.random.randn(len(df)))
+
+    meg = epochs.get_data()
+    n_samples, n_chans, n_times = meg.shape
+    for t in range(n_times):
+        meg[:, :, t] = scale(meg[:, :, t])
+
+    names = ['word_length', 'word_freq', 'word_func', 'word_dummy']
+
+    features = scale(df[names].values)
+
+    return meg, epochs.times, features, names, df.word

### 2020-03-04 code/synthetic_iclr/main.py

@@ -1,0 +1,123 @@
+import numpy as np
+from models import B2B, Forward, Backward, GridCCA, GridPLS, GridRegCCA
+from models import score_knockout
+from sklearn.metrics import roc_auc_score
+from sklearn.preprocessing import scale
+# from tqdm import trange
+import time
+# import pandas as pd
+# from tqdm import trange
+# from itertools import product
+# import pickle
+# import submitit
+# import seaborn as sns
+# import matplotlib.pyplot as plt
+
+
+class Synthetic(object):
+    def __init__(self,
+                 dim_x=50,         # number of features
+                 dim_y=30,         # number of sensors
+                 nc=5,             # number of selected features
+                 snr=1.0,          # signal-to-noise ratio
+                 nonlinear=False):  # number of selected features
+
+        # linear transformation
+        self.F = np.random.randn(dim_x, dim_y) / np.sqrt(dim_x)
+
+        # masking transformation
+        self.E = np.array([0] * (dim_x - nc) + [1] * (nc))
+
+        # features covariance
+        self.cov_X = np.random.randn(dim_x, dim_x) / np.sqrt(dim_x)
+        self.cov_X = self.cov_X @ self.cov_X.T
+
+        # noise covariance
+        self.cov_N = np.random.randn(dim_x, dim_x) / np.sqrt(dim_x)
+        self.cov_N = self.cov_N @ self.cov_N.T
+
+        self.dim_x = dim_x
+        self.dim_y = dim_y
+        self.nonlinear = nonlinear
+        self.snr = snr
+
+    def sample(self, n_samples=1000):
+        X = np.random.multivariate_normal(np.zeros(self.dim_x),
+                                          self.cov_X, n_samples)
+        N = np.random.multivariate_normal(np.zeros(self.dim_x),
+                                          self.cov_N, n_samples)
+
+        # observed sensor data
+        Y = (self.snr * X @ np.diag(self.E) + N) @ self.F
+
+        if self.nonlinear:
+            Y = 1. / (1. + np.exp(-Y))
+
+        # return inputs, outputs, and solution
+        return scale(X), scale(Y)
+
+
+def run(args=dict()):
+    import warnings
+    warnings.filterwarnings("ignore")
+
+    models = {
+        "Forward": Forward(),
+        "Backward": Backward(),
+        "PLS": GridPLS(),
+        "CCA": GridCCA(),
+        "RegCCA": GridRegCCA(),
+        "B2B": B2B(),
+    }
+
[…]

### 2020-03-04 code/synthetic_iclr/models.py

@@ -1,0 +1,757 @@
+import numpy as np
+from scipy import linalg
+from copy import deepcopy
+from scipy.stats import pearsonr
+from sklearn.cross_decomposition import CCA as SkCCA
+from sklearn.cross_decomposition import PLSRegression as SkPLS
+from sklearn.model_selection import GridSearchCV
+from sklearn.model_selection import ShuffleSplit
+from sklearn.metrics import r2_score
+from sklearn.base import BaseEstimator, RegressorMixin
+import warnings
+warnings.simplefilter("ignore")
+
+alphas = np.logspace(-4, 4, 20)
+components = np.linspace(0., 1., 20)
+
+
+def r_score(X, Y, multioutput='uniform_average'):
+    """column-wise correlation coefficients"""
+
+    assert multioutput in ('raw', 'raw_values', 'uniform_average',
+                           'variance_weighted')
+    if X.ndim == 1:
+        X = X[:, None]
+    if Y.ndim == 1:
+        Y = Y[:, None]
+    assert len(X) >= 2
+    np.testing.assert_equal(X.shape, Y.shape)
+
+    R = np.zeros(X.shape[1])
+    for idx, (x, y) in enumerate(zip(X.T, Y.T)):
+        R[idx] = pearsonr(x, y)[0]
+
+    if multioutput == 'uniform_average':
+        R = R.mean()
+    elif multioutput == 'variance_weighted':
+        std = np.r_[X, Y].std(0)
+        R = np.average(R, weights=std)
+    return R
+
+
+def rn_score(X, Y, scoring='r', multioutput='uniform_average'):
+    assert scoring in ('r', 'r2')
+    assert multioutput in ('raw', 'raw_values',
+                           'uniform_average', 'variance_weighted')
+
+    if scoring == 'r':
+        return r_score(X, Y, multioutput=multioutput)
+
+    elif scoring == 'r2':
+        return r2_score(X, Y, multioutput=multioutput)
+
+
+class B2B(BaseEstimator, RegressorMixin):
+    def __init__(self, alphas=alphas,
+                 independent_alphas=True, ensemble=None,
+                 scoring='r'):
+        self.alphas = alphas
+        self.independent_alphas = independent_alphas
+        self.ensemble = ensemble
+        self.scoring = scoring
+        self.__name__ = 'B2B'
+
+    def fit(self, X, Y):
+
+        self.G_ = list()
+        self.H_ = list()
+
+        # Prepare ensembling
[…]

### 2020-03-04 code/synthetic_iclr/time_series.py

@@ -1,0 +1,69 @@
+import numpy as np
+from sklearn.model_selection import KFold
+from sklearn.preprocessing import scale
+import matplotlib.pyplot as plt
+import mne
+from mne.datasets import sample
+from models import RegCCA, B2B, score_knockout
+
+
+def fetch_data():
+    # Preprocess data
+    data_path = sample.data_path() + '/MEG/sample/'
+    raw_fname = data_path + 'sample_audvis_filt-0-40_raw.fif'
+    events_fname = data_path + 'sample_audvis_filt-0-40_raw-eve.fif'
+    raw = mne.io.read_raw_fif(raw_fname, preload=True)
+    picks = mne.pick_types(raw.info, meg=True, exclude='bads')
+    raw.filter(1., 30., fir_design='firwin')
+    events_samples, _, events_values = mne.read_events(events_fname).T
+
+    Y = raw.get_data()[picks]
+    X = np.zeros((4, raw.n_times))
+    for idx, value in enumerate(set(events_values)):
+        sel = np.where(events_values == value)[0]
+        X[idx, events_samples[sel]] = 1.
+
+    # Robust scaling
+    low, high = np.percentile(Y, [.01, 99.99], axis=1)
+    for ch, (l, h) in enumerate(zip(low, high)):
+        Y[ch] = np.clip(Y[ch], l, h)
+
+    Y = scale(Y.T)
+    X = scale(X.T)
+
+    return X, Y
+
+
+def time_lag(X, tois):
+
+    n_times, n_events = X.shape
+    Xs = np.zeros((n_times, n_events, len(tois)))
+
+    for i, t in enumerate(tois):
+        Xs[:, :, i] = np.roll(X, t, axis=0)
+
+    return Xs.reshape(n_times, -1)
+
+
+if __name__ == '__main__':
+
+    X, Y = fetch_data()
+
+    models = dict(B2B=B2B(ensemble=20), CCA=RegCCA())
+    tois = range(-10, 10)
+
+    axes = iter(plt.subplots(len(models), 1)[1])
+
+    for m, model in models.items():
+        Xs = time_lag(X, tois)
+
+        cv = KFold(5)
+        scores = np.zeros((cv.n_splits, Xs.shape[1]))
+        for split, (train, test) in enumerate(cv.split(X)):
+            print('.', end='')
+            model.fit(Xs[train], Y[train])
+            scores[split] = score_knockout(model, Xs[test], Y[test])
+
[…]

### 2020-03-04 paper/cover_letter_neuroimage.tex
up

\documentclass[11pt,a4paper,roman]{moderncv}        % possible options include font size ('10pt', '11pt' and '12pt'), paper size ('a4paper', 'letterpaper', 'a5paper', 'legalpaper', 'executivepaper' and 'landscape') and font family ('sans' and 'roman')
\usepackage[english,en-lcroman]{babel}

% moderncv themes
\moderncvstyle{classic}                            % style options are 'casual' (default), 'classic', 'oldstyle' and 'banking'
\moderncvcolor{green}                              % color options 'blue' (default), 'orange', 'green', 'red', 'purple', 'grey' and 'black'
%\renewcommand{\familydefault}{\sfdefault}         % to set the default font; use '\sfdefault' for the default sans serif font, '\rmdefault' for the default roman one, or any tex font name
%\nopagenumbers{}                                  % uncomment to suppress automatic page numbering for CVs longer than one page

% character encoding
\usepackage[utf8]{inputenc}                       % if you are not using xelatex ou lualatex, replace by the encoding you are using

% adjust the page margins
\usepackage[scale=0.75]{geometry}
%\setlength{\hintscolumnwidth}{3cm}                % if you want to change the width of the column with the dates
%\setlength{\makecvtitlenamewidth}{10cm}           % for the 'classic' style, if you want to force the width allocated to your name and avoid line breaks. be careful though, the length is normally calculated to avoid any overlap with your personal info; use this at your own typographical risks...

% personal data
\name{Jean-Rémi}{King}
\extrainfo{École Normale Supérieure, PSL University}
\extrainfo{Facebook AI}
%\title{Resumé title}                               % optional, remove / comment the line if not wanted
% \address{29 rue d'Ulm, 75005}{Paris}                % optional, remove / comment the line if not wanted; the "postcode city" and and "country" arguments can be omitted or provided empty
% \phone[mobile]{+33 6 64 25 56 54}                   % optional, remove / comment the line if not wanted
\email{jeanremi@fb.com}                               % optional, remove / comment the line if not wanted

\begin{document}
% \recipient{Destinatario}{Departamento, Empresa}
\date{\today}
\opening{Dear Editor,}
\closing{Thank you for your consideration,}
\makelettertitle

I am pleased to submit our latest work entitled “Discriminating the Influence of
Correlated Factors from Multivariate Observations: the Back-to-Back Regression”
to \emph{Neuroimage}.

The past years have been marked by an ongoing debate about the benefits and
drawbacks of decoding and encoding methods in neuroimaging.

\emph{Encoding} has been shown to disentangle the contribution of correlated factors
onto brain activity (e.g. Weichwald et al Neuroimage 2015, Nasarelis et al Neuroimage 2011)
and is thus the gold standard method for neuroscientific investigation.

Nonetheless, \emph{decoding} has become increasingly popular since its introduction in
the mid 2000s (e.g. Kamitani and Tong Nature Neuroscience 2005), even though it
is not valid when the decoded features are correlated: e.g. one may decode the
category of a stimulus because of its contrast, even if contrast but not
category modulates brain activity.

Why, then, is decoding so popular? Presumably because, unlike encoding, it
efficiently combines multiple measurements (e.g. voxels of M/EEG sensors) and
allows the detection of weak effects corrupted by strong but structured noise (e.g. movement,
eye blinks etc).

Here, we introduce “back-to-back” regression (B2B), a linear model that combines the
benefits of decoding and encoding approaches. We show with i) a proof, ii)
synthetic data and iii) magnetoencephalography data that our method reliably provides a single,
unbiased, unidimensional coefficient for each factor. This coefficient fluctuates
around 0 if the factor is not causal, and tends towards 1 if it is.

An earlier version of this work was reviewed for ICLR on the
OpenReview platform: https://openreview.net/forum?id=B1lKDlHtwS. As you can see
there, the method was considered new and valid. However, the paper was
ultimately rejected because its applicability was only demonstrated for
neuroimaging.

Overall, we thus believe that this method will help resolve the existing tension
between encoding and decoding in neuroimaging and will provide a reliable means
to discriminate causal from non-causal features.

\vspace{0.5cm}

\makeletterclosing

\end{document}

### 2020-03-04 paper/introduction.tex
up

@@
 
-First, causal and non-causal factors may be numerous and collinear.
+First, causal and non-causal factors may be numerous and partially correlated.
 % In physics,
@@
 Specifically, forward modeling disentangles the independent contribution of
-collinear factors, but does not combine multidimensional observations. By
+correlated factors, but does not combine multidimensional observations. By
 contrast, backward modeling combines multiple observations, but does not
-disentangle collinear factors \citep{weichwald2015causal,
+disentangle factors that are linearly correlated \citep{weichwald2015causal,
 hebart2018deconstructing, king2018encoding}. To combine some of the benefits of forward
@@
 we apply B2B to a large neuroimaging dataset and reveal that distinct but
-collinear word features lead to distinguishable brain representations
+linearly-correlated word features lead to distinguishable brain representations
 (Section~\ref{sec:experiment_real}).

### 2020-03-04 paper/meg_exps.tex
up

@@
 We aim to identify the word features that cause a variation in brain responses.
-We consider four distinct but collinear features.
+We consider four distinct but linearly-correlated features.
 %
@@
 We compared the ability of Forward regression, Backward regression, CCA, PLS and
-B2B to estimate the causal contribution of four distinct but collinear features
+B2B to estimate the causal contribution of four distinct but linearly-correlated features
 on brain evoked responses to words.
@@
 These results illustrate that backward modeling cannot be used to estimate the
-causal contribution of collinear features.
+causal contribution of correlated features.
 

### 2020-03-04 paper/paper.tex
up

@@
 \usepackage{lineno,hyperref}
-\modulolinenumbers[5]
+% \modulolinenumbers[5]
 
@@
 
-\title{Discriminating the Linear Influences of Collinear Factors \\from Multivariate Observations: the Back-to-Back Regression}
+\title{Discriminating the Influence of Correlated Factors \\from Multivariate Observations: the Back-to-Back Regression}
 
@@
 Identifying causes solely from observations can be particularly challenging when
-i) potential factors are difficult to manipulate individually and ii)
+i) potential factors are difficult to manipulate independently and ii)
 observations are multi-dimensional. To address this issue, we
 introduce ``Back-to-Back'' regression (B2B), a linear method designed to efficiently
-measure, from a set of collinear factors, those that most plausibly account for
+measure, from a set of correlated factors, those that most plausibly account for
 multidimensional observations. First, we prove the consistency of B2B, its links to
@@
 correlation analysis and partial least squares) on causal identification when
-both the factors and the observations are collinear.
+the factors and the observations are partially collinear.
 Finally, we apply B2B to magneto-encephalography of 102 subjects
 recorded during a reading task to test whether our method appropriately disentangles
-the respective contribution of word length and word frequency - two collinear
+the respective contribution of word length and word frequency - two correlated
 factors known to cause early and late brain responses respectively. The results
 show that these two factors are better disentangled with B2B than with
-other standard techniques. We finally discuss the limits of B2B.
+other standard techniques.
 \end{abstract}
@@
 Forward and cross-decomposition models have been used to identify the causal
-contribution of collinear features onto multi-dimensional observations (e.g.
+contribution of correlated factors onto multi-dimensional observations (e.g.
 \citep{naselaris2011encoding}). These approaches typically lead to multiple
@@
 
-\clearpage
-\newpage
+\section{Acknowledgements}
+We are thankful to Gael Varoquaux and Alexandre Gramfort for the valuable
+feedback. This work was supported by ANR-17-EURE-0017 and the Fyssen Foundation
+to JRK.
 
@@
 
+\newpage
+\clearpage
 \section{Appendices}

### 2020-03-04 paper/supplementary.tex
supp
up

@@
 
-     Consider the B2B model from Equation $$Y = (XE + N)F$$ $N$ centred and full rank noise.
+     Consider the B2B model from Equation $$Y = (XE + N)F$$ $N$ centered and full rank noise.
 
-     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimise
+     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimize
 
@@
 
-\section{Additional Figures}
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_supp.pdf}
-  \caption{}
-  \label{fig:meg_supp}
-\end{figure}
-
 \end{document}

### 2020-03-05 paper/paper.tex
fig

@@
 
-\journal{Neuroimage}
+% \journal{Neuroimage}
 
@@
 \section{Introduction}
-
 \input{introduction.tex}
@@
 \begin{figure}[t!]
-  \include{fig_problem.tex}
+
+  \centering
+  \begin{tikzpicture}
+  \newcommand\posY{0}
+  \newcommand\posX{3}
+  \newcommand\posE{5}
+  \newcommand\posN{7}
+  \newcommand\posF{10}
+
+  \node[thick, draw=black, minimum height=3cm, minimum width=2cm, fill=yellow!10] (Y) at (\posY, 0){};
+  \node[] (eq) at (1.5, 0){$=$};
+  \node[] (times) at (4, 0){$\times$};
+  \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=yellow!10] (X) at (\posX, 0){};
+  \node[thick, draw=black, minimum height=1cm, minimum width=1cm, fill=red!10] (E) at (\posE, 0){};
+
+  \draw[fill=white] (\posE - 0.5 + 0.0, 0.5 - 0.0) rectangle (\posE - 0.5 + 0.0 + 0.1, 0.5 - 0.0 - 0.1);
+  \draw[fill=black] (\posE - 0.5 + 0.1, 0.5 - 0.1) rectangle (\posE - 0.5 + 0.1 + 0.1, 0.5 - 0.1 - 0.1);
+  \draw[fill=white] (\posE - 0.5 + 0.2, 0.5 - 0.2) rectangle (\posE - 0.5 + 0.2 + 0.1, 0.5 - 0.2 - 0.1);
+  \draw[fill=white] (\posE - 0.5 + 0.3, 0.5 - 0.3) rectangle (\posE - 0.5 + 0.3 + 0.1, 0.5 - 0.3 - 0.1);
+  \draw[fill=white] (\posE - 0.5 + 0.4, 0.5 - 0.4) rectangle (\posE - 0.5 + 0.4 + 0.1, 0.5 - 0.4 - 0.1);
+  \draw[fill=white] (\posE - 0.5 + 0.5, 0.5 - 0.5) rectangle (\posE - 0.5 + 0.5 + 0.1, 0.5 - 0.5 - 0.1);
+  \draw[fill=white] (\posE - 0.5 + 0.6, 0.5 - 0.6) rectangle (\posE - 0.5 + 0.6 + 0.1, 0.5 - 0.6 - 0.1);
+  \draw[fill=black] (\posE - 0.5 + 0.7, 0.5 - 0.7) rectangle (\posE - 0.5 + 0.7 + 0.1, 0.5 - 0.7 - 0.1);
+  \draw[fill=white] (\posE - 0.5 + 0.8, 0.5 - 0.8) rectangle (\posE - 0.5 + 0.8 + 0.1, 0.5 - 0.8 - 0.1);
+  \draw[fill=black] (\posE - 0.5 + 0.9, 0.5 - 0.9) rectangle (\posE - 0.5 + 0.9 + 0.1, 0.5 - 0.9 - 0.1);
+
+  \node[] (plus) at (6, 0){$+$};
+  \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=blue!10] (N) at (\posN, 0){};
+  \node[thick, draw=black, minimum height=1cm, minimum width=2cm, fill=red!10] (F) at (\posF, 0){};
+  \draw [thick] (2.25, -1.5) to [round left paren ] (2.25, 1.5);
+  \draw [thick] (7.75, -1.5) to [round right paren ] (7.75, 1.5);
+
+  \node[] (times2) at (8.5, 0){$\times$};
+
+  \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{n \times d_y}$}};
+  \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{n \times d_x}$}};
+  \node[] (annE) at (\posE, -1.8){\scalebox{0.85}{$E \in \mathbb{D}^{d_x \times d_x}$}};
+  \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{n \times d_x}$}};
+  \node[] (annF) at (\posF, -1.8){\scalebox{0.85}{$F \in \mathbb{R}^{d_x \times d_y}$}};
+
+  \node[] (labY) at (\posY, 2){Observations};
+  \node[] (labX) at (\posX, 2){Factors};
+  \node[] (labE) at (\posE, 2){Cause};
+  \node[] (labE) at (\posE, 1.7){selection};
+  \node[] (labN) at (\posN, 2){Noise};
+  \node[] (labF) at (\posF, 2){Cause-effect};
+  \node[] (labF) at (\posF, 1.7){mapping};
+
+  \node[] (sim1) at (0,-3.25) {\scalebox{0.85}{$X \sim P(X)$}};
+  \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
+  \node[] (reg1) at (6,-3.25) {$\hat{E} = \text{diag}(\underbrace{(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}})$};
+  \end{tikzpicture}
+  \caption{Back-to-back regression identifies the subset of factors $E_{ii} = 1$ in $X$ that influence some observations $Y$ by 1) regressing from $Y$ to $X$ to obtain $\hat{X}$, and 2) returning the diagonal of the regression coefficients from $X$ to $\hat{X}$.}
 \label{fig:b2b}
@@
 \input{synthetic.tex}
+\input{meg_exps.tex}
 
@@
 \label{sec:experiment_real}
-
-\input{meg_exps.tex}
 

### 2020-05-18 paper/__latexindent_temp.tex
fmri

@@
-@article{cardinali2013data,
-  title={Data assimilation: Observation influence diagnostic of a data assimilation system},
-  author={Cardinali, C},
-  journal={Shinfield Park, England: European Centre for Medium-Range Weather Forecasts: Research Department},
-  year={2013}
+@article{fmriprep1,
+    author = {Esteban, Oscar and Markiewicz, Christopher and Blair, Ross W and Moodie, Craig and Isik, Ayse Ilkay and Erramuzpe Aliaga, Asier and Kent, James and Goncalves, Mathias and DuPre, Elizabeth and Snyder, Madeleine and Oya, Hiroyuki and Ghosh, Satrajit and Wright, Jessey and Durnez, Joke and Poldrack, Russell and Gorgolewski, Krzysztof Jacek},
+    title = {{fMRIPrep}: a robust preprocessing pipeline for functional {MRI}},
+    year = {2018},
+    doi = {10.1038/s41592-018-0235-4},
+    journal = {Nature Methods}
 }
+
+@article{fmriprep2,
+    author = {Esteban, Oscar and Blair, Ross and Markiewicz, Christopher J.
+and Berleant, Shoshana L.
+and Moodie, Craig and Ma, Feilong and Isik, Ayse Ilkay and Erramuzpe, Asier and Kent, James D.
+andGoncalves, Mathias and DuPre, Elizabeth and Sitek, Kevin R.
+and Gomez, Daniel E.
+P.
+and Lurie, Daniel J.
+and Ye, Zhifang and Poldrack, Russell A.
+and Gorgolewski, Krzysztof J.},
+    title = {fMRIPrep},
+    year = 2018,
+    doi = {10.5281/zenodo.852659},
+    publisher = {Zenodo},
+    journal = {Software}
+}
+
+@article{nipype1,
+    author = {Gorgolewski, K. and Burns, C. D. and Madison, C. and Clark, D. and Halchenko, Y. O. and Waskom, M. L. and Ghosh, S.},
+    doi = {10.3389/fninf.2011.00013},
+    journal = {Frontiers in Neuroinformatics},
+    pages = 13,
+    shorttitle = {Nipype},
+    title = {Nipype: a flexible, lightweight and extensible neuroimaging data processing framework in Python},
+    volume = 5,
+    year = 2011
+}
+
+@article{nipype2,
+    author = {Gorgolewski, Krzysztof J.
+and Esteban, Oscar and Markiewicz, Christopher J.
+and Ziegler, Erik and Ellis, David Gage and Notter, Michael Philipp and Jarecka, Dorota and Johnson, Hans and Burns, Christopher and Manhães-Savio, Alexandre and Hamalainen, Carlo and Yvernault, Benjamin and Salo, Taylor and Jordan, Kesshi and Goncalves, Mathias and Waskom, Michael and Clark, Daniel and Wong, Jason and Loney, Fred and Modat, Marc and Dewey, Blake E and Madison, Cindee and Visconti di Oleggio Castello, Matteo and Clark, Michael G.
+and Dayan, Michael and Clark, Dav and Keshavan, Anisha and Pinsard, Basile and Gramfort, Alexandre and Berleant, Shoshana and Nielson, Dylan M.
+and Bougacha, Salma and Varoquaux, Gael and Cipollini, Ben and Markello, Ross and Rokem, Ariel and Moloney, Brendan and Halchenko, Yaroslav O.
+and Wassermann , Demian and Hanke, Michael and Horea, Christian and Kaczmarzyk, Jakub and Gilles de Hollander and DuPre, Elizabeth and Gillman, Ashley and Mordom, David and Buchanan, Colin and Tungaraza, Rosalia and Pauli, Wolfgang M.
+and Iqbal, Shariq and Sikka, Sharad and Mancini, Matteo and Schwartz, Yannick and Malone, Ian B.
+and Dubois, Mathieu and Frohlich, Caroline and Welch, David and Forbes, Jessica and Kent, James and Watanabe, Aimi and Cumba, Chad and Huntenburg, Julia M.
+and Kastman, Erik and Nichols, B.
+Nolan and Eshaghi, Arman and Ginsburg, Daniel and Schaefer, Alexander and Acland, Benjamin and Giavasis, Steven and Kleesiek, Jens and Erickson, Drew and Küttner, René and Haselgrove, Christian and Correa, Carlos and Ghayoor, Ali and Liem, Franz and Millman, Jarrod and Haehn, Daniel and Lai, Jeff and Zhou, Dale and Blair, Ross and Glatard, Tristan and Renfro, Mandy and Liu, Siqi and Kahn, Ari E.
+and Pérez-García, Fernando and Triplett, William and Lampe, Leonie and Stadler, Jörg and Kong, Xiang-Zhen and Hallquist, Michael and Chetverikov, Andrey and Salvatore, John and Park, Anne and Poldrack, Russell and Craddock, R.
+Cameron and Inati, Souheil and Hinds, Oliver and Cooper, Gavin and Perkins, L.
+Nathan and Marina, Ana and Mattfeld, Aaron and Noel, Maxime and Lukas Snoek and Matsubara, K and Cheung, Brian and Rothmei, Simon and Urchs, Sebastian and Durnez, Joke and Mertz, Fred and Geisler, Daniel and Floren, Andrew and Gerhard, Stephan and Sharp, Paul and Molina-Romero, Miguel and Weinstein, Alejandro and Broderick, William and Saase, Victor and Andberg, Sami Kristian and Harms, Robbert and Schlamp, Kai and Arias, Jaime and Papadopoulos Orfanos, Dimitri and Tarbert, Claire and Tambini, Arielle and De La Vega, Alejandro and Nickson, Thomas and Brett, Matthew and Falkiewicz, Marcel and Podranski, Kornelius and Linkersdörfer, Janosch and Flandin, Guillaume and Ort, Eduard and Shachnev, Dmitry and McNamee, Daniel and Davison, Andrew and Varada, Jan and Schwabacher, Isaac and Pellman, John and Perez-Guevara, Martin and Khanuja, Ranjeet and Pannetier, Nicolas and McDermottroe, Conor and Ghosh, Satrajit},
+    title = {Nipype},
+    year = 2018,
+    doi = {10.5281/zenodo.596855},
+    publisher = {Zenodo},
+    journal = {Software}
+}
+
+@article{n4,
+    author = {Tustison, N. J. and Avants, B. B. and Cook, P. A. and Zheng, Y. and Egan, A. and Yushkevich, P. A. and Gee, J. C.},
+    doi = {10.1109/TMI.2010.2046908},
+    issn = {0278-0062},
+    journal = {IEEE Transactions on Medical Imaging},
+    number = 6,
+    pages = {1310-1320},
+    shorttitle = {N4ITK},
+    title = {N4ITK: Improved N3 Bias Correction},
+    volume = 29,
+    year = 2010
+}
+
+@article{fs_reconall,
+    author = {Dale, Anders M. and Fischl, Bruce and Sereno, Martin I.},
+    doi = {10.1006/nimg.1998.0395},
+    issn = {1053-8119},
+    journal = {NeuroImage},
+    number = 2,
+    pages = {179-194},
+    shorttitle = {Cortical Surface-Based Analysis},
+    title = {Cortical Surface-Based Analysis: I. Segmentation and Surface Reconstruction},
+    url = {http://www.sciencedirect.com/science/article/pii/S1053811998903950},
+    volume = 9,
+    year = 1999
+}
+
+@article{mindboggle,
+    author = {Klein, Arno and Ghosh, Satrajit S.
+and Bao, Forrest S.
+and Giard, Joachim and Häme, Yrjö and Stavsky, Eliezer and Lee, Noah and Rossa, Brian and Reuter, Martin and Neto, Elias Chaibub and Keshavan, Anisha},
+    doi = {10.1371/journal.pcbi.1005350},
+    issn = {1553-7358},
+    journal = {PLOS Computational Biology},
+    number = 2,
+    pages = {e1005350},
+    title = {Mindboggling morphometry of human brains},
+    url = {http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005350},
+    volume = 13,
+    year = 2017
+}
+
+@article{mni152lin,
+    title = {A {Probabilistic} {Atlas} of the {Human} {Brain}: {Theory} and {Rationale} for {Its} {Development}: {The} {International} {Consortium} for {Brain} {Mapping} ({ICBM})},
+    author = {Mazziotta, John C. and Toga, Arthur W. and Evans, Alan and Fox, Peter and Lancaster, Jack},
+    volume = {2},
+    issn = {1053-8119},
+    shorttitle = {A {Probabilistic} {Atlas} of the {Human} {Brain}},
+    doi = {10.1006/nimg.1995.1012},
+    number = {2, Part A},
+    journal = {NeuroImage},
+    year = {1995},
+    pages = {89--101}
+}
+
+@article{mni152nlin2009casym,
+    title = {Unbiased nonlinear average age-appropriate brain templates from birth to adulthood},
+    author = {Fonov, VS and Evans, AC and McKinstry, RC and Almli, CR and Collins, DL},
+    doi = {10.1016/S1053-8119(09)70884-5},
+    journal = {NeuroImage},
+    pages = {S102},
+    volume = {47, Supplement 1},
+    year = 2009
+}
+
+@article{mni152nlin6asym,
+    author = {Evans, AC and Janke, AL and Collins, DL and Baillet, S},
+    title = {Brain templates and atlases},
+    doi = {10.1016/j.neuroimage.2012.01.024},
+    journal = {NeuroImage},
+    volume = {62},
+    number = {2},
+    pages = {911--922},
+    year = 2012
+}
+
+@article{ants,
+    author = {Avants, B.B. and Epstein, C.L. and Grossman, M. and Gee, J.C.},
+    doi = {10.1016/j.media.2007.06.004},
+    issn = {1361-8415},
+    journal = {Medical Image Analysis},
+    number = 1,
+    pages = {26-41},
+    shorttitle = {Symmetric diffeomorphic image registration with cross-correlation},
+    title = {Symmetric diffeomorphic image registration with cross-correlation: Evaluating automated labeling of elderly and neurodegenerative brain},
+    url = {http://www.sciencedirect.com/science/article/pii/S1361841507000606},
+    volume = 12,
+    year = 2008
+}
+
+@article{fsl_fast,
+    author = {Zhang, Y. and Brady, M. and Smith, S.},
+    doi = {10.1109/42.906424},
+    issn = {0278-0062},
+    journal = {IEEE Transactions on Medical Imaging},
+    number = 1,
+    pages = {45-57},
+    title = {Segmentation of brain {MR} images through a hidden Markov random field model and the expectation-maximization algorithm},
+    volume = 20,
+    year = 2001
+}
+
+@article{fieldmapless1,
+    author = {Wang, Sijia and Peterson, Daniel J. and Gatenby, J. C. and Li, Wenbin and Grabowski, Thomas J. and Madhyastha, Tara M.},
+    doi = {10.3389/fninf.2017.00017},
+    issn = {1662-5196},
+    journal = {Frontiers in Neuroinformatics},
+    language = {English},
+    title = {Evaluation of Field Map and Nonlinear Registration Methods for Correction of Susceptibility Artifacts in Diffusion {MRI}},
+    url = {http://journal.frontiersin.org/article/10.3389/fninf.2017.00017/full},
+    volume = 11,
+    year = 2017
+}
+
+@phdthesis{fieldmapless2,
+    address = {Berlin},
+    author = {Huntenburg, Julia M.},
+    language = {eng},
+    school = {Freie Universität},
+    title = {Evaluating nonlinear coregistration of {BOLD} {EPI} and T1w images},
+    type = {Master's Thesis},
+    url = {http://hdl.handle.net/11858/00-001M-0000-002B-1CB5-A},
+    year = 2014
+}
+
+@article{fieldmapless3,
+    author = {Treiber, Jeffrey Mark and White, Nathan S.
+and Steed, Tyler Christian and Bartsch, Hauke and Holland, Dominic and Farid, Nikdokht and McDonald, Carrie R.
+and Carter, Bob S.
+and Dale, Anders Martin and Chen, Clark C.},
+    doi = {10.1371/journal.pone.0152472},
+    issn = {1932-6203},
+    journal = {PLOS ONE},
+    number = 3,
+    pages = {e0152472},
+    title = {Characterization and Correction of Geometric Distortions in 814 Diffusion Weighted Images},
+    url = {http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0152472},
+    volume = 11,
+    year = 2016
+}
+
+@article{flirt,
+    title = {A global optimisation method for robust affine registration of brain images},
+    volume = {5},
+    issn = {1361-8415},
+    url = {http://www.sciencedirect.com/science/article/pii/S1361841501000366},
+    doi = {10.1016/S1361-8415(01)00036-6},
+    number = {2},
+    urldate = {2018-07-27},
+    journal = {Medical Image Analysis},
+    author = {Jenkinson, Mark and Smith, Stephen},
+    year = {2001},
+    keywords = {Affine transformation, flirt, fsl, Global optimisation, Multi-resolution search, Multimodal registration, Robustness},
+    pages = {143--156}
+}
+
+@article{mcflirt,
+    author = {Jenkinson, Mark and Bannister, Peter and Brady, Michael and Smith, Stephen},
+    doi = {10.1006/nimg.2002.1132},
+    issn = {1053-8119},
+    journal = {NeuroImage},
+    number = 2,
+    pages = {825-841},
+    title = {Improved Optimization for the Robust and Accurate Linear Registration and Motion Correction of Brain Images},
+    url = {http://www.sciencedirect.com/science/article/pii/S1053811902911328},
+    volume = 17,
+    year = 2002
+}
+
+@article{bbr,
+    author = {Greve, Douglas N and Fischl, Bruce},
+    doi = {10.1016/j.neuroimage.2009.06.060},
+    issn = {1095-9572},
+    journal = {NeuroImage},
+    number = 1,
+    pages = {63-72},
+    title = {Accurate and robust brain image alignment using boundary-based registration},
+    volume = 48,
+    year = 2009
+}
+
+@article{aroma,
+    author = {Pruim, Raimon H. R. and Mennes, Maarten and van Rooij, Daan and Llera, Alberto and Buitelaar, Jan K. and Beckmann, Christian F.},
+    doi = {10.1016/j.neuroimage.2015.02.064},
+    issn = {1053-8119},
+    journal = {NeuroImage},
+    number = {Supplement C},
+    pages = {267-277},
+    shorttitle = {ICA-AROMA},
+    title = {ICA-{AROMA}: A robust {ICA}-based strategy for removing motion artifacts from fMRI data},
+    url = {http://www.sciencedirect.com/science/article/pii/S1053811915001822},
+    volume = 112,
+    year = 2015
+}
+
+@article{power_fd_dvars,
+    author = {Power, Jonathan D. and Mitra, Anish and Laumann, Timothy O. and Snyder, Abraham Z. and Schlaggar, Bradley L. and Petersen, Steven E.},
+    doi = {10.1016/j.neuroimage.2013.08.048},
+    issn = {1053-8119},
+    journal = {NeuroImage},
+    number = {Supplement C},
+    pages = {320-341},
+    title = {Methods to detect, characterize, and remove motion artifact in resting state fMRI},
+    url = {http://www.sciencedirect.com/science/article/pii/S1053811913009117},
+    volume = 84,
+    year = 2014
+}
+
+@article{confounds_satterthwaite_2013,
+    author = {Satterthwaite, Theodore D.
+and Elliott, Mark A.
+and Gerraty, Raphael T.
+and Ruparel, Kosha and Loughead, James and Calkins, Monica E.
+and Eickhoff, Simon B.
+and Hakonarson, Hakon and Gur, Ruben C.
+and Gur, Raquel E.
+and Wolf, Daniel H.},
+    doi = {10.1016/j.neuroimage.2012.08.052},
+    issn = {10538119},
+    journal = {NeuroImage},
+    number = 1,
+    pages = {240--256},
+    title = {{An improved framework for confound regression and filtering for control of motion artifact in the preprocessing of resting-state functional connectivity data}},
+    url = {http://linkinghub.elsevier.com/retrieve/pii/S1053811912008609},
+    volume = 64,
+    year = 2013
+}
+
+@article{nilearn,
+    author = {Abraham, Alexandre and Pedregosa, Fabian and Eickenberg, Michael and Gervais, Philippe and Mueller, Andreas and Kossaifi, Jean and Gramfort, Alexandre and Thirion, Bertrand and Varoquaux, Gael},
+    doi = {10.3389/fninf.2014.00014},
+    issn = {1662-5196},
+    journal = {Frontiers in Neuroinformatics},
+    language = {English},
+    title = {Machine learning for neuroimaging with scikit-learn},
+    url = {https://www.frontiersin.org/articles/10.3389/fninf.2014.00014/full},
+    volume = 8,
+    year = 2014
+}
+
+@article{lanczos,
+    author = {Lanczos, C.},
+    doi = {10.1137/0701007},
+    issn = {0887-459X},
+    journal = {Journal of the Society for Industrial and Applied Mathematics Series B Numerical Analysis},
+    number = 1,
+    pages = {76-85},
+    title = {Evaluation of Noisy Data},
+    url = {http://epubs.siam.org/doi/10.1137/0701007},
+    volume = 1,
+    year = 1964
+}
+
+@article{compcor,
+    author = {Behzadi, Yashar and Restom, Khaled and Liau, Joy and Liu, Thomas T.},
+    doi = {10.1016/j.neuroimage.2007.04.042},
+    issn = {1053-8119},
+    journal = {NeuroImage},
+    number = 1,
+    pages = {90-101},
+    title = {A component based noise correction method ({CompCor}) for {BOLD} and perfusion based fMRI},
+    url = {http://www.sciencedirect.com/science/article/pii/S1053811907003837},
+    volume = 37,
+    year = 2007
+}
+
+@article{hcppipelines,
+    author = {Glasser, Matthew F.
+and Sotiropoulos, Stamatios N.
+and Wilson, J.
+Anthony and Coalson, Timothy S.
+and Fischl, Bruce and Andersson, Jesper L.
+and Xu, Junqian and Jbabdi, Saad and Webster, Matthew and Polimeni, Jonathan R.
+and Van Essen, David C.
+and Jenkinson, Mark},
+    doi = {10.1016/j.neuroimage.2013.04.127},
+    issn = {1053-8119},
+    journal = {NeuroImage},
+    pages = {105-124},
+    series = {Mapping the Connectome},
+    title = {The minimal preprocessing pipelines for the Human Connectome Project},
+    url = {http://www.sciencedirect.com/science/article/pii/S1053811913005053},
+    volume = 80,
+    year = 2013
+}
+
+@article{fs_template,
+    author = {Reuter, Martin and Rosas, Herminia Diana and Fischl, Bruce},
+    doi = {10.1016/j.neuroimage.2010.07.020},
+    journal = {NeuroImage},
+    number = 4,
+    pages = {1181-1196},
+    title = {Highly accurate inverse consistent registration: A robust approach},
+    volume = 53,
+    year = 2010
+}
+
+@article{afni,
+    author = {Cox, Robert W. and Hyde, James S.},
+    doi = {10.1002/(SICI)1099-1492(199706/08)10:4/5<171::AID-NBM453>3.0.CO;2-L},
+    journal = {NMR in Biomedicine},
+    number = {4-5},
+    pages = {171-178},
+    title = {Software tools for analysis and visualization of fMRI data},
+    volume = 10,
+    year = 1997
+}
+
+@article{posse_t2s,
+    author = {Posse, Stefan and Wiese, Stefan and Gembris, Daniel and Mathiak, Klaus and Kessler, Christoph and Grosse-Ruyken, Maria-Liisa and Elghahwagi, Barbara and Richards, Todd and Dager, Stephen R.
+and Kiselev, Valerij G.},
+    doi = {10.1002/(SICI)1522-2594(199907)42:1<87::AID-MRM13>3.0.CO;2-O},
+    journal = {Magnetic Resonance in Medicine},
+    number = 1,
+    pages = {87-97},
+    title = {Enhancement of {BOLD}-contrast sensitivity by single-shot multi-echo functional {MR} imaging},
+    volume = 42,
+    year = 1999
+}

### 2020-05-18 paper/appendix.tex
fmri

@@
 
+\begin{wrapfigure}{r}{.5\textwidth}
+  \vspace{-11ex}
+  \begin{center}
+    \includegraphics[width=0.48\textwidth,
+                     trim=0cm 0cm 0cm 0cm,
+                     clip=True]{figures/fmri_decod.pdf}
+    \includegraphics[width=0.48\textwidth,
+                     trim=0cm 0cm 0cm 0cm,
+                     clip=True]{figures/fmri_betas.pdf}
+
+    \label{fig:fmri_strip}
+  \end{center}
+  \caption{Top. P-values across subjects derived from a Backward decoding of
+  each feature (columns). Bottom. P-values across subjects derived from
+  the $\beta$ coefficients of a Forward Model and the $\hat E$ of an unbiased
+  B2B (i.e. B2B where $H$ is not regularized). Each cell displays the left
+  hemisphere view laterally (top) or ventrally (bottom)}
+  \vspace{-6ex}
+\end{wrapfigure}
+
 \subsection{Robustness to increasing number of factors}

### 2020-05-18 paper/fmri_exps.tex
fmri

%\newpage
%\subsection{Dataset}

Next, we apply our method to brain imaging data from the anonymized multimodal
neuroimaging ``Mother Of all Unification Studies'' (MOUS) dataset
\citep{schoffelen2019204}. The dataset contains functional Magnetic Resonance
Imaging (fMRI) and magneto-encephalography (MEG)
recordings of each 102 healthy native-Dutch adults who participated in a reading
task. Thirteen subjects were excluded from the analysis (12 MEG and 1 fMRI)
because of missing files or corrupted headers.
%
Subjects were exposed to a rapid serial visual presentation of Dutch words. The
word lists consisted of 120 sentences, and scrambled lists of the same words.
Each word was presented on the computer screen for 351ms on average (min: 300ms,
max: 1400ms). Successive words were separated by a blank screen for 300ms, and
successive sentences were separated by an empty screen for a few (3-4) seconds.

\subsubsection{fMRI preprocessing}
Results included in this manuscript come from preprocessing performed
using \emph{fMRIPrep} 20.0.7 (\citet{fmriprep1}; \citet{fmriprep2};
RRID:SCR\_016216), which is based on \emph{Nipype} 1.4.2
(\citet{nipype1}; \citet{nipype2}; RRID:SCR\_002502).

\begin{description}
\item[Anatomical data preprocessing]
A total of 2 T1-weighted (T1w) images were found within the input BIDS
dataset. All of them were corrected for intensity non-uniformity (INU)
with \texttt{N4BiasFieldCorrection} \citep{n4}, distributed with ANTs
2.2.0 \citep[RRID:SCR\_004757]{ants}. The T1w-reference was then
skull-stripped with a \emph{Nipype} implementation of the
\texttt{antsBrainExtraction.sh} workflow (from ANTs), using OASIS30ANTs
as target template. Brain tissue segmentation of cerebrospinal fluid
(CSF), white-matter (WM) and gray-matter (GM) was performed on the
brain-extracted T1w using \texttt{fast} \citep[FSL 5.0.9,
RRID:SCR\_002823,][]{fsl_fast}. A T1w-reference map was computed after
registration of 2 T1w images (after INU-correction) using
\texttt{mri\_robust\_template} \citep[FreeSurfer 6.0.1,][]{fs_template}.
Brain surfaces were reconstructed using \texttt{recon-all}
\citep[FreeSurfer 6.0.1, RRID:SCR\_001847,][]{fs_reconall}, and the
brain mask estimated previously was refined with a custom variation of
the method to reconcile ANTs-derived and FreeSurfer-derived
segmentations of the cortical gray-matter of Mindboggle
\citep[RRID:SCR\_002438,][]{mindboggle}. Volume-based spatial
normalization to two standard spaces (MNI152NLin2009cAsym,
MNI152NLin6Asym) was performed through nonlinear registration with
\texttt{antsRegistration} (ANTs 2.2.0), using brain-extracted versions
of both T1w reference and the T1w template. The following templates were
selected for spatial normalization: \emph{ICBM 152 Nonlinear
Asymmetrical template version 2009c} {[}\citet{mni152nlin2009casym},
RRID:SCR\_008796; TemplateFlow ID: MNI152NLin2009cAsym{]}, \emph{FSL's
MNI ICBM 152 non-linear 6th Generation Asymmetric Average Brain
Stereotaxic Registration Model} {[}\citet{mni152nlin6asym},
RRID:SCR\_002823; TemplateFlow ID: MNI152NLin6Asym{]},
\item[Functional data preprocessing]
For each of the 2 BOLD runs found per subject (across all tasks and
sessions), the following preprocessing was performed. First, a reference
volume and its skull-stripped version were generated using a custom
methodology of \emph{fMRIPrep}. Susceptibility distortion correction
(SDC) was omitted. The BOLD reference was then co-registered to the T1w
reference using \texttt{bbregister} (FreeSurfer) which implements
boundary-based registration \citep{bbr}. Co-registration was configured
with six degrees of freedom. Head-motion parameters with respect to the
BOLD reference (transformation matrices, and six corresponding rotation
and translation parameters) are estimated before any spatiotemporal
filtering using \texttt{mcflirt} \citep[FSL 5.0.9,][]{mcflirt}. BOLD
runs were slice-time corrected using \texttt{3dTshift} from AFNI
20160207 \citep[RRID:SCR\_005927]{afni}. The BOLD time-series were
resampled onto the following surfaces (FreeSurfer reconstruction
nomenclature): \emph{fsnative}, \emph{fsaverage5}. The BOLD time-series
(including slice-timing correction when applied) were resampled onto
their original, native space by applying the transforms to correct for
head-motion. These resampled BOLD time-series will be referred to as
\emph{preprocessed BOLD in original space}, or just \emph{preprocessed
BOLD}. The BOLD time-series were resampled into standard space,
generating a \emph{preprocessed BOLD run in MNI152NLin2009cAsym space}.
First, a reference volume and its skull-stripped version were generated
using a custom methodology of \emph{fMRIPrep}. Automatic removal of
motion artifacts using independent component analysis
\citep[ICA-AROMA,][]{aroma} was performed on the \emph{preprocessed BOLD
on MNI space} time-series after removal of non-steady state volumes and
spatial smoothing with an isotropic, Gaussian kernel of 6mm FWHM
(full-width half-maximum). Corresponding ``non-aggresively'' denoised
runs were produced after such smoothing.
% Additionally, the
% ``aggressive'' noise-regressors were collected and placed in the
% corresponding confounds file.
% Several confounding time-series were
% calculated based on the \emph{preprocessed BOLD}: framewise displacement
% (FD), DVARS and three region-wise global signals. FD and DVARS are
% calculated for each functional run, both using their implementations in
% \emph{Nipype} \citep[following the definitions by][]{power_fd_dvars}.
% The three global signals are extracted within the CSF, the WM, and the
% whole-brain masks. Additionally, a set of physiological regressors were
% extracted to allow for component-based noise correction
% \citep[\emph{CompCor},][]{compcor}. Principal components are estimated
% after high-pass filtering the \emph{preprocessed BOLD} time-series
% (using a discrete cosine filter with 128s cut-off) for the two
% \emph{CompCor} variants: temporal (tCompCor) and anatomical (aCompCor).
% tCompCor components are then calculated from the top 5\% variable voxels
% within a mask covering the subcortical regions. This subcortical mask is
% obtained by heavily eroding the brain mask, which ensures it does not
% include cortical GM regions. For aCompCor, components are calculated
% within the intersection of the aforementioned mask and the union of CSF
% and WM masks calculated in T1w space, after their projection to the
% native space of each functional run (using the inverse BOLD-to-T1w
% transformation). Components are also calculated separately within the WM
% and CSF masks. For each CompCor decomposition, the \emph{k} components
% with the largest singular values are retained, such that the retained
% components' time series are sufficient to explain 50 percent of variance
% across the nuisance mask (CSF, WM, combined, or temporal). The remaining
% components are dropped from consideration. The head-motion estimates
% calculated in the correction step were also placed within the
% corresponding confounds file. The confound time series derived from head
% motion estimates and global signals were expanded with the inclusion of
% temporal derivatives and quadratic terms for each
% \citep{confounds_satterthwaite_2013}. Frames that exceeded a threshold
% of 0.5 mm FD or 1.5 standardised DVARS were annotated as motion
% outliers.
% All resamplings can be performed with \emph{a single
% interpolation step} by composing all the pertinent transformations
% (i.e.~head-motion transform matrices, susceptibility distortion
% correction when available, and co-registrations to anatomical and output
% spaces). Gridded (volumetric) resamplings were performed using
% \texttt{antsApplyTransforms} (ANTs), configured with Lanczos
% interpolation to minimize the smoothing effects of other kernels
% \citep{lanczos}. Non-gridded (surface) resamplings were performed using
% \texttt{mri\_vol2surf} (FreeSurfer).
\end{description}

Many internal operations of \emph{fMRIPrep} use \emph{Nilearn} 0.6.2
\citep[RRID:SCR\_001362]{nilearn}, mostly within the functional
processing workflow. For more details of the pipeline, see
\href{https://fmriprep.readthedocs.io/en/latest/workflows.html}{the
section corresponding to workflows in \emph{fMRIPrep}'s documentation}.

% The above boilerplate text was automatically generated by fMRIPrep with
% the express intention that users should copy and paste this text into
% their manuscripts \emph{unchanged}. It is released under the
% \href{https://creativecommons.org/publicdomain/zero/1.0/}{CC0} license.

\subsubsection{fMRI analysis}

The preprocessed volumetric fMRI data was linearly projected to its closest surface
using nilearn $vol_to_surf$ function with a 3 mm radius. A surface search-light
analysis was then implemented by concatenating the surface
fMRI data within a 8mm radius of each vertex. For each subject and vertex, we
thus build an observation matrix $Y \in \mathbb{R}^{n \times d_y}$ of $n\approx$ 1400 words
by $d_y\approx40$ vertices. Each
of the columns of $Y$ is normalized to have zero mean and unit variance.

These multidimensional brain observations were to be accounted for (or decoded by) four features.

\subsubsection{Feature definition}

We aim to identify the word features that cause a variation in brain responses. We consider four distinct but colinear features.
%
First, 'Word Length' refers to the total number of letters. Word Length is expected to specifically cause a variation in the early evoked MEG responses (i.e. from 100 ms after stimulus onset) elicited by the retinotopically-tuned visual cortices (e.g. \citep{pegado2014timing}.).
%
Second, 'Word Frequency' indexes how frequently each word appears in Dutch and was derived with the the Zipf logarithmic scale of \citep{van2014subtlex} provided by the WordFreq package \citep{speerwordfreq}. Word Frequency is expected to specifically cause a variation in the late evoked MEG responses (i.e. from 400 ms), because it variably engages semantic processes in the temporal cortices \citep{kutas2011thirty}.
%
Third, 'Word Function' indicates whether each word is a content word (i.e. a noun, a verb, an adjective or an adverb) or a function word (i.e. a preposition, a conjunction, a determinant, a pronoun or a numeral), and was derived from Spacy's part of speech tagger \citep{spacy2}. To our knowledge, this feature has not been thouroughly investigated with MEG. Its causal contribution to reading processes in the brain thus remains unclear.
%
Finally, to verify that B2B and other methods would not inadequately identify non-causal features, we added a dummy feature, constructed from a noisy combination of Word Length and Word Frequency:
$dummy = z(length) + z(frequency) + \mathcal{N}$, where $z$ normalizes features and $\mathcal{N}$ is a random vector sampling Gaussian distribution (all terms thus have a zero-mean and a unit-variance). This procedure yields an $X \in \mathbb{R}^{n \times d_x}$ matrix of $n\approx$ 2700 words by
$d_x=4$ features for each subject. Each of the columns of $X$ is normalized to
have a mean and a standard deviation of 0 and 1 respectively.

To account for the delay of blood oxygenation level dependent responses, the
features were convolved using the default hemodynamic response function Glover
of nilearn \citep{nilearn}.

\subsubsection{Models and statistics}

We compare B2B to four standard methods: Forward regression, Backward
regression, Regularized CCA and PLS, as implemented in scikit-learn
\citep{sklearn} and pyrcca \citep{bilenko2016pyrcca}, and optimized with nested cross-validation over twenty $l2$
regularization parameters logarithmically spaced between $10^{-4}$ and $10^4$
(for regression methods) or 1 to 4 canonical components (for cross-decomposition
methods).

We used the feature importance described in Algorithm \ref{algorithm:b2b_fi} to assess the extent to which each feature $X_i$ specifically improves the prediction of held-out $Y$ data, using a 5-fold cross-validation (with shuffled trials to homogeneize the distributions between the training and testing splits).

Each model was implemented for each subject and each time sample independently. Pairwise comparison between models were performed using a two-sided Wilcoxon test across subjects (n=90) using the average $\Delta R$ across time. Corresponding effect sizes are shown in Figure~\ref{fig:meg_results}, and p-values are reported below.

\begin{figure}
  \begin{center}
          \includegraphics[width=.8\textwidth,
                       trim=1cm 1cm 2cm 2cm,
                       clip=True]{figures/fmri_delta_r.png}
          \includegraphics[width=.8\textwidth,
                       trim=0cm 0cm 0cm 0cm,
                       clip=True]{figures/fmri_strip.png}

      \label{fig:fmri_delta_r}
  \end{center}
  \caption{Top. Multiple models (rows) are compared on their ability to
  reliably predict out-of-sample fMRI signals evoked by words by quantifying
  the improvement of correlation coefficient $\Delta R$ for each of the four
  features (columns). Each cell displays the left hemisphere view laterally
  (top) or ventrally (bottom). Bottom left. Region
  of Interest (ROI), defined by the cortex vertices that were reliably
  predicted for by the gold standard Forward Model using all features and
  thresholded with a Wilcoxon at $p<0.001$ (uncorrected) across subjects. Bottom
  Right. Average $\Delta R$ within the ROI for each subject
  (dot); top horizontal lines indicate when B2B significantly outperforms
  other methods across subjects.}
\end{figure}

\subsubsection{Results}
We compared the ability of Forward regression, Backward regression, CCA, PLS
and B2B to estimate the causal contribution of four distinct but collinear
features on brain responses to words.

Supplementary Figure \ref{fig:fmri_supp} shows that the Backward model decodes
the dummy variable well above chance level. In addition, it decodes both Word
Length and Word Frequency across large sections of the cortex, even though
these features are known to specifically influence early visual cortex and
associative cortices respectively. These results illustrate that backward
modeling cannot be used to estimate the causal contribution of collinear
features.

On the contrary, the $\beta$ coefficients of the Forward model reveals the
expected spatial specificity of Word Length and Word Frequency. However, these
coefficients are moderately consistent across subjects, and can thus
dramatically limit the possibility to achieve high levels of
significance, as evaluated with a Wilcoxon tests across subjects at each vertex.

In principle, the inter-subject variability could be less impactful with
multivariate observations methods such as CCA and PLS. However, these
cross-decompositions do not provide a single and interpretable coefficient
for each feature at each vertex.

On the contrary, the $\hat E$ coefficients of an unbiased B2B (i.e. a B2B where
$H$ is not regularized) derive from multivariate fMRI observations but are
unidimensional for each feature. In fact, $\hat E$ reaches higher levels of
significance, as evaluated with a Wilcoxon test across subjects at each vertex
than the Forward $\beta$ \ref{fig:fmri_supp}. Note that a direct
comparison between the Forward coefficients $\beta$ and those of B2B ($\hat E$)
is not feasible, as these two do not have the same units.

To fairly compare the Forward, CCA, PLS and B2B methods on a common metric,
without being over-dependent on the
potentially invalid assumptions of each model, we compare them on their ability
to improve Y
predictions of out-of-sample data when one factor is introduced into the model.
This analysis leads to a $\Delta R$ as described in Algorithm
\ref{algorithm:b2b_fi}). As expected, the Forward, CCA, PLS
and B2B method predicted that the Dummy Variable does not improve the $Y$
prediction.

The average $\Delta R$ across subjects is displayed for each method and each
feature in \ref{fig:fmri_r}, and suggest that B2B surpasses other methods.
To formally quantify these comparisons, we compare, for each subject separately,
the average $\Delta R$ across vertices. To avoid including irrelevant brain
regions, we restrict this analysis to vertices which could be reliably accounted
for by the Forward model ($p<.001$, not corrected across vertices)
\ref{fig:fmri_r}. The results show that B2B significantly outperforms all
other methods on all features (all $p<0.0001$) but the dummy variable
(min $p=0.0852$).

### 2020-05-18 paper/meg_exps.tex
fig
fmri

@@
 %\subsection{Dataset}
-
-Next, we apply our method to brain imaging data from the anonymized multimodal
-neuroimaging ``Mother Of all Unification Studies'' (MOUS) dataset
-\citep{schoffelen2019204}. The dataset contains magneto-encephalography (MEG)
-recordings of 102 healthy native-Dutch adults who participated in a reading
-task. Twelve subjects were excluded from the analysis because of corrupted file headers.
-%
-Subjects were exposed to a rapid serial visual presentation of Dutch words. The
-word lists consisted of 120 sentences, and scrambled lists of the same words.
-Each word was presented on the computer screen for 351ms on average (min: 300ms,
-max: 1400ms). Successive words were separated by a blank screen for 300ms, and
-successive sentences were separated by an empty screen for a few (3-4) seconds.
 
@@
 of the columns of $Y$ is normalized to have zero mean and unit variance.
+
+We use the same features as for the fMRI experiments, except that the features
+were not convolved by an hemodynamic response function. This yields an $X \in \mathbb{R}^{n \times d_x}$
+matrix of $n\approx$ 2,700 words by $d_x=4$ features for each subject. Each of
+the columns of $X$ is normalized to have a mean and a standard deviation of 0
+and 1 respectively.
+
+We compare B2B against other methods following the same procedure as for the
+fMRI experiments, except that the search light is applied across time samples.
 
@@
 
-\subsubsection{Feature definition}
-
-We aim to identify the word features that cause a variation in brain responses.
-We consider four distinct but linearly-correlated features.
-%
-First, 'Word Length' refers to the total number of letters. Word Length is
-expected to specifically cause a variation in the early evoked MEG responses
-(i.e. from 100 ms after stimulus onset) elicited by the retinotopically-tuned
-visual cortices (e.g. \citep{pegado2014timing}.).
-%
-Second, 'Word Frequency' indexes how frequently each word appears in Dutch and
-was derived with the the Zipf logarithmic scale of \citep{van2014subtlex}
-provided by the WordFreq package \citep{speerwordfreq}. Word Frequency is
-expected to specifically cause a variation in the late evoked MEG responses
-(i.e. from 400 ms), because it variably engages semantic processes in the
-temporal cortices \citep{kutas2011thirty}.
-%
-Third, 'Word Function' indicates whether each word is a content word (i.e. a
-noun, a verb, an adjective or an adverb) or a function word (i.e. a preposition,
-a conjunction, a determinant, a pronoun or a numeral), and was derived from
-Spacy's part of speech tagger \citep{spacy2}. To our knowledge, this feature has
-not been thoroughly investigated with MEG. Its causal contribution to reading
-processes in the brain thus remains unclear.
-%
-Finally, to verify that B2B and other methods would not inadequately identify
-non-causal features, we added a dummy feature, constructed from a noisy
-combination of Word Length and Word Frequency: $dummy = z(length) + z(frequency)
-+ \mathcal{N}$, where $z$ normalizes features and $\mathcal{N}$ is a random
-vector sampling Gaussian distribution (all terms thus have a zero-mean and a
-unit-variance). This procedure yields an $X \in \mathbb{R}^{n \times d_x}$
-matrix of $n\approx$ 2,700 words by $d_x=4$ features for each subject. Each of
-the columns of $X$ is normalized to have a mean and a standard deviation of 0
-and 1 respectively.
-
-\subsubsection{Models and statistics}
-
-We compare B2B to four standard methods: Forward regression, Backward
-regression, CCA and PLS, as implemented in scikit-learn \citep{sklearn} and
-\citep{bilenko2016pyrcca}, and optimized with nested cross-validation over
-twenty $l2$ regularization parameters logarithmically spaced between $10^{-4}$
-and $10^4$ (for regression and CCA methods) or 1 to 4 canonical components (for
-PLS).
-
-We used the feature importance described in Algorithm \ref{algorithm:b2b_fi} to
-assess the extent to which each feature $X_i$ specifically improves the
-prediction of held-out $Y$ data, using a five-fold cross-validation (with shuffled
-trials to homogeneize the distributions between the training and testing
-splits).
-
-Each model was implemented for each subject and each time sample independently.
-Pairwise comparison between models were performed using a two-sided Wilcoxon
-test across subjects (n=90) using the average $\Delta R$ across time.
-Corresponding effect sizes are shown in Figure~\ref{fig:meg_results}, and
-p-values are reported below.
-
-\subsubsection{Results}
-We compared the ability of Forward regression, Backward regression, CCA, PLS and
-B2B to estimate the causal contribution of four distinct but linearly-correlated features
-on brain evoked responses to words.
-
-\begin{wrapfigure}{r}{.6\textwidth}
+\begin{figure}
   %\vspace{-12ex}
   \begin{center}
-    \includegraphics[width=0.6\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg.pdf}
+    \includegraphics[width=0.6\textwidth,
+                     trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg.pdf}
   \end{center}
@@
 }
-\end{wrapfigure}
+\end{figure}
+
+\subsubsection{Results}
+We compared the ability of Forward regression, Backward regression, CCA, PLS and
+B2B to estimate the causal contribution of four distinct but linearly-correlated features
+on brain evoked responses to words.
 

### 2020-05-18 paper/paper.tex
fmri

@@
 the factors and the observations are partially collinear.
-Finally, we apply B2B to magneto-encephalography of 102 subjects
-recorded during a reading task to test whether our method appropriately disentangles
+Finally, we apply B2B to magneto-encephalography and fMRI of 102 subjects
+recorded during a one hour long reading task to test whether our method appropriately disentangles
 the respective contribution of word length and word frequency - two correlated
-factors known to cause early and late brain responses respectively. The results
+factors known to cause early visual and late associative cortical responses respectively. The results
 show that these two factors are better disentangled with B2B than with
@@
   Magnetoencephalography \sep
+  fMRI \sep
   Decoding \sep
@@
   Reading \sep
-  N400
 \end{keyword}
@@
 \label{sec:experiment_synthetic}
+\input{synthetic.tex}
 
-\input{synthetic.tex}
+\subsection{functional Magnetic Resonance Imaging}
+\label{sec:experiment_fmri}
+\input{fmri_exps.tex}
+
+\subsection{Magnetoencephalography}
+\label{sec:experiment_real}
 \input{meg_exps.tex}
-
-\subsection{Magnetoencephalography data}
-\label{sec:experiment_real}
 

### 2020-05-19 paper/introduction.tex
intro

@@
 (e.g. brain activity) from a minimal combination of competing factors (e.g.
-word length, word frequency), and analytically investigate,
+word length, word frequency), and analytically investigate
 the estimated contribution of each factor from the model's coefficients
@@
 The second challenge to measuring causal influence is that observations can be
-multidimensional. The relationship between causes and effects is thus often
-considered in a \emph{backward} manner, by training models to maximally predict causes
-from multidimensional observations. For example, brain activity is often
+multidimensional.
+For example, brain activity is often
 recorded with hundreds or thousands of simultaneous measurements via functional
@@
 physiological probes \citep{friston1994statistical,steinmetz2018challenges}.
-As simultaneous measurements may be affected by common noise sources, it is
-common to use backward
-modeling, by, for example, fitting a support vector machine across multiple
-sensors to decode the category of a stimulus \citep{norman2006beyond,cichy2014resolving,
-kriegeskorte2008representational}.
+The relationship between putative causes and observations is thus often
+done by training models in a \emph{backward}
+fashion: i.e. from observations to putative causal factors. For example, it is
+common to fit a support vector machine across multiple
+brain voxels or multiple electrodes to detect the
+category of a stimulus independently of common noise sources (e.g. head
+movements, eye blink)
+\citep{norman2006beyond,cichy2014resolving,
+kriegeskorte2008representational, king2018encoding}.
 
@@
 techniques such as Partial Least Squares (PLS) and Canonical Correlation
-Analysis (CCA) \citep{de2019multiway}. CCA and PLS aim to find, from two sets of
-data $X$ and $Y$, the components $H$ and $G$ were $XH$ and $YG$ are maximally
-correlated or maximally covarying respectively. Because CCA and PLS are based on
-a generalized eigen decomposition, their resulting coefficients are mixing the
-features of $X$ and $Y$ in a way that makes them notoriously difficult to
+Analysis (CCA) \citep{de2019multiway, bilenko2016pyrcca}. CCA and PLS aim to find, from two sets of
+data $X$ and $Y$, the components $H$ and $G$ such that $XH$ and $YG$ are maximally
+correlated or maximally covarying respectively.
+
+While CCA and PLS can make use
+of multidimensional features and observations, they are not designed
+for feature discovery. First, these methods are not not directional: observations
+and factors can be assigned to either $X$ or $Y$. Second, these method project $X$
+and $Y$ onto a reduced but nonetheless multidimensional space. Third, because
+CCA and PLS are based on a generalized eigen decomposition, their resulting
+coefficients mix the features of $X$ and $Y$ in a way that makes them notoriously difficult to
 interpret \citep{lebart1995statistique}.
@@
 the benefits of forward and backward modeling (Section~\ref{sec:algorithm}), but
-also provides robust, interpretable, unidimensional and unbiased coefficients for
-each of tested factor.
+can also provide robust, interpretable, unidimensional and unbiased coefficients for
+each factor.
 
-The present paper focuses on the restricted issue of disentangling the
-influence of linearly correlated predictors ($X$) onto noisy multivariate
-observations ($Y$). The present approach thus differs from other causal
-discovery algorithms based on temporal-delays and/or nonlinear interactions in
-systems where the directionality of causation (from X to Y or vice versa) is
-unknown (e.g. \citep{peters2017elements, granger1969investigating,
-janzing2013quantifying, scholkopf2016modeling}.
-
-After detailing B2B method and proving its convergence
+After detailing B2B and proving its convergence
 (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms
@@
 disentangling causal factors (Section~\ref{sec:experiment_synthetic}). Finally,
-we apply B2B to a large neuroimaging dataset and reveal that distinct but
+we apply B2B to large neuroimaging datasets and reveal that distinct but
 linearly-correlated word features lead to distinguishable brain representations

### 2020-05-19 paper/paper.tex
abstract

@@
 the factors and the observations are partially collinear.
-Finally, we apply B2B to magneto-encephalography and fMRI of 102 subjects
+Finally, we apply B2B to magneto-encephalography and functional Magnetic
+Resonance Imaging of 102 subjects
 recorded during a one hour long reading task to test whether our method appropriately disentangles
-the respective contribution of word length and word frequency - two correlated
-factors known to cause early visual and late associative cortical responses respectively. The results
-show that these two factors are better disentangled with B2B than with
+the respective contribution of collinear factors such as word length, word
+frequency respectively known to cause early visual and late associative cortical
+responses. The results
+show that these factors are better disentangled with B2B than with
 other standard techniques.
@@
   Feature Discovery \sep
-  Magnetoencephalography \sep
+  MEG \sep
   fMRI \sep
@@
 
+% Here, we focus on the restricted issue of disentangling the
+% influence of linearly correlated predictors ($X$) onto noisy multivariate
+% observations ($Y$). The present approach thus differs from other causal
+% discovery algorithms based on temporal-delays and/or nonlinear interactions in
+% systems where the directionality of causation (from X to Y or vice versa) is
+% unknown (e.g. \citep{peters2017elements, granger1969investigating,
+% janzing2013quantifying, scholkopf2016modeling}.
+
 Finally, CCA has been used in neuroimaging for a variety of purposes such as

### 2020-05-19 paper/synthetic.tex
synthetic

@@
 %
-The purpose of these experiments is to evaluate the ability of B2B in terms of
-prediction of independent and identically distributed data, as well as a method
-to recover causal factors.
+The purpose of these experiments is to evaluate the ability of B2B on its
+ability to 1) accurately predict independent and identically distributed data,
+and 2) recover causal factors when the ground truth is known.
 
@@
 
-We compare the performance of B2B against four competing methods, all
-implemented in scikit-learn \citep{sklearn} and pyrcca \citep{bilenko2016pyrcca}:
+We compare the performance of B2B against four baseline methods.
 %
@@
 \subsection{Baseline models}
+
+All baseline method were based on the implementations of scikit-learn \citep{sklearn} and pyrcca
+\citep{bilenko2016pyrcca}. For pedagogical purposes, we briefly summarize them below.
 
@@
 
-B2B leads to unbiased (i.e. zeros-centered) scalar coefficients for non-causal
-features. In contrast, the Forward, Backward, CCA and PLS models lead to a
-loading vector $H_i$ per feature $i$ (or one vector $G^i$ for the backward
-model). To transform such vector into an estimated causal contribution $\hat E$,
-we take the sum of square coefficients:
+%
+B2B leads to \emph{scalar} coefficients for non-causal
+features. The diagonal of this matrix, $\hat E$, can thus be directly used
+as a causal contribution estimate. Note that is estimate is unbiased (i.e.
+zeros-centered) if the second regression $H$ is not regularized.
+
+In contrast, the Forward, Backward, CCA and PLS models lead to a
+loading \emph{vector} for each feature: each vector indicates how to it
+relates to each dimension of the observations (for forward or backward
+models) or each canonical projection of the observations
+To an estimated causal contribution, we must thus transform such vector into a
+scalar, by e.g. taking the sum of square coefficients:
 % \begin{equation}
@@
 % \end{equation}
+Note that in such B2B cases, the estimates are necessarily biased: i.e. positive.
 
-To estimate whether models accurately identify causal factors, we compute the
+Finally, to estimate whether each model accurately identifies causal factors independently of
+their potential biases, we compute the
 area-under-the-curve (AUC) across factors $AUC(E, \hat E)$.
@@
 %\hat{E}_{k-1}) / 2 \end{equation}
-The AUC allows evaluating the capacity of models at detecting the causal
-importance of factors when ground truth labels are available, as is the case in
-this setup.
+By definition, this AUC evaluation can only be done when ground truth labels are available, as is the case in
+this synthetic setup.
 
@@
   \vspace{-4ex}
-  \caption{Synthetic experiments.
-Average AUC (top) and Feature Importance $\Delta R$ (bottom) when varying experimental conditions individually.
-Higher is better.
-B2B compares favorably in all cases.
+  \caption{Synthetic experiments. Average AUC (top) and Feature Importance
+  $\Delta R$ (bottom) when varying experimental conditions individually.
+  Higher is better. B2B compares favorably in all cases.
   \label{fig:percondition}}
@@
 
-In most cases, $E$ is not known and AUC can thus not be estimated. To address this issue, we assess the ability of each model to reliably predict
+In most cases, $E$ is not known. The above AUC can thus not be estimated.
+To address this issue, we assess the ability of each model to reliably predict
 independent and identically distributed data from $Y$, given all of the $X$
-features versus all-but-ones feature $X_{-i}$ (i.e. 'knock-out X'). This procedure
-results in two correlation metrics $R_{full}$ and $R_{knockout}$, whose
-difference $\Delta R_i = R_{full}-R_{knockout}$ indicates how much each $X_i$
-improves the prediction of $Y$.
-%beyond what can be predicted from all other features $X_{-i}$.
-In our figures, $\Delta R$ is the average of $\Delta R_i$. A higher score means that for prediction, the model relies on individual features rather than combinations of features.
+features versus all-but-one feature $X_{-i}$ (i.e. 'knock-out X'). This
+procedure results in two correlation metrics $R_{full}$ and $R_{knockout}$ for
+each feature (for the B2B and Backward models) for each dimension of $Y$ (for
+the Forward model) or each canonical dimension of $Y$ (for CCA and PLS). The
+difference $\Delta R_i = R_{full}-R_{knockout}$ indicates how much each
+$X_i$ improves the prediction of (a pre-defined dimension or projection of) $Y$.
 
-We show in Appendix~\ref{appendix:feature_importance} pseudo-code to assess feature importance for our algorithm as well as baselines.
-For the Backward Model, feature importance cannot be assessed as the $X$ collinearity is never taken into account.
+We show in Appendix~\ref{appendix:feature_importance} pseudo-code to assess
+feature importance for each model. For the Backward Model, feature importance
+cannot be assessed as the $X$ collinearity is never
+taken into account.
 
-% Feature importance can be computed similarly for the CCA, PLS and Forward baselines (see Appendix \ref{appendix:feature_importance}).
-For the Backward Model, feature importance cannot be assessed as there is no prediction.
+In our synthetic figures, we do report each feature $i$ separately, as they are
+randomly generated. $\Delta R$ thus represents the average of $\Delta R_i$. A
+higher score means that for prediction, the model relies on individual features
+rather than combinations of features.
 
-We show in Figures~\ref{fig:percondition} (bottom) and \ref{fig:auc_plots} (right, in Appendix) that our method outperforms baselines.
-
-%
-% \iffalse
-% For CCA and PLS, feature importance is assessed as follow:
-% \begin{enumerate}
-%   \item Fit $H$ and $G$ given $X_{train}$ and $Y_{train}$
-%   \item For each feature $i$:
-%   \begin{itemize}
-%     \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
-%     \item Fit $H^i_k$ and $G_k$ given $X_{train} K$ and $Y_{train}$
-%   \end{itemize}
-% \end{enumerate}
-% \fi
-% For the Backward Model, feature importance cannot be assessed because there is no prediction.
-% We thus simply
-% report the model performance $R_i = corr(YG_i, X_i)$.
-
-% \iffalse
-% \begin{enumerate}
-%
-%   \item Fit $G$ given $X_{train}$ and $Y_{train}$
-%   \item Fit $H$ given $X_{train}$ and $Y_{train} G^T$
-%   \item For each feature $i$:
-%   \begin{itemize}
-%     \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
-%     \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train} G^T$
-%   \end{itemize}
-%
-%   \item Compute $R_{full}$ given $Y_{test} G^T$ and $X_{test}$
-%   \item For all $i \in d_x$:
-%   \begin{itemize}
-%     \item Compute $R^i_{knockout}$ given $Y_{test} G^T$ and $X_{test} K {H^i_k}^T$
-%     \item Compute $R^i_{delta}=R_{full} - R^i_{knockout}$
-%   \end{itemize}
-%   \end{enumerate}
-% \fi
-
-% \iffalse Ridge regression \citep{hoerl1959optimum}, multitask Lasso
-% \citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls,
-% tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and
-% Reduced Rank Regression or RRR \citep{Izenman_rrr}.
-% %
-% Each of these methods estimates a matrix of coefficients $\hat{W} \in
-% \mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$
-% using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
-% %
-% For B2B, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
-% %
-%
-%
-% We evaluate seven metrics:
-% %
-% test error in-distribution, test error out-distribution, test error-in
-% distribution after weighting with $\hat{E}$, test error out-distribution after
-% weighting with $\hat{E}$, Sonquist-Morgan false positives on $\hat{E}$,
-% Sonquist-Morgan false negatives on $\hat{E}$, and the area under the ROC curve
-% (AUC) between $\hat{E}$ and $E$.
-% %
-% The ``in-distribution'' metrics measure the test error of each predictor before
-% and after weighting their features by the estimated (continuous!) mask
-% $\hat{E}$.
-% %
-% The ``out-distribution'' metrics are similar, measuring the test error at
-% samples drawn using two different random covariance matrices for both $X$ and
-% $N$.
-% %
-% The false positives, false negatives, and AUC statistics measure the quality of
-% identification of causal factors between $\hat{E}$ and the true, binary $E$.
-%
-% \begin{figure}[htpb] \centering
-% \includegraphics[width=\textwidth]{synthetic.pdf} \caption{Results of synthetic
-% experiments. Each dot depicts the average value of a metric for B2B ($x$-axis)
-% against a competitor ($y$-axis) for each experiment configuration averaged over
-% $20$ random seeds. The prediction error on $Y$ within (in) and outside (out) of
-% the training distribution are weakly but consistently larger for all of the
-% models as compared to B2B with and without the Sonquist binarization (E). The
-% AUC, measuring the ability of each model to separate causal from noncausal
-% factors, is consistently larger in B2B as compared to the other models.}
-% \label{table:synthetic} \end{figure} \fi
-
-% Figure~\ref{table:synthetic} summarizes the results for all experimental
-% conditions, metrics, and methods.
-% %
-% B2B is the method obtaining best test errors, lowest false positive and false
-% negative discovery about $E$, and overall the best AUC when it comes to
-% detecting causal influences.
+Figures~\ref{fig:percondition} (bottom) and \ref{fig:auc_plots} (right, in
+Appendix) show that B2B outperforms baseline models.

### 2020-05-19 paper/synthetic.tex
synthetic

@@
 
-In our synthetic figures, we do report each feature $i$ separately, as they are
-randomly generated. $\Delta R$ thus represents the average of $\Delta R_i$. A
-higher score means that for prediction, the model relies on individual features
-rather than combinations of features.
+\subsection{Results of the synthetic experiment}
 
 Figures~\ref{fig:percondition} (bottom) and \ref{fig:auc_plots} (right, in
-Appendix) show that B2B outperforms baseline models.
+Appendix) show the results of our synthetic experiment. Note that the figure
+does not display each feature $i$ separately, as they are
+randomly generated. $\Delta R$ represents the average of $\Delta R_i$. A
+higher score means that for prediction, the model relies on individual features
+rather than combinations of features. Overall both the AUC and the held-out
+prediction reliability evaluations show that B2B outperforms baseline models.

### 2020-05-19 paper/synthetic.tex
synthetic

@@
 The purpose of these experiments is to evaluate the ability of B2B on its
-ability to 1) accurately predict independent and identically distributed data,
-and 2) recover causal factors when the ground truth is known.
+ability to 1) recover causal factors when the ground truth is known and 2)
+accurately predict independent and identically distributed data otherwise.
 
@@
 
-\subsection{Baseline models}
+\subsubsection{Baseline models}
 
-All baseline method were based on the implementations of scikit-learn \citep{sklearn} and pyrcca
+All baseline methods were based on the implementations of scikit-learn \citep{sklearn} and pyrcca
 \citep{bilenko2016pyrcca}. For pedagogical purposes, we briefly summarize them below.
@@
 
-\subsection{Evaluating Causal Discovery from models' coefficients}
+\subsubsection{Evaluating Causal Discovery from models' coefficients}
 
@@
 features. The diagonal of this matrix, $\hat E$, can thus be directly used
-as a causal contribution estimate. Note that is estimate is unbiased (i.e.
-zeros-centered) if the second regression $H$ is not regularized.
+as a causal contribution estimate. Note that this estimate is unbiased (i.e.
+zeros-centered) only if the second regression $H$ is not regularized.
 
 In contrast, the Forward, Backward, CCA and PLS models lead to a
-loading \emph{vector} for each feature: each vector indicates how to it
-relates to each dimension of the observations (for forward or backward
-models) or each canonical projection of the observations
-To an estimated causal contribution, we must thus transform such vector into a
+loading \emph{vector} for each feature: each vector indicates how to a feature
+relates to each dimension of the observations (for Forward or Backward
+models) or each canonical projection of the observations.
+To estimate causal contribution, we must thus transform such vector into a
 scalar, by e.g. taking the sum of square coefficients:
@@
 % \end{equation}
-Note that in such B2B cases, the estimates are necessarily biased: i.e. positive.
+Note that in such B2B cases, the estimates are  thus necessarily positive and thus biased.
 
@@
 By definition, this AUC evaluation can only be done when ground truth labels are available, as is the case in
-this synthetic setup.
-
-We report AUC results in Figures~\ref{fig:percondition}~(top) and ~\ref{fig:auc_plots}~(left, in Appendix), and compare favorably to all baselines.
+this synthetic setup, but not in the neuroimaging experiments below.
 
@@
 
-\subsection{Evaluating Causal Discovery with held-out prediction reliability}
+\subsubsection{Evaluating Causal Discovery with held-out prediction reliability}
 
-In most cases, $E$ is not known. The above AUC can thus not be estimated.
+In most real-world cases, $E$ is not known. Consequently, the above AUC evaluation can not be estimated.
 To address this issue, we assess the ability of each model to reliably predict
@@
 
-\subsection{Results of the synthetic experiment}
+\subsubsection{Results of the synthetic experiment}
 
@@
 Appendix) show the results of our synthetic experiment. Note that the figure
-does not display each feature $i$ separately, as they are
-randomly generated. $\Delta R$ represents the average of $\Delta R_i$. A
-higher score means that for prediction, the model relies on individual features
-rather than combinations of features. Overall both the AUC and the held-out
-prediction reliability evaluations show that B2B outperforms baseline models.
+does not display each feature separately, as they are
+randomly generated. $\Delta R$ represents the average of $\Delta R_i$ across
+features $i$.
+Overall both the AUC and the held-out prediction reliability evaluations
+show that B2B compares favorably to the baseline models.

### 2020-05-19 paper/fmri_exps.tex
up

@@
 Imaging (fMRI) and magneto-encephalography (MEG)
-recordings of each 102 healthy native-Dutch adults who participated in a reading
-task. Thirteen subjects were excluded from the analysis (12 MEG and 1 fMRI)
-because of missing files or corrupted headers.
+recordings of each 102 healthy Dutch adults who performed a reading
+task in the scanner. Thirteen subjects were excluded from the analysis (12 MEG and 1 fMRI)
+because of technical difficulties reading the files.
 %
@@
 Results included in this manuscript come from preprocessing performed
-using \emph{fMRIPrep} 20.0.7 (\citet{fmriprep1}; \citet{fmriprep2};
+using \emph{fMRIPrep} 20.0.7 (\citep{fmriprep1}; \citep{fmriprep2};
 RRID:SCR\_016216), which is based on \emph{Nipype} 1.4.2
-(\citet{nipype1}; \citet{nipype2}; RRID:SCR\_002502).
+(\citep{nipype1}; \citep{nipype2}; RRID:SCR\_002502).
 
@@
 \item[Anatomical data preprocessing]
-A total of 2 T1-weighted (T1w) images were found within the input BIDS
+A total of two T1-weighted (T1w) images per subject were found within the input BIDS
 dataset. All of them were corrected for intensity non-uniformity (INU)
@@
 selected for spatial normalization: \emph{ICBM 152 Nonlinear
-Asymmetrical template version 2009c} {[}\citet{mni152nlin2009casym},
+Asymmetrical template version 2009c} {[}\citep{mni152nlin2009casym},
 RRID:SCR\_008796; TemplateFlow ID: MNI152NLin2009cAsym{]}, \emph{FSL's
 MNI ICBM 152 non-linear 6th Generation Asymmetric Average Brain
-Stereotaxic Registration Model} {[}\citet{mni152nlin6asym},
+Stereotaxic Registration Model} {[}\citep{mni152nlin6asym},
 RRID:SCR\_002823; TemplateFlow ID: MNI152NLin6Asym{]},
 \item[Functional data preprocessing]
-For each of the 2 BOLD runs found per subject (across all tasks and
-sessions), the following preprocessing was performed. First, a reference
+For each of the 2 BOLD runs found per subject, the following preprocessing was performed. First, a reference
 volume and its skull-stripped version were generated using a custom
@@
 
-\subsubsection{fMRI analysis}
-
 The preprocessed volumetric fMRI data was linearly projected to its closest surface
-using nilearn $vol_to_surf$ function with a 3 mm radius. A surface search-light
+using nilearn \texttt{vol\_to\_surf} function with a 3 mm radius. A surface searchlight
 analysis was then implemented by concatenating the surface
@@
 thus build an observation matrix $Y \in \mathbb{R}^{n \times d_y}$ of $n\approx$ 1400 words
-by $d_y\approx40$ vertices. Each
+by $d_y\approx40$ vertices per searchlight sphere. Each
 of the columns of $Y$ is normalized to have zero mean and unit variance.
@@
 First, 'Word Length' refers to the total number of letters.
-Word Length is expected to specifically cause a variation in the early evoked MEG responses (i.e.
+Word Length is expected to primarily cause a variation in the early evoked MEG responses (i.e.
 from 100 ms after stimulus onset) elicited by the retinotopically-tuned visual cortices (e.g.
@@
 %
-Second, 'Word Frequency' indexes how frequently each word appears in Dutch and was derived with the the Zipf logarithmic scale of \citep{van2014subtlex} provided by the WordFreq package \citep{speerwordfreq}.
-Word Frequency is expected to specifically cause a variation in the late evoked MEG responses (i.e.
-from 400 ms), because it variably engages semantic processes in the temporal cortices \citep{kutas2011thirty}.
+Second, 'Word Frequency' indexes how frequently each word appears in Dutch and
+was derived with the the Zipf logarithmic scale of \citep{van2014subtlex}
+provided by the WordFreq package \citep{speerwordfreq}. Word Frequency is
+expected to primarily cause a variation in the late evoked MEG responses (i.e.
+from 400 ms) in the left frontal, temporal and parietal cortices
+cortices \citep{kutas2011thirty,mccandliss2003visual}.
 %
-Third, 'Word Function' indicates whether each word is a content word (i.e.
-a noun, a verb, an adjective or an adverb) or a function word (i.e.
-a preposition, a conjunction, a determinant, a pronoun or a numeral), and was derived from Spacy's part of speech tagger \citep{spacy2}.
-To our knowledge, this feature has not been thouroughly investigated with MEG.
-Its causal contribution to reading processes in the brain thus remains unclear.
+Third, 'Word Function' indicates whether each word is a content word (i.e. a
+noun, a verb, an adjective or an adverb) or a function word (i.e. a preposition,
+a conjunction, a determinant, a pronoun or a numeral), and was derived from
+Spacy's part of speech tagger \citep{spacy2}. To our knowledge, this feature has
+not been thoroughly investigated with fMRI and MEG. While its causal contribution to
+reading processes in the brain thus remains unclear, this lexical feature can
+nonetheless be expected to present similar brain patterns to word frequency.
 %
 Finally, to verify that B2B and other methods would not inadequately identify non-causal features, we added a dummy feature, constructed from a noisy combination of Word Length and Word Frequency:
-$dummy = z(length) + z(frequency) + \mathcal{N}$, where $z$ normalizes features and $\mathcal{N}$ is a random vector sampling Gaussian distribution (all terms thus have a zero-mean and a unit-variance).
-This procedure yields an $X \in \mathbb{R}^{n \times d_x}$ matrix of $n\approx$ 2700 words by
-$d_x=4$ features for each subject. Each of the columns of $X$ is normalized to
-have a mean and a standard deviation of 0 and 1 respectively.
+$dummy = z(length) + z(frequency) + \mathcal{N}$, where $z$ normalizes features
+and $\mathcal{N}$ is a random vector sampling Gaussian distribution (all terms
+thus have a zero-mean and a unit-variance).
 
 To account for the delay of blood oxygenation level dependent responses, the
-features were convolved using the default hemodynamic response function Glover
-of nilearn \citep{nilearn}.
+features were convolved using the Glover hemodynamic response function
+of nilearn's \texttt{compute\_regressor} function an oversampling of 16 and default parameters \citep{nilearn}.
+
+This procedure yields an $X \in \mathbb{R}^{n \times d_x}$ matrix of $n\approx$
+840 TR (Repetition Time: 2 sec.) by
+$d_x=4$ features for each subject. Each of the columns of $X$ is normalized to
+have a zero-mean and a unit variance.
 
@@
           \includegraphics[width=.8\textwidth,
-                       trim=1cm 1cm 2cm 2cm,
+                       trim=1cm 1cm 1cm 2cm,
                        clip=True]{figures/fmri_delta_r.png}
@@
 
-\subsubsection{Results}
+\subsubsection{fMRI Results}
 We compared the ability of Forward regression, Backward regression, CCA, PLS
@@
 Length and Word Frequency across large sections of the cortex, even though
-these features are known to specifically influence early visual cortex and
-associative cortices respectively. These results illustrate that backward
-modeling cannot be used to estimate the causal contribution of collinear
-features.
+these features are known to primarily influence early visual cortex and
+associative cortices respectively. These results illustrate that while Backward
+modeling achieve high sensitivity, it is not valid to estimate the specific
+causal contribution of collinear features.
 
-On the contrary, the $\beta$ coefficients of the Forward model reveals the
-expected spatial specificity of Word Length and Word Frequency. However, these
-coefficients are moderately consistent across subjects, and can thus
-dramatically limit the possibility to achieve high levels of
-significance, as evaluated with a Wilcoxon tests across subjects at each vertex.
-
-In principle, the inter-subject variability could be less impactful with
-multivariate observations methods such as CCA and PLS. However, these
-cross-decompositions do not provide a single and interpretable coefficient
+On the contrary, the $\beta$
+coefficients of the Forward model reveals the expected spatial specificity of
+Word Length and Word Frequency, peaking the visual and temporal cortices
+respectively. However such Forward coefficients may be variable across
+subjects, and could thus underestimate the significance of each factor at each
+vertex, as evaluated with a Wilcoxon tests across subjects. In principle, such
+inter-subject variability could be less impactful with
+multivariate observations methods such as CCA and PLS \citep{bilenko2016pyrcca,
+king2018encoding}, but they fail to provide a single and interpretable coefficient
 for each feature at each vertex.
 
-On the contrary, the $\hat E$ coefficients of an unbiased B2B (i.e. a B2B where
-$H$ is not regularized) derive from multivariate fMRI observations but are
-unidimensional for each feature. In fact, $\hat E$ reaches higher levels of
-significance, as evaluated with a Wilcoxon test across subjects at each vertex
-than the Forward $\beta$ \ref{fig:fmri_supp}. Note that a direct
+In contrast, the $\hat E$ of an unbiased B2B (i.e. a B2B where $H$ is not
+regularized) reaches higher levels of significance at each vertex
+than the Forward $\beta$ \ref{fig:fmri_supp}. (Note that a direct
 comparison between the Forward coefficients $\beta$ and those of B2B ($\hat E$)
-is not feasible, as these two do not have the same units.
+is not feasible, as these two do not have the same units.)
 
-To fairly compare the Forward, CCA, PLS and B2B methods on a common metric,
-without being over-dependent on the
-potentially invalid assumptions of each model, we compare them on their ability
-to improve Y
+To fairly compare the Forward, CCA, PLS and B2B methods on a common and robust
+evaluation metric we compare them on their ability to improve $Y$
 predictions of out-of-sample data when one factor is introduced into the model.
@@
 and B2B method predicted that the Dummy Variable does not improve the $Y$
-prediction.
+prediction. This confirms that these methods accurately rule out the known
+non-causal factor.
 
-The average $\Delta R$ across subjects is displayed for each method and each
-feature in \ref{fig:fmri_r}, and suggest that B2B surpasses other methods.
-To formally quantify these comparisons, we compare, for each subject separately,
-the average $\Delta R$ across vertices. To avoid including irrelevant brain
-regions, we restrict this analysis to vertices which could be reliably accounted
+The average $\Delta R$ across subjects is displayed for each model and each
+feature in \ref{fig:fmri_delta_r}. Overall, B2B favorably compares to baseline models.
+To quantify this assessment, we compare for each subject separately
+the average $\Delta R$ across vertices obtained between B2B and each
+baseline model. To limit the inclusion of uninformative brain regions in this
+summary, we restrict the analysis to vertices which can be reliably accounted
 for by the Forward model ($p<.001$, not corrected across vertices)
-\ref{fig:fmri_r}. The results show that B2B significantly outperforms all
-other methods on all features (all $p<0.0001$) but the dummy variable
+\ref{fig:fmri_delta_r}. The results show that B2B significantly outperforms all
+baseline models on all factors (all $p<0.0001$) but the dummy variable
 (min $p=0.0852$).

### 2020-05-19 paper/meg_exps.tex
up

@@
 addition, the same decoding time course was observed for the dummy variable.
-These results illustrate that backward modeling cannot be used to estimate the
+Once again, these results illustrate that Backward modeling cannot be used to estimate the
 causal contribution of correlated features.
@@
 PLS~($p=0.7989$). Overall, these results confirm that Word Length, Word
-Frequency and Word Function causally influence specific periods of brain
-responses to words.
+Frequency and Word Function specifically improve the prediction of specific periods of brain
+responses to words, and thus form plausible independent causal contributors.
 
@@
 compared B2B to other models across subjects (Figure \ref{fig:meg_results}
-right). For Word Length B2B outperforms all models (all $p < 0.00001$) but CCA
+right). For Word Length, B2B outperforms all models (all $p < 0.00001$) but CCA
 ($p=0.0678$). For Word Frequency, B2B outperforms all models (all $p < 0.0006$).
 For "Word Function", B2B outperforms all models (all $p < 0.0015$). Overall,
-these results show that B2B reliably outperforms standard methods, especially
-when the effects are difficult to detect.
+these results show that B2B favorably compares against baseline models.

### 2020-05-19 paper/paper.tex
up

@@
 
+\subsection{Problem setup}
+
 We consider the measurement of multivariate signal $Y \in \mathbb{R}^{n \times
@@
 Given the model in Equation~\cref{eq:model}, the goal of Back-to-Back
-Regression (B2B) is to estimate the matrix of $E$, i.e. to identify the factors
+Regression (B2B) is to estimate the matrix $E$, i.e. to identify the factors
 that most reliably account for the multivariate observations.
@@
 %
-Then, the first regression is performed using $(X_1, Y_1)$, and the second
+The first regression is performed using $(X_1, Y_1)$, and the second
 regression is performed using $(X_2, Y_2)$.
 %
-To compensate for the reduction in sample size caused by the split, B2B is
+To compensate for the reduction in sample size caused by the split, the two
+successive regressions are
 repeated over many random splits, and the final estimate $\hat E$ of the causal
@@
 %
-To accelerate this ensembling procedure, we implemented an efficient
+To accelerate this ensembling procedure, we use an efficient
 leave-one-out cross-validation scheme as detailed in \citep{rifkin2007notes}
@@
 elements of $\text{diag}(\hat H)$. The methods to recover them are presented in
-the Appendix \ref{recovering}.
+the Appendix (\ref{recovering}).
 
@@
 
-\subsection{Magnetoencephalography}
+\subsection{Magneto-encephalography}
 \label{sec:experiment_real}
@@
 
-Forward and cross-decomposition models have been used to identify the causal
-contribution of correlated factors onto multi-dimensional observations (e.g.
-\citep{naselaris2011encoding}). These approaches typically lead to multiple
-coefficients for each features (i.e. one per dimension of $Y$ or one per
-component respectively). Furthermore, these coefficients can be difficult to
-summarize into a single causal estimate. By contrast, B2B quickly (Fig.
-\ref{fig:duration}) leads to a single unbiased scalar values $\hat E$ tending
-towards 1 and 0 for causal and non-causal features respectively.
+% Forward and cross-decomposition models have been used to identify the causal
+% contribution of correlated factors onto multi-dimensional observations (e.g.
+% \citep{naselaris2011encoding}). These approaches typically lead to multiple
+% coefficients for each features (i.e. one per dimension of $Y$ or one per
+% component respectively). Furthermore, these coefficients can be difficult to
+% summarize into a single causal estimate. By contrast, B2B quickly (Fig.
+% \ref{fig:duration}) leads to a single unbiased scalar values $\hat E$ tending
+% towards 1 and 0 for causal and non-causal features respectively.
 
-A variety of other statistical methods applied to neuroimaging data have been
+A variety of statistical methods applied to neuroimaging have been
 proposed to clarify what is being represented in brain responses - i.e. what
-feature causes specific brain activity. One of the popular linear method is
-Representational Similarity Analysis (RSA)
-\citep{kriegeskorte2008representational}, and consists in analyzing the
-similarity of brain responses associated with specific categorical conditions
-(e.g. distinct images), by (1) fitting one-against-all classifiers on each
-condition and (2) testing whether these classifiers can discriminate all other
-conditions. The resulting confusion matrix is then analyzed in an unsupervised
-manner to reveal which conditions lead to similar brain activity patterns. B2B
-differs from RSA in that (1) it uses regressions instead of classifications, and
-can thus generalize to new items and new contexts and (2) it is fully
-supervised.
-
-% Here, we focus on the restricted issue of disentangling the
-% influence of linearly correlated predictors ($X$) onto noisy multivariate
-% observations ($Y$). The present approach thus differs from other causal
-% discovery algorithms based on temporal-delays and/or nonlinear interactions in
-% systems where the directionality of causation (from X to Y or vice versa) is
-% unknown (e.g. \citep{peters2017elements, granger1969investigating,
-% janzing2013quantifying, scholkopf2016modeling}.
-
-Finally, CCA has been used in neuroimaging for a variety of purposes such as
+feature causes specific brain activity. For example, CCA and PLS have been used in
+neuroimaging for a variety of purposes such as
 denoising and subject alignment \citep{cca_hotelling, de2019multiway}. While CCA
-relates to B2B, these two methods diverge in several ways. First, CCA and B2B
-have different objectives: CCA aims to find the potentially numerous and poorly
-interpretable components where X and Y are maximally correlated, whereas B2B
+and PLS relates to B2B, these methods diverge in several ways. First, they have different objectives: CCA aims to find the potentially numerous and poorly
+interpretable components where $X$ and $Y$ are maximally correlated, whereas B2B
 aims to recover the causal factors from X to Y. Second, B2B is not symmetric
 between $X$ and $Y$: it aims to identify specific causal features by first
-optimizing over the decoders $G$ and then over $H$. By contrast, CCA is
+optimizing over the decoders $G$ and then over $H$. By contrast, CCA and PLS are
 symmetric between $X$ and $Y$, and aims to find $G$ and $H$ such that they
@@
 contrast B2B assesses each feature $X^j$ on a single $Y$ component specifically
-selected to maximize signal-to-noise ratio of that feature $j$. Fourth, and
-unlike B2B, CCA does not separately optimize two distinct regularization
-parameters for $G$ and $H$. Finally, CCA does not use different data splits to
+selected to maximize signal-to-noise ratio of that feature $j$. Fourth CCA does
+not separately optimize two distinct regularization
+parameters for $G$ and $H$, where B2B does. Finally, CCA does not use different data splits to
 estimate $G$ and $H$. Together, these differences may explain why B2B reliably
@@
 
-% I find a bit surprising to mention causality literature in ML at the end. I
-% would have expected this in the introduction to convince reviewer that know
-% your stuff.
+One popular multivariate analysis of brain activity is
+Representational Similarity Analysis (RSA)
+\citep{kriegeskorte2008representational}, and consists in analyzing the
+similarity of brain responses associated with specific categorical conditions
+(e.g. distinct images), by (1) fitting one-against-all classifiers on each
+condition and (2) testing whether these classifiers can discriminate all other
+conditions. The resulting confusion matrix is then analyzed, generally in an unsupervised
+manner, to reveal which conditions lead to similar brain activity patterns. B2B
+differs from RSA in that it uses regressions instead of one-hot classifications.
+Consequently, and unlike RSA, B2B (1) is fully supervised (2) provides
+interpretable coefficients and (3) can generalize to new items and new contexts.
+In practice, these elements allow B2B to apply to event-related paradigms and
+latent variable analyses.
 
-\section{Conclusion}
-In this work, we proposed Back-to-Back (B2B) regression, a linear method to
-disentangle confounded factors from multidimensional observations.
-%
-B2B repeatedly performs two successive multidimensional regressions on
-independent subsets of the data: the first regression is applied on the output
-domain (as in backward decoding), whereas the second regression is applied on
-the input domain (as in forward encoding).
-%
-We provided a theoretical guarantee about the consistency of B2B, and compared
-it to several baselines in controlled synthetic experiments.
-%
-We also applied B2B to a recent brain imaging dataset, analyzing the timing of
-brain responses and their connection to word features.
-%
-We obtained results consistent with prior work in neuroscience literature,
-confirming the reliability of B2B for real data analysis.
-%
+The present empirical results, together with their theoretical foundations,
+suggest that B2B may come as a tool of choice for causal discovery when
+feature-wise intervention is limited.
 

### 2020-05-19 paper/appendix.tex
E -> S

--- 2020-03-04 paper/appendix.tex
@@
 
-     Consider the B2B model from equation \ref{eq:model} $$Y = (XE + N)F$$ with
+     Consider the B2B model from equation \ref{eq:model} $$Y = (XS + N)F$$ with
      $N$ centered and full rank noise.
 
-     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat
+     If $F$ and $X$ are full-rank on $Img(S)$, then, the solution of B2B, $\hat
      H$ minimizes
@@
 
-     $$E\hat H = \hat H$$
+     $$S\hat H = \hat H$$
 \end{theorem}
@@
  \begin{align*}
-    \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2]
+    \hat G = \arg \min_G \mathbb{S}[\left \| YG - X \right \|^2]
 \end{align*}
-Replacing $Y$ by its model definition $Y = (XE+N)F$, we have
+Replacing $Y$ by its model definition $Y = (XS+N)F$, we have
  \begin{align*}
-    \hat G &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2] =\arg \min_G \mathbb{E}[\left \| X - XEFG + NFG \right\|^2]
+    \hat G &=   \arg \min_G \mathbb{S}[\left \| X - (XS + N)FG \right\|^2] =\arg \min_G \mathbb{S}[\left \| X - XSFG + NFG \right\|^2]
   \end{align*}
@@
   \begin{align}
-    	  \hat G &=  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
+    	  \hat G &=  \arg \min_G \left \| X - XSFG\right\| ^2  + \left \| NFG\right \| ^2
      \label{eq:Gdoublenorm}
@@
 \begin{align*}
-    \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
-    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
-    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]
+    \hat H = \arg \min_H \mathbb{S}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{S}[\| XH - (XS + N)F \hat G \|^2] \\
+    &=\arg \min_H \mathbb{S}[\| X(H - SF \hat G) \| ^2] + \mathbb{S}[\| NF\hat G \| ^2]\\
+    &= \arg \min_H \mathbb{S}[\| X(H - SF \hat G) \| ^2]
  \end{align*}
@@
  \begin{align}
-    \hat H = EF \hat G
+    \hat H = SF \hat G
     \label{eq:Hdoublenom}
@@
 
-Let us now prove that $EF\hat G = F\hat G$.
+Let us now prove that $SF\hat G = F\hat G$.
 
-Let $F^\dagger$ be the pseudo inverse of $F$, and $Z=F^\dagger EF\hat G$, we
-have $FZ = FF^\dagger EF \hat G$
+Let $F^\dagger$ be the pseudo inverse of $F$, and $Z=F^\dagger SF\hat G$, we
+have $FZ = FF^\dagger SF \hat G$
 
-Since $F$ is full rank on $Img(E)$, we have $FF^\dagger E =E$, and $FZ = EF\hat
+Since $F$ is full rank on $Img(S)$, we have $FF^\dagger S =S$, and $FZ = SF\hat
 G$
 
-As $E$ is a binary diagonal matrix, it is an orthogonal projection and therefore
+As $S$ is a binary diagonal matrix, it is an orthogonal projection and therefore
 a contraction, thus
- $$ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$$ and
- $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
+ $$ \| NSF\hat G\|^2 \leq \| NF\hat G \|^2$$ and
+ $$\left \| X - XSFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XSF\hat G \| ^2  + \| NSF\hat G \| ^2 \leq \| X - XSF\hat G \| ^2  + \| NF\hat G \| ^2$$
 
-But since $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, we also have
-$$\left \| X - XEF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 \leq \left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2$$
+But since $\hat G =  \arg \min_G \left \| X - XSFG\right\| ^2  + \left \| NFG\right \| ^2$, we also have
+$$\left \| X - XSF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 \leq \left \| X - XSFZ\right \| ^2  + \left \| NFZ\right \| ^2$$
 
 Summarizing the above,
-$$\left \| X - XEF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
-$$\left \| X - XEF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2$$
-$$\left \| NF\hat G\right \| ^2 =  \| NEF\hat G \| ^2$$
+$$\left \| X - XSF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 \leq \| X - XSF\hat G \| ^2  + \| NSF\hat G \| ^2 \leq \| X - XSF\hat G \| ^2  + \| NF\hat G \| ^2$$
+$$\left \| X - XSF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 = \| X - XSF\hat G \| ^2  + \| NSF\hat G \| ^2$$
+$$\left \| NF\hat G\right \| ^2 =  \| NSF\hat G \| ^2$$
 
-$N$ being full rank, this yields $EF\hat G = F\Hat G$.
+$N$ being full rank, this yields $SF\hat G = F\Hat G$.
 
-Replacing into $\eqref{eq:Gdoublenorm}$, and setting $H = EFG$, we have
+Replacing into $\eqref{eq:Gdoublenorm}$, and setting $H = SFG$, we have
 \begin{align*}
-	\hat G &=  \arg \min_G  \left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 \\
-	&=   \arg \min_G \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 \\
+	\hat G &=  \arg \min_G  \left \| X - XSFG\right \| ^2  + \left \| NFG\right \| ^2 \\
+	&=   \arg \min_G \left \| X - XSFG\right \| ^2  + \left \| NSFG\right \| ^2 \\
 	\hat H &=  \arg \min_H \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2
@@
 
-Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$, since $E$, a binary
+Finally, $S\hat H = S SF\hat G = SF\hat G = \hat H$, since $S$, a binary
 diagonal matrix, is involutive. This completes the proof.
@@
 \begin{equation}
-  Y = (XE+N)F+M
+  Y = (XS+N)F+M
   \label{eq:model_noisy_measure}
@@
 hypotheses. Indeed, we can rewrite $M = MF^{-1}F$ over $Img(F)$, which leads to:
-$$Y = (XE+N)F+M = (XE+N+MF^{-1})F = (XE+N')F$$
+$$Y = (XS+N)F+M = (XS+N+MF^{-1})F = (XS+N')F$$
 
-Consequently, assuming that $F$ is full rank on $Img(XE)$, B2B yields the same
+Consequently, assuming that $F$ is full rank on $Img(XS)$, B2B yields the same
 solutions to equations \ref{eq:model} and \ref{eq:model_noisy_measure}.
@@
 
-\subsection{Recovering E}
+\subsection{Recovering S}
 \label{recovering}
 
-In case of noise, B2B yields non binary $\hat E$. Three thresholding rules can
+In case of noise, B2B yields non binary $\hat S$. Three thresholding rules can
 be used to binarize its values thus explicitly recover "causal" features.
@@
 $\ref{eq:diagk}$ implies that the $k$ first diagonal elements of $\hat H$ are
-bounded:  % we assume that $F$ has full rank over $Img(E)$.
+bounded:  % we assume that $F$ has full rank over $Img(S)$.
 
@@
 
+\begin{figure}
+  \begin{center}
+    \includegraphics[width=0.7\textwidth,
+                     trim=0cm 0cm 0cm 0cm,
+                     clip=True]{figures/fmri_decod.pdf}
+    \includegraphics[width=0.7\textwidth,
+                     trim=0cm 0cm 0cm 0cm,
+                     clip=True]{figures/fmri_betas.pdf}
+
+    \label{fig:fmri_supp}
+  \end{center}
+  \caption{Top. P-values across subjects derived from a Backward decoding of
+  each feature (columns). Bottom. P-values across subjects derived from
+  the $\beta$ coefficients of a Forward Model and the $\hat S$ of an unbiased
+  B2B (i.e. B2B where $H$ is not regularized). Each cell displays the left
+  hemisphere view laterally (top) or ventrally (bottom)}
+\end{figure}
+
 \subsection{Robustness to increasing number of factors}

### 2020-05-19 paper/fmri_exps.tex
E -> S

@@
 
-We aim to identify the word features that cause a variation in brain responses. We consider four distinct but colinear features.
+We aim to identify the word features that cause a variation in brain responses. We consider four distinct but collinear features.
 %
@@
 
-In contrast, the $\hat E$ of an unbiased B2B (i.e. a B2B where $H$ is not
+In contrast, the $\hat S$ of an unbiased B2B (i.e. a B2B where $H$ is not
 regularized) reaches higher levels of significance at each vertex
 than the Forward $\beta$ \ref{fig:fmri_supp}. (Note that a direct
-comparison between the Forward coefficients $\beta$ and those of B2B ($\hat E$)
+comparison between the Forward coefficients $\beta$ and those of B2B ($\hat S$)
 is not feasible, as these two do not have the same units.)

### 2020-05-19 paper/paper.tex
E -> S

@@
   \newcommand\posX{3}
-  \newcommand\posE{5}
+  \newcommand\posS{5}
   \newcommand\posN{7}
@@
   \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=yellow!10] (X) at (\posX, 0){};
-  \node[thick, draw=black, minimum height=1cm, minimum width=1cm, fill=red!10] (E) at (\posE, 0){};
+  \node[thick, draw=black, minimum height=1cm, minimum width=1cm, fill=red!10] (S) at (\posS, 0){};
 
-  \draw[fill=white] (\posE - 0.5 + 0.0, 0.5 - 0.0) rectangle (\posE - 0.5 + 0.0 + 0.1, 0.5 - 0.0 - 0.1);
-  \draw[fill=black] (\posE - 0.5 + 0.1, 0.5 - 0.1) rectangle (\posE - 0.5 + 0.1 + 0.1, 0.5 - 0.1 - 0.1);
-  \draw[fill=white] (\posE - 0.5 + 0.2, 0.5 - 0.2) rectangle (\posE - 0.5 + 0.2 + 0.1, 0.5 - 0.2 - 0.1);
-  \draw[fill=white] (\posE - 0.5 + 0.3, 0.5 - 0.3) rectangle (\posE - 0.5 + 0.3 + 0.1, 0.5 - 0.3 - 0.1);
-  \draw[fill=white] (\posE - 0.5 + 0.4, 0.5 - 0.4) rectangle (\posE - 0.5 + 0.4 + 0.1, 0.5 - 0.4 - 0.1);
-  \draw[fill=white] (\posE - 0.5 + 0.5, 0.5 - 0.5) rectangle (\posE - 0.5 + 0.5 + 0.1, 0.5 - 0.5 - 0.1);
-  \draw[fill=white] (\posE - 0.5 + 0.6, 0.5 - 0.6) rectangle (\posE - 0.5 + 0.6 + 0.1, 0.5 - 0.6 - 0.1);
-  \draw[fill=black] (\posE - 0.5 + 0.7, 0.5 - 0.7) rectangle (\posE - 0.5 + 0.7 + 0.1, 0.5 - 0.7 - 0.1);
-  \draw[fill=white] (\posE - 0.5 + 0.8, 0.5 - 0.8) rectangle (\posE - 0.5 + 0.8 + 0.1, 0.5 - 0.8 - 0.1);
-  \draw[fill=black] (\posE - 0.5 + 0.9, 0.5 - 0.9) rectangle (\posE - 0.5 + 0.9 + 0.1, 0.5 - 0.9 - 0.1);
+  \draw[fill=white] (\posS - 0.5 + 0.0, 0.5 - 0.0) rectangle (\posS - 0.5 + 0.0 + 0.1, 0.5 - 0.0 - 0.1);
+  \draw[fill=black] (\posS - 0.5 + 0.1, 0.5 - 0.1) rectangle (\posS - 0.5 + 0.1 + 0.1, 0.5 - 0.1 - 0.1);
+  \draw[fill=white] (\posS - 0.5 + 0.2, 0.5 - 0.2) rectangle (\posS - 0.5 + 0.2 + 0.1, 0.5 - 0.2 - 0.1);
+  \draw[fill=white] (\posS - 0.5 + 0.3, 0.5 - 0.3) rectangle (\posS - 0.5 + 0.3 + 0.1, 0.5 - 0.3 - 0.1);
+  \draw[fill=white] (\posS - 0.5 + 0.4, 0.5 - 0.4) rectangle (\posS - 0.5 + 0.4 + 0.1, 0.5 - 0.4 - 0.1);
+  \draw[fill=white] (\posS - 0.5 + 0.5, 0.5 - 0.5) rectangle (\posS - 0.5 + 0.5 + 0.1, 0.5 - 0.5 - 0.1);
+  \draw[fill=white] (\posS - 0.5 + 0.6, 0.5 - 0.6) rectangle (\posS - 0.5 + 0.6 + 0.1, 0.5 - 0.6 - 0.1);
+  \draw[fill=black] (\posS - 0.5 + 0.7, 0.5 - 0.7) rectangle (\posS - 0.5 + 0.7 + 0.1, 0.5 - 0.7 - 0.1);
+  \draw[fill=white] (\posS - 0.5 + 0.8, 0.5 - 0.8) rectangle (\posS - 0.5 + 0.8 + 0.1, 0.5 - 0.8 - 0.1);
+  \draw[fill=black] (\posS - 0.5 + 0.9, 0.5 - 0.9) rectangle (\posS - 0.5 + 0.9 + 0.1, 0.5 - 0.9 - 0.1);
 
@@
   \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{n \times d_x}$}};
-  \node[] (annE) at (\posE, -1.8){\scalebox{0.85}{$E \in \mathbb{D}^{d_x \times d_x}$}};
+  \node[] (annS) at (\posS, -1.8){\scalebox{0.85}{$S \in \mathbb{D}^{d_x \times d_x}$}};
   \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{n \times d_x}$}};
@@
   \node[] (labX) at (\posX, 2){Factors};
-  \node[] (labE) at (\posE, 2){Cause};
-  \node[] (labE) at (\posE, 1.7){selection};
+  \node[] (labS) at (\posS, 2){Cause};
+  \node[] (labS) at (\posS, 1.7){selection};
   \node[] (labN) at (\posN, 2){Noise};
@@
   \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
-  \node[] (reg1) at (6,-3.25) {$\hat{E} = \text{diag}(\underbrace{(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}})$};
+  \node[] (reg1) at (6,-3.25) {$\hat{S} = \text{diag}(\underbrace{(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}})$};
   \end{tikzpicture}
-  \caption{Back-to-back regression identifies the subset of factors $E_{ii} = 1$ in $X$ that influence some observations $Y$ by 1) regressing from $Y$ to $X$ to obtain $\hat{X}$, and 2) returning the diagonal of the regression coefficients from $X$ to $\hat{X}$.}
+  \caption{Back-to-back regression identifies the subset of factors $S_{ii} = 1$ in $X$ that influence some observations $Y$ by 1) regressing from $Y$ to $X$ to obtain $\hat{X}$, and 2) returning the diagonal of the regression coefficients from $X$ to $\hat{X}$.}
 \label{fig:b2b}
@@
 %
-By considering a square binary diagonal matrix of \emph{causal influences} $E
-\in \mathbb{D}^{d_x \times d_x}$, we denote by $XE$ the causal factors of $Y$.
+By considering a square binary diagonal matrix of \emph{causal influences} $S
+\in \mathbb{D}^{d_x \times d_x}$, we denote by $XS$ the causal factors of $Y$.
 %
@@
 \begin{equation}
-    y_i = (x_i E + n_i) F
+    y_i = (x_i S + n_i) F
     \label{eq:model}
@@
 Given the model in Equation~\cref{eq:model}, the goal of Back-to-Back
-Regression (B2B) is to estimate the matrix $E$, i.e. to identify the factors
+Regression (B2B) is to estimate the matrix $S$, i.e. to identify the factors
 that most reliably account for the multivariate observations.
@@
 %
-The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} =
+The diagonal of the regression coefficients $\hat H$, denoted by $\hat{S} =
 \text{diag}(\hat{H})$, is the desired estimate of the causal influence matrix
-$E$, as detailed in the \ref{appendix:theorem_proof}.
+$S$, as detailed in the \ref{appendix:theorem_proof}.
 
@@
 successive regressions are
-repeated over many random splits, and the final estimate $\hat E$ of the causal
+repeated over many random splits, and the final estimate $\hat S$ of the causal
 influence matrix is the average over the estimates associated to each split
@@
 %
-% After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$.
+% After obtaining $\hat{S}$, we can fit a final regression from $X \hat{S}$ to $Y$.
 
@@
     \KwIn{input data $X \in \mathbb{R}^{n \times d_x}$, output data $Y \in \mathbb{R}^{n\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
-    \KwOut{estimate of causal influences $\hat{E} \in \mathbb{D}^{d_x \times d_x}$.}
-    $\hat{E} \leftarrow 0$\;
+    \KwOut{estimate of causal influences $\hat{S} \in \mathbb{D}^{d_x \times d_x}$.}
+    $\hat{S} \leftarrow 0$\;
     \For{$i = 1, \ldots, m$}{
@@
         $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$ \Comment*[r]{$\hat H=(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2 \hat G$}
-        $\hat{E} \leftarrow \hat{E} + \text{diag}(\hat{H})$\;
+        $\hat{S} \leftarrow \hat{S} + \text{diag}(\hat{H})$\;
     }
-    $\hat{E} \leftarrow \hat{E} / m$\;
-    $\hat{W} \leftarrow \text{LinearRegression}(X \hat{E}, Y)$\;
-    \Return{$\hat{E}$, $\hat{W}$}
+    $\hat{S} \leftarrow \hat{S} / m$\;
+    $\hat{W} \leftarrow \text{LinearRegression}(X \hat{S}, Y)$\;
+    \Return{$\hat{S}$, $\hat{W}$}
     \caption{Back-to-back regression.}
@@
 
-     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centered and full
+     Consider the B2B model from Equation $Y = (XS + N)F$, $N$ centered and full
      rank noise.
@@
      Let $Img(M)$ refers to the image of the matrix $M$. If $F$ and $X$ are
-     full-rank on the $Img(E)$, then, the solution of B2B, $\hat H$, will
+     full-rank on the $Img(S)$, then, the solution of B2B, $\hat H$, will
      minimize
      %
-     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $E\hat H = \hat H$
+     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $S\hat H = \hat H$
 \end{theorem}
@@
 
-Since  $E\hat H = \hat H$, we have
+Since  $S\hat H = \hat H$, we have
 \begin{equation}
-  \hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top.
+  \hat H = \arg \min_H  \left \| X - XSH\right\| ^2  + \left \| NSH\right \| ^2 = (S X^\top XS +SN^\top NS) ^\dagger SXX^\top.
 \end{equation}
-% So, $\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top$.
+% So, $\hat H = \arg \min_H  \left \| X - XSH\right\| ^2  + \left \| NSH\right \| ^2 = (S X^\top XS +SN^\top NS) ^\dagger SXX^\top$.
 
-Assuming, without loss of generality, that the active features in $E$ are the $k
+Assuming, without loss of generality, that the active features in $S$ are the $k
 \in \mathbb{Z}: k \in [0, d_x]$ first features, and rewriting $X=(X_1,X_2)$ and
@@
 In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so
-$\text{diag}_k(\hat H)=I$, and $$\text{diag}(\hat H) = \text{diag}(E)$$
-Therefore, we recover $E$ from $\hat H$.
+$\text{diag}_k(\hat H)=I$, and $$\text{diag}(\hat H) = \text{diag}(S)$$
+Therefore, we recover $S$ from $\hat H$.
 
-In the presence of noise, the causal factors of $E$ correspond to the positive
+In the presence of noise, the causal factors of $S$ correspond to the positive
 elements of $\text{diag}(\hat H)$. The methods to recover them are presented in
@@
 % summarize into a single causal estimate. By contrast, B2B quickly (Fig.
-% \ref{fig:duration}) leads to a single unbiased scalar values $\hat E$ tending
+% \ref{fig:duration}) leads to a single unbiased scalar values $\hat S$ tending
 % towards 1 and 0 for causal and non-causal features respectively.

### 2020-05-19 paper/synthetic.tex
E -> S

@@
 The data generating process for each experiment constructs $n=1000$ training examples
-according to the model $Y = (\text{h} XE + N)F$, where $\text{h}$ is a
+according to the model $Y = (\text{h} XS + N)F$, where $\text{h}$ is a
 scalar that modulates the signal-to-noise ratio.
@@
 %contains rows drawn from $\mathcal{N}(0, \Sigma_X)$, \item $N \in \mathbb{R}^{n
-%\times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, \item $E \in
+%\times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, \item $S \in
 %\mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
@@
 drawn from $\mathcal{N}(0, \Sigma_X)$, $N \in \mathbb{R}^{n \times d_x}$
-contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, $E \in \mathbb{R}^{d_x
+contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, $S \in \mathbb{R}^{d_x
 \times d_x}$ is a binary diagonal matrix containing $n_c$ ones, $\Sigma_X =
@@
 B2B leads to \emph{scalar} coefficients for non-causal
-features. The diagonal of this matrix, $\hat E$, can thus be directly used
+features. The diagonal of this matrix, $\hat S$, can thus be directly used
 as a causal contribution estimate. Note that this estimate is unbiased (i.e.
@@
 % \begin{equation}
-  $\hat E_i = \sum_j {H^j_i}^2 $
+  $\hat S_i = \sum_j {H^j_i}^2 $
 % \end{equation}
@@
 their potential biases, we compute the
-area-under-the-curve (AUC) across factors $AUC(E, \hat E)$.
-%\begin{equation} AUC = 1 - \sum_1^n (E_k - E_{k-1}) ( \hat{E}_k +
-%\hat{E}_{k-1}) / 2 \end{equation}
+area-under-the-curve (AUC) across factors $AUC(S, \hat S)$.
+%\begin{equation} AUC = 1 - \sum_1^n (S_k - S_{k-1}) ( \hat{S}_k +
+%\hat{S}_{k-1}) / 2 \end{equation}
 By definition, this AUC evaluation can only be done when ground truth labels are available, as is the case in
@@
 
-In most real-world cases, $E$ is not known. Consequently, the above AUC evaluation can not be estimated.
+In most real-world cases, $S$ is not known. Consequently, the above AUC evaluation can not be estimated.
 To address this issue, we assess the ability of each model to reliably predict

### 2020-05-19 paper/appendix.tex
n -> m

@@
 \end{equation}
-  with $M\in \mathbb{R}^{n \times d_y}$.
+  with $M\in \mathbb{R}^{m \times d_y}$.
 
@@
     %\SetAlgoLined
-    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
-\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
-\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
+    \KwIn{$X_{train} \in \mathbb{R}^{m \times d_x}$, $X_{test} \in
+\mathbb{R}^{m' \times d_x}$, $Y_{train} \in \mathbb{R}^{m\times d_y}$, $Y_{test}
+\in \mathbb{R}^{m'\times d_y}$, } \KwOut{estimate of prediction improvement
 $\Delta{R} \in \mathbb{D}^{d_x}$.} $H, G = \text{B2B}(X_{train}, Y_{train})$\;
@@
     %\SetAlgoLined
-    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
-\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
-\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
+    \KwIn{$X_{train} \in \mathbb{R}^{m \times d_x}$, $X_{test} \in
+\mathbb{R}^{m' \times d_x}$, $Y_{train} \in \mathbb{R}^{m\times d_y}$, $Y_{test}
+\in \mathbb{R}^{m'\times d_y}$, } \KwOut{estimate of prediction improvement
 $\Delta{R} \in \mathbb{D}^{d_x, d_y}$.}
@@
     %\SetAlgoLined
-    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
-\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
-\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
+    \KwIn{$X_{train} \in \mathbb{R}^{m \times d_x}$, $X_{test} \in
+\mathbb{R}^{m' \times d_x}$, $Y_{train} \in \mathbb{R}^{m\times d_y}$, $Y_{test}
+\in \mathbb{R}^{m'\times d_y}$, } \KwOut{estimate of prediction improvement
 $\Delta{R} \in \mathbb{D}^{d_x, d_z}$.} $H, G = \text{CCA}(X_{train},

### 2020-05-19 paper/fmri_exps.tex
n -> m

@@
 fMRI data within a 8mm radius of each vertex. For each subject and vertex, we
-thus build an observation matrix $Y \in \mathbb{R}^{n \times d_y}$ of $n\approx$ 1400 words
+thus build an observation matrix $Y \in \mathbb{R}^{m \times d_y}$ of $m\approx$ 1400 words
 by $d_y\approx40$ vertices per searchlight sphere. Each
@@
 
-This procedure yields an $X \in \mathbb{R}^{n \times d_x}$ matrix of $n\approx$
+This procedure yields an $X \in \mathbb{R}^{m \times d_x}$ matrix of $m\approx$
 840 TR (Repetition Time: 2 sec.) by
@@
 Each model was implemented for each subject and each time sample independently.
-Pairwise comparison between models were performed using a two-sided Wilcoxon test across subjects (n=90) using the average $\Delta R$ across time.
+Pairwise comparison between models were performed using a two-sided Wilcoxon test across subjects (m=90) using the average $\Delta R$ across time.
 Corresponding effect sizes are shown in Figure~\ref{fig:meg_results}, and p-values are reported below.

### 2020-05-19 paper/meg_exps.tex
n -> m

@@
 For each subject and each time sample relative to word onset, we
-build an observation matrix $Y \in \mathbb{R}^{n \times d_y}$ of $n\approx$ 2,700 words
+build an observation matrix $Y \in \mathbb{R}^{m \times d_y}$ of $m\approx$ 2,700 words
 by $d_y=301$ MEG channels (273 magnetometers and 28 compensation channels). Each
@@
 We use the same features as for the fMRI experiments, except that the features
-were not convolved by an hemodynamic response function. This yields an $X \in \mathbb{R}^{n \times d_x}$
-matrix of $n\approx$ 2,700 words by $d_x=4$ features for each subject. Each of
+were not convolved by an hemodynamic response function. This yields an $X \in \mathbb{R}^{m \times d_x}$
+matrix of $m\approx$ 2,700 words by $d_x=4$ features for each subject. Each of
 the columns of $X$ is normalized to have a mean and a standard deviation of 0

### 2020-05-19 paper/paper.tex
n -> m

@@
 
-  \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{n \times d_y}$}};
-  \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{n \times d_x}$}};
+  \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{m \times d_y}$}};
+  \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{m \times d_x}$}};
   \node[] (annS) at (\posS, -1.8){\scalebox{0.85}{$S \in \mathbb{D}^{d_x \times d_x}$}};
-  \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{n \times d_x}$}};
+  \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{m \times d_x}$}};
   \node[] (annF) at (\posF, -1.8){\scalebox{0.85}{$F \in \mathbb{R}^{d_x \times d_y}$}};
@@
 
-We consider the measurement of multivariate signal $Y \in \mathbb{R}^{n \times
-d_y}$ (the dependent variables), generated from a set of putative causes $X \in \mathbb{R}^{n \times
+We consider the measurement of multivariate signal $Y \in \mathbb{R}^{m \times
+d_y}$ (the dependent variables), generated from a set of putative causes $X \in \mathbb{R}^{m \times
 d_x}$ (the independent variables), via some unknown linear apparatus $F \in \mathbb{R}^{d_x \times d_y}$.
@@
 %
-where $i$ is a given sample, and $n_i$ is a sample-specific noise drawn from a
+where $i$ is a given sample, and $n_i$ is a sample-specific additive noise drawn from a
 centered distribution.
@@
 %
-In practice, we observe $n$ samples $(X, Y)$ from the model.
+In practice, we observe $m$ samples $(X, Y)$ from the model.
 %
@@
     %\SetAlgoLined
-    \KwIn{input data $X \in \mathbb{R}^{n \times d_x}$, output data $Y \in \mathbb{R}^{n\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
+    \KwIn{input data $X \in \mathbb{R}^{m \times d_x}$, output data $Y \in \mathbb{R}^{m\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
     \KwOut{estimate of causal influences $\hat{S} \in \mathbb{D}^{d_x \times d_x}$.}
@@
 
-In the presence of noise, the causal factors of $S$ correspond to the positive
+In the presence of additive noise, the causal factors of $S$ correspond to the positive
 elements of $\text{diag}(\hat H)$. The methods to recover them are presented in

### 2020-05-19 paper/synthetic.tex
n -> m

@@
 
-The data generating process for each experiment constructs $n=1000$ training examples
+The data generating process for each experiment constructs $m=1000$ training examples
 according to the model $Y = (\text{h} XS + N)F$, where $\text{h}$ is a
@@
 %\begin{itemize} \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries
-%drawn from $\mathcal{N}(0, \sigma^2)$, \item $X \in \mathbb{R}^{n \times d_x}$
-%contains rows drawn from $\mathcal{N}(0, \Sigma_X)$, \item $N \in \mathbb{R}^{n
+%drawn from $\mathcal{N}(0, \sigma^2)$, \item $X \in \mathbb{R}^{m \times d_x}$
+%contains rows drawn from $\mathcal{N}(0, \Sigma_X)$, \item $N \in \mathbb{R}^{m
 %\times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, \item $S \in
@@
 $\mathcal{N}(0, \sigma^2)$ where $\sigma^2$ is inversely proportional to $d_x$,
-$X \in \mathbb{R}^{n \times d_x}$ contains rows
-drawn from $\mathcal{N}(0, \Sigma_X)$, $N \in \mathbb{R}^{n \times d_x}$
+$X \in \mathbb{R}^{m \times d_x}$ contains rows
+drawn from $\mathcal{N}(0, \Sigma_X)$, $N \in \mathbb{R}^{m \times d_x}$
 contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, $S \in \mathbb{R}^{d_x
@@
 area-under-the-curve (AUC) across factors $AUC(S, \hat S)$.
-%\begin{equation} AUC = 1 - \sum_1^n (S_k - S_{k-1}) ( \hat{S}_k +
+%\begin{equation} AUC = 1 - \sum_1^m (S_k - S_{k-1}) ( \hat{S}_k +
 %\hat{S}_{k-1}) / 2 \end{equation}

### 2020-05-20 paper/response.tex
up

Reviewers' comments:

Reviewer #1: The authors introduce a novel machine learning approach for the
    prediction of word length and word frequency in MEG time series data. The
    authors begin with an introduction to the problem, their solution, and a
    simulation demonstrating their approach is effective relative to other methods
    focused on the same problem. They demonstrate their approach using a real-world
    example. In short, I have relatively few remarks on their execution, which is
    satisfactory for my tastes. Overall, I think that many researchers in
    neuroimaging will find this technique useful.

    There is one major limitation for the broader application of this approach and
    that's demonstrating its effectiveness in fMRI data. The majority of
    neuroimaging studies are done using fMRI (and I do not say this to discount the
    other valuable tools of EEG, MEG, NIRS, etc.). To get other neuroscientists to
    grab onto this approach, I recommend the authors perform another analysis using
    fMRI data. My suggestion for the authors would be to use data from the Human
    Connectome Project (http://www.humanconnectomeproject.org/). The outcome
    variable used with this approach is not important as so much the demonstration
    that their technique is effective for fMRI data. Without including fMRI data,
    the approach comes off as useful but speculative for fMRI data, which
    drastically limits its impact potential.

We thank Reviewer #1 for this remark. We now applied the same set of analyses to
fMRI recordings of 100 subjects performing a reading experiment similar to the
MEG study described in the paper. We kept the exact same methods and parameters, and
used the fmriprep package to perform the preprocessing, so as to minimize the
risk of p-hacking.

Overall, the results consistently show that B2B compares favorably to other methods.
Furthermore, these novel analyses complements the temporal results obtained with MEG in
that the effects of Word Length are predominantly observed over the early visual
cortices, and the Word Frequency are predominantly observed over the
fronto-temporal areas.

# Reviewer 2

First, the authors justify their approach of using backward modeling by the
    ability to estimate common noise sources. However, this would also be possible
    using multivariate forward modeling. For that reason, I'm wondering how
    multivariate forward modeling would fare under these circumstances. I would find
    it highly informative to include a (similarly cross-validated) multivariate GLM
    to the analyses and results.

We thank Reviewer 2 for this remark. The multivariate GLM / MANOVA is indeed
a relevant analysis. We now indicate in supplementary analyses:

Following the recommendations of one of our reviewers, we also implemented a
multivariate variant of the forward model, i.e. a MANOVA, using the statsmodels
implementation \citep{seabold2010statsmodels}. MANOVA is primarily used as a
inferential statistics, and does not trivially convert to a predicting method.
Consequently, we did not find a way to compare MANOVA against B2B with the
$\Delta R$ evaluation. However, the effects of MANOVA comes with its own effect
size statistics: the Wilk's Lambda. For each searchlight, we thus use the $F-values$ of the
Wilk's Lambda statistics as proxy for $\hat S$ and feeds it to a second-level
Wilcoxon tests across subjects, like we did for the other models.

The results show that all factors, including the Dummy variable, are
systematically above chance level in all recorded brain regions \ref{fig:fmri_supp}. This result
can be explained by the large dimensionality of $Y$. Indeed, limiting the
searchlight to a 1mm radius did not lead to these spurious effects but provided
results similar to the Forward model.

Nonetheless, MANOVA does appear to capture some plausible effects. Indeed, the
$F-values$ obtained for both Word Length and Word Frequency were weakly but
significantly higher than those obtained with the Dummy variable in the
occipital and temporal brain areas \ref{fig:fmri_supp}.

Overall, this suggests that MANOVA (1) can lead to positively biased estimates of
$\hat S$ (2) appears weaker than B2B in terms of second-level analysis across
subjects (3) misses the effect of Word Function detected with the Forward model,
and (4) does not trivially translate into a prediction tool. Together these
elements thus suggest that MANOVA is less suitable to the present objective than
B2B.

> Likewise, while I applaud the authors efforts in conducting these analyses,
    I am wondering whether the impact of this method could not be made even
    stronger by adding more experimental findings. Are there any other openly
    available datasets that could be used to demonstrate the ability to
    disentangle known factors?

As indicated to Reviewer 1, we now successfully apply the same analyses to the fMRI
recordings of 100 subjects reading word sequences.

Other comments:

- Rather than using the feature importance delta_R, is it possible to just
    report the parameter estimate E_hat at each time point? What would be the
    downside of this?

It is possible to report E_hat (now relabelled S_hat). We now added a Supplementary Figure
to that effect. Note, however, that E_hat statistics are not in the same unit as the forward H_hat
coefficients (a.k.a betas values in GLM). Consequently, it is not straightforward to
compare these two metrics directly. As a proxy, we thus report their respective
second-level p-values across subjects. This figure shows show that B2B S_hat
coefficients compare favorably to the Forward H_hat coefficient and the MANOVA coefficients.

- The authors describe their method as unbiased but then introduce two
    regularization parameters, which will bias the estimated coefficients towards
    zero. While the statement "B2B leads to unbiased (i.e. zeros-centered) scalar
    coefficients for non-causal features" is correct for the mean of those
    coefficients, the variance will also be biased. I think these are important
    criteria to consider.

We agree with this remark. We now clarify the meaning and implication of the bias.

Note that $\S hat$ is unbiased, in the sense that it is centered around zero when there is no
effect, only if the second regression $H$ is not regularized. Second-level
statistics testing whether $\S hat$ is superior to 0 are thus only valid if $H$
is not regularized.

- The results of backward regression are not shown in Figure 4, but it would be
    nice to see those results. Is there a possibility to add them to Figure 4 in a
    separate panel?

We now added the decoding performance to a supplementary figure.

- I think the authors could better explain the dimensions of the data (dy
    referring to the number of measurement channels, and dx to the number of
    independent variables). Perhaps the authors could add this nomenclature to their
    graphical illustration? Likewise, for the regression of Y onto X, it is unclear
    if this is a mass-univariate regression or a multivariate regression (i.e.
    taking the error covariance between Ys into account or not). This might be
    obvious to them, but is not necessarily obvious to the reader.

We corrected the text to clarify this issue. However, we did not find an elegant way to
modify the figures.

- I find the choice of the letter E for one of the causal factors unfortunate
    since it is typically reserved for the true error component of a multivariate
    model. Likewise, the choice of N is unfortunate since the letter n has multiple
    meanings in the text: one particular instance of n, or the number of samples.
    This is a very minor comment though.

We now changed $E$ into $S$ for Selection matrix. We also changed the number of
samples into $m$, and keep $N$ as the Noise matrix. Thank you for these suggestions

- Typo line 28: "were" should read "where"?
Corrected.

- Line 181: do the authors refer to the Wilcoxon signed-rank test?
Yes, corrected

### 2020-05-20 paper/appendix.tex
up

@@
 
+\subsection{Supplementary comparison of models' coefficients}
+
+Following the recommendations of one of our reviewers, we also implemented a
+multivariate variant of the forward model, i.e. a MANOVA, using the statsmodels
+implementation \citep{seabold2010statsmodels}. MANOVA is primarily used as a
+inferential statistics, and does not trivially convert to a predicting method.
+Consequently, we did not find a way to compare MANOVA against B2B with the
+$\Delta R$ evaluation. However, the effects of MANOVA comes with its own effect
+size statistics: the Wilk's Lambda. For each searchlight, we thus use the $F-values$ of the
+Wilk's Lambda statistics as proxy for $\hat S$ and feeds it to a second-level
+Wilcoxon signed-rank tests across subjects, like we did for the other models.
+
+The results show that all factors, including the Dummy variable, are
+systematically above chance level in all recorded brain regions \ref{fig:fmri_supp}. This result
+can be explained by the large dimensionality of $Y$. Indeed, limiting the
+searchlight to a 1mm radius did not lead to these spurious effects but provided
+results similar to the Forward model.
+
+Nonetheless, MANOVA does appear to capture some plausible effects. Indeed, the
+$F-values$ obtained for both Word Length and Word Frequency were weakly but
+significantly higher than those obtained with the Dummy variable in the
+occipital and temporal brain areas \ref{fig:fmri_supp}.
+
+Overall, this suggests that MANOVA (1) can lead to positively biased estimates of
+$\hat S$ (2) appears weaker than B2B in terms of second-level analysis across
+subjects (3) misses the effect of Word Function detected with the Forward model,
+and (4) does not trivially translate into a prediction tool. Together these
+elements thus suggest that MANOVA is less suitable to the present objective than
+B2B.
+
 \begin{figure}
@@
                      trim=0cm 0cm 0cm 0cm,
-                     clip=True]{figures/fmri_decod.pdf}
-    \includegraphics[width=0.7\textwidth,
-                     trim=0cm 0cm 0cm 0cm,
-                     clip=True]{figures/fmri_betas.pdf}
+                     clip=True]{figures/fmri_controls.pdf}
 
@@
   \end{center}
-  \caption{Top. P-values across subjects derived from a Backward decoding of
-  each feature (columns). Bottom. P-values across subjects derived from
-  the $\beta$ coefficients of a Forward Model and the $\hat S$ of an unbiased
-  B2B (i.e. B2B where $H$ is not regularized). Each cell displays the left
-  hemisphere view laterally (top) or ventrally (bottom)}
+  \caption{Top rows. Pearson $R$ correlation obtained for a Backward decoding model and
+  a Forward encoding model trained with one factor at a time. These models can
+  not take into account the factor covariance, and thus lead to spurious effects
+  (e.g. visual cortex effect for the dummy variable). Bottom rows. Second-
+  level p-values across subjects for a Forward, a B2B and a Manova trained with
+  all factors. The B2B achieves better p-values, without leading to spurious effects for
+  the Dummy variable. The Manova leads to biased estimates due to its inability
+  to deal with overfitting. The last row shows where the Manova F-values of each factor differs
+  significantly from the F-values of the Dummy variable.}
 \end{figure}

### 2020-05-20 paper/fmri_exps.tex
up

@@
 Each model was implemented for each subject and each time sample independently.
-Pairwise comparison between models were performed using a two-sided Wilcoxon test across subjects (m=90) using the average $\Delta R$ across time.
-Corresponding effect sizes are shown in Figure~\ref{fig:meg_results}, and p-values are reported below.
+Pairwise comparison between models were performed using a two-sided Wilcoxon
+signed-rank test across subjects (m=90) using the average $\Delta R$ across time.
+Corresponding effect sizes are shown in Figure~\ref{fig:meg_results}, and
+p-values are reported below.
 
@@
   predicted for by the gold standard Forward Model using all features and
-  thresholded with a Wilcoxon at $p<0.001$ (uncorrected) across subjects. Bottom
+  thresholded with a Wilcoxon signed-rank test at $p<0.001$ (uncorrected) across subjects. Bottom
   Right. Average $\Delta R$ within the ROI for each subject
@@
 
-On the contrary, the $\beta$
+On the contrary, the $H$
 coefficients of the Forward model reveals the expected spatial specificity of
@@
 subjects, and could thus underestimate the significance of each factor at each
-vertex, as evaluated with a Wilcoxon tests across subjects. In principle, such
+vertex, as evaluated with a  signed-rank tests across subjects. In principle, such
 inter-subject variability could be less impactful with
@@
 regularized) reaches higher levels of significance at each vertex
-than the Forward $\beta$ \ref{fig:fmri_supp}. (Note that a direct
-comparison between the Forward coefficients $\beta$ and those of B2B ($\hat S$)
+than the Forward $H$ \ref{fig:fmri_supp}. (Note that a direct
+comparison between the Forward coefficients $H$ and those of B2B ($\hat S$)
 is not feasible, as these two do not have the same units.)
@@
 To fairly compare the Forward, CCA, PLS and B2B methods on a common and robust
-evaluation metric we compare them on their ability to improve $Y$
+evaluation metric, robust to coefficient biases, we thus compare them on their ability to improve $Y$
 predictions of out-of-sample data when one factor is introduced into the model.

### 2020-05-20 paper/paper.tex
up

@@
   \end{tikzpicture}
-  \caption{Back-to-back regression identifies the subset of factors $S_{ii} = 1$ in $X$ that influence some observations $Y$ by 1) regressing from $Y$ to $X$ to obtain $\hat{X}$, and 2) returning the diagonal of the regression coefficients from $X$ to $\hat{X}$.}
+  \caption{Back-to-back regression identifies the subset of factors $S_{ii} = 1$ in $X$ that influence $m$ multidimensional observations $Y$ by 1) regressing from $Y$ to $X$ to obtain $\hat{X}$, and 2) returning the diagonal of the regression coefficients from $X$ to $\hat{X}$.}
 \label{fig:b2b}
@@
 %
-While the triplet of variables $X$ and $N$ are independent, we allow each of
+While $X$ and $N$ are independent, we allow each of
 them to have any form of covariance.
@@
 regularization, respectively.
-
-%
-% After obtaining $\hat{S}$, we can fit a final regression from $X \hat{S}$ to $Y$.
 
@@
 
+Note that $\S hat$ is unbiased, in the sense that it is centered around zero when there is no
+effect, only if the second regression $H$ is not regularized. Second-level
+statistics testing whether $\S hat$ is superior to 0 are thus only valid if $H$
+is not regularized.
+
 \section{Experiments}
@@
 
-\section{Related work}
+\section{Discussion}
 
-% Forward and cross-decomposition models have been used to identify the causal
-% contribution of correlated factors onto multi-dimensional observations (e.g.
-% \citep{naselaris2011encoding}). These approaches typically lead to multiple
-% coefficients for each features (i.e. one per dimension of $Y$ or one per
-% component respectively). Furthermore, these coefficients can be difficult to
-% summarize into a single causal estimate. By contrast, B2B quickly (Fig.
-% \ref{fig:duration}) leads to a single unbiased scalar values $\hat S$ tending
-% towards 1 and 0 for causal and non-causal features respectively.
+Here, we introduce B2B, a method to disentangle the causal contribution of
+collinear factors from multidimensional observations. After proving the validity
+of B2B, we show that it compares favorably against baseline models, such as
+Backward, Forward, CCA and PLS models both on a wide spectrum of synthetic data,
+and on two large neuroimaging datasets.
 
-A variety of statistical methods applied to neuroimaging have been
-proposed to clarify what is being represented in brain responses - i.e. what
-feature causes specific brain activity. For example, CCA and PLS have been used in
-neuroimaging for a variety of purposes such as
-denoising and subject alignment \citep{cca_hotelling, de2019multiway}. While CCA
-and PLS relates to B2B, these methods diverge in several ways. First, they have different objectives: CCA aims to find the potentially numerous and poorly
+In addition, B2B can be very fast to compute as long as both $H$ and $G$ are based
+on l2 optimization with l2 regularization \ref{fig:duration}. Furthermore, B2B
+need not be limited to these standard optimizations: the $H$ and $G$ operators
+can in principle be found with e.g. support vector or random forest regression.
+We reserve the study of these computational variants to future work.
+
+Like Forward and cross-decomposition models, B2B is limited by the correlations
+between factors. At the extreme, if two factors are identical, and thus fully
+correlated, none of these statistical method can disentangle the causal
+contribution of each of them separately. In practice, this implies that, like
+general linear models, B2B will best work with high signal-to-noise ratio and/or
+orthogonal factors, and will see its sensitivity diminish when the
+signal-to-noise ratio of collinear factors drops.
+
+In the present neuroimaging context, B2B follows a long series of statistical
+methods designed to characterize brain representations - i.e. to identify what
+sensory feature causes specific brain responses \citep{naselaris2011encoding}.
+In this regard, CCA and PLS have been used in electrophysiology and neuroimaging
+for tracking representations (e.g. \citep{musch2020transformation}) as well as
+for denoising and subject alignment \citep{cca_hotelling, de2019multiway}. While
+CCA and PLS relates to B2B, these methods diverge in several important ways.
+First, they have different objectives: CCA aims to find the potentially numerous and poorly
 interpretable components where $X$ and $Y$ are maximally correlated, whereas B2B
@@
 selected to maximize signal-to-noise ratio of that feature $j$. Fourth CCA does
-not separately optimize two distinct regularization
-parameters for $G$ and $H$, where B2B does. Finally, CCA does not use different data splits to
+not separately optimize two distinct regularization parameters for $G$ and $H$,
+whereas B2B does. Finally, CCA does not use different data splits to
 estimate $G$ and $H$. Together, these differences may explain why B2B reliably
@@
 
-One popular multivariate analysis of brain activity is
+One popular method to investigate multidimensional patterns of brain activity is
 Representational Similarity Analysis (RSA)
-\citep{kriegeskorte2008representational}, and consists in analyzing the
+\citep{kriegeskorte2008representational}. RSA quantifies the
 similarity of brain responses associated with specific categorical conditions
 (e.g. distinct images), by (1) fitting one-against-all classifiers on each
-condition and (2) testing whether these classifiers can discriminate all other
-conditions. The resulting confusion matrix is then analyzed, generally in an unsupervised
-manner, to reveal which conditions lead to similar brain activity patterns. B2B
-differs from RSA in that it uses regressions instead of one-hot classifications.
-Consequently, and unlike RSA, B2B (1) is fully supervised (2) provides
-interpretable coefficients and (3) can generalize to new items and new contexts.
-In practice, these elements allow B2B to apply to event-related paradigms and
-latent variable analyses.
+category and (2) testing whether these classifiers discriminates all other
+categories. The resulting categories $\times$ categories confusion matrix is
+then analyzed, generally in an unsupervised manner, to reveal the categories
+that present similar brain activity patterns. B2B
+subsumes RSA in that (1) it uses regressions instead of one-hot
+classifications and (2) it is fully supervised.
+Consequently, and unlike RSA, B2B (1) provides interpretable coefficients and
+(2) can generalize to new items and new contexts. In practice, these elements
+allow B2B to apply to event-related paradigms and latent variable analyses,
+whereas RSA can only be applied when the same one-hot-encoded condition is
+repeated multiple times.
 
-The present empirical results, together with their theoretical foundations,
-suggest that B2B may come as a tool of choice for causal discovery when
-feature-wise intervention is limited.
+More generally, the present empirical results, together with their theoretical
+foundations, suggest that B2B may serve as a method of choice for causal
+discovery when feature-wise intervention is limited.
 
@@
 feedback. This work was supported by ANR-17-EURE-0017 and the Fyssen Foundation
-to JRK.
+to JRK for his work at PSL.
 

### 2020-05-20 paper/introduction.tex
up

@@
 minimal subset that suffices to reliably predict novel observations. This
-endeavor is impeded by two major challenges.
-
-First, causal and non-causal factors may be numerous and partially correlated.
+endeavor is impeded by two major challenges. First, causal and non-causal
+factors may be numerous and partially correlated.
 % In physics,
@@
 
-The second challenge to measuring causal influence is that observations can be
+The second challenge for measuring causal influence is that observations can be
 multidimensional.
@@
 brain voxels or multiple electrodes to detect the
-category of a stimulus independently of common noise sources (e.g. head
-movements, eye blink)
-\citep{norman2006beyond,cichy2014resolving,
-kriegeskorte2008representational, king2018encoding}.
+category of a stimulus \citep{norman2006beyond,cichy2014resolving,
+kriegeskorte2008representational, king2018encoding}. Decoding has become
+particularly popular in neuroscience, because brain recordings are typically
+corrupted by major physiological noise, such as muscle movements, eye blinks,
+displacements etc. As these noises sources often lie along specific
+components of the multidimensional recordings, the remaining neural signals can
+be robustly picked up by multiple decoders \citep{king2018encoding}.
 
@@
 Specifically, forward modeling disentangles the independent contribution of
-correlated factors, but does not combine multidimensional observations. By
-contrast, backward modeling combines multiple observations, but does not
+correlated factors but does not efficiently combine high-dimensional observations. By
+contrast, backward modeling combines multiple observations but does not
 disentangle factors that are linearly correlated \citep{weichwald2015causal,
@@
 techniques such as Partial Least Squares (PLS) and Canonical Correlation
-Analysis (CCA) \citep{de2019multiway, bilenko2016pyrcca}. CCA and PLS aim to find, from two sets of
-data $X$ and $Y$, the components $H$ and $G$ such that $XH$ and $YG$ are maximally
+Analysis (CCA) \citep{de2019multiway, bilenko2016Pyrcca}. CCA and PLS aim to find, from two sets of
+data $X$ and $Y$, the matrices $H$ and $G$ such that $XH$ and $YG$ are maximally
 correlated or maximally covarying respectively.
@@
 for feature discovery. First, these methods are not not directional: observations
-and factors can be assigned to either $X$ or $Y$. Second, these method project $X$
+and factors can be assigned to either $X$ or $Y$. Second, these methods project $X$
 and $Y$ onto a reduced but nonetheless multidimensional space. Third, because
@@
 Here, we introduce the `back-to-back regression' (B2B), which not only combines
-the benefits of forward and backward modeling (Section~\ref{sec:algorithm}), but
+the benefits of forward and backward modeling (Section~\ref{sec:algorithm}) but
 can also provide robust, interpretable, unidimensional and unbiased coefficients for
@@
 disentangling causal factors (Section~\ref{sec:experiment_synthetic}). Finally,
-we apply B2B to large neuroimaging datasets and reveal that distinct but
-linearly-correlated word features lead to distinguishable brain representations
-(Section~\ref{sec:experiment_real}).
+we apply B2B to large MEG and fMRI datasets acquired during a simple reading
+task and reveal that B2B efficiently distinguishes the respective effects of
+covarying word features (Section~\ref{sec:experiment_real}).

### 2020-05-20 paper/paper.tex
up

@@
 introduce ``Back-to-Back'' regression (B2B), a linear method designed to efficiently
-measure, from a set of correlated factors, those that most plausibly account for
+estimate, from a set of correlated factors, those that most plausibly account for
 multidimensional observations. First, we prove the consistency of B2B, its links to
-other linear approaches, and show how it provides a robust, unbiased and
+other linear approaches, and show how it can provide a robust, unbiased and
 interpretable scalar estimate for each factor.
 Second, we use a variety of simulated data to show that B2B outperforms
-least-squares regression and cross-decomposition techniques (e.g. canonical
+forward modeling ("encoding"), backward modeling ("decoding") as well as
+cross-decomposition modeling (i.e.. canonical
 correlation analysis and partial least squares) on causal identification when
-the factors and the observations are partially collinear.
-Finally, we apply B2B to magneto-encephalography and functional Magnetic
-Resonance Imaging of 102 subjects
-recorded during a one hour long reading task to test whether our method appropriately disentangles
+the factors and the observations are not orthogonal.
+Finally, we apply B2B to a hundred magneto-encephalography recordings and to a hundred functional Magnetic
+Resonance Imaging recordings acquired during a one hour reading task. B2B successfully disentangles
 the respective contribution of collinear factors such as word length, word
-frequency respectively known to cause early visual and late associative cortical
-responses. The results
-show that these factors are better disentangled with B2B than with
-other standard techniques.
+frequency in the early visual and late associative cortical
+responses respectively. These factors are better disentangled with B2B than with
+other standard techniques. We discuss how the speed and generality of B2B
+sets the foundations of an omnibus test for disentanglement of covarying
+factors in multidimensional observations.
 \end{abstract}
@@
 \begin{keyword}
-  Cross-Decomposition \sep
   Feature Discovery \sep
@@
   Encoding \sep
+  Cross-Decomposition \sep
   Reading \sep
@@
 We consider the measurement of multivariate signal $Y \in \mathbb{R}^{m \times
-d_y}$ (the dependent variables), generated from a set of putative causes $X \in \mathbb{R}^{m \times
-d_x}$ (the independent variables), via some unknown linear apparatus $F \in \mathbb{R}^{d_x \times d_y}$.
+d_y}$ (the dependent variables, e.g. the neural responses),
+generated from a set of putative causes $X \in \mathbb{R}^{m \times
+d_x}$ (the independent variables, e.g. the features of a stimulus),
+via some unknown linear apparatus $F \in \mathbb{R}^{d_x \times d_y}$ (e.g. the
+projection of neural activity onto MEG channels or fMRI voxels).
 %
@@
 This problem space, along with the sizes of all variables involved, is
-illustrated in Figure~\ref{fig:b2b}.
+illustrated in Fig.~\ref{fig:b2b}.
 %
@@
 Regression (B2B) is to estimate the matrix $S$, i.e. to identify the factors
-that most reliably account for the multivariate observations.
+that reliably account for the multivariate observations.
 
@@
 %
-To avoid this issue, we split our sample $(X, Y)$ into two splits $(X_1, Y_1)$
+To avoid this issue, we split our sample $(X, Y)$ into two disjoint sets $(X_1, Y_1)$
 and $(X_2, Y_2)$.
@@
 repeated over many random splits, and the final estimate $\hat S$ of the causal
-influence matrix is the average over the estimates associated to each split
+influence matrix is the average over the estimates associated with each split
 \citep{breiman1996bagging}.
@@
 To accelerate this ensembling procedure, we use an efficient
-leave-one-out cross-validation scheme as detailed in \citep{rifkin2007notes}
+leave-one-out (LOO) cross-validation scheme as detailed in \citep{rifkin2007notes}
 % http://cbcl.mit.edu/publications/ps/MIT-CSAIL-TR-2007-025.pdf,
@@
 \begin{equation}
-\hat{Y} = (\Sigma_X G Y - \text{diag}(\Sigma_X G) Y) \;/\; \text{diag}(I - \Sigma_X G) \qquad \text{(element-wise division)}
+\hat{Y}_{LOO} = (\Sigma_X G Y - \text{diag}(\Sigma_X G) Y) \;/\; \text{diag}(I - \Sigma_X G) \qquad \text{(element-wise division)}
 \end{equation}
@@
 
-Note that $\S hat$ is unbiased, in the sense that it is centered around zero when there is no
+Note that $hat \S $ is unbiased, in the sense that it is centered around zero when there is no
 effect, only if the second regression $H$ is not regularized. Second-level
-statistics testing whether $\S hat$ is superior to 0 are thus only valid if $H$
+statistics testing whether $hat \S$ is superior to 0 are thus only valid if $H$
 is not regularized.
@@
 
-We perform two sets of experiments to evaluate B2B: one on controlled synthetic
-data, and a second one on a real, large-scale magneto-encephalography (MEG)
+We perform three sets of experiments to evaluate B2B: one on controlled synthetic
+data, a second one on a real, large-scale functional Magnetic Resonance Imaging
+(fMRI) dataset and a third one on a real, large-scale magneto-encephalography (MEG)
 dataset.
@@
 We use scikit-learn's PLS and RidgeCV \citep{sklearn} as well as Pyrcca's
-regularized canonical component analysis (RegCCA, \citep{bilenko2016pyrcca})
-objects to compare B2B against the standard baselines.
+regularized canonical component analysis (RegCCA, \citep{bilenko2016Pyrcca})
+objects to compare B2B against the standard baselines, with common
+hyper-parameter optimizations.
 
-\subsection{Synthetic data}
+\subsection{Synthetic Experiment}
 \label{sec:experiment_synthetic}
@@
 
-\subsection{functional Magnetic Resonance Imaging}
+\subsection{functional Magnetic Resonance Imaging Experiment}
 \label{sec:experiment_fmri}
@@
 
-\subsection{Magneto-encephalography}
+\subsection{Magneto-encephalograph Experiment}
 \label{sec:experiment_real}
@@
 collinear factors from multidimensional observations. After proving the validity
-of B2B, we show that it compares favorably against baseline models, such as
-Backward, Forward, CCA and PLS models both on a wide spectrum of synthetic data,
-and on two large neuroimaging datasets.
+of B2B, we show that it generally compares favorably against baseline models
+both on a wide spectrum of synthetic data and on two large neuroimaging datasets.
 
 In addition, B2B can be very fast to compute as long as both $H$ and $G$ are based
-on l2 optimization with l2 regularization \ref{fig:duration}. Furthermore, B2B
-need not be limited to these standard optimizations: the $H$ and $G$ operators
+on l2 optimization with l2 regularization (Supplementary
+Fig.~\ref{fig:duration}), as is done in the above experiments. However, B2B
+does not need not be limited to l2 optimization and regularization: the $H$ and $G$ operators
 can in principle be found with e.g. support vector or random forest regression.
-We reserve the study of these computational variants to future work.
+We reserve the study of these computational variants for future work.
 
-Like Forward and cross-decomposition models, B2B is limited by the correlations
+Like forward and cross-decomposition models, B2B is limited by the correlations
 between factors. At the extreme, if two factors are identical, and thus fully
-correlated, none of these statistical method can disentangle the causal
-contribution of each of them separately. In practice, this implies that, like
+correlated, no statistical method can disentangle their relative causal
+contribution, and intervention is thus mandatory. In practice, this implies that, like
 general linear models, B2B will best work with high signal-to-noise ratio and/or
@@
 In this regard, CCA and PLS have been used in electrophysiology and neuroimaging
-for tracking representations (e.g. \citep{musch2020transformation}) as well as
-for denoising and subject alignment \citep{cca_hotelling, de2019multiway}. While
+to track representations (e.g. \citep{musch2020transformation}) as well as
+to denoise recordings as well as to align subjects \citep{cca_hotelling, de2019multiway}. While
 CCA and PLS relates to B2B, these methods diverge in several important ways.
@@
 optimizing over the decoders $G$ and then over $H$. By contrast, CCA and PLS are
-symmetric between $X$ and $Y$, and aims to find $G$ and $H$ such that they
+symmetric between $X$ and $Y$, and aims to find $G$ and $H$ jointly such that they
 project $X$ and $Y$ on maximally correlated dimensions. Third, CCA is based an
@@
 potentially dilute the impact of each feature onto multiple components. In
-contrast B2B assesses each feature $X^j$ on a single $Y$ component specifically
-selected to maximize signal-to-noise ratio of that feature $j$. Fourth CCA does
+contrast B2B assesses each feature $X_i$ on a single $Y$ component specifically
+selected to maximize signal-to-noise ratio of that feature $i$. Fourth CCA does
 not separately optimize two distinct regularization parameters for $G$ and $H$,
 whereas B2B does. Finally, CCA does not use different data splits to
-estimate $G$ and $H$. Together, these differences may explain why B2B reliably
-outperform CCA on estimating causal influences (Figs. \ref{fig:percondition} and
+estimate $G$ and $H$. Together, these differences may explain why B2B generally
+outperform CCA on estimating causal influences (Figs.~\ref{fig:percondition} and
 \ref{fig:auc_plots}).
@@
 that present similar brain activity patterns. B2B
-subsumes RSA in that (1) it uses regressions instead of one-hot
+subsumes RSA in that (1) it can use regressions instead of one-hot
 classifications and (2) it is fully supervised.

### 2020-05-20 paper/synthetic.tex
up

@@
 
-All baseline methods were based on the implementations of scikit-learn \citep{sklearn} and pyrcca
-\citep{bilenko2016pyrcca}. For pedagogical purposes, we briefly summarize them below.
+All baseline methods were based on the implementations of scikit-learn \citep{sklearn} and Pyrcca
+\citep{bilenko2016Pyrcca}. For pedagogical purposes, we briefly summarize them below.
 
@@
 
-We employ five-fold cross-validation to select the optimal number of components
+We employ five-fold nested cross-validation to select the optimal number of components
 for CCA and PLS. Regressions were $\ell2$-regularized with a $\lambda$ regularization
@@
 B2B leads to \emph{scalar} coefficients for non-causal
-features. The diagonal of this matrix, $\hat S$, can thus be directly used
+features. The diagonal of this matrix, $\hat S \in \mathbb{R}^{d_x}$, can thus be directly used
 as a causal contribution estimate. Note that this estimate is unbiased (i.e.
@@
 
-In contrast, the Forward, Backward, CCA and PLS models lead to a
-loading \emph{vector} for each feature: each vector indicates how to a feature
-relates to each dimension of the observations (for Forward or Backward
-models) or each canonical projection of the observations.
-To estimate causal contribution, we must thus transform such vector into a
-scalar, by e.g. taking the sum of square coefficients:
+In contrast, the loading coefficients of the Forward ($H_i \in \mathbb{R}^{d_y}$),
+Backward ($G^i \in \mathbb{R}^{d_y}$), CCA and PLS models ($H_i \in \mathbb{R}^{d_z}$) lead to a
+loading \emph{vector} for each feature $i$.
+To estimate causal contribution, we must thus transform such vectors into
+scalars, by e.g. taking the sum of square coefficients:
 % \begin{equation}
@@
 % \end{equation}
-Note that in such B2B cases, the estimates are  thus necessarily positive and thus biased.
+Note that in such B2B cases, the estimates are thus positive and would thus bias
+a second-level statistical analysis against 0.
 
@@
 this synthetic setup, but not in the neuroimaging experiments below.
+
+Figures~\ref{fig:percondition} (top) and~\ref{fig:auc_plots} (top) show the
+results of this AUC evaluation. Note that the figure does not display each
+feature separately, as they are randomly generated. The results show that B2B
+compares favorably to other methods on these synthetic data.
 
@@
 
-\subsubsection{Evaluating Causal Discovery with held-out prediction reliability}
+\subsubsection{Evaluating Causal Discovery through the reliability of held-out prediction}
 
-In most real-world cases, $S$ is not known. Consequently, the above AUC evaluation can not be estimated.
+In most real-world cases, $S$ is not known. Consequently, the above AUC
+evaluation cannot be estimated.
 To address this issue, we assess the ability of each model to reliably predict
@@
 difference $\Delta R_i = R_{full}-R_{knockout}$ indicates how much each
-$X_i$ improves the prediction of (a pre-defined dimension or projection of) $Y$.
-
+$X_i$ improves the prediction of
+a) the target dimension (i.e. $G^iY \in \mathbb{R}$ for B2B,
+b) the average across all of the dimensions $j$ of $Y$
+($\frac{1}{d_y}\sum_{j}^{d_y}\Delta R_i^{j}$) for the
+Forward model or
+c) the average across the canonical dimensions $j$ of $Y$
+($\frac{1}{d_z}\sum_{j}^{d_z}\Delta R_i^{j}$) for CCA and PLS.
 We show in Appendix~\ref{appendix:feature_importance} pseudo-code to assess
@@
 
-\subsubsection{Results of the synthetic experiment}
-
-Figures~\ref{fig:percondition} (bottom) and \ref{fig:auc_plots} (right, in
-Appendix) show the results of our synthetic experiment. Note that the figure
-does not display each feature separately, as they are
-randomly generated. $\Delta R$ represents the average of $\Delta R_i$ across
-features $i$.
+Figures~\ref{fig:percondition} (bottom) and~\ref{fig:auc_plots} (right, in
+Appendix) show the results of this evaluation on held-out data.
 Overall both the AUC and the held-out prediction reliability evaluations

### 2020-05-20 paper/response.tex
response

@@
 $\Delta R$ evaluation. However, the effects of MANOVA comes with its own effect
-size statistics: the Wilk's Lambda. For each searchlight, we thus use the $F-values$ of the
+size statistics: the Wilk's Lambda. For each searchlight, we thus use the $F$-values of the
 Wilk's Lambda statistics as proxy for $\hat S$ and feeds it to a second-level
@@
 The results show that all factors, including the Dummy variable, are
-systematically above chance level in all recorded brain regions \ref{fig:fmri_supp}. This result
+systematically above chance level in all recorded brain regions (Supplementary Fig.~\ref{fig:fmri_supp}). This result
 can be explained by the large dimensionality of $Y$. Indeed, limiting the
@@
 Nonetheless, MANOVA does appear to capture some plausible effects. Indeed, the
-$F-values$ obtained for both Word Length and Word Frequency were weakly but
+$F$-values obtained for both Word Length and Word Frequency were weakly but
 significantly higher than those obtained with the Dummy variable in the
-occipital and temporal brain areas \ref{fig:fmri_supp}.
+occipital and temporal brain areas~\ref{fig:fmri_supp}.
 
@@
 
-Note that $\S hat$ is unbiased, in the sense that it is centered around zero when there is no
+Note that $\hat S$ is unbiased, in the sense that it is centered around zero when there is no
 effect, only if the second regression $H$ is not regularized. Second-level
-statistics testing whether $\S hat$ is superior to 0 are thus only valid if $H$
+statistics testing whether $\hat S$ is superior to 0 are thus only valid if $H$
 is not regularized.

### 2020-05-20 paper/response.tex
clean up

--- 2020-05-20 paper/response.tex
@@
 
-Reviewer #1: The authors introduce a novel machine learning approach for the
-    prediction of word length and word frequency in MEG time series data. The
-    authors begin with an introduction to the problem, their solution, and a
-    simulation demonstrating their approach is effective relative to other methods
-    focused on the same problem. They demonstrate their approach using a real-world
-    example. In short, I have relatively few remarks on their execution, which is
-    satisfactory for my tastes. Overall, I think that many researchers in
-    neuroimaging will find this technique useful.
-
+Reviewer #1:
     There is one major limitation for the broader application of this approach and
@@
 
-We thank Reviewer #1 for this remark. We now applied the same set of analyses to
+We thank Reviewer #1 for this remark. We now apply the same analyses to
 fMRI recordings of 100 subjects performing a reading experiment similar to the
-MEG study described in the paper. We kept the exact same methods and parameters, and
-used the fmriprep package to perform the preprocessing, so as to minimize the
-risk of p-hacking.
+MEG study described in original manuscript. We kept the exact same methods and parameters, and
+used the fmriprep package to perform the preprocessing.
 
 Overall, the results consistently show that B2B compares favorably to other methods.
-Furthermore, these novel analyses complements the temporal results obtained with MEG in
-that the effects of Word Length are predominantly observed over the early visual
-cortices, and the Word Frequency are predominantly observed over the
-fronto-temporal areas.
+Furthermore, these novel analyses complements the temporal results obtained with
+MEG. Specifically, the effects of Word Length are predominantly observed over the early visual
+cortices, and the Word Frequency are predominantly observed over the fronto-temporal areas.
 
-# Reviewer 2
-
+# Reviewer 2:
     First, the authors justify their approach of using backward modeling by the
@@
 We thank Reviewer 2 for this remark. The multivariate GLM / MANOVA is indeed
-a relevant analysis. We now indicate in supplementary analyses:
+a relevant analysis. We now indicate in supplementary materials the following analyses:
 
-Following the recommendations of one of our reviewers, we also implemented a
+"
+Following the recommendations of one of our reviewers, we implemented a
 multivariate variant of the forward model, i.e. a MANOVA, using the statsmodels
@@
 Consequently, we did not find a way to compare MANOVA against B2B with the
-$\Delta R$ evaluation. However, the effects of MANOVA comes with its own effect
-size statistics: the Wilk's Lambda. For each searchlight, we thus use the $F-values$ of the
+$\Delta R$ evaluation. However, the effects of MANOVA are generally summarized
+with the Wilk's Lambda stastistics or its transformation into an $F$-value.
+For each searchlight, we thus use the $F$-values of the
 Wilk's Lambda statistics as proxy for $\hat S$ and feeds it to a second-level
-Wilcoxon tests across subjects, like we did for the other models.
+Wilcoxon signed-rank tests across subjects, like we did for the other models.
 
 The results show that all factors, including the Dummy variable, are
-systematically above chance level in all recorded brain regions \ref{fig:fmri_supp}. This result
+systematically above chance level in all recorded brain regions (Fig.~\ref{fig:fmri_supp}). This result
 can be explained by the large dimensionality of $Y$. Indeed, limiting the
@@
 Nonetheless, MANOVA does appear to capture some plausible effects. Indeed, the
-$F-values$ obtained for both Word Length and Word Frequency were weakly but
+$F$-values obtained for both Word Length and Word Frequency were weakly but
 significantly higher than those obtained with the Dummy variable in the
-occipital and temporal brain areas \ref{fig:fmri_supp}.
+occipital and temporal brain areas (Fig.~\ref{fig:fmri_supp}). This results
+suggests that the effect size of the MANOVA can be biased and, thus, is not
+valid for second-level statistics.
 
@@
 B2B.
+"
 
@@
 
-As indicated to Reviewer 1, we now successfully apply the same analyses to the fMRI
-recordings of 100 subjects reading word sequences.
+We agree with Reviewer 2. As indicated to Reviewer 1, we now successfully apply
+the same analyses to the fMRI recordings of 100 subjects reading word sequences.
 
@@
 
-It is possible to report E_hat (now relabelled S_hat). We now added a Supplementary Figure
-to that effect. Note, however, that E_hat statistics are not in the same unit as the forward H_hat
+It is possible to report $\hat E$ (now relabelled $\hat S$), however they are
+interpretable only if the second regression $H$ is not regularized.
+We now added a Supplementary Figure to that effect.
+
+Note, however, that E_hat statistics are not in the same unit as the forward H_hat
 coefficients (a.k.a betas values in GLM). Consequently, it is not straightforward to
 compare these two metrics directly. As a proxy, we thus report their respective
-second-level p-values across subjects. This figure shows show that B2B S_hat
-coefficients compare favorably to the Forward H_hat coefficient and the MANOVA coefficients.
+second-level p-values across subjects. The associated figure shows that B2B's $\hat S$
+coefficients compare favorably to the Forward H_hat coefficient and the MANOVA
+coefficients.
 
@@
 
-We agree with this remark. We now clarify the meaning and implication of the bias.
+We agree with this remark. We now clarify the meaning and implication of the
+bias in the problem statement:
 
-Note that $\S hat$ is unbiased, in the sense that it is centered around zero when there is no
+"Note that $\hat S$ is unbiased, in the sense that it is centered around zero when there is no
 effect, only if the second regression $H$ is not regularized. Second-level
-statistics testing whether $\S hat$ is superior to 0 are thus only valid if $H$
-is not regularized.
+statistics testing whether $\hat S$ is superior to 0 are thus only valid if $H$
+is not regularized."
 
@@
 
-We now added the decoding performance to a supplementary figure.
+We now added the decoding performance to a supplementary figure. The latter show
+that approximately the same cortical network can be used to decode Word Length
+and Word Frequency. It also shows that the Dummy Variable can be decoded from
+various brain areas, and from the visual cortex in particular.
 
@@
 
-We corrected the text to clarify this issue. However, we did not find an elegant way to
-modify the figures.
+We amended the text in multiple places to clarify this issue.
 
@@
 - Typo line 28: "were" should read "where"?
-Corrected.
+Yes this is now corrected.
 
 - Line 181: do the authors refer to the Wilcoxon signed-rank test?
-Yes, corrected
+Yes, this is now corrected.

### 2020-05-20 paper/_check_can_be_removed_supplementary.tex
clean up

--- 2020-05-19 paper/appendix.tex
@@
-\newpage
+\documentclass{article}
+
+\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref, amsthm}
+\usepackage[ruled, linesnumbered]{algorithm2e}
+\usepackage[]{neurips_2019}
+\newtheorem{theorem}{Theorem}
+
+\title{Measuring causal influence with\\ back-to-back regression: the linear case - supplementary material}
+
+\begin{document}
+
 \appendix
-\section{Appendix}
 
-\subsection{Proof of consistency theorem}
-\label{appendix:theorem_proof}
+\maketitle
 
-Proof of the theorem in \ref{sec:theorem}:
+\section{Theorem - detailed proof}
+\label{sec:theorem}
 
@@
 
-     Consider the B2B model from equation \ref{eq:model} $$Y = (XS + N)F$$ with
-     $N$ centered and full rank noise.
+     Consider the B2B model from Equation $$Y = (XS + N)F$$ $N$ centered and full rank noise.
 
-     If $F$ and $X$ are full-rank on $Img(S)$, then, the solution of B2B, $\hat
-     H$ minimizes
+     If $F$ and $X$ are full-rank on $Img(S)$, then, the solution of B2B, $\hat H$, will minimize
 
-     $$\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$$ and satisfies
+     $$\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$$ and satisfy
 
@@
 
- Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions
- of B2B.
+ Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B.
 
@@
 
-In the same way, for $\hat H$, we have
+Samely, for $\hat H$, we have
 \begin{align*}
@@
 
-Let $F^\dagger$ be the pseudo inverse of $F$, and $Z=F^\dagger SF\hat G$, we
-have $FZ = FF^\dagger SF \hat G$
+Let $F^\dagger$ be the pseudo inverse of $F$, and $Z=F^\dagger SF\hat G$, we have $FZ = FF^\dagger SF \hat G$
 
-Since $F$ is full rank on $Img(S)$, we have $FF^\dagger S =S$, and $FZ = SF\hat
-G$
+Since $F$ is full rank on $Img(S)$, we have $FF^\dagger S =S$, and $FZ = SF\hat G$
 
-As $S$ is a binary diagonal matrix, it is an orthogonal projection and therefore
-a contraction, thus
+As $S$ is a binary diagonal matrix, it is an orthogonal projection and therefore a contraction, thus
  $$ \| NSF\hat G\|^2 \leq \| NF\hat G \|^2$$ and
@@
 
-Finally, $S\hat H = S SF\hat G = SF\hat G = \hat H$, since $S$, a binary
-diagonal matrix, is involutive. This completes the proof.
+Finally, $S\hat H = S SF\hat G = SF\hat G = \hat H$, since $S$, a binary diagonal matrix, is involutive. This completes the proof.
 \end{proof}
 
-\newpage
-\subsection{Modeling measurement noise}
-
-Equation \ref{eq:model} does not explicitly contain a measurement noise term.
-Yet, in most experimental cases, the problem is best described as:
-\begin{equation}
-  Y = (XS+N)F+M
-  \label{eq:model_noisy_measure}
-\end{equation}
-  with $M\in \mathbb{R}^{m \times d_y}$.
-
-This equation is actually equivalent to Equation \ref{eq:model} given our
-hypotheses. Indeed, we can rewrite $M = MF^{-1}F$ over $Img(F)$, which leads to:
-$$Y = (XS+N)F+M = (XS+N+MF^{-1})F = (XS+N')F$$
-
-Consequently, assuming that $F$ is full rank on $Img(XS)$, B2B yields the same
-solutions to equations \ref{eq:model} and \ref{eq:model_noisy_measure}.
-
-\subsection{Feature importance}
-\label{appendix:feature_importance}
-
-For B2B, feature importance is assessed as follows:
-
-\begin{algorithm}[H]
-    %\SetAlgoLined
-    \KwIn{$X_{train} \in \mathbb{R}^{m \times d_x}$, $X_{test} \in
-\mathbb{R}^{m' \times d_x}$, $Y_{train} \in \mathbb{R}^{m\times d_y}$, $Y_{test}
-\in \mathbb{R}^{m'\times d_y}$, } \KwOut{estimate of prediction improvement
-$\Delta{R} \in \mathbb{D}^{d_x}$.} $H, G = \text{B2B}(X_{train}, Y_{train})$\;
-$R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;
-
-    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\;
-    $R_{k} =
-\text{corr}(X_{test} K H, Y_{test} G_i)$\; $\Delta R_i = R_{full} - R_{k}$\; }
-\Return{$\Delta R$} \caption{B2B feature importance.} \label{algorithm:b2b_fi}
-\end{algorithm}
-
-For the Forward Model, the feature importance is assessed as follows:
-
-\begin{algorithm}[H]
-    %\SetAlgoLined
-    \KwIn{$X_{train} \in \mathbb{R}^{m \times d_x}$, $X_{test} \in
-\mathbb{R}^{m' \times d_x}$, $Y_{train} \in \mathbb{R}^{m\times d_y}$, $Y_{test}
-\in \mathbb{R}^{m'\times d_y}$, } \KwOut{estimate of prediction improvement
-$\Delta{R} \in \mathbb{D}^{d_x, d_y}$.}
-
-$H = \text{LinearRegression}(X_{train}, Y_{train})\;
-R_{full} = \text{corr}(X_{test} K, Y_{test})$\;
-
-    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $R_{k} = \text{corr}(X_{test}
-K H, Y_{test})$\; $\Delta R_i = R_{full} - R_{k}$\; } \Return{$\Delta R$}
-\caption{Forward feature importance.} \label{algorithm:fwd_fi} \end{algorithm}
-
-% \iffalse
-% \begin{enumerate}
-%   \item Fit $H$ given $X_{train}$ and $Y_{train}$
-%   \item For each feature $i$:
-%   \begin{itemize}
-%     \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
-%     \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train}$
-%   \end{itemize}
-%   \end{enumerate}
-% \fi
-
-For the CCA and PLS models, the feature importance is assessed as follows:
-
-\begin{algorithm}[H]
-    %\SetAlgoLined
-    \KwIn{$X_{train} \in \mathbb{R}^{m \times d_x}$, $X_{test} \in
-\mathbb{R}^{m' \times d_x}$, $Y_{train} \in \mathbb{R}^{m\times d_y}$, $Y_{test}
-\in \mathbb{R}^{m'\times d_y}$, } \KwOut{estimate of prediction improvement
-$\Delta{R} \in \mathbb{D}^{d_x, d_z}$.} $H, G = \text{CCA}(X_{train},
-Y_{train})$\; $R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;
-
-    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $R_{k} = \text{corr}(X_{test} K H,
-Y_{test} G)$\; $\Delta R_i = R_{full} - R_{k}$\; } \Return{$\Delta R$}
-\caption{CCA and PLS feature importance.} \label{algorithm:cdp_fi}
-\end{algorithm}
-
-For the Backward Model, feature importance cannot be assessed because there is
-no prediction.
-
-\subsection{Recovering S}
-\label{recovering}
-
-In case of noise, B2B yields non binary $\hat S$. Three thresholding rules can
-be used to binarize its values thus explicitly recover "causal" features.
-
-First, given known signal-to-noise ratio, the threshold above which a feature
-should considered to be "causal" can be derived analytically. Indeed, Equation
-$\ref{eq:diagk}$ implies that the $k$ first diagonal elements of $\hat H$ are
-bounded:  % we assume that $F$ has full rank over $Img(S)$.
-
-$$0 \leq \frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}} \leq
-\text{diag}_k(\hat{H})\leq  \frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$$
-where $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote
-the largest and smallest eigenvalues of $\Sigma_{X_1 X_1}$ and $\Sigma_{N_1
-N_1}$.
-
-The average value $\mu$ of non-zero coefficients of $\text{diag}(\hat H)$ is the
-trace of $\hat H$ divided by $k$, and can be computed as
-\begin{equation}
-\mu = \frac{Var(X)}{Var(X)+Var(N)}
-\end{equation}
-
-The decision threshold between "causal" and "non-causal" elements is thus a fraction
-$\mu$, whose proportion arbitrarily depends on the necessity to favor type I and
-type II errors. In practice, we cannot use this procedure for our MEG study,
-because signal-to-noise ratio is unknown.
-
-Second, $\text{diag}(\hat H)$ can be binarized with the Sonquist-Morgan
-criterion \citep{sonquist_morgan}, a non-parametric clustering procedure
-separating small and large values in a given set. This procedure maximizes the
-ratio of inter-group variance while minimizing the intra-group variance, over
-all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest
-values. Let $m_0$ and $m_1$ be the average values of the two clusters, $p$ and
-$d_x-p$ their size, and $v$ the total variance of the sample, Sonquist-Morgan
-criterion maximizes \citep{kass75}:
-\begin{equation}
-  \frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}
-\end{equation}
-
-This procedure assumes that there exists at least one causal and at least one
-non-causal feature.
-%
-Third, second-order statistics across multiple datasets can be used to identify
-the elements of $\text{diag}(\hat H)$ that are significantly different from 0.
-This procedure is detailed in the method section of our MEG experiment.
-
-Overall, these three procedures thus vary in their additional assumptions: i.e.
-(1) a known signal-to-noise ratio, (2) the existence of both causal and
-non-causal factors or (3) independent repetitions of the experiment.
-
-\section{Additional Figures}
-
-% \begin{figure}[t!]
-%   \centering
-%   \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_supp.pdf}
-%   \caption{Average $\Delta R$ for each $Y$ dimension (Forward model), canonical components (PLS and CCA) or feature (B2B) across subjects.
-Formally, Backward model cannot have a $\Delta R$ because it never unmixes the multiple $X$ features.
-In such cases, we thus simply report the decoding score.}
-%   \label{fig:meg_supp}
-% \end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{figures/delta_auc.pdf}
-  \caption{Synthetic experiments. Distribution (over conditions) of AUC (top) and Feature
-  Importance $\Delta R$ (bottom) metrics between our method (y-axis) and the baselines
-  (x-axis). Each dot is a distinct synthetic experiment. Dots below the diagonal indicates
-  that B2B outperform the tested model. \label{fig:auc_plots}}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.7\linewidth]{figures/duration}
-  \caption{Wall-clock run-time for our method B2B and for the baselines. Each dot is a
-  distinct synthetic experiment. B2B runs much faster than cross-decomposition baselines.
-  \label{fig:duration}}
-\end{figure}
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=0.7\textwidth,
-                     trim=0cm 0cm 0cm 0cm,
-                     clip=True]{figures/fmri_decod.pdf}
-    \includegraphics[width=0.7\textwidth,
-                     trim=0cm 0cm 0cm 0cm,
-                     clip=True]{figures/fmri_betas.pdf}
-
-    \label{fig:fmri_supp}
-  \end{center}
-  \caption{Top. P-values across subjects derived from a Backward decoding of
-  each feature (columns). Bottom. P-values across subjects derived from
-  the $\beta$ coefficients of a Forward Model and the $\hat S$ of an unbiased
-  B2B (i.e. B2B where $H$ is not regularized). Each cell displays the left
-  hemisphere view laterally (top) or ventrally (bottom)}
-\end{figure}
-
-\subsection{Robustness to increasing number of factors}
-
-To test whether each of the methods robustly scales to an increasingly
-large number of potential causes $X$, we enhanced the four ad-hoc features
-(word length, word frequency, word function, dummy variable) with another
-ten features. These additional features corresponds to the first dimensions
-of word embedding as provided by Spacy \citep{spacy2}. The results shown
-in Figure \ref{fig:embeddings}, show that the feature importance of ad-hoc
-features as derived by B2B remain unchanged and are actually improved.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.5\linewidth]{figures/compare_embeddings.pdf}
-  \caption{Comparison of $\Delta R$ when the models are tested on four
-  variables (top) and when the models are tested on an these four variables
-  as well as another 10 word-embedding features (bottom). These results
-  illustrate that, unlike Regularized CCA, B2B remains robust even when
-  the number of tested factors increases.
-  \label{fig:embeddings}}
-\end{figure}
+\end{document}

### 2020-05-20 paper/response.tex
pdf

@@
-Reviewers' comments:
+\documentclass[preprint,12pt,3p]{elsarticle}
 
-Reviewer #1:
+\usepackage{lineno,hyperref}
+% \modulolinenumbers[5]
+
+\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, amsthm}
+\usepackage[ruled, linesnumbered]{algorithm2e}
+\usepackage{times,wrapfig}
+\usepackage{cleveref}
+
+\begin{document}
+\title{Response to reviewers}
+
+\section{Reviewer 1}
+\begin{quote}\textbf{
     There is one major limitation for the broader application of this approach and
@@
     drastically limits its impact potential.
+}
+\end{quote}
 
-We thank Reviewer #1 for this remark. We now apply the same analyses to
+We thank Reviewer 1 for this remark. We now apply the same analyses to
 fMRI recordings of 100 subjects performing a reading experiment similar to the
@@
 
-# Reviewer 2:
+\section{Reviewer 2}
+\begin{quote}\textbf{
     First, the authors justify their approach of using backward modeling by the
@@
     to the analyses and results.
+}
+\end{quote}
 
@@
 
-"
-Following the recommendations of one of our reviewers, we implemented a
+\begin{quote}
+    "Following the recommendations of one of our reviewers, we implemented a
 multivariate variant of the forward model, i.e. a MANOVA, using the statsmodels
-implementation \citep{seabold2010statsmodels}. MANOVA is primarily used as a
+implementation. MANOVA is primarily used as a
 inferential statistics, and does not trivially convert to a predicting method.
@@
 The results show that all factors, including the Dummy variable, are
-systematically above chance level in all recorded brain regions (Fig.~\ref{fig:fmri_supp}). This result
+systematically above chance level in all recorded brain regions (Supp Fig.). This result
 can be explained by the large dimensionality of $Y$. Indeed, limiting the
@@
 significantly higher than those obtained with the Dummy variable in the
-occipital and temporal brain areas (Fig.~\ref{fig:fmri_supp}). This results
+occipital and temporal brain areas (Supp Fig.). This results
 suggests that the effect size of the MANOVA can be biased and, thus, is not
@@
 elements thus suggest that MANOVA is less suitable to the present objective than
-B2B.
-"
+B2B."
 
-> Likewise, while I applaud the authors efforts in conducting these analyses,
+\end{quote}
+
+\begin{quote}\textbf{
+Likewise, while I applaud the authors efforts in conducting these analyses,
     I am wondering whether the impact of this method could not be made even
@@
     disentangle known factors?
+}
+\end{quote}
 
@@
 
-Other comments:
-
-- Rather than using the feature importance delta_R, is it possible to just
-    report the parameter estimate E_hat at each time point? What would be the
+\begin{quote}\textbf{
+Rather than using the feature importance delta R, is it possible to just
+    report the parameter estimate E hat at each time point? What would be the
     downside of this?
+}
+\end{quote}
 
@@
 
-Note, however, that E_hat statistics are not in the same unit as the forward H_hat
+Note, however, that $\hat E$ statistics are not in the same unit as the forward $\hat H$
 coefficients (a.k.a betas values in GLM). Consequently, it is not straightforward to
@@
 second-level p-values across subjects. The associated figure shows that B2B's $\hat S$
-coefficients compare favorably to the Forward H_hat coefficient and the MANOVA
+coefficients compare favorably to the Forward $\hat H$ coefficient and the MANOVA
 coefficients.
 
-- The authors describe their method as unbiased but then introduce two
+\begin{quote}\textbf{
+The authors describe their method as unbiased but then introduce two
     regularization parameters, which will bias the estimated coefficients towards
@@
     criteria to consider.
+}
+\end{quote}
 
@@
 
-- The results of backward regression are not shown in Figure 4, but it would be
+\begin{quote}\textbf{
+The results of backward regression are not shown in Figure 4, but it would be
     nice to see those results. Is there a possibility to add them to Figure 4 in a
     separate panel?
-
+}
+\end{quote}
 We now added the decoding performance to a supplementary figure. The latter show
@@
 
-- I think the authors could better explain the dimensions of the data (dy
+\begin{quote}\textbf{
+I think the authors could better explain the dimensions of the data (dy
     referring to the number of measurement channels, and dx to the number of
@@
     obvious to them, but is not necessarily obvious to the reader.
-
+}
+\end{quote}
 We amended the text in multiple places to clarify this issue.
 
-- I find the choice of the letter E for one of the causal factors unfortunate
+\begin{quote}\textbf{
+I find the choice of the letter E for one of the causal factors unfortunate
     since it is typically reserved for the true error component of a multivariate
@@
     This is a very minor comment though.
-
+}
+\end{quote}
 We now changed $E$ into $S$ for Selection matrix. We also changed the number of
@@
 
-- Typo line 28: "were" should read "where"?
+\begin{quote}\textbf{
+Typo line 28: "were" should read "where"?
+}
+\end{quote}
 Yes this is now corrected.
 
-- Line 181: do the authors refer to the Wilcoxon signed-rank test?
+\begin{quote}\textbf{
+Line 181: do the authors refer to the Wilcoxon signed-rank test?
+}
+\end{quote}
 Yes, this is now corrected.
+
+\end{document}

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/coreg.py
up nimg

@@ -1,0 +1,426 @@
+"""
+:Module: coreg.py
+:Synopsis: Histogram-based co-registration of 3D MRI images.
+:Author: DOHMATOB Elvis Dopgima <gmdopp@gmail.com>
+
+"""
+
+from scipy.ndimage import gaussian_filter
+from scipy.optimize import fmin_powell
+import scipy.special
+from scipy.signal import sepfir2d
+import numpy as np
+import nibabel
+
+from nilearn._utils.compat import _basestring
+
+from .affine_transformations import (spm_matrix, apply_realignment,
+                                     nibabel2spm_affine)
+from .io_utils import loaduint8, get_basenames, save_vols
+from .kernel_smooth import fwhm2sigma, centered_smoothing_kernel
+from .histograms import (trilinear_interp, joint_histogram,
+                         _correct_voxel_samp, make_sampled_grid)
+
+# 'texture' of floats in machine precision
+EPS = np.finfo(float).eps
+
+
+def compute_similarity_from_jhist(jh, fwhm=None, cost_fun='nmi'):
+    """
+    Computes an information-theoretic similarity from a joint histogram.
+
+    Parameters
+    ----------
+    jh: 2D array
+        joint histogram of the two random variables being compared
+
+    fwhm: float or pair of float
+        kernel width for smoothing the joint histogram
+
+    cost_fun: string, optional (default 'nmi')
+        smilarity model to use; possible values are:
+        'mi': Mutual Information
+        'nmi': Normalized Mutual Information
+        'ecc': Entropic Cross-Correlation
+
+    Returns
+    -------
+    o: float
+        the computed similariy measure
+    """
+    # sanitize input
+    if fwhm is None:
+        fwhm = [7., 7.]
+    if len(np.shape(jh)) != 2:
+        raise ValueError("jh must be 2D array, got %s" % jh)
+
+    if len(np.shape(fwhm)) == 0:
+        fwhm = [fwhm] * 2
+    fwhm = fwhm[:2]
+
+    # create separable filter for smoothing the joint-histogram
+    lim = np.ceil(fwhm * 2)
+    krn1 = centered_smoothing_kernel(fwhm[0],
+                                     np.linspace(-1 * lim[0], lim[0],
+                                                 num=int(2 * lim[0])))
+    krn1 = krn1 / np.sum(krn1)
+    krn2 = centered_smoothing_kernel(fwhm[1],
+                                     np.linspace(-1 * lim[1], lim[1],
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/histograms.py

@@ -1,0 +1,333 @@
+"""
+:Module: histograms
+:Synopsis: building sampled meshgrids, masking meshgrids, performing trilinear
+interpolation, computation of joint histograms between fixed and moving
+3D images, etc.
+:Author: DOHMATOB Elvis Dopgima <gmdopp@gmail.com>
+
+"""
+
+import numpy as np
+from .affine_transformations import get_physical_coords
+from .io_utils import is_niimg
+
+
+# magic table from SPM
+ran = np.array([0.656619, 0.891183, 0.488144, 0.992646, 0.373326,
+                0.531378, 0.181316, 0.501944, 0.422195,
+                0.660427, 0.673653, 0.95733, 0.191866, 0.111216,
+                0.565054, 0.969166, 0.0237439, 0.870216,
+                0.0268766, 0.519529, 0.192291, 0.715689, 0.250673,
+                0.933865, 0.137189, 0.521622, 0.895202,
+                0.942387, 0.335083, 0.437364, 0.471156, 0.14931,
+                0.135864, 0.532498, 0.725789, 0.398703,
+                0.358419, 0.285279, 0.868635, 0.626413, 0.241172,
+                0.978082, 0.640501, 0.229849, 0.681335,
+                0.665823, 0.134718, 0.0224933, 0.262199, 0.116515,
+                0.0693182, 0.85293, 0.180331, 0.0324186,
+                0.733926, 0.536517, 0.27603, 0.368458, 0.0128863,
+                0.889206, 0.866021, 0.254247, 0.569481,
+                0.159265, 0.594364, 0.3311, 0.658613, 0.863634,
+                0.567623, 0.980481, 0.791832, 0.152594,
+                0.833027, 0.191863, 0.638987, 0.669, 0.772088,
+                0.379818, 0.441585, 0.48306, 0.608106,
+                0.175996, 0.00202556, 0.790224, 0.513609, 0.213229,
+                0.10345, 0.157337, 0.407515, 0.407757,
+                0.0526927, 0.941815, 0.149972, 0.384374, 0.311059,
+                0.168534, 0.896648
+                ])
+
+
+def _correct_voxel_samp(affine, samp):
+    """
+    Corrects a sampling rate (samp) according to the image's resolution.
+
+    Parameters
+    ----------
+    affine: 2D array of shape (4, 4)
+        affine matrix of the underlying image
+
+    samp: 1D array_like of 3 floats, optional (default [1, 1, 1])
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/io_utils.py

@@ -1,0 +1,1002 @@
+"""
+:Module: utils
+:Synopsis: routine business related to image i/o manips
+:Author: dohmatob elvis dopgima
+
+"""
+
+import os
+import warnings
+import re
+import shutil
+import tempfile
+import filecmp
+import numpy as np
+import nibabel
+from sklearn.externals import joblib
+from nipype.interfaces.dcm2nii import Dcm2nii
+from nipype.caching import Memory
+from nilearn.image import iter_img
+from nilearn.image.image import check_niimg, check_niimg_4d
+from nilearn._utils.compat import _basestring
+
+DICOM_EXTENSIONS = [".dcm", ".ima", ".dicom"]
+
+
+def is_niimg(img):
+    """
+    Checks whether given img is nibabel image object.
+
+    """
+    if isinstance(img, (nibabel.Nifti1Image,
+                        nibabel.Nifti1Pair,
+                        nibabel.Spm2AnalyzeImage,
+                        nibabel.Spm99AnalyzeImage,
+                        nibabel.MGHImage,
+                        # add other supported image types below (e.g
+                        # AnalyseImage, etc.)
+                        )):
+        return type(img)
+    else:
+        return False
+
+
+def load_vols(niimgs):
+    """Loads a nifti image (or a bail of) into a list qof 3D volumes.
+
+    Parameters
+    ----------
+    niimgs: 3 or 4D Niimg-like object
+        If niimgs is an iterable, checks if data is really 4D. Then,
+        considering that it is a list of niimg and load them one by one.
+        If niimg is a string, consider it as a path to Nifti image and
+        call nibabel.load on it. If it is an object, check if get_data
+        and get_affine methods are present, raise an Exception otherwise.
+
+    Returns
+    -------
+    niimgs_: list of nifti image objects
+        The loaded volumes.
+    """
+    # try loading 4d
+    try:
+        niimgs = list(check_niimg_4d(niimgs, return_iterator=True))
+    except TypeError:
+        # probably not 4d
+        niimgs = [check_niimg(niimgs)]
+    except ValueError:
+        # probably inconsisten affines
+        pass
+    try:
+        # try loading volumes one-by-one
+        if isinstance(niimgs, _basestring):
+            niimgs = [niimgs]
+        return [check_niimg(niimg, ensure_ndim=3) for niimg in niimgs]
+    except TypeError:
+        pass
+
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/kernel_smooth.py

@@ -1,0 +1,417 @@
+"""
+:Module: kernel_smooth
+:Synopsis: assorted utilities for smoothing images (to absorb noise before
+computing a gradient, etc.). A good starting point for understanding this code
+is the smooth_image(..) wrapper function
+:Author: DOHMATOB Elvis Dopgima, adapted from nipy source code. Credits
+to nipy dev.
+
+"""
+
+import numpy as np
+import numpy.fft as npfft
+import nibabel as ni
+import scipy.linalg
+import gc
+from nilearn._utils.compat import _basestring
+
+from .affine_transformations import get_physical_coords
+from .io_utils import is_niimg
+
+# 'texture' of floats in machine precision
+EPS = np.finfo(float).eps
+
+
+def fwhm2sigma(fwhm):
+    """Convert a FWHM value to sigma in a Gaussian kernel.
+
+    Parameters
+    ----------
+    fwhm: array-like
+       FWHM value or values
+
+    Returns
+    -------
+    sigma: array or float
+       sigma values corresponding to `fwhm` values
+
+    Examples
+    --------
+    >>> sigma = fwhm2sigma(6)
+    >>> sigmae = fwhm2sigma([6, 7, 8])
+    >>> sigma == sigmae[0]
+    True
+
+    """
+
+    fwhm = np.asarray(fwhm)
+
+    return fwhm / np.sqrt(8 * np.log(2))
+
+
+def sigma2fwhm(sigma):
+    """Convert a sigma in a Gaussian kernel to a FWHM value.
+
+    Parameters
+    ----------
+    sigma: array-like
+       sigma value or values
+
+    Returns
+    -------
+    fwhm: array or float
+       fwhm values corresponding to `sigma` values
+
+    Examples
+    --------
+    >>> fwhm = sigma2fwhm(3)
+    >>> fwhms = sigma2fwhm([3, 4, 5])
+    >>> fwhm == fwhms[0]
+    True
+
+    """
+
+    sigma = np.asarray(sigma)
+
+    return sigma * np.sqrt(8 * np.log(2))
+
+
+def _crop(X, tol=1.0e-10):
+    """Find a bounding box for support of fabs(X) > tol and returned
+    crop region.
+
+    Parameters
+    ----------
+    X: array_like
+        data to be cropped
+    tol: float, optional (default 1e-10)
+       tolerance for threholding the box
+
+    Returns
+    -------
+    cropped version of X
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/realign.py

@@ -1,0 +1,735 @@
+"""
+:Module: spm_realign
+:Synopsis: MRI within-modality Motion Correction, SPM style
+:Author: DOHMATOB Elvis Dopgima
+
+"""
+
+import os
+import numpy as np
+import scipy.ndimage as ndimage
+import scipy.linalg
+import nibabel
+from nilearn._utils.compat import _basestring
+from sklearn.externals.joblib import Parallel, delayed
+from .kernel_smooth import smooth_image
+from .affine_transformations import (
+    get_initial_motion_params, transform_coords, apply_realignment_to_vol,
+    apply_realignment)
+from .reslice import reslice_vols
+from .io_utils import save_vol, save_vols, get_basenames, is_niimg, load_vols
+
+# useful constants
+INFINITY = np.inf
+
+
+def _single_volume_fit(moving_vol, fixed_vol_affine, fixed_vol_A0,
+                       affine_correction, b, x1, x2, x3, fwhm, n_iterations,
+                       interp, lkp, tol, log=lambda x: None):
+    """
+    Realigns moving_vol to fixed_vol.
+
+    Paremeters
+    ----------
+    moving_vol: Nibabel object
+        coregistration source
+
+    fixed_vol_affine: 2D array
+        coregistration target
+
+    fixed_vol_A0: 2D array
+        rate of change of chi2 w.r.t. parameter changes
+
+    affine_correction: 2D array of shape (4, 4), optional (default None)
+        affine transformation to be applied to vols before realignment
+        (this is useful in multi-session realignment)
+
+    b: array
+        intercept vector
+
+    x1, x2, x3: arrays
+        grid
+
+    fwhm: float
+        the FWHM of the Gaussian smoothing kernel (mm) applied to the
+        images before estimating the realignment parameters.
+
+    n_iterations: int
+        max number of Gauss-Newton iterations when solving LSP for
+        registering a volume to the reference
+
+    interp: int
+        B-spline degree used for interpolation
+
+    lkp: arry_like of ints
+        affine transformation parameters sought-for. Possible values of
+        elements of the list are:
+        0  - x translation
+        1  - y translation
+        2  - z translation
+        3  - x rotation about - {pitch} (radians)
+        4  - y rotation about - {roll}  (radians)
+        5  - z rotation about - {yaw}   (radians)
+        6  - x scaling
+        7  - y scaling
+        8  - z scaling
+        9  - x shear
+        10 - y shear
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/reslice.py

@@ -1,0 +1,181 @@
+"""
+:Module: spm_reslice
+:Synopsis: Routine functions for reslicing volumes after affine registration
+(as in motion correction, coregistration, etc)
+:Author: DOHMATOB Elvis Dopgima
+
+"""
+
+import numpy as np
+import scipy.ndimage
+import scipy.linalg
+import nibabel
+from nilearn.image.image import check_niimg
+from .affine_transformations import get_physical_coords
+
+
+def _get_mask(M, coords, dim, wrp=[1, 1, 0], tiny=5e-2):
+    """
+    Wrapper for get_physical_coords(...) with optional wrapping of dimensions.
+
+    Parameters
+    ----------
+    M: 2D array of shape (4, 4)
+        affine transformation describing voxel-to-world mapping
+    coords: array_like of shape (3, n_voxels)
+        voxel(s) under consideration
+    dim: list of 3 ints
+        dimensions (nx, ny, nz) of the voxel space (for example [64, 56, 21])
+    wrp: list of 3 bools, optional (default [1, 1, 0])
+        each coordinate value indicates whether wrapping should be done in the
+        corresponding dimension or not. Possible values are:
+        [0, 0, 0]: no wrapping; use this value for PET data
+        [1, 1, 0]: wrap all except z (slice-wise) dimension; use this value for
+        fMRI data
+    tiny: float, optional (default 5e-2)
+        threshold for filtering voxels that have fallen out of the FOV
+
+    Returns
+    -------
+    Tuple (fov_mask, physical_coords), where:
+    fov_mask: 1D array_like of len voxel.shape[1]
+        mask for filtering voxels that are still in the FOV. 1 means 'OK',
+        0 means 'fell out of FOV'
+    physical_coords: array of same shape as input coords
+        transformed coords
+
+    """
+
+    physical_coords = get_physical_coords(M, coords)
+    fov_mask = np.ones(physical_coords.shape[-1]).astype('bool')
+
+    for j in range(3):
+        if not wrp[j]:
+            fov_mask = fov_mask & (physical_coords[j] >= -tiny
+                                   ) & (physical_coords[j] < dim[j] + tiny)
+
+    return fov_mask, physical_coords
+
+
+def reslice_vols(vols, target_affine=None, interp_order=3,
+                 interp_mode='constant', mask=True, wrp=None, log=None,
+                 verbose=True):
+    """
+    Uses B-spline interpolation to reslice (i.e resample) all other
+    volumes to have thesame affine header matrix as the first (0th) volume.
+
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/slice_timing.py

@@ -1,0 +1,497 @@
+"""
+:module: st
+:synopsis: module for STC (Slice-Timing Correction) in fMRI data
+:author: elvis[dot]dohmatob[at]inria[dot]fr
+
+"""
+
+import os
+import nibabel
+import scipy
+import numpy as np
+from sklearn.utils.validation import check_is_fitted
+from nilearn.image.image import check_niimg
+from nilearn._utils.compat import _basestring
+from .io_utils import is_niimg, save_vols, get_basenames
+
+
+def get_slice_indices(n_slices, slice_order='ascending',
+                      interleaved=False, return_final=False):
+    """Function computes the (unique permutation on) slice indices, consistent
+    with the specified slice order.
+
+    Parameters
+    ----------
+    n_slices: int
+        The number of slices there're altogether.
+
+    slice_order: string ('ascending', 'descending'), or array of ints or
+                 length n_slices slice order of acquisitions in a TR.
+        'ascending': slices were acquired from bottommost to topmost
+        'descending': slices were acquired from topmost to bottommost.
+        If list of integers, 0-based (i.e Python!) indexing is assumed.
+
+    interleaved: bool (optional, default False)
+        Ff set, then slices were acquired in interleaved order, odd-numbered
+        slices first, and then even-numbered slices.
+
+    Returns
+    -------
+    slice_indices: 1D array of length n_slices
+        Slice indices consistent with slice order (i.e, slice_indices[k]
+        is the corrected index of slice k according to the slice order).
+
+    Raises
+    ------
+    ValueError
+    """
+
+    if isinstance(slice_order, _basestring):
+        slice_indices = list(range(n_slices))
+        if interleaved:
+            # python indexing begins from 0 (MATLAB begins from 1)
+            slice_indices = slice_indices[0::2] + slice_indices[1::2]
+        if slice_order.lower() == 'ascending':
+            pass
+        elif slice_order.lower() == 'descending':
+            slice_indices = np.flipud(slice_indices)
+        else:
+            raise ValueError("Unknown slice order '%s'!" % slice_order)
+    else:
+        if interleaved:
+            raise ValueError(
+                ("Since you have specified an explicit slice order, I don't "
+                 "expecting you to set the 'interleaved' flag."))
+
+        # here, I'm assuming an explicitly specified slice order as a
+        # permutation on n symbols
+        slice_order = np.array(slice_order, dtype='int')
+
+        assert len(slice_order) == n_slices
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/subject_data.py

@@ -1,0 +1,855 @@
+"""
+Encapsulation of subject data. Handles subject data logic
+like filetypes, extensions, .nii.gz -> .nii conversions, setup and
+teardown for reports, etc., general sanitization, etc. Also handles
+progressive report generation for subject during preprocessing.
+
+"""
+# Author: DOHMATOB Elvis Dopgima
+
+import os
+import time
+import warnings
+import numpy as np
+from matplotlib.pyplot import cm
+from sklearn.externals.joblib import Memory
+from nilearn.image import mean_img
+from nilearn._utils.compat import _basestring
+from .io_utils import (niigz2nii as do_niigz2nii, dcm2nii as do_dcm2nii,
+                       nii2niigz as do_nii2niigz,
+                       isdicom, delete_orientation, hard_link, get_shape,
+                       is_4D, is_3D, get_basenames, is_niimg)
+from .reporting.base_reporter import (
+    commit_subject_thumnbail_to_parent_gallery,
+    ResultsGallery, Thumbnail, a, img, copy_web_conf_files,
+    ProgressReport, get_subject_report_html_template,
+    get_subject_report_preproc_html_template, copy_failed_png)
+
+from .reporting.preproc_reporter import (generate_tsdiffana_thumbnail,
+                                         generate_realignment_thumbnails,
+                                         generate_coregistration_thumbnails,
+                                         make_nipype_execution_log_html)
+
+# tooltips for thumbnails in report pages
+mc_tooltip = ("Motion parameters estimated during motion-"
+              "correction. If motion is less than half a "
+              "voxel, it's generally OK. Moreover, it's "
+              "recommended to include these estimated motion "
+              "parameters as confounds (nuissance regressors) "
+              "in the the GLM.")
+segment_acronyms = ("Acronyms: TPM means Tissue Probability Map; GM means "
+                    "Grey-Matter;"
+                    " WM means White-Matter; CSF means Cerebro-Spinal Fuild")
+reg_tooltip = ("The red contours should"
+               " match the background image well. Otherwise, something might"
+               " have gone wrong. Typically things that can go wrong include: "
+               "lesions (missing brain tissue); bad orientation headers; "
+               "non-brain tissue in anatomical image, etc. In rare cases, it "
+               "might be that the registration algorithm simply didn't "
+               "succeed.")
+segment_tooltip = ("%s. The TPM contours shoud match the background image "
+                   "well. Otherwise, something might have gone wrong. "
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/tests/_test_utils.py

@@ -1,0 +1,122 @@
+import os
+import nibabel
+import numpy as np
+from nilearn._utils.compat import _basestring
+from pyaffineprep.subject_data import SubjectData
+
+DATA_DIR = "test_tmp_data"
+if not os.path.exists(DATA_DIR):
+    os.makedirs(DATA_DIR)
+
+
+def create_random_image(shape=None,
+                        ndim=3,
+                        n_scans=None,
+                        affine=np.eye(4),
+                        parent_class=nibabel.Nifti1Image):
+    """
+    Creates a random image of prescribed shape
+
+    """
+
+    np.random.RandomState(0)
+
+    if shape is None:
+        shape = np.random.random_integers(20, size=ndim)
+
+    ndim = len(shape)
+
+    ndim = len(shape)
+    if n_scans is not None and ndim == 4:
+        shape[-1] = n_scans
+
+    return parent_class(np.random.randn(*shape), affine)
+
+
+def make_dataset(n_subjects=1, n_scans=10, n_sessions=1,
+                 threeD_filenames=False, dataset_name="test_dataset",
+                 output_dir=DATA_DIR, ext="nii.gz"):
+
+    output_dir = os.path.join(output_dir, dataset_name)
+    if not os.path.exists(output_dir):
+        os.makedirs(output_dir)
+
+    dataset = []
+    for i in range(n_subjects):
+        subject_data = {"subject_id": "sub%03i" % (i + 1), "func": [],
+                        'anat': '%s/anat.nii.gz' % DATA_DIR}
+        nibabel.save(create_random_image(ndim=3),
+                     subject_data['anat'])
+        subject_data_dir = os.path.join(output_dir, subject_data["subject_id"])
+        if not os.path.exists(subject_data_dir):
+            os.makedirs(subject_data_dir)
+
+        for j in range(n_sessions):
+            session_dir = os.path.join(subject_data_dir,
+                                       "session%03i" % (j + 1))
+            if not os.path.exists(session_dir):
+                os.makedirs(session_dir)
+            sfunc = []
+            if threeD_filenames:
+                for k in range(n_scans):
+                    func_filename = os.path.join(session_dir,
+                                                 "func%03i.%s" % (j + 1, ext))
+                    nibabel.save(create_random_image(ndim=3),
+                                 func_filename)
+                    sfunc.append(func_filename)
+                subject_data['func'].append(sfunc)
+            else:
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/tests/test_affine_transformations.py

@@ -1,0 +1,108 @@
+import numpy as np
+from nose.tools import assert_equal
+import nibabel
+from ..affine_transformations import (
+    get_initial_motion_params, spm_matrix, spm_imatrix, transform_coords,
+    apply_realignment, nibabel2spm_affine, get_physical_coords,
+    extract_realignment_params, extract_realignment_matrix)
+from ._test_utils import create_random_image
+
+
+def test_get_initial_motion_params():
+    # params for zero motion
+    p = get_initial_motion_params()
+    np.testing.assert_array_equal(p, [0, 0, 0, 0, 0, 0, 1, 1, 1, 0, 0, 0])
+
+
+def test_spm_matrix():
+    p = get_initial_motion_params()
+
+    # identity
+    np.testing.assert_array_equal(spm_matrix(p),
+                                  np.eye(4))
+
+    # induce translations
+    p[:3] += [.2, .3, .4]
+
+    M = np.array([[1., 0., 0., .2],
+                  [0., 1., 0., .3],
+                  [0., 0., 1., .4],
+                  [0., 0., 0., 1.]])
+
+    np.testing.assert_array_equal(M, spm_matrix(p))
+
+
+def test_spm_imatrix():
+    p = get_initial_motion_params()
+
+    # spm_matrix and spm_imatrix should be inverses of one another
+    np.testing.assert_array_equal(spm_imatrix(spm_matrix(p)), p)
+
+
+def test_transform_coords():
+    p = get_initial_motion_params()
+    M1 = np.eye(4)
+    M2 = np.eye(4)
+    coords = (0., 0., 1.)
+
+    # rigidly move the voxel
+    new_coords = transform_coords(p, M1, M2, coords)
+
+    # coords shouldn't change
+    assert_equal(new_coords.shape, (3, 1))
+    np.testing.assert_array_equal(new_coords.ravel(), coords)
+
+
+def test_apply_realignment_3D_niimg():
+    # create 3D niimg
+    vol = create_random_image(shape=(7, 11, 13))
+
+    # apply realignment to vol
+    apply_realignment(vol, [1, 2, 3, 4, 5, 6])
+
+
+def test_nibabel2spm_affine():
+    affine = np.eye(4)
+    np.testing.assert_array_equal(nibabel2spm_affine(affine)[:, -1],
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/tests/test_coreg.py

@@ -1,0 +1,69 @@
+import os
+import numpy as np
+from nose.tools import assert_true, nottest
+import numpy.testing
+import nibabel
+import scipy.io
+from ..coreg import compute_similarity_from_jhist, Coregister
+from ..affine_transformations import apply_realignment_to_vol
+from .test_histograms import test_joint_histogram
+
+# global setup
+THIS_FILE = os.path.abspath(__file__).split('.')[0]
+THIS_DIR = os.path.dirname(THIS_FILE)
+OUTPUT_DIR = "/tmp/%s" % os.path.basename(THIS_FILE)
+
+
+def test_compute_similarity_from_jhist():
+    jh = test_joint_histogram()
+
+    for cost_fun in ['mi', 'nmi', 'ecc']:
+        s = compute_similarity_from_jhist(jh, cost_fun=cost_fun)
+        assert_true(s <= 1)
+
+
+@nottest
+def test_coregister_on_toy_data():
+    shape = (23, 29, 31)
+    ref = nibabel.Nifti1Image(np.arange(np.prod(shape)).reshape(shape),
+                              np.eye(4)
+                              )
+
+    # rigidly move reference vol to get a new volume: the source vol
+    src = apply_realignment_to_vol(ref, [1, 1, 1,  # translations
+                                         0, .01, 0,  # rotations
+                                         ])
+
+    # learn realignment params for coregistration: src -> ref
+    c = Coregister(sep=[4, 2, 1]).fit(ref, src)
+
+    # compare estimated realigment parameters with ground-truth
+    numpy.testing.assert_almost_equal(-c.params_[4], .01, decimal=2)
+    numpy.testing.assert_array_almost_equal(-c.params_[[3, 5]],
+                                            [0, 0], decimal=2)
+    numpy.testing.assert_array_equal(np.round(-c.params_)[[0, 1, 2]],
+                                     [1., 1., 1.])
+
+
+@nottest
+def test_coregister_on_real_data():
+    # load data
+    _tmp = scipy.io.loadmat(
+        os.path.join(THIS_DIR, "test_data/some_anat.mat"),
+        squeeze_me=True, struct_as_record=False)
+    ref = nibabel.Nifti1Image(_tmp['data'], _tmp['affine'])
+
+    # rigidly move reference vol to get a new volume: the source vol
+    src = apply_realignment_to_vol(ref, [1, 2, 3,  # translations
+                                         0, .01, 0,  # rotations
+                                         ])
+
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/tests/test_histograms.py

@@ -1,0 +1,67 @@
+import numpy as np
+import nibabel
+import numpy.testing
+import nose
+import nose.tools
+from ..histograms import (_correct_voxel_samp, make_sampled_grid,
+                          trilinear_interp, joint_histogram)
+
+
+def test_correct_voxel_samp():
+    numpy.testing.assert_array_equal(
+        _correct_voxel_samp(np.eye(4), 2), [2., 2., 2.])
+
+    numpy.testing.assert_array_equal(
+        _correct_voxel_samp(np.eye(4), [3, 2, 1]), [3., 2., 1.])
+
+    numpy.testing.assert_array_equal(_correct_voxel_samp(
+            np.array([[-1., 0., 0., 128.],
+                      [0., 1., 0., -168.],
+                      [0., 0., 3., -75.],
+                      [0., 0., 0., 1.]]),
+            4), [4., 4., 4. / 3])
+
+
+def test_make_sampled_grid_without_spm_magic():
+    for samp in [1., [1.], [1.] * 3]:
+        numpy.testing.assert_array_equal(make_sampled_grid([3, 5, 7],
+                                                           samp=samp,
+                                                           magic=False),
+                                         np.array(
+                [[1., 1., 1., 1., 1., 1., 1., 1., 1., 1., 1., 1., 1., 1., 1.],
+                 [1., 1., 1., 1., 1., 2., 2., 2., 2., 2., 3., 3., 3., 3., 3.],
+                 [1., 2., 3., 4., 5., 1., 2., 3., 4., 5., 1., 2., 3., 4., 5.]
+                 ]))
+
+
+def test_trilinear_interp():
+    shape = (23, 29, 31)
+    f = np.arange(np.prod(shape))
+
+    nose.tools.assert_equal(trilinear_interp(f, shape, 1, 1, 1), 0.)
+    nose.tools.assert_equal(trilinear_interp(f, shape, 2, 1, 1), 1.)
+    nose.tools.assert_equal(trilinear_interp(f, shape, 1 + shape[0], 1, 1),
+                            shape[0])
+    nose.tools.assert_true(0. < trilinear_interp(f, shape, 1.5, 1., 1.) < 1.)
+
+
+def test_joint_histogram():
+    ref_shape = (23, 29, 61)
+    src_shape = (13, 51, 19)
+    ref = np.arange(np.prod(ref_shape)).reshape(ref_shape)
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/tests/test_io_utils.py

@@ -1,0 +1,517 @@
+import os
+import tempfile
+import inspect
+import numpy as np
+from nose.tools import assert_equal, assert_true, assert_false
+import nibabel
+from nilearn.image.image import check_niimg_4d
+from nilearn._utils.compat import _basestring
+from numpy.testing import assert_array_equal
+
+from pyaffineprep.io_utils import delete_orientation
+
+from ..io_utils import (
+    do_3Dto4D_merge, load_vols, save_vols, save_vol, hard_link, nii2niigz,
+    get_basename, get_basenames, is_niimg, is_4D, is_3D, get_vox_dims,
+    niigz2nii, _expand_path, isdicom, get_shape, get_relative_path,
+    loaduint8)
+
+# global setup
+this_file = os.path.basename(os.path.abspath(__file__)).split('.')[0]
+OUTPUT_DIR = "/tmp/%s" % this_file
+IMAGE_EXTENSIONS = [".nii", ".nii.gz", ".img"]
+
+
+def create_random_image(shape=None, ndim=3, n_scans=None, affine=np.eye(4),
+                        parent_class=nibabel.Nifti1Image):
+    """
+    Creates a random image of prescribed shape
+
+    """
+    if n_scans is not None:
+        ndim = 4
+    if shape is None:
+        shape = np.random.random_integers(20, size=ndim)
+    ndim = len(shape)
+    if n_scans is not None and ndim == 4:
+        shape[-1] = n_scans
+    return parent_class(np.random.randn(*shape), affine)
+
+
+def test_save_vol():
+    output_dir = os.path.join(OUTPUT_DIR, inspect.stack()[0][3])
+    if not os.path.exists(output_dir):
+        os.makedirs(output_dir)
+    vol = create_random_image(ndim=3)
+    output_filename = save_vol(vol, output_dir=output_dir,
+                               basename='123.nii.gz')
+    assert_equal(os.path.basename(output_filename),
+                 '123.nii.gz')
+    output_filename = save_vol(vol, output_dir=output_dir, basename='123',
+                               prefix='s', ext=".img")
+    assert_equal(os.path.basename(output_filename),
+                 's123.img')
+
+
+def test_save_vols():
+    # setup
+    n_scans = 10
+    output_dir = os.path.join(OUTPUT_DIR, inspect.stack()[0][3])
+    if not os.path.exists(output_dir):
+        os.makedirs(output_dir)
+
+    # create 4D film
+    film = create_random_image(ndim=4, n_scans=n_scans)
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/tests/test_kernel_smooth.py

@@ -1,0 +1,68 @@
+import numpy as np
+import nibabel
+from nose.tools import assert_equal
+import numpy.testing
+from ..kernel_smooth import fwhm2sigma, sigma2fwhm, smooth_image
+
+
+def create_random_image(shape=None,
+                        ndim=3,
+                        n_scans=None,
+                        affine=np.eye(4),
+                        parent_class=nibabel.Nifti1Image):
+    """
+    Creates a random image of prescribed shape
+
+    """
+    rng = np.random.RandomState(0)
+    if shape is None:
+        shape = rng.random_integers(20, size=ndim)
+    ndim = len(shape)
+    ndim = len(shape)
+    if n_scans is not None and ndim == 4:
+        shape[-1] = n_scans
+    return parent_class(np.random.randn(*shape), affine)
+
+
+def test_fwhm2sigma():
+    fwhm = [1, 2, 3]
+    for _fwhm in fwhm:
+        numpy.testing.assert_array_equal(
+            fwhm2sigma(_fwhm), np.array(_fwhm) / np.sqrt(8. * np.log(2)))
+    for j in range(3):
+        _fwhm = fwhm[j:]
+        numpy.testing.assert_array_equal(
+            fwhm2sigma(_fwhm), np.array(_fwhm) / np.sqrt(8. * np.log(2)))
+
+
+def test_sigma2sigma():
+    sigma = [7, 2, 3]
+    for _sigma in sigma:
+        numpy.testing.assert_array_equal(sigma2fwhm(_sigma),
+                                         np.array(
+                                             _sigma) * np.sqrt(8. * np.log(2)))
+    for j in range(3):
+        _sigma = sigma[j:]
+        numpy.testing.assert_array_equal(sigma2fwhm(_sigma),
+                                         np.array(
+                                             _sigma) * np.sqrt(8. * np.log(2)))
+
+
+def test_fwhm2sigma_and_sigma2fwhm_are_inverses():
+    toto = [5, 7, 11.]
+    numpy.testing.assert_array_equal(toto, sigma2fwhm(fwhm2sigma(toto)))
+    numpy.testing.assert_array_almost_equal(toto, fwhm2sigma(sigma2fwhm(toto)))
+
+
+def test_smooth_image_for_3D_vol():
+    vol = create_random_image()
+    svol = smooth_image(vol, [5, 7, 11.])
+    assert_equal(svol.shape, vol.shape)
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/tests/test_realign.py

@@ -1,0 +1,228 @@
+import os
+import inspect
+import numpy as np
+import scipy.io
+from nose.tools import assert_true, assert_equal, nottest
+import nibabel
+from nilearn.image import index_img
+from ..realign import _compute_rate_of_change_of_chisq, MRIMotionCorrection
+from ..affine_transformations import (
+    apply_realignment, extract_realignment_params, get_initial_motion_params)
+from ._test_utils import create_random_image
+
+# global setup
+THIS_FILE = os.path.abspath(__file__).split('.')[0]
+THIS_DIR = os.path.dirname(THIS_FILE)
+OUTPUT_DIR = "/tmp/%s" % os.path.basename(THIS_FILE)
+
+
+def _make_vol_specific_translation(translation, n_scans, t):
+    """
+    translation: constant part plus vol-dependent part
+
+    """
+
+    return (t > 0) * (translation + 1. * t / n_scans)
+
+
+def _make_vol_specific_rotation(rotation, n_scans, t):
+    """
+    rotation: constant part plus vol-dependent part
+
+    """
+
+    return (t > 0) * (rotation + 1. * t / n_scans
+                      ) * np.pi / 180.
+
+
+def test_compute_rate_of_change_of_chisq():
+    # XXX please document strange variables !!!
+    # setup
+    true_A = np.array(
+        [[-0., -0., -0., -0., -0., -0.],
+         [1., -0., -0., -0., 1.000001, 1.000001],
+         [1., -0., -0., -0., 1.0000015, 1.0000015],
+         [-2., 1., -0., 1.000001, -2.000001, -5.],
+         [1., -0., -0., -0., 1.000001, 2.000001],
+         [1., -0., -0., -0., 1.0000015, 2.0000015],
+         [-2., 1., -0., 1.0000015, -2.000001, -6.9999995],
+         [1., -0., -0., -0., 1.000001, 3.000001],
+         [1., -0., -0., -0., 1.0000015, 3.0000015],
+         [-2., 1., -0., 1.000002, -2.000001, -8.999999],
+         [1., -0., -0., -0., 1.000001, 4.000001],
+         [1., -0., -0., -0., 1.0000015, 4.0000015],
+         [-2., -3., 1., -7.0000005, -5., 0.9999975],
+         [1., -0., -0., -0., 2.000001, 1.000001],
+         [1., -0., -0., -0., 2.0000015, 1.0000015],
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/tests/test_reslice.py

@@ -1,0 +1,44 @@
+import numpy as np
+from nose.tools import assert_false
+import nibabel
+from ..reslice import reslice_vols
+from ..affine_transformations import (
+    get_initial_motion_params, apply_realignment)
+
+
+def test_reslice_vols():
+    # create basic L pattern
+    n_scans = 3
+    film = np.zeros((10, 10, 10, n_scans))
+    film[-3:, 5:-1, :, :] = 1
+    film[:, 2:5, :, :] = 1
+    affine = np.array(
+        [[-2.99256921e+00, -1.12436414e-01, -2.23214120e-01,
+          1.01544670e+02],
+         [-5.69147766e-02, 2.87465930e+00, -1.07026458e+00,
+          -8.77408752e+01],
+         [-2.03200281e-01, 8.50703299e-01, 3.58708930e+00,
+          -7.10269012e+01],
+         [0.00000000e+00, 0.00000000e+00, 0.00000000e+00,
+          1.00000000e+00]])
+    film = nibabel.Nifti1Image(film, affine)
+
+    # rigidly move other volumes w.r.t. the first
+    rp = np.array([get_initial_motion_params()
+                   for _ in range(n_scans)])
+    for t in range(film.shape[-1]):
+        rp[t, ...][:3] += t / n_scans
+        rp[t, ...][3:6] += np.pi * t
+
+    film = apply_realignment(film, rp)
+
+    # affines are not the same
+    assert_false(np.all(
+        film[1].get_affine() == film[0].get_affine()))
+
+    # reslice vols
+    film = list(reslice_vols(film))
+
+    # affines are now the same
+    np.testing.assert_array_equal(film[1].get_affine(),
+                                  film[0].get_affine())

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/tests/test_slice_timing.py

@@ -1,0 +1,268 @@
+import os
+import inspect
+import numpy as np
+import nibabel
+from nose.tools import assert_equal, assert_true, raises
+from ..slice_timing import STC, fMRISTC, get_slice_indices
+from ..io_utils import save_vols
+
+# global setup
+this_file = os.path.basename(os.path.abspath(__file__)).split('.')[0]
+OUTPUT_DIR = "/tmp/%s" % this_file
+
+
+def test_get_slice_indices_ascending():
+    np.testing.assert_array_equal(
+        get_slice_indices(5, slice_order="ascending", return_final=True),
+        [0, 1, 2, 3, 4])
+
+
+def test_get_slice_indices_ascending_interleaved():
+    np.testing.assert_array_equal(
+        get_slice_indices(5, slice_order="ascending", interleaved=True,
+                          return_final=True), [0, 3, 1, 4, 2])
+
+
+def test_get_slice_indices_descending():
+    # descending
+    np.testing.assert_array_equal(
+        get_slice_indices(5, slice_order="descending", return_final=True),
+        [4, 3, 2, 1, 0])
+
+
+def test_get_slice_indices_descending_interleaved():
+    # descending and interleaved
+    np.testing.assert_array_equal(
+        get_slice_indices(5, slice_order="descending", interleaved=True,
+                          return_final=True), [4, 1, 3, 0, 2])
+
+
+def test_get_slice_indices_explicit():
+    slice_order = [1, 4, 3, 2, 0]
+    np.testing.assert_array_equal(
+        get_slice_indices(5, slice_order=slice_order, return_final=True),
+        [4, 0, 3, 2, 1])
+
+
+@raises(ValueError)
+def test_get_slice_indices_explicit_interleaved():
+    slice_order = [1, 4, 3, 2, 0]
+    np.testing.assert_array_equal(
+        get_slice_indices(5, slice_order=slice_order,
+                          interleaved=True), [2, 0, 4, 1, 3])
+
+
+def test_STC_constructor():
+    stc = STC()
+    assert_equal(stc.ref_slice, 0)
+    assert_equal(stc.interleaved, False)
+    assert_true(stc.verbose == 1)
+
+
+def test_fMRISTC_constructor():
+    fmristc = fMRISTC()
+    assert_equal(fmristc.ref_slice, 0)
+    assert_equal(fmristc.interleaved, False)
+    assert_true(fmristc.verbose == 1)
+
+
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/tests/test_tsdiffana.py

@@ -1,0 +1,152 @@
+"""
+
+"""
+
+import numpy as np
+import nibabel
+from numpy.testing import assert_array_almost_equal
+from ..time_diff import (time_slice_diffs, multi_session_time_slice_diffs,
+                         plot_tsdiffs)
+from ._test_utils import _make_sd
+from nose.tools import assert_true, assert_false
+
+
+def make_test_data(n_scans=3):
+    shape = (7, 8, 9, n_scans)
+    film = np.zeros(shape)
+    film[-3:, 5:-1, :, :] = 1
+    film[:, 2:5, :, :] = 1
+    scal = np.sum(film) * 1. / film.size
+    affine = np.eye(4)
+    return nibabel.Nifti1Image(film, affine), scal
+
+
+def test_ts_diff_ana_null():
+    """ Run ts_diff_ana on constant image sequence """
+    # create basic L pattern
+    n_scans = 2
+    film, scal = make_test_data(n_scans=n_scans)
+    film.to_filename('/tmp/plop.nii')
+    shape = film.shape
+    report = time_slice_diffs(film)
+
+    assert_array_almost_equal(report['volume_means'],
+                              scal * np.ones(n_scans))
+    assert_array_almost_equal(report['volume_mean_diff2'],
+                              np.zeros(n_scans - 1))
+    assert_array_almost_equal(report['slice_mean_diff2'],
+                              np.zeros((n_scans - 1, shape[2])))
+    assert_array_almost_equal(report['diff2_mean_vol'].get_data(),
+                              np.zeros(shape[:3]))
+    assert_array_almost_equal(report['slice_diff2_max_vol'].get_data(),
+                              np.zeros(shape[:3]))
+
+
+def test_ts_diff_ana():
+    """ Run ts_diff_ana on changing image sequence """
+    # create basic L pattern
+    n_scans = 2
+    shape = (7, 8, 9, n_scans)
+    film = np.zeros(shape)
+    film[-3:, 5:-1, :, :] = 1
+    film[:, 2:5, :, :] = 1
+    ref = film.copy()
+    scal = np.sum(ref) * 1. / ref.size
+    film = np.dot(film, np.diag(np.arange(n_scans)))
+
+    affine = np.eye(4)
+    film = nibabel.Nifti1Image(film, affine)
+
+    report = time_slice_diffs(film)
+
+    assert_array_almost_equal(report['volume_means'],
+                              scal * np.arange(n_scans))
[…]

### 2020-08-03 code/fmri_experiment/externals/pypreprocess/time_diff.py

@@ -1,0 +1,281 @@
+# emacs: -*- mode: python; py-indent-offset: 4; indent-tabs-mode: nil -*-
+# vi: set ft=python sts=4 ts=4 sw=4 et:
+''' Time series diagnostics
+
+These started life as ``tsdiffana.m`` - see
+http://imaging.mrc-cbu.cam.ac.uk/imaging/DataDiagnostics
+
+Oliver Josephs (FIL) gave Matthew Brett the idea of time-point to time-point
+subtraction as a diagnostic for motion and other sudden image changes.
+This has been implemented in the Nipy package.
+
+We give here a simpler implementation with modified dependences
+
+'''
+import numpy as np
+import nibabel as nib
+from nilearn.plotting import plot_stat_map
+from nilearn.image.image import check_niimg_4d
+from nilearn.image import mean_img, reorder_img
+
+
+def multi_session_time_slice_diffs(img_list):
+    """ time slice difference on several 4D images
+
+    Parameters
+    ----------
+    img_list: list of 4D Niimg-like
+        Input multi-session images
+
+    returns
+    -------
+    results : dict
+        see time_slice_diffs docstring for details.
+
+    note
+    ----
+    The results are accumulated across sessions
+    """
+    results = {}
+    for i, img in enumerate(img_list):
+        results_ = time_slice_diffs(img)
+        if i == 0:
+            for key, val in results_.items():
+                # special case for 'session_length' to make
+                # aggregation easier later on
+                results[key] = val if key != 'session_length' else [val]
+        else:
+            results['volume_mean_diff2'] = np.hstack((
+                    results['volume_mean_diff2'],
+                    results_['volume_mean_diff2']))
+            results['slice_mean_diff2'] = np.vstack((
+                    results['slice_mean_diff2'],
+                    results_['slice_mean_diff2']))
+            results['volume_means'] = np.hstack((
+                    results['volume_means'],
+                    results_['volume_means']))
+            results['diff2_mean_vol'] = mean_img(
+                [results['diff2_mean_vol'], results_['diff2_mean_vol']])
+            results['slice_diff2_max_vol'] = nib.Nifti1Image(
+                np.maximum(results_['slice_diff2_max_vol'].get_data(),
[…]

### 2020-08-03 code/fmri_experiment/paths.py

@@ -1,0 +1,26 @@
+import os
+
+# Setup paths and file names
+this_path = '/' + os.path.join(*os.path.realpath(__file__).split('/')[:-1])
+
+
+def setup_datapath():
+    fname_datapath = '/' + this_path + '/data_path.txt'
+
+    if not os.path.isfile(fname_datapath):
+        data_path = input("Enter data path (or create a data_path.txt):")
+        if data_path[-1] != '/':
+            data_path += '/'
+        with open(fname_datapath, 'w') as f:
+            f.write(data_path)
+    with open(fname_datapath, 'r') as f:
+        data_path = f.read()
+    return data_path
+
+
+data_path = setup_datapath()
+preproc_path = data_path[:-1] + '_preproc/'
+
+# Where preprocessed fMRI files will be stored
+if not os.path.isdir(preproc_path):
+    os.makedirs(preproc_path)

### 2020-08-03 code/fmri_experiment/preproc.py

@@ -1,0 +1,225 @@
+# Author: Jean-Remi King
+#
+# Adapted from MNE-Python, Dipy and Pypreprocess (BSL Licence)
+
+from copy import deepcopy
+
+import nibabel as nib
+import numpy as np
+import pandas as pd
+from dipy.align import imaffine, imwarp, metrics, transforms, vector_fields
+from dipy.align.reslice import reslice
+from h5io import read_hdf5, write_hdf5
+from nilearn import image
+
+from externals.pypreprocess.coreg import Coregister
+from externals.pypreprocess.realign import MRIMotionCorrection
+from externals.pypreprocess.slice_timing import fMRISTC
+
+
+def do_preproc_fmri(anat, func, fsaverage):
+    params = dict()
+    stages = dict()
+
+    # Slice-Timing Correction
+    stc = fMRISTC(verbose=False)
+    func = stc.fit(raw_data=func).transform(func)
+    params['slice_timing'] = stc.kernel_
+
+    stages['stc'] = deepcopy(func)
+
+    # Motion correction
+    mc = MRIMotionCorrection(n_sessions=1, verbose=False)
+    mc_output = mc.fit([func, ]).transform(concat=True)
+    func = mc_output['realigned_images'][0]
+    params['motion'] = mc_output['realignment_parameters'][0]
+
+    stages['mc'] = deepcopy(func)
+
+    # Coregister to anatomy
+    coreg = Coregister(verbose=False)
+    coreg = coreg.fit(target=anat, source=func)
+    func = coreg.transform(func)
+    params['coreg'] = coreg.params_
+
+    stages['coreg'] = deepcopy(func)
+
+    # Concatenate
+    func = image.concat_imgs(func)
+
+    # Fit Morphing to FSaverage
+    morpher = SDR(zooms=3.5,
+                  niter_affine=(100, 50, 10),
+                  niter_sdr=(100, 50, 10))
+    morpher.fit(nib.load(anat), nib.load(fsaverage))
+    attrs = ('affine_idx_in_', 'comp_', 'backward_',
+             'mri_from_affine_', 'mri_to_affine_', 'shape_',
+             'zooms', 'niter_affine', 'niter_sdr')
+    params['morph'] = dict((k, getattr(morpher, k)) for k in attrs)
+
+    return func, stages, params
+
+
+class SDR():
+    "Symmetric Diffeomorphic Registration"
+    def __init__(self, zooms=2.,
[…]

### 2020-08-03 paper/appendix.tex

@@
 For the Backward Model, feature importance cannot be assessed because there is
-no prediction.
+no prediction combining multiple factors.
 
@@
 $\mu$, whose proportion arbitrarily depends on the necessity to favor type I and
-type II errors. In practice, we cannot use this procedure for our MEG study,
+type II errors. In practice, we cannot use this procedure for our fMRI and MEG experiments,
 because signal-to-noise ratio is unknown.
@@
 
-Following the recommendations of one of our reviewers, we also implemented a
+Following the recommendations of one of our reviewers, we implemented a
 multivariate variant of the forward model, i.e. a MANOVA, using the statsmodels
@@
 Consequently, we did not find a way to compare MANOVA against B2B with the
-$\Delta R$ evaluation. However, the effects of MANOVA comes with its own effect
-size statistics: the Wilk's Lambda. For each searchlight, we thus use the $F-values$ of the
+$\Delta R$ evaluation. However, the effects of MANOVA are generally summarized
+with the Wilk's Lambda stastistics or its transformation into an $F$-value.
+For each searchlight, we thus use the $F$-values of the
 Wilk's Lambda statistics as proxy for $\hat S$ and feeds it to a second-level
@@
 The results show that all factors, including the Dummy variable, are
-systematically above chance level in all recorded brain regions \ref{fig:fmri_supp}. This result
+systematically above chance level in all recorded brain regions (Fig.~\ref{fig:fmri_supp}). This result
 can be explained by the large dimensionality of $Y$. Indeed, limiting the
@@
   \end{center}
-  \caption{Top rows. Pearson $R$ correlation obtained for a Backward decoding model and
+  \caption{Top two rows. Pearson $R$ correlation obtained for a Backward decoding model and
   a Forward encoding model trained with one factor at a time. These models can
   not take into account the factor covariance, and thus lead to spurious effects
-  (e.g. visual cortex effect for the dummy variable). Bottom rows. Second-
-  level p-values across subjects for a Forward, a B2B and a Manova trained with
-  all factors. The B2B achieves better p-values, without leading to spurious effects for
+  (e.g. visual cortex effect for the dummy variable). Bottom five rows. Second-
+  level p-values across subjects for the coefficients of the Forward, a B2B and a Manova trained with
+  all factors, as well as for the $\Delta R$ of the B2B, for comparison.
+  B2B achieves better p-values, without leading to spurious effects for
   the Dummy variable. The Manova leads to biased estimates due to its inability
-  to deal with overfitting. The last row shows where the Manova F-values of each factor differs
-  significantly from the F-values of the Dummy variable.}
+  to deal with overfitting. The last row shows where the Manova's $F$-values of each factor differs
+  significantly from the $F$-values estimated for the Dummy variable.}
 \end{figure}
@@
 ten features. These additional features corresponds to the first dimensions
-of word embedding as provided by Spacy \citep{spacy2}. The results shown
-in Figure \ref{fig:embeddings}, show that the feature importance of ad-hoc
+of word embedding as provided by Spacy \citep{spacy2}. The MEG results shown
+in Fig.~\ref{fig:embeddings}, show that the feature importance of ad-hoc
 features as derived by B2B remain unchanged and are actually improved.

### 2020-08-03 paper/fmri_exps.tex

@@
 Next, we apply our method to brain imaging data from the anonymized multimodal
-neuroimaging ``Mother Of all Unification Studies'' (MOUS) dataset
+neuroimaging ``Mother Of all Unification Studies'' (MOUS) study
 \citep{schoffelen2019204}. The dataset contains functional Magnetic Resonance
 Imaging (fMRI) and magneto-encephalography (MEG)
-recordings of each 102 healthy Dutch adults who performed a reading
-task in the scanner. Thirteen subjects were excluded from the analysis (12 MEG and 1 fMRI)
+recordings of 102 healthy Dutch adults who performed a reading
+task in the scanner. Ten subjects were excluded from the analysis (9/102 MEG and 1/102 fMRI)
 because of technical difficulties reading the files.
@@
 Many internal operations of \emph{fMRIPrep} use \emph{Nilearn} 0.6.2
-\citep[RRID:SCR\_001362]{nilearn}, mostly within the functional
+\citep[RRID:SCR\_001362]{Nilearn}, mostly within the functional
 processing workflow. For more details of the pipeline, see
@@
 The preprocessed volumetric fMRI data was linearly projected to its closest surface
-using nilearn \texttt{vol\_to\_surf} function with a 3 mm radius. A surface searchlight
+using Nilearn \texttt{vol\_to\_surf} function with a 3 mm radius and the
+Freesurfer 'fsaverage5' surface. A surface searchlight
 analysis was then implemented by concatenating the surface
@@
 thus build an observation matrix $Y \in \mathbb{R}^{m \times d_y}$ of $m\approx$ 1400 words
-by $d_y\approx40$ vertices per searchlight sphere. Each
-of the columns of $Y$ is normalized to have zero mean and unit variance.
+by $d_y\approx40$ vertices per searchlight sphere. Each of the columns of $Y$ is normalized to have zero mean and unit variance.
 
@@
 %
-First, 'Word Length' refers to the total number of letters.
-Word Length is expected to primarily cause a variation in the early evoked MEG responses (i.e.
-from 100 ms after stimulus onset) elicited by the retinotopically-tuned visual cortices (e.g.
-\citep{pegado2014timing}.).
+First, 'Word Length' refers to the total number of letters. Word Length is
+expected to primarily cause a variation in the early evoked MEG responses
+(i.e. from 100 ms after stimulus onset) primarily elicited by the
+retinotopically-tuned visual cortices (e.g. \citep{pegado2014timing}.) but
+expected across the full visual hierarchy (e.g. \citep{hagler2006spatial}).
 %
 Second, 'Word Frequency' indexes how frequently each word appears in Dutch and
-was derived with the the Zipf logarithmic scale of \citep{van2014subtlex}
+here derives from the Zipf logarithmic scale of \citep{van2014subtlex}
 provided by the WordFreq package \citep{speerwordfreq}. Word Frequency is
 expected to primarily cause a variation in the late evoked MEG responses (i.e.
-from 400 ms) in the left frontal, temporal and parietal cortices
+from 400 ms) in the left frontal, temporal and parietal
 cortices \citep{kutas2011thirty,mccandliss2003visual}.
@@
 noun, a verb, an adjective or an adverb) or a function word (i.e. a preposition,
-a conjunction, a determinant, a pronoun or a numeral), and was derived from
+a conjunction, a determinant, a pronoun or a numeral), and here derives from
 Spacy's part of speech tagger \citep{spacy2}. To our knowledge, this feature has
@@
 
-To account for the delay of blood oxygenation level dependent responses, the
+To account for the delay of blood oxygenation level dependent responses, these four
 features were convolved using the Glover hemodynamic response function
-of nilearn's \texttt{compute\_regressor} function an oversampling of 16 and default parameters \citep{nilearn}.
+of Nilearn's \texttt{compute\_regressor} function an oversampling of 16 and default parameters \citep{Nilearn}.
 
 This procedure yields an $X \in \mathbb{R}^{m \times d_x}$ matrix of $m\approx$
-840 TR (Repetition Time: 2 sec.) by
-$d_x=4$ features for each subject. Each of the columns of $X$ is normalized to
-have a zero-mean and a unit variance.
+840 TR (Repetition Time: 2 sec.) by $d_x=4$ factors for each subject.
+Each of the columns of $X$ is normalized to have a zero-mean and a unit variance.
 
@@
 regression, Regularized CCA and PLS, as implemented in scikit-learn
-\citep{sklearn} and pyrcca \citep{bilenko2016pyrcca}, and optimized with nested cross-validation over twenty $l2$
+\citep{sklearn} and Pyrcca \citep{bilenko2016Pyrcca}, and optimized with nested cross-validation over twenty $l2$
 regularization parameters logarithmically spaced between $10^{-4}$ and $10^4$
@@
 
-We used the feature importance described in Algorithm \ref{algorithm:b2b_fi} to assess the extent to which each feature $X_i$ specifically improves the prediction of held-out $Y$ data, using a 5-fold cross-validation (with shuffled trials to homogeneize the distributions between the training and testing splits).
+We used the feature importance described in Algorithm \ref{algorithm:b2b_fi} to assess the extent to which each feature $X_i$ specifically improves the prediction of held-out $Y$ data, using a 5-fold cross-validation (with shuffled trials to homogenize the distributions between the training and testing splits).
 
@@
 Pairwise comparison between models were performed using a two-sided Wilcoxon
-signed-rank test across subjects (m=90) using the average $\Delta R$ across time.
-Corresponding effect sizes are shown in Figure~\ref{fig:meg_results}, and
+signed-rank test across subjects using the average $\Delta R$ across time.
+Corresponding effect sizes are shown in Fig.~\ref{fig:meg_results}, and
 p-values are reported below.
@@
 
-Supplementary Figure \ref{fig:fmri_supp} shows that the Backward model decodes
+Supplementary Fig.~\ref{fig:fmri_supp} shows that the Backward model decodes
 the dummy variable well above chance level. In addition, it decodes both Word
@@
 these features are known to primarily influence early visual cortex and
-associative cortices respectively. These results illustrate that while Backward
-modeling achieve high sensitivity, it is not valid to estimate the specific
+associative cortices respectively. In consequence, while Backward
+modeling achieves high sensitivity, it is not valid to estimate the specific
 causal contribution of collinear features.
@@
 coefficients of the Forward model reveals the expected spatial specificity of
-Word Length and Word Frequency, peaking the visual and temporal cortices
-respectively. However such Forward coefficients may be variable across
+Word Length and Word Frequency, peaking in the visual and temporal cortices
+respectively. However, such Forward coefficients may be variable across
 subjects, and could thus underestimate the significance of each factor at each
@@
 inter-subject variability could be less impactful with
-multivariate observations methods such as CCA and PLS \citep{bilenko2016pyrcca,
+multivariate observations methods such as CCA and PLS \citep{bilenko2016Pyrcca,
 king2018encoding}, but they fail to provide a single and interpretable coefficient
@@
 regularized) reaches higher levels of significance at each vertex
-than the Forward $H$ \ref{fig:fmri_supp}. (Note that a direct
+than the Forward $H$ (Supplementary Fig.~\ref{fig:fmri_supp}). (Note that a direct
 comparison between the Forward coefficients $H$ and those of B2B ($\hat S$)
-is not feasible, as these two do not have the same units.)
+is challenging, because these two metrics do not have the same units.)
 
 To fairly compare the Forward, CCA, PLS and B2B methods on a common and robust
-evaluation metric, robust to coefficient biases, we thus compare them on their ability to improve $Y$
+evaluation metrics, robust to coefficient biases, we thus compare them on their ability to improve $Y$
 predictions of out-of-sample data when one factor is introduced into the model.
-This analysis leads to a $\Delta R$ as described in Algorithm
-\ref{algorithm:b2b_fi}). As expected, the Forward, CCA, PLS
+This analysis leads to a $\Delta R$ as described in the synthetic experiment.
+As expected, the Forward, CCA, PLS
 and B2B method predicted that the Dummy Variable does not improve the $Y$
@@
 The average $\Delta R$ across subjects is displayed for each model and each
-feature in \ref{fig:fmri_delta_r}. Overall, B2B favorably compares to baseline models.
+feature in~\ref{fig:fmri_delta_r}. Overall, B2B favorably compares to baseline models.
 To quantify this assessment, we compare for each subject separately
@@
 for by the Forward model ($p<.001$, not corrected across vertices)
-\ref{fig:fmri_delta_r}. The results show that B2B significantly outperforms all
-baseline models on all factors (all $p<0.0001$) but the dummy variable
-(min $p=0.0852$).
+The results, displayed in Fig.~\ref{fig:fmri_delta_r}, show that B2B
+is significantly better than baseline models on all factors (all $p<0.0001$)
+but the dummy variable (min $p=0.0852$).

### 2020-08-03 paper/meg_exps.tex

@@
 
-The raw MEG data was bandpass-filtered between 0.1 and 40Hz using MNE-Python
+One hundred and two subjects performed a similar reading task to the one
+described above in the MEG scanner. The raw MEG data was bandpass-filtered between 0.1 and 40Hz using MNE-Python
 default parameters \citep{gramfort2013meg, gramfort2014mne}. Specifically, we
@@
 thus led a sampling frequency of 240Hz. The average responses across words is
-displayed in Figure \ref{fig:meg_evoked}.
+displayed in Fig.~\ref{fig:meg_evoked}.
 For each subject and each time sample relative to word onset, we
@@
 We compare B2B against other methods following the same procedure as for the
-fMRI experiments, except that the search light is applied across time samples.
+fMRI experiments, except that the searchlight swipes across time samples, and
+is trained across all MEG channels.
 
@@
 
-Figure \ref{fig:meg_results} shows, for each model, the effects obtained across
+Figure~\ref{fig:meg_results} shows, for each model, the effects obtained across
 time (left) and subjects (right).
@@
 
-To assess which model would be most sensitive to these causal discoveries, we
-compared B2B to other models across subjects (Figure \ref{fig:meg_results}
-right). For Word Length, B2B outperforms all models (all $p < 0.00001$) but CCA
-($p=0.0678$). For Word Frequency, B2B outperforms all models (all $p < 0.0006$).
-For "Word Function", B2B outperforms all models (all $p < 0.0015$). Overall,
-these results show that B2B favorably compares against baseline models.
+We compare B2B to other models across subjects (Fig.~\ref{fig:meg_results}
+right). For both Word Length and Word Frequency, B2B outperforms all models (all $p < 0.0001$).
+For "Word Function", B2B outperforms all models (all $p < 0.0001$) but CCA ($p<0.0001$).
+Overall, these results show that B2B generally compares favorably against baseline models.

### 2020-08-03 paper/paper.tex

@@
 other standard techniques. We discuss how the speed and generality of B2B
-sets the foundations of an omnibus test for disentanglement of covarying
+sets the foundations of an omnibus test to disentangle the causal contributions of covarying
 factors in multidimensional observations.

### 2020-08-03 code/public/main.py
public

@@ -54,14 +54,19 @@ def rn_score(X, Y, scoring='r', multioutput='uniform_average'):
 class B2B(BaseEstimator, RegressorMixin):
     def __init__(self, alphas=alphas,
                  independent_alphas=True, ensemble=None,
+                 G='ridge_cv',
+                 H='ridge_cv',
                  scoring='r'):
         self.alphas = alphas
         self.independent_alphas = independent_alphas
         self.ensemble = ensemble
         self.scoring = scoring
         self.__name__ = 'B2B'
+        self.G = G
+        self.H = H
 
     def fit(self, X, Y):
+        import copy
 
         self.G_ = list()
         self.H_ = list()
@@ -80,24 +85,40 @@ class B2B(BaseEstimator, RegressorMixin):
         for train, test in ensemble:
 
             # Fit decoder
-            G, G_alpha, YG = ridge_cv(Y[train], X[train],
-                                      self.alphas,
-                                      self.independent_alphas)
+            if self.G == 'ridge_cv':
+                G, G_alpha, YG = ridge_cv(Y[train], X[train],
+                                          self.alphas,
+                                          self.independent_alphas)
+                if len(X[train]) != len(X):
+                    YG = Y @ G.T
+            else:
+                G = copy.deepcopy(self.G).fit(Y[train], X[train])
+                YG = np.empty((len(Y), X.shape[1]))
+                YG[test] = G.predict(Y[test])
+
             self.G_.append(G)
 
-            if len(X[train]) != len(X):
-                YG = Y @ G.T
-
             # Fit encoder
-            H, H_alpha, _ = ridge_cv(X[test], YG[test],
-                                     self.alphas,
-                                     self.independent_alphas)
+            if self.H == 'ridge_cv':
+                H, H_alpha, _ = ridge_cv(X[test], YG[test],
+                                         self.alphas,
+                                         self.independent_alphas)
+            else:
+                H = copy.deepcopy(self.H).fit(X[test], YG[test])
             self.H_.append(H)
 
         # Aggregate ensembling
-        self.G_ = np.mean(self.G_, 0)
-        self.H_ = np.mean(self.H_, 0)
-        self.E_ = np.diag(self.H_)
+        if self.G == 'ridge_cv':
[…]
